### 2019-04-22 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Start on LaTeX manuscript
Add latex file based on MDPI template and MDPI latex class definitions.
See https://www.mdpi.com/authors/latex

%  LaTeX support: latex@mdpi.com 
%  In case you need support, please attach all files that are necessary for compiling as well as the log file, and specify the details of your LaTeX setup (which operating system and LaTeX version / tools you are using).

%  =================================================================
\documentclass[fire,article,submit,moreauthors,pdftex]{Definitions/mdpi}

%\usepackage[utf8]{inputenc}

% If you would like to post an early version of this manuscript as a preprint, you may use preprint as the journal and change 'submit' to 'accept'. The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

%--------------------
% Class Options:
%--------------------
%----------
% journal
%----------
% Choose between the following MDPI journals:
% acoustics, actuators, addictions, admsci, aerospace, agriculture, agriengineering, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, asc, asi, atmosphere, atoms, axioms, batteries, bdcc, behavsci , beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci , buildings, cancers, carbon , catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, children, cleantechnol, climate, clockssleep, cmd, coatings, colloids, computation, computers, condensedmatter, cosmetics, cryptography, crystals, dairy, data, dentistry, designs , diagnostics, diseases, diversity, drones, econometrics, economies, education, electrochem, electronics, energies, entropy, environments, epigenomes, est, fermentation, fibers, fire, fishes, fluids, foods, forecasting, forests, fractalfract, futureinternet, futurephys, galaxies, games, gastrointestdisord, gels, genealogy, genes, geohazards, geosciences, geriatrics, hazardousmatters, healthcare, heritage, highthroughput, horticulturae, humanities, hydrology, ijerph, ijfs, ijgi, ijms, ijns, ijtpp, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jcdd, jcm, jcp, jcs, jdb, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmse, jnt, jof, joitmc, jpm, jrfm, jsan, land, languages, laws, life, literature, logistics, lubricants, machines, magnetochemistry, make, marinedrugs, materials, mathematics, mca, medicina, medicines, medsci, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, modelling, molbank, molecules, mps, mti, nanomaterials, ncrna, neuroglia, nitrogen, notspecified, nutrients, ohbm, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, physics, plants, plasma, polymers, polysaccharides, preprints , proceedings, processes, proteomes, psych, publications, quantumrep, quaternary, qubs, reactions, recycling, religions, remotesensing, reports, resources, risks, robotics, safety, sci, scipharm, sensors, separations, sexes, signals, sinusitis, smartcities, sna, societies, socsci, soilsystems, sports, standards, stats, surfaces, surgeries, sustainability, symmetry, systems, technologies, test, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vehicles, vetsci, vibration, viruses, vision, water, wem, wevj

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by: 
% abstract, addendum, article, benchmark, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, conferenceproceedings, correction, conferencereport, expressionofconcern, extendedabstract, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimages, letter, meetingreport, newbookreceived, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, supfile, technicalnote, viewpoint
% supfile = supplementary materials

%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.

%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.

%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. If eps figures are used, remove the option pdftex and use LaTeX and dvi2pdf.

%=================================================================
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{xx}
\issuenum{1}
\articlenumber{5}
\pubyear{2019}
\copyrightyear{2019}
%\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}
%\updates{yes} % If there is an update available, un-comment this line

%% MDPI internal command: uncomment if new journal that already uses continuous page numbers 
%\continuouspages{yes}

%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, calc, indentfirst, fancyhdr, graphicx, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, amsthm, hyphenat, natbib, hyperref, footmisc, geometry, caption, url, mdframed, tabto, soul, multirow, microtype, tikz

%=================================================================
%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{Moisture effects on leaf litter are species-specific and result in non-additive effects flammability in mixed conifer forest}

% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0000-0000-000-000X} % Add \orcidA{} behind the author's name
%\newcommand{\orcidauthorB}{0000-0000-000-000X} % Add \orcidB{} behind the author's name

% Authors, for the paper (add full first names)
\Author{Rita Quinones de Magalhaes $^{1,\dagger,\ddagger}$\orcidA{} and Dylan W. Schwilk $^{2,\ddagger}$}

% Authors, for metadata in PDF
\AuthorNames{Rita Magalhaes, Dylan W. Schwilk}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
$^{2}$ \quad Texas Tech UnNiversity; dylan.schwilk@ttu.edu}

% Contact information of the corresponding author
\corres{Correspondence: dylan.schwilk@ttu.edu, DWS}

% Current address and/or shared authorship
\firstnote{Current address: Affiliation 3} 
\secondnote{These authors contributed equally to this work.}
% The commands \thirdnote{} till \eighthnote{} are available for further notes

%\simplesumm{} % Simple summary

%\conference{} % An extended version of a conference paper

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{ Moisture content is a strong determinant of forest fuel
  flammability. But less is known about the mechanisms by which litter traits
  influence litter moisture dynamics, particularly the processes of moisture
  absorption and desorption. Characterizing how species’ leaf litter retain and
  release moisture through time is critical to the assessment of fuel’s
  availability to burn. Natural forests are often comprised of mixed stands,
  leading to mixtures of leaf litter on the surface fuelling the fires. Two
  studies have already established the existence of non-additive effects in the
  flammability of dry litter, but we lack a mechanistic understanding of
  moisture synergies in mixed litter and of moisture effects on the
  flammability of mixed litter. The litter from eight species of the
  mixed-conifer forest was saturated and allowed to dry to determine moisture
  absorption and desorption. Burn trials were performed in moist litter beds of
  single species and of mixtures, to establish flammability response to the
  dry-down process. Moisture absorption is determined by traits pertaining to
  leaf geometry while moisture desorption is influenced by emergent properties
  of the litterbed. Species that produce more aerated litter beds (such as Q.
  kelloggii and the pines) have faster rates of litter desiccation. My results
  show a relationship between time since wetting and time to ignition and
  spread rate, with significant interaction with genus. This indicates a
  genus-specific behavior of flammability and time since wetting. I found
  non-additive effects in litter mixtures for both moisture content and
  flammability. Mixtures containing Q. kelloggii and P. jeffreyi, species that
  dry faster, produced more flammable behavior (lower time to ignition, faster
  spread rate, taller flames). Fire behavior of mixtures depended on moisture
  content, although mixtures without Q. kelloggii had always higher moisture
  content than predicted by the average fire behavior of the species in the
  mixture. At lower moisture contents (<75\%) the mixtures exhibit positive
  non-additive behavior and at higher moisture contents (> 200\%) they exhibit
  negative non-additive behavior.}

% A single paragraph of about 200 words maximum. For research articles,
% abstracts should give a pertinent overview of the work. We strongly encourage
% authors to use the following style of structured abstracts, but without
% headings: (1) Background: Place the question addressed in a broad context and
% highlight the purpose of the study; (2) Methods: Describe briefly the main
% methods or treatments applied; (3) Results: Summarize the article's main
% findings; and (4) Conclusion: Indicate the main conclusions or
% interpretations. The abstract should be an objective representation of the
% article, it must not contain results which are not presented and
% substantiated in the main text and should not exaggerate the main
% conclusions.

% Keywords
\keyword{flammability, litter, flame spread rate}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
%\MSC{}
%\JEL{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Diversity
%\LSID{\url{http://}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Applied Sciences:
%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data:
%\dataset{DOI number or link to the deposited data set in cases where the data set is published or set to be published separately. If the data set is submitted and will be published as a supplement to this paper in the journal Data, this field will be filled by the editors of the journal. In this case, please make sure to submit the data set as a supplement when entering your manuscript into our manuscript editorial system.}

%\datasetlicense{license under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Toxins
%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}

%\setcounter{secnumdepth}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{section}{-1} %% Remove this when starting to work on the template.
\section{How to Use this Template}
The template details the sections that can be used in a manuscript. Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section). Please check the instructions for authors page of the journal to verify the correct order and names. For any questions, please contact the editorial office of the journal or support@mdpi.com. For LaTeX related questions please contact latex@mdpi.com.
%The order of the section titles is: Introduction, Materials and Methods, Results, Discussion, Conclusions for these journals: aerospace,algorithms,antibodies,antioxidants,atmosphere,axioms,biomedicines,carbon,crystals,designs,diagnostics,environments,fermentation,fluids,forests,fractalfract,informatics,information,inventions,jfmk,jrfm,lubricants,neonatalscreening,neuroglia,particles,pharmaceutics,polymers,processes,technologies,viruses,vision

\section{Introduction}
% The introduction should briefly place the study in a broad context and highlight why it is important. It should define the purpose of the work and its significance. The current state of the research field should be reviewed carefully and key publications cited. Please highlight controversial and diverging hypotheses when necessary. Finally, briefly mention the main aim of the work and highlight the principal conclusions. As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research. Citing a journal paper \cite{ref-journal}. And now citing a book reference \cite{ref-book}. Please use the command \citep{ref-journal} for the following MDPI journals, which use author-date citation: Administrative Sciences, Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Languages, Laws, Religions, Risks, Social Sciences.

Fuel moisture content is an important, and often driving, determinant of surface fire behavior. Environmental conditions such as temperature, relative humidity, and wind, drive the moisture content of the fuel available to burn [CITE]. In particular, fine dead fuels respond very readily to environmental changes [CITE]. In mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fuelled by plant litter, leaf traits can influence landscape scale fire behavior (Schwilk \& Caprio 2011). Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape (Fonda 1998, Kane et al. 2008, Schwilk \& Caprio 2011, Kreye et al. 2013). Fine fuel moisture content strongly reduces ignition probability and flame spread rate (Gisborne 1936, Fons 1946, Anderson \& Rothermel 1965, Xanthopoulos \& Wakimoto 1993, Dimitrakopoulos \& Papaioannou 2001, Weise et al. 2005, Pellizzaro et al. 2007). However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be characterized by leaf traits, litter traits, or both.

Fuel moisture influences flammability and fire behavior through effects on ignition, spread rate, flame height, fuel consumption, and heat release, due to the high specific heat content of water (Rothermel 1972). Specifically, high moisture contents decrease the likelihood of ignition thereby reducing fire spread (Rothermel 1983, Nelson 2001). Moisture acts as a heat-sink resulting in less energy available for propagation, and the resulting water vapor will both cool the flaming front and dilute flammable gases (Albini 1976, Shafizadeh et al. 1977). Variation in moisture of living fuels has strong effects on flammability, but moisture of dead excised litter fuels can also vary. Moisture of litter fuels varies with climate, with seasons and with short term weather within a season.  Therefore, the rate of drying and flammability at any particular moisture content are key parameters that influence the fuel flammability throughout the fire season.  If species differ in the rate of drying following a precipitation event, this may strongly influence relatively flammability across species at any particular time post-wetting.  Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires (Pickett et al. 1999, Keeley et al. 2009). Fuel moisture dynamics of leaf litter are probably dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds (Swift et a. 1979, Hille \& Stephens 2005, Nelson \& Hiers 2008, Matthews 2014).

Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss (Kreye et al. 2013). Absorptive capacity represents the potential maximum fuel moisture and is evaluated by the moisture content of litter when saturated. After saturation, the differences in moisture content indicate different abilities to retain moisture. The ability of litter to absorb and desorb water should be dependent on leaf and/or litter geometry, however, which traits control these processes are largely unknown. I propose two different sets of traits responsible for variation in the two processes mentioned: leaf particle geometry and litter bulk density. Litter characteristics such as surface-area-to-volume ratio, litter bulk density (Gibos 2010), mass density, particle size and shape, and chemical traits (Nelson 2001) influence moisture diffusion through litter. Litter packing has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits (such as leaf size, litter bulk density). Fuelbeds composed of larger leaves with lower moisture retainability and decomposition rates favor drying, thus increasing the probability of ignition and fire spread rate. 
Species traits may interact such that the flammability of multi-species mixtures is not the average of the constituent species (van Altena et al. 2012, de Magalhães \& Schwilk 2012, Blauw et al. 2015).  Such non-additive effects have been shown to be synergistic in dry litter fuels where the flammability of a mixture is driven by the most flammable species (de Magalhães \& Schwilk 2012, van Altena et al. 2012). In most forests, leaf litter will be a mixture representative of the species in the stand. It is possible that similar non-additive effects may influence litter moisture content, especially if different leaf traits influence adsorptive capacity and drying rate. Recent work by Blauw et al. (2015) suggests that moisture content can contribute to non-additive effects on heathland flammability, but the mechanism of this effect is not known. I investigated non-additive effects on both moisture content and the flammability of litter mixtures varying in time since wetting and in moisture content. I expect that litter mixtures will differ in their bulk density and thus, will dry at different rates.

I investigated species trait effects on litter moisture dynamics and consequent litter flammability across eight tree species in a temperate fire-prone forest in California. I investigated the relationship between traits such as leaf thinness and litter bulk density with the processes of moisture absorption and desorption. I determined the rate of moisture loss, the moisture of extinction (the threshold representing the moisture value above which fire will not ignite or spread) and the influence of moisture on the flammability of eight species from a mixed-conifer forest in California. I performed burn trials on leaf litter at various moisture levels and determined how time since wetting influenced time to ignition and flame spread rate. The burn trials were also performed with mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects. I studied whether the non-additive effects on flammability reported in previous studies (de Magalhães \& Schwilk 2012, van Altena et al. 2012) occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Materials and Methods}

\subsection{Site description and species selection}

Field sites for this study were located in Sequoia and Kings Canyon National Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in mixed-conifer forest. I chose eight tree species that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \& Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws., \emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin, \emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J. Buchholz. In this study, I consider litter the top layer of leaves and small twigs less than 0.625 cm diameter (= 1-hour fuel, Fosberg 1970) that have fallen in the previous year (mostly undecomposed). I collected the leaf litter in the summer (mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the study area. Effort was made to collect from different populations (minimum of four, each population separated from the others by 9 to 32 km) for each species to control for plastic and ecotypic effects. An exception to this involved the litter from A. magnifica which is restricted to the higher elevations and could only be collected from one area, due to logistical constraints. Within that area, \emph{A. magnifica} litter was collected from over 10 different individuals. For the remaining species, the collection involved 2-4 individual trees at each site, separated by at least 10 m. The litter was collect about 2 m away from the trunk to obtain a more uniform sample, because bark and heavier branches tend to fall closer to the tree trunk.

\subsection{Moisture absorption and desorption}

I produced dry-down curves to determine maximum water retainability (moisture absorption) and the rate of moisture content loss (moisture desorption) for each species and for four distinct mixtures. These curves allowed me to determine the rate at which a species loses moisture (approximated as exponential decay), and determine maximum water content following saturation and gravity draining.  Samples from the eight species (six replicates per species) and four mixtures (5 replicates) were placed in baskets constructed of aluminium screen. I used 450 g of litter (dry weight) per trial, and controlled for depth, by changing the size of the basket. Weight was measured by using a balance sensitive to 0.1 g (model XS16001L, Mettler Toledo, Columbus, OH). Samples were immersed in clear plastic storage bins filled with water (Figure 2.1) for 24 hours to fully saturate them. After saturation, the baskets were removed from the water, allowed to drain for 3 minutes, and samples weighed to measure maximum moisture content. The samples were then allowed to dry in a partially controlled environment kept at 21 C temperature and 30\% relative humidity, assessed via an iButton sensor (Maxim Integrated) placed over the samples.  During the dry-down process three subsamples were taken at 24 hour intervals and weighed. These subsamples were then oven dried for 24 hours at 100 C and re-weighed to assess moisture content. This was repeated until the final sample from a basket was less than10\% moisture.

 The dry-down curves indicated that the eight species varied in their moisture behavior (Figure 2). To determine a subset of species to use in the mixture trials, I performed analysis of contrast using lsmeans (package: lsmeans; Lenth 2016) to determine similarity between the species, which fell into four groups. I then selected one species from each group to include in the mixtures resulting in four distinct 3-species mixtures.

Leaf trait data
We investigated the influence of leaf thinness and litter bulk density on the moisture dynamics of the species in this study. Leaf thinness data (leaf length over thickness) was obtained for subsamples of the litter collected for this project.  Bulk density of the litter was obtained for each basket during the dry-down experiment by dividing the dry mass (450 g) by the estimated litter volume (basket width x length x litter depth).

\subsection{Flammability assessment}

I performed flammability tests across multiple moisture levels on monospecific litter beds and on mixed litter beds replicated 5 times. I measured flammability using a 150 cm long burn table, 15 cm wide and 15 cm tall, in which leaf litter was placed and the table gently shaken to allow settling. In my experience, this design allows fire behaviour to reach constant flame spread rates (de Magalhães and Schwilk 2012).  Two tall metal rulers equally spaced along the apparatus allow visual determination of maximum flame height and nine type K thermocouples attached to data loggers measured temperature every second at three distances from the ignition point and at three different depths below the litter surface. Litter was ignited with a single ignition source (propane torch; maximum adiabatic flame temperature of 1899 C, http://www.benzomatic.com/products/fuel.aspx) at one end of the apparatus and allowed to burn until extinction. If a species at a specific moisture level failed to ignite after three attempts, no further attempts were made.  A timer was used to record time to ignition and duration of flaming combustion. The flammability trials followed the methodology detailed on de Magalhães and Schwilk (2012).

During flammability trials, I measured time to ignition (s), duration of flaming combustion (s), spread rate (mm/s), calculated by dividing the length of the burned surface by the time it takes the flaming front to reach the end of the apparatus, maximal flame height (mm), and mass loss (\%), calculated as the difference between fuel mass before and after flame extinction (Anderson 1970, Cornelissen et al. 2003, Ormeño et al. 2009). I also measured duration above 100 C, or the length of time temperature exceeded 100 C, and temperature integration, which integrates duration of combustion and temperature above 100 C and serves as a proxy for heat release (de Magalhães and Schwilk 2012, Cornwell et al. 2015). By performing the flammability tests at several moisture levels I were able to determine moisture of extinction (a threshold representing the moisture value above which litter will not ignite or spread) for all species in this study, using a logistic regression.

Burn trials occurred in a cement structure used to simulate house fires at the Fire Department of the City of Lubbock.  Though not completely air tight, this structure minimizes wind and helps regulate temperature and relative humidity. Temperature and relative humidity were measured before every trial using a Kestrel 3000 (Nielsen-Kellerman, Boothwyn, PA). To minimize temperature and relative humidity variation, I conducted trials between 10 am and 3 pm.

\subsection{Statistical analyses}

Dry-down curves (Figure 2.2) were created based on all data points collected, fitted with an exponential curve to each species:

%m = mmax e-λt								Eq. 1

Were m is moisture content at time t, and $\lambda$ is the exponential decay constant. I used a liner mixed-effects model taking the natural logarithm of moisture content, with time since wetting and species as fixed effects and nested within replicate. I extracted the coefficients of the fitted model to describe maximum water retainability (mmax; g of water per g of dry mass) and desiccation index ($\lambda$) for each species. I used post-hoc Tukey tests (package: lsmeans; Lenth 2016) to discriminate species’ behavior in terms of maximum water retainability and desiccation index. I analysed the influence of leaf size and litter bulk density on moisture dynamics (maximum water retainability and desiccation index) using linear nested models and used anova to determine which traits influenced the two processes. The effect of time since wetting on flammability was modelled using linear regression with the interaction between time since wetting and genus specified and with average vapor pressure deficit in the burning room during that trial as predictor. Differences across species were tested with post-hoc Tukey tests. A binomial logistic regression was fitted to the probability of ignition to obtain moisture of extinction.
I investigated non-additive effects on moisture and on flammability by testing if moisture content and flammability of mixtures were predicted by the average of the constituent species. I created the predicted values using the models that determine the relationship between moisture content/flammability and time since wetting for individual species, and then averaging the results obtained for the species in a mixture. This procedure was done for both moisture content and flammability parameter values of the constituent species. The expectation from the null model is that the difference between observed and predicted values of flammability will be zero. The null model is created by averaging the moisture content and the flammability of the individual species comprising the mixture for each time since wetting measured. I tested whether there were significant departures from the null (zero) with a mixed effects linear model (package: lme4; Bates et al. 2015). All analyses were conducted in R v.3.3.1 (R Core Team 2016).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}

\subsection{Leaf traits influence on moisture dynamics }

There are marked differences in the extent of moisture species retain, seen as the maximum moisture content at time 0 (y-axis intercept, Fig. 2.2). Q. kelloggii, the only broadleaf, is the species that retains more moisture. Tukey tests on the linear mixed-effects model discriminated the species into three different maximum water retention groups: 1) Q. kelloggii; 2) P. lambertiana and C. decurrens; and 3) A. magnifica, S. giganteum, A. concolor and P. jeffreyi. P. ponderosa could not be discriminated between the last two groups. There are also marked differences in the desiccation rate of the eight species and the Tukey tests discriminated species into three groups: 1) Pinus; 2) the Cupressaceae; and 3) Q. kelloggii in between. C. decurrens is among the species with the lowest moisture retainability and is the species with the slower desiccation. Species with lower litter bulk density, such as Q. kelloggii and the pines, have higher desiccation rates than all other species.

The regression between maximum water retention and leaf litter traits showed that maximum water retention was significantly affected by a measure of leaf size (particle length/thickness; p=0.0369) but not by a characteristic of the litter bed (bulk density), indicating that the ability to absorb water is particle-dependent and a feature of a leaf’s morphology and/or chemistry (Table 2.1). Desiccation rate was significantly affected by litter bulk density but not leaf thinness (p=0.8794), indicating that desiccation rate is controlled by the emergent properties of the fuelbed. This is explained by the fact that more aerated fuel beds will facilitate evaporation, due to increased diffusion of water molecules from the leaf surface.

\subsection{Time since wetting and its effect on flammability}

Time since wetting was modelled against time to ignition and spread rate using genus as a fixed effect, because in a previous study (de Magalhães and Schwilk 2012) I observed that species flammability grouped around genus. Both models showed significant effects

of time since wetting and genus on time to ignition and spread rate (F10, 171=24.51, p < 2.2e-16 and F10, 171=55. 85, p<2.2e-16 respectively; Figures 2.3 and 2.4). There were also significant interactions of time since wetting and genus on the response variables (time to ignition and spread rate; both p < 2.2e-16). For time since ignition, the negative relationship is driven by Sequoiadendron and Calocedrus, as evidenced by the steeper slopes of the curves. However, if I removed Sequoiadendron and Calocedrus from the models, the interaction of time since wetting and genus disappears for time to ignition (p=0.1643) but is still significant for spread rate (p = 2.05e-6). After drying for 72 hours, Q. kelloggi is always more flammable and more responsive to changes in moisture content, while C. decurrens and S. giganteum are not (Figures 2.3 and 2.4).

We used binomial logistic regression to model the relationship between moisture content and the probability of ignition to determine moisture of extinction for the eight species. Q. kelloggii ignited at all moisture levels tested (10--65\%) while C. decurrens and A. concolor failed to do so across most moisture levels. I assumed the inflection point on the curve at 50\% to allow for the determination of moisture of extinction for the remaining species (15\% for A. concolor and S. giganteum, 20\% for A. magnifica, 35\% for P. lambertiana, and 60\% for P. ponderosa and P. jeffreyi).

\subsection{Effects of moisture and flammability in litter mixtures}

Moisture effects on litter mixtures

The mixture that does not contain Q. kelloggii (AbCaPi; Fig. 2.5) absorbs less moisture. This result is to be expected considering the result obtained from the single species trials, and reveals that Q. kelloggii, as the species that absorbs more moisture, drives the water retainability behavior of the mixtures it appears in. In terms of maximum water retainability, the mixtures discriminate into three groups: 1) mixture without Q. kelloggii (AbCaPi), 2) mixtures with both A. concolor and Q. kelloggii (AbCaQu and AbPiQu), and 3) mixture without A. concolor (CaPiQu). There is no discrimination between the mixtures in terms of slopes. Though the mixtures with both C. decurrens and P. jeffreyi (AbCaPi and CaPiQu) are significantly different from one another, the remaining two mixtures (AbCaQu and AbPiQu) cannot be distinguished from one another nor from AbCaPi and CaPiQu. The moisture content of a mixture at any given time was also dependent upon the bulk density of the mixture (anova analysis resulted in a better model if bulk density was included as a fixed effect, p=0.01135).

Time since wetting effect on mixture flammability

In general, the flammability of mixtures was higher as time since wetting went up as expected (Table 2.2). The absence of Q. kelloggii in a mixture, in general, lead to slower flame spread rates, longer flaming combustion, and longer ignition times. The combination of both Q. kelloggii and P. jeffreyi in a mixture generated taller flames and shorter flaming combustion.

Non-additivity in moisture content

We examined the relationship between observed moisture content of a mixture, and the moisture content predicted based on the constituent species to determine the existence of non-additive effects in moisture content (Fig. 2.6). The mixture that does not contain Q. kelloggii (AbCaPi) consistently retains more moisture that what would be predicted by the average of its constituent species. In general, mixtures start the drying process drier than expected and after 48 hours are wetter than expected, indicating lower moisture absorption and slower desiccation rates. The non-additive behavior is dependent on the moisture content of the mixture. When moisture content is above 200\%, I observe negative non-additivity for all mixtures containing Q. kelloggii. There is additive behavior when moisture content is between 75 and 150\%, and positive non-additivity below 75\% moisture content.

Non-additivity in flammability

We determined the presence of non-additivity in the flammability of litter mixtures by comparing the flammability of the mixtures to a null model created by the arithmetic average of the predicted individual flammability of each species in the mixture (Figs. 2.7 and 2.8). Both spread rate and time to ignition exhibited non-additivity, mediated by time since wetting. Observed spread rate values for the mixtures were higher than predicted by the null model, though the non-additivity was stronger for mixtures that contained Q. kelloggii. This result is likely driven by the bulk density of the mixtures, which was lower than predicted, making the mixtures’ litter beds more aerated. Non-additivity in time to ignition is strongly influenced by the presence of C. decurrens in the mixture. This species failed to ignite several times in the individual trials and was assigned the maximum number of seconds for the ignition attempt, 180.

This section may be divided by subheadings. It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.
\begin{quote}
This section may be divided by subheadings. It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.
\end{quote}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Subsection}
\unskip
\subsubsection{Subsubsection}

Bulleted lists look like this:
\begin{itemize}[leftmargin=*,labelsep=5.8mm]
\item	First bullet
\item	Second bullet
\item	Third bullet
\end{itemize}

Numbered lists can be added as follows:
\begin{enumerate}[leftmargin=*,labelsep=4.9mm]
\item	First item 
\item	Second item
\item	Third item
\end{enumerate}

The text continues here.

\subsection{Figures, Tables and Schemes}

All figures and tables should be cited in the main text as Figure 1, Table 1, etc.

\begin{figure}[H]
\centering
\includegraphics[width=2 cm]{Definitions/logo-mdpi}
\caption{This is a figure, Schemes follow the same formatting. If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel. (\textbf{b}) Description of what is contained in the second panel. Figures should be placed in the main text near to the first time they are cited. A caption on a single line should be centered.}
\end{figure}

Text

Text

\begin{table}[H]
\caption{This is a table caption. Tables should be placed in the main text near to the first time they are cited.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
\midrule
entry 1		& data			& data\\
entry 2		& data			& data\\
\bottomrule
\end{tabular}
\end{table}

Text

Text

%\begin{listing}[H]
%\caption{Title of the listing}
%\rule{\textwidth}{1pt}
%\raggedright Text of the listing. In font size footnotesize, small, or normalsize. Preferred format: left aligned and single spaced. Preferred border format: top border line and bottom border line.
%\rule{\textwidth}{1pt}
%\end{listing}

\subsection{Formatting of Mathematical Components}

This is an example of an equation:

\begin{equation}
a + b = c
\end{equation}
%% If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments). This ensures correct linenumbering. The blank line should be removed when the documentclass option is changed to "accept" because the text following an equation should not be a new paragraph.

Please punctuate equations as regular text. Theorem-type environments (including propositions, lemmas, corollaries etc.) can be formatted as follows:
%% Example of a theorem:
\begin{Theorem}
Example text of a theorem.
\end{Theorem}

The text continues here. Proofs must be formatted as follows:

%% Example of a proof:
\begin{proof}[Proof of Theorem 1]
Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
\end{proof}
The text continues here.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}

Leaf traits influence on moisture dynamics 
The species in this study displayed different patterns of both moisture absorption and desorption (with Q. kelloggii and the pines exhibiting higher values for both), which were controlled by different litter traits (leaf thinness and litter bulk density, respectively). Q. kelloggii was the species that absorbed the most moisture, followed by the pines. The species that have a greater absorptive capacity are also the species that are usually more flammable when dry. This indicates that following a rain event, the most flammable species in my study would not be available as fuel immediately, but cannot be discounted as potential fuel. A species capacity to absorb moisture is controlled by leaf particle geometry, specifically the ratio between leaf length and leaf thickness, and not litter bed arrangement. The eight species differed in their desorption capacity. Q. kelloggii and the pines had higher desiccation rates, while C. decurrens exhibited lower desiccation rates. The ANOVA results indicate that leaf thinness best explains variation in maximum water retention, while litter bulk density best explains variation in desiccation rates. This result parallels bulk density results for these and other species (van Wagtendonk, Benedict and Sydoriak 1998; Stephens, Finney and Schantz 2004) de Magalhães and Schwilk 2012). So, the overall dry-down process of litter is being controlled by both traits intrinsic to the leaf particles (leaf geometry) and traits intrinsic to the litter bed (bulk density). Decomposition can also affect moisture content in litter. Van Wagner (1969), investigating fuel weathering, found that pine needles had a variation in moisture content 3 to 6 times greater when they lacked their protective wax coating when compared to intact needles.  
Time since wetting and flammability
I was interested in investigating how flammability, particularly time to ignition and spread rate, changed with time since wetting, as this is a more meaningful interpretation of moisture influences on flammability. After a rain event the litters of different species will exhibit different moisture contents. My results show that both time to ignition and spread rate were significantly affected by time since wetting, and the interaction of time since wetting and genus was also significant. This indicates that there is a species-specific response to flammability based on time since wetting, and the differences are due to different moisture contents at each time interval. Therefore, mere moisture content will not inform on the fire behavior measures of a litter bed. Species composition of the litter must be considered, and time since the last rain event should allow the determination of the potential flammability of an area. 
Time since wetting, as expected, lowered the flammability of the eight species in this study and its effect overwhelms the effect of volatile content. Q. kelloggii and the pines are more flammable at higher moisture contents when compared to volatile-rich species such as C. decurrens, Abies spp. or S. giganteum (Chapter 4 of this dissertation). This effect is influenced by litter bulk density and is likely to persist in fuel mixtures where the constituent species have different desiccation indices and therefore will have different moisture contents. I have determined moisture of extinction for the eight species in this

tudy. As expected, species that have higher flammability values (Pinus spp.) had higher moisture of extinction than species with lower flammability (Abies spp. and S. giganteum). I used a higher cut-off moisture for the burn trials than the moisture of extinction found in the literature (65\% vs. 27-30\%; Rothermel 1972, Dickinson et al. 2016) in order to be conservative, as moisture content studies as sparse. However, Q. kelloggii was still flammable at ~80-90\% moisture content (values at 72 hours used for the species mixtures), which I did not anticipate. Due to its high moisture content, Q. kelloggii was not burnt at 24 and 48 hours since wetting, which explains the gap shown in the Figures. Though I cannot comment on the exact behavior of Q. kelloggii at 24 and 48 hours, it is unlikely to be more flammable than the pines. This result indicates that though Q. kelloggii might be too wet to start a fire right after a rain event, Q. kelloggii litter dries faster than the other species in this study and will become very flammable within two days. 
Effects of moisture and flammability in litter mixtures

I tested the dry-down process and flammability of four unique three-species litter mixtures. These mixtures contained species representative of all dry-down behavior types observed in single species trials, as assessed by the Tukey test. As expected from the single species results, mixtures that contained Q. kelloggii absorbed more water during the saturation period, but there was no difference in desiccation rate among the mixtures. However, when investigating the non-additivity in moisture content, the mixture that did not contain Q. kelloggii retained more moisture than would be predicted by the null model (arithmetic average of the three constituent species in the mixture). In general, non-additivity varied with moisture content, with positive non-additive behavior at lower moisture contents (<75\%, 72-96 hours since wetting occurred) and negative non-additive behavior at very high moisture contents (>200\%, 0 hours since wetting occurred). This result indicates that immediately after a rain event, the moisture content of a mixture will be lower than expected, however these high values for moisture content are unlikely to be found in natural litters. As the mixture dries, the dry-down behavior will approximate the null model, and three to four days after the rain event the dry-down behavior will become non-additive again. The positive non-additivity seems influenced by litter bed characteristics, since the strongest non-additivity was observed for the mixture with two species that have less dense litter: Q. kelloggii and P. jeffreyi (CaPiQu).

The flammability results for the mixtures followed closely the results obtained for the individual species. Mixtures containing Q. kelloggii and P. jeffreyi had faster spread rates and lower times to ignition, resulting in higher overall flammability. These results are influenced by the lower bulk densities of litter mixtures containing those two species. There was a non-additive effect of moisture on flammability for the four mixtures in this study, reinforcing results shown for dry litter (de Magalhães and Schwilk 2012, van Altena et al. 2012). Spread rate and time to ignition had significant non-additivity, and were strongly influenced by the species present in the mixture. Presence of C. decurrens in a mixture marked a big difference between the null model and the prediction. This was due to several failed ignitions for this species when burned individually, which conditioned the predicted values, while all the mixtures ignited. There were marked differences in the non-additivity results from this study and the results from a previous study investigating dry litter samples. The presence of non-additive effects in moisture content leads to non-additive effects in the flammability of moist litter beds. When accounting for this effect, the non-additivity on the flammability of moist litter beds is subtler, when compared to species effects. This result, together with previously published work, suggests that litter flammability is prone to non-additive effects, reinforcing the idea that we cannot approach flammability by lumping species into fuel categories, as is common in fire behavior models. Community flammability is dependent on and particular to its constituent species and cannot be addressed by simply averaging the flammability of all species it comprises. This dynamic view is likely to be increasingly relevant considering ongoing climate change, which predicts changes in community assembly.

I have found non-additive effects (both positive and negative) in moisture content and in the flammability of wet litter mixtures of mixed-conifer forest. These results complement the efforts of Blauw et al. (2015) for moist litter and add to the literature investigating non-additivity in litter flammability (de Magalhães and Schwilk 2012, van Altena et al. 2012). However, these studies are limited in the scope of the species studied and suggest that non-additivity needs to be investigated at a larger scale, within different systems and comparatively across a larger number of species. Moisture content is considered the single most important driver of litter flammability and the results from my study indicate how leaf and litter traits affect moisture dynamics, and how moisture effects on litter flammability are species-specific. A trait-based perspective provides a better mechanistic understanding of moisture dynamics of litter fuel flammability and fire behavior and is especially important for predicting and understanding fire in plant communities that are changing.

% Authors should discuss the results and how they can be interpreted in perspective of previous studies and of the working hypotheses. The findings and their implications should be discussed in the broadest context possible. Future research directions may also be highlighted.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{Conclusions}

% This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{Patents}
% This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{6pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
%\supplementary{The following are available online at \linksupplementary{s1}, Figure S1: title, Table S1: title, Video S1: title.}

% Only for the journal Methods and Protocols:
% If you wish to submit a video article, please do so with any other supplementary material.
% \supplementary{The following are available at \linksupplementary{s1}, Figure S1: title, Table S1: title, Video S1: title. A supporting video article is available at doi: link.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{For research articles with several authors, a short paragraph specifying their individual contributions must be provided. The following statements should be used ``conceptualization, X.X. and Y.Y.; methodology, X.X.; software, X.X.; validation, X.X., Y.Y. and Z.Z.; formal analysis, X.X.; investigation, X.X.; resources, X.X.; data curation, X.X.; writing--original draft preparation, X.X.; writing--review and editing, X.X.; visualization, X.X.; supervision, X.X.; project administration, X.X.; funding acquisition, Y.Y.'', please turn to the  \href{http://img.mdpi.org/data/contributor-role-instruction.pdf}{CRediT taxonomy} for the term explanation. Authorship must be limited to those who have contributed substantially to the work reported.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\funding{Please add: ``This research received no external funding'' or ``This research was funded by NAME OF FUNDER grant number XXX.'' and  and ``The APC was funded by XXX''. Check carefully that the details given are accurate and use the standard spelling of funding agency names at \url{https://search.crossref.org/funding}, any errors may affect your future funding.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections. This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{Declare conflicts of interest or state ``The authors declare no conflict of interest.'' Authors must identify and declare any personal circumstances or interest that may be perceived as inappropriately influencing the representation or interpretation of reported research results. Any role of the funders in the design of the study; in the collection, analyses or interpretation of data; in the writing of the manuscript, or in the decision to publish the results must be declared in this section. If there is no role, please state ``The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, or in the decision to publish the results''.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\abbreviations{The following abbreviations are used in this manuscript:\\

\noindent 
\begin{tabular}{@{}ll}
MDPI & Multidisciplinary Digital Publishing Institute\\
DOAJ & Directory of open access journals\\
TLA & Three letter acronym\\
LD & linear dichroism
\end{tabular}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\appendixtitles{no} %Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
\appendix
\section{}
\unskip
\subsection{}
The appendix is an optional section that can contain details and data supplemental to the main text. For example, explanations of experimental details that would disrupt the flow of the main text, but nonetheless remain crucial to understanding and reproducing the research shown; figures of replicates for experiments of which representative data is shown in the main text can be added here if brief, or as Supplementary data. Mathematical proofs of results not central to the paper can be added as an appendix.

\section{}
All appendix sections must be cited in the main text. In the appendixes, Figures, Tables, etc. should be labeled starting with `A', e.g., Figure A1, Figure A2, etc.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here.

%=====================================
% References, variant A: internal bibliography
%=====================================
\reftitle{References}
\begin{thebibliography}{999}
% Reference 1
\bibitem[Author1(year)]{ref-journal}
Author1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
% Reference 2
\bibitem[Author2(year)]{ref-book}
Author2, L. The title of the cited contribution. In {\em The Book Title}; Editor1, F., Editor2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
\end{thebibliography}

% The following MDPI journals use author-date citation: Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Laws, Religions, Risks, Social Sciences. For those journals, please follow the formatting guidelines on http://www.mdpi.com/authors/references
% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p. 328; \citeyear{ref-journal-3b}, p.475). This produces: Wong (1999, p. 328; 2000, p. 475)

%=====================================
% References, variant B: external bibliography
%=====================================
%\externalbibliography{yes}
%\bibliography{your_external_BibTeX_file}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\sampleavailability{Samples of the compounds ...... are available from the authors.}

%% for journal Sci
%\reviewreports{\\
%Reviewer 1 comments and authors’ response\\
%Reviewer 2 comments and authors’ response\\
%Reviewer 3 comments and authors’ response
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}

### 2019-04-22 ms/template.tex
Start on LaTeX manuscript
Add latex file based on MDPI template and MDPI latex class definitions.
See https://www.mdpi.com/authors/latex

--- 2019-04-22 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
@@
 
-%  =================================================================
-\documentclass[fire,article,submit,moreauthors,pdftex]{Definitions/mdpi}
-
-%\usepackage[utf8]{inputenc}
+%=================================================================
+\documentclass[journal,article,submit,moreauthors,pdftex]{Definitions/mdpi}
 
@@
 % Full title of the paper (Capitalized)
-\Title{Moisture effects on leaf litter are species-specific and result in non-additive effects flammability in mixed conifer forest}
+\Title{Title}
 
@@
 % Authors, for the paper (add full first names)
-\Author{Rita Quinones de Magalhaes $^{1,\dagger,\ddagger}$\orcidA{} and Dylan W. Schwilk $^{2,\ddagger}$}
+\Author{Firstname Lastname $^{1,\dagger,\ddagger}$\orcidA{}, Firstname Lastname $^{1,\ddagger}$ and Firstname Lastname $^{2,}$*}
 
 % Authors, for metadata in PDF
-\AuthorNames{Rita Magalhaes, Dylan W. Schwilk}
+\AuthorNames{Firstname Lastname, Firstname Lastname and Firstname Lastname}
 
@@
 $^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
-$^{2}$ \quad Texas Tech UnNiversity; dylan.schwilk@ttu.edu}
+$^{2}$ \quad Affiliation 2; e-mail@e-mail.com}
 
 % Contact information of the corresponding author
-\corres{Correspondence: dylan.schwilk@ttu.edu, DWS}
+\corres{Correspondence: e-mail@e-mail.com; Tel.: (optional; include country code; if there are multiple corresponding authors, add author initials) +xx-xxxx-xxx-xxxx (F.L.)}
 
@@
 % Abstract (Do not insert blank lines, i.e. \\) 
-\abstract{ Moisture content is a strong determinant of forest fuel
-  flammability. But less is known about the mechanisms by which litter traits
-  influence litter moisture dynamics, particularly the processes of moisture
-  absorption and desorption. Characterizing how species’ leaf litter retain and
-  release moisture through time is critical to the assessment of fuel’s
-  availability to burn. Natural forests are often comprised of mixed stands,
-  leading to mixtures of leaf litter on the surface fuelling the fires. Two
-  studies have already established the existence of non-additive effects in the
-  flammability of dry litter, but we lack a mechanistic understanding of
-  moisture synergies in mixed litter and of moisture effects on the
-  flammability of mixed litter. The litter from eight species of the
-  mixed-conifer forest was saturated and allowed to dry to determine moisture
-  absorption and desorption. Burn trials were performed in moist litter beds of
-  single species and of mixtures, to establish flammability response to the
-  dry-down process. Moisture absorption is determined by traits pertaining to
-  leaf geometry while moisture desorption is influenced by emergent properties
-  of the litterbed. Species that produce more aerated litter beds (such as Q.
-  kelloggii and the pines) have faster rates of litter desiccation. My results
-  show a relationship between time since wetting and time to ignition and
-  spread rate, with significant interaction with genus. This indicates a
-  genus-specific behavior of flammability and time since wetting. I found
-  non-additive effects in litter mixtures for both moisture content and
-  flammability. Mixtures containing Q. kelloggii and P. jeffreyi, species that
-  dry faster, produced more flammable behavior (lower time to ignition, faster
-  spread rate, taller flames). Fire behavior of mixtures depended on moisture
-  content, although mixtures without Q. kelloggii had always higher moisture
-  content than predicted by the average fire behavior of the species in the
-  mixture. At lower moisture contents (<75\%) the mixtures exhibit positive
-  non-additive behavior and at higher moisture contents (> 200\%) they exhibit
-  negative non-additive behavior.}
-
-% A single paragraph of about 200 words maximum. For research articles,
-% abstracts should give a pertinent overview of the work. We strongly encourage
-% authors to use the following style of structured abstracts, but without
-% headings: (1) Background: Place the question addressed in a broad context and
-% highlight the purpose of the study; (2) Methods: Describe briefly the main
-% methods or treatments applied; (3) Results: Summarize the article's main
-% findings; and (4) Conclusion: Indicate the main conclusions or
-% interpretations. The abstract should be an objective representation of the
-% article, it must not contain results which are not presented and
-% substantiated in the main text and should not exaggerate the main
-% conclusions.
+\abstract{A single paragraph of about 200 words maximum.
+For research articles, abstracts should give a pertinent overview of the work.
+We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: Place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: Describe briefly the main methods or treatments applied; (3) Results: Summarize the article's main findings; and (4) Conclusion: Indicate the main conclusions or interpretations.
+The abstract should be an objective representation of the article, it must not contain results which are not presented and substantiated in the main text and should not exaggerate the main conclusions.}
 
 % Keywords
-\keyword{flammability, litter, flame spread rate}
+\keyword{keyword 1; keyword 2; keyword 3 (list three to ten pertinent keywords specific to the article, yet reasonably common within the subject discipline.)}
 
@@
 \section{Introduction}
-% The introduction should briefly place the study in a broad context and highlight why it is important.
+The introduction should briefly place the study in a broad context and highlight why it is important.
 It should define the purpose of the work and its significance.
@@
 
-Fuel moisture content is an important, and often driving, determinant of surface fire behavior.
-Environmental conditions such as temperature, relative humidity, and wind, drive the moisture content of the fuel available to burn [CITE].
-In particular, fine dead fuels respond very readily to environmental changes [CITE].
-In mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fuelled by plant litter, leaf traits can influence landscape scale fire behavior (Schwilk \& Caprio 2011).
-Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape (Fonda 1998, Kane et al.
-2008, Schwilk \& Caprio 2011, Kreye et al.
-2013).
-Fine fuel moisture content strongly reduces ignition probability and flame spread rate (Gisborne 1936, Fons 1946, Anderson \& Rothermel 1965, Xanthopoulos \& Wakimoto 1993, Dimitrakopoulos \& Papaioannou 2001, Weise et al.
-2005, Pellizzaro et al.
-2007).
-However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be characterized by leaf traits, litter traits, or both.
-
-Fuel moisture influences flammability and fire behavior through effects on ignition, spread rate, flame height, fuel consumption, and heat release, due to the high specific heat content of water (Rothermel 1972).
-Specifically, high moisture contents decrease the likelihood of ignition thereby reducing fire spread (Rothermel 1983, Nelson 2001).
-Moisture acts as a heat-sink resulting in less energy available for propagation, and the resulting water vapor will both cool the flaming front and dilute flammable gases (Albini 1976, Shafizadeh et al.
-1977).
-Variation in moisture of living fuels has strong effects on flammability, but moisture of dead excised litter fuels can also vary.
-Moisture of litter fuels varies with climate, with seasons and with short term weather within a season.
-Therefore, the rate of drying and flammability at any particular moisture content are key parameters that influence the fuel flammability throughout the fire season.
-If species differ in the rate of drying following a precipitation event, this may strongly influence relatively flammability across species at any particular time post-wetting.
-Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires (Pickett et al.
-1999, Keeley et al.
-2009).
-Fuel moisture dynamics of leaf litter are probably dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds (Swift et a.
-1979, Hille \& Stephens 2005, Nelson \& Hiers 2008, Matthews 2014).
-
-Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss (Kreye et al.
-2013).
-Absorptive capacity represents the potential maximum fuel moisture and is evaluated by the moisture content of litter when saturated.
-After saturation, the differences in moisture content indicate different abilities to retain moisture.
-The ability of litter to absorb and desorb water should be dependent on leaf and/or litter geometry, however, which traits control these processes are largely unknown.
-I propose two different sets of traits responsible for variation in the two processes mentioned: leaf particle geometry and litter bulk density.
-Litter characteristics such as surface-area-to-volume ratio, litter bulk density (Gibos 2010), mass density, particle size and shape, and chemical traits (Nelson 2001) influence moisture diffusion through litter.
-Litter packing has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits (such as leaf size, litter bulk density).
-Fuelbeds composed of larger leaves with lower moisture retainability and decomposition rates favor drying, thus increasing the probability of ignition and fire spread rate.
-Species traits may interact such that the flammability of multi-species mixtures is not the average of the constituent species (van Altena et al.
-2012, de Magalhães \& Schwilk 2012, Blauw et al.
-2015).
-Such non-additive effects have been shown to be synergistic in dry litter fuels where the flammability of a mixture is driven by the most flammable species (de Magalhães \& Schwilk 2012, van Altena et al.
-2012).
-In most forests, leaf litter will be a mixture representative of the species in the stand.
-It is possible that similar non-additive effects may influence litter moisture content, especially if different leaf traits influence adsorptive capacity and drying rate.
-Recent work by Blauw et al.
-(2015) suggests that moisture content can contribute to non-additive effects on heathland flammability, but the mechanism of this effect is not known.
-I investigated non-additive effects on both moisture content and the flammability of litter mixtures varying in time since wetting and in moisture content.
-I expect that litter mixtures will differ in their bulk density and thus, will dry at different rates.
-
-I investigated species trait effects on litter moisture dynamics and consequent litter flammability across eight tree species in a temperate fire-prone forest in California.
-I investigated the relationship between traits such as leaf thinness and litter bulk density with the processes of moisture absorption and desorption.
-I determined the rate of moisture loss, the moisture of extinction (the threshold representing the moisture value above which fire will not ignite or spread) and the influence of moisture on the flammability of eight species from a mixed-conifer forest in California.
-I performed burn trials on leaf litter at various moisture levels and determined how time since wetting influenced time to ignition and flame spread rate.
-The burn trials were also performed with mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects.
-I studied whether the non-additive effects on flammability reported in previous studies (de Magalhães \& Schwilk 2012, van Altena et al.
-2012) occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Materials and Methods}
-
-\subsection{Site description and species selection}
-
-Field sites for this study were located in Sequoia and Kings Canyon National Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in mixed-conifer forest.
-I chose eight tree species that are representative of this type of forest: \emph{Pinus jeffreyi} Grev.
-\& Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl.
-ex Laws., \emph{Abies concolor} (Gord.
-\& Glend.) Lindl.
-ex Hildebr., \emph{Abies magnifica} A.
-Murr., \emph{Calocedrus decurrens} (Torr.) Florin, \emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
-Buchholz.
-In this study, I consider litter the top layer of leaves and small twigs less than 0.625 cm diameter (= 1-hour fuel, Fosberg 1970) that have fallen in the previous year (mostly undecomposed).
-I collected the leaf litter in the summer (mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the study area.
-Effort was made to collect from different populations (minimum of four, each population separated from the others by 9 to 32 km) for each species to control for plastic and ecotypic effects.
-An exception to this involved the litter from A.
-magnifica which is restricted to the higher elevations and could only be collected from one area, due to logistical constraints.
-Within that area, \emph{A.
-magnifica} litter was collected from over 10 different individuals.
-For the remaining species, the collection involved 2-4 individual trees at each site, separated by at least 10 m.
-The litter was collect about 2 m away from the trunk to obtain a more uniform sample, because bark and heavier branches tend to fall closer to the tree trunk.
-
-\subsection{Moisture absorption and desorption}
-
-I produced dry-down curves to determine maximum water retainability (moisture absorption) and the rate of moisture content loss (moisture desorption) for each species and for four distinct mixtures.
-These curves allowed me to determine the rate at which a species loses moisture (approximated as exponential decay), and determine maximum water content following saturation and gravity draining.
-Samples from the eight species (six replicates per species) and four mixtures (5 replicates) were placed in baskets constructed of aluminium screen.
-I used 450 g of litter (dry weight) per trial, and controlled for depth, by changing the size of the basket.
-Weight was measured by using a balance sensitive to 0.1 g (model XS16001L, Mettler Toledo, Columbus, OH).
-Samples were immersed in clear plastic storage bins filled with water (Figure 2.1) for 24 hours to fully saturate them.
-After saturation, the baskets were removed from the water, allowed to drain for 3 minutes, and samples weighed to measure maximum moisture content.
-The samples were then allowed to dry in a partially controlled environment kept at 21 C temperature and 30\% relative humidity, assessed via an iButton sensor (Maxim Integrated) placed over the samples.
-During the dry-down process three subsamples were taken at 24 hour intervals and weighed.
-These subsamples were then oven dried for 24 hours at 100 C and re-weighed to assess moisture content.
-This was repeated until the final sample from a basket was less than10\% moisture.
-
- The dry-down curves indicated that the eight species varied in their moisture behavior (Figure 2).
-To determine a subset of species to use in the mixture trials, I performed analysis of contrast using lsmeans (package: lsmeans; Lenth 2016) to determine similarity between the species, which fell into four groups.
-I then selected one species from each group to include in the mixtures resulting in four distinct 3-species mixtures.
-
-Leaf trait data
-We investigated the influence of leaf thinness and litter bulk density on the moisture dynamics of the species in this study.
-Leaf thinness data (leaf length over thickness) was obtained for subsamples of the litter collected for this project.
-Bulk density of the litter was obtained for each basket during the dry-down experiment by dividing the dry mass (450 g) by the estimated litter volume (basket width x length x litter depth).
-
-\subsection{Flammability assessment}
-
-I performed flammability tests across multiple moisture levels on monospecific litter beds and on mixed litter beds replicated 5 times.
-I measured flammability using a 150 cm long burn table, 15 cm wide and 15 cm tall, in which leaf litter was placed and the table gently shaken to allow settling.
-In my experience, this design allows fire behaviour to reach constant flame spread rates (de Magalhães and Schwilk 2012).
-Two tall metal rulers equally spaced along the apparatus allow visual determination of maximum flame height and nine type K thermocouples attached to data loggers measured temperature every second at three distances from the ignition point and at three different depths below the litter surface.
-Litter was ignited with a single ignition source (propane torch; maximum adiabatic flame temperature of 1899 C, http://www.benzomatic.com/products/fuel.aspx) at one end of the apparatus and allowed to burn until extinction.
-If a species at a specific moisture level failed to ignite after three attempts, no further attempts were made.
-A timer was used to record time to ignition and duration of flaming combustion.
-The flammability trials followed the methodology detailed on de Magalhães and Schwilk (2012).
-
-During flammability trials, I measured time to ignition (s), duration of flaming combustion (s), spread rate (mm/s), calculated by dividing the length of the burned surface by the time it takes the flaming front to reach the end of the apparatus, maximal flame height (mm), and mass loss (\%), calculated as the difference between fuel mass before and after flame extinction (Anderson 1970, Cornelissen et al.
-2003, Ormeño et al.
-2009).
-I also measured duration above 100 C, or the length of time temperature exceeded 100 C, and temperature integration, which integrates duration of combustion and temperature above 100 C and serves as a proxy for heat release (de Magalhães and Schwilk 2012, Cornwell et al.
-2015).
-By performing the flammability tests at several moisture levels I were able to determine moisture of extinction (a threshold representing the moisture value above which litter will not ignite or spread) for all species in this study, using a logistic regression.
-
-Burn trials occurred in a cement structure used to simulate house fires at the Fire Department of the City of Lubbock.
-Though not completely air tight, this structure minimizes wind and helps regulate temperature and relative humidity.
-Temperature and relative humidity were measured before every trial using a Kestrel 3000 (Nielsen-Kellerman, Boothwyn, PA).
-To minimize temperature and relative humidity variation, I conducted trials between 10 am and 3 pm.
-
-\subsection{Statistical analyses}
-
-Dry-down curves (Figure 2.2) were created based on all data points collected, fitted with an exponential curve to each species:
-
-%m = mmax e-λt								Eq. 1
-
-Were m is moisture content at time t, and $\lambda$ is the exponential decay constant.
-I used a liner mixed-effects model taking the natural logarithm of moisture content, with time since wetting and species as fixed effects and nested within replicate.
-I extracted the coefficients of the fitted model to describe maximum water retainability (mmax; g of water per g of dry mass) and desiccation index ($\lambda$) for each species.
-I used post-hoc Tukey tests (package: lsmeans; Lenth 2016) to discriminate species’ behavior in terms of maximum water retainability and desiccation index.
-I analysed the influence of leaf size and litter bulk density on moisture dynamics (maximum water retainability and desiccation index) using linear nested models and used anova to determine which traits influenced the two processes.
-The effect of time since wetting on flammability was modelled using linear regression with the interaction between time since wetting and genus specified and with average vapor pressure deficit in the burning room during that trial as predictor.
-Differences across species were tested with post-hoc Tukey tests.
-A binomial logistic regression was fitted to the probability of ignition to obtain moisture of extinction.
-I investigated non-additive effects on moisture and on flammability by testing if moisture content and flammability of mixtures were predicted by the average of the constituent species.
-I created the predicted values using the models that determine the relationship between moisture content/flammability and time since wetting for individual species, and then averaging the results obtained for the species in a mixture.
-This procedure was done for both moisture content and flammability parameter values of the constituent species.
-The expectation from the null model is that the difference between observed and predicted values of flammability will be zero.
-The null model is created by averaging the moisture content and the flammability of the individual species comprising the mixture for each time since wetting measured.
-I tested whether there were significant departures from the null (zero) with a mixed effects linear model (package: lme4; Bates et al.
-2015).
-All analyses were conducted in R v.3.3.1 (R Core Team 2016).
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Results}
-
-\subsection{Leaf traits influence on moisture dynamics }
-
-There are marked differences in the extent of moisture species retain, seen as the maximum moisture content at time 0 (y-axis intercept, Fig.
-2.2).
-Q.
-kelloggii, the only broadleaf, is the species that retains more moisture.
-Tukey tests on the linear mixed-effects model discriminated the species into three different maximum water retention groups: 1) Q.
-kelloggii; 2) P.
-lambertiana and C.
-decurrens; and 3) A.
-magnifica, S.
-giganteum, A.
-concolor and P.
-jeffreyi.
-P.
-ponderosa could not be discriminated between the last two groups.
-There are also marked differences in the desiccation rate of the eight species and the Tukey tests discriminated species into three groups: 1) Pinus; 2) the Cupressaceae; and 3) Q.
-kelloggii in between.
-C.
-decurrens is among the species with the lowest moisture retainability and is the species with the slower desiccation.
-Species with lower litter bulk density, such as Q.
-kelloggii and the pines, have higher desiccation rates than all other species.
-
-The regression between maximum water retention and leaf litter traits showed that maximum water retention was significantly affected by a measure of leaf size (particle length/thickness; p=0.0369) but not by a characteristic of the litter bed (bulk density), indicating that the ability to absorb water is particle-dependent and a feature of a leaf’s morphology and/or chemistry (Table 2.1).
-Desiccation rate was significantly affected by litter bulk density but not leaf thinness (p=0.8794), indicating that desiccation rate is controlled by the emergent properties of the fuelbed.
-This is explained by the fact that more aerated fuel beds will facilitate evaporation, due to increased diffusion of water molecules from the leaf surface.
-
-\subsection{Time since wetting and its effect on flammability}
-
-Time since wetting was modelled against time to ignition and spread rate using genus as a fixed effect, because in a previous study (de Magalhães and Schwilk 2012) I observed that species flammability grouped around genus.
-Both models showed significant effects
-
-of time since wetting and genus on time to ignition and spread rate (F10, 171=24.51, p < 2.2e-16 and F10, 171=55.
-85, p<2.2e-16 respectively; Figures 2.3 and 2.4).
-There were also significant interactions of time since wetting and genus on the response variables (time to ignition and spread rate; both p < 2.2e-16).
-For time since ignition, the negative relationship is driven by Sequoiadendron and Calocedrus, as evidenced by the steeper slopes of the curves.
-However, if I removed Sequoiadendron and Calocedrus from the models, the interaction of time since wetting and genus disappears for time to ignition (p=0.1643) but is still significant for spread rate (p = 2.05e-6).
-After drying for 72 hours, Q.
-kelloggi is always more flammable and more responsive to changes in moisture content, while C.
-decurrens and S.
-giganteum are not (Figures 2.3 and 2.4).
-
-We used binomial logistic regression to model the relationship between moisture content and the probability of ignition to determine moisture of extinction for the eight species.
-Q.
-kelloggii ignited at all moisture levels tested (10--65\%) while C.
-decurrens and A.
-concolor failed to do so across most moisture levels.
-I assumed the inflection point on the curve at 50\% to allow for the determination of moisture of extinction for the remaining species (15\% for A.
-concolor and S.
-giganteum, 20\% for A.
-magnifica, 35\% for P.
-lambertiana, and 60\% for P.
-ponderosa and P.
-jeffreyi).
-
-\subsection{Effects of moisture and flammability in litter mixtures}
-
-Moisture effects on litter mixtures
-
-The mixture that does not contain Q.
-kelloggii (AbCaPi; Fig.
-2.5) absorbs less moisture.
-This result is to be expected considering the result obtained from the single species trials, and reveals that Q.
-kelloggii, as the species that absorbs more moisture, drives the water retainability behavior of the mixtures it appears in.
-In terms of maximum water retainability, the mixtures discriminate into three groups: 1) mixture without Q.
-kelloggii (AbCaPi), 2) mixtures with both A.
-concolor and Q.
-kelloggii (AbCaQu and AbPiQu), and 3) mixture without A.
-concolor (CaPiQu).
-There is no discrimination between the mixtures in terms of slopes.
-Though the mixtures with both C.
-decurrens and P.
-jeffreyi (AbCaPi and CaPiQu) are significantly different from one another, the remaining two mixtures (AbCaQu and AbPiQu) cannot be distinguished from one another nor from AbCaPi and CaPiQu.
-The moisture content of a mixture at any given time was also dependent upon the bulk density of the mixture (anova analysis resulted in a better model if bulk density was included as a fixed effect, p=0.01135).
-
-Time since wetting effect on mixture flammability
-
-In general, the flammability of mixtures was higher as time since wetting went up as expected (Table 2.2).
-The absence of Q.
-kelloggii in a mixture, in general, lead to slower flame spread rates, longer flaming combustion, and longer ignition times.
-The combination of both Q.
-kelloggii and P.
-jeffreyi in a mixture generated taller flames and shorter flaming combustion.
-
-Non-additivity in moisture content
-
-We examined the relationship between observed moisture content of a mixture, and the moisture content predicted based on the constituent species to determine the existence of non-additive effects in moisture content (Fig.
-2.6).
-The mixture that does not contain Q.
-kelloggii (AbCaPi) consistently retains more moisture that what would be predicted by the average of its constituent species.
-In general, mixtures start the drying process drier than expected and after 48 hours are wetter than expected, indicating lower moisture absorption and slower desiccation rates.
-The non-additive behavior is dependent on the moisture content of the mixture.
-When moisture content is above 200\%, I observe negative non-additivity for all mixtures containing Q.
-kelloggii.
-There is additive behavior when moisture content is between 75 and 150\%, and positive non-additivity below 75\% moisture content.
-
-Non-additivity in flammability
-
-We determined the presence of non-additivity in the flammability of litter mixtures by comparing the flammability of the mixtures to a null model created by the arithmetic average of the predicted individual flammability of each species in the mixture (Figs.
-2.7 and 2.8).
-Both spread rate and time to ignition exhibited non-additivity, mediated by time since wetting.
-Observed spread rate values for the mixtures were higher than predicted by the null model, though the non-additivity was stronger for mixtures that contained Q.
-kelloggii.
-This result is likely driven by the bulk density of the mixtures, which was lower than predicted, making the mixtures’ litter beds more aerated.
-Non-additivity in time to ignition is strongly influenced by the presence of C.
-decurrens in the mixture.
-This species failed to ignite several times in the individual trials and was assigned the maximum number of seconds for the ignition attempt, 180.
 
@@
 
-Leaf traits influence on moisture dynamics 
-The species in this study displayed different patterns of both moisture absorption and desorption (with Q.
-kelloggii and the pines exhibiting higher values for both), which were controlled by different litter traits (leaf thinness and litter bulk density, respectively).
-Q.
-kelloggii was the species that absorbed the most moisture, followed by the pines.
-The species that have a greater absorptive capacity are also the species that are usually more flammable when dry.
-This indicates that following a rain event, the most flammable species in my study would not be available as fuel immediately, but cannot be discounted as potential fuel.
-A species capacity to absorb moisture is controlled by leaf particle geometry, specifically the ratio between leaf length and leaf thickness, and not litter bed arrangement.
-The eight species differed in their desorption capacity.
-Q.
-kelloggii and the pines had higher desiccation rates, while C.
-decurrens exhibited lower desiccation rates.
-The ANOVA results indicate that leaf thinness best explains variation in maximum water retention, while litter bulk density best explains variation in desiccation rates.
-This result parallels bulk density results for these and other species (van Wagtendonk, Benedict and Sydoriak 1998; Stephens, Finney and Schantz 2004) de Magalhães and Schwilk 2012).
-So, the overall dry-down process of litter is being controlled by both traits intrinsic to the leaf particles (leaf geometry) and traits intrinsic to the litter bed (bulk density).
-Decomposition can also affect moisture content in litter.
-Van Wagner (1969), investigating fuel weathering, found that pine needles had a variation in moisture content 3 to 6 times greater when they lacked their protective wax coating when compared to intact needles.
-Time since wetting and flammability
-I was interested in investigating how flammability, particularly time to ignition and spread rate, changed with time since wetting, as this is a more meaningful interpretation of moisture influences on flammability.
-After a rain event the litters of different species will exhibit different moisture contents.
-My results show that both time to ignition and spread rate were significantly affected by time since wetting, and the interaction of time since wetting and genus was also significant.
-This indicates that there is a species-specific response to flammability based on time since wetting, and the differences are due to different moisture contents at each time interval.
-Therefore, mere moisture content will not inform on the fire behavior measures of a litter bed.
-Species composition of the litter must be considered, and time since the last rain event should allow the determination of the potential flammability of an area.
-Time since wetting, as expected, lowered the flammability of the eight species in this study and its effect overwhelms the effect of volatile content.
-Q.
-kelloggii and the pines are more flammable at higher moisture contents when compared to volatile-rich species such as C.
-decurrens, Abies spp.
-or S.
-giganteum (Chapter 4 of this dissertation).
-This effect is influenced by litter bulk density and is likely to persist in fuel mixtures where the constituent species have different desiccation indices and therefore will have different moisture contents.
-I have determined moisture of extinction for the eight species in this
-
-tudy.
-As expected, species that have higher flammability values (Pinus spp.) had higher moisture of extinction than species with lower flammability (Abies spp.
-and S.
-giganteum).
-I used a higher cut-off moisture for the burn trials than the moisture of extinction found in the literature (65\% vs.
-27-30\%; Rothermel 1972, Dickinson et al.
-2016) in order to be conservative, as moisture content studies as sparse.
-However, Q.
-kelloggii was still flammable at ~80-90\% moisture content (values at 72 hours used for the species mixtures), which I did not anticipate.
-Due to its high moisture content, Q.
-kelloggii was not burnt at 24 and 48 hours since wetting, which explains the gap shown in the Figures.
-Though I cannot comment on the exact behavior of Q.
-kelloggii at 24 and 48 hours, it is unlikely to be more flammable than the pines.
-This result indicates that though Q.
-kelloggii might be too wet to start a fire right after a rain event, Q.
-kelloggii litter dries faster than the other species in this study and will become very flammable within two days.
-Effects of moisture and flammability in litter mixtures
-
-I tested the dry-down process and flammability of four unique three-species litter mixtures.
-These mixtures contained species representative of all dry-down behavior types observed in single species trials, as assessed by the Tukey test.
-As expected from the single species results, mixtures that contained Q.
-kelloggii absorbed more water during the saturation period, but there was no difference in desiccation rate among the mixtures.
-However, when investigating the non-additivity in moisture content, the mixture that did not contain Q.
-kelloggii retained more moisture than would be predicted by the null model (arithmetic average of the three constituent species in the mixture).
-In general, non-additivity varied with moisture content, with positive non-additive behavior at lower moisture contents (<75\%, 72-96 hours since wetting occurred) and negative non-additive behavior at very high moisture contents (>200\%, 0 hours since wetting occurred).
-This result indicates that immediately after a rain event, the moisture content of a mixture will be lower than expected, however these high values for moisture content are unlikely to be found in natural litters.
-As the mixture dries, the dry-down behavior will approximate the null model, and three to four days after the rain event the dry-down behavior will become non-additive again.
-The positive non-additivity seems influenced by litter bed characteristics, since the strongest non-additivity was observed for the mixture with two species that have less dense litter: Q.
-kelloggii and P.
-jeffreyi (CaPiQu).
-
-The flammability results for the mixtures followed closely the results obtained for the individual species.
-Mixtures containing Q.
-kelloggii and P.
-jeffreyi had faster spread rates and lower times to ignition, resulting in higher overall flammability.
-These results are influenced by the lower bulk densities of litter mixtures containing those two species.
-There was a non-additive effect of moisture on flammability for the four mixtures in this study, reinforcing results shown for dry litter (de Magalhães and Schwilk 2012, van Altena et al.
-2012).
-Spread rate and time to ignition had significant non-additivity, and were strongly influenced by the species present in the mixture.
-Presence of C.
-decurrens in a mixture marked a big difference between the null model and the prediction.
-This was due to several failed ignitions for this species when burned individually, which conditioned the predicted values, while all the mixtures ignited.
-There were marked differences in the non-additivity results from this study and the results from a previous study investigating dry litter samples.
-The presence of non-additive effects in moisture content leads to non-additive effects in the flammability of moist litter beds.
-When accounting for this effect, the non-additivity on the flammability of moist litter beds is subtler, when compared to species effects.
-This result, together with previously published work, suggests that litter flammability is prone to non-additive effects, reinforcing the idea that we cannot approach flammability by lumping species into fuel categories, as is common in fire behavior models.
-Community flammability is dependent on and particular to its constituent species and cannot be addressed by simply averaging the flammability of all species it comprises.
-This dynamic view is likely to be increasingly relevant considering ongoing climate change, which predicts changes in community assembly.
-
-I have found non-additive effects (both positive and negative) in moisture content and in the flammability of wet litter mixtures of mixed-conifer forest.
-These results complement the efforts of Blauw et al.
-(2015) for moist litter and add to the literature investigating non-additivity in litter flammability (de Magalhães and Schwilk 2012, van Altena et al.
-2012).
-However, these studies are limited in the scope of the species studied and suggest that non-additivity needs to be investigated at a larger scale, within different systems and comparatively across a larger number of species.
-Moisture content is considered the single most important driver of litter flammability and the results from my study indicate how leaf and litter traits affect moisture dynamics, and how moisture effects on litter flammability are species-specific.
-A trait-based perspective provides a better mechanistic understanding of moisture dynamics of litter fuel flammability and fire behavior and is especially important for predicting and understanding fire in plant communities that are changing.
-
-% Authors should discuss the results and how they can be interpreted in perspective of previous studies and of the working hypotheses.
+Authors should discuss the results and how they can be interpreted in perspective of previous studies and of the working hypotheses.
 The findings and their implications should be discussed in the broadest context possible.
@@
 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section{Conclusions}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Materials and Methods}
 
-% This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.
+Materials and Methods should be described with sufficient details to allow others to replicate and build on published results.
+Please note that publication of your manuscript implicates that you must make all materials, data, computer code, and protocols associated with the publication available to readers.
+Please disclose at the submission stage any restrictions on the availability of materials or information.
+New methods and protocols should be described in detail while well-established methods can be briefly described and appropriately cited.
+
+Research manuscripts reporting large datasets that are deposited in a publicly available database should specify where the data have been deposited and provide the relevant accession numbers.
+If the accession numbers have not yet been obtained at the time of submission, please state that they will be provided during review.
+They must be provided prior to publication.
+
+Interventionary studies involving animals or humans, and other studies require ethical approval must list the authority that provided approval and the corresponding ethical approval code.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section{Patents}
-% This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.
+\section{Conclusions}
+
+This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Patents}
+This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.
 

### 2019-04-22 scripts/moisture-ms-figs-tables.R

@@ -1,43 +1,44 @@
 # moistsingles-plots
 
+RESULTS <- "../ms/figs_tables/"
+
 source("theme-opts.R")
 source("dry-down.R")
 source("burn-moist.R")
 
 ## Figure 1: Plotting dry-down
-
 xbreaks <- seq(0, 144, 24)
 ybreaks <- seq(0, 700, 50)
 
 ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
   geom_point(size=1.2, alpha=0.7, stroke=0) +
+  geom_smooth(method="glm",
+              method.args=list(family=gaussian(link="log")), se=FALSE, size =0.6) +
   scale_colour_manual(values=schwilkcolors) +
   xlab("Hours since wetting") + ylab("Moisture by dry weight (%)") +
   scale_x_continuous(breaks=xbreaks) +
   scale_y_continuous(breaks=ybreaks) +
   pubtheme.nogridlines +
-  geom_smooth(method="glm",
-              method.args=list(family=gaussian(link="log")), se=FALSE, size =0.6) +
   theme(legend.position=c(0.75, 0.86),
         legend.title=element_blank(),
         legend.text = element_text(family=fontfamily, size=smsize, face="italic"))
  #       legend.key.height=unit(smsize,"pt"))
 
-ggsave("../results/moisture_ms/fig1_drydown-curves.pdf", width=col1, height=col1, units="cm")
+ggsave(file.path(RESULTS, "fig1_drydown-curves.pdf"), width=col1, height=col1, units="cm")
 
 ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
   geom_point(size=1.2, alpha=0.7, stroke=0) +
+  geom_smooth(method="lm", se=FALSE, size =0.6) +
   scale_colour_manual(values=schwilkcolors) +
   xlab("Hours since wetting") + ylab("Moisture by dry weight (%)") +
   scale_x_continuous(breaks=xbreaks) +
   scale_y_log10() +
   pubtheme.nogridlines +
-  geom_smooth(method="lm", se=FALSE, size =0.6) +
   theme(legend.position=c(0.75, 0.86),
         legend.title=element_blank(),
         legend.text = element_text(family=fontfamily, size=smsize, face="italic"))
 
-ggsave("../results/moisture_ms/fig1_drydown-curves_logged.pdf", width=col1, height=col1, units="cm")
[…]

### 2019-05-13 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Get extrafont working, edits

@@
+%% reftex-default-bibliography: ("/home/schwilk/write/bib/schwilk.bib")
+
 % LaTeX support: latex@mdpi.com 
@@
 
+%\bibliography{/home/schwilk/write/bib/schwilk.bib}
+
 %=================================================================
@@
 $^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
-$^{2}$ \quad Texas Tech UnNiversity; dylan.schwilk@ttu.edu}
+$^{2}$ \quad Texas Tech University; dylan.schwilk@ttu.edu}
 
@@
   availability to burn. Natural forests are often comprised of mixed stands,
-  leading to mixtures of leaf litter on the surface fuelling the fires. Two
-  studies have already established the existence of non-additive effects in the
+  leading to mixtures of leaf litter on the surface fuelling the fires. Multiple
+  studies have established the existence of non-additive effects in the
   flammability of dry litter, but we lack a mechanistic understanding of
   moisture synergies in mixed litter and of moisture effects on the
-  flammability of mixed litter. The litter from eight species of the
+  flammability of mixed litter. The litter from eight species of a
   mixed-conifer forest was saturated and allowed to dry to determine moisture
@@
   leaf geometry while moisture desorption is influenced by emergent properties
-  of the litterbed. Species that produce more aerated litter beds (such as Q.
-  kelloggii and the pines) have faster rates of litter desiccation. My results
+  of the litterbed. Species that produce more aerated litter beds (such as \emph{Q.
+  kelloggii} and \emph{Pinus} species) have faster rates of litter desiccation. My results
   show a relationship between time since wetting and time to ignition and
   spread rate, with significant interaction with genus. This indicates a
-  genus-specific behavior of flammability and time since wetting. I found
-  non-additive effects in litter mixtures for both moisture content and
-  flammability. Mixtures containing Q. kelloggii and P. jeffreyi, species that
+  genus-specific behavior of flammability and time since wetting.
+  We found non-additive effects in litter mixtures for both moisture content and
+  flammability. Mixtures containing \emph{Q. kelloggii} and \emph{P. jeffreyi}, species that
   dry faster, produced more flammable behavior (lower time to ignition, faster
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\setcounter{section}{-1} %% Remove this when starting to work on the template.
-\section{How to Use this Template}
-The template details the sections that can be used in a manuscript.
+% \setcounter{section}{-1} %% Remove this when starting to work on the template.
+% \section{How to Use this Template}
+% The template details the sections that can be used in a manuscript.
 Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section).
@@
 In particular, fine dead fuels respond very readily to environmental changes [CITE].
-In mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fuelled by plant litter, leaf traits can influence landscape scale fire behavior (Schwilk \& Caprio 2011).
-Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape (Fonda 1998, Kane et al.
-2008, Schwilk \& Caprio 2011, Kreye et al.
-2013).
-Fine fuel moisture content strongly reduces ignition probability and flame spread rate (Gisborne 1936, Fons 1946, Anderson \& Rothermel 1965, Xanthopoulos \& Wakimoto 1993, Dimitrakopoulos \& Papaioannou 2001, Weise et al.
-2005, Pellizzaro et al.
-2007).
+In mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fuelled by plant litter, leaf traits can influence landscape scale fire behavior \cite{Schwilk+Caprio-2011}.
+Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape \cite{Fonda+Belanger+etal-1998, Kane+Varner+etal-2008, Schwilk+Caprio-2011, Kreye+Varner+etal-2013}.
+Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936, Fons-1946, Anderson+Rothermal-1965}.
 However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be characterized by leaf traits, litter traits, or both.
 
-Fuel moisture influences flammability and fire behavior through effects on ignition, spread rate, flame height, fuel consumption, and heat release, due to the high specific heat content of water (Rothermel 1972).
-Specifically, high moisture contents decrease the likelihood of ignition thereby reducing fire spread (Rothermel 1983, Nelson 2001).
-Moisture acts as a heat-sink resulting in less energy available for propagation, and the resulting water vapor will both cool the flaming front and dilute flammable gases (Albini 1976, Shafizadeh et al.
-1977).
+Fuel moisture influences flammability and fire behavior through effects on ignition, spread rate, flame height, fuel consumption, and heat release, due to the high specific heat content of water \cite{Rothermel-1972}.
+Specifically, high moisture contents decrease the likelihood of ignition thereby reducing fire spread \cite{Nelson-2011}.
+Moisture acts as a heat-sink resulting in less energy available for propagation, and the resulting water vapor will both cool the flaming front and dilute flammable gases \cite{Albini-1976, Shafizadeh-1977}.
 Variation in moisture of living fuels has strong effects on flammability, but moisture of dead excised litter fuels can also vary.
 Moisture of litter fuels varies with climate, with seasons and with short term weather within a season.
-Therefore, the rate of drying and flammability at any particular moisture content are key parameters that influence the fuel flammability throughout the fire season.
+Therefore, the rate of drying and flammability at any particular moisture content are key parameters that influence the fuel flammability throughout the fire season \cite{Viney-1991}.
 If species differ in the rate of drying following a precipitation event, this may strongly influence relatively flammability across species at any particular time post-wetting.
-Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires (Pickett et al.
-1999, Keeley et al.
-2009).
-Fuel moisture dynamics of leaf litter are probably dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds (Swift et a.
-1979, Hille \& Stephens 2005, Nelson \& Hiers 2008, Matthews 2014).
+Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires.
+Fuel moisture dynamics of leaf litter are dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds \cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}.
 
-Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss (Kreye et al.
-2013).
+Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss \citep{Kreye-2013}.
 Absorptive capacity represents the potential maximum fuel moisture and is evaluated by the moisture content of litter when saturated.
@@
 The ability of litter to absorb and desorb water should be dependent on leaf and/or litter geometry, however, which traits control these processes are largely unknown.
-I propose two different sets of traits responsible for variation in the two processes mentioned: leaf particle geometry and litter bulk density.
-Litter characteristics such as surface-area-to-volume ratio, litter bulk density (Gibos 2010), mass density, particle size and shape, and chemical traits (Nelson 2001) influence moisture diffusion through litter.
+We propose two different sets of traits responsible for variation in the two processes mentioned: leaf particle geometry and litter bulk density.
+Litter characteristics such as surface-area-to-volume ratio, litter bulk density, mass density, particle size and shape, and chemical traits influence moisture diffusion through litter.
 Litter packing has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits (such as leaf size, litter bulk density).
 Fuelbeds composed of larger leaves with lower moisture retainability and decomposition rates favor drying, thus increasing the probability of ignition and fire spread rate.
-Species traits may interact such that the flammability of multi-species mixtures is not the average of the constituent species (van Altena et al.
-2012, de Magalhães \& Schwilk 2012, Blauw et al.
-2015).
-Such non-additive effects have been shown to be synergistic in dry litter fuels where the flammability of a mixture is driven by the most flammable species (de Magalhães \& Schwilk 2012, van Altena et al.
-2012).
+
+Species traits may interact such that the flammability of multi-species mixtures is not the average of the constituent species.
+Such non-additive effects have been shown to be synergistic in dry litter fuels where the flammability of a mixture is driven by the most flammable species  \cite{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}.
 In most forests, leaf litter will be a mixture representative of the species in the stand.
 It is possible that similar non-additive effects may influence litter moisture content, especially if different leaf traits influence adsorptive capacity and drying rate.
-Recent work by Blauw et al.
-(2015) suggests that moisture content can contribute to non-additive effects on heathland flammability, but the mechanism of this effect is not known.
-I investigated non-additive effects on both moisture content and the flammability of litter mixtures varying in time since wetting and in moisture content.
+Blauw et al.\cite{Blauw+Wensink+etal-2015} suggests that moisture content can contribute to non-additive effects on heathland flammability, but the mechanism of this effect is not known.
+
+% I investigated non-additive effects on both moisture content and the flammability of litter mixtures varying in time since wetting and in moisture content.
 I expect that litter mixtures will differ in their bulk density and thus, will dry at different rates.
 
-I investigated species trait effects on litter moisture dynamics and consequent litter flammability across eight tree species in a temperate fire-prone forest in California.
-I investigated the relationship between traits such as leaf thinness and litter bulk density with the processes of moisture absorption and desorption.
-I determined the rate of moisture loss, the moisture of extinction (the threshold representing the moisture value above which fire will not ignite or spread) and the influence of moisture on the flammability of eight species from a mixed-conifer forest in California.
-I performed burn trials on leaf litter at various moisture levels and determined how time since wetting influenced time to ignition and flame spread rate.
-The burn trials were also performed with mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects.
-I studied whether the non-additive effects on flammability reported in previous studies (de Magalhães \& Schwilk 2012, van Altena et al.
-2012) occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.
+We investigated species trait effects on litter moisture dynamics and consequent litter flammability across eight tree species in a temperate fire-prone forest in California.
+We determined the rate of moisture loss, the moisture of extinction (the threshold representing the moisture value above which fire will not ignite or spread) and the influence of moisture on the flammability of eight species from a mixed-conifer forest in California.
+Experimental burn trials of leaf litter at various moisture levels determined how time since wetting influenced time to ignition and flame spread rate.
+The burn trials were also performed with multi-species litter mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects.
+We compared leaf litter traits with the measured moisture absorption and desorption rates.
+Finally, we tested if the non-additive effects on flammability reported in previous studies occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.
 
@@
 Buchholz.
-In this study, I consider litter the top layer of leaves and small twigs less than 0.625 cm diameter (= 1-hour fuel, Fosberg 1970) that have fallen in the previous year (mostly undecomposed).
+In this study, I consider litter the top layer of leaves and small twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the previous year (mostly undecomposed).
 I collected the leaf litter in the summer (mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the study area.
@@
 
-I produced dry-down curves to determine maximum water retainability (moisture absorption) and the rate of moisture content loss (moisture desorption) for each species and for four distinct mixtures.
-These curves allowed me to determine the rate at which a species loses moisture (approximated as exponential decay), and determine maximum water content following saturation and gravity draining.
-Samples from the eight species (six replicates per species) and four mixtures (5 replicates) were placed in baskets constructed of aluminium screen.
-I used 450 g of litter (dry weight) per trial, and controlled for depth, by changing the size of the basket.
-Weight was measured by using a balance sensitive to 0.1 g (model XS16001L, Mettler Toledo, Columbus, OH).
-Samples were immersed in clear plastic storage bins filled with water (Figure 2.1) for 24 hours to fully saturate them.
-After saturation, the baskets were removed from the water, allowed to drain for 3 minutes, and samples weighed to measure maximum moisture content.
-The samples were then allowed to dry in a partially controlled environment kept at 21 C temperature and 30\% relative humidity, assessed via an iButton sensor (Maxim Integrated) placed over the samples.
-During the dry-down process three subsamples were taken at 24 hour intervals and weighed.
-These subsamples were then oven dried for 24 hours at 100 C and re-weighed to assess moisture content.
-This was repeated until the final sample from a basket was less than10\% moisture.
+We produced dry-down curves to determine maximum water retainability and the rate of moisture content loss (dessication rate) for each species and for four distinct mixtures.
+These curves allowed us to  determine the rate at which a species loses moisture (approximated as
+exponential decay), and determine maximum water content following saturation
+and gravity draining. Samples from the eight species (six replicates per
+species) and four mixtures (5 replicates per mixture type) were placed in
+baskets (approximately 45 x 45 cm by 15 cm tall) constructed of 1 cm aluminum
+screen. Samples were 450 g of litter (dry weight) per trial, and litter depth
+was standardized to 10 cm by changing the horizontal dimensions of the basket.
+Weight was measured by using a balance sensitive to 0.1 g (model XS16001L,
+Mettler Toledo, Columbus, OH). Samples were immersed in plastic storage bins
+filled with water for 24 hours to fully saturate them. After saturation, the
+baskets were removed from the water, allowed to drain for 3 minutes, and
+samples weighed to measure maximum moisture content. The samples were then
+allowed to dry in a partially controlled environment kept at 21 C temperature
+and 30\% relative humidity, assessed via an iButton sensor (Maxim Integrated)
+placed over the samples. During the dry-down process three subsamples were
+taken at 24 hour intervals and weighed. These subsamples were then oven dried
+for 24 hours at 100 C and re-weighed to assess moisture content. This was
+repeated until the final sample from a basket was less than10\% moisture.
 
- The dry-down curves indicated that the eight species varied in their moisture behavior (Figure 2).
-To determine a subset of species to use in the mixture trials, I performed analysis of contrast using lsmeans (package: lsmeans; Lenth 2016) to determine similarity between the species, which fell into four groups.
+The dry-down curves indicated that the eight species varied in their moisture behavior [FIG 2?].
+To determine a subset of species to use in the mixture trials, we performed analysis of contrasts using lsmeans package in R  (TODO Lenth 2016) to determine similarity between the species, which fell into four groups.
 I then selected one species from each group to include in the mixtures resulting in four distinct 3-species mixtures.
@@
  Leaf trait data
+
 We investigated the influence of leaf thinness and litter bulk density on the moisture dynamics of the species in this study.
@@
 
-I performed flammability tests across multiple moisture levels on monospecific litter beds and on mixed litter beds replicated 5 times.
-I measured flammability using a 150 cm long burn table, 15 cm wide and 15 cm tall, in which leaf litter was placed and the table gently shaken to allow settling.
-In my experience, this design allows fire behaviour to reach constant flame spread rates (de Magalhães and Schwilk 2012).
-Two tall metal rulers equally spaced along the apparatus allow visual determination of maximum flame height and nine type K thermocouples attached to data loggers measured temperature every second at three distances from the ignition point and at three different depths below the litter surface.
-Litter was ignited with a single ignition source (propane torch; maximum adiabatic flame temperature of 1899 C, http://www.benzomatic.com/products/fuel.aspx) at one end of the apparatus and allowed to burn until extinction.
+We performed flammability tests across multiple moisture levels on monospecific litter beds and on mixed litter beds with eaach species or mixture replicated 5 times at each moisture level.
+Flammability was assessed using a 150 cm long burn table, 15 cm wide and 15 cm tall, in which leaf litter was placed and the table gently shaken to allow settling.
+This design allows fire behaviour to reach constant flame spread rates \cite{Magalhaes+Schwilk-2012}.
+Two graduated metal rulers   equally spaced along the apparatus allowed visual determination of maximum flame height.
+Litter was ignited with a propane torch; maximum adiabatic flame temperature of 1899 C, http://www.benzomatic.com/products/fuel.aspx) at one end of the apparatus and allowed to burn until extinction.
 If a species at a specific moisture level failed to ignite after three attempts, no further attempts were made.
 A timer was used to record time to ignition and duration of flaming combustion.
-The flammability trials followed the methodology detailed on de Magalhães and Schwilk (2012).
+The flammability trials followed the methodology detailed in de Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.
 
-During flammability trials, I measured time to ignition (s), duration of flaming combustion (s), spread rate (mm/s), calculated by dividing the length of the burned surface by the time it takes the flaming front to reach the end of the apparatus, maximal flame height (mm), and mass loss (\%), calculated as the difference between fuel mass before and after flame extinction (Anderson 1970, Cornelissen et al.
-2003, Ormeño et al.
-2009).
+% and nine type-K thermocouples attached to data loggers measured temperature every second at three distances from the ignition point and at three different depths below the litter surface.
+
+During flammability trials, we recorded time to ignition (s), duration of flaming combustion (s), spread rate (mm/s), calculated by dividing the length of the burned surface by the time it takes the flaming front to reach the end of the apparatus, maximal flame height (mm), and mass loss (\%), calculated as the difference between fuel mass before and after flame extinction.
 I also measured duration above 100 C, or the length of time temperature exceeded 100 C, and temperature integration, which integrates duration of combustion and temperature above 100 C and serves as a proxy for heat release (de Magalhães and Schwilk 2012, Cornwell et al.
@@
 
-Burn trials occurred in a cement structure used to simulate house fires at the Fire Department of the City of Lubbock.
-Though not completely air tight, this structure minimizes wind and helps regulate temperature and relative humidity.
+Burn trials occurred in a cement structure used to simulate house fires at the Fire Department of the City of Lubbock.This structure minimizes wind and helps regulate temperature and relative humidity.
 Temperature and relative humidity were measured before every trial using a Kestrel 3000 (Nielsen-Kellerman, Boothwyn, PA).
-To minimize temperature and relative humidity variation, I conducted trials between 10 am and 3 pm.
+To minimize temperature and relative humidity variation, all trials were conducted on clear days between 10 am and 3 pm.
 
@@
 
-Dry-down curves (Figure 2.2) were created based on all data points collected, fitted with an exponential curve to each species:
+Dry-down curves were created based on the drying experiment, fitted with an exponential decay curve to each species.:
 
-%m = mmax e-λt								Eq. 1
+\begin{equation}
+m = m_{max} e^{-\lambda t}
+\end{equation}
 
-Were m is moisture content at time t, and $\lambda$ is the exponential decay constant.
-I used a liner mixed-effects model taking the natural logarithm of moisture content, with time since wetting and species as fixed effects and nested within replicate.
-I extracted the coefficients of the fitted model to describe maximum water retainability (mmax; g of water per g of dry mass) and desiccation index ($\lambda$) for each species.
-I used post-hoc Tukey tests (package: lsmeans; Lenth 2016) to discriminate species’ behavior in terms of maximum water retainability and desiccation index.
-I analysed the influence of leaf size and litter bulk density on moisture dynamics (maximum water retainability and desiccation index) using linear nested models and used anova to determine which traits influenced the two processes.
-The effect of time since wetting on flammability was modelled using linear regression with the interaction between time since wetting and genus specified and with average vapor pressure deficit in the burning room during that trial as predictor.
+Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential decay coefficient.
+To estimate these c oefficents we fit a linear mixed-effects model by first taking the natural logarithm of moisture content, with time since wetting and species as fixed effects nested within replicate basket.
+We extracted the coefficients of the fitted model to describe maximum water retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate ($\lambda$) for each species.
+All analyses were conducted in R \cite{RCoreTeam-2019}.
+We fit models with R packages `lme4' \cite{Bates_Machler_etal-2015} and `afex' \cite{Singmann_Bolker_etal-2017}.
+Approximate degrees of freedom and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997} which is recommended by Luke \cite{Luke-2017} as producing acceptable Type 1 error rates.
+
+We tested if litter bulk density influenced moisture dynamics (maximum water retainability and desiccation rate) by fitting linear models with species means as observations.
+
+The effect of time since wetting on flammability was modelled using linear regression with the interaction between time since wetting and genus as fixed effects and average vapor pressure deficit in the burning room during that trial as a nuisance covariate.
 Differences across species were tested with post-hoc Tukey tests.
 A binomial logistic regression was fitted to the probability of ignition to obtain moisture of extinction.
-I investigated non-additive effects on moisture and on flammability by testing if moisture content and flammability of mixtures were predicted by the average of the constituent species.
-I created the predicted values using the models that determine the relationship between moisture content/flammability and time since wetting for individual species, and then averaging the results obtained for the species in a mixture.
+
+We investigated non-additive effects on moisture and on flammability by testing if moisture content and flammability of mixtures were predicted by the average of the constituent species.
+To choose mixtures we first grouped the species according to their maximum water retainability and dessication rate according to estimated marginal means by species using the R package `emmeans' \cite{Lenth-2019}.
+This resulted in four broad groups of species and we then selected one representative species from each litter type group and produced all four possible three-species litter mixtures using those four species.
+For each mixture and time since wetting, we predicted moisture content and flame spread rate using the models fit to single species data described above and then calculated an expected moisture content and spread rate as the average of these three values for the mixture.
 This procedure was done for both moisture content and flammability parameter values of the constituent species.
-The expectation from the null model is that the difference between observed and predicted values of flammability will be zero.
-The null model is created by averaging the moisture content and the flammability of the individual species comprising the mixture for each time since wetting measured.
-I tested whether there were significant departures from the null (zero) with a mixed effects linear model (package: lme4; Bates et al.
-2015).
-All analyses were conducted in R v.3.3.1 (R Core Team 2016).
+The expectation from the null model was that the difference between observed and predicted values of flammability will be zero.
+We tested if there were significant departures from the null (zero) with a mixed effects linear model.
 
@@
 
-There are marked differences in the extent of moisture species retain, seen as the maximum moisture content at time 0 (y-axis intercept, Fig.
-2.2).
-Q.
-kelloggii, the only broadleaf, is the species that retains more moisture.
-Tukey tests on the linear mixed-effects model discriminated the species into three different maximum water retention groups: 1) Q.
-kelloggii; 2) P.
-lambertiana and C.
-decurrens; and 3) A.
-magnifica, S.
-giganteum, A.
-concolor and P.
-jeffreyi.
-P.
-ponderosa could not be discriminated between the last two groups.
-There are also marked differences in the desiccation rate of the eight species and the Tukey tests discriminated species into three groups: 1) Pinus; 2) the Cupressaceae; and 3) Q.
-kelloggii in between.
-C.
-decurrens is among the species with the lowest moisture retainability and is the species with the slower desiccation.
-Species with lower litter bulk density, such as Q.
-kelloggii and the pines, have higher desiccation rates than all other species.
+There are marked differences in the extent of moisture species retain initially following wetting and draining by gravity (all p < 0.0001, Fig.
+\ref{fig1-drydown} and Table \ref{tab1-drydown}).
+\emph{Q.
+kelloggii}, the only broadleaf, is the species that retains the most moisture but has had a rapid dessication rate (.
+Calculating estimated marginal means following the linear mixed-effects model discriminated the species into three different maximum water retention groups: 1) \emph{Q.
+kelloggii}; 2) \emph{P.
+lambertiana} and \emph{C.
+decurrens}; and 3) \emph{A.
+magnifica}, \emph{S.
+giganteum}, \emph{A.
+concolor}, \emph{P.
+jeffreyi} and \emph{P.
+ponderosa}.
+There are also differences in desiccation rates across the eight species and estimated mixed linear model slopes discriminated species into three groups from lowest to most rapid: 1) (\emph{C.decurrens}, 2) \emph{Abies spp} and \emph{S.
+gianteum}, and 3) \emph{Q.
+kelloggii} and \emph{Pinus spp}.
 
-The regression between maximum water retention and leaf litter traits showed that maximum water retention was significantly affected by a measure of leaf size (particle length/thickness; p=0.0369) but not by a characteristic of the litter bed (bulk density), indicating that the ability to absorb water is particle-dependent and a feature of a leaf’s morphology and/or chemistry (Table 2.1).
-Desiccation rate was significantly affected by litter bulk density but not leaf thinness (p=0.8794), indicating that desiccation rate is controlled by the emergent properties of the fuelbed.
-This is explained by the fact that more aerated fuel beds will facilitate evaporation, due to increased diffusion of water molecules from the leaf surface.
+\begin{figure}[H]
+  \centering
+  \label{fig-drydown}
+\includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
+\caption{Dry down curves for eight litter types.
+Lines show best-fit exponential decay curves fit to moisture content on a dry mass basis.
+Litter samples we saturated by immersion for 24 hours and then allowed to drain by gravity for 3 min before initial weighing.
+Colors indicate genera and there were significant differences in both }
+\end{figure}
+
+\begin{table}[H]
+  \caption{Mixed model results for moisture content as a function of time since wetting.
+Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
+  \label{tab1-drydown}
+\centering
+%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
+\input{figs_tables/drydown-tab.ltx}
+\end{table}
+
+Across these eight species litter bulk density had a negative effect on desiccation rate (Fig. \ref{fig-bd-di}, linear model p = 0.011).
+
+\begin{figure}[H]
+  \centering
+  \label{fig-bd-di}
+\includegraphics[width=8cm]{figs_tables/fig3_di_by_bd.pdf}
+\caption{ }
+\end{figure}
 
@@
 
-Time since wetting was modelled against time to ignition and spread rate using genus as a fixed effect, because in a previous study (de Magalhães and Schwilk 2012) I observed that species flammability grouped around genus.
-Both models showed significant effects
+Time to ignition and spread rate were modeled against time since wetting and using genus as a fixed effect.
+
+[TODO]?  redo in terms of groups.
 
@@
 Moisture effects on litter mixtures
+
+Based on visual inspection of the dry down curves and the estimated marginal means for maximum water retainability and desiccation rate, selected four species to represent the full range of litter moisture dynamics for use in the litter mixtures: \emph{Abies concolor}, \
 
@@
 
-\begin{figure}[H]
-\centering
-\includegraphics[width=2 cm]{Definitions/logo-mdpi}
-\caption{This is a figure, Schemes follow the same formatting.
-If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel.
-(\textbf{b}) Description of what is contained in the second panel.
-Figures should be placed in the main text near to the first time they are cited.
-A caption on a single line should be centered.}
-\end{figure}
-
-Text
-
-Text
-
-\begin{table}[H]
-\caption{This is a table caption. Tables should be placed in the main text near to the first time they are cited.}
-\centering
-%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-\begin{tabular}{ccc}
-\toprule
-\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
-\midrule
-entry 1		& data			& data\\
-entry 2		& data			& data\\
-\bottomrule
-\end{tabular}
-\end{table}
-
 Text
@@
 
-\subsection{Formatting of Mathematical Components}
-
-This is an example of an equation:
-
-\begin{equation}
-a + b = c
-\end{equation}
-%% If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments).
-This ensures correct linenumbering.
-The blank line should be removed when the documentclass option is changed to "accept" because the text following an equation should not be a new paragraph.
-
-Please punctuate equations as regular text. Theorem-type environments (including propositions, lemmas, corollaries etc.) can be formatted as follows:
-%% Example of a theorem:
-\begin{Theorem}
-Example text of a theorem.
-\end{Theorem}
-
-The text continues here. Proofs must be formatted as follows:
-
-%% Example of a proof:
-\begin{proof}[Proof of Theorem 1]
-Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
-\end{proof}
-The text continues here.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Leaf traits influence on moisture dynamics 
+\subsection{Leaf traits influence on moisture dynamics }
+
 The species in this study displayed different patterns of both moisture absorption and desorption (with Q.
@@
 The ANOVA results indicate that leaf thinness best explains variation in maximum water retention, while litter bulk density best explains variation in desiccation rates.
-This result parallels bulk density results for these and other species (van Wagtendonk, Benedict and Sydoriak 1998; Stephens, Finney and Schantz 2004) de Magalhães and Schwilk 2012).
-So, the overall dry-down process of litter is being controlled by both traits intrinsic to the leaf particles (leaf geometry) and traits intrinsic to the litter bed (bulk density).
+
+Bulk densities measured here closely match those for these and other species \cite{van_Wagtendonk+Sydoriak+etal-1998,Stephens+Finney+etal-2004}
+
+The overall dry-down process of litter is being controlled by both traits intrinsic to the leaf particles (leaf geometry) and traits intrinsic to the litter bed (bulk density).
 Decomposition can also affect moisture content in litter.
 Van Wagner (1969), investigating fuel weathering, found that pine needles had a variation in moisture content 3 to 6 times greater when they lacked their protective wax coating when compared to intact needles.
-Time since wetting and flammability
-I was interested in investigating how flammability, particularly time to ignition and spread rate, changed with time since wetting, as this is a more meaningful interpretation of moisture influences on flammability.
-After a rain event the litters of different species will exhibit different moisture contents.
-My results show that both time to ignition and spread rate were significantly affected by time since wetting, and the interaction of time since wetting and genus was also significant.
+
+TODO
+
+With eight species, power to detect species-specific trait effects on these moisture patterns was weak, but we examined one litter trait known to directly influence flammability: litter bulk density.
+Litter density is largely driven by leaf size in this system \cite{Magalhaes+Schwilk-2012}.
+
+\subsection{Time since wetting and flammability}
+
+After a rain event, the litter of different species will exhibit different moisture contents.
+These results show that both time to ignition and spread rate were affected by time since wetting, and the interaction of time since wetting and genus was also significant.
+
 This indicates that there is a species-specific response to flammability based on time since wetting, and the differences are due to different moisture contents at each time interval.
@@
 Species composition of the litter must be considered, and time since the last rain event should allow the determination of the potential flammability of an area.
+
 Time since wetting, as expected, lowered the flammability of the eight species in this study and its effect overwhelms the effect of volatile content.
@@
 or S.
-giganteum (Chapter 4 of this dissertation).
+giganteum.
 This effect is influenced by litter bulk density and is likely to persist in fuel mixtures where the constituent species have different desiccation indices and therefore will have different moisture contents.
-I have determined moisture of extinction for the eight species in this
-
-tudy.
+I have determined moisture of extinction for the eight species in this study.
 As expected, species that have higher flammability values (Pinus spp.) had higher moisture of extinction than species with lower flammability (Abies spp.
@@
 I used a higher cut-off moisture for the burn trials than the moisture of extinction found in the literature (65\% vs.
-27-30\%; Rothermel 1972, Dickinson et al.
-2016) in order to be conservative, as moisture content studies as sparse.
+27-30\%; \citep{ Rothermel-1972} [TODO Dickinson et al.
+2016] in order to be conservative, as moisture content studies as sparse.
 However, Q.
@@
 kelloggii litter dries faster than the other species in this study and will become very flammable within two days.
-Effects of moisture and flammability in litter mixtures
 
-I tested the dry-down process and flammability of four unique three-species litter mixtures.
+\subsection{Effects of moisture and flammability in litter mixtures}
+
+We tested the dry-down process and flammability of four unique three-species litter mixtures.
 These mixtures contained species representative of all dry-down behavior types observed in single species trials, as assessed by the Tukey test.
@@
 These results are influenced by the lower bulk densities of litter mixtures containing those two species.
-There was a non-additive effect of moisture on flammability for the four mixtures in this study, reinforcing results shown for dry litter (de Magalhães and Schwilk 2012, van Altena et al.
-2012).
+There was a non-additive effect of moisture on flammability for the four mixtures in this study, reinforcing results shown for dry litter \citep{Magalhaes+Schwilk-2012, VanAltena+Logtestjin+etal-2012, }.
 Spread rate and time to ignition had significant non-additivity, and were strongly influenced by the species present in the mixture.
@@
 %=====================================
-\reftitle{References}
-\begin{thebibliography}{999}
-% Reference 1
-\bibitem[Author1(year)]{ref-journal}
-Author1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
-% Reference 2
-\bibitem[Author2(year)]{ref-book}
-Author2, L. The title of the cited contribution. In {\em The Book Title}; Editor1, F., Editor2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
-\end{thebibliography}
+% \reftitle{References}
+% \begin{thebibliography}{999}
+% % Reference 1
+% \bibitem[Author1(year)]{ref-journal}
+% Author1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
+% % Reference 2
+% \bibitem[Author2(year)]{ref-book}
+% Author2, L. The title of the cited contribution. In {\em The Book Title}; Editor1, F., Editor2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
+% \end{thebibliography}
 
@@
 % References, variant B: external bibliography
-%=====================================
-%\externalbibliography{yes}
-%\bibliography{your_external_BibTeX_file}
+% =====================================
+\externalbibliography{yes}
+\bibliography{/home/schwilk/write/bib/schwilk}
 
@@
 %% optional
-\sampleavailability{Samples of the compounds ...... are available from the authors.}
+%\sampleavailability{Samples of the compounds ...... are available from the authors.}
 

### 2019-05-13 scripts/burn-moist.R

@@ -41,13 +41,13 @@ modt2ignit2 <- lm(t2ignit ~ hour*genus + vpd, data=filter(burnt, genus!="Caloced
 summary(modt2ignit2)
 anova(modt2ignit2)
 
-library(agricolae)
+## library(agricolae)
 
-tuk <- HSD.test(modspread, "genus", group=TRUE)
-tuk
+## tuk <- HSD.test(modspread, "genus", group=TRUE)
+## tuk
 
-tuk <- HSD.test(modt2ignit, "genus", group=TRUE)
-tuk
+## tuk <- HSD.test(modt2ignit, "genus", group=TRUE)
+## tuk
 
 ##################################
 ## Binomial analysis of ignition

### 2019-05-13 scripts/dry-down.R

@@ -54,18 +54,33 @@ dry.mod <- lmer(log(MC_dry) ~ hour*spcode + (1 | tray), data=mc)
 summary(dry.mod)
 anova(dry.mod)
 
-library(afex)
-dry.mod <- mixed(log(MC_dry) ~ hour*spcode + (1 | tray), data=mc)
-dry.mod.tab <- summary(dry.mod)$coefficients
-dry.mod.anova.tab <- anova(dry.mod)
-dry.mod.tab
-dry.mod.anova.tab
-
 # pairwise tests:
 
 # test for pairwise differences and number of distinct groups
-#library(lsmeans)
-#cld(lstrends(dry.mod,~ spcode, var = "hour"))
+## library(emmeans)
+## dry.emm <- emmeans(dry.mod, ~ hour | spcode)
+## CLD(dry.mod)
+
+library(afex)
+dry.mixed <- mixed(log(MC_dry) ~ hour*spcode + (1 | tray), data=mc, method="KR")
+dry.mixed.tab <- summary(dry.mod)$coefficients
+dry.mixed.anova.tab <- dry.mixed
+dry.mixed.tab
+dry.mixed.anova.tab
+
+
+dry.mixed.genus <- mixed(log(MC_dry) ~ hour*genus + (hour | spcode) + (1 | tray), data=mc, method="KR")
+
+
+dry.emm <- emmeans(dry.mod, "spcode")
+plot(dry.emm, comparisons=TRUE)
+
+dry.trends <- emtrends(dry.mod, "spcode", var="hour")
+plot(dry.trends, comparisons=TRUE)
+
+#emmip(dry.trends,  spcode ~  hour, cov.reduce= hour)
+
+
 # so there are three slope groups
 #cld(lsmeans(dry.mod, ~ spcode))
 # and 4 intercept groups (but slopes differ so . . .)
@@ -99,17 +114,23 @@ decomp.sum3 <- decomp[, c(2, 5:11)] %>% group_by(spcode, year) %>%
 
 newmctr <- merge(decomp.sum3[, c(1, 3:14)], newmc, by="spcode", sort=F)
 
+## newmctr_allrows <- right_join(decomp.sum3[, c(1, 3:14)], mc, by="spcode")
+## newmctr_allrows <- right_join(mc.sum, newmctr_allrows, by="spcode")
+
+
 ###########################################################
 # Establishing the influence of leaf traits on dry-down
[…]

### 2019-05-13 scripts/moisture-ms-figs-tables.R

@@ -1,4 +1,9 @@
-# moistsingles-plots
+# Figures for moisture manuscript
+# 2019-05-13
+
+
+library(xtable)
+library(ggrepel)
 
 RESULTS <- "../ms/figs_tables/"
 
@@ -40,11 +45,145 @@ ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
 
 ggsave(file.path(RESULTS, "fig1_drydown-curves_logged.pdf"), width=col1, height=col1, units="cm")
 
+# stats for fig 1
+tab1 <- nice(dry.mixed)
+names(tab1)[4] <- "p value"
+#tab1 <- tab1[ ,2:4]
+tab1 <- xtable(tab1)
+
+print(tab1, file=file.path(RESULTS, "drydown-tab.ltx"),  booktabs=TRUE, floating=FALSE,
+      include.rownames=FALSE)
+
+
+tabS1 <- summary(dry.mod)$coefficient
+print(xtable(tabS1), file=file.path(RESULTS, "drydown-coef-tab.ltx"),  booktabs=TRUE, floating=FALSE)
+
+tabS2 <- dry.emm
+print(xtable(tabS2), file=file.path(RESULTS, "drydown-emmeans.ltx"),  booktabs=TRUE, floating=FALSE,
+      include.rownames=FALSE)
+
+
+## Figure S1
+dry.mod <- lmer(log(MC_dry) ~ hour*display.name + (1 | tray), data=mc)
+dry.emm <- emmeans(dry.mod, "display.name")
+figS1 <- plot(dry.emm, comparisons=TRUE) +
+  xlab("Estimated marginal mean (log water content)") +
+  ylab("Species") +
+  pubtheme.nogridlines +
+  theme(axis.text.y = element_text(face = "italic"))
+figS1  
+ggsave(file.path(RESULTS, "figS1_max_water_emmeans.pdf"), width=col2, height=col1, units="cm")
+
+dry.trends <- emtrends(dry.mod, "display.name", var="hour")
+figS2 <- plot(dry.trends, comparisons=TRUE) +
+  xlab(expression(paste("Estimated marginal mean dessication rate (", hr^-1, ")"))) +
+  ylab("Species") +
+  pubtheme.nogridlines +
+  theme(axis.text.y = element_text(face = "italic"))
+figS2
[…]

### 2019-05-13 ms/SI.tex
Add supp mat file

%% SI
\documentclass[letterpaper]{article}

\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{fullpage}

\title{Magalhaes and Schwilk Supplementary Materials}
\author{Rita Quinones de Magalhaes and Dylan W. Schwilk}
%\date{\today}

% Hint: \title{what ever}, \author{who care} and \date{when ever} could stand 
% before or after the \begin{document} command 
% BUT the \maketitle command MUST come AFTER the \begin{document} command!

\begin{document}

\maketitle

\section{Tables}

\begin{table}[h]
  \caption{Mixed model coefficients for drydown curves. Table shows linear model results of model fit to  natural log of dry-mass based fuel moisture content as a function of time and species.} 
  \label{tabS1}
\centering
\input{figs_tables/drydown-coef-tab.ltx}
\end{table}

\section{Figures}

\begin{figure}[h]
  \centering
  \label{fig-S1}

  \includegraphics[width=17cm]{figs_tables/figS1_max_water_emmeans.pdf}
\caption{Estimated marginal means by species for maximum water retainability based on mixed model results.}

\end{figure}

\begin{figure}[h]
  \centering
  \label{fig-S2}
\includegraphics[width=17cm]{figs_tables/figS2_dessication_emmeans.pdf}
\caption{Estimated marginal means by species for dessication rate based on mixed model results.}
\end{figure}

\end{document}

### 2019-05-13 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Add di and maxmc trait models

@@
 % Full title of the paper (Capitalized)
-\Title{Moisture effects on leaf litter are species-specific and result in non-additive effects flammability in mixed conifer forest}
+\Title{Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest}
 
@@
 % Authors, for the paper (add full first names)
-\Author{Rita Quinones de Magalhaes $^{1,\dagger,\ddagger}$\orcidA{} and Dylan W. Schwilk $^{2,\ddagger}$}
+\Author{Rita Quinones de Magalhaes $^{\dagger}$ and Dylan W. Schwilk}
 
 % Authors, for metadata in PDF
-\AuthorNames{Rita Magalhaes, Dylan W. Schwilk}
+\AuthorNames{Rita Magalh\~{a}es, Dylan W. Schwilk}
 
 % Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
-\address{%
-$^{1}$ \quad Affiliation 1; e-mail@e-mail.com\\
-$^{2}$ \quad Texas Tech University; dylan.schwilk@ttu.edu}
+\address[1]{Texas Tech University}
 
@@
 % Current address and/or shared authorship
-\firstnote{Current address: Affiliation 3} 
-\secondnote{These authors contributed equally to this work.}
+\firstnote{Current address: Rochester Institute of Technology, Rochester, NY 14623} 
+%\secondnote{These authors contributed equally to this work.}
 % The commands \thirdnote{} till \eighthnote{} are available for further notes
@@
 
-We tested if litter bulk density influenced moisture dynamics (maximum water retainability and desiccation rate) by fitting linear models with species means as observations.
+We tested if one leaf trait (specific leaf area) and one litter trait (litter
+bulk density) influenced moisture dynamics (maximum water retainability and
+desiccation rate) by fitting linear models with species means as observations.
+For each response variable we fit a single linear model with two predictors and
+an interaction term.
 
@@
 
-\subsection{Leaf traits influence on moisture dynamics }
+\subsection{Species specific moisture dynamics and leaf traits}
 
@@
   \centering
-  \label{fig-drydown}
+  \label{fig1-drydown}
 \includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
-\caption{Dry down curves for eight litter types.
-Lines show best-fit exponential decay curves fit to moisture content on a dry mass basis.
+\caption[Dry down curves for eight litter types.]{Lines show best-fit exponential decay curves fit to moisture content on a dry mass basis.
 Litter samples we saturated by immersion for 24 hours and then allowed to drain by gravity for 3 min before initial weighing.
@@
 
-Across these eight species litter bulk density had a negative effect on desiccation rate (Fig. \ref{fig-bd-di}, linear model p = 0.011).
+Across these eight species specific leaf area had a positive effect on maximum water retention (Fig.
+\ref{maxmc}, linear model p = 0.003 and litter bulk density had a negative effect on desiccation rate (Fig.
+\ref{fig-bd-di}, linear model p = 0.029).
+Anova tables are shown in in Fig.
+\ref{tab-mc-anova} and Fig.
+\ref{tab-di-anova}.
+
+\begin{figure}[H]
+  \centering
+  \label{fig-maxmc-di}
+\includegraphics[width=16cm]{figs_tables/fig3-SLA-maxMC.pdf}
+\caption[Maximum water retention by specific leaf area]{Maximum water retention by specific leaf area (SLA) across eight species.  TODO }
+\end{figure}
+
+\begin{table}[H]
+  \caption{Linear model results for maximum moisture content as a function of specific leaf area and litter bulk density.}
+  \label{tab2-mc-anova}
+\centering
+%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
+\input{figs_tables/mc-anova.ltx}
+\end{table}
 
@@
   \label{fig-bd-di}
-\includegraphics[width=8cm]{figs_tables/fig3_di_by_bd.pdf}
-\caption{ }
+\includegraphics[width=16cm]{figs_tables/fig4_di_by_bd.pdf}
+\caption[Litter density effect on dessication rate.]{  TODO }
 \end{figure}
 
-\subsection{Time since wetting and its effect on flammability}
+\begin{table}[H]
+  \caption{Linear model results for estimated dessication rate ($hr^{-1}$) as a function of specific leaf area and litter bulk density.}
+  \label{tab2-di-anova}
+\centering
+%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
+\input{figs_tables/di-anova.ltx}
+\end{table}
+
+\subsection{Time since wetting effect on flammability}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Subsection}
-\unskip
-\subsubsection{Subsubsection}
-
-Bulleted lists look like this:
-\begin{itemize}[leftmargin=*,labelsep=5.8mm]
-\item	First bullet
-\item	Second bullet
-\item	Third bullet
-\end{itemize}
-
-Numbered lists can be added as follows:
-\begin{enumerate}[leftmargin=*,labelsep=4.9mm]
-\item	First item 
-\item	Second item
-\item	Third item
-\end{enumerate}
-
-The text continues here.
-
-\subsection{Figures, Tables and Schemes}
-
-All figures and tables should be cited in the main text as Figure 1, Table 1, etc.
-
-Text
-
-Text
-
 %\begin{listing}[H]
@@
 These results are influenced by the lower bulk densities of litter mixtures containing those two species.
-There was a non-additive effect of moisture on flammability for the four mixtures in this study, reinforcing results shown for dry litter \citep{Magalhaes+Schwilk-2012, VanAltena+Logtestjin+etal-2012, }.
+There was a non-additive effect of moisture on flammability for the four mixtures in this study, reinforcing results shown for dry litter \citep{Magalhaes+Schwilk-2012, VanAltena+Logtestjin+etal-2012}.
 Spread rate and time to ignition had significant non-additivity, and were strongly influenced by the species present in the mixture.
@@
 
-I have found non-additive effects (both positive and negative) in moisture content and in the flammability of wet litter mixtures of mixed-conifer forest.
+% I have found non-additive effects (both positive and negative) in moisture content and in the flammability of wet litter mixtures of mixed-conifer forest.
 These results complement the efforts of Blauw et al.

### 2019-05-13 ms/SI.tex
Add di and maxmc trait models

@@
   \includegraphics[width=17cm]{figs_tables/figS1_max_water_emmeans.pdf}
-\caption{Estimated marginal means by species for maximum water retainability based on mixed model results.}
+\caption{Estimated marginal means by species for maximum water retainability based on mixed model results. Shaded regions indicate 95\% confidence intervals for the marginal means.}
 
@@
 \includegraphics[width=17cm]{figs_tables/figS2_dessication_emmeans.pdf}
-\caption{Estimated marginal means by species for dessication rate based on mixed model results.}
+\caption{Estimated marginal means by species for dessication rate based on mixed model results. Shaded regions indicate 95\% confidence intervals for the marginal means.}
 \end{figure}

### 2019-05-13 scripts/dry-down.R

@@ -95,13 +95,14 @@ coefunc <- function(d){
     return(data.frame(logmaxMC = res[1,1], logmaxMC.se = res[1,2],  di= res[2,1], di.se = res[2,2]))
 }
 
+mc.sum <- mc %>% group_by(spcode) %>% 
+      summarise(bd.mean=mean(bd), bd.sd=sd(bd)) 
+
 mcdis <- mc %>% group_by(spcode) %>% do(coefunc(.)) %>% mutate(maxMC = exp(logmaxMC), maxMC.se=exp(logmaxMC.se))
 
-mc.sum <- mc %>% group_by(spcode) %>% 
-      summarise(MC_dry.mean=mean(MC_dry),
-                MC_dry.sd=sd(MC_dry),
-                bd.mean=mean(bd),
-                bd.sd=sd(bd)) 
+mc.sum <- left_join(mc.sum, mcdis)
+
+
   
 newmc <- merge(mc.sum, mcdis, by="spcode")
 
@@ -124,8 +125,8 @@ newmctr <- merge(decomp.sum3[, c(1, 3:14)], newmc, by="spcode", sort=F)
 
 # models with eight observations (species means)
 
-modmaxMCbulk <- lm(maxMC ~ bd.mean , data=newmctr)
-summary(modmaxMCbulk)
+#modmaxMCbulk <- lm(maxMC ~ bd.mean , data=newmctr)
+#summary(modmaxMCbulk)
 
 ## modmaxMClt <- lmer(MC_dry.mean ~ bd * lt_mean + (1 | spcode), data=newmctr_allrows)
 ## summary(modmaxMClt)
@@ -134,10 +135,10 @@ summary(modmaxMCbulk)
 
 #anova(modmaxMCbulk, modmaxMClt)
 
-moddibulk <- lm(di ~ bd.mean, data=newmctr)
-summary(moddibulk)
+#moddibulk <- lm(di ~ bd.mean, data=newmctr)
+#summary(moddibulk)
 
-moddilt <- lm(di ~ lt_mean + bd.mean, data=newmctr)
-summary(moddilt)
+#moddilt <- lm(di ~ lt_mean + bd.mean, data=newmctr)
+#summary(moddilt)
 
-anova(moddibulk, moddilt)
+#anova(moddibulk, moddilt)

### 2019-05-13 scripts/moisture-ms-figs-tables.R

@@ -31,21 +31,21 @@ ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
 
 ggsave(file.path(RESULTS, "fig1_drydown-curves.pdf"), width=col1, height=col1, units="cm")
 
-ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
-  geom_point(size=1.2, alpha=0.7, stroke=0) +
-  geom_smooth(method="lm", se=FALSE, size =0.6) +
-  scale_colour_manual(values=schwilkcolors) +
-  xlab("Hours since wetting") + ylab("Moisture by dry weight (%)") +
-  scale_x_continuous(breaks=xbreaks) +
-  scale_y_log10() +
-  pubtheme.nogridlines +
-  theme(legend.position=c(0.75, 0.86),
-        legend.title=element_blank(),
-        legend.text = element_text(family=fontfamily, size=smsize, face="italic"))
+## ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
+##   geom_point(size=1.2, alpha=0.7, stroke=0) +
+##   geom_smooth(method="lm", se=FALSE, size =0.6) +
+##   scale_colour_manual(values=schwilkcolors) +
+##   xlab("Hours since wetting") + ylab("Moisture by dry weight (%)") +
+##   scale_x_continuous(breaks=xbreaks) +
+##   scale_y_log10() +
+##   pubtheme.nogridlines +
+##   theme(legend.position=c(0.75, 0.86),
+##         legend.title=element_blank(),
+##         legend.text = element_text(family=fontfamily, size=smsize, face="italic"))
 
-ggsave(file.path(RESULTS, "fig1_drydown-curves_logged.pdf"), width=col1, height=col1, units="cm")
+## ggsave(file.path(RESULTS, "fig1_drydown-curves_logged.pdf"), width=col1, height=col1, units="cm")
 
-# stats for fig 1
+## # stats for fig 1
 tab1 <- nice(dry.mixed)
 names(tab1)[4] <- "p value"
 #tab1 <- tab1[ ,2:4]
@@ -54,7 +54,7 @@ tab1 <- xtable(tab1)
 print(tab1, file=file.path(RESULTS, "drydown-tab.ltx"),  booktabs=TRUE, floating=FALSE,
       include.rownames=FALSE)
 
-
[…]

### 2019-05-13 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
edits

@@
 
-The dry-down curves indicated that the eight species varied in their moisture behavior [FIG 2?].
-To determine a subset of species to use in the mixture trials, we performed analysis of contrasts using lsmeans package in R  (TODO Lenth 2016) to determine similarity between the species, which fell into four groups.
-I then selected one species from each group to include in the mixtures resulting in four distinct 3-species mixtures.
+% The dry-down curves indicated that the eight species varied in their moisture behavior (Fig \ref{fig-drydown}).
+To determine a subset of species to use in the mixture trials, we performed analysis of contrasts using lsmeans package in R (TODO Lenth 2016) to guide choosing a subset of species for the litter mixtures.
+Based on the marginal means analysis, we chose four species to maximize variation in water retention and drying rate.
+Creating all possible three species mixtures from these four species resulted in four distinct three-species mixtures.
 
- Leaf trait data
+% Leaf trait data
 
-We investigated the influence of leaf thinness and litter bulk density on the moisture dynamics of the species in this study.
-Leaf thinness data (leaf length over thickness) was obtained for subsamples of the litter collected for this project.
-Bulk density of the litter was obtained for each basket during the dry-down experiment by dividing the dry mass (450 g) by the estimated litter volume (basket width x length x litter depth).
+With only eight species in this study, our power to test leaf trait effects on the observed moisture dynamics is low, so we decided a priori to test the effect of one leaf trait likely to influence moisture absoorption, specific leaf area, and one litter trait known to directly effect flammability and likely to influence drying rate, litter bulk density.
+Specific leaf area was taken from leaf fresh leaf area and leaf dry mass from X individuals per species and was collected as part of an ealier study in 2011 (CITE).
+Litter bulk density was obtained in this  experiment for each basket by dividing the dry mass (450 g) by the estimated litter volume (basket width x length x litter depth).
 
@@
 
-The effect of time since wetting on flammability was modelled using linear regression with the interaction between time since wetting and genus as fixed effects and average vapor pressure deficit in the burning room during that trial as a nuisance covariate.
+The effect of time since wetting on flammability parameters was modelled using linear regression with the interaction between time since wetting and genus as fixed effects and average vapor pressure deficit in the burning room during that trial as a nuisance covariate.
 Differences across species were tested with post-hoc Tukey tests.

### 2019-05-16 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Edits to ms

@@
 
-We performed flammability tests across multiple moisture levels on monospecific litter beds and on mixed litter beds with eaach species or mixture replicated 5 times at each moisture level.
+We performed flammability tests across multiple moisture levels on monospecific litter beds and on mixed litter beds with each species or mixture replicated 5 times at each moisture level.
 Flammability was assessed using a 150 cm long burn table, 15 cm wide and 15 cm tall, in which leaf litter was placed and the table gently shaken to allow settling.
@@
 
-Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential decay coefficient.
-To estimate these c oefficents we fit a linear mixed-effects model by first taking the natural logarithm of moisture content, with time since wetting and species as fixed effects nested within replicate basket.
-We extracted the coefficients of the fitted model to describe maximum water retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate ($\lambda$) for each species.
-All analyses were conducted in R \cite{RCoreTeam-2019}.
-We fit models with R packages `lme4' \cite{Bates_Machler_etal-2015} and `afex' \cite{Singmann_Bolker_etal-2017}.
-Approximate degrees of freedom and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997} which is recommended by Luke \cite{Luke-2017} as producing acceptable Type 1 error rates.
+Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
+decay coefficient. To estimate these c oefficents we fit a linear mixed-effects
+model by first taking the natural logarithm of moisture content, with time
+since wetting and species as fixed effects nested within replicate basket. We
+extracted the coefficients of the fitted model to describe maximum water
+retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
+($\lambda$) for each species. All analyses were conducted in R
+\cite{RCoreTeam-2019}. We fit models with R packages `lme4'
+\cite{Bates_Machler_etal-2015} and `afex' \cite{Singmann_Bolker_etal-2017}.
+Approximate degrees of freedom and p-values were calculated by the
+Kenward-Roger approximation \cite{Kenward_Roger-1997} which is recommended by
+Luke \cite{Luke-2017} as producing acceptable Type 1 error rates.
 
@@
 
-The effect of time since wetting on flammability parameters was modelled using linear regression with the interaction between time since wetting and genus as fixed effects and average vapor pressure deficit in the burning room during that trial as a nuisance covariate.
-Differences across species were tested with post-hoc Tukey tests.
-A binomial logistic regression was fitted to the probability of ignition to obtain moisture of extinction.
+Our past work \cite{Magalhaes+Schwilk-2012} demonstrated that dry litter
+flammability across these eight species was driven by leaf traits and litter
+packing and therefore species within a genus tended to behave similarly with
+the two scale-leaved genera in the Cupressaceae, \emph{Calocedrus} and
+\emph{Sequoiadendron} also behaving similarly. Therefore, we used taxon
+(\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as a fixed in
+our models of moisture effects on flammability parameters. The effect of time
+since wetting on flammability parameters was modelled using linear regression
+with the interaction between time since wetting and genus as fixed effects and
+average vapor pressure deficit in the burning room during that trial as a
+nuisance covariate. Differences across species were tested with post-hoc Tukey
+tests. A binomial logistic regression was fitted to the probability of ignition
+to obtain moisture of extinction.
 
-We investigated non-additive effects on moisture and on flammability by testing if moisture content and flammability of mixtures were predicted by the average of the constituent species.
-To choose mixtures we first grouped the species according to their maximum water retainability and dessication rate according to estimated marginal means by species using the R package `emmeans' \cite{Lenth-2019}.
-This resulted in four broad groups of species and we then selected one representative species from each litter type group and produced all four possible three-species litter mixtures using those four species.
-For each mixture and time since wetting, we predicted moisture content and flame spread rate using the models fit to single species data described above and then calculated an expected moisture content and spread rate as the average of these three values for the mixture.
-This procedure was done for both moisture content and flammability parameter values of the constituent species.
-The expectation from the null model was that the difference between observed and predicted values of flammability will be zero.
-We tested if there were significant departures from the null (zero) with a mixed effects linear model.
+We investigated non-additive effects on moisture and on flammability by testing
+if moisture content and flammability of mixtures were predicted by the average
+of the constituent species. To choose mixtures we first grouped the species
+according to their maximum water retainability and dessication rate according
+to estimated marginal means by species using the R package `emmeans'
+\cite{Lenth-2019}. This resulted in four broad groups of species and we then
+selected one representative species from each litter type group and produced
+all four possible three-species litter mixtures using those four species. For
+each mixture and time since wetting, we predicted moisture content and flame
+spread rate using the models fit to single species data described above and
+then calculated an expected moisture content and spread rate as the average of
+these three values for the mixture. This procedure was done for both moisture
+content and flammability parameter values of the constituent species. The
+expectation from the null model was that the difference between observed and
+predicted values of flammability will be zero. We tested if there were
+significant departures from the null (zero) with a mixed effects linear model.
 
@@
 There are marked differences in the extent of moisture species retain initially following wetting and draining by gravity (all p < 0.0001, Fig.
-\ref{fig1-drydown} and Table \ref{tab1-drydown}).
+\ref{fig:drydown} and Table \ref{tab:drydown}).
 \emph{Q.
-kelloggii}, the only broadleaf, is the species that retains the most moisture but has had a rapid dessication rate (.
+kelloggii}, the only broadleaf, is the species that retains the most moisture initially.
+\emph{Quercus}, however, like the pines has a rapid dessication rate (Fig.
+\ref{fig:drydown-logged}).
 Calculating estimated marginal means following the linear mixed-effects model discriminated the species into three different maximum water retention groups: 1) \emph{Q.
@@
 jeffreyi} and \emph{P.
-ponderosa}.
+ponderosa} (SI 1).
 There are also differences in desiccation rates across the eight species and estimated mixed linear model slopes discriminated species into three groups from lowest to most rapid: 1) (\emph{C.decurrens}, 2) \emph{Abies spp} and \emph{S.
 gianteum}, and 3) \emph{Q.
-kelloggii} and \emph{Pinus spp}.
+kelloggii} and \emph{Pinus spp} (SI 2).
 
@@
   \centering
-  \label{fig1-drydown}
 \includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
@@
 Colors indicate genera and there were significant differences in both }
+ \label{fig:drydown}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/fig2_drydown-curves_logged.pdf}
+\caption[Semi-log scale dry down curves for eight litter types.]{Lines show best-fit lines fit to the natural log of moisture content on a dry mass basis.
+The slope of the line shows the dessication rate ($hr^{-1}$).
+Litter samples we saturated by immersion for 24 hours and then allowed to drain by gravity for 3 min before initial weighing.
+Colors indicate genera and there were significant differences in both }
+ \label{fig:drydown-logged}
 \end{figure}
@@
 Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
-  \label{tab1-drydown}
+  \label{tab:drydown}
 \centering
@@
 Across these eight species specific leaf area had a positive effect on maximum water retention (Fig.
-\ref{maxmc}, linear model p = 0.003 and litter bulk density had a negative effect on desiccation rate (Fig.
-\ref{fig-bd-di}, linear model p = 0.029).
+\ref{fig:maxmc-di}, linear model p = 0.003) and litter bulk density had a negative effect on desiccation rate (Fig.
+\ref{fig:bd-di}, linear model p = 0.029).
 Anova tables are shown in in Fig.
 \ref{tab-mc-anova} and Fig.
-\ref{tab-di-anova}.
+\ref{tab:di-anova}.
 
@@
   \centering
-  \label{fig-maxmc-di}
 \includegraphics[width=16cm]{figs_tables/fig3-SLA-maxMC.pdf}
 \caption[Maximum water retention by specific leaf area]{Maximum water retention by specific leaf area (SLA) across eight species.  TODO }
+  \label{fig:maxmc-di}
 \end{figure}
@@
   \caption{Linear model results for maximum moisture content as a function of specific leaf area and litter bulk density.}
-  \label{tab2-mc-anova}
+  \label{tab2:mc-anova}
 \centering
@@
   \centering
-  \label{fig-bd-di}
 \includegraphics[width=16cm]{figs_tables/fig4_di_by_bd.pdf}
 \caption[Litter density effect on dessication rate.]{  TODO }
+  \label{fig:bd-di}
 \end{figure}
@@
   \caption{Linear model results for estimated dessication rate ($hr^{-1}$) as a function of specific leaf area and litter bulk density.}
-  \label{tab2-di-anova}
+  \label{tab2:di-anova}
 \centering
@@
 
-Time to ignition and spread rate were modeled against time since wetting and using genus as a fixed effect.
+Time to ignition and spread rate were modeled against time since wetting and using genus as a fixed effect and species as a random effect.
+Time since wetting had a strong effect on spread rate (p < 0.0001) and the slope of this effect varied by genus (Fig.
+\ref{fig:spreadrate}, Table \ref{tab:spreadrate-anova}).
+The pines were able to ignite and burn much sooner after wetting than the other genera, none of which had measurable spread rates until burning was attempted 72 hours after wetting.
+At that point, the \emph{Abies}, \emph{Calocedrus}, and \emph{Sequiadendron} had much drying litter than did the \emph{Quercus}, yet the spread rates were slow.
+\emph{Quercus} had the steepest response to time since wetting (Fig.
+\ref{fig5:spreadrate}).
 
-[TODO]?  redo in terms of groups.
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/fig5_spread_by_time.pdf}
+\caption[Flame spread rate by time since wetting]{TODO }
+  \label{fig:spreadrate}
+\end{figure}
 
-of time since wetting and genus on time to ignition and spread rate (F10, 171=24.51, p < 2.2e-16 and F10, 171=55.
-85, p<2.2e-16 respectively; Figures 2.3 and 2.4).
-There were also significant interactions of time since wetting and genus on the response variables (time to ignition and spread rate; both p < 2.2e-16).
-For time since ignition, the negative relationship is driven by Sequoiadendron and Calocedrus, as evidenced by the steeper slopes of the curves.
-However, if I removed Sequoiadendron and Calocedrus from the models, the interaction of time since wetting and genus disappears for time to ignition (p=0.1643) but is still significant for spread rate (p = 2.05e-6).
-After drying for 72 hours, Q.
-kelloggi is always more flammable and more responsive to changes in moisture content, while C.
-decurrens and S.
-giganteum are not (Figures 2.3 and 2.4).
+\begin{table}[H]
+  \caption[Mixed linear model of flame spread rate.]{Mixed model results for flame spread rate  as a function of time since wetting with genus as a fixed effect and species as a nested random effect.
+Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
+  \label{tab:spreadrate-anova}
+\centering
+%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
+\input{figs_tables/spreadrate-anova.ltx}
+\end{table}
 
@@
 The ANOVA results indicate that leaf thinness best explains variation in maximum water retention, while litter bulk density best explains variation in desiccation rates.
+
+NOTE: P lambertiana higher water absorption, presumably due to higher SLA.
 

### 2019-05-16 scripts/moisture-ms-figs-tables.R
Add flam figures

@@ -1,6 +1,5 @@
 # Figures for moisture manuscript
-# 2019-05-13
-
+# 2019-05-16
 
 library(xtable)
 library(ggrepel)
@@ -31,19 +30,19 @@ ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
 
 ggsave(file.path(RESULTS, "fig1_drydown-curves.pdf"), width=col1, height=col1, units="cm")
 
-## ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
-##   geom_point(size=1.2, alpha=0.7, stroke=0) +
-##   geom_smooth(method="lm", se=FALSE, size =0.6) +
-##   scale_colour_manual(values=schwilkcolors) +
-##   xlab("Hours since wetting") + ylab("Moisture by dry weight (%)") +
-##   scale_x_continuous(breaks=xbreaks) +
-##   scale_y_log10() +
-##   pubtheme.nogridlines +
-##   theme(legend.position=c(0.75, 0.86),
-##         legend.title=element_blank(),
-##         legend.text = element_text(family=fontfamily, size=smsize, face="italic"))
+ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
+  geom_point(size=1.2, alpha=0.7, stroke=0) +
+  geom_smooth(method="lm", se=FALSE, size =0.6) +
+  scale_colour_manual(values=schwilkcolors) +
+  xlab("Hours since wetting") + ylab("Moisture by dry weight (%)") +
+  scale_x_continuous(breaks=xbreaks) +
+  scale_y_log10() +
+  pubtheme.nogridlines +
+  theme(legend.position=c(0.75, 0.86),
+        legend.title=element_blank(),
+        legend.text = element_text(family=fontfamily, size=smsize, face="italic"))
 
-## ggsave(file.path(RESULTS, "fig1_drydown-curves_logged.pdf"), width=col1, height=col1, units="cm")
+ggsave(file.path(RESULTS, "fig2_drydown-curves_logged.pdf"), width=col1, height=col1, units="cm")
 
 ## # stats for fig 1
 tab1 <- nice(dry.mixed)
@@ -61,7 +60,6 @@ print(xtable(tabS1), file=file.path(RESULTS, "drydown-coef-tab.ltx"),  booktabs=
 tabS2 <- dry.emm
 print(xtable(tabS2), file=file.path(RESULTS, "drydown-emmeans.ltx"),  booktabs=TRUE, floating=FALSE,
[…]

### 2019-05-16 scripts/moisture-ms-figs-tables_full.R
Add file for full set of figures
Need to clean up file naming, this is fulls et of figures named by type,
etc. The moisture-ms-figs-tables.R file will then have only those
figures in the ms and si and will have figures and tables named by
number.

@@ -1,0 +1,376 @@
+# Figures for moisture manuscript
+# 2019-05-13
+
+
+library(xtable)
+library(ggrepel)
+
+RESULTS <- "../ms/figs_tables/"
+
+source("theme-opts.R")
+source("dry-down.R")
+source("burn-moist.R")
+
+## Figure 1: Plotting dry-down
+xbreaks <- seq(0, 144, 24)
+ybreaks <- seq(0, 700, 50)
+
+ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
+  geom_point(size=1.2, alpha=0.7, stroke=0) +
+  geom_smooth(method="glm",
+              method.args=list(family=gaussian(link="log")), se=FALSE, size =0.6) +
+  scale_colour_manual(values=schwilkcolors) +
+  xlab("Hours since wetting") + ylab("Moisture by dry weight (%)") +
+  scale_x_continuous(breaks=xbreaks) +
+  scale_y_continuous(breaks=ybreaks) +
+  pubtheme.nogridlines +
+  theme(legend.position=c(0.75, 0.86),
+        legend.title=element_blank(),
+        legend.text = element_text(family=fontfamily, size=smsize, face="italic"))
+ #       legend.key.height=unit(smsize,"pt"))
+
+ggsave(file.path(RESULTS, "fig1_drydown-curves.pdf"), width=col1, height=col1, units="cm")
+
+ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
+  geom_point(size=1.2, alpha=0.7, stroke=0) +
+  geom_smooth(method="lm", se=FALSE, size =0.6) +
+  scale_colour_manual(values=schwilkcolors) +
+  xlab("Hours since wetting") + ylab("Moisture by dry weight (%)") +
+  scale_x_continuous(breaks=xbreaks) +
+  scale_y_log10() +
+  pubtheme.nogridlines +
+  theme(legend.position=c(0.75, 0.86),
+        legend.title=element_blank(),
+        legend.text = element_text(family=fontfamily, size=smsize, face="italic"))
+
+ggsave(file.path(RESULTS, "fig2_drydown-curves_logged.pdf"), width=col1, height=col1, units="cm")
+
+## # stats for fig 1
+tab1 <- nice(dry.mixed)
+names(tab1)[4] <- "p value"
+#tab1 <- tab1[ ,2:4]
+tab1 <- xtable(tab1)
+
[…]

### 2019-05-17 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Getting there

@@
 With only eight species in this study, our power to test leaf trait effects on the observed moisture dynamics is low, so we decided a priori to test the effect of one leaf trait likely to influence moisture absoorption, specific leaf area, and one litter trait known to directly effect flammability and likely to influence drying rate, litter bulk density.
-Specific leaf area was taken from leaf fresh leaf area and leaf dry mass from X individuals per species and was collected as part of an ealier study in 2011 (CITE).
+Specific leaf area was taken from leaf fresh leaf area and leaf dry mass from X individuals per species and was collected as part of an ealier study \cite{Magalhaes+Schwilk-2012}.
 Litter bulk density was obtained in this experiment for each basket by dividing the dry mass (450 g) by the estimated litter volume (basket width x length x litter depth).
@@
 
-We performed flammability tests across multiple moisture levels on monospecific litter beds and on mixed litter beds with each species or mixture replicated 5 times at each moisture level.
-Flammability was assessed using a 150 cm long burn table, 15 cm wide and 15 cm tall, in which leaf litter was placed and the table gently shaken to allow settling.
-This design allows fire behaviour to reach constant flame spread rates \cite{Magalhaes+Schwilk-2012}.
-Two graduated metal rulers   equally spaced along the apparatus allowed visual determination of maximum flame height.
-Litter was ignited with a propane torch; maximum adiabatic flame temperature of 1899 C, http://www.benzomatic.com/products/fuel.aspx) at one end of the apparatus and allowed to burn until extinction.
-If a species at a specific moisture level failed to ignite after three attempts, no further attempts were made.
-A timer was used to record time to ignition and duration of flaming combustion.
-The flammability trials followed the methodology detailed in de Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.
+We performed flammability tests across multiple moisture levels on monospecific
+litter beds and on mixed litter beds with each species or mixture replicated 5
+times at each moisture level. We used the initial dry-down experiment to guide
+our timeing so as to attempt to burn samples over the full range of moisture
+contents from 100\% to below 10\% moisture. To expand our data on the dry end
+of this gradient, we supplemented these data with data from an earlier
+experiment on well-dried litter \cite{Magalhaes+Schwilk-2012}. This
+supplemental data added five replicates per species at an oven dried moisture
+content of $<$ 5\%. We assigned these replicates a time since wetting value of
+144 hours equivalent to the longest time since drying for the wetted samples.
+Flammability was assessed using a 150 cm long burn table, 15 cm wide and 15 cm
+tall, in which leaf litter was placed and the table gently shaken to allow
+settling. This design allows fire to reach constant flame spread rates
+\cite{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced along
+the apparatus allowed visual determination of maximum flame height. Litter was
+ignited with a propane torch; maximum adiabatic flame temperature of 1899 C,
+http://www.benzomatic.com/products/fuel.aspx) at one end of the apparatus and
+allowed to burn until extinction. If a species at a specific moisture level
+failed to ignite after three attempts, no further attempts were made. A timer
+was used to record time to ignition and duration of flaming combustion. The
+flammability trials followed the methodology detailed in de Magalhães and
+Schwilk \cite{Magalhaes+Schwilk-2012}.
 
@@
 
-During flammability trials, we recorded time to ignition (s), duration of flaming combustion (s), spread rate (mm/s), calculated by dividing the length of the burned surface by the time it takes the flaming front to reach the end of the apparatus, maximal flame height (mm), and mass loss (\%), calculated as the difference between fuel mass before and after flame extinction.
-I also measured duration above 100 C, or the length of time temperature exceeded 100 C, and temperature integration, which integrates duration of combustion and temperature above 100 C and serves as a proxy for heat release (de Magalhães and Schwilk 2012, Cornwell et al.
-2015).
-By performing the flammability tests at several moisture levels I were able to determine moisture of extinction (a threshold representing the moisture value above which litter will not ignite or spread) for all species in this study, using a logistic regression.
+During flammability trials, we recorded time to ignition (s), duration of
+flaming combustion (s), spread rate (mm/s), calculated by dividing the length
+of the burned surface by the time it takes the flaming front to reach the end
+of the apparatus, maximal flame height (mm), and mass loss (\%), calculated as
+the difference between fuel mass before and after flame extinction. Our
+thermocouple system failed or data was lost for two days of burning trials,
+therefore we have omitted those data from this analysis. Our past work
+\cite{Magalhaes+Schwilk-2012} demonstrated that flame spread rate captures one
+important axis of flammability variation \citep{Schwilk-2015,
+  Pausas+Keeley+etal-2017} and percent fuel consumed captures a portion of the
+second major axis of flammability variation (although not as well as
+integrating temperature over time does). Our analysis, therefore, focused on
+these two relatively orthogonal measures of flammability.
 
-Burn trials occurred in a cement structure used to simulate house fires at the Fire Department of the City of Lubbock.This structure minimizes wind and helps regulate temperature and relative humidity.
-Temperature and relative humidity were measured before every trial using a Kestrel 3000 (Nielsen-Kellerman, Boothwyn, PA).
-To minimize temperature and relative humidity variation, all trials were conducted on clear days between 10 am and 3 pm.
+Burn trials occurred in a cement structure used to simulate house fires at the
+Fire Department of the City of Lubbock. This structure minimizes wind and helps
+regulate temperature and relative humidity. Temperature and relative humidity
+were measured before every trial using a Kestrel 3000 (Nielsen-Kellerman,
+Boothwyn, PA). To minimize temperature and relative humidity variation, all
+trials were conducted on clear days between 10 am and 3 pm.
 
@@
 flammability across these eight species was driven by leaf traits and litter
-packing and therefore species within a genus tended to behave similarly with
-the two scale-leaved genera in the Cupressaceae, \emph{Calocedrus} and
-\emph{Sequoiadendron} also behaving similarly. Therefore, we used taxon
-(\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as a fixed in
-our models of moisture effects on flammability parameters. The effect of time
-since wetting on flammability parameters was modelled using linear regression
-with the interaction between time since wetting and genus as fixed effects and
+packing. Due to similar leaf traits, species within a genus tended to behave
+similarly and the two scale-leaved genera in the Cupressaceae,
+\emph{Calocedrus} and \emph{Sequoiadendron} also behaved similarly to one
+another. Therefore, we used taxonomic groups (\emph{Abies}, Cupressaceae,
+\emph{Quercus}, and \emph{Pinus}) as our flammability functional groups and
+treated these as fixed effects in our models. The effect of fuel moisture and
+time since wetting on flammability parameters were modeled using mixed effect
+models with fuel moisture or time since wetting and taxon as fixed effects and
 average vapor pressure deficit in the burning room during that trial as a
-nuisance covariate. Differences across species were tested with post-hoc Tukey
-tests. A binomial logistic regression was fitted to the probability of ignition
-to obtain moisture of extinction.
+nuisance covariate.
 
-We investigated non-additive effects on moisture and on flammability by testing
-if moisture content and flammability of mixtures were predicted by the average
-of the constituent species. To choose mixtures we first grouped the species
-according to their maximum water retainability and dessication rate according
-to estimated marginal means by species using the R package `emmeans'
+We investigated possible non-additive effects on moisture and on flammability
+by testing if moisture content and flammability of mixtures were predicted by
+the average of the constituent species. To choose mixtures we first grouped the
+species according to their maximum water retainability and dessication rate
+according to estimated marginal means by species using the R package `emmeans'
 \cite{Lenth-2019}. This resulted in four broad groups of species and we then
@@
 
-There are marked differences in the extent of moisture species retain initially following wetting and draining by gravity (all p < 0.0001, Fig.
-\ref{fig:drydown} and Table \ref{tab:drydown}).
-\emph{Q.
-kelloggii}, the only broadleaf, is the species that retains the most moisture initially.
+There are marked differences in the extent of moisture species retain initially
+following wetting and draining by gravity (all p < 0.0001, Fig.
+\ref{fig:drydown} and Table \ref{tab:drydown}). \emph{Q. kelloggii}, the only
+broadleaf, is the species that retains the most moisture initially.
 \emph{Quercus}, however, like the pines has a rapid dessication rate (Fig.
-\ref{fig:drydown-logged}).
-Calculating estimated marginal means following the linear mixed-effects model discriminated the species into three different maximum water retention groups: 1) \emph{Q.
-kelloggii}; 2) \emph{P.
-lambertiana} and \emph{C.
-decurrens}; and 3) \emph{A.
-magnifica}, \emph{S.
-giganteum}, \emph{A.
-concolor}, \emph{P.
-jeffreyi} and \emph{P.
-ponderosa} (SI 1).
-There are also differences in desiccation rates across the eight species and estimated mixed linear model slopes discriminated species into three groups from lowest to most rapid: 1) (\emph{C.decurrens}, 2) \emph{Abies spp} and \emph{S.
-gianteum}, and 3) \emph{Q.
-kelloggii} and \emph{Pinus spp} (SI 2).
+\ref{fig:drydown-logged}). Calculating estimated marginal means following the
+linear mixed-effects model discriminated the species into three different
+maximum water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P.
+  lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S.
+  giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa}
+(SI Fig. 1). There are also differences in desiccation rates across the eight
+species and estimated mixed linear model slopes discriminated species into
+three groups from lowest to most rapid: 1) (\emph{C.decurrens}, 2) \emph{Abies
+  spp} and \emph{S. gianteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp}
+(SI Fig. 2).
 
@@
 \includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
-\caption[Dry down curves for eight litter types.]{Lines show best-fit exponential decay curves fit to moisture content on a dry mass basis.
-Litter samples we saturated by immersion for 24 hours and then allowed to drain by gravity for 3 min before initial weighing.
-Colors indicate genera and there were significant differences in both }
+\caption{Dry down curves for eight litter types. Lines show best-fit
+  exponential decay curves fit to moisture content on a dry mass basis. Litter
+  samples we saturated by immersion for 24 hours and then allowed to drain by
+  gravity for 3 min before initial weighing. Colors indicate genera and there
+  were significant differences in both }
  \label{fig:drydown}
@@
 
-\begin{figure}[H]
-  \centering
-\includegraphics[width=8cm]{figs_tables/fig2_drydown-curves_logged.pdf}
-\caption[Semi-log scale dry down curves for eight litter types.]{Lines show best-fit lines fit to the natural log of moisture content on a dry mass basis.
-The slope of the line shows the dessication rate ($hr^{-1}$).
-Litter samples we saturated by immersion for 24 hours and then allowed to drain by gravity for 3 min before initial weighing.
-Colors indicate genera and there were significant differences in both }
- \label{fig:drydown-logged}
-\end{figure}
-
 \begin{table}[H]
-  \caption{Mixed model results for moisture content as a function of time since wetting.
-Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
+  \caption{Mixed model results for moisture content as a function of time since
+    wetting. Approximate degrees of freedom, pseudo F statistics and p-values
+    were calculated by the Kenward-Roger approximation
+    \cite{Kenward_Roger-1997}.}
   \label{tab:drydown}
@@
 %% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-\input{figs_tables/drydown-tab.ltx}
+\input{figs_tables/tab1_drydown_anova.ltx}
 \end{table}
 
-Across these eight species specific leaf area had a positive effect on maximum water retention (Fig.
-\ref{fig:maxmc-di}, linear model p = 0.003) and litter bulk density had a negative effect on desiccation rate (Fig.
-\ref{fig:bd-di}, linear model p = 0.029).
-Anova tables are shown in in Fig.
-\ref{tab-mc-anova} and Fig.
-\ref{tab:di-anova}.
+Across these eight species specific leaf area had a positive effect on maximum
+water retention (Fig. \ref{fig:maxmc-di}, linear model p = 0.003) and litter
+bulk density had a negative effect on desiccation rate (Fig. \ref{fig:bd-di},
+linear model p = 0.029). Anova tables are shown in in Fig. \ref{tab-mc-anova}
+and Fig. \ref{tab:di-anova}.
 
@@
   \centering
-\includegraphics[width=16cm]{figs_tables/fig3-SLA-maxMC.pdf}
-\caption[Maximum water retention by specific leaf area]{Maximum water retention by specific leaf area (SLA) across eight species.  TODO }
+\includegraphics[width=16cm]{figs_tables/fig2_SLA_maxMC.pdf}
+\caption{Maximum water retention by specific leaf area (SLA) across eight species.}
   \label{fig:maxmc-di}
@@
 %% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-\input{figs_tables/mc-anova.ltx}
+\input{figs_tables/tab_mc_anova.ltx}
 \end{table}
@@
   \centering
-\includegraphics[width=16cm]{figs_tables/fig4_di_by_bd.pdf}
-\caption[Litter density effect on dessication rate.]{  TODO }
+\includegraphics[width=16cm]{figs_tables/fig3_di_bd.pdf}
+\caption{Linear model predicting dessication rate as a function of specific leaf area and litter bulk density across eight species.}
   \label{fig:bd-di}
 \end{figure}
+
+\begin{table}[H]
+  \caption{Linear model results for estimated maximum water retention (\%) as a function of specific leaf area and litter bulk density.}
+  \label{tab2:mc-anova}
+\centering
+%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
+\input{figs_tables/tab2_mc_anova.ltx}
+\end{table}
 
@@
 %% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-\input{figs_tables/di-anova.ltx}
+\input{figs_tables/tab3_di_anova.ltx}
 \end{table}
 
-\subsection{Time since wetting effect on flammability}
+\subsection{Moisture effects on flammability}
 
-Time to ignition and spread rate were modeled against time since wetting and using genus as a fixed effect and species as a random effect.
-Time since wetting had a strong effect on spread rate (p < 0.0001) and the slope of this effect varied by genus (Fig.
+Flame spread rate declined with fuel moisture (Fig.
+\ref{fig:spread-moist}.
+\emph{Quercus} had higher spread rates across all fuel moistures and had a steeper decline in response to moisture than did the other groups (taxon by time since wetting interaction p = 0.022, Table \ref{tab:spread-moist-anova}).
+Fuel moisture negatively influenced fuel consumption and overall  fuel consumption as well as the rate of change with moisture content varied across the taxonomic group (Fig.
+\ref{fig:consume-moist}, Table \ref{tab:consume-moist-anova}).
+Although the overall ranking of groups was simlar for fuel consumption and for flame spread rate, in contrast to the pattern with spread rate, for fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in flammability with fuel moisture than did the other groups.
+
+%% spread rate by moisture
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/fig4_spread_actualMC.pdf}
+\caption{Flame spread rate by moisture content (model fit F= 61.15, d.f = 9, 148,  p < 0.0001). Model ANOVA table in Table \ref{tab:spread-moist-anova} and coefficients in SI Table 2.
+  \label{fig:spread-moist}
+\end{figure}
+
+\begin{table}[H]
+  \caption{Mixed model results for flame spread rate as a function of fuel moisture with taxonomic group as a fixed effect and species as a nested random effect.
+Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.
+Coefficient estimates are in SI Table 2.}
+  \label{tab:spreadrate-anova}
+\centering
+\input{figs_tables/tab3_spread_moist_anova.ltx}
+\end{table}
+
+%% fuel consumption by moisture:
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/fig5_consume_actualMC.pdf}
+\caption{Percent fuel consumed by moisture content. Model ANOVA table in Table \ref{tab:consume-moist-anova} and coefficients in SI Table 3.
+  \label{fig:consume-moist}
+\end{figure}
+
+\begin{table}[H]
+  \caption{Mixed model results for fuel consumption as a function of fuel moisture with taxonomic group as a fixed effect and species as a nested random effect.
+Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.
+Coefficient estimates are in SI Table 3.}
+  \label{tab:consume-moist-anova}
+\centering
+\input{figs_tables/tab4_consume_moist_anova.ltx}
+\end{table}
+
+As a consequnece of the relationships shown above, time since wetting had a strong effect on spread rate (p < 0.0001) and the slope of this effect varied by taxonimc (Fig.
 \ref{fig:spreadrate}, Table \ref{tab:spreadrate-anova}).
@@
 \ref{fig5:spreadrate}).
-
-\begin{figure}[H]
-  \centering
-\includegraphics[width=8cm]{figs_tables/fig5_spread_by_time.pdf}
-\caption[Flame spread rate by time since wetting]{TODO }
-  \label{fig:spreadrate}
-\end{figure}
 

### 2019-05-17 ms/SI.tex
Edits to ms
Getting there

@@
 \centering
-\input{figs_tables/drydown-coef-tab.ltx}
+\input{figs_tables/SI_tab1_drydown_coef.ltx}
+\end{table}
+
+\begin{table}[h]
+  \caption{Mixed model coefficients for flame spread rate. Table shows mixed linear model results  TODO.} 
+  \label{tabS3}
+\centering
+\input{figs_tables/spreadrate-coef-tab.ltx}
 \end{table}
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
-  \label{fig-S1}
-
-  \includegraphics[width=17cm]{figs_tables/figS1_max_water_emmeans.pdf}
-\caption{Estimated marginal means by species for maximum water retainability based on mixed model results. Shaded regions indicate 95\% confidence intervals for the marginal means.}
+  \includegraphics[width=8cm]{figs_tables/SI_fig1_drydown_curves_logged.pdf}
+   \label{SI_fig1}
+\caption[Semi-log scale dry down curves for eight litter types.]{Lines show best-fit lines fit to the natural log of moisture content on a dry mass basis.
+The slope of the line shows the dessication rate ($hr^{-1}$).
+Litter samples we saturated by immersion for 24 hours and then allowed to drain by gravity for 3 min before initial weighing.
+Colors indicate genera and there were significant differences in both }
 
@@
   \centering
+  \label{SI_fig2}
+  \includegraphics[width=17cm]{figs_tables/SI_fig2_max_water_emmeans.pdf}
+\caption{Estimated marginal means by species for maximum water retainability based on mixed model results. Shaded regions indicate 95\% confidence intervals for the marginal means.}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
   \label{fig-S2}
-\includegraphics[width=17cm]{figs_tables/figS2_dessication_emmeans.pdf}
+\includegraphics[width=17cm]{figs_tables/SI_fig3_dessication_emmeans.pdf}
 \caption{Estimated marginal means by species for dessication rate based on mixed model results. Shaded regions indicate 95\% confidence intervals for the marginal means.}

### 2019-05-17 scripts/moisture-ms-figs-tables.R

@@ -10,11 +10,13 @@ source("theme-opts.R")
 source("dry-down.R")
 source("burn-moist.R")
 
-## Figure 1: Plotting dry-down
+###############################################################################
+## Figure 1: dry-down
+###############################################################################
 xbreaks <- seq(0, 144, 24)
 ybreaks <- seq(0, 700, 50)
 
-ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
+fig1 <- ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
   geom_point(size=1.2, alpha=0.7, stroke=0) +
   geom_smooth(method="glm",
               method.args=list(family=gaussian(link="log")), se=FALSE, size =0.6) +
@@ -26,11 +28,16 @@ ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
   theme(legend.position=c(0.75, 0.86),
         legend.title=element_blank(),
         legend.text = element_text(family=fontfamily, size=smsize, face="italic"))
- #       legend.key.height=unit(smsize,"pt"))
+#       legend.key.height=unit(smsize,"pt"))
 
-ggsave(file.path(RESULTS, "fig1_drydown-curves.pdf"), width=col1, height=col1, units="cm")
+ggsave(file.path(RESULTS, "fig1_drydown-curves.pdf"), plot=fig1,
+       width=col1, height=col1, units="cm")
 
-ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
+###############################################################################
+# SI fig 1: dry down on semi log scale
+###############################################################################
+
+SI_fig1 <- ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
   geom_point(size=1.2, alpha=0.7, stroke=0) +
   geom_smooth(method="lm", se=FALSE, size =0.6) +
   scale_colour_manual(values=schwilkcolors) +
@@ -42,48 +49,59 @@ ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
         legend.title=element_blank(),
         legend.text = element_text(family=fontfamily, size=smsize, face="italic"))
 
-ggsave(file.path(RESULTS, "fig2_drydown-curves_logged.pdf"), width=col1, height=col1, units="cm")
+ggsave(file.path(RESULTS, "SI_fig1_drydown-curves_logged.pdf"), plot=SI_fig1,
[…]

### 2019-05-20 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Add beta regression model for Fig 7
This should be final figure needed for monoculture flammability trials.

@@
   availability to burn. Natural forests are often comprised of mixed stands,
-  leading to mixtures of leaf litter on the surface fuelling the fires. Multiple
-  studies have established the existence of non-additive effects in the
-  flammability of dry litter, but we lack a mechanistic understanding of
+  leading to mixtures of leaf litter on the surface fuelling the fires.
+  Multiple studies have established the existence of non-additive effects in
+  the flammability of dry litter, but we lack a mechanistic understanding of
   moisture synergies in mixed litter and of moisture effects on the
@@
   single species and of mixtures, to establish flammability response to the
-  dry-down process. Moisture absorption is determined by traits pertaining to
-  leaf geometry while moisture desorption is influenced by emergent properties
-  of the litterbed. Species that produce more aerated litter beds (such as \emph{Q.
-  kelloggii} and \emph{Pinus} species) have faster rates of litter desiccation. My results
-  show a relationship between time since wetting and time to ignition and
-  spread rate, with significant interaction with genus. This indicates a
-  genus-specific behavior of flammability and time since wetting.
-  We found non-additive effects in litter mixtures for both moisture content and
-  flammability. Mixtures containing \emph{Q. kelloggii} and \emph{P. jeffreyi}, species that
-  dry faster, produced more flammable behavior (lower time to ignition, faster
-  spread rate, taller flames). Fire behavior of mixtures depended on moisture
-  content, although mixtures without Q. kelloggii had always higher moisture
-  content than predicted by the average fire behavior of the species in the
-  mixture. At lower moisture contents (<75\%) the mixtures exhibit positive
-  non-additive behavior and at higher moisture contents (> 200\%) they exhibit
-  negative non-additive behavior.}
+  dry-down process. Moisture dynamics vary across species with higher specific
+  leaf area being associated with higher maximum absorption and lower litter
+  bulk denisty being associated with faster drying rates. Species that produce
+  more aerated litter beds (such as \emph{Q. kelloggii} and \emph{Pinus}
+  species) have faster rates of litter desiccation. These differing moisture
+  dynamics result in time since wetting having strongly differing effects on
+  flammability across the major litter types.}
+
+  % show a
+  % relationship between time since wetting and time to ignition and spread rate,
+  % with significant interaction with genus. This indicates a genus-specific
+  % behavior of flammability and time since wetting. We found non-additive
+  % effects in litter mixtures for both moisture content and flammability.
+  % Mixtures containing \emph{Q. kelloggii} and \emph{P. jeffreyi}, species that
+  % dry faster, produced more flammable behavior (lower time to ignition, faster
+  % spread rate, taller flames). Fire behavior of mixtures depended on moisture
+  % content, although mixtures without Q. kelloggii had always higher moisture
+  % content than predicted by the average fire behavior of the species in the
+  % mixture. At lower moisture contents (<75\%) the mixtures exhibit positive
+  % non-additive behavior and at higher moisture contents (> 200\%) they exhibit
+  % negative non-additive behavior.}
 
@@
 During flammability trials, we recorded time to ignition (s), duration of
-flaming combustion (s), spread rate (mm/s), calculated by dividing the length
-of the burned surface by the time it takes the flaming front to reach the end
-of the apparatus, maximal flame height (mm), and mass loss (\%), calculated as
-the difference between fuel mass before and after flame extinction. Our
-thermocouple system failed or data was lost for two days of burning trials,
-therefore we have omitted those data from this analysis. Our past work
-\cite{Magalhaes+Schwilk-2012} demonstrated that flame spread rate captures one
-important axis of flammability variation \citep{Schwilk-2015,
-  Pausas+Keeley+etal-2017} and percent fuel consumed captures a portion of the
-second major axis of flammability variation (although not as well as
-integrating temperature over time does). Our analysis, therefore, focused on
-these two relatively orthogonal measures of flammability.
+flaming combustion (s), flame spread rate (mm/s), calculated by dividing the
+length of the burned surface by the time it takes the flaming front to reach
+the end of the apparatus, maximal flame height (mm), and mass loss (\%),
+calculated as the difference between fuel mass before and after flame
+extinction. Some trials resulted only in smoldering combustion; for those
+trials, a time to ignition and mass loss were recorded, but flame spread rate
+and flame height were recorded as zero. Our thermocouple system failed or data
+was lost for two days of burning trials, therefore we have omitted those data
+from this analysis. Our past work \cite{Magalhaes+Schwilk-2012} demonstrated
+that flame spread rate captures one important axis of flammability variation
+\citep{Schwilk-2015, Pausas+Keeley+etal-2017} and percent fuel consumed
+captures a portion of the second major axis of flammability variation (although
+not as well as integrating temperature over time does). Our analysis,
+therefore, focused on these two relatively orthogonal measures of flammability.
 
@@
 As a consequnece of the relationships shown above, time since wetting had a strong effect on spread rate (p < 0.0001) and the slope of this effect varied by taxonimc (Fig.
-\ref{fig:spreadrate}, Table \ref{tab:spreadrate-anova}).
+\ref{fig:spread_time}, Table \ref{tab:spreadrate-anova}).
 The pines were able to ignite and burn much sooner after wetting than the other genera, none of which had measurable spread rates until burning was attempted 72 hours after wetting.
@@
 \emph{Quercus} had the steepest response to time since wetting (Fig.
-\ref{fig5:spreadrate}).
+\ref{fig:spread_time}).
+
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/fig6_spread_time.pdf}
+\caption{Flame spread rate by time since wetting for litter of eight species. Model ANOVA table in Table \ref{tab:tab5_spread_time_anova} and coefficients in SI Table 4.
+  \label{fig:spread_time}
+\end{figure}
 
@@
 Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
-  \label{tab:spreadrate-anova}
+  \label{tab:spread_time_anova}
 \centering
 %% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-\input{figs_tables/spreadrate-anova.ltx}
+\input{figs_tables/tab5_spread_time_anova.ltx}
 \end{table}

### 2019-05-20 scripts/moisture-ms-figs-tables.R

@@ -10,6 +10,8 @@ source("theme-opts.R")
 source("dry-down.R")
 source("burn-moist.R")
 
+set.seed(100)
+
 ###############################################################################
 ## Figure 1: dry-down
 ###############################################################################
@@ -194,7 +196,9 @@ b <- oldflam %>% filter(nchar(as.character(spcode)) < 5) %>% bind_rows(burnt)
 b <- b %>% mutate(consum = case_when(consum==100 ~ 0, TRUE ~ consum),
                   didburn = !(t2ignit==0 & spread==0 & consum <0.000000001),
                   taxon = case_when(genus=="Sequoiadendron" |
-                                      genus=="Calocedrus" ~ "Cupressaceae", TRUE ~ genus)
+                                    genus=="Calocedrus" ~ "Cupressaceae",
+                       #             spcode=="Pila" ~ "P. lambertiana",
+                                    TRUE ~ genus)
                   )
 b.scaled <- b %>% mutate_if(is.numeric, scale)
 
@@ -211,7 +215,7 @@ fig4 <- ggplot(b, aes(actualMC_dry, spread, color=taxon)) +
   scale_colour_manual(values=schwilkcolors) +
   xlab("Moisture content (%)") + ylab("Spread rate (mm/s)") +
   pubtheme.nogridlines +
-  stat_smooth(data=b.burned, method="lm", se=FALSE, size=0.8, drop=TRUE) +
+  stat_smooth(method="lm", se=FALSE, size=0.8, drop=TRUE) +
   theme(legend.position=c(.75, .86),
         legend.spacing.y=unit(0,"cm"),
         legend.text = element_text(family=fontfamily, size=smsize-1, face="italic"),
@@ -220,7 +224,8 @@ ggsave(file.path(RESULTS, "fig4_spread_actualMC.pdf"), plot=fig4,
        width=col1, height=col1, unit="cm")
 
 fig4
-spread.moist.mod <- lmer(spread ~ actualMC_dry + taxon + taxon:hour + vpd + (1 | spcode),
+
+spread.moist.mod <- lmer(spread ~ actualMC_dry + taxon + taxon:actualMC_dry + vpd + (1 | spcode),
                          data=b)
 summary(spread.moist.mod)
 anova(spread.moist.mod)
@@ -228,7 +233,7 @@ anova(spread.moist.mod)
 ###############################################################################
 ## Table 3: Spread rate and moisture content ANOVA
[…]

### 2019-05-20 scripts/theme-opts.R

@@ -39,10 +39,12 @@ pubtheme <- theme_grey() +
         axis.text.y  = element_text(family=fontfamily, size=axissz, color = "black"),
         ## strip.text.x = element_text(family=fontfamily, size = axissz, face="italic"),
         ## strip.text.y = element_text(family=fontfamily, size = axissz, face="italic"),
+        legend.background = element_rect(fill = "transparent"),
         legend.title = element_text(family=fontfamily, size=axissz),
         legend.text = element_text(family=fontfamily, size=smsize),
-        legend.key = element_rect(fill=NA),
-        legend.spacing.y = NULL,  
+        legend.key = element_rect(fill="transparent"),
+        legend.spacing.y = NULL,
+        legend.margin=margin(c(1,1,1,1)),
         legend.key.height = unit(smsize, "pt"),
         panel.grid.major = element_line(colour = "grey90", size = 0.2),
         panel.grid.minor = element_line(colour = "grey95", size =0.5),

### 2019-05-21 scripts/moisture-ms-figs-tables.R
Add beta regression for fuel consumption models
Use glmmadmb

@@ -19,7 +19,7 @@ xbreaks <- seq(0, 144, 24)
 ybreaks <- seq(0, 700, 50)
 
 fig1 <- ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
-  geom_point(size=1.2, alpha=0.7, stroke=0) +
+  geom_jitter(height=0, width=0.5, size=1.2, alpha=0.7, stroke=0) +
   geom_smooth(method="glm",
               method.args=list(family=gaussian(link="log")), se=FALSE, size =0.6) +
   scale_colour_manual(values=schwilkcolors) +
@@ -188,7 +188,7 @@ print(xtable(tab3), file=file.path(RESULTS, "tab3_di_anova.ltx"),  booktabs=TRUE
 ###############################################################################
 ## Flammability plots and tables
 ###############################################################################
-oldflam <- read.csv("../data/burn-trials/flamdt.csv") %>% left_join(species) %>%
+oldflam <- read.csv("../data/burn-trials/flamdt.csv") %>% right_join(species) %>%
   mutate(hour=144, vpd=2.44, t2ignit=ignit, combust=combus) #, actualMC_dry=5)
 b <- oldflam %>% filter(nchar(as.character(spcode)) < 5) %>% bind_rows(burnt)
 
@@ -251,33 +251,66 @@ print(xtable(tab.spread.moist.coef), file=file.path(RESULTS, "SI_tab2_spread_moi
 ###############################################################################
 ## Fig 5: Fuel consumption by moisture content
 ###############################################################################
+b.consume <- b %>% mutate(spcode=factor(spcode), taxon=factor(taxon), consum=consum/100)
+b.consume <- mutate(b.consume, consum.t = case_when(consum == 0 ~ consum+0.00001, TRUE ~ consum),
+            consum=consum.t) %>% filter(!is.na(actualMC_dry))
+
+opts <-admbControl(impSamp=0,maxfn=500,imaxfn=500,maxph=5,noinit=FALSE,shess=TRUE,
+            run=TRUE, ZI_kluge=FALSE, poiss_prob_bound=TRUE)
+
+consume.moist.mod <- glmmadmb(consum ~ actualMC_dry + taxon + taxon:actualMC_dry + (1 | spcode),
+                             family="beta", link="logit", zeroInflation=FALSE, mcmc=TRUE,
+                             data=b.consume, admb.opts=opts, verbose=TRUE)
+summary(consume.moist.mod)
+
+
[…]

### 2019-05-21 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Add all monoculture plots and tables to text.

@@
   gravity for 3 min before initial weighing. Colors indicate genera and there
-  were significant differences in both }
+  were significant differences in both. A small amount of horizontal jitter
+  (0.5 hr) was added to this plot to aid in visualizing because, although
+  collections were staggered, there were often multiple trays collected at a
+  single time point.}
  \label{fig:drydown}
@@
 \begin{table}[H]
-  \caption{Linear model results for maximum moisture content as a function of specific leaf area and litter bulk density.}
+  \caption{Linear model results for maximum moisture content as a function of specific leaf area and litter bulk density. Mixed models were fit as a mixed beta regression. }
   \label{tab2:mc-anova}
@@
 %% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-\input{figs_tables/tab_mc_anova.ltx}
+\input{figs_tables/tab2_mc_anova.ltx}
 \end{table}
@@
 \includegraphics[width=8cm]{figs_tables/fig4_spread_actualMC.pdf}
-\caption{Flame spread rate by moisture content (model fit F= 61.15, d.f = 9, 148,  p < 0.0001). Model ANOVA table in Table \ref{tab:spread-moist-anova} and coefficients in SI Table 2.
+\caption{Flame spread rate by moisture content (model fit F= 61.15, d.f = 9, 148,  p < 0.0001). Model ANOVA table in Table \ref{tab:spread-moist-anova} and coefficients in SI Table 2.}
   \label{fig:spread-moist}
@@
 \begin{table}[H]
-  \caption{Mixed model results for flame spread rate as a function of fuel moisture with taxonomic group as a fixed effect and species as a nested random effect.
+  \caption{Linear mixed model results for flame spread rate as a function of fuel moisture with taxonomic group as a fixed effect and species as a nested random effect.
 Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.
@@
 \includegraphics[width=8cm]{figs_tables/fig5_consume_actualMC.pdf}
-\caption{Percent fuel consumed by moisture content. Model ANOVA table in Table \ref{tab:consume-moist-anova} and coefficients in SI Table 3.
+\caption{Percent fuel consumed by moisture content. Lines indicate best fit beta regression lines. Model deviance table in Table \ref{tab:consume-moist-anova} and coefficients in SI Table 3.}
   \label{fig:consume-moist}
@@
 \begin{table}[H]
-  \caption{Mixed model results for fuel consumption as a function of fuel moisture with taxonomic group as a fixed effect and species as a nested random effect.
-Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.
+  \caption{Generalized linear mixed model results for fuel consumption as a function of fuel moisture with taxonomic group as a fixed effect and species as a nested random effect.
+The analysis of deviance table below shows deviance comparisons for the simplest null model with only intercept and random effect (model 1), for a taxon effect only (model 2), for a taxon and mosture content effect (model 3) and for the full model including the taxon by moisture interaction (model 4).
 Coefficient estimates are in SI Table 3.}
@@
 
-As a consequnece of the relationships shown above, time since wetting had a strong effect on spread rate (p < 0.0001) and the slope of this effect varied by taxonimc (Fig.
+As a consequence of the relationships shown above, time since wetting had a strong effect on spread rate (p < 0.0001) and the slope of this effect varied by taxonomic group (Fig.
 \ref{fig:spread_time}, Table \ref{tab:spreadrate-anova}).
-The pines were able to ignite and burn much sooner after wetting than the other genera, none of which had measurable spread rates until burning was attempted 72 hours after wetting.
-At that point, the \emph{Abies}, \emph{Calocedrus}, and \emph{Sequiadendron} had much drying litter than did the \emph{Quercus}, yet the spread rates were slow.
+The pines were able to ignite and burn much sooner after wetting than the other genera, none of which had measurable spread rates until burning was attempted 72 hours after wetting (note that trials that failed to even ignite smoldering combustion are ommitted completely from the model and figure).
+At that point, the \emph{Abies}, \emph{Calocedrus}, and \emph{Sequiadendron} had dryer litter than did the \emph{Quercus}, yet the spread rates were slow.
 \emph{Quercus} had the steepest response to time since wetting (Fig.
-\ref{fig:spread_time}).
+\ref{fig:spread_time}, SI Table 4)).
 
@@
 \includegraphics[width=8cm]{figs_tables/fig6_spread_time.pdf}
-\caption{Flame spread rate by time since wetting for litter of eight species. Model ANOVA table in Table \ref{tab:tab5_spread_time_anova} and coefficients in SI Table 4.
+\caption{Flame spread rate by time since wetting for litter of eight species.
+  Lines indicate best fit linear model fits. Model ANOVA table in Table
+  \ref{tab:tab5_spread_time_anova} and coefficients in SI Table 4. Horizontal
+  jitter (up to 2 hr) was added to the plot for visualization and to avoid
+  overplotting. All collections were at the times labeled on the x-axis.. A
+  lack of data at earlier time points for a taxon indicates no ignitions
+  occurred at those time points.}
   \label{fig:spread_time}
@@
   \caption[Mixed linear model of flame spread rate.]{Mixed model results for flame spread rate  as a function of time since wetting with genus as a fixed effect and species as a nested random effect.
-Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
+Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.
+Estimated coefficients are in SI Table 4}
   \label{tab:spread_time_anova}
@@
 
-We used binomial logistic regression to model the relationship between moisture content and the probability of ignition to determine moisture of extinction for the eight species.
-Q.
-kelloggii ignited at all moisture levels tested (10--65\%) while C.
-decurrens and A.
-concolor failed to do so across most moisture levels.
-I assumed the inflection point on the curve at 50\% to allow for the determination of moisture of extinction for the remaining species (15\% for A.
-concolor and S.
-giganteum, 20\% for A.
-magnifica, 35\% for P.
-lambertiana, and 60\% for P.
-ponderosa and P.
-jeffreyi).
+Fuel consumption was highly variable across trials and intermediate values were less common than very low are moderately high consumption (Fig.
+\ref{fig:spread_time}).
+All parameters significantly contributed to the model as indicated by Table \ref{tab:consume-time-anova}.
+As with flame spread rate some pines had the most rapid response to drying.
+However, unlike in the other flammability analyses, here the pines did not all behave similarly: \emph{P.
+lambertiana} had lower fuel consumption than did \emph{P.
+ponderosa} and \emph{P.
+jeffreyii}.
+We noticed this during trials and in model fitting we found that placing \emph{P.
+lambertiana} in its own group allowed model convergence that was otherwise impossible.
+
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/fig7_consume_time.pdf}
+\caption{Percent fuel consumed by time since wetting for litter of eight
+  species. Curves shown indicate best fit beta regression lines. Model ANOVA
+  table in Table \ref{tab:tab5_spread_time_anova} and coefficients in SI Table
+  4. Horizontal jitter (up to 2 hr) was added to the plot for visualization and
+  to avoid overplotting. All collections were at the times labeled on the
+  x-axis.. A lack of data at earlier time points for a taxon indicates no
+  ignitions occurred at those time points.}
+  \label{fig:spread_time}
+\end{figure}
+
+\begin{table}[H]
+  \caption{Generalized linear mixed model results for fuel consumption as a function of time since wetting with taxonomic group as a fixed effect and species as a nested random effect.
+These are beta regression models.
+The analysis of deviance table below shows deviance comparisons for the simplest null model with only intercept and random effect (model 1), for a taxon effect only (model 2), for a taxon and mosture content effect (model 3) and for the full model including the taxon by moisture interaction (model 4).
+Coefficient estimates are in SI Table 5.}
+  \label{tab:consume-time-anova}
+\centering
+\input{figs_tables/tab6_consume_time_anova.ltx}
+\end{table}
 
@@
 
-Moisture effects on litter mixtures
-
-Based on visual inspection of the dry down curves and the estimated marginal means for maximum water retainability and desiccation rate, selected four species to represent the full range of litter moisture dynamics for use in the litter mixtures: \emph{Abies concolor}, \
+Based on visual inspection of the dry down curves and the estimated marginal means for maximum water retainability and desiccation rate (SI Figs 2-3), we selected four species to represent the full range of litter moisture dynamics for use in the litter mixtures.
 
@@
 This species failed to ignite several times in the individual trials and was assigned the maximum number of seconds for the ignition attempt, 180.
-
-This section may be divided by subheadings.
-It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.
-\begin{quote}
-  This section may be divided by subheadings.
-It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.
-\end{quote}
 

### 2019-07-01 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Complete results draft.
Need to write conclusions.

@@
 all four possible three-species litter mixtures using those four species. For
-each mixture and time since wetting, we predicted moisture content and flame
-spread rate using the models fit to single species data described above and
-then calculated an expected moisture content and spread rate as the average of
-these three values for the mixture. This procedure was done for both moisture
-content and flammability parameter values of the constituent species. The
+each mixture and time since wetting, we predicted moisture content based on the
+mean content of the three consituent species at that time since wetting. We
+predicted flammability parameters based on the models fit to single species
+data described above and then calculated an expected moisture content and
+spread rate as the average of these three values for the mixture. The
 expectation from the null model was that the difference between observed and
@@
 significant departures from the null (zero) with a mixed effects linear model.
+For flammability parameters, we fit linear mixed effect models with the
+observed-predicted flammability value as the dependent variable and used Wald
+tests to test if the centered intercept was different than zero (overall non
+additivity) and if the slope with time since wetting was different from zero
+(non additivity changing with drying).
 
@@
 \emph{Quercus}, however, like the pines has a rapid dessication rate (Fig.
-\ref{fig:drydown-logged}). Calculating estimated marginal means following the
+\ref{fig:drydown} and SI Fig. 1). Calculating estimated marginal means following the
 linear mixed-effects model discriminated the species into three different
@@
 bulk density had a negative effect on desiccation rate (Fig. \ref{fig:bd-di},
-linear model p = 0.029). Anova tables are shown in in Fig. \ref{tab-mc-anova}
-and Fig. \ref{tab:di-anova}.
+linear model p = 0.029). Anova results are shown in tables \ref{tab:mc_anova}
+and \ref{tab:di_anova}.
 
@@
   \caption{Linear model results for maximum moisture content as a function of specific leaf area and litter bulk density. Mixed models were fit as a mixed beta regression. }
-  \label{tab2:mc-anova}
+  \label{tab:mc_anova}
 \centering
@@
   \caption{Linear model results for estimated maximum water retention (\%) as a function of specific leaf area and litter bulk density.}
-  \label{tab2:mc-anova}
+  \label{tab:di_anova}
 \centering
@@
 Flame spread rate declined with fuel moisture (Fig.
-\ref{fig:spread-moist}.
-\emph{Quercus} had higher spread rates across all fuel moistures and had a steeper decline in response to moisture than did the other groups (taxon by time since wetting interaction p = 0.022, Table \ref{tab:spread-moist-anova}).
+\ref{fig:spread_moist}.
+\emph{Quercus} had higher spread rates across all fuel moistures and had a steeper decline in response to moisture than did the other groups (taxon by time since wetting interaction p = 0.022, Table \ref{tab:spreadrate_anova}).
 Fuel moisture negatively influenced fuel consumption and overall  fuel consumption as well as the rate of change with moisture content varied across the taxonomic group (Fig.
-\ref{fig:consume-moist}, Table \ref{tab:consume-moist-anova}).
+\ref{fig:consume_moist}, Table \ref{tab:consume_moist_anova}).
 Although the overall ranking of groups was simlar for fuel consumption and for flame spread rate, in contrast to the pattern with spread rate, for fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in flammability with fuel moisture than did the other groups.
@@
 \includegraphics[width=8cm]{figs_tables/fig4_spread_actualMC.pdf}
-\caption{Flame spread rate by moisture content (model fit F= 61.15, d.f = 9, 148,  p < 0.0001). Model ANOVA table in Table \ref{tab:spread-moist-anova} and coefficients in SI Table 2.}
-  \label{fig:spread-moist}
+\caption{Flame spread rate by moisture content (model fit F= 61.15, d.f = 9, 148,  p < 0.0001). Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI Table 2.}
+  \label{fig:spread_moist}
 \end{figure}
@@
 Coefficient estimates are in SI Table 2.}
-  \label{tab:spreadrate-anova}
+  \label{tab:spreadrate_anova}
 \centering
@@
 \includegraphics[width=8cm]{figs_tables/fig5_consume_actualMC.pdf}
-\caption{Percent fuel consumed by moisture content. Lines indicate best fit beta regression lines. Model deviance table in Table \ref{tab:consume-moist-anova} and coefficients in SI Table 3.}
-  \label{fig:consume-moist}
+\caption{Percent fuel consumed by moisture content. Lines indicate best fit beta regression lines. Model deviance table in Table \ref{tab:consume_moist_anova} and coefficients in SI Table 3.}
+  \label{fig:consume_moist}
 \end{figure}
@@
 Coefficient estimates are in SI Table 3.}
-  \label{tab:consume-moist-anova}
+  \label{tab:consume_moist_anova}
 \centering
@@
 As a consequence of the relationships shown above, time since wetting had a strong effect on spread rate (p < 0.0001) and the slope of this effect varied by taxonomic group (Fig.
-\ref{fig:spread_time}, Table \ref{tab:spreadrate-anova}).
+\ref{fig:spread_time}, Table \ref{tab:spreadrate_anova}).
 The pines were able to ignite and burn much sooner after wetting than the other genera, none of which had measurable spread rates until burning was attempted 72 hours after wetting (note that trials that failed to even ignite smoldering combustion are ommitted completely from the model and figure).
@@
   Lines indicate best fit linear model fits. Model ANOVA table in Table
-  \ref{tab:tab5_spread_time_anova} and coefficients in SI Table 4. Horizontal
+  \ref{tab:spread_time_anova} and coefficients in SI Table 4. Horizontal
   jitter (up to 2 hr) was added to the plot for visualization and to avoid
@@
 \ref{fig:spread_time}).
-All parameters significantly contributed to the model as indicated by Table \ref{tab:consume-time-anova}.
+All parameters significantly contributed to the model as indicated by Table \ref{tab:consume_time_anova}.
 As with flame spread rate some pines had the most rapid response to drying.
@@
   species. Curves shown indicate best fit beta regression lines. Model ANOVA
-  table in Table \ref{tab:tab5_spread_time_anova} and coefficients in SI Table
+  table in Table \ref{tab:consume_time_anova} and coefficients in SI Table
   4. Horizontal jitter (up to 2 hr) was added to the plot for visualization and
@@
 Coefficient estimates are in SI Table 5.}
-  \label{tab:consume-time-anova}
+  \label{tab:consume_time_anova}
 \centering
@@
 
-Based on visual inspection of the dry down curves and the estimated marginal means for maximum water retainability and desiccation rate (SI Figs 2-3), we selected four species to represent the full range of litter moisture dynamics for use in the litter mixtures.
+Based on visual inspection of the dry down curves and the estimated marginal means for maximum water retainability and desiccation rate (SI Figs 2-3), we selected four species to represent the full range of litter moisture dynamics for use in the litter mixtures and examined all four three-species combinations of those four.
+The four species were \emph{Abies concolor}, \emph{Calocedrus decurrens}, \emph{Pinus jeffreyi} and \emph{Quercus kellogii}.
 
-The mixture that does not contain Q.
-kelloggii (AbCaPi; Fig.
-2.5) absorbs less moisture.
-This result is to be expected considering the result obtained from the single species trials, and reveals that Q.
-kelloggii, as the species that absorbs more moisture, drives the water retainability behavior of the mixtures it appears in.
-In terms of maximum water retainability, the mixtures discriminate into three groups: 1) mixture without Q.
-kelloggii (AbCaPi), 2) mixtures with both A.
-concolor and Q.
-kelloggii (AbCaQu and AbPiQu), and 3) mixture without A.
-concolor (CaPiQu).
-There is no discrimination between the mixtures in terms of slopes.
-Though the mixtures with both C.
-decurrens and P.
-jeffreyi (AbCaPi and CaPiQu) are significantly different from one another, the remaining two mixtures (AbCaQu and AbPiQu) cannot be distinguished from one another nor from AbCaPi and CaPiQu.
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/fig8_mixture_drydown-curves.pdf}
+\caption{Fuel moisture by time since saturation for four litter mixtures.
+Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.
+Moistures were measured every 24 hours but data are shown with slight horizontal jitter to reduce point overlap.}
+  \label{fig:mixture_dry_down}
+\end{figure}
+
+Mixture dry down rates and maximum water retention showed some relationship to the dry down curves for individual constituent species (Fig.
+\ref{fig:mixture_dry_down}.
+For example, the mixture that does not contain \emph{Q.
+kelloggii} (AbCaPi in Fig.
+Fig.
+\ref{fig:mixture_dry_down}) has the lowest initial retention.
+Although intiial water retention differed across mixtures (RESULTS?), there was no significant difference in dry down rates (slopes in linear model based on log of moisture content).
+
+% .
 The moisture content of a mixture at any given time was also dependent upon the bulk density of the mixture (anova analysis resulted in a better model if bulk density was included as a fixed effect, p=0.01135).
 
-Time since wetting effect on mixture flammability
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/fig9_mixture_obs_vs_pred_mc.pdf}
+\caption{Observed vs predicted fuel moisture content for litter mixtures.
+Predictions are based on the mean moisture content at each particualr time sinece wetting for the three constituent species in each mixture.
+Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
+  \label{fig:mixture_obs_pred_mc}
+\end{figure}
 
-In general, the flammability of mixtures was higher as time since wetting went up as expected (Table 2.2).
+There was some evidence for non additive in drying rate a tendency for litter mixtures.
+Mixtures tended to have lower water content than predicted from the mean of the constituent species after wetting, but higher water content than expected after long drying (Wald test shows slope of mixed linear model illustrated in Fig.
+\ref{fig:mixture_obs_pred_mc} less than 1, p < 0.0001).
+
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/fig10_mixture_obs_vs_pred_spread.pdf}
+\caption{Observed vs predicted flame spread rate for litter mixtures.
+Y axis shows relative spread rate values (predicted - observed) so that positive values indicate faster than predicted and negative slower than predicted.
+Predictions are based on the modeled spread rate at each particular time since wetting for the three constituent species in each mixture.
+Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
+  \label{fig:mixture_obs_pred_spread}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/fig11_mixture_obs_vs_pred_consume.pdf}
+\caption{Observed vs predicted fuel consumption for litter mixtures.
+Predictions are based on the modeled spread rate at each particular time since wetting for the three constituent species in each mixture.
+Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
+  \label{fig:mixture_obs_pred_consume}
+\end{figure}
+
+Flammability measures also demonstrated nonadditivity in mixtures.
+Flame spread rate was consistently slightly higher than predicted (Fig \ref{fig:mixture_obs_pred_spread}, Wald test of mixed linear model centered intercept for observed-predicted values greater than zero, p = 0.017) and the positive non additivity increased with time since wetting (slope greater than zero, p = 0.005).
+Fuel consumption, however, showed no overall bias towards positive or negative non additivity, but did show a positive relationship with time since wetting (Fig.
+\ref{fig:mixture_obs_pred_consume}, p < 0.0001).
+
+% In general, the flammability of mixtures was higher as time since wetting went up as expected (Table 2.2).
 The absence of Q.
@@
 
-Non-additivity in moisture content
+% Non-additivity in moisture content
 
-We examined the relationship between observed moisture content of a mixture, and the moisture content predicted based on the constituent species to determine the existence of non-additive effects in moisture content (Fig.
+% We examined the relationship between observed moisture content of a mixture, and the moisture content predicted based on the constituent species to determine the existence of non-additive effects in moisture content (Fig.
 2.6).
@@
 
-Non-additivity in flammability
+% Non-additivity in flammability
 
-We determined the presence of non-additivity in the flammability of litter mixtures by comparing the flammability of the mixtures to a null model created by the arithmetic average of the predicted individual flammability of each species in the mixture (Figs.
+% We determined the presence of non-additivity in the flammability of litter mixtures by comparing the flammability of the mixtures to a null model created by the arithmetic average of the predicted individual flammability of each species in the mixture (Figs.
 2.7 and 2.8).
@@
 jeffreyi (CaPiQu).
+
+The fact that mixtures tended to have lower water content than predicted from
+the mean of the constituent species after wetting, but higher water content
+than expected after long drying is a bit surprising as past work has shown that
+mixtures tend to be more aerated than expected given the bulk density of litter
+of individual species \citep{Magalhaes+Schwilk-2012}. This non additivity in
+drying rate did not fully explain non additivity in flammability, however.
 

### 2019-07-01 scripts/dry-down.R

@@ -7,7 +7,7 @@
 
 library(tidyr)
 library(lme4)
-
+library(emmeans)
 library(dplyr)
 library(stringr)
 

### 2019-07-01 scripts/moisture-ms-figs-tables.R

@@ -3,12 +3,14 @@
 
 library(xtable)
 library(ggrepel)
+library(glmmADMB)
 
 RESULTS <- "../ms/figs_tables/"
 
 source("theme-opts.R")
 source("dry-down.R")
 source("burn-moist.R")
+source("read-moistmix.R")
 
 set.seed(100)
 
@@ -463,3 +465,147 @@ tab.consume.time.coef <- summary(consume.time.mod)$coefficients
 print(xtable(tab.consume.time.coef),
       file=file.path(RESULTS, "SI_tab5_consume_time_coef.ltx"),
       booktabs=TRUE, floating=FALSE)
+
+
+##############################################################################
+## Mixtures
+##############################################################################
+
+
+###############################################################################
+## Fig 8: Drydown for mixtures
+###############################################################################
+fig8 <- ggplot(mmc, aes(hour, MC_dry, color=spcode)) +
+  geom_jitter(height=0, width=0.5, size=1.2, alpha=0.7, stroke=0) +
+  geom_smooth(method="glm",
+              method.args=list(family=gaussian(link="log")), se=FALSE, size =0.6) +
+  scale_colour_manual(values=schwilkcolors) +
+  xlab("Hours since wetting") + ylab("Moisture by dry weight (%)") +
+  scale_x_continuous(breaks=xbreaks) +
+  scale_y_continuous(breaks=ybreaks) +
+  pubtheme.nogridlines +
+  theme(legend.position=c(0.75, 0.86),
+        legend.title=element_blank())
+#        legend.text = element_text(family=fontfamily, size=smsize, face="italic"))
+#       legend.key.height=unit(smsize,"pt"))
+fig8
+ggsave(file.path(RESULTS, "fig8_mixture_drydown-curves.pdf"), plot=fig8,
+       width=col1, height=col1, units="cm")
+
+
+###############################################################################
+## Fig 9: Observed vs predicted moisture
+###############################################################################
+
+fig9 <- ggplot(mmc.sum, aes(MC_dry_pred, MC_dry.mean, color=spcode, shape=factor(hour))) +
+  geom_point(size=4) +
+  geom_pointrange(aes(ymin=MC_dry.mean-MC_dry.sd,
[…]

### 2019-07-01 scripts/read-moistmix.R

@@ -31,15 +31,15 @@ mmc <- mmc %>% mutate(mixcode = str_replace(mixcode, "Ab", "Abco"),
                       sp3 = str_sub(mixcode, 9,12))
 
 
-mmc <- mmc %>% mutate(MC_dry_pred1 = exp(predict(dry.mod$full_model, allow.new.levels=TRUE,
+mmc <- mmc %>% mutate(MC_dry_pred1 = exp(predict(dry.mod, allow.new.levels=TRUE,
                                                 newdata=data.frame(spcode=sp1,
                                                 tray = str_c(sp1,"_",rep,"NEW"),
                                                 hour = hour))),
-                      MC_dry_pred2 = exp(predict(dry.mod$full_model, allow.new.levels=TRUE,
+                      MC_dry_pred2 = exp(predict(dry.mod, allow.new.levels=TRUE,
                                                 newdata=data.frame(spcode=sp2,
                                                 tray = str_c(sp2,"_",rep,"NEW"),
                                                 hour = hour))),
-                      MC_dry_pred3 = exp(predict(dry.mod$full_model, allow.new.levels=TRUE,
+                      MC_dry_pred3 = exp(predict(dry.mod, allow.new.levels=TRUE,
                                                 newdata=data.frame(spcode=sp3,
                                                 tray = str_c(sp3,"_",rep,"NEW"),
                                                 hour = hour))),
@@ -73,28 +73,23 @@ mmc$logMC_dry <-  log(mmc$MC_dry)
 mmc <- mmc %>% mutate(tray = str_c(spcode, "_", rep))
 
 # Fit a nested model using lmer
-mdry.mod <- lmer(log(MC_dry) ~ hour*spcode + (1 + hour | tray), data=mmc)
+mdry.mod <- lmer(log(MC_dry) ~ hour*spcode + (1 | tray), data=mmc)
 summary(mdry.mod)
 anova(mdry.mod)
 
 # comparing the above model with one with bulk density added. Second model is better
-mdrybd.mod <- lmer(log(MC_dry) ~ hour*spcode + bd + (1 + hour | tray), data=mmc)
+mdrybd.mod <- lmer(log(MC_dry) ~ hour*spcode + bd + (1 | tray), data=mmc)
 summary(mdrybd.mod)
 anova(mdry.mod, mdrybd.mod)
 
[…]

### 2019-07-03 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Add author contributions, little bits to ms draft

@@
 % LaTeX support: latex@mdpi.com 
-%  In case you need support, please attach all files that are necessary for compiling as well as the log file, and specify the details of your LaTeX setup (which operating system and LaTeX version / tools you are using).
+%  In case you need support, please attach all files that are necessary for compiling as well as the log file, and specify the details of your LaTeX setup (which operating system and LaTeX version  tools you are using).
 
@@
 %----------
-% Choose between the following MDPI journals:
-% acoustics, actuators, addictions, admsci, aerospace, agriculture, agriengineering, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, asc, asi, atmosphere, atoms, axioms, batteries, bdcc, behavsci , beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci , buildings, cancers, carbon , catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, children, cleantechnol, climate, clockssleep, cmd, coatings, colloids, computation, computers, condensedmatter, cosmetics, cryptography, crystals, dairy, data, dentistry, designs , diagnostics, diseases, diversity, drones, econometrics, economies, education, electrochem, electronics, energies, entropy, environments, epigenomes, est, fermentation, fibers, fire, fishes, fluids, foods, forecasting, forests, fractalfract, futureinternet, futurephys, galaxies, games, gastrointestdisord, gels, genealogy, genes, geohazards, geosciences, geriatrics, hazardousmatters, healthcare, heritage, highthroughput, horticulturae, humanities, hydrology, ijerph, ijfs, ijgi, ijms, ijns, ijtpp, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jcdd, jcm, jcp, jcs, jdb, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmse, jnt, jof, joitmc, jpm, jrfm, jsan, land, languages, laws, life, literature, logistics, lubricants, machines, magnetochemistry, make, marinedrugs, materials, mathematics, mca, medicina, medicines, medsci, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, modelling, molbank, molecules, mps, mti, nanomaterials, ncrna, neuroglia, nitrogen, notspecified, nutrients, ohbm, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, physics, plants, plasma, polymers, polysaccharides, preprints , proceedings, processes, proteomes, psych, publications, quantumrep, quaternary, qubs, reactions, recycling, religions, remotesensing, reports, resources, risks, robotics, safety, sci, scipharm, sensors, separations, sexes, signals, sinusitis, smartcities, sna, societies, socsci, soilsystems, sports, standards, stats, surfaces, surgeries, sustainability, symmetry, systems, technologies, test, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vehicles, vetsci, vibration, viruses, vision, water, wem, wevj
-
-%---------
-% article
-%---------
-% The default type of manuscript is "article", but can be replaced by: 
-% abstract, addendum, article, benchmark, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, conferenceproceedings, correction, conferencereport, expressionofconcern, extendedabstract, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimages, letter, meetingreport, newbookreceived, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, supfile, technicalnote, viewpoint
-% supfile = supplementary materials
-
-%----------
-% submit
-%----------
-% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted.
-This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information.
-Also, line numbering will be removed.
-Journal info and pagination for accepted papers will also be assigned by the Editorial Office.
-
-%------------------
-% moreauthors
-%------------------
-% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.
-
-%---------
-% pdftex
-%---------
-% The option pdftex is for use with pdfLaTeX. If eps figures are used, remove the option pdftex and use LaTeX and dvi2pdf.
-
 %=================================================================
@@
 
-% Authors should discuss the results and how they can be interpreted in perspective of previous studies and of the working hypotheses.
-The findings and their implications should be discussed in the broadest context possible.
-Future research directions may also be highlighted.
-
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontributions{For research articles with several authors, a short paragraph specifying their individual contributions must be provided.
-The following statements should be used ``conceptualization, X.X.
-and Y.Y.; methodology, X.X.; software, X.X.; validation, X.X., Y.Y.
-and Z.Z.; formal analysis, X.X.; investigation, X.X.; resources, X.X.; data curation, X.X.; writing--original draft preparation, X.X.; writing--review and editing, X.X.; visualization, X.X.; supervision, X.X.; project administration, X.X.; funding acquisition, Y.Y.'', please turn to the  \href{http://img.mdpi.org/data/contributor-role-instruction.pdf}{CRediT taxonomy} for the term explanation.
-Authorship must be limited to those who have contributed substantially to the work reported.}
+\authorcontributions{conceptualization, RMQM and DWS; methodology,  RMQM and DWS; experiments, RMQM. coding and data analysis, DWS and RMQM; writing DWS and RMQM; writing--review and editing, DWS}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\funding{Please add: ``This research received no external funding'' or ``This research was funded by NAME OF FUNDER grant number XXX.'' and  and ``The APC was funded by XXX''.
-Check carefully that the details given are accurate and use the standard spelling of funding agency names at \url{https://search.crossref.org/funding}, any errors may affect your future funding.}
+\funding{This research received no external funding.}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\acknowledgments{In this section you can acknowledge any support given which is not covered by the author contribution or funding sections.
-This may include administrative and technical support, or donations in kind (e.g., materials used for experiments).}
+\acknowledgments{TODO.}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\conflictsofinterest{Declare conflicts of interest or state ``The authors declare no conflict of interest.'' Authors must identify and declare any personal circumstances or interest that may be perceived as inappropriately influencing the representation or interpretation of reported research results.
-Any role of the funders in the design of the study; in the collection, analyses or interpretation of data; in the writing of the manuscript, or in the decision to publish the results must be declared in this section.
-If there is no role, please state ``The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, or in the decision to publish the results''.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% optional
-\abbreviations{The following abbreviations are used in this manuscript:\\
-
-\noindent 
-\begin{tabular}{@{}ll}
-MDPI & Multidisciplinary Digital Publishing Institute\\
-DOAJ & Directory of open access journals\\
-TLA & Three letter acronym\\
-LD & linear dichroism
-\end{tabular}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% optional
-\appendixtitles{no} %Leave argument "no" if all appendix headings stay EMPTY (then no dot is printed after "Appendix A"). If the appendix sections contain a heading then change the argument to "yes".
-\appendix
-\section{}
-\unskip
-\subsection{}
-The appendix is an optional section that can contain details and data supplemental to the main text.
-For example, explanations of experimental details that would disrupt the flow of the main text, but nonetheless remain crucial to understanding and reproducing the research shown; figures of replicates for experiments of which representative data is shown in the main text can be added here if brief, or as Supplementary data.
-Mathematical proofs of results not central to the paper can be added as an appendix.
+\conflictsofinterest{The authors declare no conflict of interest.
+The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, or in the decision to publish the results.}
 
@@
 All appendix sections must be cited in the main text. In the appendixes, Figures, Tables, etc. should be labeled starting with `A', e.g., Figure A1, Figure A2, etc.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here.
-
-%=====================================
-% References, variant A: internal bibliography
-%=====================================
-% \reftitle{References}
-% \begin{thebibliography}{999}
-% % Reference 1
-% \bibitem[Author1(year)]{ref-journal}
-% Author1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
-% % Reference 2
-% \bibitem[Author2(year)]{ref-book}
-% Author2, L. The title of the cited contribution. In {\em The Book Title}; Editor1, F., Editor2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
-% \end{thebibliography}
-
-% The following MDPI journals use author-date citation: Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Laws, Religions, Risks, Social Sciences.
-For those journals, please follow the formatting guidelines on http://www.mdpi.com/authors/references
-% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
-% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p.
-328; \citeyear{ref-journal-3b}, p.475).
-This produces: Wong (1999, p.
-328; 2000, p.
-475)
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% optional
-%\sampleavailability{Samples of the compounds ...... are available from the authors.}
-
-%% for journal Sci
-%\reviewreports{\\
-%Reviewer 1 comments and authors’ response\\
-%Reviewer 2 comments and authors’ response\\
-%Reviewer 3 comments and authors’ response
-%}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \end{document}

### 2019-07-08 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
ms edits

@@
 % Abstract (Do not insert blank lines, i.e. \\) 
-\abstract{ Moisture content is a strong determinant of forest fuel
-  flammability. But less is known about the mechanisms by which litter traits
+\abstract{ Moisture content is a strong determinant of forest fuel flammability
+  and varies on small spatial scales and over short time periods. Emperical
+  modeling of equilibrium dead fine fuel moisture is well established, but less
+  is known about the mechanisms by which species-specific litter traits
   influence litter moisture dynamics, particularly the processes of moisture
@@
   more aerated litter beds (such as \emph{Q. kelloggii} and \emph{Pinus}
-  species) have faster rates of litter desiccation. These differing moisture
-  dynamics result in time since wetting having strongly differing effects on
-  flammability across the major litter types.}
-
-  % show a
-  % relationship between time since wetting and time to ignition and spread rate,
-  % with significant interaction with genus. This indicates a genus-specific
-  % behavior of flammability and time since wetting. We found non-additive
-  % effects in litter mixtures for both moisture content and flammability.
-  % Mixtures containing \emph{Q. kelloggii} and \emph{P. jeffreyi}, species that
-  % dry faster, produced more flammable behavior (lower time to ignition, faster
-  % spread rate, taller flames). Fire behavior of mixtures depended on moisture
-  % content, although mixtures without Q. kelloggii had always higher moisture
-  % content than predicted by the average fire behavior of the species in the
-  % mixture. At lower moisture contents (<75\%) the mixtures exhibit positive
-  % non-additive behavior and at higher moisture contents (> 200\%) they exhibit
-  % negative non-additive behavior.}
-
-% A single paragraph of about 200 words maximum. For research articles,
-% abstracts should give a pertinent overview of the work. We strongly encourage
-% authors to use the following style of structured abstracts, but without
-% headings: (1) Background: Place the question addressed in a broad context and
-% highlight the purpose of the study; (2) Methods: Describe briefly the main
-% methods or treatments applied; (3) Results: Summarize the article's main
-% findings; and (4) Conclusion: Indicate the main conclusions or
-% interpretations. The abstract should be an objective representation of the
-% article, it must not contain results which are not presented and
-% substantiated in the main text and should not exaggerate the main
-% conclusions.
+  species) have faster rates of litter desiccation. We examined two axes of
+  flammability (spread rate and fuel consumption) and found that these
+  differing moisture dynamics result in time since wetting having strongly
+  differing effects on flammability across the major litter types typified by
+  different tree genera. Furthermore, we found that litter mixtures exhibited
+  consistent positive non additivity in flame spread rate and that fuel
+  consumption and flame spread rate both showed increasing positive non
+  additivity with time since wetting.
 
@@
 \keyword{flammability, litter, flame spread rate}
-
-% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
-%\PACS{J0101}
-%\MSC{}
-%\JEL{}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Only for the journal Diversity
-%\LSID{\url{http://}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Only for the journal Applied Sciences:
-%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Only for the journal Data:
-%\dataset{DOI number or link to the deposited data set in cases where the data set is published or set to be published separately.
-If the data set is submitted and will be published as a supplement to this paper in the journal Data, this field will be filled by the editors of the journal.
-In this case, please make sure to submit the data set as a supplement when entering your manuscript into our manuscript editorial system.}
-
-%\datasetlicense{license under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Only for the journal Toxins
-%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \setcounter{section}{-1} %% Remove this when starting to work on the template.
-% \section{How to Use this Template}
-% The template details the sections that can be used in a manuscript.
-Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section).
-Please check the instructions for authors page of the journal to verify the correct order and names.
-For any questions, please contact the editorial office of the journal or support@mdpi.com.
-For LaTeX related questions please contact latex@mdpi.com.
-%The order of the section titles is: Introduction, Materials and Methods, Results, Discussion, Conclusions for these journals: aerospace,algorithms,antibodies,antioxidants,atmosphere,axioms,biomedicines,carbon,crystals,designs,diagnostics,environments,fermentation,fluids,forests,fractalfract,informatics,information,inventions,jfmk,jrfm,lubricants,neonatalscreening,neuroglia,particles,pharmaceutics,polymers,processes,technologies,viruses,vision
-
 \section{Introduction}
-% The introduction should briefly place the study in a broad context and highlight why it is important.
-It should define the purpose of the work and its significance.
-The current state of the research field should be reviewed carefully and key publications cited.
-Please highlight controversial and diverging hypotheses when necessary.
-Finally, briefly mention the main aim of the work and highlight the principal conclusions.
-As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research.
-Citing a journal paper \cite{ref-journal}.
-And now citing a book reference \cite{ref-book}.
-Please use the command \citep{ref-journal} for the following MDPI journals, which use author-date citation: Administrative Sciences, Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Languages, Laws, Religions, Risks, Social Sciences.
 
 Fuel moisture content is an important, and often driving, determinant of surface fire behavior.
-Environmental conditions such as temperature, relative humidity, and wind, drive the moisture content of the fuel available to burn [CITE].
-In particular, fine dead fuels respond very readily to environmental changes [CITE].
-In mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fuelled by plant litter, leaf traits can influence landscape scale fire behavior \cite{Schwilk+Caprio-2011}.
-Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape \cite{Fonda+Belanger+etal-1998, Kane+Varner+etal-2008, Schwilk+Caprio-2011, Kreye+Varner+etal-2013}.
-Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936, Fons-1946, Anderson+Rothermal-1965}.
-However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be characterized by leaf traits, litter traits, or both.
-
+Environmental conditions such as temperature, relative humidity, and wind, drive the moisture content of the fuel available to burn \cite{Kreye+Varner+Etal-2018.
+In particular, fine dead fuels respond very readily to environmental changes \cite{Nelson 2001}.
 Fuel moisture influences flammability and fire behavior through effects on ignition, spread rate, flame height, fuel consumption, and heat release, due to the high specific heat content of water \cite{Rothermel-1972}.
@@
 Moisture of litter fuels varies with climate, with seasons and with short term weather within a season.
-Therefore, the rate of drying and flammability at any particular moisture content are key parameters that influence the fuel flammability throughout the fire season \cite{Viney-1991}.
-If species differ in the rate of drying following a precipitation event, this may strongly influence relatively flammability across species at any particular time post-wetting.
+Therefore, the rate of drying and flammability at any particular moisture content are key parameters that influence the fuel flammability throughout the fire season \cite{Nelson 2001}.
+
+Fire scientists have long categorized fuel by size classes that correspond to moisture exchange rates, although such simplified categories have long been recognized as not sufficient to describe species specific and decomposition effects on moisture dynamics \cite{Anderson-1985}.
+A variety of models describing moisture dynamics of fine fuels are used by fire scientists \cite[eg]{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001} that include at least temperature and relative humidity and most aim to describe equilibrium conditions although some include precipitation and solar radiation effects as well.
+Fuel moisture dynamics of leaf litter are dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds \cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}.
+In many litter fuel driven systems, natural lightning ignitions are concurrent with precipitation and therefore fire behavior can depend on rate of drying following a wetting event.
+If species differ in the rate of drying, this may strongly influence relatively flammability across species at any particular time post-wetting.
 Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires.
-Fuel moisture dynamics of leaf litter are dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds \cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}.
+
+In mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fuelled by plant litter, leaf traits can influence landscape scale fire behavior \cite{Schwilk+Caprio-2011}.
+Similarly, strong species specific effects on litter flammability are recorded for the forests tof the southeastern US \cite{Nowacki+Abrams-2008} where species-specific litter drying rates play a strong role in flammability feedbacks \cite{Kreye+Varner+etal-2013}.
+Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape \cite{Fonda+Belanger+etal-1998, Kane+Varner+etal-2008, Schwilk+Caprio-2011, Kreye+Varner+etal-2013}.
+Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936, Fons-1946, Anderson+Rothermal-1965}.
+However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be characterized by leaf traits, litter traits, or both.
 
@@
 
+Flammability is a multi-dimensional characteristic of fuels \citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing different components of fire behavior.
+In litter driven fire, the two major axes of variation are described by measurements related to flame spread rate and measurements related to heat transfer.
+Flame spread rate is easy to measure in burning trials, but heat transfer can be more difficult and often proxies such as duration of heating are used \citep{Magalhaes+Schwilk-2012, Varner+Kane+etal-2015}.
+These different components of fire behavior have differing ecological effects.
 Species traits may interact such that the flammability of multi-species mixtures is not the average of the constituent species.
@@
 
-% I investigated non-additive effects on both moisture content and the flammability of litter mixtures varying in time since wetting and in moisture content.
-I expect that litter mixtures will differ in their bulk density and thus, will dry at different rates.
-
 We investigated species trait effects on litter moisture dynamics and consequent litter flammability across eight tree species in a temperate fire-prone forest in California.
 We determined the rate of moisture loss, the moisture of extinction (the threshold representing the moisture value above which fire will not ignite or spread) and the influence of moisture on the flammability of eight species from a mixed-conifer forest in California.
-Experimental burn trials of leaf litter at various moisture levels determined how time since wetting influenced time to ignition and flame spread rate.
+Previous work described the flammability of these species in monoculture and in mixture for dry litter \cite{Magalhaes+Schwilk-2012}.
+{TODO- more on oaks?]  Experimental burn trials of leaf litter at various moisture levels determined how time since wetting influenced two axes of flammability characterized by flame spread rate and total fuel consumed.
 The burn trials were also performed with multi-species litter mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects.
 We compared leaf litter traits with the measured moisture absorption and desorption rates.
-Finally, we tested if the non-additive effects on flammability reported in previous studies occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.
+Finally, we tested if the non-additive effects on  reported in previous studies occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.
 
@@
   ignitions occurred at those time points.}
-  \label{fig:spread_time}
+  \label{fig:consume_time}
 \end{figure}
@@
 
-\subsection{Effects of moisture and flammability in litter mixtures}
+\subsection{Non additivity in  litter mixtures}
 
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/fig9_mixture_obs_vs_pred_mc.pdf}
+\includegraphics[width=16cm]{figs_tables/fig9_mixture_obs_vs_pred_mc.pdf}
 \caption{Observed vs predicted fuel moisture content for litter mixtures.
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/fig10_mixture_obs_vs_pred_spread.pdf}
+\includegraphics[width=16cm]{figs_tables/fig10_mixture_obs_vs_pred_spread.pdf}
 \caption{Observed vs predicted flame spread rate for litter mixtures.
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/fig11_mixture_obs_vs_pred_consume.pdf}
+\includegraphics[width=16cm]{figs_tables/fig11_mixture_obs_vs_pred_consume.pdf}
 \caption{Observed vs predicted fuel consumption for litter mixtures.
@@
 
-% In general, the flammability of mixtures was higher as time since wetting went up as expected (Table 2.2).
-The absence of Q.
-kelloggii in a mixture, in general, lead to slower flame spread rates, longer flaming combustion, and longer ignition times.
-The combination of both Q.
-kelloggii and P.
-jeffreyi in a mixture generated taller flames and shorter flaming combustion.
-
-% Non-additivity in moisture content
-
-% We examined the relationship between observed moisture content of a mixture, and the moisture content predicted based on the constituent species to determine the existence of non-additive effects in moisture content (Fig.
-2.6).
-The mixture that does not contain Q.
-kelloggii (AbCaPi) consistently retains more moisture that what would be predicted by the average of its constituent species.
-In general, mixtures start the drying process drier than expected and after 48 hours are wetter than expected, indicating lower moisture absorption and slower desiccation rates.
-The non-additive behavior is dependent on the moisture content of the mixture.
-When moisture content is above 200\%, I observe negative non-additivity for all mixtures containing Q.
-kelloggii.
-There is additive behavior when moisture content is between 75 and 150\%, and positive non-additivity below 75\% moisture content.
-
-% Non-additivity in flammability
-
-% We determined the presence of non-additivity in the flammability of litter mixtures by comparing the flammability of the mixtures to a null model created by the arithmetic average of the predicted individual flammability of each species in the mixture (Figs.
-2.7 and 2.8).
-Both spread rate and time to ignition exhibited non-additivity, mediated by time since wetting.
-Observed spread rate values for the mixtures were higher than predicted by the null model, though the non-additivity was stronger for mixtures that contained Q.
-kelloggii.
-This result is likely driven by the bulk density of the mixtures, which was lower than predicted, making the mixtures’ litter beds more aerated.
-Non-additivity in time to ignition is strongly influenced by the presence of C.
-decurrens in the mixture.
-This species failed to ignite several times in the individual trials and was assigned the maximum number of seconds for the ignition attempt, 180.
-
-%\begin{listing}[H]
-%\caption{Title of the listing}
-%\rule{\textwidth}{1pt}
-%\raggedright Text of the listing. In font size footnotesize, small, or normalsize. Preferred format: left aligned and single spaced. Preferred border format: top border line and bottom border line.
-%\rule{\textwidth}{1pt}
-%\end{listing}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection{Leaf traits influence on moisture dynamics }
+\subsection{Species specific moisture dynamics and leaf traits}
 
-The species in this study displayed different patterns of both moisture absorption and desorption (with Q.
-kelloggii and the pines exhibiting higher values for both), which were controlled by different litter traits (leaf thinness and litter bulk density, respectively).
-Q.
-kelloggii was the species that absorbed the most moisture, followed by the pines.
-The species that have a greater absorptive capacity are also the species that are usually more flammable when dry.
-This indicates that following a rain event, the most flammable species in my study would not be available as fuel immediately, but cannot be discounted as potential fuel.
-A species capacity to absorb moisture is controlled by leaf particle geometry, specifically the ratio between leaf length and leaf thickness, and not litter bed arrangement.
-The eight species differed in their desorption capacity.
-Q.
-kelloggii and the pines had higher desiccation rates, while C.
-decurrens exhibited lower desiccation rates.
-The ANOVA results indicate that leaf thinness best explains variation in maximum water retention, while litter bulk density best explains variation in desiccation rates.
-
+The species in this study displayed different patterns of both moisture absorption and desorption, which were controlled by different litter traits (specific leaf area and litter bulk density, respectively).
+\emph{Q.
+kelloggii} was the species that absorbed the most moisture, followed distantly by the finest-needled pine, \emph{P.
+lambertiana}.
+These species that have a greater absorptive capacity are also among the most flammable in terms of spread rate when dry.
+One presumption of ours prior to this experiment was that the high flammability previously reported for \emph{Q.
+kelloggii} \cite{Magalhaes+Schwilk-2012} would disappear under higher fuel moisture contents.
+This was partly supported: \emph{Q.
+kelloggii} fails to ignite until after 72 hours of drying in this study (Figs.
+\ref{fig:spread_time} and \ref{fig:consume_time}).
+However, high moisture content in this oak was unable to suppress ignition of mixtures to which it contributed (Figs.
+\ref{{fig:mixture_obs_pred_spread} and \ref{{fig:mixture_obs_pred_consume}).
 NOTE: P lambertiana higher water absorption, presumably due to higher SLA.
 
-Bulk densities measured here closely match those for these and other species \cite{van_Wagtendonk+Sydoriak+etal-1998,Stephens+Finney+etal-2004}
-
+With eight species, power to detect species-specific trait effects on these moisture patterns was weak, but we examined one litter trait known to directly influence flammability: litter bulk density.
+Litter density is largely driven by leaf size in this system \cite{Magalhaes+Schwilk-2012}.
 The overall dry-down process of litter is being controlled by both traits intrinsic to the leaf particles (leaf geometry) and traits intrinsic to the litter bed (bulk density).
@@
 
-TODO
-
-With eight species, power to detect species-specific trait effects on these moisture patterns was weak, but we examined one litter trait known to directly influence flammability: litter bulk density.
-Litter density is largely driven by leaf size in this system \cite{Magalhaes+Schwilk-2012}.
-
-\subsection{Time since wetting and flammability}
+T\subsection{Time since wetting and flammability}
 
@@
 Time since wetting, as expected, lowered the flammability of the eight species in this study and its effect overwhelms the effect of volatile content.
-Q.
-kelloggii and the pines are more flammable at higher moisture contents when compared to volatile-rich species such as C.
-decurrens, Abies spp.
-or S.
-giganteum.
+The pines are more flammable at higher moisture contents when compared to volatile-rich species such as \emph{C.
+decurrens}, \emph{Abies spp}.
+or \emph{S.
+giganteum}.
 This effect is influenced by litter bulk density and is likely to persist in fuel mixtures where the constituent species have different desiccation indices and therefore will have different moisture contents.
-I have determined moisture of extinction for the eight species in this study.
+
+% I have determined moisture of extinction for the eight species in this study.
 As expected, species that have higher flammability values (Pinus spp.) had higher moisture of extinction than species with lower flammability (Abies spp.

### 2019-07-24 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Solve units issue with spread rate
It turns out that one of the metadata files stored with data had
incorrect information on units. Also, the description of some variables
was incorrect. A root issue is that the actual raw data is all missing,
the data as committed by Rita already includes calculations such as
spread rate and is missing the raw flaming start and end times, etc. I
must assume those simple calculations are correct, I am sure they are,
but it would be best to have the raw data with stopwatch times.

@@
   release moisture through time is critical to the assessment of fuel’s
-  availability to burn. Natural forests are often comprised of mixed stands,
-  leading to mixtures of leaf litter on the surface fuelling the fires.
-  Multiple studies have established the existence of non-additive effects in
-  the flammability of dry litter, but we lack a mechanistic understanding of
-  moisture synergies in mixed litter and of moisture effects on the
-  flammability of mixed litter. The litter from eight species of a
+  availability to burn. Furthermore, Natural forests are often comprised of
+  mixed stands, leading to mixtures of leaf litter on the surface fuelling the
+  fires. Multiple studies have established the existence of non-additive
+  effects in the flammability of dry litter, but we lack a mechanistic
+  understanding of moisture synergies in mixed litter and of moisture effects
+  on the flammability of mixed litter. The litter from eight species of a
   mixed-conifer forest was saturated and allowed to dry to determine moisture
-  absorption and desorption. Burn trials were performed in moist litter beds of
-  single species and of mixtures, to establish flammability response to the
-  dry-down process. Moisture dynamics vary across species with higher specific
-  leaf area being associated with higher maximum absorption and lower litter
-  bulk denisty being associated with faster drying rates. Species that produce
-  more aerated litter beds (such as \emph{Q. kelloggii} and \emph{Pinus}
-  species) have faster rates of litter desiccation. We examined two axes of
-  flammability (spread rate and fuel consumption) and found that these
+  absorption capacity and desorption rates. Burn trials were performed in moist
+  litter beds of single species and of mixtures, to establish flammability
+  response to the dry-down process. Moisture dynamics vary across species with
+  higher specific leaf area being associated with higher maximum absorption and
+  lower litter bulk density being associated with faster drying rates. Species
+  that produce more aerated litter beds (such as \emph{Q. kelloggii} and
+  \emph{Pinus} species) have faster rates of litter drying. We examined two
+  axes of flammability (spread rate and fuel consumption) and found that these
   differing moisture dynamics result in time since wetting having strongly
   differing effects on flammability across the major litter types typified by
-  different tree genera. Furthermore, we found that litter mixtures exhibited
-  consistent positive non additivity in flame spread rate and that fuel
-  consumption and flame spread rate both showed increasing positive non
-  additivity with time since wetting.
+  different tree genera. Finally, we found that litter mixtures exhibited
+  consistent non additivity in flame spread rate in fuel consumption. This non
+  additivity became more positive with fuel drying. Litter mixtures burn with
+  fire behavior more similar to that of the most flammable constituent species
+  and this effect increases as fuels dry.}
 
@@
 
-Fuel moisture content is an important, and often driving, determinant of surface fire behavior.
-Environmental conditions such as temperature, relative humidity, and wind, drive the moisture content of the fuel available to burn \cite{Kreye+Varner+Etal-2018.
-In particular, fine dead fuels respond very readily to environmental changes \cite{Nelson 2001}.
-Fuel moisture influences flammability and fire behavior through effects on ignition, spread rate, flame height, fuel consumption, and heat release, due to the high specific heat content of water \cite{Rothermel-1972}.
-Specifically, high moisture contents decrease the likelihood of ignition thereby reducing fire spread \cite{Nelson-2011}.
-Moisture acts as a heat-sink resulting in less energy available for propagation, and the resulting water vapor will both cool the flaming front and dilute flammable gases \cite{Albini-1976, Shafizadeh-1977}.
-Variation in moisture of living fuels has strong effects on flammability, but moisture of dead excised litter fuels can also vary.
-Moisture of litter fuels varies with climate, with seasons and with short term weather within a season.
-Therefore, the rate of drying and flammability at any particular moisture content are key parameters that influence the fuel flammability throughout the fire season \cite{Nelson 2001}.
+Fuel moisture content is an important, and often driving, determinant of
+surface fire behavior. Environmental conditions such as temperature, relative
+humidity, and wind drive the moisture content of the fuel available to burn
+\cite{Kreye+Varner+etal-2018}. In particular, fine dead fuels respond very
+readily to environmental changes \cite{Nelson-2001}. Fuel moisture influences
+flammability and fire behavior through effects on ignition, spread rate, flame
+height, fuel consumption, and heat release, due to the high specific heat
+content of water \cite{Rothermel-1972, Nelson-2011}. Moisture acts as a
+heat-sink resulting in less energy available for propagation, and the resulting
+water vapor will both cool the flaming front and dilute flammable gases
+\cite{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
+strong effects on flammability, but moisture of dead excised litter fuels can
+also vary. Moisture of litter fuels varies with climate, with seasons and with
+short term weather within a season. Therefore, the rate of drying and
+flammability at any particular moisture content are key parameters that
+influence fuel flammability throughout the fire season \cite{Nelson-2001}.
 
-Fire scientists have long categorized fuel by size classes that correspond to moisture exchange rates, although such simplified categories have long been recognized as not sufficient to describe species specific and decomposition effects on moisture dynamics \cite{Anderson-1985}.
-A variety of models describing moisture dynamics of fine fuels are used by fire scientists \cite[eg]{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001} that include at least temperature and relative humidity and most aim to describe equilibrium conditions although some include precipitation and solar radiation effects as well.
+Fire scientists often categorize fuel by size classes that correspond to moisture exchange rates, although such simplified categories have long been recognized as not sufficient to describe species specific and decomposition effects on moisture dynamics \cite{Anderson-1985}.
 Fuel moisture dynamics of leaf litter are dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds \cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}.
-In many litter fuel driven systems, natural lightning ignitions are concurrent with precipitation and therefore fire behavior can depend on rate of drying following a wetting event.
+A variety of models describing moisture dynamics of fine fuels are used by fire scientists \cite{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}.
+These models usually aim to describe equilibrium conditions and at least  include temperature and relative humidity although some include precipitation and solar radiation effects as well.
+Explicit models of fire behavior as a function of time since wetting could improve fire prediction, however.
+In many litter fuel driven systems, natural lightning ignitions are concurrent with precipitation and, therefore, fire behavior can depend on rate of drying following a wetting event.
 If species differ in the rate of drying, this may strongly influence relatively flammability across species at any particular time post-wetting.
-Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires.
+Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires \cite{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.
 
-In mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fuelled by plant litter, leaf traits can influence landscape scale fire behavior \cite{Schwilk+Caprio-2011}.
-Similarly, strong species specific effects on litter flammability are recorded for the forests tof the southeastern US \cite{Nowacki+Abrams-2008} where species-specific litter drying rates play a strong role in flammability feedbacks \cite{Kreye+Varner+etal-2013}.
 Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape \cite{Fonda+Belanger+etal-1998, Kane+Varner+etal-2008, Schwilk+Caprio-2011, Kreye+Varner+etal-2013}.
 Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936, Fons-1946, Anderson+Rothermal-1965}.
-However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be characterized by leaf traits, litter traits, or both.
+However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be driven by by leaf traits, ermegent litter litter traits, or both.
+Species specific effects on litter flammability are recorded for the forests of the southeastern US \cite{Nowacki+Abrams-2008} where species-specific litter drying rates play a role in flammability feedbacks \cite{Kreye+Varner+etal-2013}.
+Similarly, in mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fueled by plant litter, overstory tree composition drives variation in local fire behavior \cite{Schwilk+Caprio-2011} as a result of differential litter flammability cross tree species and as a result of synergistic interactions in multi-species litter mixtures \cite{Magalhaes+Schwilk-2012}.
+These flammability feedbacks may help explain historical fire frequency patterns, but current estimates of litter flammability in mixed conifer forests are based largely on burning dry litter.
 
@@
 After saturation, the differences in moisture content indicate different abilities to retain moisture.
-The ability of litter to absorb and desorb water should be dependent on leaf and/or litter geometry, however, which traits control these processes are largely unknown.
 We propose two different sets of traits responsible for variation in the two processes mentioned: leaf particle geometry and litter bulk density.
-Litter characteristics such as surface-area-to-volume ratio, litter bulk density, mass density, particle size and shape, and chemical traits influence moisture diffusion through litter.
-Litter packing has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits (such as leaf size, litter bulk density).
+Litter characteristics such as surface-area-to-volume ratio, litter bulk density, mass density, particle size and shape, and chemical traits influence moisture diffusion through litter [CITE].
+Litter packing has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits, especially leaf size \cite{Scarff+Westoby-2006}.
 Fuelbeds composed of larger leaves with lower moisture retainability and decomposition rates favor drying, thus increasing the probability of ignition and fire spread rate.
@@
 Such non-additive effects have been shown to be synergistic in dry litter fuels where the flammability of a mixture is driven by the most flammable species  \cite{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}.
-In most forests, leaf litter will be a mixture representative of the species in the stand.
 It is possible that similar non-additive effects may influence litter moisture content, especially if different leaf traits influence adsorptive capacity and drying rate.
-Blauw et al.\cite{Blauw+Wensink+etal-2015} suggests that moisture content can contribute to non-additive effects on heathland flammability, but the mechanism of this effect is not known.
+There is some evidence for non addiitvity of flammability in moist fuels and one study has found increasing non additivity with moisture \cite{Blauw+Wensink+etal-2015}.
 
-We investigated species trait effects on litter moisture dynamics and consequent litter flammability across eight tree species in a temperate fire-prone forest in California.
-We determined the rate of moisture loss, the moisture of extinction (the threshold representing the moisture value above which fire will not ignite or spread) and the influence of moisture on the flammability of eight species from a mixed-conifer forest in California.
-Previous work described the flammability of these species in monoculture and in mixture for dry litter \cite{Magalhaes+Schwilk-2012}.
-{TODO- more on oaks?]  Experimental burn trials of leaf litter at various moisture levels determined how time since wetting influenced two axes of flammability characterized by flame spread rate and total fuel consumed.
-The burn trials were also performed with multi-species litter mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects.
-We compared leaf litter traits with the measured moisture absorption and desorption rates.
-Finally, we tested if the non-additive effects on  reported in previous studies occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.
+We investigated species specific litter moisture dynamics across eight tree species in a temperate fire-prone forest in California.
+We determined the maximum water retention and rate of moisture loss and examined how these vary in relation to two leaf and litter bed traits.
+We then examined the influence of moisture on the flammability across these species and how flame spread rate and fuel consumption varied with time since wetting and by taxon.
+Experimental burn trials of leaf litter at various moisture levels determined how time since wetting influenced two axes of flammability characterized by flame spread rate and total fuel consumed.
+Previous work described the flammability of these species in monocultures and in mixtures for oven-dry litter \cite{Magalhaes+Schwilk-2012} and found consistent positive non additivity of flammability where fire behavior of mixtures was driven by the most flammable species.
+Here, we examined if this positive non additivity was maintained in mixtures across varying levels of fuel moisture.
+The burn trials were also performed with multi-species litter mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects as seen in oven-dried litter.
+We tested if the non-additive effects on  reported in previous studies occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.
 
@@
 Field sites for this study were located in Sequoia and Kings Canyon National Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in mixed-conifer forest.
-I chose eight tree species that are representative of this type of forest: \emph{Pinus jeffreyi} Grev.
+We chose eight dominant tree species that are representative of this type of forest: \emph{Pinus jeffreyi} Grev.
 \& Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl.
@@
 Buchholz.
-In this study, I consider litter the top layer of leaves and small twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the previous year (mostly undecomposed).
-I collected the leaf litter in the summer (mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the study area.
-Effort was made to collect from different populations (minimum of four, each population separated from the others by 9 to 32 km) for each species to control for plastic and ecotypic effects.
-An exception to this involved the litter from A.
-magnifica which is restricted to the higher elevations and could only be collected from one area, due to logistical constraints.
+In this study, we consider litter the top layer of leaves and small twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the previous year (mostly undecomposed).
+Leaf litter was collected in the summer (mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the study area.
+Effort was made to collect from different populations (minimum of four, each population separated from the others by 9 to 32\,km) for each species to control for plastic and ecotypic effects.
+An exception to this involved the litter from \emph{A.
+magnifica} which is restricted to the higher elevations and could only be collected from one area, due to logistical constraints.
 Within that area, \emph{A.
 magnifica} litter was collected from over 10 different individuals.
-For the remaining species, the collection involved 2-4 individual trees at each site, separated by at least 10 m.
-The litter was collect about 2 m away from the trunk to obtain a more uniform sample, because bark and heavier branches tend to fall closer to the tree trunk.
+For the remaining species, the collection involved 2-4 individual trees at each site, separated by at least 10\,m.
+The litter was collect about 2\,m away from the trunk to obtain a more uniform sample, because bark and heavier branches tend to fall closer to the tree trunk.
+All litter was air dried down to < 5\% fuel moisture for weighing and dividing into moisture trials.
+The moisture manipulations and subsequent burning trials were conducted in April-August 2015.
 
@@
 
-We produced dry-down curves to determine maximum water retainability and the rate of moisture content loss (dessication rate) for each species and for four distinct mixtures.
-These curves allowed us to  determine the rate at which a species loses moisture (approximated as
-exponential decay), and determine maximum water content following saturation
-and gravity draining. Samples from the eight species (six replicates per
-species) and four mixtures (5 replicates per mixture type) were placed in
-baskets (approximately 45 x 45 cm by 15 cm tall) constructed of 1 cm aluminum
-screen. Samples were 450 g of litter (dry weight) per trial, and litter depth
-was standardized to 10 cm by changing the horizontal dimensions of the basket.
-Weight was measured by using a balance sensitive to 0.1 g (model XS16001L,
-Mettler Toledo, Columbus, OH). Samples were immersed in plastic storage bins
-filled with water for 24 hours to fully saturate them. After saturation, the
-baskets were removed from the water, allowed to drain for 3 minutes, and
-samples weighed to measure maximum moisture content. The samples were then
-allowed to dry in a partially controlled environment kept at 21 C temperature
-and 30\% relative humidity, assessed via an iButton sensor (Maxim Integrated)
-placed over the samples. During the dry-down process three subsamples were
+We produced dry-down curves to determine maximum water retainability and the
+rate of moisture content loss for each species and for four distinct
+three-species mixtures. These curves allowed us to determine the rate at which
+a species loses moisture, approximated as exponential decay, and to determine
+maximum water content following saturation and gravity draining. Samples from
+the eight species (six replicates per species) and four mixtures (5 replicates
+per mixture type) were placed in baskets (approximately 45 x 45\,cm by 15\,cm
+tall) constructed of 1 cm aluminum screen. Samples were 450 g of litter (dry
+weight) per trial, and litter depth was standardized to 10\,cm by changing the
+horizontal dimensions of the basket. Weight was measured by using a balance
+sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples were
+immersed in plastic storage bins filled with water for 24 hours to fully
+saturate them. After saturation, the baskets were removed from the water,
+allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
+content. The samples were then allowed to dry in a controlled environment kept
+at 21\,C temperature and 30\% relative humidity, assessed via an iButton sensor
+(Maxim Integrated) placed over the samples. Litter baskets were arranged in a
+completely randomized design. During the dry-down process three subsamples were
 taken at 24 hour intervals and weighed. These subsamples were then oven dried
-for 24 hours at 100 C and re-weighed to assess moisture content. This was
-repeated until the final sample from a basket was less than10\% moisture.
+for 24 hours at 100\,C and re-weighed to assess moisture content. This was
+repeated until the final sample from a basket was less than 10\% moisture.
 
@@
 
-With only eight species in this study, our power to test leaf trait effects on the observed moisture dynamics is low, so we decided a priori to test the effect of one leaf trait likely to influence moisture absoorption, specific leaf area, and one litter trait known to directly effect flammability and likely to influence drying rate, litter bulk density.
-Specific leaf area was taken from leaf fresh leaf area and leaf dry mass from X individuals per species and was collected as part of an ealier study \cite{Magalhaes+Schwilk-2012}.
-Litter bulk density was obtained in this experiment for each basket by dividing the dry mass (450 g) by the estimated litter volume (basket width x length x litter depth).
+With only eight species in this study, our power to test leaf trait effects on
+the observed moisture dynamics was low, so we decided \emph{a priori} to test
+the effect of one leaf trait likely to influence moisture absoorption, specific
+leaf area, and one litter trait known to directly effect flammability and
+likely to influence drying rate, litter bulk density. Specific leaf area was
+taken from leaf fresh leaf area and leaf dry mass from 5 individuals per
+species and was collected as part of a previous study
+\cite{Magalhaes+Schwilk-2012}. Litter bulk density was obtained in this
+experiment for each basket by dividing the dry mass (450\,g) by the estimated
+litter volume (basket width x length x litter depth).
 
@@
 times at each moisture level. We used the initial dry-down experiment to guide
-our timeing so as to attempt to burn samples over the full range of moisture
-contents from 100\% to below 10\% moisture. To expand our data on the dry end
-of this gradient, we supplemented these data with data from an earlier
-experiment on well-dried litter \cite{Magalhaes+Schwilk-2012}. This
-supplemental data added five replicates per species at an oven dried moisture
-content of $<$ 5\%. We assigned these replicates a time since wetting value of
-144 hours equivalent to the longest time since drying for the wetted samples.
-Flammability was assessed using a 150 cm long burn table, 15 cm wide and 15 cm
-tall, in which leaf litter was placed and the table gently shaken to allow
-settling. This design allows fire to reach constant flame spread rates
-\cite{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced along
-the apparatus allowed visual determination of maximum flame height. Litter was
-ignited with a propane torch; maximum adiabatic flame temperature of 1899 C,
-http://www.benzomatic.com/products/fuel.aspx) at one end of the apparatus and
-allowed to burn until extinction. If a species at a specific moisture level
-failed to ignite after three attempts, no further attempts were made. A timer
-was used to record time to ignition and duration of flaming combustion. The
-flammability trials followed the methodology detailed in de Magalhães and
-Schwilk \cite{Magalhaes+Schwilk-2012}.
+our timeing so as to attempt to burn samples over a range of moisture contents.
+We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
+preliminary trials on damp fuels suggested that across these fuels ignition is
+unlikely above 100\% fuel moisture. To expand our data on the dry end of this
+gradient, we supplemented these data with data from an earlier experiment on
+oven-dried litter \cite{Magalhaes+Schwilk-2012}. This supplemental data added
+five replicates per species at an oven dried moisture content of $<$ 5\%. We
+assigned these replicates a time since wetting value of 144 hours equivalent to
+the longest time since drying for the wetted samples. Flammability was assessed
+using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which leaf
+litter was placed and the table gently shaken to allow settling. Burning trials
+all used 450\,g (dry weight) of litter. For three-species mixtures, this was
+150\,g per species. This design allows fire to reach constant flame spread
+rates \cite{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced
+along the apparatus allowed visual determination of maximum flame height.
+Litter was ignited with a propane torch; maximum adiabatic flame temperature of
+1899 C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the
+apparatus and allowed to burn until extinction.
+If a sample failed to ignite after 30 seconds, no further attempt was made and if three samples of a species at a particular mositure level failed to ignite, no further attempts were made on that species/moisture combination and the replicate was allowed to continue drying.
+A timer was used to record total time to ignition and duration of flaming combustion.
+The flammability trials followed the methodology detailed in de
+Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.
 
@@
 length of the burned surface by the time it takes the flaming front to reach
-the end of the apparatus, maximal flame height (mm), and mass loss (\%),
-calculated as the difference between fuel mass before and after flame
-extinction. Some trials resulted only in smoldering combustion; for those
-trials, a time to ignition and mass loss were recorded, but flame spread rate
-and flame height were recorded as zero. Our thermocouple system failed or data
-was lost for two days of burning trials, therefore we have omitted those data
-from this analysis. Our past work \cite{Magalhaes+Schwilk-2012} demonstrated
-that flame spread rate captures one important axis of flammability variation
+the end of the apparatus, maximal flame height (mm), and fuel consumption (\%).
+Some trials resulted only in smoldering combustion; for those trials, a time to
+ignition and mass loss were recorded, but flame spread rate and flame height
+were recorded as zero. Our thermocouple system failed or data was lost for two
+days of burning trials, therefore we have omitted those data from this
+analysis. Our past work \cite{Magalhaes+Schwilk-2012} demonstrated that flame
+spread rate captures one important axis of flammability variation
 \citep{Schwilk-2015, Pausas+Keeley+etal-2017} and percent fuel consumed
@@
 not as well as integrating temperature over time does). Our analysis,
-therefore, focused on these two relatively orthogonal measures of flammability.
+therefore, focused on these two relatively orthogonal measures of flammability
+to avoid examining multiple redundant dependent variables.
 
 Burn trials occurred in a cement structure used to simulate house fires at the
-Fire Department of the City of Lubbock. This structure minimizes wind and helps
+Fire Department of the City of Lubbock. This structure minimized wind and helped
 regulate temperature and relative humidity. Temperature and relative humidity
 were measured before every trial using a Kestrel 3000 (Nielsen-Kellerman,
-Boothwyn, PA). To minimize temperature and relative humidity variation, all
-trials were conducted on clear days between 10 am and 3 pm.
+Boothwyn, PA). All trials were conducted on clear days between 10 am and 3 pm.
 
@@
 Kenward-Roger approximation \cite{Kenward_Roger-1997} which is recommended by
-Luke \cite{Luke-2017} as producing acceptable Type 1 error rates.
-
-We tested if one leaf trait (specific leaf area) and one litter trait (litter
-bulk density) influenced moisture dynamics (maximum water retainability and
-desiccation rate) by fitting linear models with species means as observations.
-For each response variable we fit a single linear model with two predictors and
-an interaction term.
+Luke \cite{Luke-2017} as producing acceptable Type 1 error rates. We tested if
+one leaf trait (specific leaf area) and one litter trait (litter bulk density)
+influenced moisture dynamics by fitting linear models with species means as
+observations. For each response variable we fit a single linear model with two
+predictors and an interaction term.
 
@@
 flammability across these eight species was driven by leaf traits and litter
-packing. Due to similar leaf traits, species within a genus tended to behave
-similarly and the two scale-leaved genera in the Cupressaceae,
-\emph{Calocedrus} and \emph{Sequoiadendron} also behaved similarly to one
-another. Therefore, we used taxonomic groups (\emph{Abies}, Cupressaceae,
-\emph{Quercus}, and \emph{Pinus}) as our flammability functional groups and
-treated these as fixed effects in our models. The effect of fuel moisture and
-time since wetting on flammability parameters were modeled using mixed effect
+packing. These previously reported flammability trials demonstrated that species within a genus tended to behave similarly and the two scale-leaved genera in the Cupressaceae,
+\emph{Calocedrus} and \emph{Sequoiadendron}, also behaved similarly to one
+another.
+Therefore, to preserve power and focus on analyses we used taxonomic groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our flammability functional groups and treated these as fixed effects in our models.
+The effect of fuel moisture and
+time since wetting on flammability parameters were modeled using linear
 models with fuel moisture or time since wetting and taxon as fixed effects and
 average vapor pressure deficit in the burning room during that trial as a
-nuisance covariate.
+nuisance covariate. We used linear mixed models to predict spread rate, but
+fuel consumption tends to exhibit threshold effects with few intermediate
+values. Therefore, we used mixed generalized linear models with a beta
+regression link with the `glmmADMB` package \cite{Skaug_Fournier_etal-2016} to
+model percent fuel consumed as a function moisture content and of time since
+wetting.
 
@@
 by testing if moisture content and flammability of mixtures were predicted by
-the average of the constituent species. To choose mixtures we first grouped the
+the average of the constituent species. To choose mixtures, we first grouped the
 species according to their maximum water retainability and dessication rate
@@
 
-There are marked differences in the extent of moisture species retain initially
+There were marked differences in the extent of moisture species retain initially
 following wetting and draining by gravity (all p < 0.0001, Fig.
 \ref{fig:drydown} and Table \ref{tab:drydown}). \emph{Q. kelloggii}, the only
-broadleaf, is the species that retains the most moisture initially.
-\emph{Quercus}, however, like the pines has a rapid dessication rate (Fig.
+broadleaf, is the species that retained the most moisture initially.
+\emph{Quercus}, like the pines, had a rapid drying rate (Fig.
 \ref{fig:drydown} and SI Fig. 1). Calculating estimated marginal means following the
@@
   gravity for 3 min before initial weighing. Colors indicate genera and there
-  were significant differences in both. A small amount of horizontal jitter
+  were significant differences in both maximum water retention and in rate of drying (based linear model on log of water content). A small amount of horizontal jitter
   (0.5 hr) was added to this plot to aid in visualizing because, although
@@
 \begin{table}[H]
-  \caption{Mixed model results for moisture content as a function of time since
+  \caption{Linear mixed model results for moisture content as a function of time since
     wetting. Approximate degrees of freedom, pseudo F statistics and p-values
-    were calculated by the Kenward-Roger approximation
+    were calculated by the Kenward-Roger approximation.
     \cite{Kenward_Roger-1997}.}
@@
 bulk density had a negative effect on desiccation rate (Fig. \ref{fig:bd-di},
-linear model p = 0.029). Anova results are shown in tables \ref{tab:mc_anova}
-and \ref{tab:di_anova}.
+linear model p = 0.029). Anova results are shown in Rables \ref{tab:mc_di_anova}.
 
@@
 \includegraphics[width=16cm]{figs_tables/fig2_SLA_maxMC.pdf}
-\caption{Maximum water retention by specific leaf area (SLA) across eight species.}
+\caption{Maximum water retention by specific leaf area (SLA) across eight species. ANOVA table in Table \ref{tab:mc_di_anova}}
   \label{fig:maxmc-di}
 \end{figure}
-
-\begin{table}[H]
-  \caption{Linear model results for maximum moisture content as a function of specific leaf area and litter bulk density. Mixed models were fit as a mixed beta regression. }
-  \label{tab:mc_anova}
-\centering
-%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-\input{figs_tables/tab2_mc_anova.ltx}
-\end{table}
 
@@
 \includegraphics[width=16cm]{figs_tables/fig3_di_bd.pdf}
-\caption{Linear model predicting dessication rate as a function of specific leaf area and litter bulk density across eight species.}
+\caption{Linear model predicting dessication rate as a function of specific leaf area and litter bulk density across eight species. ANOVA table in Table \ref{tab:mc_di_anova}}.
   \label{fig:bd-di}
@@
 \begin{table}[H]
-  \caption{Linear model results for estimated maximum water retention (\%) as a function of specific leaf area and litter bulk density.}
-  \label{tab:di_anova}
+  \caption{Linear model results for maximum water retention (\% moisture content on dry mass basis) and for estimated desiccation rate (hr$^{-1}$) as functions of specific leaf area and litter bulk density across eight species.
+ANOVA tables shown with degrees of freedom, sum of squares, mean squares, F, and p for each effect.}
+  \label{tab:mc_di_anova}
 \centering
 %% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-\input{figs_tables/tab2_mc_anova.ltx}
+
+\begin{tabular}{ccrrrrr}
+  \toprule
+Ind. variable & Effect & df & Sum of Squares & Mean squares & F & p value \\ 
+  \midrule
+Moisture content & SLA & 1 & 250834.5 & 250834.51 & 130.13 & 0.000 \\ 
+  & density & 1 & 538.5 & 538.54 & 0.28 & 0.625 \\ 
+  & SLA:density & 1 & 5135.2 & 5135.24 & 2.66 & 0.178 \\ 
+  & Residuals & 4 & 7710.4 & 1927.61 &  &  \\ 
+  \midrule
+
+Desiccation rate & density & 1 & $3.92 \times 10^{-4}$ & $3.92 \times 10^{-4}$ & 11.01 & 0.029 \\ 
+ &  SLA & 1 & $1.88 \times 10^{-5}$ & $1.88 \times 10^{-5}$ & 0.53 & 0.507 \\ 
+ &  density:SLA & 1 & $6.89 \times 10^{-6}$ & $6.89 \times 10^{-6}$ & 0.19 & 0.683 \\ 
+ &  Residuals & 4 & $1.42 \times 10^{-4}$ & $3.56 \times 10^{-5}$ &  &  \\ 
+   \bottomrule
+
+\end{tabular}
 \end{table}
 
-\begin{table}[H]
-  \caption{Linear model results for estimated dessication rate ($hr^{-1}$) as a function of specific leaf area and litter bulk density.}
-  \label{tab2:di-anova}
-\centering
-%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-\input{figs_tables/tab3_di_anova.ltx}
-\end{table}
+% \begin{table}[H]
+%   \caption{Linear model results for estimated dessication rate ($hr^{-1}$) as a function of specific leaf area and litter bulk density.
+Anova table shown with degrees of freedom, sum of squares, mean squares, F, and p for each effect}
+%   \label{tab:di_anova}
+% \centering
+% %% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
+% \input{figs_tables/tab3_di_anova.ltx}
+% \end{table}
 
@@
 Flame spread rate declined with fuel moisture (Fig.
-\ref{fig:spread_moist}.
+\ref{fig:spread_moist}).
 \emph{Quercus} had higher spread rates across all fuel moistures and had a steeper decline in response to moisture than did the other groups (taxon by time since wetting interaction p = 0.022, Table \ref{tab:spreadrate_anova}).
-Fuel moisture negatively influenced fuel consumption and overall  fuel consumption as well as the rate of change with moisture content varied across the taxonomic group (Fig.
+Fuel moisture negatively influenced fuel consumption.
+Overall fuel consumption and the rate of change with moisture content varied across the taxonomic groups (Fig.
 \ref{fig:consume_moist}, Table \ref{tab:consume_moist_anova}).
-Although the overall ranking of groups was simlar for fuel consumption and for flame spread rate, in contrast to the pattern with spread rate, for fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in flammability with fuel moisture than did the other groups.
+Although the overall ranking of groups was similar for fuel consumption and for flame spread rate, in contrast to the pattern with spread rate, for fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in flammability with fuel moisture than did the other groups.
 
@@
 \begin{table}[H]
-  \caption{Linear mixed model results for flame spread rate as a function of fuel moisture with taxonomic group as a fixed effect and species as a nested random effect.
+  \caption{Generalized linear mixed model results for flame spread rate as a function of fuel moisture with taxonomic group as a fixed effect and species as a nested random effect.
 Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.
@@
 \centering
-\input{figs_tables/tab3_spread_moist_anova.ltx}
+\input{figs_tables/tab4_spread_moist_anova.ltx}
 \end{table}
@@
 \centering
-\input{figs_tables/tab4_consume_moist_anova.ltx}
+\input{figs_tables/tab5_consume_moist_anova.ltx}
 \end{table}
@@
 %% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-\input{figs_tables/tab5_spread_time_anova.ltx}
+\input{figs_tables/tab6_spread_time_anova.ltx}
 \end{table}
@@
 Fuel consumption was highly variable across trials and intermediate values were less common than very low are moderately high consumption (Fig.
-\ref{fig:spread_time}).
-All parameters significantly contributed to the model as indicated by Table \ref{tab:consume_time_anova}.
+\ref{fig:consume_time}).
+All parameters significantly contributed to the generalized linear mixed model as indicated by the nested model comparisons in Table \ref{tab:consume_time_anova}.
 As with flame spread rate some pines had the most rapid response to drying.
@@
 lambertiana} in its own group allowed model convergence that was otherwise impossible.
+This breaks our intent to use only \emph{a priori} taxonomic groups, but we could not obtain model convergence otherwise.
 
@@
 \begin{table}[H]
-  \caption{Generalized linear mixed model results for fuel consumption as a function of time since wetting with taxonomic group as a fixed effect and species as a nested random effect.
+  \caption{Generalized linear mixed model results for fuel consumption as a function of time since wetting with taxonomic group as a fixed effet and species as a nested random effect.
+The table shows sequential chi-square tests on nested beta regression models.
+The analysis of deviance table below shows deviance comparisons for the simplest null model with only intercept and random effect (model 1), for a taxon effect only (model 2), for a taxon and mosture content effect (model 3) and for the full model including the taxon by moisture interaction (model 4).
+Coefficient estimates for the full model are in SI Table 5.}
+  \label{tab:consume_time_anova}
+\centering
+\input{figs_tables/tab7_consume_time_anova.ltx}
+\end{table}
+
+% \begin{table}[H]
+%   \caption{Generalized linear mixed model results for fuel consumption as a function of time since wetting with taxonomic group as a fixed effect and species as a nested random effect.
 These are beta regression models.
@@
 Coefficient estimates are in SI Table 5.}
-  \label{tab:consume_time_anova}
-\centering
-\input{figs_tables/tab6_consume_time_anova.ltx}
-\end{table}
+%   \label{tab:consume_time_anova}
+% \centering
+% \input{figs_tables/tab6_consume_time_anova.ltx}
+% \end{table}
 
@@
 Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.
-Moistures were measured every 24 hours but data are shown with slight horizontal jitter to reduce point overlap.}
+Moistures were measured every 24 hours but data are shown with slight horizontal jitter to reduce point overlap.
+Mixed linear model results shown in Table \ref{tab:mixtures_drydown}.}
   \label{fig:mixture_dry_down}
@@
 \ref{fig:mixture_dry_down}) has the lowest initial retention.
-Although intiial water retention differed across mixtures (RESULTS?), there was no significant difference in dry down rates (slopes in linear model based on log of moisture content).
+Initial water retention and rates differed across mixtures (linear model based on log of moisture content, Table \ref{tab:mixtures_drydown}).
+
+\begin{table}[H]
+  \caption{Linear mixed model results for moisture content as a function of time since
+    wetting for four different litter mixture types. Approximate degrees of freedom, pseudo F statistics and p-values
+    were calculated by the Kenward-Roger approximation
+    \cite{Kenward_Roger-1997}.}
+  \label{tab:mixtures_drydown}
+\centering
+%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
+\input{figs_tables/tab8_mixtures_drydown_anova.ltx}
+\end{table}
 
@@
 
-There was some evidence for non additive in drying rate a tendency for litter mixtures.
+There was evidence for non additive effects in drying rate.
 Mixtures tended to have lower water content than predicted from the mean of the constituent species after wetting, but higher water content than expected after long drying (Wald test shows slope of mixed linear model illustrated in Fig.
@@
 \caption{Observed vs predicted flame spread rate for litter mixtures.
-Y axis shows relative spread rate values (predicted - observed) so that positive values indicate faster than predicted and negative slower than predicted.
-Predictions are based on the modeled spread rate at each particular time since wetting for the three constituent species in each mixture.
+Filled circles show observed spread rate values.
+Open diamonds indicate predicted spread rates based on the average spread rate at each particular time since wetting for the three constituent species in each mixture.
 Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
@@
 \end{figure}
+
+% \begin{figure}[H]
+%   \centering
+% \includegraphics[width=16cm]{figs_tables/fig11_mixture_obs_vs_pred_consume.pdf}
+% \caption{Observed vs predicted fuel consumption for litter mixtures.
+Predictions are based on the modeled spread rate at each particular time since wetting for the three constituent species in each mixture.
+Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
+%   \label{fig:mixture_obs_pred_consume}
+% \end{figure}
 
@@
 \caption{Observed vs predicted fuel consumption for litter mixtures.
-Predictions are based on the modeled spread rate at each particular time since wetting for the three constituent species in each mixture.
+Filled circles show observed fuel consumption percentages.
+Open diamonds indicate predicted fuel consumption based on the average fuel consumption at each particular time since wetting for the three constituent species in each mixture.
 Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
@@
 Flame spread rate was consistently slightly higher than predicted (Fig \ref{fig:mixture_obs_pred_spread}, Wald test of mixed linear model centered intercept for observed-predicted values greater than zero, p = 0.017) and the positive non additivity increased with time since wetting (slope greater than zero, p = 0.005).
-Fuel consumption, however, showed no overall bias towards positive or negative non additivity, but did show a positive relationship with time since wetting (Fig.
+Fuel consumption, however, showed no overall bias towards positive or negative non additivity, but non addivity (difference between observed and expected values) did show a positive relationship with time since wetting (Fig.
 \ref{fig:mixture_obs_pred_consume}, p < 0.0001).
@@
 
-\subsection{Species specific moisture dynamics and leaf traits}
+\subsection{Species specific moisture dynamics flammability}
 
@@
 kelloggii} \cite{Magalhaes+Schwilk-2012} would disappear under higher fuel moisture contents.
-This was partly supported: \emph{Q.
+This prediction was partly supported: \emph{Q.
 kelloggii} fails to ignite until after 72 hours of drying in this study (Figs.
@@
 However, high moisture content in this oak was unable to suppress ignition of mixtures to which it contributed (Figs.
-\ref{{fig:mixture_obs_pred_spread} and \ref{{fig:mixture_obs_pred_consume}).
-NOTE: P lambertiana higher water absorption, presumably due to higher SLA.
+\ref{fig:mixture_obs_pred_spread} and \ref{fig:mixture_obs_pred_consume}).
 
-With eight species, power to detect species-specific trait effects on these moisture patterns was weak, but we examined one litter trait known to directly influence flammability: litter bulk density.
-Litter density is largely driven by leaf size in this system \cite{Magalhaes+Schwilk-2012}.
-The overall dry-down process of litter is being controlled by both traits intrinsic to the leaf particles (leaf geometry) and traits intrinsic to the litter bed (bulk density).
-Decomposition can also affect moisture content in litter.
-Van Wagner (1969), investigating fuel weathering, found that pine needles had a variation in moisture content 3 to 6 times greater when they lacked their protective wax coating when compared to intact needles.
+    The dry down curves we measured demonstrate that after a rain event, the litter of different species will exhibit different moisture contents over time.
+Our flammability trials demonstrate that that both flame spread rate and fuel consumption increased with time since wetting but that the rate of increase varied across these genera.
+This indicates that there is a taxon-specific response to flammability based on time since wetting.
+Only the pines are able to ignite and support flame spread immediately after wetting (Fig.
+\ref{fig:spread_time}).
+\emph{Quercus} exhibited a threshold response to drying with no ignition prior to 72 hours of drying but once ignition is possible, having faster spread rates than the other genera.
 
-T\subsection{Time since wetting and flammability}
+With eight species, power to detect species-specific trait effects on these moisture patterns was weak, but we examined one litter trait known to directly influence flammability: litter bulk density and one leaf trait, specific leaf area.
+Litter density is largely driven by leaf size in this system  and in turns controls the major axis of flammability variation in dry litter \cite{Magalhaes+Schwilk-2012}.
+Our results suggest that moisture dynamics are controlled by both traits intrinsic to the leaf particles (specific leaf area) and emergent traits of the litter bed (bulk density).
+Specific leaf area here is probably capturing some joint effect of cuticle thickness and total surface area of litter particles.
+The oak has much higher initial water retention than the other species, followed by the finest-needled pine, \emph{P.
+lambertiana}.
+With only a handful of species, however, disentangling mechanism of water absorption is not possible.
+Our study examines fresh litter, but we expect that decomposition will affect moisture content.
+For example, Van Wagner \cite{Van_Wagner-1969}, investigating fuel weathering, found that pine needles had a variation in moisture content 3 to 6 times greater when they lacked their protective wax coating when compared to intact needles.
 
-After a rain event, the litter of different species will exhibit different moisture contents.
-These results show that both time to ignition and spread rate were affected by time since wetting, and the interaction of time since wetting and genus was also significant.
-
-This indicates that there is a species-specific response to flammability based on time since wetting, and the differences are due to different moisture contents at each time interval.
-Therefore, mere moisture content will not inform on the fire behavior measures of a litter bed.
+% Therefore, mere moisture content will not inform on the fire behavior measures of a litter bed.
 Species composition of the litter must be considered, and time since the last rain event should allow the determination of the potential flammability of an area.
 
-Time since wetting, as expected, lowered the flammability of the eight species in this study and its effect overwhelms the effect of volatile content.
-The pines are more flammable at higher moisture contents when compared to volatile-rich species such as \emph{C.
+% The pines are more flammable at higher moisture contents when compared to volatile-rich species such as \emph{C.
 decurrens}, \emph{Abies spp}.
@@
 We tested the dry-down process and flammability of four unique three-species litter mixtures.
-These mixtures contained species representative of all dry-down behavior types observed in single species trials, as assessed by the Tukey test.
-As expected from the single species results, mixtures that contained Q.
-kelloggii absorbed more water during the saturation period, but there was no difference in desiccation rate among the mixtures.
-However, when investigating the non-additivity in moisture content, the mixture that did not contain Q.
-kelloggii retained more moisture than would be predicted by the null model (arithmetic average of the three constituent species in the mixture).
-In general, non-additivity varied with moisture content, with positive non-additive behavior at lower moisture contents (<75\%, 72-96 hours since wetting occurred) and negative non-additive behavior at very high moisture contents (>200\%, 0 hours since wetting occurred).
-This result indicates that immediately after a rain event, the moisture content of a mixture will be lower than expected, however these high values for moisture content are unlikely to be found in natural litters.
+These mixtures contained species representative of all dry-down behavior types observed in single species trials.
+As expected from the single species results, mixtures that contained \emph{Q.
+kelloggii} absorbed more water during wetting, and the mixture without any Cupressaceae showed the most rapid dessication (Figure \ref{fig:mixture_dry_down} and Table \ref{tab:mixtures_drydown}).
+Drydown rates did not behave as the average of the constituent species and mixtures tended to be drier than expected early in the process and wetter than expected later (Fig.
+\ref{fig:mixture_obs_pred_mc}).
+The fact that mixtures tended to have lower water content than predicted from the mean of the constituent species after wetting, but higher water content
+than expected after long drying is surprising because past work has shown that
+mixtures tend to pack less densely than do litter monocultures \cite[][, Table 2]{Magalhaes+Schwilk-2012}. One would expect that the lower bulk density of mixtures would speed drying.
+
+% This result indicates that immediately after a rain event, the moisture content of a mixture will be lower than expected, however these high values for moisture content are unlikely to be found in natural litters.
 As the mixture dries, the dry-down behavior will approximate the null model, and three to four days after the rain event the dry-down behavior will become non-additive again.
-The positive non-additivity seems influenced by litter bed characteristics, since the strongest non-additivity was observed for the mixture with two species that have less dense litter: Q.
-kelloggii and P.
-jeffreyi (CaPiQu).
 
-The fact that mixtures tended to have lower water content than predicted from
-the mean of the constituent species after wetting, but higher water content
-than expected after long drying is a bit surprising as past work has shown that
-mixtures tend to be more aerated than expected given the bulk density of litter
-of individual species \citep{Magalhaes+Schwilk-2012}. This non additivity in
-drying rate did not fully explain non additivity in flammability, however.
+The overall ranking of flammability in litter mixtrures was consistent with the
+results obtained for the individual species. Mixtures containing \emph{Q.
+  kelloggii} and \emph{P. jeffreyi} had fast spread rates. These results are
+likely influenced by the lower bulk densities of litter mixtures containing
+those two species. Most interestingly, there were striking pattern of nan
+additivity in spread rate and in fuel consumption for the four mixtures in this
+study, indicating that the positive non additivity is in flammability is not
+restricted to completely dry litter mixtures \citep{Magalhaes+Schwilk-2012,
+  VanAltena+Logtestjin+etal-2012}. Non additivity in both spread rate fuel
+consumption became more positive with time since wetting and these patterns
+were consistent across mixture identities. This is a slightly different result
+than that reported by Blauw et al \citeyear{Blauw+Wensink+etal-2015}. Blauw et
+al found that non-addivity was greatest in moist mixtures of heathland fuels.
+In their work this non addiitvity was generally negative but that positive non
+addivity in mixtures was often driven by mixtures meeting ignition thresholds
+not met by the constituent species alone. To limit the number of models fit, we
+limited our investigation to two flammability parameters that represent two
+major axes of flammability variation. Therefore, our study does not fully
+capture ignition effects although we do have some results consistent with
+theirs: mixtures with \emph{Q. kellogii} could ignite at times since wetting at
+which the species in monoculture could not. Overall, our study does show a consistent pattern of increasingly positive non addiitvity in flammability with litter drying. This result, together with
+previously published work, suggests that litter flammability is prone to
+non-additive effects, reinforcing the idea that we cannot approach flammability
+by lumping species into fuel categories, as is common in fire behavior models.
+Community flammability is dependent on and particular to its constituent
+species and cannot be addressed by simply averaging the flammability of all
+species it comprises. This dynamic view is likely to be increasingly relevant
+considering ongoing climate change, which predicts changes in community
+assembly.
 
-The flammability results for the mixtures followed closely the results obtained for the individual species.
-Mixtures containing Q.
-kelloggii and P.
-jeffreyi had faster spread rates and lower times to ignition, resulting in higher overall flammability.
-These results are influenced by the lower bulk densities of litter mixtures containing those two species.
-There was a non-additive effect of moisture on flammability for the four mixtures in this study, reinforcing results shown for dry litter \citep{Magalhaes+Schwilk-2012, VanAltena+Logtestjin+etal-2012}.
-Spread rate and time to ignition had significant non-additivity, and were strongly influenced by the species present in the mixture.
-Presence of C.
-decurrens in a mixture marked a big difference between the null model and the prediction.
-This was due to several failed ignitions for this species when burned individually, which conditioned the predicted values, while all the mixtures ignited.
-There were marked differences in the non-additivity results from this study and the results from a previous study investigating dry litter samples.
-The presence of non-additive effects in moisture content leads to non-additive effects in the flammability of moist litter beds.
-When accounting for this effect, the non-additivity on the flammability of moist litter beds is subtler, when compared to species effects.
-This result, together with previously published work, suggests that litter flammability is prone to non-additive effects, reinforcing the idea that we cannot approach flammability by lumping species into fuel categories, as is common in fire behavior models.
-Community flammability is dependent on and particular to its constituent species and cannot be addressed by simply averaging the flammability of all species it comprises.
-This dynamic view is likely to be increasingly relevant considering ongoing climate change, which predicts changes in community assembly.
+[CONCLUSIONS ?]
 

### 2019-07-24 ms/SI.tex
Solve units issue with spread rate
It turns out that one of the metadata files stored with data had
incorrect information on units. Also, the description of some variables
was incorrect. A root issue is that the actual raw data is all missing,
the data as committed by Rita already includes calculations such as
spread rate and is missing the raw flaming start and end times, etc. I
must assume those simple calculations are correct, I am sure they are,
but it would be best to have the raw data with stopwatch times.

@@
 \begin{table}[h]
-  \caption{Mixed model coefficients for flame spread rate. Table shows mixed linear model results  TODO.} 
+  \caption{Mixed model coefficients for flame spread rate. Table shows mixed linear model results.} 
   \label{tabS3}
 \centering
-\input{figs_tables/spreadrate-coef-tab.ltx}
+\input{figs_tables/SI_tab2_spread_moist_coef.ltx}
 \end{table}
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
   \centering
-  \includegraphics[width=8cm]{figs_tables/SI_fig1_drydown_curves_logged.pdf}
+  \includegraphics[width=8cm]{figs_tables/SI_fig1_drydown-curves_logged.pdf}
    \label{SI_fig1}

### 2019-07-24 scripts/moisture-ms-figs-tables.R

@@ -60,7 +60,7 @@ ggsave(file.path(RESULTS, "SI_fig1_drydown-curves_logged.pdf"), plot=SI_fig1,
 ###############################################################################
 ## Table 1: drydown by species ANOVA
 ###############################################################################
-tab1 <- nice(dry.mixed)
+tab1 <- nice(dry.mixed, sig_symbols = rep("", 4))
 names(tab1)[4] <- "p value"
 #tab1 <- tab1[ ,2:4]
 tab1 <- xtable(tab1)
@@ -174,16 +174,25 @@ anova(di.mod)
 ###############################################################################
 
 tab2 <- anova(mc.mod)
-names(tab2)[5] <- "p value"
-print(xtable(tab2), file=file.path(RESULTS, "tab2_mc_anova.ltx"),  booktabs=TRUE, floating=FALSE,
+tab2$Effect <- rownames(tab2)
+tab2 <- tab2[,c(6,1,2,3,4,5)]
+names(tab2) <- c("Effect", "df", "Sum of Squares", "Mean squares", "F", "p value")
+tab2.x <- xtable(tab2, digits=c(0,0,0,1,2,2,3))
+#tab2.x[1,6] <- "<0.0001"
+print(tab2.x, file=file.path(RESULTS, "tab2_mc_anova.ltx"),  booktabs=TRUE, floating=FALSE,
       include.rownames=FALSE)
 
 ###############################################################################
 ## Table 3: dessication rateanova
 ###############################################################################
-tab3 <- anova(di.mod)
-names(tab3)[5] <- "p value"
-print(xtable(tab3), file=file.path(RESULTS, "tab3_di_anova.ltx"),  booktabs=TRUE, floating=FALSE,
+tab3 <- as.data.frame(anova(di.mod))
+tab3$Effect <- rownames(tab3)
+rownames(tab3) <- NULL
+tab3 <- tab3[,c(6,1,2,3,4,5)]
+names(tab3) <- c("Effect", "df", "Sum of Squares", "Mean squares", "F", "p value")
+tab3.x <- xtable(tab3, digits=c(0, 0 ,0,-2,-2, 2,3))
+print(tab3.x, file=file.path(RESULTS, "tab3_di_anova.ltx"), math.style.exponents=TRUE,
+      booktabs=TRUE, floating=FALSE,
       include.rownames=FALSE)
 
 
[…]

### 2019-07-24 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
edit abstract

@@
 %--------------------
+
 %----------
@@
   fires. Multiple studies have established the existence of non-additive
-  effects in the flammability of dry litter, but we lack a mechanistic
-  understanding of moisture synergies in mixed litter and of moisture effects
-  on the flammability of mixed litter. The litter from eight species of a
+  effects in the flammability of dry litter, but we lack an understanding of
+  moisture synergies in mixed litter and of moisture effects on the
+  flammability of mixed litter. The litter from eight species of a
   mixed-conifer forest was saturated and allowed to dry to determine moisture
   absorption capacity and desorption rates. Burn trials were performed in moist
-  litter beds of single species and of mixtures, to establish flammability
+  litter beds of single species and of mixtures to establish flammability
   response to the dry-down process. Moisture dynamics vary across species with
@@
   \emph{Pinus} species) have faster rates of litter drying. We examined two
-  axes of flammability (spread rate and fuel consumption) and found that these
+  axes of flammability variation (spread rate and fuel consumption) and found that these
   differing moisture dynamics result in time since wetting having strongly
@@
   fire behavior more similar to that of the most flammable constituent species
-  and this effect increases as fuels dry.}
+  and this effect increases as fuels dry. }
 
@@
 
-[CONCLUSIONS ?]
-
 % I have found non-additive effects (both positive and negative) in moisture content and in the flammability of wet litter mixtures of mixed-conifer forest.

### 2019-07-25 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Draft

@@
   fire behavior more similar to that of the most flammable constituent species
-  and this effect increases as fuels dry. }
+  and this effect increases as fuels dry. ADD Synthesis statement.}
 
@@
 strong effects on flammability, but moisture of dead excised litter fuels can
-also vary. Moisture of litter fuels varies with climate, with seasons and with
+also vary. Moisture of litter fuels varies with climate, with seasons, and with
 short term weather within a season. Therefore, the rate of drying and
@@
 
-Fire scientists often categorize fuel by size classes that correspond to moisture exchange rates, although such simplified categories have long been recognized as not sufficient to describe species specific and decomposition effects on moisture dynamics \cite{Anderson-1985}.
-Fuel moisture dynamics of leaf litter are dependent upon forest structure (shading and rain throughfall effects), litter physical structure, and chemical composition of the fuelbeds \cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}.
-A variety of models describing moisture dynamics of fine fuels are used by fire scientists \cite{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}.
-These models usually aim to describe equilibrium conditions and at least  include temperature and relative humidity although some include precipitation and solar radiation effects as well.
-Explicit models of fire behavior as a function of time since wetting could improve fire prediction, however.
-In many litter fuel driven systems, natural lightning ignitions are concurrent with precipitation and, therefore, fire behavior can depend on rate of drying following a wetting event.
-If species differ in the rate of drying, this may strongly influence relatively flammability across species at any particular time post-wetting.
-Such differential moisture content across time, may contribute to the often high spatial heterogeneity of surface fires \cite{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.
+Fire scientists often categorize fuel by size classes that correspond to
+moisture exchange rates, although such simplified categories have long been
+recognized as not sufficient to describe species-specific effects on moisture
+dynamics \cite{Anderson-1985}. Fuel moisture dynamics of leaf litter are
+dependent upon forest structure (shading and rain throughfall effects), litter
+physical structure, and chemical composition of the fuelbeds
+\cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
+models describing moisture dynamics of fine fuels are used by fire scientists
+\cite{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
+usually aim to describe equilibrium conditions and at least include temperature
+and relative humidity although some include precipitation and solar radiation
+effects as well. Explicit models of fire behavior as a function of time since
+wetting could improve fire prediction, however. In many litter fuel driven
+systems, natural lightning ignitions are concurrent with precipitation and,
+therefore, fire behavior can depend on rate of drying following a wetting
+event. If species differ in their drying rate, this may strongly influence
+relatively flammability across species at any particular time following wetting.
+Such differential moisture content across time, may contribute to the often
+high spatial heterogeneity of surface fires \cite{Knapp_Schwilk_etal-2006,
+  Kreye_Hiers_etal-2018}.
 
-Litter packing (measured as litterbed density or packing ratio) controls the flammability of dry litter fuels in seasonally dry climates (higher litter density reduces flammability) and leaf litter packing is influenced by leaf size and shape \cite{Fonda+Belanger+etal-1998, Kane+Varner+etal-2008, Schwilk+Caprio-2011, Kreye+Varner+etal-2013}.
-Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936, Fons-1946, Anderson+Rothermal-1965}.
-However, the mechanism by which both monospecific and mixed litterbeds absorb and desorb moisture is poorly known, and could be driven by by leaf traits, ermegent litter litter traits, or both.
-Species specific effects on litter flammability are recorded for the forests of the southeastern US \cite{Nowacki+Abrams-2008} where species-specific litter drying rates play a role in flammability feedbacks \cite{Kreye+Varner+etal-2013}.
-Similarly, in mixed-conifer forest of the western USA, where the fire regime is characterized by surface fires fueled by plant litter, overstory tree composition drives variation in local fire behavior \cite{Schwilk+Caprio-2011} as a result of differential litter flammability cross tree species and as a result of synergistic interactions in multi-species litter mixtures \cite{Magalhaes+Schwilk-2012}.
-These flammability feedbacks may help explain historical fire frequency patterns, but current estimates of litter flammability in mixed conifer forests are based largely on burning dry litter.
+% Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936,   Fons-1946, Anderson+Rothermal-1965}.
 
-Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss \citep{Kreye-2013}.
+Litter packing (measured as litterbed density or packing ratio) controls the
+flammability of dry litter fuels in seasonally dry climates because higher
+litter density reduces flammability, especially on the flame spread rate axis.
+Leaf litter density is driven by leaf size and shape
+\cite{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
+These litter particle effects on litter bed aeration, in turn, have effects on
+litter bed flammability \cite{Ganteaume+Marielle+etal-2011,
+  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
+are well established, however, how moisture dynamics modulate species specific
+flammability is poorly known, and could be driven by by leaf traits, emergent
+litter litter traits, or both. Species specific effects on litter flammability
+are recorded for the forests of the southeastern US \cite{Nowacki+Abrams-2008}
+where species-specific litter drying rates play a role in flammability
+feedbacks \cite{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
+the western USA, where the fire regime is characterized by surface fires fueled
+by plant litter, overstory tree composition drives variation in local fire
+behavior \cite{Schwilk+Caprio-2011} as a result of differential litter
+flammability cross tree species and as a result of synergistic interactions in
+multi-species litter mixtures \cite{Magalhaes+Schwilk-2012}. These flammability
+feedbacks may help explain historical fire frequency patterns, but current
+estimates of litter flammability in mixed conifer forests are based largely on
+burning dry litter.
+
+Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss \citep{Kreye+Kobziar+etal-2013}.
 Absorptive capacity represents the potential maximum fuel moisture and is evaluated by the moisture content of litter when saturated.
 After saturation, the differences in moisture content indicate different abilities to retain moisture.
-We propose two different sets of traits responsible for variation in the two processes mentioned: leaf particle geometry and litter bulk density.
-Litter characteristics such as surface-area-to-volume ratio, litter bulk density, mass density, particle size and shape, and chemical traits influence moisture diffusion through litter [CITE].
-Litter packing has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits, especially leaf size \cite{Scarff+Westoby-2006}.
-Fuelbeds composed of larger leaves with lower moisture retainability and decomposition rates favor drying, thus increasing the probability of ignition and fire spread rate.
+Litter particle geometry may have direct effects on maximum moisture absorption through area-mass relationships and we also expect effects of leaf morphaology, espcially leaf cuticle thickness \cite{Van_Wagner-1969}.
+Drying rates should also be influenced by leaf and litter particle traits: surface-area-to-volume ratio and litter bulk density will influence air flow through the litter bed.
+Litter packing therefore has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits, especially leaf size \cite{Scarff+Westoby-2006}.
+Fuelbeds composed of larger leaves with lower moisture retainability and should favor drying, thus increasing the probability of ignition and fire spread.
 
-Flammability is a multi-dimensional characteristic of fuels \citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing different components of fire behavior.
-In litter driven fire, the two major axes of variation are described by measurements related to flame spread rate and measurements related to heat transfer.
-Flame spread rate is easy to measure in burning trials, but heat transfer can be more difficult and often proxies such as duration of heating are used \citep{Magalhaes+Schwilk-2012, Varner+Kane+etal-2015}.
-These different components of fire behavior have differing ecological effects.
-Species traits may interact such that the flammability of multi-species mixtures is not the average of the constituent species.
-Such non-additive effects have been shown to be synergistic in dry litter fuels where the flammability of a mixture is driven by the most flammable species  \cite{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}.
-It is possible that similar non-additive effects may influence litter moisture content, especially if different leaf traits influence adsorptive capacity and drying rate.
-There is some evidence for non addiitvity of flammability in moist fuels and one study has found increasing non additivity with moisture \cite{Blauw+Wensink+etal-2015}.
+Flammability is a multi-dimensional characteristic of fuels
+\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
+different components of fire behavior. In litter driven fire, the two major
+axes of variation are described by measurements related to flame spread rate
+and measurements related to heat transfer. Flame spread rate is easy to measure
+in burning trials, but heat transfer can be more difficult and often proxies
+such as duration of heating are used \citep{Magalhaes+Schwilk-2012,
+  Varner+Kane+etal-2015}. These different components of fire behavior have
+differing ecological effects. Species traits may interact such that the
+flammability of multi-species mixtures is not the average of the constituent
+species. Such non-additive effects have been shown to be synergistic in dry
+litter fuels where the flammability of a mixture is driven by the most
+flammable species \cite{VanAltena+Logtestjin+etal-2012,
+  Magalhaes+Schwilk-2012}. It is possible that similar non-additive effects may
+influence litter moisture content, especially if different leaf traits
+influence adsorptive capacity and drying rate. There is some evidence for
+non-additive flammability in moist fuels with one study finding increasing
+non-additivity with moisture \cite{Blauw+Wensink+etal-2015}.
 
-We investigated species specific litter moisture dynamics across eight tree species in a temperate fire-prone forest in California.
-We determined the maximum water retention and rate of moisture loss and examined how these vary in relation to two leaf and litter bed traits.
-We then examined the influence of moisture on the flammability across these species and how flame spread rate and fuel consumption varied with time since wetting and by taxon.
-Experimental burn trials of leaf litter at various moisture levels determined how time since wetting influenced two axes of flammability characterized by flame spread rate and total fuel consumed.
-Previous work described the flammability of these species in monocultures and in mixtures for oven-dry litter \cite{Magalhaes+Schwilk-2012} and found consistent positive non additivity of flammability where fire behavior of mixtures was driven by the most flammable species.
-Here, we examined if this positive non additivity was maintained in mixtures across varying levels of fuel moisture.
-The burn trials were also performed with multi-species litter mixtures to evaluate if species contribute equally to litter moisture dynamics or if there are non-additive effects as seen in oven-dried litter.
-We tested if the non-additive effects on  reported in previous studies occur under ecologically-relevant moisture levels and whether fuel moisture effects strengthen or mask such non-additivity.
+We investigated species specific litter moisture dynamics across eight tree
+species in a temperate fire-prone forest in California. We determined the
+maximum water retention and rate of moisture loss and examined how these vary
+in relation to two leaf and litter bed traits. We then examined the influence
+of moisture on the flammability across these species. Experimental
+burn trials of leaf litter at various moisture levels determined how time since
+wetting influenced two axes of flammability characterized by flame spread rate
+and total fuel consumed. Previous work described the flammability of these
+species in monocultures and in mixtures for oven-dry litter
+\cite{Magalhaes+Schwilk-2012} and found consistent positive non-additivity of
+flammability where fire behavior of mixtures was driven by the most flammable
+species. Here, we examined if this positive non additivity was maintained in
+mixtures across varying levels of fuel moisture. The burn trials were also
+performed with multi-species litter mixtures to evaluate if species contribute
+equally to litter moisture dynamics or if there are non-additive effects as
+observed in oven-dried litter. We tested if the non-additive effects on reported in
+previous studies occur under ecologically-relevant moisture levels and whether
+fuel moisture effects strengthen or mask such non-additivity.
 
@@
 Leaf litter was collected in the summer (mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the study area.
-Effort was made to collect from different populations (minimum of four, each population separated from the others by 9 to 32\,km) for each species to control for plastic and ecotypic effects.
-An exception to this involved the litter from \emph{A.
-magnifica} which is restricted to the higher elevations and could only be collected from one area, due to logistical constraints.
+We collected from a minimum of four populations per species with each population 9--32\,km distant from its neighbors.
+An exception to this was \emph{A.
+magnifica} which is restricted to the higher elevations and could only be collected from one area, due to designated wilderness boundaries and logistical constraints.
 Within that area, \emph{A.
 magnifica} litter was collected from over 10 different individuals.
-For the remaining species, the collection involved 2-4 individual trees at each site, separated by at least 10\,m.
-The litter was collect about 2\,m away from the trunk to obtain a more uniform sample, because bark and heavier branches tend to fall closer to the tree trunk.
-All litter was air dried down to < 5\% fuel moisture for weighing and dividing into moisture trials.
-The moisture manipulations and subsequent burning trials were conducted in April-August 2015.
+For the remaining species, the collection involved 2--4 individual trees at each site, separated from one another by at least 10\,m.
+The litter was collected 2\,m away from the trunk to obtain a more uniform sample of leaf and fine branchlet litter, because bark and heavier branches tend to fall closer to the tree trunk.
+All litter was air dried to < 5\% fuel moisture by dry weight for weighing and dividing into moisture trials.
+The moisture manipulations and subsequent burning trials were conducted in April--August 2015.
 
@@
 the eight species (six replicates per species) and four mixtures (5 replicates
-per mixture type) were placed in baskets (approximately 45 x 45\,cm by 15\,cm
-tall) constructed of 1 cm aluminum screen. Samples were 450 g of litter (dry
-weight) per trial, and litter depth was standardized to 10\,cm by changing the
+per mixture type) were placed in baskets approximately 45 x 45\,cm by 15\,cm
+tall constructed of 1\,cm aluminum screen. Samples were 450\,g dry weight
+litter per trial and litter depth was standardized to 10\,cm by changing the
 horizontal dimensions of the basket. Weight was measured by using a balance
-sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples were
-immersed in plastic storage bins filled with water for 24 hours to fully
+sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples
+were immersed in plastic storage bins filled with water for 24 hours to fully
 saturate them. After saturation, the baskets were removed from the water,
@@
 the observed moisture dynamics was low, so we decided \emph{a priori} to test
-the effect of one leaf trait likely to influence moisture absoorption, specific
+the effect of one leaf trait likely to influence moisture absorption, specific
 leaf area, and one litter trait known to directly effect flammability and
@@
 1899 C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the
-apparatus and allowed to burn until extinction.
-If a sample failed to ignite after 30 seconds, no further attempt was made and if three samples of a species at a particular mositure level failed to ignite, no further attempts were made on that species/moisture combination and the replicate was allowed to continue drying.
-A timer was used to record total time to ignition and duration of flaming combustion.
-The flammability trials followed the methodology detailed in de
-Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.
+apparatus and allowed to burn until extinction. If a sample failed to ignite
+after 30 seconds, no further attempt was made and if three samples of a species
+at a particular mositure level failed to ignite, no further attempts were made
+on that species/moisture combination and the replicate was allowed to continue
+drying. A timer was used to record total time to ignition and duration of
+flaming combustion. The flammability trials followed the methodology detailed
+in de Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.
 
@@
 Burn trials occurred in a cement structure used to simulate house fires at the
-Fire Department of the City of Lubbock. This structure minimized wind and helped
-regulate temperature and relative humidity. Temperature and relative humidity
-were measured before every trial using a Kestrel 3000 (Nielsen-Kellerman,
-Boothwyn, PA). All trials were conducted on clear days between 10 am and 3 pm.
+Fire Department of the City of Lubbock. This structure minimized wind and
+helped regulate temperature and relative humidity. Temperature and relative
+humidity were measured before every trial using a Kestrel 3000
+(Nielsen-Kellerman, Boothwyn, PA). All trials were conducted on clear days
+between 10 am and 3 pm.
 
@@
 Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
-decay coefficient. To estimate these c oefficents we fit a linear mixed-effects
+decay coefficient. To estimate these coefficents we fit a linear mixed-effects
 model by first taking the natural logarithm of moisture content, with time
@@
 retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
-($\lambda$) for each species. All analyses were conducted in R
+($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R
 \cite{RCoreTeam-2019}. We fit models with R packages `lme4'
@@
 another.
-Therefore, to preserve power and focus on analyses we used taxonomic groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our flammability functional groups and treated these as fixed effects in our models.
+Therefore, to preserve power used taxonomic groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our flammability functional groups and treated these as fixed effects in our models.
 The effect of fuel moisture and
@@
 by testing if moisture content and flammability of mixtures were predicted by
-the average of the constituent species. To choose mixtures, we first grouped the
-species according to their maximum water retainability and dessication rate
+the average of the constituent species. To choose mixtures, we first grouped
+the species according to their maximum water retainability and dessication rate
 according to estimated marginal means by species using the R package `emmeans'
-\cite{Lenth-2019}. This resulted in four broad groups of species and we then
-selected one representative species from each litter type group and produced
-all four possible three-species litter mixtures using those four species. For
-each mixture and time since wetting, we predicted moisture content based on the
-mean content of the three consituent species at that time since wetting. We
-predicted flammability parameters based on the models fit to single species
-data described above and then calculated an expected moisture content and
-spread rate as the average of these three values for the mixture. The
-expectation from the null model was that the difference between observed and
-predicted values of flammability will be zero. We tested if there were
-significant departures from the null (zero) with a mixed effects linear model.
-For flammability parameters, we fit linear mixed effect models with the
-observed-predicted flammability value as the dependent variable and used Wald
-tests to test if the centered intercept was different than zero (overall non
-additivity) and if the slope with time since wetting was different from zero
-(non additivity changing with drying).
+\cite{Lenth-2019}. This resulted in four groups of species and we then selected
+one representative species from each litter type group and produced all four
+possible three-species litter mixtures using those four species. For each
+mixture and time since wetting, we predicted moisture content based on the mean
+content of the three consituent species at that time since wetting. We
+predicted flammability parameters based on the average flammability measures
+from individual species trials at each time since wetting and then calculated a
+predicted spread rate and fuel consumption as the average of these three values
+for the mixture. The expectation from the null model was that the difference
+between observed and predicted values of flammability will be zero. We tested
+if there were significant departures from the null (zero) with a mixed effects
+linear model. For flammability parameters, we fit linear mixed effect models
+with the observed-predicted flammability value as the dependent variable and
+used Wald tests to test if the centered intercept was different than zero (a
+measure of overall non-additivity) and if the slope with time since wetting was
+different from zero (indicating non-additivity changing with drying).
 
@@
 There were marked differences in the extent of moisture species retain initially
-following wetting and draining by gravity (all p < 0.0001, Fig.
+following wetting and draining by gravity (all $P$ < 0.0001, Fig.
 \ref{fig:drydown} and Table \ref{tab:drydown}). \emph{Q. kelloggii}, the only
@@
 species and estimated mixed linear model slopes discriminated species into
-three groups from lowest to most rapid: 1) (\emph{C.decurrens}, 2) \emph{Abies
+three groups from lowest to most rapid: 1) \emph{C.decurrens}, 2) \emph{Abies
   spp} and \emph{S. gianteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp}
@@
   (0.5 hr) was added to this plot to aid in visualizing because, although
-  collections were staggered, there were often multiple trays collected at a
+  collections were staggered, there were sometimes multiple trays collected at a
   single time point.}
@@
   \caption{Linear mixed model results for moisture content as a function of time since
-    wetting. Approximate degrees of freedom, pseudo F statistics and p-values
+    wetting. Approximate degrees of freedom (df), pseudo F statistics and p-values
     were calculated by the Kenward-Roger approximation.
@@
 %% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-\input{figs_tables/tab1_drydown_anova.ltx}
+\begin{tabular}{llll}
+  \toprule
+Effect & df & $F$ & $P$ \\ 
+  \midrule
+hour & 1, 183.73 & 1599.59 & $<$.0001 \\ 
+  spcode & 7, 89.73 & 36.15 & $<$.0001 \\ 
+  hour:spcode & 7, 184.15 & 23.06 & $<$.0001 \\ 
+   \bottomrule
+\end{tabular}
 \end{table}
@@
 Across these eight species specific leaf area had a positive effect on maximum
-water retention (Fig. \ref{fig:maxmc-di}, linear model p = 0.003) and litter
+water retention (Fig. \ref{fig:maxmc-di}, linear model $P$ = 0.003) and litter
 bulk density had a negative effect on desiccation rate (Fig. \ref{fig:bd-di},
-linear model p = 0.029). Anova results are shown in Rables \ref{tab:mc_di_anova}.
+linear model $P$ = 0.029). Anova results are shown in Tables \ref{tab:mc_di_anova}.
 
@@
 \includegraphics[width=16cm]{figs_tables/fig2_SLA_maxMC.pdf}
-\caption{Maximum water retention by specific leaf area (SLA) across eight species. ANOVA table in Table \ref{tab:mc_di_anova}}
+\caption{Maximum water retention by specific leaf area across eight species. ANOVA table in Table \ref{tab:mc_di_anova}.}
   \label{fig:maxmc-di}
@@
 \includegraphics[width=16cm]{figs_tables/fig3_di_bd.pdf}
-\caption{Linear model predicting dessication rate as a function of specific leaf area and litter bulk density across eight species. ANOVA table in Table \ref{tab:mc_di_anova}}.
+\caption{Drying rate rate as a function of specific leaf area and litter bulk density across eight species. ANOVA table in Table \ref{tab:mc_di_anova}}.
   \label{fig:bd-di}
@@
 \begin{table}[H]
-  \caption{Linear model results for maximum water retention (\% moisture content on dry mass basis) and for estimated desiccation rate (hr$^{-1}$) as functions of specific leaf area and litter bulk density across eight species.
-ANOVA tables shown with degrees of freedom, sum of squares, mean squares, F, and p for each effect.}
+  \caption{Linear model results for maximum water retention (\% moisture content on dry mass basis) and for estimated drying rate (hr$^{-1}$) as functions of specific leaf area and litter bulk density across eight species.
+ANOVA tables shown with degrees of freedom, sum of squares, mean squares, $F$, and $P$ for each effect.}
   \label{tab:mc_di_anova}
@@
   \toprule
-Ind. variable & Effect & df & Sum of Squares & Mean squares & F & p value \\ 
+Ind. variable & Effect & df & Sum of Squares & Mean squares & $F$ & $P$ \\ 
   \midrule
@@
 
-% \begin{table}[H]
-%   \caption{Linear model results for estimated dessication rate ($hr^{-1}$) as a function of specific leaf area and litter bulk density.
-Anova table shown with degrees of freedom, sum of squares, mean squares, F, and p for each effect}
-%   \label{tab:di_anova}
-% \centering
-% %% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-% \input{figs_tables/tab3_di_anova.ltx}
-% \end{table}
-
 \subsection{Moisture effects on flammability}
@@
 \ref{fig:spread_moist}).
-\emph{Quercus} had higher spread rates across all fuel moistures and had a steeper decline in response to moisture than did the other groups (taxon by time since wetting interaction p = 0.022, Table \ref{tab:spreadrate_anova}).
+\emph{Quercus} had higher spread rates across all fuel moistures and had a steeper decline in response to moisture than did the other groups (taxon by time since wetting interaction $P$ = 0.022, Table \ref{tab:spreadrate_anova}).
 Fuel moisture negatively influenced fuel consumption.
@@
 \includegraphics[width=8cm]{figs_tables/fig4_spread_actualMC.pdf}
-\caption{Flame spread rate by moisture content (model fit F= 61.15, d.f = 9, 148,  p < 0.0001). Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI Table 2.}
+\caption{Flame spread rate by moisture content (model fit $F$= 61.15, df = 9, 148,  $P$ < 0.0001). Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI Table 2.}
   \label{fig:spread_moist}
@@
 
+%% Table 3 (formerly 3 and 5)
 \begin{table}[H]
-  \caption{Generalized linear mixed model results for flame spread rate as a function of fuel moisture with taxonomic group as a fixed effect and species as a nested random effect.
-Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.
-Coefficient estimates are in SI Table 2.}
+  \caption{Generalized linear mixed model results for two models predicting flame spread rate: the first model examined spread rate as a function of fuel moisture (mc) with taxonomic group as a fixed effect and species as a nested random effect.
+The second model predicted spread rate as a function of time since wetting (time) with taxonomic group as a fixed effect and species as a nested random effect.
+Approximate degrees of freedom (df), pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.
+Estimated coefficients are in SI Tables 2 and 4}
   \label{tab:spreadrate_anova}
 \centering
-\input{figs_tables/tab4_spread_moist_anova.ltx}
+\begin{tabular}{ccrrrrr}
+  \toprule
+Model set & Effect & df & $F$ & & $P$ \\
+  \midrule
+  Effect of moisture & mc & 1, 146.20 & 21.04 & $<$.0001 \\ 
+ &  taxon & 3, 51.58 & 65.08 & $<$.0001 \\ 
+ & vpd & 1, 148.79 & 9.62 & .002 \\ 
+ & mc:taxon & 3, 147.21 & 4.83 & .003 \\
+
+  \midrule
+
+  Effect of time & hour & 1, 214.50 & 66.22 & $<$.0001 \\ 
+ &  taxon & 3, 7.59 & 61.81 & $<$.0001 \\ 
+ & vpd & 1, 215.00 & 13.49 & .0003 \\ 
+ & hour:taxon & 3, 183.97 & 3.99 & .009 \\
+
+   \bottomrule
+
+\end{tabular}
 \end{table}
@@
 \includegraphics[width=8cm]{figs_tables/fig5_consume_actualMC.pdf}
-\caption{Percent fuel consumed by moisture content. Lines indicate best fit beta regression lines. Model deviance table in Table \ref{tab:consume_moist_anova} and coefficients in SI Table 3.}
+\caption{Percent fuel consumed by moisture content. Lines indicate best fit beta regression lines. Model deviance table in Table \ref{tab:consume_anova} and coefficients are in SI Table 3.}
   \label{fig:consume_moist}
@@
 \begin{table}[H]
-  \caption{Generalized linear mixed model results for fuel consumption as a function of fuel moisture with taxonomic group as a fixed effect and species as a nested random effect.
-The analysis of deviance table below shows deviance comparisons for the simplest null model with only intercept and random effect (model 1), for a taxon effect only (model 2), for a taxon and mosture content effect (model 3) and for the full model including the taxon by moisture interaction (model 4).
-Coefficient estimates are in SI Table 3.}
-  \label{tab:consume_moist_anova}
+  \caption{Generalized linear mixed model results for two sets of models predicting fuel consumption.
+The first set modeled fuel consumption as a function of fuel moisture (mc) with taxonomic group as a fixed effect and species as a nested random effect.
+The second set modeled fuel consumption as a function of as a function of time since wetting (time) with taxonomic group as a fixed effet and species as a nested random effect.
+The analysis of deviance table below shows sequential chi-square tests on nested beta regression models: from the simplest null model within each group with only intercept and random effect, for a taxon effect only model, for a taxon and moisture content effect, and for the full model including the taxon by moisture or time interaction.
+Coefficient estimates are in SI Tables 3 and 5.}
+  \label{tab:consume_anova}
   \centering
-\input{figs_tables/tab5_consume_moist_anova.ltx}
+
+  \begin{tabular}{ccrrrr}
+\toprule
+Model set & model & Log Lik. & df & Deviance & $P$ \\ 
+\midrule
+  Effect of moisture &  consum \~{} 1 & 262.12 &  &  &  \\ 
+   & consum \~{} taxon & 268.73 & 3 & 13.23 & 0.0042 \\ 
+   & consum \~{} taxon + mc & 289.57 & 1 & 41.68 & < 0.0001 \\ 
+   & consum \~{} taxon + mc + taxon:mc & 293.87 & 3 & 8.60 & 0.0351 \\ 
+\midrule
+Effect of time &  consum \~{} 1 & 140.08 &  &  &  \\ 
+ &  consum \~{} taxon & 155.84 & 4.00 & 31.51 & < 0.001 \\ 
+ & consum \~{} taxon + time & 209.84 & 1.00 & 108.01 & <0.001 \\ 
+ & consum \~{} hour + taxon + taxon:time & 216.96 & 4.00 & 14.23 & 0.01 \\ 
+    \bottomrule
+    \end{tabular}
 \end{table}
 
-As a consequence of the relationships shown above, time since wetting had a strong effect on spread rate (p < 0.0001) and the slope of this effect varied by taxonomic group (Fig.
+As a consequence of the relationships shown above, time since wetting had a significant effect on spread rate ($P$ < 0.0001) and the slope of this effect varied by taxonomic group (Fig.
 \ref{fig:spread_time}, Table \ref{tab:spreadrate_anova}).
-The pines were able to ignite and burn much sooner after wetting than the other genera, none of which had measurable spread rates until burning was attempted 72 hours after wetting (note that trials that failed to even ignite smoldering combustion are ommitted completely from the model and figure).
+The pines were able to ignite and burn much sooner after wetting than did the other taxa, none of which had measurable spread rates until burning was attempted 72 hours after wetting (note that trials that failed to even ignite smoldering combustion are ommitted completely from the model and figure).
 At that point, the \emph{Abies}, \emph{Calocedrus}, and \emph{Sequiadendron} had dryer litter than did the \emph{Quercus}, yet the spread rates were slow.
@@
   Lines indicate best fit linear model fits. Model ANOVA table in Table
-  \ref{tab:spread_time_anova} and coefficients in SI Table 4. Horizontal
+  \ref{tab:spreadrate_anova} and coefficients in SI Table 4. Horizontal
   jitter (up to 2 hr) was added to the plot for visualization and to avoid
-  overplotting. All collections were at the times labeled on the x-axis.. A
+  overplotting. All collections were at the times labeled on the x-axis. A
   lack of data at earlier time points for a taxon indicates no ignitions
@@
 
-\begin{table}[H]
-  \caption[Mixed linear model of flame spread rate.]{Mixed model results for flame spread rate  as a function of time since wetting with genus as a fixed effect and species as a nested random effect.
-Approximate degrees of freedom, pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.
-Estimated coefficients are in SI Table 4}
-  \label{tab:spread_time_anova}
-\centering
-%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-\input{figs_tables/tab6_spread_time_anova.ltx}
-\end{table}
-
-Fuel consumption was highly variable across trials and intermediate values were less common than very low are moderately high consumption (Fig.
+Fuel consumption was highly variable across trials and intermediate values were less common than were very low or moderately high consumption (Fig.
 \ref{fig:consume_time}).
-All parameters significantly contributed to the generalized linear mixed model as indicated by the nested model comparisons in Table \ref{tab:consume_time_anova}.
+All parameters significantly contributed to the generalized linear mixed beta regression model as indicated by the nested model comparisons in Table \ref{tab:consume_time_anova}.
 As with flame spread rate some pines had the most rapid response to drying.
@@
 lambertiana} in its own group allowed model convergence that was otherwise impossible.
-This breaks our intent to use only \emph{a priori} taxonomic groups, but we could not obtain model convergence otherwise.
+This was at odds with our intent to use only \emph{a priori} taxonomic groups, but we could not obtain model convergence otherwise.
+This result should therefore be seen as preliminary because the test does not reflect an \emph{a priori} hypothesis.
 
@@
   species. Curves shown indicate best fit beta regression lines. Model ANOVA
-  table in Table \ref{tab:consume_time_anova} and coefficients in SI Table
+  table in Table \ref{tab:consume_anova} and coefficients in SI Table
   4. Horizontal jitter (up to 2 hr) was added to the plot for visualization and
@@
 
-\begin{table}[H]
-  \caption{Generalized linear mixed model results for fuel consumption as a function of time since wetting with taxonomic group as a fixed effet and species as a nested random effect.
-The table shows sequential chi-square tests on nested beta regression models.
-The analysis of deviance table below shows deviance comparisons for the simplest null model with only intercept and random effect (model 1), for a taxon effect only (model 2), for a taxon and mosture content effect (model 3) and for the full model including the taxon by moisture interaction (model 4).
-Coefficient estimates for the full model are in SI Table 5.}
-  \label{tab:consume_time_anova}
-\centering
-\input{figs_tables/tab7_consume_time_anova.ltx}
-\end{table}
+\subsection{Non-additivity in  litter mixtures}
 
-% \begin{table}[H]
-%   \caption{Generalized linear mixed model results for fuel consumption as a function of time since wetting with taxonomic group as a fixed effect and species as a nested random effect.
-These are beta regression models.
-The analysis of deviance table below shows deviance comparisons for the simplest null model with only intercept and random effect (model 1), for a taxon effect only (model 2), for a taxon and mosture content effect (model 3) and for the full model including the taxon by moisture interaction (model 4).
-Coefficient estimates are in SI Table 5.}
-%   \label{tab:consume_time_anova}
-% \centering
-% \input{figs_tables/tab6_consume_time_anova.ltx}
-% \end{table}
-
-\subsection{Non additivity in  litter mixtures}
-
-Based on visual inspection of the dry down curves and the estimated marginal means for maximum water retainability and desiccation rate (SI Figs 2-3), we selected four species to represent the full range of litter moisture dynamics for use in the litter mixtures and examined all four three-species combinations of those four.
+Based on visual inspection of the dry down curves and the estimated marginal means for maximum water retainability and desiccation rate (SI Figs 2--3), we selected four species to represent the full range of litter moisture dynamics for use in the litter mixtures and examined all four three-species combinations of those four.
 The four species were \emph{Abies concolor}, \emph{Calocedrus decurrens}, \emph{Pinus jeffreyi} and \emph{Quercus kellogii}.
@@
   \caption{Linear mixed model results for moisture content as a function of time since
-    wetting for four different litter mixture types. Approximate degrees of freedom, pseudo F statistics and p-values
+    wetting for four different litter mixture types. Approximate degrees of freedom (df), pseudo F statistics and p-values
     were calculated by the Kenward-Roger approximation
@@
 \centering
-%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-\input{figs_tables/tab8_mixtures_drydown_anova.ltx}
+
+\begin{tabular}{llll}
+  \toprule
+Effect & df & $F$ & $P$ \\ 
+  \midrule
+time & 1, 96.19 & 565.28 & $<$.0001 \\ 
+  mixture & 3, 96.08 & 23.63 & $<$.0001 \\ 
+  time:mixture & 3, 95.71 & 8.70 & $<$.0001 \\ 
+   \bottomrule
+\end{tabular}
 \end{table}
@@
 \caption{Observed vs predicted fuel moisture content for litter mixtures.
-Predictions are based on the mean moisture content at each particualr time sinece wetting for the three constituent species in each mixture.
-Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
+  Predictions are based on the mean moisture content at each time since wetting
+  for the three constituent species in each mixture. Codes for species in
+  mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi =
+  \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
   \label{fig:mixture_obs_pred_mc}
@@
 
-There was evidence for non additive effects in drying rate.
-Mixtures tended to have lower water content than predicted from the mean of the constituent species after wetting, but higher water content than expected after long drying (Wald test shows slope of mixed linear model illustrated in Fig.
-\ref{fig:mixture_obs_pred_mc} less than 1, p < 0.0001).
+There was evidence for non-additive effects in drying rate.
+Mixtures tended to have lower water content than predicted from the mean of the constituent species after wetting, but had higher water content than expected after long drying (Wald test shows slope of mixed linear model illustrated in Fig.
+\ref{fig:mixture_obs_pred_mc} less than 1, $P$ < 0.0001).
 
@@
 \includegraphics[width=16cm]{figs_tables/fig10_mixture_obs_vs_pred_spread.pdf}
-\caption{Observed vs predicted flame spread rate for litter mixtures.
+\caption{Observed vs.
+predicted flame spread rate for litter mixtures.
 Filled circles show observed spread rate values.
-Open diamonds indicate predicted spread rates based on the average spread rate at each particular time since wetting for the three constituent species in each mixture.
+Open diamonds indicate predicted spread rates based on the average spread rate at each time since wetting for the three constituent species in each mixture.
 Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
@@
 \end{figure}
-
-% \begin{figure}[H]
-%   \centering
-% \includegraphics[width=16cm]{figs_tables/fig11_mixture_obs_vs_pred_consume.pdf}
-% \caption{Observed vs predicted fuel consumption for litter mixtures.
-Predictions are based on the modeled spread rate at each particular time since wetting for the three constituent species in each mixture.
-Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
-%   \label{fig:mixture_obs_pred_consume}
-% \end{figure}
 
@@
 
-Flammability measures also demonstrated nonadditivity in mixtures.
-Flame spread rate was consistently slightly higher than predicted (Fig \ref{fig:mixture_obs_pred_spread}, Wald test of mixed linear model centered intercept for observed-predicted values greater than zero, p = 0.017) and the positive non additivity increased with time since wetting (slope greater than zero, p = 0.005).
-Fuel consumption, however, showed no overall bias towards positive or negative non additivity, but non addivity (difference between observed and expected values) did show a positive relationship with time since wetting (Fig.
-\ref{fig:mixture_obs_pred_consume}, p < 0.0001).
+Flammability measures also demonstrated non-additivity in mixtures.
+Flame spread rate was consistently higher than predicted (Fig \ref{fig:mixture_obs_pred_spread}, Wald test of mixed linear model centered intercept for observed-predicted values greater than zero, $P$ = 0.017) and the positive non-additivity increased with time since wetting (slope greater than zero, $P$ = 0.005).
+Fuel consumption, however, showed no overall bias towards positive or negative non-additivity, but non-additivity increased with time since wetting (Fig.
+\ref{fig:mixture_obs_pred_consume}, slope not equal to zero, $P$ < 0.0001).
 
@@
 
-The species in this study displayed different patterns of both moisture absorption and desorption, which were controlled by different litter traits (specific leaf area and litter bulk density, respectively).
-\emph{Q.
-kelloggii} was the species that absorbed the most moisture, followed distantly by the finest-needled pine, \emph{P.
-lambertiana}.
-These species that have a greater absorptive capacity are also among the most flammable in terms of spread rate when dry.
-One presumption of ours prior to this experiment was that the high flammability previously reported for \emph{Q.
+The species in this study displayed different patterns of both moisture absorption and desorption, apparently controlled by different litter traits.
+In this sample of species, those that have a greater absorptive capacity are also among the most flammable in terms of spread rate when dry.
+This highlights the importance of fuel moisture in modulating flammability.
+One shortcoming of past work in this system \cite{Magalhaes+Schwilk-2012} was that flammability trials tested only very dry fuels.
+We therefore expected that rankings of species flammability might change under more realistic moisture levels.
+For example, one presumption of ours prior to this experiment was that the high flammability previously reported for \emph{Q.
 kelloggii} \cite{Magalhaes+Schwilk-2012} would disappear under higher fuel moisture contents.
@@
 
-    The dry down curves we measured demonstrate that after a rain event, the litter of different species will exhibit different moisture contents over time.
-Our flammability trials demonstrate that that both flame spread rate and fuel consumption increased with time since wetting but that the rate of increase varied across these genera.
-This indicates that there is a taxon-specific response to flammability based on time since wetting.
+    The dry down curves we measured demonstrate that after a rain event, the litter of different species will exhibit different moisture contents over time as a consequence of both differential initial absorption and of differences in drying rate.
+This has consequences for flammability as a function of species and time since wetting.
+Our flammability trials demonstrate that that both flame spread rate and fuel consumption increased with time since wetting but that the rate of increase varied across these taxa.
 Only the pines are able to ignite and support flame spread immediately after wetting (Fig.
 \ref{fig:spread_time}).
-\emph{Quercus} exhibited a threshold response to drying with no ignition prior to 72 hours of drying but once ignition is possible, having faster spread rates than the other genera.
+In contrast, \emph{Quercus} exhibited a threshold response to drying with no ignition prior to 72 hours of drying but once ignition is possible, having faster spread rates than the other taxa.
+The Cupressaceae, which have dense litter but low maximum water absorption had low flammability across moisture levels.
 
 With eight species, power to detect species-specific trait effects on these moisture patterns was weak, but we examined one litter trait known to directly influence flammability: litter bulk density and one leaf trait, specific leaf area.
-Litter density is largely driven by leaf size in this system  and in turns controls the major axis of flammability variation in dry litter \cite{Magalhaes+Schwilk-2012}.
-Our results suggest that moisture dynamics are controlled by both traits intrinsic to the leaf particles (specific leaf area) and emergent traits of the litter bed (bulk density).
-Specific leaf area here is probably capturing some joint effect of cuticle thickness and total surface area of litter particles.
-The oak has much higher initial water retention than the other species, followed by the finest-needled pine, \emph{P.
+Litter density is largely driven by leaf size in this system and, in turn, controls the major axis of flammability variation in dry litter \cite{Magalhaes+Schwilk-2012}.
+Our results suggest that moisture dynamics are controlled by both traits intrinsic to the leaf particles (specific leaf area) and an emergent property of the litter bed (bulk density).
+Specific leaf area here is probably capturing some joint effect of cuticle thickness and total surface area of litter particles, but we cannot test the morphological and physical mechanisms of water absorption with these data.
+The oak has much higher initial water retention than do the other species, followed by the finest-needled pine, \emph{P.
 lambertiana}.
-With only a handful of species, however, disentangling mechanism of water absorption is not possible.
-Our study examines fresh litter, but we expect that decomposition will affect moisture content.
+Our study examines fresh litter, but we expect that decomposition will affect moisture dynamics.
 For example, Van Wagner \cite{Van_Wagner-1969}, investigating fuel weathering, found that pine needles had a variation in moisture content 3 to 6 times greater when they lacked their protective wax coating when compared to intact needles.
-
-% Therefore, mere moisture content will not inform on the fire behavior measures of a litter bed.
-Species composition of the litter must be considered, and time since the last rain event should allow the determination of the potential flammability of an area.
-
-% The pines are more flammable at higher moisture contents when compared to volatile-rich species such as \emph{C.
-decurrens}, \emph{Abies spp}.
-or \emph{S.
-giganteum}.
-This effect is influenced by litter bulk density and is likely to persist in fuel mixtures where the constituent species have different desiccation indices and therefore will have different moisture contents.
-
-% I have determined moisture of extinction for the eight species in this study.
-As expected, species that have higher flammability values (Pinus spp.) had higher moisture of extinction than species with lower flammability (Abies spp.
-and S.
-giganteum).
-I used a higher cut-off moisture for the burn trials than the moisture of extinction found in the literature (65\% vs.
-27-30\%; \citep{ Rothermel-1972} [TODO Dickinson et al.
-2016] in order to be conservative, as moisture content studies as sparse.
-However, Q.
-kelloggii was still flammable at ~80-90\% moisture content (values at 72 hours used for the species mixtures), which I did not anticipate.
-Due to its high moisture content, Q.
-kelloggii was not burnt at 24 and 48 hours since wetting, which explains the gap shown in the Figures.
-Though I cannot comment on the exact behavior of Q.
-kelloggii at 24 and 48 hours, it is unlikely to be more flammable than the pines.
-This result indicates that though Q.
-kelloggii might be too wet to start a fire right after a rain event, Q.
-kelloggii litter dries faster than the other species in this study and will become very flammable within two days.
 
@@
 
-We tested the dry-down process and flammability of four unique three-species litter mixtures.
+We measured drying rates and flammability of four unique three-species litter mixtures.
 These mixtures contained species representative of all dry-down behavior types observed in single species trials.
 As expected from the single species results, mixtures that contained \emph{Q.
-kelloggii} absorbed more water during wetting, and the mixture without any Cupressaceae showed the most rapid dessication (Figure \ref{fig:mixture_dry_down} and Table \ref{tab:mixtures_drydown}).
-Drydown rates did not behave as the average of the constituent species and mixtures tended to be drier than expected early in the process and wetter than expected later (Fig.
+kelloggii} absorbed more water during wetting, and the one mixture without any Cupressaceae had the most rapid drying rate (Figure \ref{fig:mixture_dry_down} and Table \ref{tab:mixtures_drydown}).
+Drydown rates, however, did not behave as the average of the constituent species: mixtures tended to be drier than expected early in the process and wetter than expected later (Fig.
 \ref{fig:mixture_obs_pred_mc}).
-The fact that mixtures tended to have lower water content than predicted from the mean of the constituent species after wetting, but higher water content
-than expected after long drying is surprising because past work has shown that
-mixtures tend to pack less densely than do litter monocultures \cite[][, Table 2]{Magalhaes+Schwilk-2012}. One would expect that the lower bulk density of mixtures would speed drying.
+The fact that mixtures tended to have lower water content than predicted from the mean of the constituent species after wetting, but higher water content than expected after long drying is surprising because past work has shown that
+mixtures tend to pack less densely than do litter monocultures \cite[][, Table 2]{Magalhaes+Schwilk-2012}.
+If packing density controls rate of drying, then we would expect mixtures to dry more rapidly than predicted from monoculture trials.
+This pattern in mositure contrent through time, however, did not directly predict flammability through time.
 
-% This result indicates that immediately after a rain event, the moisture content of a mixture will be lower than expected, however these high values for moisture content are unlikely to be found in natural litters.
-As the mixture dries, the dry-down behavior will approximate the null model, and three to four days after the rain event the dry-down behavior will become non-additive again.
-
-The overall ranking of flammability in litter mixtrures was consistent with the
-results obtained for the individual species. Mixtures containing \emph{Q.
-  kelloggii} and \emph{P. jeffreyi} had fast spread rates. These results are
-likely influenced by the lower bulk densities of litter mixtures containing
-those two species. Most interestingly, there were striking pattern of nan
-additivity in spread rate and in fuel consumption for the four mixtures in this
-study, indicating that the positive non additivity is in flammability is not
-restricted to completely dry litter mixtures \citep{Magalhaes+Schwilk-2012,
-  VanAltena+Logtestjin+etal-2012}. Non additivity in both spread rate fuel
+The overall ranking of flammability in litter mixtures showed some consistency
+with the results obtained for the individual species, but there were
+significant non-additive effects. Mixtures containing \emph{Q. kelloggii} and
+\emph{P. jeffreyi} tended to have fast spread rates. These results are likely
+influenced by the lower bulk densities of litter mixtures containing those two
+species. Most interestingly, there were striking pattern of non-additivity in
+spread rate and in fuel consumption for the four mixtures, indicating that
+positive non-additivity in flammability is not restricted to completely dry
+litter mixtures observed in previous work \citep{Magalhaes+Schwilk-2012,
+  VanAltena+Logtestjin+etal-2012}. Non-additivity in both spread rate and fuel
 consumption became more positive with time since wetting and these patterns
@@
 than that reported by Blauw et al \citeyear{Blauw+Wensink+etal-2015}. Blauw et
-al found that non-addivity was greatest in moist mixtures of heathland fuels.
-In their work this non addiitvity was generally negative but that positive non
-addivity in mixtures was often driven by mixtures meeting ignition thresholds
-not met by the constituent species alone. To limit the number of models fit, we
-limited our investigation to two flammability parameters that represent two
-major axes of flammability variation. Therefore, our study does not fully
-capture ignition effects although we do have some results consistent with
-theirs: mixtures with \emph{Q. kellogii} could ignite at times since wetting at
-which the species in monoculture could not. Overall, our study does show a consistent pattern of increasingly positive non addiitvity in flammability with litter drying. This result, together with
-previously published work, suggests that litter flammability is prone to
-non-additive effects, reinforcing the idea that we cannot approach flammability
-by lumping species into fuel categories, as is common in fire behavior models.
-Community flammability is dependent on and particular to its constituent
-species and cannot be addressed by simply averaging the flammability of all
-species it comprises. This dynamic view is likely to be increasingly relevant
-considering ongoing climate change, which predicts changes in community
-assembly.
+al found that non-addivity was greater in moist mixtures of heathland fuels
+than in dry fuels. In their work, non-addiitvity was generally negative but cases of 
+positive non-addivity in mixtures was often a result of mixtures meeting
+ignition thresholds not met by the constituent species alone. To limit the
+number of models fit, we limited our investigation to two flammability
+parameters that represent two major axes of flammability variation. Therefore,
+our study does not fully capture ignition effects although we do have some
+results consistent with theirs: mixtures with \emph{Q. kellogii} could ignite
+at times since wetting at which the species in monoculture could not.
 
-% I have found non-additive effects (both positive and negative) in moisture content and in the flammability of wet litter mixtures of mixed-conifer forest.
-These results complement the efforts of Blauw et al.
-(2015) for moist litter and add to the literature investigating non-additivity in litter flammability (de Magalhães and Schwilk 2012, van Altena et al.
-2012).
-However, these studies are limited in the scope of the species studied and suggest that non-additivity needs to be investigated at a larger scale, within different systems and comparatively across a larger number of species.
-Moisture content is considered the single most important driver of litter flammability and the results from my study indicate how leaf and litter traits affect moisture dynamics, and how moisture effects on litter flammability are species-specific.
-A trait-based perspective provides a better mechanistic understanding of moisture dynamics of litter fuel flammability and fire behavior and is especially important for predicting and understanding fire in plant communities that are changing.
+Overall, our study demonstrates a consistent pattern of increasingly positive
+non-additivity in flammability with litter drying across two largely orthogonal
+measures of flammability. This result, together with previously published work,
+suggests that litter flammability is prone to non-additive effects, suggesting
+caution in lumping species into fuel categories. Community flammability is
+dependent on and particular to its constituent species and cannot be addressed
+by simply averaging the flammability of all species included. In litter-driven
+fire, positive non-additivity appears to be the rule, and the relationships we
+described here can form a working hypothesis and baseline expectation to adjust
+fire behavior predictions in real communities: in mixtures with fuel moisture
+content below 50\%, observed flame spread rates approached double the predicted
+value based on monoculture trials (Fig. \ref{fig:mixture_obs_pred_spread}). In
+completely dry litter from this same system, past work demonstrates observed
+spread rates approached three times the expectation
+\cite{Magalhaes+Schwilk-2012}.
 
-% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section{Conclusions}
+Positive non-additivity, whereby more flammable species have effects on fire
+behavior disproportionate to their contribution to the litter layer, could lead
+to vegetation feedbacks. In this temperate mixed conifer forest, these effects
+are likely to include positive feedbacks whereby the shade intolerant species
+such as the long-needled pines and the large-leaved oak, are favored by the
+increased fire to which their fuels contribute \cite{Schwilk+Caprio-2011} and
+similar results are reported for southeastern US forests where species
+composition shifts have led to decreases in fire activity
+\citep{Nowacki+Abrams-2008}. Such positive feedback effects increases the
+potential for species to exert community scale effects even when relatively
+rare and that can influence community assembly and the evolutionary of
+niche-constructing traits \cite{Kerr+Schwilk+etal-1999, Schwilk+Kerr-2002}.
+Ongoing climate change and concurrent changes to fire regimes are linked
+phenomena. The patterns we demonstrate here show to potential importance of species
+traits and interactions at small scales and among subtle shifts in vegetation,
+but vegetation-fire feedbacks are important at a wide range of scales from stand-level species composition shifts to the earth system \cite{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}.
 
-% This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \section{Patents}
-% This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \vspace{6pt}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% optional
-%\supplementary{The following are available online at \linksupplementary{s1}, Figure S1: title, Table S1: title, Video S1: title.}
-
-% Only for the journal Methods and Protocols:
-% If you wish to submit a video article, please do so with any other supplementary material.
-% \supplementary{The following are available at \linksupplementary{s1}, Figure S1: title, Table S1: title, Video S1: title. A supporting video article is available at doi: link.}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\acknowledgments{TODO.}
+\acknowledgments{The authors thank the National Park Service and the employees of Sequoia and Kings Canyon National Parks.This work was carried out with permission from NPS under permit \# SEKI-2014-SCI-0040.
+We also thank students who assisted with field work and the litter collection, including Lizz Waring, Taylor Brown and Chelsea Griffin.}
 
@@
 
-\section{}
-All appendix sections must be cited in the main text. In the appendixes, Figures, Tables, etc. should be labeled starting with `A', e.g., Figure A1, Figure A2, etc.
-
-%=====================================
 % References, variant B: external bibliography

### 2019-07-25 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Normalize and spell check

@@
   \emph{Pinus} species) have faster rates of litter drying. We examined two
-  axes of flammability variation (spread rate and fuel consumption) and found that these
-  differing moisture dynamics result in time since wetting having strongly
-  differing effects on flammability across the major litter types typified by
-  different tree genera. Finally, we found that litter mixtures exhibited
-  consistent non additivity in flame spread rate in fuel consumption. This non
-  additivity became more positive with fuel drying. Litter mixtures burn with
-  fire behavior more similar to that of the most flammable constituent species
-  and this effect increases as fuels dry. ADD Synthesis statement.}
+  axes of flammability variation (spread rate and fuel consumption) and found
+  that these differing moisture dynamics result in time since wetting having
+  strongly differing effects on flammability across the major litter types
+  typified by different tree genera. Finally, we found that litter mixtures
+  exhibited consistent non-additivity in flame spread rate in fuel consumption.
+  This non-additivity became more positive with fuel drying. Litter mixtures
+  burn with fire behavior more similar to that of the most flammable
+  constituent species and this effect increases as fuels dry. ADD Synthesis
+  statement.}
 
@@
 dependent upon forest structure (shading and rain throughfall effects), litter
-physical structure, and chemical composition of the fuelbeds
+physical structure, and chemical composition of the fuel-beds
 \cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
@@
 event. If species differ in their drying rate, this may strongly influence
-relatively flammability across species at any particular time following wetting.
-Such differential moisture content across time, may contribute to the often
-high spatial heterogeneity of surface fires \cite{Knapp_Schwilk_etal-2006,
-  Kreye_Hiers_etal-2018}.
+relatively flammability across species at any particular time following
+wetting. Such differential moisture content across time, may contribute to the
+often high spatial heterogeneity of surface fires
+\cite{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.
 
@@
 
-Litter packing (measured as litterbed density or packing ratio) controls the
+Litter packing (measured as litter-bed density or packing ratio) controls the
 flammability of dry litter fuels in seasonally dry climates because higher
@@
 
-Fuel moisture content through time is controlled by two main processes, absorptive capacity and rate of moisture loss \citep{Kreye+Kobziar+etal-2013}.
-Absorptive capacity represents the potential maximum fuel moisture and is evaluated by the moisture content of litter when saturated.
-After saturation, the differences in moisture content indicate different abilities to retain moisture.
-Litter particle geometry may have direct effects on maximum moisture absorption through area-mass relationships and we also expect effects of leaf morphaology, espcially leaf cuticle thickness \cite{Van_Wagner-1969}.
-Drying rates should also be influenced by leaf and litter particle traits: surface-area-to-volume ratio and litter bulk density will influence air flow through the litter bed.
-Litter packing therefore has two pathways of effect on fire behavior: directly by its effect on oxygen availability for combustion, and indirectly by its effect on dry-down rate, and both these pathways are affected by leaf litter traits, especially leaf size \cite{Scarff+Westoby-2006}.
-Fuelbeds composed of larger leaves with lower moisture retainability and should favor drying, thus increasing the probability of ignition and fire spread.
+Fuel moisture content through time is controlled by two main processes,
+absorptive capacity and rate of moisture loss \citep{Kreye+Kobziar+etal-2013}.
+Absorptive capacity represents the potential maximum fuel moisture and is
+evaluated by the moisture content of litter when saturated. After saturation,
+the differences in moisture content indicate different abilities to retain
+moisture. Litter particle geometry may have direct effects on maximum moisture
+absorption through area-mass relationships and we also expect effects of leaf
+morphology, especially leaf cuticle thickness \cite{Van_Wagner-1969}. Drying
+rates should also be influenced by leaf and litter particle traits:
+surface-area-to-volume ratio and litter bulk density will influence air flow
+through the litter bed. Litter packing therefore has two pathways of effect on
+fire behavior: directly by its effect on oxygen availability for combustion,
+and indirectly by its effect on dry-down rate, and both these pathways are
+affected by leaf litter traits, especially leaf size
+\cite{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves with lower
+moisture retainability and should favor drying, thus increasing the probability
+of ignition and fire spread.
 
@@
 influence litter moisture content, especially if different leaf traits
-influence adsorptive capacity and drying rate. There is some evidence for
+influence absorptive capacity and drying rate. There is some evidence for
 non-additive flammability in moist fuels with one study finding increasing
@@
 in relation to two leaf and litter bed traits. We then examined the influence
-of moisture on the flammability across these species. Experimental
-burn trials of leaf litter at various moisture levels determined how time since
-wetting influenced two axes of flammability characterized by flame spread rate
-and total fuel consumed. Previous work described the flammability of these
-species in monocultures and in mixtures for oven-dry litter
+of moisture on the flammability across these species. Experimental burn trials
+of leaf litter at various moisture levels determined how time since wetting
+influenced two axes of flammability characterized by flame spread rate and
+total fuel consumed. Previous work described the flammability of these species
+in monocultures and in mixtures for oven-dry litter
 \cite{Magalhaes+Schwilk-2012} and found consistent positive non-additivity of
@@
 equally to litter moisture dynamics or if there are non-additive effects as
-observed in oven-dried litter. We tested if the non-additive effects on reported in
-previous studies occur under ecologically-relevant moisture levels and whether
-fuel moisture effects strengthen or mask such non-additivity.
+observed in oven-dried litter. We tested if the non-additive effects on
+reported in previous studies occur under ecologically-relevant moisture levels
+and whether fuel moisture effects strengthen or mask such non-additivity.
 
@@
 
-Field sites for this study were located in Sequoia and Kings Canyon National Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in mixed-conifer forest.
-We chose eight dominant tree species that are representative of this type of forest: \emph{Pinus jeffreyi} Grev.
-\& Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl.
-ex Laws., \emph{Abies concolor} (Gord.
-\& Glend.) Lindl.
-ex Hildebr., \emph{Abies magnifica} A.
-Murr., \emph{Calocedrus decurrens} (Torr.) Florin, \emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
-Buchholz.
-In this study, we consider litter the top layer of leaves and small twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the previous year (mostly undecomposed).
-Leaf litter was collected in the summer (mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the study area.
-We collected from a minimum of four populations per species with each population 9--32\,km distant from its neighbors.
-An exception to this was \emph{A.
-magnifica} which is restricted to the higher elevations and could only be collected from one area, due to designated wilderness boundaries and logistical constraints.
-Within that area, \emph{A.
-magnifica} litter was collected from over 10 different individuals.
-For the remaining species, the collection involved 2--4 individual trees at each site, separated from one another by at least 10\,m.
-The litter was collected 2\,m away from the trunk to obtain a more uniform sample of leaf and fine branchlet litter, because bark and heavier branches tend to fall closer to the tree trunk.
-All litter was air dried to < 5\% fuel moisture by dry weight for weighing and dividing into moisture trials.
-The moisture manipulations and subsequent burning trials were conducted in April--August 2015.
+Field sites for this study were located in Sequoia and Kings Canyon National
+Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in
+mixed-conifer forest. We chose eight dominant tree species that are
+representative of this type of forest: \emph{Pinus jeffreyi} Grev. \& Balf.,
+\emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
+\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
+  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
+\emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
+Buchholz. In this study, we consider litter the top layer of leaves and small
+twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the
+previous year (mostly undecomposed). Leaf litter was collected in the summer
+(mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the
+study area. We collected from a minimum of four populations per species with
+each population 9--32\,km distant from its neighbors. An exception to this was
+\emph{A. magnifica} which is restricted to the higher elevations and could only
+be collected from one area, due to designated wilderness boundaries and
+logistical constraints. Within that area, \emph{A. magnifica} litter was
+collected from over 10 different individuals. For the remaining species, the
+collection involved 2--4 individual trees at each site, separated from one
+another by at least 10\,m. The litter was collected 2\,m away from the trunk to
+obtain a more uniform sample of leaf and fine branchlet litter, because bark
+and heavier branches tend to fall closer to the tree trunk. All litter was air
+dried to < 5\% fuel moisture by dry weight for weighing and dividing into
+moisture trials. The moisture manipulations and subsequent burning trials were
+conducted in April--August 2015.
 
@@
 
-% The dry-down curves indicated that the eight species varied in their moisture behavior (Fig \ref{fig-drydown}).
-To determine a subset of species to use in the mixture trials, we performed analysis of contrasts using lsmeans package in R (TODO Lenth 2016) to guide choosing a subset of species for the litter mixtures.
-Based on the marginal means analysis, we chose four species to maximize variation in water retention and drying rate.
-Creating all possible three species mixtures from these four species resulted in four distinct three-species mixtures.
-
-% Leaf trait data
-
 With only eight species in this study, our power to test leaf trait effects on
@@
 times at each moisture level. We used the initial dry-down experiment to guide
-our timeing so as to attempt to burn samples over a range of moisture contents.
+our timing so as to attempt to burn samples over a range of moisture contents.
 We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
@@
 after 30 seconds, no further attempt was made and if three samples of a species
-at a particular mositure level failed to ignite, no further attempts were made
+at a particular moisture level failed to ignite, no further attempts were made
 on that species/moisture combination and the replicate was allowed to continue
@@
 in de Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.
-
-% and nine type-K thermocouples attached to data loggers measured temperature every second at three distances from the ignition point and at three different depths below the litter surface.
 
@@
 
-Dry-down curves were created based on the drying experiment, fitted with an exponential decay curve to each species.:
+Dry-down curves were created based on the drying experiment, fitted with an
+exponential decay curve to each species.:
 
@@
 Where $m$ is moisture content at time $t$, and $\lambda$ is the exponential
-decay coefficient. To estimate these coefficents we fit a linear mixed-effects
+decay coefficient. To estimate these coefficients we fit a linear mixed-effects
 model by first taking the natural logarithm of moisture content, with time
@@
 flammability across these eight species was driven by leaf traits and litter
-packing. These previously reported flammability trials demonstrated that species within a genus tended to behave similarly and the two scale-leaved genera in the Cupressaceae,
-\emph{Calocedrus} and \emph{Sequoiadendron}, also behaved similarly to one
-another.
-Therefore, to preserve power used taxonomic groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our flammability functional groups and treated these as fixed effects in our models.
-The effect of fuel moisture and
-time since wetting on flammability parameters were modeled using linear
-models with fuel moisture or time since wetting and taxon as fixed effects and
-average vapor pressure deficit in the burning room during that trial as a
-nuisance covariate. We used linear mixed models to predict spread rate, but
-fuel consumption tends to exhibit threshold effects with few intermediate
-values. Therefore, we used mixed generalized linear models with a beta
-regression link with the `glmmADMB` package \cite{Skaug_Fournier_etal-2016} to
-model percent fuel consumed as a function moisture content and of time since
-wetting.
+packing. These previously reported flammability trials demonstrated that
+species within a genus tended to behave similarly and the two scale-leaved
+genera in the Cupressaceae, \emph{Calocedrus} and \emph{Sequoiadendron}, also
+behaved similarly to one another. Therefore, to preserve power used taxonomic
+groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our
+flammability functional groups and treated these as fixed effects in our
+models. The effect of fuel moisture and time since wetting on flammability
+parameters were modeled using linear models with fuel moisture or time since
+wetting and taxon as fixed effects and average vapor pressure deficit in the
+burning room during that trial as a nuisance covariate. We used linear mixed
+models to predict spread rate, but fuel consumption tends to exhibit threshold
+effects with few intermediate values. Therefore, we used mixed generalized
+linear models with a beta regression link with the `glmmADMB` package
+\cite{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function
+moisture content and of time since wetting.
 
@@
 mixture and time since wetting, we predicted moisture content based on the mean
-content of the three consituent species at that time since wetting. We
+content of the three constituent species at that time since wetting. We
 predicted flammability parameters based on the average flammability measures
@@
 
-There were marked differences in the extent of moisture species retain initially
-following wetting and draining by gravity (all $P$ < 0.0001, Fig.
+There were marked differences in the extent of moisture species retain
+initially following wetting and draining by gravity (all $P$ < 0.0001, Fig.
 \ref{fig:drydown} and Table \ref{tab:drydown}). \emph{Q. kelloggii}, the only
 broadleaf, is the species that retained the most moisture initially.
-\emph{Quercus}, like the pines, had a rapid drying rate (Fig.
-\ref{fig:drydown} and SI Fig. 1). Calculating estimated marginal means following the
-linear mixed-effects model discriminated the species into three different
-maximum water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P.
-  lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S.
-  giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa}
-(SI Fig. 1). There are also differences in desiccation rates across the eight
-species and estimated mixed linear model slopes discriminated species into
-three groups from lowest to most rapid: 1) \emph{C.decurrens}, 2) \emph{Abies
-  spp} and \emph{S. gianteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp}
-(SI Fig. 2).
+\emph{Quercus}, like the pines, had a rapid drying rate (Fig. \ref{fig:drydown}
+and SI Fig. 1). Calculating estimated marginal means following the linear
+mixed-effects model discriminated the species into three different maximum
+water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P. lambertiana} and
+\emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S. giganteum}, \emph{A.
+  concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa} (SI Fig. 1). There are
+also differences in desiccation rates across the eight species and estimated
+mixed linear model slopes discriminated species into three groups from lowest
+to most rapid: 1) \emph{C.decurrens}, 2) \emph{Abies spp} and \emph{S.
+  giganteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp} (SI Fig. 2).
 
@@
   gravity for 3 min before initial weighing. Colors indicate genera and there
-  were significant differences in both maximum water retention and in rate of drying (based linear model on log of water content). A small amount of horizontal jitter
-  (0.5 hr) was added to this plot to aid in visualizing because, although
-  collections were staggered, there were sometimes multiple trays collected at a
-  single time point.}
+  were significant differences in both maximum water retention and in rate of
+  drying (based linear model on log of water content). A small amount of
+  horizontal jitter (0.5 hr) was added to this plot to aid in visualizing
+  because, although collections were staggered, there were sometimes multiple
+  trays collected at a single time point.}
  \label{fig:drydown}
@@
 \begin{table}[H]
-  \caption{Linear mixed model results for moisture content as a function of time since
-    wetting. Approximate degrees of freedom (df), pseudo F statistics and p-values
-    were calculated by the Kenward-Roger approximation.
+  \caption{Linear mixed model results for moisture content as a function of
+    time since wetting. Approximate degrees of freedom (df), pseudo F
+    statistics and p-values were calculated by the Kenward-Roger approximation.
     \cite{Kenward_Roger-1997}.}
@@
 bulk density had a negative effect on desiccation rate (Fig. \ref{fig:bd-di},
-linear model $P$ = 0.029). Anova results are shown in Tables \ref{tab:mc_di_anova}.
+linear model $P$ = 0.029). Anova results are shown in Tables
+\ref{tab:mc_di_anova}.
 
@@
 \includegraphics[width=16cm]{figs_tables/fig2_SLA_maxMC.pdf}
-\caption{Maximum water retention by specific leaf area across eight species. ANOVA table in Table \ref{tab:mc_di_anova}.}
+\caption{Maximum water retention by specific leaf area across eight species.
+  ANOVA table in Table \ref{tab:mc_di_anova}.}
   \label{fig:maxmc-di}
@@
 \includegraphics[width=16cm]{figs_tables/fig3_di_bd.pdf}
-\caption{Drying rate rate as a function of specific leaf area and litter bulk density across eight species. ANOVA table in Table \ref{tab:mc_di_anova}}.
+\caption{Drying rate rate as a function of specific leaf area and litter bulk
+  density across eight species. ANOVA table in Table \ref{tab:mc_di_anova}}.
   \label{fig:bd-di}
@@
 \begin{table}[H]
-  \caption{Linear model results for maximum water retention (\% moisture content on dry mass basis) and for estimated drying rate (hr$^{-1}$) as functions of specific leaf area and litter bulk density across eight species.
-ANOVA tables shown with degrees of freedom, sum of squares, mean squares, $F$, and $P$ for each effect.}
+  \caption{Linear model results for maximum water retention (\% moisture
+    content on dry mass basis) and for estimated drying rate (hr$^{-1}$) as
+    functions of specific leaf area and litter bulk density across eight
+    species. ANOVA tables shown with degrees of freedom, sum of squares, mean
+    squares, $F$, and $P$ for each effect.}
   \label{tab:mc_di_anova}
@@
 
-Flame spread rate declined with fuel moisture (Fig.
-\ref{fig:spread_moist}).
-\emph{Quercus} had higher spread rates across all fuel moistures and had a steeper decline in response to moisture than did the other groups (taxon by time since wetting interaction $P$ = 0.022, Table \ref{tab:spreadrate_anova}).
-Fuel moisture negatively influenced fuel consumption.
-Overall fuel consumption and the rate of change with moisture content varied across the taxonomic groups (Fig.
-\ref{fig:consume_moist}, Table \ref{tab:consume_moist_anova}).
-Although the overall ranking of groups was similar for fuel consumption and for flame spread rate, in contrast to the pattern with spread rate, for fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in flammability with fuel moisture than did the other groups.
+Flame spread rate declined with fuel moisture (Fig. \ref{fig:spread_moist}).
+\emph{Quercus} had higher spread rates across all fuel moistures and had a
+steeper decline in response to moisture than did the other groups (taxon by
+time since wetting interaction $P$ = 0.022, Table \ref{tab:spreadrate_anova}).
+Fuel moisture negatively influenced fuel consumption. Overall fuel consumption
+and the rate of change with moisture content varied across the taxonomic groups
+(Fig. \ref{fig:consume_moist}, Table \ref{tab:consume_moist_anova}). Although
+the overall ranking of groups was similar for fuel consumption and for flame
+spread rate, in contrast to the pattern with spread rate, for fuel consumption,
+\emph{Pinus} and \emph{Abies} has a steeper decline in flammability with fuel
+moisture than did the other groups.
 
@@
 \includegraphics[width=8cm]{figs_tables/fig4_spread_actualMC.pdf}
-\caption{Flame spread rate by moisture content (model fit $F$= 61.15, df = 9, 148,  $P$ < 0.0001). Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI Table 2.}
+\caption{Flame spread rate by moisture content (model fit $F$= 61.15, df = 9,
+  148, $P$ < 0.0001). Model ANOVA table in Table \ref{tab:spreadrate_anova} and
+  coefficients in SI Table 2.}
   \label{fig:spread_moist}
@@
 \begin{table}[H]
-  \caption{Generalized linear mixed model results for two models predicting flame spread rate: the first model examined spread rate as a function of fuel moisture (mc) with taxonomic group as a fixed effect and species as a nested random effect.
-The second model predicted spread rate as a function of time since wetting (time) with taxonomic group as a fixed effect and species as a nested random effect.
-Approximate degrees of freedom (df), pseudo F statistics and p-values were calculated by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.
-Estimated coefficients are in SI Tables 2 and 4}
+  \caption{Generalized linear mixed model results for two models predicting
+    flame spread rate: the first model examined spread rate as a function of
+    fuel moisture (mc) with taxonomic group as a fixed effect and species as a
+    nested random effect. The second model predicted spread rate as a function
+    of time since wetting (time) with taxonomic group as a fixed effect and
+    species as a nested random effect. Approximate degrees of freedom (df),
+    pseudo F statistics and p-values were calculated by the Kenward-Roger
+    approximation \cite{Kenward_Roger-1997}. Estimated coefficients are in SI
+    Tables 2 and 4}
   \label{tab:spreadrate_anova}
@@
 \includegraphics[width=8cm]{figs_tables/fig5_consume_actualMC.pdf}
-\caption{Percent fuel consumed by moisture content. Lines indicate best fit beta regression lines. Model deviance table in Table \ref{tab:consume_anova} and coefficients are in SI Table 3.}
+\caption{Percent fuel consumed by moisture content. Lines indicate best fit
+  beta regression lines. Model deviance table in Table \ref{tab:consume_anova}
+  and coefficients are in SI Table 3.}
   \label{fig:consume_moist}
@@
 \begin{table}[H]
-  \caption{Generalized linear mixed model results for two sets of models predicting fuel consumption.
-The first set modeled fuel consumption as a function of fuel moisture (mc) with taxonomic group as a fixed effect and species as a nested random effect.
-The second set modeled fuel consumption as a function of as a function of time since wetting (time) with taxonomic group as a fixed effet and species as a nested random effect.
-The analysis of deviance table below shows sequential chi-square tests on nested beta regression models: from the simplest null model within each group with only intercept and random effect, for a taxon effect only model, for a taxon and moisture content effect, and for the full model including the taxon by moisture or time interaction.
-Coefficient estimates are in SI Tables 3 and 5.}
+  \caption{Generalized linear mixed model results for two sets of models
+    predicting fuel consumption. The first set modeled fuel consumption as a
+    function of fuel moisture (mc) with taxonomic group as a fixed effect and
+    species as a nested random effect. The second set modeled fuel consumption
+    as a function of as a function of time since wetting (time) with taxonomic
+    group as a fixed effect and species as a nested random effect. The analysis
+    of deviance table below shows sequential chi-square tests on nested beta
+    regression models: from the simplest null model within each group with only
+    intercept and random effect, for a taxon effect only model, for a taxon and
+    moisture content effect, and for the full model including the taxon by
+    moisture or time interaction. Coefficient estimates are in SI Tables 3 and
+    5.}
   \label{tab:consume_anova}
@@
 
-As a consequence of the relationships shown above, time since wetting had a significant effect on spread rate ($P$ < 0.0001) and the slope of this effect varied by taxonomic group (Fig.
-\ref{fig:spread_time}, Table \ref{tab:spreadrate_anova}).
-The pines were able to ignite and burn much sooner after wetting than did the other taxa, none of which had measurable spread rates until burning was attempted 72 hours after wetting (note that trials that failed to even ignite smoldering combustion are ommitted completely from the model and figure).
-At that point, the \emph{Abies}, \emph{Calocedrus}, and \emph{Sequiadendron} had dryer litter than did the \emph{Quercus}, yet the spread rates were slow.
-\emph{Quercus} had the steepest response to time since wetting (Fig.
-\ref{fig:spread_time}, SI Table 4)).
+As a consequence of the relationships shown above, time since wetting had a
+significant effect on spread rate ($P$ < 0.0001) and the slope of this effect
+varied by taxonomic group (Fig. \ref{fig:spread_time}, Table
+\ref{tab:spreadrate_anova}). The pines were able to ignite and burn much sooner
+after wetting than did the other taxa, none of which had measurable spread
+rates until burning was attempted 72 hours after wetting (note that trials that
+failed to even ignite smoldering combustion are omitted completely from the
+model and figure). At that point, the \emph{Abies}, \emph{Calocedrus}, and
+\emph{Sequoiadendron} had dryer litter than did the \emph{Quercus}, yet the
+spread rates were slow. \emph{Quercus} had the steepest response to time since
+wetting (Fig. \ref{fig:spread_time}, SI Table 4)).
 
@@
   Lines indicate best fit linear model fits. Model ANOVA table in Table
-  \ref{tab:spreadrate_anova} and coefficients in SI Table 4. Horizontal
-  jitter (up to 2 hr) was added to the plot for visualization and to avoid
-  overplotting. All collections were at the times labeled on the x-axis. A
-  lack of data at earlier time points for a taxon indicates no ignitions
-  occurred at those time points.}
+  \ref{tab:spreadrate_anova} and coefficients in SI Table 4. Horizontal jitter
+  (up to 2 hr) was added to the plot for visualization and to avoid
+  overplotting. All collections were at the times labeled on the x-axis. A lack
+  of data at earlier time points for a taxon indicates no ignitions occurred at
+  those time points.}
 \label{fig:spread_time}
@@
 
-Fuel consumption was highly variable across trials and intermediate values were less common than were very low or moderately high consumption (Fig.
-\ref{fig:consume_time}).
-All parameters significantly contributed to the generalized linear mixed beta regression model as indicated by the nested model comparisons in Table \ref{tab:consume_time_anova}.
-As with flame spread rate some pines had the most rapid response to drying.
-However, unlike in the other flammability analyses, here the pines did not all behave similarly: \emph{P.
-lambertiana} had lower fuel consumption than did \emph{P.
-ponderosa} and \emph{P.
-jeffreyii}.
-We noticed this during trials and in model fitting we found that placing \emph{P.
-lambertiana} in its own group allowed model convergence that was otherwise impossible.
-This was at odds with our intent to use only \emph{a priori} taxonomic groups, but we could not obtain model convergence otherwise.
-This result should therefore be seen as preliminary because the test does not reflect an \emph{a priori} hypothesis.
+Fuel consumption was highly variable across trials and intermediate values were
+less common than were very low or moderately high consumption (Fig.
+\ref{fig:consume_time}). All parameters significantly contributed to the
+generalized linear mixed beta regression model as indicated by the nested model
+comparisons in Table \ref{tab:consume_time_anova}. As with flame spread rate
+some pines had the most rapid response to drying. However, unlike in the other
+flammability analyses, here the pines did not all behave similarly: \emph{P.
+  lambertiana} had lower fuel consumption than did \emph{P. ponderosa} and
+\emph{P. jeffreyi}. We noticed this during trials and in model fitting we
+found that placing \emph{P. lambertiana} in its own group allowed model
+convergence that was otherwise impossible. This was at odds with our intent to
+use only \emph{a priori} taxonomic groups, but we could not obtain model
+convergence otherwise. This result should therefore be seen as preliminary
+because the test does not reflect an \emph{a priori} hypothesis.
 
@@
   species. Curves shown indicate best fit beta regression lines. Model ANOVA
-  table in Table \ref{tab:consume_anova} and coefficients in SI Table
-  4. Horizontal jitter (up to 2 hr) was added to the plot for visualization and
-  to avoid overplotting. All collections were at the times labeled on the
-  x-axis.. A lack of data at earlier time points for a taxon indicates no
-  ignitions occurred at those time points.}
+  table in Table \ref{tab:consume_anova} and coefficients in SI Table 4.
+  Horizontal jitter (up to 2 hr) was added to the plot for visualization and to
+  avoid overplotting. All collections were at the times labeled on the x-axis..
+  A lack of data at earlier time points for a taxon indicates no ignitions
+  occurred at those time points.}
   \label{fig:consume_time}
@@
 
-Based on visual inspection of the dry down curves and the estimated marginal means for maximum water retainability and desiccation rate (SI Figs 2--3), we selected four species to represent the full range of litter moisture dynamics for use in the litter mixtures and examined all four three-species combinations of those four.
-The four species were \emph{Abies concolor}, \emph{Calocedrus decurrens}, \emph{Pinus jeffreyi} and \emph{Quercus kellogii}.
+Based on visual inspection of the dry down curves and the estimated marginal
+means for maximum water retainability and desiccation rate (SI Figs 2--3), we
+selected four species to represent the full range of litter moisture dynamics
+for use in the litter mixtures and examined all four three-species combinations
+of those four. The four species were \emph{Abies concolor}, \emph{Calocedrus
+  decurrens}, \emph{Pinus jeffreyi} and \emph{Quercus kelloggii}.
 
@@
 
-Mixture dry down rates and maximum water retention showed some relationship to the dry down curves for individual constituent species (Fig.
-\ref{fig:mixture_dry_down}.
-For example, the mixture that does not contain \emph{Q.
-kelloggii} (AbCaPi in Fig.
-Fig.
-\ref{fig:mixture_dry_down}) has the lowest initial retention.
-Initial water retention and rates differed across mixtures (linear model based on log of moisture content, Table \ref{tab:mixtures_drydown}).
+Mixture dry down rates and maximum water retention showed some relationship to
+the dry down curves for individual constituent species (Fig.
+\ref{fig:mixture_dry_down}. For example, the mixture that does not contain
+\emph{Q. kelloggii} (AbCaPi in Fig. Fig. \ref{fig:mixture_dry_down}) has the
+lowest initial retention. Initial water retention and rates differed across
+mixtures (linear model based on log of moisture content, Table
+\ref{tab:mixtures_drydown}).
 
 \begin{table}[H]
-  \caption{Linear mixed model results for moisture content as a function of time since
-    wetting for four different litter mixture types. Approximate degrees of freedom (df), pseudo F statistics and p-values
-    were calculated by the Kenward-Roger approximation
-    \cite{Kenward_Roger-1997}.}
+  \caption{Linear mixed model results for moisture content as a function of
+    time since wetting for four different litter mixture types. Approximate
+    degrees of freedom (df), pseudo F statistics and p-values were calculated
+    by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
   \label{tab:mixtures_drydown}
@@
 
-There was evidence for non-additive effects in drying rate.
-Mixtures tended to have lower water content than predicted from the mean of the constituent species after wetting, but had higher water content than expected after long drying (Wald test shows slope of mixed linear model illustrated in Fig.
+There was evidence for non-additive effects in drying rate. Mixtures tended to
+have lower water content than predicted from the mean of the constituent
+species after wetting, but had higher water content than expected after long
+drying (Wald test shows slope of mixed linear model illustrated in Fig.
 \ref{fig:mixture_obs_pred_mc} less than 1, $P$ < 0.0001).
@@
 \includegraphics[width=16cm]{figs_tables/fig10_mixture_obs_vs_pred_spread.pdf}
-\caption{Observed vs.
-predicted flame spread rate for litter mixtures.
-Filled circles show observed spread rate values.
-Open diamonds indicate predicted spread rates based on the average spread rate at each time since wetting for the three constituent species in each mixture.
-Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
+\caption{Observed vs. predicted flame spread rate for litter mixtures. Filled
+  circles show observed spread rate values. Open diamonds indicate predicted
+  spread rates based on the average spread rate at each time since wetting for
+  the three constituent species in each mixture. Codes for species in mixtures:
+  Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi =
+  \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
   \label{fig:mixture_obs_pred_spread}
@@
 \includegraphics[width=16cm]{figs_tables/fig11_mixture_obs_vs_pred_consume.pdf}
-\caption{Observed vs predicted fuel consumption for litter mixtures.
-Filled circles show observed fuel consumption percentages.
-Open diamonds indicate predicted fuel consumption based on the average fuel consumption at each particular time since wetting for the three constituent species in each mixture.
-Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
+\caption{Observed vs predicted fuel consumption for litter mixtures. Filled
+  circles show observed fuel consumption percentages. Open diamonds indicate
+  predicted fuel consumption based on the average fuel consumption at each
+  particular time since wetting for the three constituent species in each
+  mixture. Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca =
+  \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu =
+  \emph{Quercus kellogii}.}
   \label{fig:mixture_obs_pred_consume}
@@
 
-Flammability measures also demonstrated non-additivity in mixtures.
-Flame spread rate was consistently higher than predicted (Fig \ref{fig:mixture_obs_pred_spread}, Wald test of mixed linear model centered intercept for observed-predicted values greater than zero, $P$ = 0.017) and the positive non-additivity increased with time since wetting (slope greater than zero, $P$ = 0.005).
-Fuel consumption, however, showed no overall bias towards positive or negative non-additivity, but non-additivity increased with time since wetting (Fig.
-\ref{fig:mixture_obs_pred_consume}, slope not equal to zero, $P$ < 0.0001).
+Flammability measures also demonstrated non-additivity in mixtures. Flame
+spread rate was consistently higher than predicted (Fig
+\ref{fig:mixture_obs_pred_spread}, Wald test of mixed linear model centered
+intercept for observed-predicted values greater than zero, $P$ = 0.017) and the
+positive non-additivity increased with time since wetting (slope greater than
+zero, $P$ = 0.005). Fuel consumption, however, showed no overall bias towards
+positive or negative non-additivity, but non-additivity increased with time
+since wetting (Fig. \ref{fig:mixture_obs_pred_consume}, slope not equal to
+zero, $P$ < 0.0001).
 
@@
 
-The species in this study displayed different patterns of both moisture absorption and desorption, apparently controlled by different litter traits.
-In this sample of species, those that have a greater absorptive capacity are also among the most flammable in terms of spread rate when dry.
-This highlights the importance of fuel moisture in modulating flammability.
-One shortcoming of past work in this system \cite{Magalhaes+Schwilk-2012} was that flammability trials tested only very dry fuels.
-We therefore expected that rankings of species flammability might change under more realistic moisture levels.
-For example, one presumption of ours prior to this experiment was that the high flammability previously reported for \emph{Q.
-kelloggii} \cite{Magalhaes+Schwilk-2012} would disappear under higher fuel moisture contents.
-This prediction was partly supported: \emph{Q.
-kelloggii} fails to ignite until after 72 hours of drying in this study (Figs.
-\ref{fig:spread_time} and \ref{fig:consume_time}).
-However, high moisture content in this oak was unable to suppress ignition of mixtures to which it contributed (Figs.
-\ref{fig:mixture_obs_pred_spread} and \ref{fig:mixture_obs_pred_consume}).
+The species in this study displayed different patterns of both moisture
+absorption and desorption, apparently controlled by different litter traits. In
+this sample of species, those that have a greater absorptive capacity are also
+among the most flammable in terms of spread rate when dry. This highlights the
+importance of fuel moisture in modulating flammability. One shortcoming of past
+work in this system \cite{Magalhaes+Schwilk-2012} was that flammability trials
+tested only very dry fuels. We therefore expected that rankings of species
+flammability might change under more realistic moisture levels. For example,
+one presumption of ours prior to this experiment was that the high flammability
+previously reported for \emph{Q. kelloggii} \cite{Magalhaes+Schwilk-2012} would
+disappear under higher fuel moisture contents. This prediction was partly
+supported: \emph{Q. kelloggii} fails to ignite until after 72 hours of drying
+in this study (Figs. \ref{fig:spread_time} and \ref{fig:consume_time}).
+However, high moisture content in this oak was unable to suppress ignition of
+mixtures to which it contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
+\ref{fig:mixture_obs_pred_consume}).
 
-    The dry down curves we measured demonstrate that after a rain event, the litter of different species will exhibit different moisture contents over time as a consequence of both differential initial absorption and of differences in drying rate.
-This has consequences for flammability as a function of species and time since wetting.
-Our flammability trials demonstrate that that both flame spread rate and fuel consumption increased with time since wetting but that the rate of increase varied across these taxa.
-Only the pines are able to ignite and support flame spread immediately after wetting (Fig.
-\ref{fig:spread_time}).
-In contrast, \emph{Quercus} exhibited a threshold response to drying with no ignition prior to 72 hours of drying but once ignition is possible, having faster spread rates than the other taxa.
-The Cupressaceae, which have dense litter but low maximum water absorption had low flammability across moisture levels.
+The dry down curves we measured demonstrate that after a rain event, the litter
+of different species will exhibit different moisture contents over time as a
+consequence of both differential initial absorption and of differences in
+drying rate. This has consequences for flammability as a function of species
+and time since wetting. Our flammability trials demonstrate that that both
+flame spread rate and fuel consumption increased with time since wetting but
+that the rate of increase varied across these taxa. Only the pines are able to
+ignite and support flame spread immediately after wetting (Fig.
+\ref{fig:spread_time}). In contrast, \emph{Quercus} exhibited a threshold
+response to drying with no ignition prior to 72 hours of drying but once
+ignition is possible, having faster spread rates than the other taxa. The
+Cupressaceae, which have dense litter but low maximum water absorption had low
+flammability across moisture levels.
 
-With eight species, power to detect species-specific trait effects on these moisture patterns was weak, but we examined one litter trait known to directly influence flammability: litter bulk density and one leaf trait, specific leaf area.
-Litter density is largely driven by leaf size in this system and, in turn, controls the major axis of flammability variation in dry litter \cite{Magalhaes+Schwilk-2012}.
-Our results suggest that moisture dynamics are controlled by both traits intrinsic to the leaf particles (specific leaf area) and an emergent property of the litter bed (bulk density).
-Specific leaf area here is probably capturing some joint effect of cuticle thickness and total surface area of litter particles, but we cannot test the morphological and physical mechanisms of water absorption with these data.
-The oak has much higher initial water retention than do the other species, followed by the finest-needled pine, \emph{P.
-lambertiana}.
-Our study examines fresh litter, but we expect that decomposition will affect moisture dynamics.
-For example, Van Wagner \cite{Van_Wagner-1969}, investigating fuel weathering, found that pine needles had a variation in moisture content 3 to 6 times greater when they lacked their protective wax coating when compared to intact needles.
+With eight species, power to detect species-specific trait effects on these
+moisture patterns was weak, but we examined one litter trait known to directly
+influence flammability: litter bulk density and one leaf trait, specific leaf
+area. Litter density is largely driven by leaf size in this system and, in
+turn, controls the major axis of flammability variation in dry litter
+\cite{Magalhaes+Schwilk-2012}. Our results suggest that moisture dynamics are
+controlled by both traits intrinsic to the leaf particles (specific leaf area)
+and an emergent property of the litter bed (bulk density). Specific leaf area
+here is probably capturing some joint effect of cuticle thickness and total
+surface area of litter particles, but we cannot test the morphological and
+physical mechanisms of water absorption with these data. The oak has much
+higher initial water retention than do the other species, followed by the
+finest-needled pine, \emph{P. lambertiana}. Our study examines fresh litter,
+but we expect that decomposition will affect moisture dynamics. For example,
+Van Wagner \cite{Van_Wagner-1969}, investigating fuel weathering, found that
+pine needles had a variation in moisture content 3 to 6 times greater when they
+lacked their protective wax coating when compared to intact needles.
 
@@
 
-We measured drying rates and flammability of four unique three-species litter mixtures.
-These mixtures contained species representative of all dry-down behavior types observed in single species trials.
-As expected from the single species results, mixtures that contained \emph{Q.
-kelloggii} absorbed more water during wetting, and the one mixture without any Cupressaceae had the most rapid drying rate (Figure \ref{fig:mixture_dry_down} and Table \ref{tab:mixtures_drydown}).
-Drydown rates, however, did not behave as the average of the constituent species: mixtures tended to be drier than expected early in the process and wetter than expected later (Fig.
-\ref{fig:mixture_obs_pred_mc}).
-The fact that mixtures tended to have lower water content than predicted from the mean of the constituent species after wetting, but higher water content than expected after long drying is surprising because past work has shown that
-mixtures tend to pack less densely than do litter monocultures \cite[][, Table 2]{Magalhaes+Schwilk-2012}.
-If packing density controls rate of drying, then we would expect mixtures to dry more rapidly than predicted from monoculture trials.
-This pattern in mositure contrent through time, however, did not directly predict flammability through time.
+We measured drying rates and flammability of four unique three-species litter
+mixtures. These mixtures contained species representative of all dry-down
+behavior types observed in single species trials. As expected from the single
+species results, mixtures that contained \emph{Q. kelloggii} absorbed more
+water during wetting, and the one mixture without any Cupressaceae had the most
+rapid drying rate (Figure \ref{fig:mixture_dry_down} and Table
+\ref{tab:mixtures_drydown}). Drydown rates, however, did not behave as the
+average of the constituent species: mixtures tended to be drier than expected
+early in the process and wetter than expected later (Fig.
+\ref{fig:mixture_obs_pred_mc}). The fact that mixtures tended to have lower
+water content than predicted from the mean of the constituent species after
+wetting, but higher water content than expected after long drying is surprising
+because past work has shown that mixtures tend to pack less densely than do
+litter monocultures \cite[][, Table 2]{Magalhaes+Schwilk-2012}. If packing
+density controls rate of drying, then we would expect mixtures to dry more
+rapidly than predicted from monoculture trials. This pattern in moisture
+content through time, however, did not directly predict flammability through
+time.
 
@@
 than that reported by Blauw et al \citeyear{Blauw+Wensink+etal-2015}. Blauw et
-al found that non-addivity was greater in moist mixtures of heathland fuels
-than in dry fuels. In their work, non-addiitvity was generally negative but cases of 
-positive non-addivity in mixtures was often a result of mixtures meeting
+al found that non-additivity was greater in moist mixtures of heathland fuels
+than in dry fuels. In their work, non-additivity was generally negative but cases of 
+positive non-additivity in mixtures was often a result of mixtures meeting
 ignition thresholds not met by the constituent species alone. To limit the
@@
 Ongoing climate change and concurrent changes to fire regimes are linked
-phenomena. The patterns we demonstrate here show to potential importance of species
-traits and interactions at small scales and among subtle shifts in vegetation,
-but vegetation-fire feedbacks are important at a wide range of scales from stand-level species composition shifts to the earth system \cite{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}.
+phenomena. Vegetation-fire feedbacks are important at a wide range of scales
+from stand-level species composition shifts to the earth system
+\cite{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}. The patterns we
+demonstrate here show to potential importance of species traits and
+interactions at local and stand scales where management and fire operations
+meet species' biology.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\acknowledgments{The authors thank the National Park Service and the employees of Sequoia and Kings Canyon National Parks.This work was carried out with permission from NPS under permit \# SEKI-2014-SCI-0040.
-We also thank students who assisted with field work and the litter collection, including Lizz Waring, Taylor Brown and Chelsea Griffin.}
+\acknowledgments{The authors thank the National Park Service and the employees
+  of Sequoia and Kings Canyon National Parks.This work was carried out with
+  permission from NPS under permit \# SEKI-2014-SCI-0040. We also thank
+  students who assisted with field work and the litter collection, including
+  Lizz Waring, Taylor Brown and Chelsea Griffin. We thank the Lubbock Fire
+  Department for their ongoing help and cooperation.}
 

### 2019-07-25 ms/jae/jae-example.tex
Add J of Animal Ecology latex style

\documentclass[a4paper,12pt]{article}
\usepackage{jae}

\title{Un superbe titre qui peut être un peu long mais c'est pas très génant parce que
  c'est bien géré.}
\running{Le titre en plus court.}

\author{Machin Bidule$^{1,2,3,*}$, \and Truc Chouette$^{2}$, \and Bob Morane$^{3}$, \and
  John Doe$^{3,4}$, \and \& Joseph Ratzinger$^{4}$.}

\affiliations{
\item Université de Lyon, F-69\,000, Lyon ; Université Lyon 1 ; CNRS, UMR 5558,
  Laboratoire de Biométrie et Biologie Évolutive, F-69\,622, Villeurbanne, France.
\item Centre for Conservation Biology, Department of Biology; Norwegian University of
  Science and Technology; NO-7491 Trondheim, Norway.
\item Norwegian Institute for Nature Research; Tungasletta 2, NO-7485 Trondheim, Norway.
\item Northern Research Institute Tromsø, Postboks 6434 NO-9294 Tromsø, Norway.}

\nwords{3450}
\ntables{4}
\nfig{4}
\nref{59}

\corr{\url{bidule@truc.com}}

\begin{document}

\maketitle

\begin{abstract}
  \noindent \begin{enumerate}
  \item Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Sed non
    risus. Suspendisse lectus tortor, dignissim sit amet, adipiscing nec, ultricies sed,
    dolor. Cras elementum ultrices diam.
  \item Maecenas ligula massa, varius a, semper congue, euismod non, mi. Proin porttitor,
    orci nec nonummy molestie, enim est eleifend mi, non fermentum diam nisl sit amet
    erat. Duis semper. Duis arcu massa, scelerisque vitae, consequat in, pretium a,
    enim. Pellentesque congue.
  \item Ut in risus volutpat libero pharetra tempor. Cras vestibulum bibendum
    augue. Praesent egestas leo in pede. Praesent blandit odio eu enim. Pellentesque sed
    dui ut augue blandit sodales.
  \item \emph{Synthesis and applications.} Vestibulum ante ipsum primis in faucibus orci
    luctus et ultrices posuere cubilia Curae; Aliquam nibh. Mauris ac mauris sed pede
    pellentesque fermentum. Maecenas adipiscing ante non diam sodales hendrerit.
  \end{enumerate}
\end{abstract}

\noindent \textbf{Keywords:} Lorem ipsum, dolor sit amet, consectetuer, adipiscing,
suspendisse

\newpage

\section*{Introduction}

Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Sed non risus. Suspendisse
lectus tortor, dignissim sit amet, adipiscing nec, ultricies sed, dolor. Cras elementum
ultrices diam. Maecenas ligula massa, varius a, semper congue, euismod non, mi. Proin
porttitor, orci nec nonummy molestie, enim est eleifend mi, non fermentum diam nisl sit
amet erat. Duis semper \citep{Bivand_2007,Broeseth_2007_REP}. Duis arcu massa, scelerisque
vitae, consequat in, pretium a, enim. Pellentesque congue. Ut in risus volutpat libero
pharetra tempor. Cras vestibulum bibendum augue. Praesent egestas leo in pede. Praesent
blandit odio eu enim. Pellentesque sed dui ut augue blandit sodales. Vestibulum ante ipsum
primis in faucibus orci luctus et ultrices posuere cubilia Curae; Aliquam nibh. Mauris ac
mauris sed pede pellentesque fermentum. Maecenas adipiscing ante non diam sodales
hendrerit.

Ut velit mauris, egestas sed, gravida nec, ornare ut, mi. Aenean ut orci vel massa
suscipit pulvinar. Nulla sollicitudin. Fusce varius, ligula non tempus aliquam, nunc
turpis ullamcorper nibh, in tempus sapien eros vitae ligula. Pellentesque rhoncus nunc et
augue. Integer id felis. Curabitur aliquet pellentesque diam. Integer quis metus vitae
elit lobortis egestas. Lorem ipsum dolor sit amet, consectetuer adipiscing elit. Morbi vel
erat non mauris convallis vehicula. Nulla et sapien. Integer tortor tellus, aliquam
faucibus, convallis id, congue eu, quam. Mauris ullamcorper felis vitae erat. Proin
feugiat, augue non elementum posuere \citep{Bullard_1991_MTh}, metus purus iaculis lectus,
et tristique ligula justo vitae magna.

Aliquam convallis sollicitudin purus. Praesent aliquam, enim at fermentum mollis, ligula
massa adipiscing nisl, ac euismod nibh nisl eu lectus. Fusce vulputate sem at
sapien. Vivamus leo. Aliquam euismod libero eu enim. Nulla nec felis sed leo placerat
imperdiet. Aenean suscipit nulla in justo
\citep{Morrison_2006_Bk,Pimlott_1967,Quinn_2003}. Suspendisse cursus rutrum augue. Nulla
tincidunt tincidunt mi. Curabitur iaculis, lorem vel rhoncus faucibus, felis magna
fermentum augue, et ultricies lacus lorem varius purus. Curabitur eu amet.

\section*{Material and methods}

\subsection*{Study area}

\citet{Fournier_1982} sed ut perspiciatis, unde omnis iste natus error sit voluptatem
accusantium doloremque laudantium, totam rem aperiam eaque ipsa, quae ab illo inventore
veritatis et quasi architecto beatae vitae dicta sunt, explicabo. Nemo enim ipsam
voluptatem, quia voluptas sit, aspernatur aut odit aut fugit, sed quia consequuntur magni
dolores eos, qui ratione voluptatem sequi nesciunt, neque porro quisquam est, qui dolorem
ipsum, quia dolor sit, amet, consectetur, adipisci velit, sed quia non numquam eius modi
tempora incidunt, ut labore et dolore magnam aliquam quaerat voluptatem. Ut enim ad minima
veniam, quis nostrum exercitationem ullam corporis suscipit laboriosam, nisi ut aliquid ex
ea commodi consequatur? Quis autem vel eum iure reprehenderit, qui in ea voluptate velit
esse, quam nihil molestiae consequatur, vel illum, qui dolorem eum fugiat, quo voluptas
nulla pariatur?

At vero eos et accusamus et iusto odio dignissimos ducimus, qui blanditiis praesentium
voluptatum deleniti atque corrupti, quos dolores et quas molestias excepturi sint,
\citet{Hirzel_2003_PROC,Holand_1998,Looijen_1998_PhD} obcaecati cupiditate non provident,
similique sunt in culpa, qui officia deserunt mollitia animi, id est laborum et dolorum
fuga. Et harum quidem rerum facilis est et expedita distinctio. Nam libero tempore, cum
soluta nobis est eligendi optio, cumque nihil impedit, quo minus id, quod maxime placeat,
facere possimus, omnis voluptas assumenda est, omnis dolor repellendus. Temporibus autem
quibusdam et aut officiis debitis aut rerum necessitatibus saepe eveniet, ut et voluptates
repudiandae sint et molestiae non recusandae. Itaque earum rerum hic tenetur a sapiente
delectus, ut aut reiciendis voluptatibus maiores alias consequatur aut perferendis
doloribus asperiores repellat.

\subsection*{Statistical analyses}

\subsubsection*{First ones}

Sed ut perspiciatis, unde omnis iste natus error sit voluptatem accusantium doloremque
laudantium, totam rem aperiam eaque ipsa, quae ab illo inventore veritatis et quasi
architecto beatae vitae dicta sunt, explicabo. Nemo enim ipsam voluptatem, quia voluptas
sit, aspernatur aut odit aut fugit, sed quia consequuntur magni dolores eos, qui ratione
voluptatem sequi nesciunt, neque porro quisquam est, qui dolorem ipsum, quia dolor sit,
amet \citep{Ray_2005_Bk}, consectetur, adipisci velit, sed quia non numquam eius modi
tempora incidunt, ut labore et dolore magnam aliquam quaerat voluptatem. Ut enim ad minima
veniam, quis nostrum exercitationem ullam corporis suscipit laboriosam, nisi ut aliquid ex
ea commodi consequatur? Quis autem vel eum iure reprehenderit, qui in ea voluptate velit
esse, quam nihil molestiae consequatur, vel illum, qui dolorem eum fugiat, quo voluptas
nulla pariatur?

\subsubsection*{Next ones}

At vero eos et accusamus et iusto odio dignissimos ducimus, qui blanditiis praesentium
voluptatum deleniti atque corrupti (Table~\ref{Tab1}), quos dolores et quas molestias
excepturi sint, obcaecati cupiditate non provident, similique sunt in culpa, qui officia
deserunt mollitia animi, id est laborum et dolorum fuga. Et harum quidem rerum facilis est
et expedita distinctio. Nam libero tempore, cum soluta nobis est eligendi optio, cumque
nihil impedit, quo minus id, quod maxime placeat, facere possimus, omnis voluptas
assumenda est, omnis dolor repellendus \citep{Gibeau_1998}. Temporibus autem quibusdam et
aut officiis debitis aut rerum necessitatibus saepe eveniet, ut et voluptates repudiandae
sint et molestiae non recusandae. Itaque earum rerum hic tenetur a sapiente delectus, ut
aut reiciendis voluptatibus maiores alias consequatur aut perferendis doloribus asperiores
repellat.

\section*{Results}

Hanc ego cum teneam sententiam, quid est cur verear, ne ad eam non possim accommodare
Torquatos nostros? quos tu paulo ante cum memoriter, tum etiam erga nos amice et benivole
collegisti, nec me tamen laudandis maioribus meis corrupisti nec segniorem ad respondendum
reddidisti. quorum facta quem ad modum, quaeso, interpretaris? sicine eos censes aut in
armatum hostem impetum fecisse aut in liberos atque in sanguinem suum tam crudelis fuisse,
nihil ut de utilitatibus, nihil ut de commodis suis cogitarent (see Fig.~\ref{Fig1})?
at id ne ferae quidem faciunt, ut ita ruant itaque turbent, ut earum motus et impetus quo
pertineant non intellegamus, tu tam egregios viros censes tantas res gessisse sine causa?
[35] Quae fuerit causa, mox videro; interea hoc tenebo, si ob aliquam causam ista, quae
sine dubio praeclara sunt, fecerint, virtutem iis per se ipsam causam non fuisse. --
Torquem detraxit hosti. -- Et quidem se texit, ne interiret. -- At magnum periculum
adiit. -- In oculis quidem exercitus. -- Quid ex eo est consecutus? -- Laudem et
caritatem, quae sunt vitae sine metu degendae praesidia firmissima. -- Filium morte
multavit. -- Si sine causa, nollem me ab eo ortum, tam inportuno tamque crudeli; sin, ut
dolore suo sanciret militaris imperii disciplinam exercitumque in gravissimo bello
animadversionis metu contineret, saluti prospexit civium, qua intellegebat contineri
suam. atque haec ratio late patet. [36] In quo enim maxime consuevit iactare vestra se
oratio, tua praesertim, qui studiose antiqua persequeris, claris et fortibus viris
commemorandis eorumque factis non emolumento aliquo, sed ipsius honestatis decore
laudandis, id totum evertitur eo delectu rerum, quem modo dixi, constituto, ut aut
voluptates omittantur maiorum voluptatum adipiscendarum causa aut dolores suscipiantur
maiorum dolorum effugiendorum gratia.

\section*{Discussion}

Sed de clarorum hominum factis illustribus et gloriosis satis hoc loco dictum sit. erit
enim iam de omnium virtutum cursu ad voluptatem proprius disserendi locus. nunc autem
explicabo, voluptas ipsa quae qualisque sit, ut tollatur error omnis imperitorum
intellegaturque ea, quae voluptaria, delicata, mollis habeatur disciplina, quam gravis,
quam continens, quam severa sit. Non enim hanc solam sequimur, quae suavitate aliqua
naturam ipsam movet et cum iucunditate quadam percipitur sensibus, sed maximam voluptatem
illam habemus, quae percipitur omni dolore detracto, nam quoniam, cum privamur dolore,
ipsa liberatione et vacuitate omnis molestiae gaudemus, omne autem id, quo gaudemus,
voluptas est, ut omne, quo offendimur, dolor, doloris omnis privatio recte nominata est
voluptas. ut enim, cum cibo et potione fames sitisque depulsa est, ipsa detractio
molestiae consecutionem affert voluptatis, sic in omni re doloris amotio successionem
efficit voluptatis (Fig.~\ref{Fig2}).

Itaque non placuit Epicuro medium esse se quiddam inter dolorem et voluptatem; illud enim
ipsum, quod quibusdam medium videretur, cum omni dolore careret, non modo voluptatem esse,
verum etiam summam voluptatem. quisquis enim sentit, quem ad modum sit affectus, eum
necesse est aut in voluptate esse aut in dolore. omnis autem privatione doloris putat
Epicurus terminari summam voluptatem, ut postea variari voluptas distinguique possit,
augeri amplificarique non possit.

At etiam Athenis, ut e patre audiebam facete et urbane Stoicos irridente, statua est in
Ceramico Chrysippi sedentis porrecta manu, quae manus significet illum in hae esse
rogatiuncula delectatum: 'Numquidnam manus tua sic affecta, quem ad modum affecta nunc
est, desiderat?' -- Nihil sane. -- 'At, si voluptas esset bonum, desideraret.' -- Ita
credo. -- 'Non est igitur voluptas bonum.' Hoc ne statuam quidem dicturam pater aiebat, si
loqui posset. conclusum est enim contra Cyrenaicos satis acute, nihil ad Epicurum. nam si
ea sola voluptas esset, quae quasi titillaret sensus, ut ita dicam, et ad eos cum
suavitate afflueret et illaberetur, nec manus esse contenta posset nec ulla pars vacuitate
doloris sine iucundo motu voluptatis. sin autem summa voluptas est, ut Epicuro placet,
nihil dolere, primum tibi recte, Chrysippe, concessum est nihil desiderare manum, cum ita
esset affecta, secundum non recte, si voluptas esset bonum, fuisse desideraturam. idcirco
enim non desideraret, quia, quod dolore caret, id in voluptate est.

\section*{Acknowledgments}

A lot of people are to thank here.

\newpage

\bibliography{BiblioExample}

\newpage

\section*{Tables}

\begin{table}[h!]
  \caption{A first table caption.}
  \label{Tab1}
  \begin{center}
    \begin{tabular}{p{3cm}p{10cm}}
      Name & Description \\
      \hline
      Agri & Proportion of agricultural areas \\
      Alpine & Proportion of alpine areas \\
      Bare & Proportion of bare ground \\
      DEM & Mean elevation \\
      DEMslope & Mean slope \\
      \hline
    \end{tabular}
  \end{center}
\end{table}

\newpage

\begin{table}[h!]
  \caption{A second table caption, longer than the first one that was
quite short. Indeed, it was supposed to be short, at the contrary of this one which is
much more informative than the previous one.}
  \label{Tab2}
  \begin{center}
    \begin{tabular}{lrrr}
      Name & Mar & Spe1 & Spe2 \\
      \hline
      Agri & -0.050 & 0.026 & 0.173 \\
      Alpine & -0.874 & -0.139 & 0.184 \\
      Bare & -0.555 & -0.922 & 0.084 \\
      DEM & -0.796 & -0.100 & 0.095 \\
      DEMslope & -0.167 & -0.205 & 0.013 \\
      \hline
    \end{tabular}
  \end{center}
\end{table}

\newpage

\section*{Figures}

\begin{figure}[h!]
  \caption{What a nice figure\dots}
  \label{Fig1}
  \begin{center}
    \includegraphics[width=6cm]{Fig1}
  \end{center}
\end{figure}

\newpage

\begin{figure}[h!]
  \caption{Even nicer!}
  \label{Fig2}
  \begin{center}
    \includegraphics[width=6cm]{Fig2}
  \end{center}
\end{figure}

\end{document}

### 2019-07-25 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Format for Journal of ecology
Still some things to do:

1. Check journal limits on number of figures, move to SI if necessary
2. Check figures for font sizes, correct "dessication rate", etc.
3. Correct some bib isse (Nelson Jr, etc)
4. Check table formatting and spacing
5. Finnish and proof SI

@@
-%% reftex-default-bibliography: ("/home/schwilk/write/bib/schwilk.bib")
+\documentclass[letterpaper,12pt]{article}
+\usepackage{jae/jae}
+\usepackage{booktabs}
 
-% LaTeX support: latex@mdpi.com 
-%  In case you need support, please attach all files that are necessary for compiling as well as the log file, and specify the details of your LaTeX setup (which operating system and LaTeX version  tools you are using).
+\title{Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest}
 
-%  =================================================================
-\documentclass[fire,article,submit,moreauthors,pdftex]{Definitions/mdpi}
+\running{Leaf litter flammability}
 
-%\usepackage[utf8]{inputenc}
+\author{Rita Quinones de Magalhaes$^{1,2}$ and Dylan W. Schwilk$^1$}
 
-% If you would like to post an early version of this manuscript as a preprint, you may use preprint as the journal and change 'submit' to 'accept'.
-The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}.
-This is especially recommended for submission to arXiv, where line numbers should be removed before posting.
-For preprints.org, the editorial staff will make this change immediately prior to posting.
+\affiliations{
+    \item Texas Tech University, Lubbock, Texas 79411, USA.
+    \item Current address: Rochester Institute of Technology, Rochester, NY 14623, USA.
+}
 
-%--------------------
-% Class Options:
-%--------------------
+\nwords{X}
+\ntables{X}
+\nfig{X}
+\nref{X}
 
-%----------
-% journal
-%----------
-%=================================================================
-\firstpage{1} 
-\makeatletter 
-\setcounter{page}{\@firstpage} 
-\makeatother
-\pubvolume{xx}
-\issuenum{1}
-\articlenumber{5}
-\pubyear{2019}
-\copyrightyear{2019}
-%\externaleditor{Academic Editor: name}
-\history{Received: date; Accepted: date; Published: date}
-%\updates{yes} % If there is an update available, un-comment this line
+\corr{\url{dylan.schwilk@ttu.edu}}
 
-%% MDPI internal command: uncomment if new journal that already uses continuous page numbers 
-%\continuouspages{yes}
+\begin{document}
 
-%------------------------------------------------------------------
-% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
-%\pdfoutput=1
+\maketitle
 
-%=================================================================
-% Add packages and commands here.
-The following packages are loaded in our class file: fontenc, calc, indentfirst, fancyhdr, graphicx, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, amsthm, hyphenat, natbib, hyperref, footmisc, geometry, caption, url, mdframed, tabto, soul, multirow, microtype, tikz
+\begin{abstract}
+  \noindent \begin{enumerate}
 
-%\bibliography{/home/schwilk/write/bib/schwilk.bib}
-
-%=================================================================
-%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition
-%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).
-
-%=================================================================
-% Full title of the paper (Capitalized)
-\Title{Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest}
-
-% Author Orchid ID: enter ID or remove command
-\newcommand{\orcidauthorA}{0000-0000-000-000X} % Add \orcidA{} behind the author's name
-%\newcommand{\orcidauthorB}{0000-0000-000-000X} % Add \orcidB{} behind the author's name
-
-% Authors, for the paper (add full first names)
-\Author{Rita Quinones de Magalhaes $^{\dagger}$ and Dylan W. Schwilk}
-
-% Authors, for metadata in PDF
-\AuthorNames{Rita Magalh\~{a}es, Dylan W. Schwilk}
-
-% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
-\address[1]{Texas Tech University}
-
-% Contact information of the corresponding author
-\corres{Correspondence: dylan.schwilk@ttu.edu, DWS}
-
-% Current address and/or shared authorship
-\firstnote{Current address: Rochester Institute of Technology, Rochester, NY 14623} 
-%\secondnote{These authors contributed equally to this work.}
-% The commands \thirdnote{} till \eighthnote{} are available for further notes
-
-%\simplesumm{} % Simple summary
-
-%\conference{} % An extended version of a conference paper
-
-% Abstract (Do not insert blank lines, i.e. \\) 
-\abstract{ Moisture content is a strong determinant of forest fuel flammability
-  and varies on small spatial scales and over short time periods. Emperical
+  \item Moisture content is a strong determinant of forest fuel flammability
+  and varies on small spatial scales and over short time periods. Empirical
   modeling of equilibrium dead fine fuel moisture is well established, but less
@@
   release moisture through time is critical to the assessment of fuel’s
-  availability to burn. Furthermore, Natural forests are often comprised of
+  availability to burn.
+  \item Natural forests are often comprised of
   mixed stands, leading to mixtures of leaf litter on the surface fuelling the
@@
   moisture synergies in mixed litter and of moisture effects on the
-  flammability of mixed litter. The litter from eight species of a
+  flammability of mixed litter.
+  \item The litter from eight species of a
   mixed-conifer forest was saturated and allowed to dry to determine moisture
@@
   litter beds of single species and of mixtures to establish flammability
-  response to the dry-down process. Moisture dynamics vary across species with
+  response to the dry-down process.
+  \item Moisture dynamics varied across species with
   higher specific leaf area being associated with higher maximum absorption and
@@
   that produce more aerated litter beds (such as \emph{Q. kelloggii} and
-  \emph{Pinus} species) have faster rates of litter drying. We examined two
+  \emph{Pinus} species) have faster rates of litter drying.
+  \item We examined two
   axes of flammability variation (spread rate and fuel consumption) and found
@@
   strongly differing effects on flammability across the major litter types
-  typified by different tree genera. Finally, we found that litter mixtures
+  typified by different tree genera. We found that litter mixtures
   exhibited consistent non-additivity in flame spread rate in fuel consumption.
-  This non-additivity became more positive with fuel drying. Litter mixtures
-  burn with fire behavior more similar to that of the most flammable
-  constituent species and this effect increases as fuels dry. ADD Synthesis
-  statement.}
+  This non-additivity became more positive with fuel drying.
+    \item \emph{Synthesis:} Litter mixtures burn with fire
+  behavior more similar to that of the most flammable constituent species and
+  this effect increases as fuels dry. statement. Such positive feedback effects
+  increases the potential for species to exert community scale effects even
+  when relatively rare and that can influence community assembly and the
+  evolutionary of niche-constructing traits.
+  \end{enumerate}
+\end{abstract}
 
-% Keywords
-\keyword{flammability, litter, flame spread rate}
+\noindent \textbf{Keywords:} California, flame spread rate, flammability, fuel consumption, fuel moisture, leaf litter, non-additivity
 
-%\setcounter{secnumdepth}{4}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{document}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\newpage
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 humidity, and wind drive the moisture content of the fuel available to burn
-\cite{Kreye+Varner+etal-2018}. In particular, fine dead fuels respond very
-readily to environmental changes \cite{Nelson-2001}. Fuel moisture influences
+\citep{Kreye+Varner+etal-2018}. In particular, fine dead fuels respond very
+readily to environmental changes \citep{Nelson-2001}. Fuel moisture influences
 flammability and fire behavior through effects on ignition, spread rate, flame
 height, fuel consumption, and heat release, due to the high specific heat
-content of water \cite{Rothermel-1972, Nelson-2011}. Moisture acts as a
+content of water \citep{Rothermel-1972, Nelson-2011}. Moisture acts as a
 heat-sink resulting in less energy available for propagation, and the resulting
 water vapor will both cool the flaming front and dilute flammable gases
-\cite{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
+\citep{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
 strong effects on flammability, but moisture of dead excised litter fuels can
@@
 flammability at any particular moisture content are key parameters that
-influence fuel flammability throughout the fire season \cite{Nelson-2001}.
+influence fuel flammability throughout the fire season \citep{Nelson-2001}.
 
@@
 recognized as not sufficient to describe species-specific effects on moisture
-dynamics \cite{Anderson-1985}. Fuel moisture dynamics of leaf litter are
+dynamics \citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are
 dependent upon forest structure (shading and rain throughfall effects), litter
 physical structure, and chemical composition of the fuel-beds
-\cite{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
+\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
 models describing moisture dynamics of fine fuels are used by fire scientists
-\cite{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
+\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
 usually aim to describe equilibrium conditions and at least include temperature
@@
 often high spatial heterogeneity of surface fires
-\cite{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.
+\citep{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.
 
-% Fine fuel moisture content reduces ignition probability and flame spread rate \cite{Gisborne-1936,   Fons-1946, Anderson+Rothermal-1965}.
+% Fine fuel moisture content reduces ignition probability and flame spread rate \citep{Gisborne-1936,   Fons-1946, Anderson+Rothermal-1965}.
 
@@
 Leaf litter density is driven by leaf size and shape
-\cite{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
+\citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
 These litter particle effects on litter bed aeration, in turn, have effects on
-litter bed flammability \cite{Ganteaume+Marielle+etal-2011,
+litter bed flammability \citep{Ganteaume+Marielle+etal-2011,
   Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
@@
 litter litter traits, or both. Species specific effects on litter flammability
-are recorded for the forests of the southeastern US \cite{Nowacki+Abrams-2008}
+are recorded for the forests of the southeastern US \citep{Nowacki+Abrams-2008}
 where species-specific litter drying rates play a role in flammability
-feedbacks \cite{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
+feedbacks \citep{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
 the western USA, where the fire regime is characterized by surface fires fueled
 by plant litter, overstory tree composition drives variation in local fire
-behavior \cite{Schwilk+Caprio-2011} as a result of differential litter
+behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
 flammability cross tree species and as a result of synergistic interactions in
-multi-species litter mixtures \cite{Magalhaes+Schwilk-2012}. These flammability
+multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These flammability
 feedbacks may help explain historical fire frequency patterns, but current
@@
 absorption through area-mass relationships and we also expect effects of leaf
-morphology, especially leaf cuticle thickness \cite{Van_Wagner-1969}. Drying
+morphology, especially leaf cuticle thickness \citep{Van_Wagner-1969}. Drying
 rates should also be influenced by leaf and litter particle traits:
@@
 affected by leaf litter traits, especially leaf size
-\cite{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves with lower
+\citep{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves with lower
 moisture retainability and should favor drying, thus increasing the probability
@@
 litter fuels where the flammability of a mixture is driven by the most
-flammable species \cite{VanAltena+Logtestjin+etal-2012,
+flammable species \citep{VanAltena+Logtestjin+etal-2012,
   Magalhaes+Schwilk-2012}. It is possible that similar non-additive effects may
@@
 non-additive flammability in moist fuels with one study finding increasing
-non-additivity with moisture \cite{Blauw+Wensink+etal-2015}.
+non-additivity with moisture \citep{Blauw+Wensink+etal-2015}.
 
@@
 in monocultures and in mixtures for oven-dry litter
-\cite{Magalhaes+Schwilk-2012} and found consistent positive non-additivity of
+\citep{Magalhaes+Schwilk-2012} and found consistent positive non-additivity of
 flammability where fire behavior of mixtures was driven by the most flammable
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Materials and Methods}
+\section*{Materials and Methods}
 
-\subsection{Site description and species selection}
+\subsection*{Site description and species selection}
 
@@
 
-\subsection{Moisture absorption and desorption}
+\subsection*{Moisture absorption and desorption}
 
@@
 species and was collected as part of a previous study
-\cite{Magalhaes+Schwilk-2012}. Litter bulk density was obtained in this
+\citep{Magalhaes+Schwilk-2012}. Litter bulk density was obtained in this
 experiment for each basket by dividing the dry mass (450\,g) by the estimated
@@
 
-\subsection{Flammability assessment}
+\subsection*{Flammability assessment}
 
@@
 gradient, we supplemented these data with data from an earlier experiment on
-oven-dried litter \cite{Magalhaes+Schwilk-2012}. This supplemental data added
+oven-dried litter \citep{Magalhaes+Schwilk-2012}. This supplemental data added
 five replicates per species at an oven dried moisture content of $<$ 5\%. We
@@
 150\,g per species. This design allows fire to reach constant flame spread
-rates \cite{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced
+rates \citep{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced
 along the apparatus allowed visual determination of maximum flame height.
@@
 flaming combustion. The flammability trials followed the methodology detailed
-in de Magalhães and Schwilk \cite{Magalhaes+Schwilk-2012}.
+in de Magalhães and Schwilk \citep{Magalhaes+Schwilk-2012}.
 
@@
 days of burning trials, therefore we have omitted those data from this
-analysis. Our past work \cite{Magalhaes+Schwilk-2012} demonstrated that flame
+analysis. Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that flame
 spread rate captures one important axis of flammability variation
@@
 
-\subsection{Statistical analyses}
+\subsection*{Statistical analyses}
 
@@
 ($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R
-\cite{RCoreTeam-2019}. We fit models with R packages `lme4'
-\cite{Bates_Machler_etal-2015} and `afex' \cite{Singmann_Bolker_etal-2017}.
+\citep{RCoreTeam-2019}. We fit models with R packages `lme4'
+\citep{Bates_Machler_etal-2015} and `afex' \citep{Singmann_Bolker_etal-2017}.
 Approximate degrees of freedom and p-values were calculated by the
-Kenward-Roger approximation \cite{Kenward_Roger-1997} which is recommended by
-Luke \cite{Luke-2017} as producing acceptable Type 1 error rates. We tested if
+Kenward-Roger approximation \citep{Kenward_Roger-1997} which is recommended by
+Luke \citep{Luke-2017} as producing acceptable Type 1 error rates. We tested if
 one leaf trait (specific leaf area) and one litter trait (litter bulk density)
@@
 
-Our past work \cite{Magalhaes+Schwilk-2012} demonstrated that dry litter
+Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that dry litter
 flammability across these eight species was driven by leaf traits and litter
@@
 linear models with a beta regression link with the `glmmADMB` package
-\cite{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function
+\citep{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function
 moisture content and of time since wetting.
@@
 according to estimated marginal means by species using the R package `emmeans'
-\cite{Lenth-2019}. This resulted in four groups of species and we then selected
+\citep{Lenth-2019}. This resulted in four groups of species and we then selected
 one representative species from each litter type group and produced all four
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Results}
+\section*{Results}
 
-\subsection{Species specific moisture dynamics and leaf traits}
+\subsection*{Species specific moisture dynamics and leaf traits}
 
@@
 
-\begin{figure}[H]
+\begin{figure}[h]
   \centering
@@
 
-\begin{table}[H]
+\begin{table}[h]
   \caption{Linear mixed model results for moisture content as a function of
@@
     statistics and p-values were calculated by the Kenward-Roger approximation.
-    \cite{Kenward_Roger-1997}.}
+    \citep{Kenward_Roger-1997}.}
   \label{tab:drydown}
@@
 
-\begin{table}[H]
+\begin{table}[h]
   \caption{Linear model results for maximum water retention (\% moisture
@@
 
-\subsection{Moisture effects on flammability}
+\subsection*{Moisture effects on flammability}
 
@@
 %% spread rate by moisture
-\begin{figure}[H]
+\begin{figure}[h]
   \centering
@@
 %% Table 3 (formerly 3 and 5)
-\begin{table}[H]
+\begin{table}[h]
   \caption{Generalized linear mixed model results for two models predicting
@@
     pseudo F statistics and p-values were calculated by the Kenward-Roger
-    approximation \cite{Kenward_Roger-1997}. Estimated coefficients are in SI
+    approximation \citep{Kenward_Roger-1997}. Estimated coefficients are in SI
     Tables 2 and 4}
@@
 %% fuel consumption by moisture:
-\begin{figure}[H]
+\begin{figure}[h]
   \centering
@@
 
-\begin{table}[H]
+\begin{table}[h]
   \caption{Generalized linear mixed model results for two sets of models
@@
 
-\subsection{Non-additivity in  litter mixtures}
+\subsection*{Non-additivity in  litter mixtures}
 
@@
     degrees of freedom (df), pseudo F statistics and p-values were calculated
-    by the Kenward-Roger approximation \cite{Kenward_Roger-1997}.}
+    by the Kenward-Roger approximation \citep{Kenward_Roger-1997}.}
   \label{tab:mixtures_drydown}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Discussion}
+\section*{Discussion}
 
-\subsection{Species specific moisture dynamics flammability}
+\subsection*{Species-specific moisture dynamics flammability}
 
@@
 importance of fuel moisture in modulating flammability. One shortcoming of past
-work in this system \cite{Magalhaes+Schwilk-2012} was that flammability trials
+work in this system \citep{Magalhaes+Schwilk-2012} was that flammability trials
 tested only very dry fuels. We therefore expected that rankings of species
@@
 one presumption of ours prior to this experiment was that the high flammability
-previously reported for \emph{Q. kelloggii} \cite{Magalhaes+Schwilk-2012} would
+previously reported for \emph{Q. kelloggii} \citep{Magalhaes+Schwilk-2012} would
 disappear under higher fuel moisture contents. This prediction was partly
@@
 turn, controls the major axis of flammability variation in dry litter
-\cite{Magalhaes+Schwilk-2012}. Our results suggest that moisture dynamics are
+\citep{Magalhaes+Schwilk-2012}. Our results suggest that moisture dynamics are
 controlled by both traits intrinsic to the leaf particles (specific leaf area)
@@
 but we expect that decomposition will affect moisture dynamics. For example,
-Van Wagner \cite{Van_Wagner-1969}, investigating fuel weathering, found that
+Van Wagner \citep{Van_Wagner-1969}, investigating fuel weathering, found that
 pine needles had a variation in moisture content 3 to 6 times greater when they
@@
 
-\subsection{Effects of moisture and flammability in litter mixtures}
+\subsection*{Effects of moisture and flammability in litter mixtures}
 
@@
 because past work has shown that mixtures tend to pack less densely than do
-litter monocultures \cite[][, Table 2]{Magalhaes+Schwilk-2012}. If packing
+litter monocultures \citep[][, Table 2]{Magalhaes+Schwilk-2012}. If packing
 density controls rate of drying, then we would expect mixtures to dry more
@@
 spread rates approached three times the expectation
-\cite{Magalhaes+Schwilk-2012}.
+\citep{Magalhaes+Schwilk-2012}.
 
@@
 such as the long-needled pines and the large-leaved oak, are favored by the
-increased fire to which their fuels contribute \cite{Schwilk+Caprio-2011} and
+increased fire to which their fuels contribute \citep{Schwilk+Caprio-2011} and
 similar results are reported for southeastern US forests where species
@@
 rare and that can influence community assembly and the evolutionary of
-niche-constructing traits \cite{Kerr+Schwilk+etal-1999, Schwilk+Kerr-2002}.
+niche-constructing traits \citep{Kerr+Schwilk+etal-1999, Schwilk+Kerr-2002}.
 Ongoing climate change and concurrent changes to fire regimes are linked
@@
 from stand-level species composition shifts to the earth system
-\cite{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}. The patterns we
+\citep{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}. The patterns we
 demonstrate here show to potential importance of species traits and
@@
 
-\vspace{6pt}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontributions{conceptualization, RMQM and DWS; methodology,  RMQM and DWS; experiments, RMQM. coding and data analysis, DWS and RMQM; writing DWS and RMQM; writing--review and editing, DWS}
+\section*{Acknowledgments}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\funding{This research received no external funding.}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\acknowledgments{The authors thank the National Park Service and the employees
+The authors thank the National Park Service and the employees
   of Sequoia and Kings Canyon National Parks.This work was carried out with
@@
   Lizz Waring, Taylor Brown and Chelsea Griffin. We thank the Lubbock Fire
-  Department for their ongoing help and cooperation.}
+  Department for their ongoing help and cooperation.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\conflictsofinterest{The authors declare no conflict of interest.
+This research received no external funding.
+
+\section*{Author contributions}
+  conceptualization, RMQM and DWS; methodology, RMQM and
+  DWS; experiments, RMQM. coding and data analysis, DWS and RMQM; writing DWS
+  and RMQM; writing--review and editing, DWS
+
+\section*{Data and code accessibility}
+All data and code for this project is at \url{https://github.com/schwilklab/trait-flam}
+
+\newpage
+\bibliography{/home/schwilk/write/bib/schwilk}
+
+% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% \conflictsofinterest{The authors declare no conflict of interest.
 The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, or in the decision to publish the results.}
@@
 % =====================================
-\externalbibliography{yes}
-\bibliography{/home/schwilk/write/bib/schwilk}
 

### 2019-07-26 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Move figures to the SI, make jecol.sty

@@
 \documentclass[letterpaper,12pt]{article}
-\usepackage{jae/jae}
-\usepackage{booktabs}
+\usepackage{jecol}
 
@@
 
-\nwords{X}
+\nwords{7600}
 \ntables{X}
@@
 and heavier branches tend to fall closer to the tree trunk. All litter was air
-dried to < 5\% fuel moisture by dry weight for weighing and dividing into
+dried to $< 5$\% fuel moisture by dry weight for weighing and dividing into
 moisture trials. The moisture manipulations and subsequent burning trials were
@@
 likely to influence drying rate, litter bulk density. Specific leaf area was
-taken from leaf fresh leaf area and leaf dry mass from 5 individuals per
-species and was collected as part of a previous study
-\citep{Magalhaes+Schwilk-2012}. Litter bulk density was obtained in this
+taken from leaf freshly fallen litter leaf area and leaf dry mass from 5 individuals per
+species.
+This measurement included petiole and for the scale-leaved species included the full excised branchlet, so this is not directly comparable to typical specific leaf area measurements on fresh leaves.
+Litter bulk density was obtained in this
 experiment for each basket by dividing the dry mass (450\,g) by the estimated
@@
   hour & 1, 183.73 & 1599.59 & $<$.0001 \\ 
-  spcode & 7, 89.73 & 36.15 & $<$.0001 \\ 
-  hour:spcode & 7, 184.15 & 23.06 & $<$.0001 \\ 
+  species & 7, 89.73 & 36.15 & $<$.0001 \\ 
+  hour:species & 7, 184.15 & 23.06 & $<$.0001 \\ 
    \bottomrule
@@
 Across these eight species specific leaf area had a positive effect on maximum
-water retention (Fig. \ref{fig:maxmc-di}, linear model $P$ = 0.003) and litter
-bulk density had a negative effect on desiccation rate (Fig. \ref{fig:bd-di},
+water retention (SI Fig 4, linear model $P$ = 0.003) and litter
+bulk density had a negative effect on desiccation rate (SI Fig 5,
 linear model $P$ = 0.029). Anova results are shown in Tables
 \ref{tab:mc_di_anova}.
-
-\begin{figure}[h]
-  \centering
-\includegraphics[width=16cm]{figs_tables/fig2_SLA_maxMC.pdf}
-\caption{Maximum water retention by specific leaf area across eight species.
-  ANOVA table in Table \ref{tab:mc_di_anova}.}
-  \label{fig:maxmc-di}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-\includegraphics[width=16cm]{figs_tables/fig3_di_bd.pdf}
-\caption{Drying rate rate as a function of specific leaf area and litter bulk
-  density across eight species. ANOVA table in Table \ref{tab:mc_di_anova}}.
-  \label{fig:bd-di}
-\end{figure}
 
@@
     species. ANOVA tables shown with degrees of freedom, sum of squares, mean
-    squares, $F$, and $P$ for each effect.}
+    squares, $F$, and $P$ for each effect. Model coefficients in SI tables 2 and 3}
   \label{tab:mc_di_anova}
@@
   148, $P$ < 0.0001). Model ANOVA table in Table \ref{tab:spreadrate_anova} and
-  coefficients in SI Table 2.}
+  coefficients in SI Table 4.}
   \label{fig:spread_moist}
@@
 
-%% Table 3 (formerly 3 and 5)
+%% Table 3 
 \begin{table}[h]
@@
   \label{tab:spreadrate_anova}
+
 \centering
@@
 
-  \begin{tabular}{ccrrrr}
+  \begin{tabular}{llrrrr}
 \toprule
@@
 \midrule
-  Effect of moisture &  consum \~{} 1 & 262.12 &  &  &  \\ 
-   & consum \~{} taxon & 268.73 & 3 & 13.23 & 0.0042 \\ 
-   & consum \~{} taxon + mc & 289.57 & 1 & 41.68 & < 0.0001 \\ 
-   & consum \~{} taxon + mc + taxon:mc & 293.87 & 3 & 8.60 & 0.0351 \\ 
+  Effect of moisture &   \~{} 1 & 262.12 &  &  &  \\ 
+   &  \~{} taxon & 268.73 & 3 & 13.23 & 0.0042 \\ 
+   & \~{} taxon + mc & 289.57 & 1 & 41.68 & $<$ 0.0001 \\ 
+   & \~{} taxon + mc + taxon:mc & 293.87 & 3 & 8.60 & 0.0351 \\ 
 \midrule
 Effect of time &  consum \~{} 1 & 140.08 &  &  &  \\ 
- &  consum \~{} taxon & 155.84 & 4.00 & 31.51 & < 0.001 \\ 
- & consum \~{} taxon + time & 209.84 & 1.00 & 108.01 & <0.001 \\ 
- & consum \~{} hour + taxon + taxon:time & 216.96 & 4.00 & 14.23 & 0.01 \\ 
+ & \~{} taxon & 155.84 & 4.00 & 31.51 & $<$ 0.001 \\ 
+ & \~{} taxon + time & 209.84 & 1.00 & 108.01 & $<$ 0.001 \\ 
+ & \~{} hour + taxon + taxon:time & 216.96 & 4.00 & 14.23 & 0.01 \\ 
     \bottomrule
@@
 
-\begin{figure}[h]
-  \centering
-\includegraphics[width=8cm]{figs_tables/fig8_mixture_drydown-curves.pdf}
-\caption{Fuel moisture by time since saturation for four litter mixtures.
-Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.
-Moistures were measured every 24 hours but data are shown with slight horizontal jitter to reduce point overlap.
-Mixed linear model results shown in Table \ref{tab:mixtures_drydown}.}
-  \label{fig:mixture_dry_down}
-\end{figure}
-
 Mixture dry down rates and maximum water retention showed some relationship to
-the dry down curves for individual constituent species (Fig.
-\ref{fig:mixture_dry_down}. For example, the mixture that does not contain
-\emph{Q. kelloggii} (AbCaPi in Fig. Fig. \ref{fig:mixture_dry_down}) has the
-lowest initial retention. Initial water retention and rates differed across
-mixtures (linear model based on log of moisture content, Table
-\ref{tab:mixtures_drydown}).
-
-\begin{table}[h]
-  \caption{Linear mixed model results for moisture content as a function of
-    time since wetting for four different litter mixture types. Approximate
-    degrees of freedom (df), pseudo F statistics and p-values were calculated
-    by the Kenward-Roger approximation \citep{Kenward_Roger-1997}.}
-  \label{tab:mixtures_drydown}
-\centering
-
-\begin{tabular}{llll}
-  \toprule
-Effect & df & $F$ & $P$ \\ 
-  \midrule
-time & 1, 96.19 & 565.28 & $<$.0001 \\ 
-  mixture & 3, 96.08 & 23.63 & $<$.0001 \\ 
-  time:mixture & 3, 95.71 & 8.70 & $<$.0001 \\ 
-   \bottomrule
-\end{tabular}
-\end{table}
+the dry down curves for individual constituent species (see SI Fig. 6). For
+example, the mixture that does not contain \emph{Q. kelloggii} (AbCaPi in SI
+Fig. 6) has the lowest initial retention. Initial water retention and rates
+differed across mixtures (linear model based on log of moisture content, SI Table 8)
 

### 2019-07-26 ms/SI.tex
Move figures to the SI, make jecol.sty

@@
 \usepackage{fullpage}
+\usepackage{float}
 
@@
 \author{Rita Quinones de Magalhaes and Dylan W. Schwilk}
-%\date{\today}
+\date{}
 
@@
 
-\begin{table}[h]
+\begin{table}[H]
   \caption{Mixed model coefficients for drydown curves. Table shows linear model results of model fit to  natural log of dry-mass based fuel moisture content as a function of time and species.} 
@@
 
-\begin{table}[h]
-  \caption{Mixed model coefficients for flame spread rate. Table shows mixed linear model results.} 
+\begin{table}[H]
+  \caption{ANOVA model coefficients for moisture content as a function of species traits} 
+  \label{tabS2}
+\centering
+\input{figs_tables/SI_tab2_mc_coef.ltx}
+\end{table}
+
+\begin{table}[H]
+  \caption{ANOVA model coefficients for drying rate as a function of species traits} 
   \label{tabS3}
 \centering
-\input{figs_tables/SI_tab2_spread_moist_coef.ltx}
+\input{figs_tables/SI_tab3_di_coef.ltx}
+\end{table}
+
+\begin{table}[H]
+  \caption{Mixed model coefficients for flame spread rate as a function of taxon and moisture content. Table shows mixed linear model results.} 
+  \label{tabS4}
+\centering
+\input{figs_tables/SI_tab4_spread_moist_coef.ltx}
+\end{table}
+
+\begin{table}[H]
+  \caption{Mixed model coefficients for fuel consumption as a function of taxon and moisture content. Table shows mixed linear model results.} 
+  \label{tabS5}
+\centering
+\input{figs_tables/SI_tab5_consume_moist_coef.ltx}
+\end{table}
+
+\begin{table}[H]
+  \caption{Mixed model coefficients for flame spread rate as a function of taxon and time since wetting. Table shows mixed linear model results.} 
+  \label{tabS6}
+\centering
+\input{figs_tables/SI_tab6_spread_time_coef.ltx}
+\end{table}
+
+\begin{table}[H]
+  \caption{Mixed model coefficients for fuel consumption as a function of taxon and time since wetting. Table shows mixed linear model results.} 
+  \label{tabS6}
+\centering
+\input{figs_tables/SI_tab7_consume_time_coef.ltx}
+\end{table}
+
+% table 7
+\begin{table}[H]
+  \caption{Linear mixed model results for moisture content as a function of
+    time since wetting for four different litter mixture types. Approximate
+    degrees of freedom (df), pseudo F statistics and p-values were calculated
+    by the Kenward-Roger approximation}.
+  \label{tab:mixtures_drydown}
+\centering
+
+\begin{tabular}{llll}
+  \toprule
+Effect & df & $F$ & $P$ \\ 
+  \midrule
+time & 1, 96.19 & 565.28 & $<$0.0001 \\ 
+  mixture & 3, 96.08 & 23.63 & $<$0.0001 \\ 
+  time:mixture & 3, 95.71 & 8.70 & $<$0.0001 \\ 
+   \bottomrule
+\end{tabular}
 \end{table}
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
-  \label{fig-S2}
+  \label{SI_fig3}
 \includegraphics[width=17cm]{figs_tables/SI_fig3_dessication_emmeans.pdf}
@@
 
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/SI_fig4_mixture_drydown-curves.pdf}
+\caption{Fuel moisture by time since saturation for four litter mixtures.
+Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.
+Moistures were measured every 24 hours but data are shown with slight horizontal jitter to reduce point overlap.
+Mixed linear model results shown in SI Table 7.}
+  \label{SI_fig4}
+\end{figure}
+
+% SI fig 5
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/fig2_SLA_maxMC.pdf}
+\caption{Maximum water retention by specific leaf area across eight species.
+  % ANOVA table in Table \ref{tab:mc_di_anova}.
+}
+  \label{fig:maxmc-di}
+\end{figure}
+
+% SI fig 6
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/fig3_di_bd.pdf}
+\caption{Drying rate rate as a function of specific leaf area and litter bulk
+  density across eight species.
+  % ANOVA table in Table \ref{tab:mc_di_anova}
+}.
+  \label{fig:bd-di}
+\end{figure}
+
 \end{document}

### 2019-07-26 scripts/moisture-ms-figs-tables.R

@@ -94,7 +94,7 @@ ggsave(file.path(RESULTS, "SI_fig2_max_water_emmeans.pdf"), plot=SI_fig2,
 ###############################################################################
 dry.trends <- emtrends(dry.mod, "display.name", var="hour")
 SI_fig3 <- plot(dry.trends) + #, comparisons=TRUE) +
-  xlab(expression(paste("Estimated marginal mean dessication rate (", hr^-1, ")"))) +
+  xlab(expression(paste("Estimated marginal mean drying rate (", hr^-1, ")"))) +
   ylab("Species") +
   pubtheme.nogridlines +
   theme(axis.text.y = element_text(face = "italic"))
@@ -109,7 +109,7 @@ ggsave(file.path(RESULTS, "SI_fig3_dessication_emmeans.pdf"), plot=SI_fig3,
 
 # 2011 trait data:
 oldtraits <- read.csv("../data/moisture/traits.csv", stringsAsFactors=FALSE)
-
+oldtraits$SLA <- oldtraits$SLA/1000
 #mtrait <- left_join(newmctr, species)
 mtrait <- left_join(mc.sum, flam.sp.avg)
 mtrait <- left_join(mtrait, oldtraits)
@@ -124,35 +124,42 @@ fig2 <- ggplot(mtrait, aes(SLA, maxMC)) +
 #  geom_errorbar(aes(ymin=maxMC+maxMC.se, ymax=maxMC-maxMC.se))+
 #  geom_errorbarh(aes(xmin=leaf.area.mean+(leaf.area.sd/sqrt(6)), xmax=leaf.area.mean-(leaf.area.sd/sqrt(6))))+
   geom_label_repel(aes(label=display.name),
-                  nudge_x=0.02,
-                  nudge_y=-0.05,
-                  fontface="italic") +
+                   hjust= 0,
+                  # direction="y",
+                   nudge_x= 2,
+                   nudge_y=100,
+                  fontface="italic",
+                  size=smsize-6,
+                  max.iter=5000) +
 #  scale_y_log10() + 
 #  scale_colour_manual(values=schwilkcolors) +
   xlab(expression(paste("Specific leaf area (", cm^3/g, ")" ))) +
-  ylab("Maximum moisture retention by dry weight ln(%)") +
+  ylab("Maximum moisture content (%)") +
   pubtheme.nogridlines
+
+fig2
[…]

### 2019-07-26 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Draft sent to Rita for approval

@@
 \nwords{7600}
-\ntables{X}
-\nfig{X}
-\nref{X}
+\ntables{4}
+\nfig{8}
+\nref{39}
 
@@
 
-  \item Moisture content is a strong determinant of forest fuel flammability
-  and varies on small spatial scales and over short time periods. Empirical
-  modeling of equilibrium dead fine fuel moisture is well established, but less
-  is known about the mechanisms by which species-specific litter traits
-  influence litter moisture dynamics, particularly the processes of moisture
-  absorption and desorption. Characterizing how species’ leaf litter retain and
-  release moisture through time is critical to the assessment of fuel’s
-  availability to burn.
-  \item Natural forests are often comprised of
-  mixed stands, leading to mixtures of leaf litter on the surface fuelling the
-  fires. Multiple studies have established the existence of non-additive
-  effects in the flammability of dry litter, but we lack an understanding of
-  moisture synergies in mixed litter and of moisture effects on the
-  flammability of mixed litter.
-  \item The litter from eight species of a
-  mixed-conifer forest was saturated and allowed to dry to determine moisture
-  absorption capacity and desorption rates. Burn trials were performed in moist
-  litter beds of single species and of mixtures to establish flammability
-  response to the dry-down process.
-  \item Moisture dynamics varied across species with
-  higher specific leaf area being associated with higher maximum absorption and
-  lower litter bulk density being associated with faster drying rates. Species
-  that produce more aerated litter beds (such as \emph{Q. kelloggii} and
-  \emph{Pinus} species) have faster rates of litter drying.
-  \item We examined two
-  axes of flammability variation (spread rate and fuel consumption) and found
-  that these differing moisture dynamics result in time since wetting having
-  strongly differing effects on flammability across the major litter types
-  typified by different tree genera. We found that litter mixtures
-  exhibited consistent non-additivity in flame spread rate in fuel consumption.
-  This non-additivity became more positive with fuel drying.
-    \item \emph{Synthesis:} Litter mixtures burn with fire
-  behavior more similar to that of the most flammable constituent species and
-  this effect increases as fuels dry. statement. Such positive feedback effects
-  increases the potential for species to exert community scale effects even
-  when relatively rare and that can influence community assembly and the
-  evolutionary of niche-constructing traits.
+      \item Moisture content is a strong determinant of forest fuel
+    flammability and varies on small spatial scales and over short time
+    periods. Empirical modeling of equilibrium dead fine fuel moisture is well
+    established, but less is known about the mechanisms by which
+    species-specific litter traits influence litter moisture dynamics.
+    Characterizing how species’ leaf litter retain and release moisture through
+    time is critical to the assessment of fuel’s availability to burn.
+      \item Natural forests are often comprised of mixed stands, leading to
+    mixtures of leaf litter on the surface fuelling the fires. Multiple studies
+    have established the existence of non-additive effects in the flammability
+    of dry litter, but we lack an understanding of moisture synergies in mixed
+    litter and of moisture effects on the flammability of mixed litter.
+      \item The litter from eight species of a mixed-conifer forest was
+    saturated and allowed to dry to determine moisture absorption capacity and
+    drying rates. Burn trials were performed in moist litter beds of single
+    species and of mixtures to establish flammability response to the dry-down
+    process.
+      \item Moisture dynamics varied across species with higher specific leaf
+    area being associated with higher maximum absorption and lower litter bulk
+    density being associated with faster drying rates. Species that produce
+    more aerated litter beds (such as \emph{Q. kelloggii} and \emph{Pinus}
+    species) have faster rates of litter drying.
+      \item We examined two axes of flammability variation (spread rate and
+    fuel consumption) and found that these differing moisture dynamics result
+    in time since wetting having strongly differing effects on flammability
+    across the major litter types typified by different tree taxa.
+      \item We found that litter mixtures exhibited consistent non-additivity
+    in flame spread rate in fuel consumption. This non-additivity became more
+    positive with fuel drying. Litter mixtures burn with fire behavior more
+    similar to that of the most flammable constituent species and this effect
+    increases as fuels dry.
+      \item \emph{Synthesis:} In this temperate mixed conifer forest, the shade
+    intolerant species are favored by the increased fire to which their fuels
+    contribute Such positive feedback effects increases the potential for
+    species to exert community scale effects even when relatively rare and that
+    can influence community assembly and the evolutionary of niche-constructing
+    traits.
   \end{enumerate}
@@
 height, fuel consumption, and heat release, due to the high specific heat
-content of water \citep{Rothermel-1972, Nelson-2011}. Moisture acts as a
+content of water \citep{Rothermel-1972, Nelson-2001}. Moisture acts as a
 heat-sink resulting in less energy available for propagation, and the resulting
@@
   Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
-are well established, however, how moisture dynamics modulate species specific
+are well established. However, how moisture dynamics modulate species specific
 flammability is poorly known, and could be driven by by leaf traits, emergent
@@
 behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
-flammability cross tree species and as a result of synergistic interactions in
+flammability across tree species and as a result of synergistic interactions in
 multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These flammability
@@
 flammability of multi-species mixtures is not the average of the constituent
-species. Such non-additive effects have been shown to be synergistic in dry
-litter fuels where the flammability of a mixture is driven by the most
-flammable species \citep{VanAltena+Logtestjin+etal-2012,
-  Magalhaes+Schwilk-2012}. It is possible that similar non-additive effects may
-influence litter moisture content, especially if different leaf traits
-influence absorptive capacity and drying rate. There is some evidence for
-non-additive flammability in moist fuels with one study finding increasing
-non-additivity with moisture \citep{Blauw+Wensink+etal-2015}.
+species. Such non-additive effects are synergistic in dry litter fuels where
+the flammability of a mixture is driven by the most flammable species
+\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. It is possible
+that similar non-additive effects may influence litter moisture content,
+especially if different leaf traits influence absorptive capacity and drying
+rate. There is some evidence for non-additive flammability in moist fuels with
+one study finding increasing non-additivity with moisture
+\citep{Blauw+Wensink+etal-2015}.
 
@@
 flammability where fire behavior of mixtures was driven by the most flammable
-species. Here, we examined if this positive non additivity was maintained in
+species. Here, we examined if this positive non-additivity was maintained in
 mixtures across varying levels of fuel moisture. The burn trials were also
@@
 reported in previous studies occur under ecologically-relevant moisture levels
-and whether fuel moisture effects strengthen or mask such non-additivity.
+and whether fuel moisture effects strengthen or moderate such non-additivity.
 
@@
 previous year (mostly undecomposed). Leaf litter was collected in the summer
-(mid-June to mid-July) in 2011 and in 2012 from 21 separate sites across the
+(mid-June to mid-July) in 2012 from 21 separate sites across the
 study area. We collected from a minimum of four populations per species with
@@
 (Maxim Integrated) placed over the samples. Litter baskets were arranged in a
-completely randomized design. During the dry-down process three subsamples were
-taken at 24 hour intervals and weighed. These subsamples were then oven dried
-for 24 hours at 100\,C and re-weighed to assess moisture content. This was
-repeated until the final sample from a basket was less than 10\% moisture.
+completely randomized design. During the dry-down process three subsamples of
+5--10\,g were taken at 24 hour intervals and weighed. These subsamples were
+then oven dried for 24 hours at 100\,C and re-weighed to assess moisture
+content. This was repeated until the final sample from a basket was less than
+10\% moisture.
 
@@
 likely to influence drying rate, litter bulk density. Specific leaf area was
-taken from leaf freshly fallen litter leaf area and leaf dry mass from 5 individuals per
-species.
-This measurement included petiole and for the scale-leaved species included the full excised branchlet, so this is not directly comparable to typical specific leaf area measurements on fresh leaves.
-Litter bulk density was obtained in this
+measured on freshly fallen litter by sampling 5 individuals per species. This
+measurement included petiole and for the scale-leaved species included the full
+excised branchlet. Therefore, this is not directly comparable to typical specific leaf
+area measurements on fresh leaves. Litter bulk density was obtained in this
 experiment for each basket by dividing the dry mass (450\,g) by the estimated
@@
 We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
-preliminary trials on damp fuels suggested that across these fuels ignition is
-unlikely above 100\% fuel moisture. To expand our data on the dry end of this
-gradient, we supplemented these data with data from an earlier experiment on
-oven-dried litter \citep{Magalhaes+Schwilk-2012}. This supplemental data added
-five replicates per species at an oven dried moisture content of $<$ 5\%. We
-assigned these replicates a time since wetting value of 144 hours equivalent to
-the longest time since drying for the wetted samples. Flammability was assessed
-using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which leaf
-litter was placed and the table gently shaken to allow settling. Burning trials
-all used 450\,g (dry weight) of litter. For three-species mixtures, this was
-150\,g per species. This design allows fire to reach constant flame spread
+preliminary trials on damp fuels suggested that across these fuels ignition
+will not occur above 100\% fuel moisture. To expand our data on the dry end of
+this gradient, we supplemented these data with data from an earlier experiment
+on oven-dried litter \citep{Magalhaes+Schwilk-2012}. This supplemental data
+added five replicates per species at an oven dried moisture content of $<$ 5\%.
+We assigned these replicates a time since wetting value of 144 hours equivalent
+to the longest time since drying for the wetted samples. Flammability was
+assessed using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which
+leaf litter was placed and the table gently shaken to allow settling. Burning
+trials all used 450\,g (dry weight) of litter. For three-species mixtures, this
+was 150\,g per species. This design allows fire to reach constant flame spread
 rates \citep{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced
@@
 Kenward-Roger approximation \citep{Kenward_Roger-1997} which is recommended by
-Luke \citep{Luke-2017} as producing acceptable Type 1 error rates. We tested if
+Luke \citeyear{Luke-2017} as producing acceptable Type 1 error rates. We tested if
 one leaf trait (specific leaf area) and one litter trait (litter bulk density)
@@
 genera in the Cupressaceae, \emph{Calocedrus} and \emph{Sequoiadendron}, also
-behaved similarly to one another. Therefore, to preserve power used taxonomic
+behaved similarly to one another. Therefore, to preserve power we used taxonomic
 groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our
@@
 linear models with a beta regression link with the `glmmADMB` package
-\citep{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function
+\citep{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function of
 moisture content and of time since wetting.
@@
 the average of the constituent species. To choose mixtures, we first grouped
-the species according to their maximum water retainability and dessication rate
+the species according to their maximum water retention and drying rate
 according to estimated marginal means by species using the R package `emmeans'
@@
   concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa} (SI Fig. 1). There are
-also differences in desiccation rates across the eight species and estimated
+also differences in drying rates across the eight species and estimated
 mixed linear model slopes discriminated species into three groups from lowest
-to most rapid: 1) \emph{C.decurrens}, 2) \emph{Abies spp} and \emph{S.
+to most rapid: 1) \emph{C. decurrens}, 2) \emph{Abies spp} and \emph{S.
   giganteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp} (SI Fig. 2).
@@
   because, although collections were staggered, there were sometimes multiple
-  trays collected at a single time point.}
+  samples collected at a single time point.}
  \label{fig:drydown}
@@
   \caption{Linear mixed model results for moisture content as a function of
-    time since wetting. Approximate degrees of freedom (df), pseudo F
+    time since wetting and species. Approximate degrees of freedom (df), pseudo F
     statistics and p-values were calculated by the Kenward-Roger approximation.
-    \citep{Kenward_Roger-1997}.}
+    \citep{Kenward_Roger-1997}. Model coefficients reported in SI Table 1.}
   \label{tab:drydown}
@@
 bulk density had a negative effect on desiccation rate (SI Fig 5,
-linear model $P$ = 0.029). Anova results are shown in Tables
+linear model $P$ = 0.029). ANOVA results are shown in Tables
 \ref{tab:mc_di_anova}.
@@
 
-Desiccation rate & density & 1 & $3.92 \times 10^{-4}$ & $3.92 \times 10^{-4}$ & 11.01 & 0.029 \\ 
+Drying rate & density & 1 & $3.92 \times 10^{-4}$ & $3.92 \times 10^{-4}$ & 11.01 & 0.029 \\ 
  &  SLA & 1 & $1.88 \times 10^{-5}$ & $1.88 \times 10^{-5}$ & 0.53 & 0.507 \\ 
@@
 and the rate of change with moisture content varied across the taxonomic groups
-(Fig. \ref{fig:consume_moist}, Table \ref{tab:consume_moist_anova}). Although
+(Fig. \ref{fig:consume_moist}, Table \ref{tab:consume_anova}). Although
 the overall ranking of groups was similar for fuel consumption and for flame
@@
 generalized linear mixed beta regression model as indicated by the nested model
-comparisons in Table \ref{tab:consume_time_anova}. As with flame spread rate
+comparisons in Table \ref{tab:consume_anova}. As with flame spread rate
 some pines had the most rapid response to drying. However, unlike in the other
@@
 water during wetting, and the one mixture without any Cupressaceae had the most
-rapid drying rate (Figure \ref{fig:mixture_dry_down} and Table
-\ref{tab:mixtures_drydown}). Drydown rates, however, did not behave as the
+rapid drying rate (SI Figure 4 and Si Table 8). Drydown rates, however, did not behave as the
 average of the constituent species: mixtures tended to be drier than expected
@@
   conceptualization, RMQM and DWS; methodology, RMQM and
-  DWS; experiments, RMQM. coding and data analysis, DWS and RMQM; writing DWS
+  DWS; experiments, RMQM; coding and data analysis, DWS and RMQM; writing DWS
   and RMQM; writing--review and editing, DWS

### 2019-08-05 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Include Rita's comments 2019-08-05

@@
     species to exert community scale effects even when relatively rare and that
-    can influence community assembly and the evolutionary of niche-constructing
+    can influence community assembly and the evolution of niche-constructing
     traits.
@@
 flammability of dry litter fuels in seasonally dry climates because higher
-litter density reduces flammability, especially on the flame spread rate axis.
-Leaf litter density is driven by leaf size and shape
-\citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
-These litter particle effects on litter bed aeration, in turn, have effects on
-litter bed flammability \citep{Ganteaume+Marielle+etal-2011,
-  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
-are well established. However, how moisture dynamics modulate species specific
-flammability is poorly known, and could be driven by by leaf traits, emergent
-litter litter traits, or both. Species specific effects on litter flammability
-are recorded for the forests of the southeastern US \citep{Nowacki+Abrams-2008}
-where species-specific litter drying rates play a role in flammability
-feedbacks \citep{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
-the western USA, where the fire regime is characterized by surface fires fueled
-by plant litter, overstory tree composition drives variation in local fire
+litter density reduces flammability. Leaf litter density is driven by leaf size
+and shape \citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006,
+  Kane+Varner+etal-2008}. These litter particle effects on litter bed aeration,
+in turn, have effects on litter bed flammability
+\citep{Ganteaume+Marielle+etal-2011, Schwilk+Caprio-2011,
+  Kreye+Varner+etal-2013}. These species-specific effects are well established.
+However, how moisture dynamics modulate species specific flammability is poorly
+known, and could be driven by by leaf traits, emergent litter litter traits, or
+both. Species specific effects on litter flammability are recorded for the
+forests of the southeastern US \citep{Nowacki+Abrams-2008} where
+species-specific litter drying rates play a role in flammability feedbacks
+\citep{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of the
+western USA, where the fire regime is characterized by surface fires fueled by
+plant litter, overstory tree composition drives variation in local fire
 behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
 flammability across tree species and as a result of synergistic interactions in
-multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These flammability
-feedbacks may help explain historical fire frequency patterns, but current
-estimates of litter flammability in mixed conifer forests are based largely on
-burning dry litter.
+multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
+flammability feedbacks may help explain historical fire frequency patterns, but
+current estimates of litter flammability in mixed conifer forests are based
+largely on burning dry litter.
 
@@
 influenced by the lower bulk densities of litter mixtures containing those two
-species. Most interestingly, there were striking pattern of non-additivity in
+species. Most interestingly, there were striking patterns of non-additivity in
 spread rate and in fuel consumption for the four mixtures, indicating that
@@
 
-The authors thank the National Park Service and the employees
-  of Sequoia and Kings Canyon National Parks.This work was carried out with
-  permission from NPS under permit \# SEKI-2014-SCI-0040. We also thank
-  students who assisted with field work and the litter collection, including
-  Lizz Waring, Taylor Brown and Chelsea Griffin. We thank the Lubbock Fire
+The authors thank the National Park Service and the employees of Sequoia and
+Kings Canyon National Parks. This work was carried out with permission from NPS
+under permit \# SEKI-2014-SCI-0040. The authors thank Russell Lackey who
+contributed essential assistance with the field work and Gabrielle Plata and
+Leticia Plata who assisted with the burning trials. We thank the Lubbock Fire
 Department for their ongoing help and cooperation.

### 2019-08-06 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Correct SI fig and table numbers, add SI figures

@@
 
-% Fine fuel moisture content reduces ignition probability and flame spread rate \citep{Gisborne-1936,   Fons-1946, Anderson+Rothermal-1965}.
-
-Litter packing (measured as litter-bed density or packing ratio) controls the
+Litter packing, measured as litter-bed density or packing ratio, controls the
 flammability of dry litter fuels in seasonally dry climates because higher
-litter density reduces flammability. Leaf litter density is driven by leaf size
-and shape \citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006,
-  Kane+Varner+etal-2008}. These litter particle effects on litter bed aeration,
-in turn, have effects on litter bed flammability
-\citep{Ganteaume+Marielle+etal-2011, Schwilk+Caprio-2011,
-  Kreye+Varner+etal-2013}. These species-specific effects are well established.
-However, how moisture dynamics modulate species specific flammability is poorly
-known, and could be driven by by leaf traits, emergent litter litter traits, or
-both. Species specific effects on litter flammability are recorded for the
-forests of the southeastern US \citep{Nowacki+Abrams-2008} where
-species-specific litter drying rates play a role in flammability feedbacks
-\citep{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of the
-western USA, where the fire regime is characterized by surface fires fueled by
-plant litter, overstory tree composition drives variation in local fire
+litter density reduces flammability as a consequence of oxygen limitation. Leaf
+litter density is driven by leaf size and shape
+\citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
+These litter particle effects on litter bed aeration, in turn, have effects on
+litter bed flammability \citep{Ganteaume+Marielle+etal-2011,
+  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
+are well established. However, how moisture dynamics modulate species specific
+flammability is poorly known and could be driven by by leaf traits, emergent
+litter litter traits, or both. Species specific effects on litter flammability
+are recorded for the forests of the southeastern US \citep{Nowacki+Abrams-2008}
+where species-specific litter drying rates play a role in flammability
+feedbacks \citep{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
+the western USA, where the fire regime is characterized by surface fires fueled
+by plant litter, overstory tree composition drives variation in local fire
 behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
@@
 previous year (mostly undecomposed). Leaf litter was collected in the summer
-(mid-June to mid-July) in 2012 from 21 separate sites across the
-study area. We collected from a minimum of four populations per species with
-each population 9--32\,km distant from its neighbors. An exception to this was
-\emph{A. magnifica} which is restricted to the higher elevations and could only
-be collected from one area, due to designated wilderness boundaries and
-logistical constraints. Within that area, \emph{A. magnifica} litter was
-collected from over 10 different individuals. For the remaining species, the
-collection involved 2--4 individual trees at each site, separated from one
+(mid-June to mid-July) in 2012 from 21 separate sites across the study area. We
+collected from a minimum of four populations per species with each population
+9--32\,km distant from its neighbors. An exception to this was \emph{A.
+  magnifica} which is restricted to the higher elevations and could only be
+collected from one area due to designated wilderness boundaries and logistical
+constraints. Within that area, \emph{A. magnifica} litter was collected from
+over 10 different individuals. For the remaining species, the collection
+involved 2--4 individual trees at each site which were separated from one
 another by at least 10\,m. The litter was collected 2\,m away from the trunk to
@@
 retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
-($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R
+($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R 3.6.1
 \citep{RCoreTeam-2019}. We fit models with R packages `lme4'
@@
 Kenward-Roger approximation \citep{Kenward_Roger-1997} which is recommended by
-Luke \citeyear{Luke-2017} as producing acceptable Type 1 error rates. We tested if
+Luke \citeyear{Luke-2017} as producing acceptable Type I error rates. We tested if
 one leaf trait (specific leaf area) and one litter trait (litter bulk density)
@@
 effects with few intermediate values. Therefore, we used mixed generalized
-linear models with a beta regression link with the `glmmADMB` package
+linear models with a beta regression link with the `glmmADMB' package
 \citep{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function of
@@
 according to estimated marginal means by species using the R package `emmeans'
-\citep{Lenth-2019}. This resulted in four groups of species and we then selected
-one representative species from each litter type group and produced all four
-possible three-species litter mixtures using those four species. For each
-mixture and time since wetting, we predicted moisture content based on the mean
-content of the three constituent species at that time since wetting. We
+\citep{Lenth-2019}. This resulted in four groups of species and we then
+selected one representative species from each litter type group and produced
+all four possible three-species litter mixtures using those four species. For
+each mixture and time since wetting, we predicted moisture content based on the
+mean content of the three constituent species at that time since wetting. We
 predicted flammability parameters based on the average flammability measures
@@
 between observed and predicted values of flammability will be zero. We tested
-if there were significant departures from the null (zero) with a mixed effects
-linear model. For flammability parameters, we fit linear mixed effect models
-with the observed-predicted flammability value as the dependent variable and
-used Wald tests to test if the centered intercept was different than zero (a
-measure of overall non-additivity) and if the slope with time since wetting was
-different from zero (indicating non-additivity changing with drying).
+if there were significant departures from the null with a mixed effects linear
+model. For flammability parameters, we fit linear mixed effect models with the
+observed-predicted flammability value as the dependent variable and used Wald
+tests to test if the centered intercept was different than zero (a measure of
+overall non-additivity) and if the slope with time since wetting was different
+from zero (indicating non-additivity changing with drying).
 
@@
 There were marked differences in the extent of moisture species retain
-initially following wetting and draining by gravity (all $P$ < 0.0001, Fig.
+initially following wetting and draining by gravity (all $P < 0.0001$, Fig.
 \ref{fig:drydown} and Table \ref{tab:drydown}). \emph{Q. kelloggii}, the only
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/fig4_spread_actualMC.pdf}
+\includegraphics[width=8cm]{figs_tables/fig2_spread_actualMC.pdf}
 \caption{Flame spread rate by moisture content (model fit $F$= 61.15, df = 9,
@@
     approximation \citep{Kenward_Roger-1997}. Estimated coefficients are in SI
-    Tables 2 and 4}
+    Tables 4 and 6}
   \label{tab:spreadrate_anova}
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/fig5_consume_actualMC.pdf}
+\includegraphics[width=8cm]{figs_tables/fig3_consume_actualMC.pdf}
 \caption{Percent fuel consumed by moisture content. Lines indicate best fit
   beta regression lines. Model deviance table in Table \ref{tab:consume_anova}
-  and coefficients are in SI Table 3.}
+  and coefficients are in SI Table 5.}
   \label{fig:consume_moist}
@@
     moisture content effect, and for the full model including the taxon by
-    moisture or time interaction. Coefficient estimates are in SI Tables 3 and
-    5.}
+    moisture or time interaction. Coefficient estimates are in SI Tables 5 and
+    7.}
   \label{tab:consume_anova}
@@
 As a consequence of the relationships shown above, time since wetting had a
-significant effect on spread rate ($P$ < 0.0001) and the slope of this effect
+significant effect on spread rate ($P < 0.0001$) and the slope of this effect
 varied by taxonomic group (Fig. \ref{fig:spread_time}, Table
@@
 spread rates were slow. \emph{Quercus} had the steepest response to time since
-wetting (Fig. \ref{fig:spread_time}, SI Table 4)).
+wetting (Fig. \ref{fig:spread_time}, SI Table 6)).
 
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/fig6_spread_time.pdf}
+\includegraphics[width=8cm]{figs_tables/fig4_spread_time.pdf}
 \caption{Flame spread rate by time since wetting for litter of eight species.
   Lines indicate best fit linear model fits. Model ANOVA table in Table
-  \ref{tab:spreadrate_anova} and coefficients in SI Table 4. Horizontal jitter
+  \ref{tab:spreadrate_anova} and coefficients in SI Table 6. Horizontal jitter
   (up to 2 hr) was added to the plot for visualization and to avoid
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/fig7_consume_time.pdf}
+\includegraphics[width=8cm]{figs_tables/fig5_consume_time.pdf}
 \caption{Percent fuel consumed by time since wetting for litter of eight
   species. Curves shown indicate best fit beta regression lines. Model ANOVA
-  table in Table \ref{tab:consume_anova} and coefficients in SI Table 4.
+  table in Table \ref{tab:consume_anova} and coefficients in SI Table 7.
   Horizontal jitter (up to 2 hr) was added to the plot for visualization and to
@@
 Fig. 6) has the lowest initial retention. Initial water retention and rates
-differed across mixtures (linear model based on log of moisture content, SI Table 8)
+differed across mixtures (linear model based on log of moisture content, SI Tables 8 and 9).
 
@@
   \centering
-\includegraphics[width=16cm]{figs_tables/fig9_mixture_obs_vs_pred_mc.pdf}
+\includegraphics[width=16cm]{figs_tables/fig6_mixture_obs_vs_pred_mc.pdf}
 \caption{Observed vs predicted fuel moisture content for litter mixtures.
@@
   \centering
-\includegraphics[width=16cm]{figs_tables/fig10_mixture_obs_vs_pred_spread.pdf}
+\includegraphics[width=16cm]{figs_tables/fig7_mixture_obs_vs_pred_spread.pdf}
 \caption{Observed vs. predicted flame spread rate for litter mixtures. Filled
@@
   \centering
-\includegraphics[width=16cm]{figs_tables/fig11_mixture_obs_vs_pred_consume.pdf}
+\includegraphics[width=16cm]{figs_tables/fig8_mixture_obs_vs_pred_consume.pdf}
 \caption{Observed vs predicted fuel consumption for litter mixtures. Filled
@@
 since wetting (Fig. \ref{fig:mixture_obs_pred_consume}, slope not equal to
-zero, $P$ < 0.0001).
+zero, $P < 0.0001$).
 

### 2019-08-06 ms/SI.tex
Correct SI fig and table numbers, add SI figures

@@
   \caption{Mixed model coefficients for fuel consumption as a function of taxon and time since wetting. Table shows mixed linear model results.} 
-  \label{tabS6}
+  \label{tabS7}
 \centering
@@
 
-% table 7
+% table 8
 \begin{table}[H]
@@
   time:mixture & 3, 95.71 & 8.70 & $<$0.0001 \\ 
+   \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[H]
+  \caption{Linear mixed model coefficients for moisture content as a function of
+    time since wetting for four different litter mixture types.}.
+  \label{tab:mixtures_drydown}
+\centering
+\begin{tabular}{rrrrrr}
+  \toprule
+ & Estimate & Std. Error & df & t value & Pr($>$$|$t$|$) \\ 
+  \midrule
+(Intercept) & 4.59 & 0.06 & 71.04 & 77.18 & 0.00 \\ 
+  hour & -0.01 & 0.00 & 95.45 & -13.96 & 0.00 \\ 
+  spcodeAbCaQu & 0.26 & 0.08 & 95.35 & 3.28 & 0.00 \\ 
+  spcodeAbPiQu & 0.54 & 0.08 & 96.53 & 7.04 & 0.00 \\ 
+  spcodeCaPiQu & 0.57 & 0.08 & 95.35 & 7.25 & 0.00 \\ 
+  hour:spcodeAbCaQu & 0.00 & 0.00 & 95.40 & 3.67 & 0.00 \\ 
+  hour:spcodeAbPiQu & -0.00 & 0.00 & 95.78 & -0.70 & 0.49 \\ 
+  hour:spcodeCaPiQu & 0.00 & 0.00 & 95.40 & 2.64 & 0.01 \\ 
    \bottomrule
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/SI_fig4_mixture_drydown-curves.pdf}
+\includegraphics[width=16cm]{figs_tables/SI_fig4_SLA_maxMC.pdf}
+\caption{Maximum water retention by specific leaf area (SLA) across eight species.}
+  \label{fig:maxmc-di}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+\includegraphics[width=16cm]{figs_tables/SI_fig5_di_bd.pdf}
+\caption{Linear model predicting dessication rate as a function of specific leaf area and litter bulk density across eight species.}.
+  \label{fig:bd-di}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+\includegraphics[width=8cm]{figs_tables/SI_fig6_mixture_drydown-curves.pdf}
 \caption{Fuel moisture by time since saturation for four litter mixtures.
@@
 Mixed linear model results shown in SI Table 7.}
-  \label{SI_fig4}
+  \label{SI_fig6}
 \end{figure}

### 2019-08-06 scripts/moisture-ms-figs-tables.R

@@ -116,10 +116,10 @@ mtrait <- left_join(mtrait, oldtraits)
 
 
 ###############################################################################
-## Fig. 2: leaf area and max retention
+## SI Fig. 4: leaf area and max retention
 ###############################################################################
 
-fig2 <- ggplot(mtrait, aes(SLA, maxMC)) +
+SI_fig4 <- ggplot(mtrait, aes(SLA, maxMC)) +
   geom_point(size=3) +
 #  geom_errorbar(aes(ymin=maxMC+maxMC.se, ymax=maxMC-maxMC.se))+
 #  geom_errorbarh(aes(xmin=leaf.area.mean+(leaf.area.sd/sqrt(6)), xmax=leaf.area.mean-(leaf.area.sd/sqrt(6))))+
@@ -137,14 +137,14 @@ fig2 <- ggplot(mtrait, aes(SLA, maxMC)) +
   ylab("Maximum moisture content (%)") +
   pubtheme.nogridlines
 
-fig2
-ggsave(file.path(RESULTS, "fig2_SLA_maxMC.pdf"), plot=fig2,
+SI_fig4
+ggsave(file.path(RESULTS, "SI_fig4_SLA_maxMC.pdf"), plot=SI_fig4,
        width=col1, height=col1, units="cm")
 
 ###############################################################################
-## Fig. 3: leaf traits and dessication rate
+## SI Fig 5: leaf traits and dessication rate
 ###############################################################################
-fig3 <- ggplot(mtrait, aes(bd.mean, abs(di))) +
+SI_fig5 <- ggplot(mtrait, aes(bd.mean, abs(di))) +
   geom_point(size=3, stroke=0, alpha=0.8) +
 #  geom_errorbar(aes(ymin=abs(di)+di.se, ymax=abs(di)-di.se))+
 #  geom_errorbarh(aes(xmin=bd.mean+(bd.sd/sqrt(6)), xmax=bd.mean-(bd.sd/sqrt(6))))+
@@ -157,8 +157,8 @@ fig3 <- ggplot(mtrait, aes(bd.mean, abs(di))) +
                     size=smsize-6,
                   fontface="italic") +
   pubtheme.nogridlines
-fig3
-ggsave(file.path(RESULTS, "fig3_di_bd.pdf"), plot=fig3,
+SI_fig5
+ggsave(file.path(RESULTS, "SI_fig5_di_bd.pdf"), plot=SI_fig5,
        width=col1, height=col1, units="cm")
 
[…]

### 2019-08-06 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Move citations to shorter custom bib file
Other minor corrections

@@
 \begin{document}
-
 \maketitle
@@
 
-% .
-The moisture content of a mixture at any given time was also dependent upon the bulk density of the mixture (anova analysis resulted in a better model if bulk density was included as a fixed effect, p=0.01135).
-
 \begin{figure}[h]
@@
 water during wetting, and the one mixture without any Cupressaceae had the most
-rapid drying rate (SI Figure 4 and Si Table 8). Drydown rates, however, did not behave as the
-average of the constituent species: mixtures tended to be drier than expected
-early in the process and wetter than expected later (Fig.
+rapid drying rate (SI Figure 4 and Si Table 8). Drydown rates, however, did not
+behave as the average of the constituent species: mixtures tended to be drier
+than expected early in the process and wetter than expected later (Fig.
 \ref{fig:mixture_obs_pred_mc}). The fact that mixtures tended to have lower
@@
 al found that non-additivity was greater in moist mixtures of heathland fuels
-than in dry fuels. In their work, non-additivity was generally negative but cases of 
-positive non-additivity in mixtures was often a result of mixtures meeting
-ignition thresholds not met by the constituent species alone. To limit the
-number of models fit, we limited our investigation to two flammability
+than in dry fuels. In their work, non-additivity was generally negative but
+cases of positive non-additivity in mixtures was often a result of mixtures
+meeting ignition thresholds not met by the constituent species alone. To limit
+the number of models fit, we limited our investigation to two flammability
 parameters that represent two major axes of flammability variation. Therefore,
@@
 \section*{Author contributions}
-  conceptualization, RMQM and DWS; methodology, RMQM and
-  DWS; experiments, RMQM; coding and data analysis, DWS and RMQM; writing DWS
-  and RMQM; writing--review and editing, DWS
+
+conceptualization, RMQM and DWS; methodology, RMQM and DWS; experiments, RMQM;
+coding and data analysis, DWS and RMQM; writing DWS and RMQM; writing--review
+and editing, DWS
 
@@
 \newpage
-\bibliography{/home/schwilk/write/bib/schwilk}
+\bibliography{Magalhaes_Schwilk_litter_flam_moisture.bib}
 

### 2019-08-06 scripts/moisture-ms-figs-tables.R

@@ -462,8 +462,8 @@ fig5 <- ggplot(b, aes(hour, consum, color=taxon)) +
         legend.text = element_text(family=fontfamily, size=smsize-1, face="italic"),
         legend.title=element_blank()) +
   geom_jitter(width=2, size=1.5, alpha=0.7, stroke=0)
-
-ggsave(file.path(RESULTS, "fig5_consume_time.pdf"), plot=fig7,
+fig5
+ggsave(file.path(RESULTS, "fig5_consume_time.pdf"), plot=fig5,
        width=col1, height=col1, unit="cm")
 
 

### 2019-12-10 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
joeR1

@@
 
-\noindent \textbf{Keywords:} California, flame spread rate, flammability, fuel consumption, fuel moisture, leaf litter, non-additivity
+\noindent \textbf{Keywords:} California, fire, flame spread rate, flammability, fuel consumption, fuel moisture, leaf litter, non-additivity
 
@@
 but we expect that decomposition will affect moisture dynamics. For example,
-Van Wagner \citep{Van_Wagner-1969}, investigating fuel weathering, found that
+Van Wagner \citeyear{Van_Wagner-1969}, investigating fuel weathering, found that
 pine needles had a variation in moisture content 3 to 6 times greater when they
@@
 \section*{Data and code accessibility}
-All data and code for this project is at \url{https://github.com/schwilklab/trait-flam}
+All data and code for this project is at
+\url{https://github.com/schwilklab/trait-flam}. DOI to be provided. For review,
+use ``joe'' branch which contains this manuscript and code needed to create the
+figures and tables.
 

### 2019-12-10 ms/SI.tex
joeR1

@@
   \label{SI_fig2}
-  \includegraphics[width=17cm]{figs_tables/SI_fig2_max_water_emmeans.pdf}
+  \includegraphics[width=16cm]{figs_tables/SI_fig2_max_water_emmeans.pdf}
 \caption{Estimated marginal means by species for maximum water retainability based on mixed model results. Shaded regions indicate 95\% confidence intervals for the marginal means.}
@@
   \label{SI_fig3}
-\includegraphics[width=17cm]{figs_tables/SI_fig3_dessication_emmeans.pdf}
+\includegraphics[width=16cm]{figs_tables/SI_fig3_dessication_emmeans.pdf}
 \caption{Estimated marginal means by species for dessication rate based on mixed model results. Shaded regions indicate 95\% confidence intervals for the marginal means.}
@@
   \centering
-\includegraphics[width=16cm]{figs_tables/SI_fig4_SLA_maxMC.pdf}
+\includegraphics[width=8cm]{figs_tables/SI_fig4_SLA_maxMC.pdf}
 \caption{Maximum water retention by specific leaf area (SLA) across eight species.}
@@
   \centering
-\includegraphics[width=16cm]{figs_tables/SI_fig5_di_bd.pdf}
+\includegraphics[width=8cm]{figs_tables/SI_fig5_di_bd.pdf}
 \caption{Linear model predicting dessication rate as a function of specific leaf area and litter bulk density across eight species.}.
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/fig2_SLA_maxMC.pdf}
+\includegraphics[width=8cm]{figs_tables/SI_fig4_SLA_maxMC.pdf}
 \caption{Maximum water retention by specific leaf area across eight species.
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/fig3_di_bd.pdf}
+\includegraphics[width=8cm]{figs_tables/SI_fig5_di_bd.pdf}
 \caption{Drying rate rate as a function of specific leaf area and litter bulk

### 2019-12-10 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Start on JoE revisions

@@
+%% Revision 2019-12-10
 \documentclass[letterpaper,12pt]{article}
@@
 
-\title{Moisture effects on leaf litter are species-specific and result in non-additive flammability in mixed conifer forest}
+\title{Moisture effect on leaf litter alters non-additive flammability in mixed conifer forest}
 
@@
 Fuel moisture content is an important, and often driving, determinant of
-surface fire behavior. Environmental conditions such as temperature, relative
-humidity, and wind drive the moisture content of the fuel available to burn
-\citep{Kreye+Varner+etal-2018}. In particular, fine dead fuels respond very
-readily to environmental changes \citep{Nelson-2001}. Fuel moisture influences
-flammability and fire behavior through effects on ignition, spread rate, flame
-height, fuel consumption, and heat release, due to the high specific heat
-content of water \citep{Rothermel-1972, Nelson-2001}. Moisture acts as a
-heat-sink resulting in less energy available for propagation, and the resulting
-water vapor will both cool the flaming front and dilute flammable gases
-\citep{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
+surface fire behavior. Moisture acts as a heat-sink resulting in less energy
+available for propagation, and the resulting water vapor will both cool the
+flaming front and dilute flammable gases \citep{Albini-1976, Shafizadeh-1977}.
+Environmental conditions such as temperature, relative humidity, and wind drive
+the moisture content of the fuel available to burn
+\citep{Kreye+Varner+etal-2018}. Variation in moisture of living fuels has
 strong effects on flammability, but moisture of dead excised litter fuels can
-also vary. Moisture of litter fuels varies with climate, with seasons, and with
-short term weather within a season. Therefore, the rate of drying and
-flammability at any particular moisture content are key parameters that
-influence fuel flammability throughout the fire season \citep{Nelson-2001}.
+also vary. In particular, fine dead fuels respond very readily to environmental
+changes \citep{Nelson-2001}. Fuel moisture influences flammability and fire
+behavior through effects on ignition, spread rate, flame height, fuel
+consumption, and heat release, due to the high specific heat content of water
+\citep{Rothermel-1972, Nelson-2001}. Moisture of litter fuels varies with
+climate, with seasons, and with short term weather within a season. Therefore,
+the rate of drying and flammability as a function of moisture content are key
+parameters that influence fuel flammability throughout the fire season
+\citep{Nelson-2001}.
 
@@
 effects as well. Explicit models of fire behavior as a function of time since
-wetting could improve fire prediction, however. In many litter fuel driven
+wetting could improve fire prediction, however.  TODO REWRITE: In many litter fuel driven
 systems, natural lightning ignitions are concurrent with precipitation and,
 therefore, fire behavior can depend on rate of drying following a wetting
-event. If species differ in their drying rate, this may strongly influence
-relatively flammability across species at any particular time following
-wetting. Such differential moisture content across time, may contribute to the
+event. If species differ in their drying rate, this may influence
+relatively flammability across species and over time following
+wetting. Such differential moisture content across time may contribute to the
 often high spatial heterogeneity of surface fires
@@
 flammability is poorly known and could be driven by by leaf traits, emergent
-litter litter traits, or both. Species specific effects on litter flammability
+litter traits, or both. Species specific effects on litter flammability
 are recorded for the forests of the southeastern US \citep{Nowacki+Abrams-2008}
 where species-specific litter drying rates play a role in flammability
-feedbacks \citep{Kreye+Varner+etal-2013}. Similarly, in mixed-conifer forest of
-the western USA, where the fire regime is characterized by surface fires fueled
+feedbacks \citep{Kreye+Varner+etal-2013}.
+Fire behavior consequences of individual species traits have been recorded for western US forests as well: in mixed conifer forest in California, where the fire regime is characterized by surface fires fueled
 by plant litter, overstory tree composition drives variation in local fire
@@
 Fuel moisture content through time is controlled by two main processes,
-absorptive capacity and rate of moisture loss \citep{Kreye+Kobziar+etal-2013}.
+absorption and retention \citep{Kreye+Kobziar+etal-2013}.
 Absorptive capacity represents the potential maximum fuel moisture and is
-evaluated by the moisture content of litter when saturated. After saturation,
-the differences in moisture content indicate different abilities to retain
+evaluated by the moisture content of litter when saturated. Differences in moisture content over time since saturation can indicate different abilities to retain
 moisture. Litter particle geometry may have direct effects on maximum moisture
 absorption through area-mass relationships and we also expect effects of leaf
-morphology, especially leaf cuticle thickness \citep{Van_Wagner-1969}. Drying
-rates should also be influenced by leaf and litter particle traits:
+morphology, especially leaf cuticle thickness \citep{Van_Wagner-1969}. Like absorption, mmoisture retention should also be influenced by leaf and litter bed traits:
 surface-area-to-volume ratio and litter bulk density will influence air flow
-through the litter bed. Litter packing therefore has two pathways of effect on
-fire behavior: directly by its effect on oxygen availability for combustion,
-and indirectly by its effect on dry-down rate, and both these pathways are
-affected by leaf litter traits, especially leaf size
+through the litter bed. Litter packing therefore effects fire behavior in two ways: directly by its effect on oxygen availability for combustion,
+and indirectly by its effect on dry-down rate. Both these pathways are
+affected by leaf particle traits, especially leaf size
 \citep{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves with lower
-moisture retainability and should favor drying, thus increasing the probability
+moisture retainability should favor drying, thus increasing the probability
 of ignition and fire spread.
@@
 axes of variation are described by measurements related to flame spread rate
-and measurements related to heat transfer. Flame spread rate is easy to measure
+and measurements related to heat transfer [CITATIONS]  . Flame spread rate is easy to measure
 in burning trials, but heat transfer can be more difficult and often proxies
@@
   Varner+Kane+etal-2015}. These different components of fire behavior have
-differing ecological effects. Species traits may interact such that the
+differing ecological effects.
+
+TODO
+
+Species traits may interact such that the
 flammability of multi-species mixtures is not the average of the constituent
-species. Such non-additive effects are synergistic in dry litter fuels where
+species. Such non-additive effects are often synergistic in dry litter fuels where
 the flammability of a mixture is driven by the most flammable species
-\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. It is possible
+\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. TODO cite negative flammability. Zhao et al 20198]
+
+It is possible
 that similar non-additive effects may influence litter moisture content,
-especially if different leaf traits influence absorptive capacity and drying
-rate. There is some evidence for non-additive flammability in moist fuels with
+
+% especially if different leaf traits influence absorptive capacity and drying
+% rate.
+
+There is some evidence for non-additive flammability in moist fuels with
 one study finding increasing non-additivity with moisture
@@
 maximum water retention and rate of moisture loss and examined how these vary
-in relation to two leaf and litter bed traits. We then examined the influence
+in relation to two leaf and litter bed traits [CHECK AND FIX]. We then examined the influence
 of moisture on the flammability across these species. Experimental burn trials
@@
 species. Here, we examined if this positive non-additivity was maintained in
-mixtures across varying levels of fuel moisture. The burn trials were also
-performed with multi-species litter mixtures to evaluate if species contribute
-equally to litter moisture dynamics or if there are non-additive effects as
+mixtures of varying levels of fuel moisture. The burn trials also
+performed with multi-species litter mixtures to evaluate if there are non-additive effects as
 observed in oven-dried litter. We tested if the non-additive effects on
-reported in previous studies occur under ecologically-relevant moisture levels
+reported in previous studies occur under [TODO CHECK} ecologically-relevant moisture levels,
 and whether fuel moisture effects strengthen or moderate such non-additivity.
@@
 Field sites for this study were located in Sequoia and Kings Canyon National
-Parks, California, USA (36 36N, 118 42 W) between 1600 and 2400 m elevation in
-mixed-conifer forest. We chose eight dominant tree species that are
-representative of this type of forest: \emph{Pinus jeffreyi} Grev. \& Balf.,
-\emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
+Parks, California, USA (N$36^\circ 36'$, W$118^\circ 42'$) between 1600 and
+2400\,m elevation in mixed-conifer forest. We chose eight dominant tree species
+that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \&
+Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
 \emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
@@
 Buchholz. In this study, we consider litter the top layer of leaves and small
-twigs less than 0.625 cm diameter (= 1-hour fuel) that have fallen in the
-previous year (mostly undecomposed). Leaf litter was collected in the summer
+twigs less than 0.625\,cm diameter (= 1-hour fuel) that have fallen in the
+previous year (mostly undecomposed or lightly decomposed). Leaf litter was collected in the summer
 (mid-June to mid-July) in 2012 from 21 separate sites across the study area. We
@@
 obtain a more uniform sample of leaf and fine branchlet litter, because bark
-and heavier branches tend to fall closer to the tree trunk. All litter was air
-dried to $< 5$\% fuel moisture by dry weight for weighing and dividing into
-moisture trials. The moisture manipulations and subsequent burning trials were
-conducted in April--August 2015.
+and heavier branches tend to fall closer to the tree trunk. The top 2.5-5\,cm
+of litter were collected by raking and shoveling into large bags. All litter
+was air dried to $< 5$\% fuel moisture by dry weight, then weighed and divided into moisture trials. The collected litter was dried in large batches, then
+stored in sealed plastic bags inside cardboard boxes in a lab storage space. We
+minimized handling and moving the bagged litter to avoid mechanical damage and
+changes to the particle size distributions. The moisture manipulations and
+subsequent burning trials were conducted in April--August 2015.
 
@@
 per mixture type) were placed in baskets approximately 45 x 45\,cm by 15\,cm
-tall constructed of 1\,cm aluminum screen. Samples were 450\,g dry weight
-litter per trial and litter depth was standardized to 10\,cm by changing the
-horizontal dimensions of the basket. Weight was measured by using a balance
-sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples
-were immersed in plastic storage bins filled with water for 24 hours to fully
-saturate them. After saturation, the baskets were removed from the water,
-allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
-content. The samples were then allowed to dry in a controlled environment kept
-at 21\,C temperature and 30\% relative humidity, assessed via an iButton sensor
-(Maxim Integrated) placed over the samples. Litter baskets were arranged in a
-completely randomized design. During the dry-down process three subsamples of
-5--10\,g were taken at 24 hour intervals and weighed. These subsamples were
-then oven dried for 24 hours at 100\,C and re-weighed to assess moisture
-content. This was repeated until the final sample from a basket was less than
-10\% moisture.
+tall constructed of aluminum mesh with 0.52\,cm square openings. Samples were
+450\,g dry weight litter per trial and litter depth was standardized to 10\,cm
+by changing the horizontal dimensions of the basket. Weight was measured by
+using a balance sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus,
+OH). Samples were immersed in plastic storage bins filled with water for 24
+hours to fully saturate them. After saturation, the baskets were removed from
+the water, allowed to drain for 3 minutes, and samples weighed to measure
+maximum moisture content. The samples were then allowed to dry in a controlled
+environment kept at $21^\circ$C temperature and 30\% relative humidity, assessed via
+an iButton sensor (Maxim Integrated) placed over the samples. Litter baskets
+were arranged in a completely randomized design. During the dry-down process
+three subsamples of 5--10\,g were taken at 24 hour intervals and weighed. These
+subsamples were then oven dried for 24 hours at $100^\circ$C and re-weighed to assess
+moisture content. This was repeated until the final sample from a basket was
+less than 10\% moisture.
 
@@
 the observed moisture dynamics was low, so we decided \emph{a priori} to test
-the effect of one leaf trait likely to influence moisture absorption, specific
-leaf area, and one litter trait known to directly effect flammability and
-likely to influence drying rate, litter bulk density. Specific leaf area was
-measured on freshly fallen litter by sampling 5 individuals per species. This
-measurement included petiole and for the scale-leaved species included the full
-excised branchlet. Therefore, this is not directly comparable to typical specific leaf
-area measurements on fresh leaves. Litter bulk density was obtained in this
-experiment for each basket by dividing the dry mass (450\,g) by the estimated
-litter volume (basket width x length x litter depth).
+the effect of one leaf trait likely to influence moisture absorption (specific
+leaf area), and one litter trait known to directly affect flammability and
+likely to influence drying rate and litter packing density (litter bulk
+density). [TODO check]  Specific leaf area was measured on freshly fallen litter by sampling
+5 individuals per species. This measurement included petiole and for the
+scale-leaved species included the full excised branchlet. Therefore, this is
+not directly comparable to typical specific leaf area measurements on fresh
+leaves. Litter bulk density was obtained in this experiment for each basket by
+dividing the dry mass (450\,g) by the estimated litter volume (basket width x
+length x litter depth).
 
@@
 trials all used 450\,g (dry weight) of litter. For three-species mixtures, this
-was 150\,g per species. This design allows fire to reach constant flame spread
-rates \citep{Magalhaes+Schwilk-2012}. Two graduated metal rulers equally spaced
-along the apparatus allowed visual determination of maximum flame height.
-Litter was ignited with a propane torch; maximum adiabatic flame temperature of
-1899 C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the
-apparatus and allowed to burn until extinction. If a sample failed to ignite
-after 30 seconds, no further attempt was made and if three samples of a species
-at a particular moisture level failed to ignite, no further attempts were made
-on that species/moisture combination and the replicate was allowed to continue
-drying. A timer was used to record total time to ignition and duration of
-flaming combustion. The flammability trials followed the methodology detailed
-in de Magalhães and Schwilk \citep{Magalhaes+Schwilk-2012}.
+was 150\,g per species. [TODO give depths] This design allows fire to reach
+constant flame spread rates \citep{Magalhaes+Schwilk-2012}. Two graduated metal
+rulers equally spaced along the apparatus allowed visual determination of
+maximum flame height. Litter was ignited with a propane torch; maximum
+adiabatic flame temperature of 1899$^\circ$C,
+\url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the apparatus
+and allowed to burn until extinction. If a sample failed to ignite after 30
+seconds, no further attempt was made and if three samples of a species at a
+particular moisture level failed to ignite, no further attempts were made on
+that species/moisture combination and the replicate was allowed to continue
+drying until the next scheduled checkpoint. A timer was used to record total
+time to ignition and duration of flaming combustion. The flammability trials
+followed the methodology detailed in 
+\citet{Magalhaes+Schwilk-2012}.
 
@@
 Some trials resulted only in smoldering combustion; for those trials, a time to
-ignition and mass loss were recorded, but flame spread rate and flame height
+ignition and fuel consumption were recorded, but flame spread rate and flame height
 were recorded as zero. Our thermocouple system failed or data was lost for two
@@
 spread rate captures one important axis of flammability variation
-\citep{Schwilk-2015, Pausas+Keeley+etal-2017} and percent fuel consumed
-captures a portion of the second major axis of flammability variation (although
-not as well as integrating temperature over time does). Our analysis,
+\citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal02018} and percent fuel consumed
+captures a portion of the second major axis of flammability variation. Our analysis,
 therefore, focused on these two relatively orthogonal measures of flammability
@@
 Luke \citeyear{Luke-2017} as producing acceptable Type I error rates. We tested if
-one leaf trait (specific leaf area) and one litter trait (litter bulk density)
+one litter particle trait (specific leaf area) and one litter bed trait (litter bulk density)
 influenced moisture dynamics by fitting linear models with species means as
@@
 
-Across these eight species specific leaf area had a positive effect on maximum
+Across these eight species specific leaf area had a positive association with maximum
 water retention (SI Fig 4, linear model $P$ = 0.003) and litter
@@
 Effect of time &  consum \~{} 1 & 140.08 &  &  &  \\ 
- & \~{} taxon & 155.84 & 4.00 & 31.51 & $<$ 0.001 \\ 
- & \~{} taxon + time & 209.84 & 1.00 & 108.01 & $<$ 0.001 \\ 
- & \~{} hour + taxon + taxon:time & 216.96 & 4.00 & 14.23 & 0.01 \\ 
+ & \~{} taxon & 155.84 & 4 & 31.51 & $<$ 0.001 \\ 
+ & \~{} taxon + time & 209.84 & 1 & 108.01 & $<$ 0.001 \\ 
+ & \~{} hour + taxon + taxon:time & 216.96 & 4 & 14.23 & 0.01 \\ 
     \bottomrule
@@
 failed to even ignite smoldering combustion are omitted completely from the
-model and figure). At that point, the \emph{Abies}, \emph{Calocedrus}, and
+model and figure). At 72 hours, the \emph{Abies}, \emph{Calocedrus}, and
 \emph{Sequoiadendron} had dryer litter than did the \emph{Quercus}, yet the
@@
 example, the mixture that does not contain \emph{Q. kelloggii} (AbCaPi in SI
-Fig. 6) has the lowest initial retention. Initial water retention and rates
+Fig. 6) has the lowest initial retention. Initial water retention and drying rates
 differed across mixtures (linear model based on log of moisture content, SI Tables 8 and 9).
@@
 drying (Wald test shows slope of mixed linear model illustrated in Fig.
-\ref{fig:mixture_obs_pred_mc} less than 1, $P$ < 0.0001).
+\ref{fig:mixture_obs_pred_mc} less than 1, $P < 0.0001$).
 
@@
 
-\subsection*{Species-specific moisture dynamics flammability}
+\subsection*{Genus-specific moisture dynamics and flammability}
 
-The species in this study displayed different patterns of both moisture
-absorption and desorption, apparently controlled by different litter traits. In
+The genera in this study displayed different patterns of both moisture
+absorption and desorption, associated with different litter traits. In
 this sample of species, those that have a greater absorptive capacity are also
@@
 work in this system \citep{Magalhaes+Schwilk-2012} was that flammability trials
-tested only very dry fuels. We therefore expected that rankings of species
-flammability might change under more realistic moisture levels. For example,
+tested only dry fuels. We therefore expected that rankings of species
+flammability might change under more [TODO FIX] realistic moisture levels. For example,
 one presumption of ours prior to this experiment was that the high flammability
@@
 drying rate. This has consequences for flammability as a function of species
-and time since wetting. Our flammability trials demonstrate that that both
+and time since wetting. Our flammability trials demonstrate that both
 flame spread rate and fuel consumption increased with time since wetting but
@@
 ignition is possible, having faster spread rates than the other taxa. The
-Cupressaceae, which have dense litter but low maximum water absorption had low
+Cupressaceae, which in our two species have dense litter but low maximum water absorption had low
 flammability across moisture levels.
+This result is consistent with the low flammability found in an Australian Cupressaceae that have small litter particles and dense packing citep{Scarff+Westoby-2006, Bowman+Haverkamp-2018}.
+However, the densely packed litter is not a consistent trait across the family because \citet{Cornwell+Elvira+etal-2015} found relatively high litter flammability and low litter density for some Asian Cupressaceae.
+
+[TODO] [more?]
 
@@
 mixtures. These mixtures contained species representative of all dry-down
-behavior types observed in single species trials. As expected from the single
-species results, mixtures that contained \emph{Q. kelloggii} absorbed more
-water during wetting, and the one mixture without any Cupressaceae had the most
-rapid drying rate (SI Figure 4 and Si Table 8). Drydown rates, however, did not
-behave as the average of the constituent species: mixtures tended to be drier
-than expected early in the process and wetter than expected later (Fig.
-\ref{fig:mixture_obs_pred_mc}). The fact that mixtures tended to have lower
-water content than predicted from the mean of the constituent species after
-wetting, but higher water content than expected after long drying is surprising
-because past work has shown that mixtures tend to pack less densely than do
-litter monocultures \citep[][, Table 2]{Magalhaes+Schwilk-2012}. If packing
-density controls rate of drying, then we would expect mixtures to dry more
-rapidly than predicted from monoculture trials. This pattern in moisture
-content through time, however, did not directly predict flammability through
-time.
+behavior types observed in single species trials. Mixtures that contained
+\emph{Quercus} absorbed more water during wetting, and the one mixture
+without any Cupressaceae had the most rapid drying rate (SI Figure 4 and SI
+Table 8). Drydown rates, however, did not behave as the average of the
+constituent species: mixtures tended to be drier than expected early in the
+process and wetter than expected later (Fig. \ref{fig:mixture_obs_pred_mc}).
+The fact that mixtures tended to have lower water content than predicted from
+the mean of the constituent species after wetting, but higher water content
+than expected after long drying is surprising because past work has shown that
+mixtures tend to pack less densely than do litter monocultures \citep[][Table
+2]{Magalhaes+Schwilk-2012}. If packing density controls rate of drying, then we
+would expect mixtures to dry more rapidly than predicted from monoculture
+trials. This pattern in moisture content through time, however, did not
+directly predict flammability through time.
 
@@
 influenced by the lower bulk densities of litter mixtures containing those two
-species. Most interestingly, there were striking patterns of non-additivity in
+species. Most interestingly, there were patterns of non-additivity in
 spread rate and in fuel consumption for the four mixtures, indicating that
-positive non-additivity in flammability is not restricted to completely dry
-litter mixtures observed in previous work \citep{Magalhaes+Schwilk-2012,
+positive non-additivity in flammability is not restricted to the completely dry
+litter mixtures studied in previous work \citep{Magalhaes+Schwilk-2012,
   VanAltena+Logtestjin+etal-2012}. Non-additivity in both spread rate and fuel
@@
 cases of positive non-additivity in mixtures was often a result of mixtures
-meeting ignition thresholds not met by the constituent species alone. To limit
+meeting ignition thresholds not met by the constituent species alone [TODO: check wording]. To limit
 the number of models fit, we limited our investigation to two flammability
@@
 
-Overall, our study demonstrates a consistent pattern of increasingly positive
+Overall, we report a consistent pattern of increasingly positive
 non-additivity in flammability with litter drying across two largely orthogonal
-measures of flammability. This result, together with previously published work,
+measures of flammability \citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal02018}. This result, together with previously published work,
 suggests that litter flammability is prone to non-additive effects, suggesting
@@
 \citep{Magalhaes+Schwilk-2012}.
+
+[TODO rewrite below]
 
@@
 \citep{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}. The patterns we
-demonstrate here show to potential importance of species traits and
+demonstrate here show the potential importance of species traits and
 interactions at local and stand scales where management and fire operations

### 2019-12-10 ms/response_to_reviews.tex
Start on JoE revisions

 
\documentclass[letterpaper, 12pt]{letter}
\usepackage{times, fullpage}
\usepackage{url}

\address{Department of Biological Sciences\\Flint and Main\\Texas Tech University\\
  Lubbock, TX 79409}

\name{Dylan W. Schwilk}

\begin{document}
\begin{letter}{}

\opening{Dear Dr. David Gibson,}

Enclosed is our revised manuscript research article, ``''.

We thank the editor and reviewer for their helpful comments and suggestions. We
have itemized each comment below and followed each with our reply and a list of
changes we made in response to the comment.

{\bf Responses to Senior Editor's comments:}

\begin{quote}
 I support the AE's recommendation to reject but allow resubmission. However, the bar for a resubmission will be high. My feeling is that your study confirms the essential findings from your 2012 study published in Journal of Ecology with the addition that your methodology of drying down adds some realism. Indeed, the confirmatory nature of this study was pointed out by reviewer 2 in their point 6. In addition to the points raised by the AE and reviewers, I will be looking in any resubmission for more evidence that this study advances your earlier work (apart from your last paragraph, your discussion is largely an explanation of your results with little context from other studies), that the findings have real ecological relevance (i.e., does it matter to the mixed-conifer communities?) and how is the new dry down treatment related to dry down conditions in litter mixtures in the field. If you can adequately address all of these concerns then I'll be happy to consider a resubmission.
\end{quote}

TODO.

Add context also suggested by reviewer 1.

\begin{quote}

\end{quote}

TODO

{\bf Responses to Associate Editor's comments:}

\begin{quote}
I support the AE's recommendation to reject but allow resubmission. However, the bar for a resubmission will be high. My feeling is that your study confirms the essential findings from your 2012 study published in Journal of Ecology with the addition that your methodology of drying down adds some realism. Indeed, the confirmatory nature of this study was pointed out by reviewer 2 in their point 6. In addition to the points raised by the AE and reviewers, I will be looking in any resubmission for more evidence that this study advances your earlier work (apart from your last paragraph, your discussion is largely an explanation of your results with little context from other studies), that the findings have real ecological relevance (i.e., does it matter to the mixed-conifer communities?) and how is the new dry down treatment related to dry down conditions in litter mixtures in the field. If you can adequately address all of these concerns then I'll be happy to consider a resubmission.
\end{quote}

Thank you. Add context. Yes, several comments converged on the absolutely correct criticism that our discussion did not engage adequately with the literature.  Wer have added X and X...

inclduing cite Zhoa 2019

\begin{quote}
The authors mention leaf size explicitly as an important parameter in non-additivity (with some references including their own 2012 paper) but they do not include it in the current study. Especially at low moisture, the small individual needles of Abies should cause low ventilation and a high likelihood of negative non-additivity on flammability, as was shown by Zhao et al 2016 Ecol. Evol. Figure 4). In the Methods, as observed also by reviewer 2, there is very little detail on the extent to which litter of the respective species was picked up as individual leaves, branchlets without leaves versus branchlets with leaves still attached. This is really important, also because positive versus negative non-additivity of leaf-twig mixtures have been reported recently (Zhao et al. 2019 Plant and Soil). I feel that data and/or photos indicating the size and shape of the different litter types (and if possible variability in these) are important to add and to be used in the interpretation of the results.
\end{quote}

We have added a figure showing particle size distributions for each of these species. Yes, Abies litter is a a combination of still attached needes and those that have detached from the branchlet.

We

\begin{quote}
In the Methods, it was unclear how the various treatments and replicates were grouped over time, given that temperature and humidity in the fire facility could vary. Were there blocks in time with one replicate for each of the treatments (in random burning order) within a block? Or was there a different statistical design in time? 
\end{quote}

Guided by the drying rates measured in the initail trials, we staggered saturation times across species aiming to allow completely blocked design in time. This staggering was mostly successful but Because some ignition attempts were unsuccessful, some days lacked the full complemnt of species while others later in the porcess sometimes had more than one replicate per species burned.

TODO

\begin{quote}
Line 196 (see also reviewer 2): how can we assume that the quality of the stored litter is the same as that used in the previous study 3 years earlier, i.e. can these two datasets be safely combined?
\end{quote}

We have added detail as requested by reviewer 2. See lines (X-X). The litter was dried in the lab (very low hmidity, but not oven dried), for several weeks on open racks. Then the litter was bagged and sealed and stored within garbage bags inside cardboard boxes in a lab storage area.  We saw no signs of mold or ongoing decomposition. We minimized handling of the bags to avoid mechanical damage and changes to little particle size.

\begin{quote}
Line 365 (see also reviewer 1): on the other hand Cornwell et al. 2015 New Phyt reported mostly high flammability in a range of Cupressaceae.
\end{quote}

TODO.  Mention.  Maybe due to very small samples that miminimize effects of packing in Cornwell et al 2015?

{\bf Responses to the Reviewer 1's comments}

\begin{quote}
This work compares mixtures of fuels, relative to their individual components, in terms of moisture retention and flammability. Moisture dynamics were also investigated in relation to properties of leaves of the species forming litter. This information is important in understanding flammability of surface fuels in the field.

The study has been well thought through and well executed, and the ms is generally well written. Stylistically, the main problem is that commas are missing in many sentences. This might sound trivial, but will really help the reader.
\end{quote}

We thank the reviewer for their comments and apologize for the failure in proofreading. The reviewer should not have had to do such copy editing and I I sincerely apologize for that. We have taken heed of both reviewers' comments and fixed all specific grammar errors and fixed the issue we had with missing symbols. For brevity, we've omited response to the very simple suggested grammar fixes and simply made all requested changes. We respond to other points individually, below

\begin{quote}
line 72 - the comment about natural lightning ignitions comes from nowhere, and is a  bit of a jolt for the reader - rephrase slightly, e.g. many fires result from lightning ignitions, which are concurrent with ....
\end{quote}

TODO.

We have rewritten this in response to reviewer 2s point that dry lightning is also amajor ignition source.

TODO

\begin{quote}
line 103 - 'litter packing therefore affects fire behaviour in two ways' is easier to read
\end{quote}

Yes, agreed. We have tried to clean up the writing throughout and make the prose more active and precise.

\begin{quote}
line 133 - 'in mixtures of litter with varying levels..' is more explicit
\end{quote}

Agreed. fixed.

\begin{quote}
line 182 - affect, not effect
\end{quote}

Yes, an embarrassing error.

\begin{quote}
line 220 - see also Prior et al (2018)  Conceptualizing ecological flammability ....Fire 1, article 14
\end{quote}

Thank you. This was an oversight. We have also added more context from the
literature to our Discussion

TODO

\begin{quote}
lines 243/244 I was a bit puzzled here -It would help to be more specific here - which aspects of moisture dynamics? Which response variables did you use?  (or 'as described below')
\end{quote}

TODO. We have rewritten that section to be more precise.

\begin{quote}
line 290 - you show an association, not necessarily an effect
\end{quote}

True. We have edited this as suggested.

\begin{quote}
line 307 and line 326 - a minor inconsistency is that you analyse present species groups, yet discuss the results in terms of individual species
\end{quote}

TODO

Yes. We have made edits to

TODO. Check wording.

\begin{quote}
lines 348-349 - logically, this does not necessarily follow - factors such as large leaves or high SLA could independently affect both absorptive capacity and flammability - although fuel moisture does of course have a strong effect 
\end{quote}

TODO.

\begin{quote}
Table 4 - get rid of the unnecessary zeroes in the df column
\end{quote}

Fixed.

\begin{quote}
line 365 - wording is problematic - change to '...ignition is possible, it had faster spread rates ...'
\end{quote}

Agreed. Fixed.

\begin{quote}
line 365 - The low flammability of litter from your Cupressaceae species is interesting - Scarff and Westoby (2006) also found this for the Australian Cupressaceae genus Callitris, as did Bowman et al. (2018) J. Ecol 106, 1010 -1022
\end{quote}

We have added some better discussion of past results and included discussion of these papers, as well as contrasting with the Cornwell et al 2015 result as pointed out by the Associate Editor.

\begin{quote}
line 387 - capitalise 'I' in SI Table  8
\end{quote}

Fixed.

\begin{quote}
line 403 - change to 'to the completely dry litter mixtures studied in previous work'
\end{quote}

TODO

\begin{quote}
Figures 7 and 8 - is there a reason that you don't present the information in the same way as in Figure 6, if your main point is the discrepancy between the mixtures and the averaged individual values?
\end{quote}

We agree this does lack consistency and an early draft used the same concise Fig 6 format for all of them. But we decided that having time on the x-axis was worth the space and allows the reader more insight into the

Reviewer 2 pointed out some exceptions to the overall statistical pattern and those exceptions would not have been easily visible without an explicit time axis.

\begin{quote}
line 416 - Prior et al (2018) also show that rate of spread and consumability are largely independent in litter fuels
\end{quote}

TODO

Yes. We have added citations here.

{\bf Responses to the Reviewer 2's comments}

\begin{quote}

My major concerns are:
1. One motivation for the study is not valid. I do not completely agree with the authors that natural lightning ignitions are concurrent with precipitation and, therefore, fire behavior can depend on rate of drying following a wetting event. The efficiency of individual lighting strikes in igniting a fire is affected by the moisture properties of forest fuel at the moment of lighting (Krawchuket et al. 2006, Ecology). Dry lighting that occurs without any accompanied rainfall nearby is most likely to cause wildfire because the fuel is dry enough to be ignited.
\end{quote}

TODO.  Get data.

\begin{quote}
2. The research question of the research is not explicitly stated both in the abstract and in the introduction. In the end of the introduction, the authors focus on specific data-collection goals, while leave the reader to figure out what the research question is. 
\end{quote}

TODO.

\begin{quote}
3. The authors concluded that positive non-additivity appeared to be the rule in litter-driven fire and this non-additivity became more positive with fuel drying. But if we look closely at figure 8, non-additivity was negative within 50 hours since wetting, non-additivity was both positive and negative at about 75 hours since wetting, and non-additivity was positive at about 100 hours since wetting. The high variability of non-additive effects at about 75 hours since wetting might have important ecological effects on fire ignition and behavior, and are important information cannot be ignored.
\end{quote}

Yes, this is a good point and we have added some discussion of this.  We tried to lmit our statements, hwoever, to those supported by the statistical tests, and .....

TODO.

\begin{quote}
4. The writing of this manuscript needs to be improved for clarity and accuracy. There are many typos in the text, and the using to scientific terms lack consistency through the text.
\end{quote}

We have made extensive edits in response to both reviewer's suggestions

\begin{quote}
5. The title does not precisely communicate the key message of the study. I suggest it is more precise to say “moisture effect … and alter non-additive flammability…”
\end{quote}

TODO

\begin{quote}
6. The last paragraph of the discussion has a significant overlap with the last paragraph of the discussion in the authors’ previous paper (de Magalhães \& Schwilk 2012) in terms of the main content.
\end{quote}

We have rewritten much of the discussion to better include findings from the literature.....

\begin{quote}
The abstract can be shorter. It exceeds 350 words limit. 
\end{quote}

TODO

\begin{quote}
Line 50: add reference to the first sentence.
\end{quote}

TODO

We disagree in this case.

\begin{quote}
Lines 50-61: in this paragraph, after the first sentence I think it is more logic to first explain the mechanism of moisture’s negative effects on fire, then explain the reason of the driving importance of dead fuel moisture content to surface fire ignition and behavior because fine dead fuels respond more readily to environmental changes than living fuels.
\end{quote}

TODO

\begin{quote}
Line 72: revise the description “ litter fuel driven system”.
\end{quote}

TODO

\begin{quote}
Line 75: “at any particular time”?
\end{quote}

\begin{quote}
Line 85: revise “driven by by” and “emergent litter litter”.
\end{quote}

\begin{quote}
Line 88: “similarly”?
\end{quote}

\begin{quote}
Line 95: “absorption capacity” and “rate of moisture loss” are not correct descriptions of processes. Better say “absorption and retention”.
\end{quote}

\begin{quote}
Line 98: revise “after saturation” to “dried certain time after saturation” for clarity.
\end{quote}

\begin{quote}
Line 101: delete “also”.
\end{quote}

\begin{quote}
Lines 102-103: litter bulk density is not a litter particle trait. Delete “will influence air flow through the litter bed”.
\end{quote}

\begin{quote}
Lines 103-105: leaf size is not leaf litter trait.
\end{quote}

\begin{quote}
Line 107: “fuel-beds composed of larger leaves with lower moisture retainability”?
\end{quote}

\begin{quote}
Line 115: start a new paragraph here.
\end{quote}

\begin{quote}
Lines 117-119: better specify the system, because negative non-additive effects on litter
flammability were also found.
\end{quote}

\begin{quote}
Line 120: “especially if different leaf traits influence absorptive capacity and drying rate”?
\end{quote}

\begin{quote}
Line 125: revise “two leaf and litter bed traits”, and better directly say “SLA and packing density”.
\end{quote}

\begin{quote}
Line 133: delete “also”.
\end{quote}

\begin{quote}
Line 134: “if species contribute equally to litter moisture dynamics”?
\end{quote}

\begin{quote}
Line 136: delete “on”. Revise “ecologically-relevant”.
\end{quote}

\begin{quote}
Line 141: revise “(36 36N, 118 42 W)”.
\end{quote}

\begin{quote}
Line 147: at least for Quercus, leaves fallen in the previous year will be partly decomposed in the summer.
\end{quote}

\begin{quote}
Line 155: add more details about how the litter was collected?
\end{quote}

\begin{quote}
Line 157: is it possible that all litter has less than 5% fuel moisture when air-dried?
\end{quote}

\begin{quote}
Line 158: the collecting of the leaf litter is in 2012. The conduction of the experiment is in 2015. Can the author explain how do they save the litter?
\end{quote}

\begin{quote}
Line 167: 1 cm ? aluminum screen.
\end{quote}

Yes, this was a typo. Thanks. The aluminum mesh openings were 0.203 inches. We have clarified and corrected to ``0.52\,cm''

\begin{quote}
Line 173: revise “21C temperature”.
\end{quote}

\begin{quote}
Line 182: revise “effect” to “affect”. 	Revise “litter bulk density” to “litter packing density”
\end{quote}

\begin{quote}
Line 185: I do not think this sentence is necessary because whether or not petioles should be included in SLA measurements depends on the research question at hand.
\end{quote}

\begin{quote}
Line 201: the fuel depth was controlled or not?
\end{quote}

\begin{quote}
Line 205: revise “1899C”.
\end{quote}

\begin{quote}
Line 208: revise “species/moisture combination” to “species and moisture combination”.
\end{quote}

\begin{quote}
Line 209: “the replicate was allowed to continue drying for …”
\end{quote}

\begin{quote}
Line 216: revise “mass loss” to “fuel consumption”.
\end{quote}

\begin{quote}
Line 224: if not all flammability data was used in the analysis, then why measure them?
\end{quote}

\begin{quote}
Line 242: revise “one leaf trait and one litter trait” to “one litter particle trait and one litter bed traits”.
\end{quote}

\begin{quote}
Line 307: revise “at that point”.
\end{quote}

\begin{quote}
Lines 314-320: this indicates that even in the same genus, species vary in flammability.
\end{quote}

\begin{quote}
Line 330: “rates”?
\end{quote}

\begin{quote}
Line 344: “moisture dynamics flammability” is difficult to understand.
\end{quote}

\begin{quote}
Line 346: revise “apparently controlled”.
\end{quote}

\begin{quote}
Line 350: revise “very dry”.
\end{quote}

\begin{quote}
Line 351: revise “more realistic moisture levels”.
\end{quote}

\begin{quote}
Line 361: revise “that that”.
\end{quote}

\begin{quote}
Line 366: “but”?
\end{quote}

\begin{quote}
Line 401: “striking”？
\end{quote}

\begin{quote}
Line 410: “not met by the constituent species alone”?
\end{quote}

\begin{quote}
Line 415: “demonstrates”?
\end{quote}

\begin{quote}
Line 439: revise “to”.
\end{quote}

\closing{Sincerely,}

\end{letter}
\end{document}

### 2019-12-10 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
fixup

@@
 the flammability of a mixture is driven by the most flammable species
-\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. TODO cite negative flammability. Zhao et al 20198]
+\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. TODO cite negative flammability. Zhao et al 2019]
 
@@
 observed in oven-dried litter. We tested if the non-additive effects on
-reported in previous studies occur under [TODO CHECK} ecologically-relevant moisture levels,
+reported in previous studies occur under [TODO CHECK]
+ecologically-relevant moisture levels,
 and whether fuel moisture effects strengthen or moderate such non-additivity.

### 2019-12-11 ms/response_to_reviews.tex
Work on edits to introduction
in response to reviews

@@
 
-We have rewritten this in response to reviewer 2s point that dry lightning is also amajor ignition source.
+This portion was poorly written.
+We have rewritten this in response to reviewer 2's comment that most fires (or, at elast burned area) result from dry lightning strikes across the west.
+See our responses below.
 
@@
 \end{quote}
+
+Yes, this sentence was poorly written and oversimplified. We thank the reviewer
+for pointing this out and we have rewritten this section. However, our badly
+made point was: lighting ignitions can occur during the same general period as
+precipitation. As the reviewer points out, dry strikes that find dry fuel are
+more likely to lead to fires (Abatzoglou et al 2016). That the total number
+lightning strikes does not well predict burned area and that ignitions occur
+often from dry strikes argues for the importance of fuel moisture. Although
+synoptic weather conditions that create dry and wet storms differ, the
+potential for precipitation and lightning during the same overall period of
+thunderstorm activity often exists (eg wet and dry convection days can occur
+under overlapping conditions, Rorig and Ferguson 1999).
+
+--  we should have said that any any precipoation that occurs during seasons when lighting is likely can make ingitions less likely
+
+Cite:
+
+Article{rorig1999characteristics,
+  title={Characteristics of lightning and wildland fire ignition in the Pacific Northwest},
+  author={Rorig, Miriam L and Ferguson, Sue A},
+  journal={Journal of Applied Meteorology},
+  volume={38},
+  number={11},
+  pages={1565--1575},
+  year={1999}
+}
+
+@article{abatzoglou2016controls,
+  title={Controls on interannual variability in lightning-caused fire activity in the western US},
+  author={Abatzoglou, John T and Kolden, Crystal A and Balch, Jennifer K and Bradley, Bethany A},
+  journal={Environmental Research Letters},
+  volume={11},
+  number={4},
+  pages={045005},
+  year={2016},
+  publisher={IOP Publishing}
+}
+
+http://www.publish.csiro.au/wf/WF08117
+
+@article{van2008temporal,
+  title={Temporal and spatial distribution of lightning strikes in California in relation to large-scale weather patterns},
+  author={Van Wagtendonk, Jan W and Cayan, Daniel R},
+  journal={Fire Ecology},
+  volume={4},
+  number={1},
+  pages={34--56},
+  year={2008},
+  publisher={Springer}
+}
 

### 2019-12-11 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Work on edits to introduction
in response to reviews

@@
       \item Moisture content is a strong determinant of forest fuel
-    flammability and varies on small spatial scales and over short time
-    periods. Empirical modeling of equilibrium dead fine fuel moisture is well
-    established, but less is known about the mechanisms by which
-    species-specific litter traits influence litter moisture dynamics.
-    Characterizing how species’ leaf litter retain and release moisture through
-    time is critical to the assessment of fuel’s availability to burn.
-      \item Natural forests are often comprised of mixed stands, leading to
-    mixtures of leaf litter on the surface fuelling the fires. Multiple studies
-    have established the existence of non-additive effects in the flammability
-    of dry litter, but we lack an understanding of moisture synergies in mixed
-    litter and of moisture effects on the flammability of mixed litter.
-      \item The litter from eight species of a mixed-conifer forest was
-    saturated and allowed to dry to determine moisture absorption capacity and
-    drying rates. Burn trials were performed in moist litter beds of single
-    species and of mixtures to establish flammability response to the dry-down
-    process.
+    flammability, however, little is known about the mechanisms by which
+    species-specific litter traits influence litter moisture absorption and
+    retention. Characterizing how species’ leaf litter retain moisture through
+    time and how fuel from multiple species interacts are critical to the
+    assessment of fuel’s availability to burn and key steps to adding species
+    or taxon specific information to fire behavior prediction.
+
+      \item To fill these gaps in knowledge, determined
+    moisture absorption and retention characteristics of the litter from
+    multiple tree species, investigated what litter traits influenced these
+    processes, and tested for nonadditive moisture dynamics and nonadditive
+    flammability in multi-species mixtures. The litter from eight tree species of a
+    mixed-conifer forest was saturated and allowed to dry to determine moisture
+    absorption capacity and drying rates. Burn trials were performed in moist
+    litter beds of single species and of mixtures to establish flammability
+    response to the dry-down process.
+
       \item Moisture dynamics varied across species with higher specific leaf
     area being associated with higher maximum absorption and lower litter bulk
-    density being associated with faster drying rates. Species that produce
-    more aerated litter beds (such as \emph{Q. kelloggii} and \emph{Pinus}
-    species) have faster rates of litter drying.
-      \item We examined two axes of flammability variation (spread rate and
-    fuel consumption) and found that these differing moisture dynamics result
-    in time since wetting having strongly differing effects on flammability
-    across the major litter types typified by different tree taxa.
-      \item We found that litter mixtures exhibited consistent non-additivity
-    in flame spread rate in fuel consumption. This non-additivity became more
-    positive with fuel drying. Litter mixtures burn with fire behavior more
-    similar to that of the most flammable constituent species and this effect
-    increases as fuels dry.
-      \item \emph{Synthesis:} In this temperate mixed conifer forest, the shade
-    intolerant species are favored by the increased fire to which their fuels
-    contribute Such positive feedback effects increases the potential for
-    species to exert community scale effects even when relatively rare and that
-    can influence community assembly and the evolution of niche-constructing
-    traits.
+    density being associated with faster drying rates. We examined two axes of
+    flammability variation and found that these differing moisture dynamics
+    result in time since wetting having strongly differing effects on
+    flammability across the major litter types typified by different tree taxa.
+    Litter mixtures exhibited consistent nonadditivity in flame spread rate and
+    in fuel consumption. This nonadditivity became more positive with fuel
+    drying.
+
+      \item \emph{Synthesis:} Taxon specific patterns of moisture absorption
+    and retention can change flammability rankings under moist compared with dry fuels.
+    Litter mixtures generally burn with fire behavior more similar to that of
+    the most flammable constituent species and this effect increases as fuels
+    dry. [TODO more]
+
   \end{enumerate}
@@
 
-\noindent \textbf{Keywords:} California, fire, flame spread rate, flammability, fuel consumption, fuel moisture, leaf litter, non-additivity
+\noindent \textbf{Keywords:} California, fire, flame spread rate, flammability, fuel consumption, fuel moisture, leaf litter, nonadditivity
 
@@
 Fuel moisture content is an important, and often driving, determinant of
-surface fire behavior. Moisture acts as a heat-sink resulting in less energy
-available for propagation, and the resulting water vapor will both cool the
-flaming front and dilute flammable gases \citep{Albini-1976, Shafizadeh-1977}.
-Environmental conditions such as temperature, relative humidity, and wind drive
-the moisture content of the fuel available to burn
-\citep{Kreye+Varner+etal-2018}. Variation in moisture of living fuels has
-strong effects on flammability, but moisture of dead excised litter fuels can
-also vary. In particular, fine dead fuels respond very readily to environmental
-changes \citep{Nelson-2001}. Fuel moisture influences flammability and fire
-behavior through effects on ignition, spread rate, flame height, fuel
-consumption, and heat release, due to the high specific heat content of water
-\citep{Rothermel-1972, Nelson-2001}. Moisture of litter fuels varies with
-climate, with seasons, and with short term weather within a season. Therefore,
-the rate of drying and flammability as a function of moisture content are key
-parameters that influence fuel flammability throughout the fire season
-\citep{Nelson-2001}.
+surface fire behavior \citep{Rothermel-1972}. Moisture acts as a heat-sink
+resulting in less energy available for propagation, and the resulting water
+vapor will both cool the flaming front and dilute flammable gases
+\citep{Albini-1976, Shafizadeh-1977}. Environmental conditions such as
+temperature, relative humidity, and wind drive the moisture content of the fuel
+available to burn \citep{Kreye+Varner+etal-2018}. Variation in moisture of
+living fuels has strong effects on flammability, but moisture of dead excised
+litter fuels can also vary. In particular, fine dead fuels respond very readily
+to environmental changes \citep{Nelson-2001}. Fuel moisture influences
+flammability and fire behavior through effects on ignition, spread rate, flame
+height, fuel consumption, and heat release, due to the high specific heat
+content of water \citep{Rothermel-1972, Nelson-2001}. Moisture of litter fuels
+varies with climate, with seasons, and with short term weather within a season.
+Therefore, the rate of drying and flammability as a function of moisture
+content are key parameters that influence fuel flammability throughout the fire
+season \citep{Nelson-2001}.
+
+Flammability is a multi-dimensional characteristic of fuels
+\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
+different components of fire behavior. In many forests that experience surface
+fire, litter properties influence fire behavior
+\citep{Ganteaume+Marielle+etal-2011, Schwilk+Caprio-2011,
+  Varner+Kane+etal-2015}. In litter driven fire, the two major axes of
+variation are described by measurements related to flame spread rate and
+measurements related to heat release or heat transfer
+\citep{Magalhaes+Schwilk-2012, Prior+Murphy+etal-2018}. These different
+components of fire behavior have differing ecological effects with spread rate
+being associated with rates of heat release and total burned area and with heat
+transfer to soils or tissues predicting soil impacts of fire and the mortality
+of better insulated plant tissues. A shortcoming in our current knowledge is
+our integration of specific species and plant traits into fuel hazard
+assessment, fire modeling, and fire effects prediction. Developing quantitative
+yet robust and general ways to integrate actual plant communities into fuel
+assessment will require understanding the magnitude of flammability variability
+across species in space and time and parameterizing based on plant
+characteristics. [TODO: get opinion on final sentences]
+
+Species differ in their litter flammability \cite{Fonda+Belanger+etal-1998,
+  Fonda-2001, Scarf+Westoby-2006, Kane+Varner+etal-2008,
+  Cornwell+Elvira+etal-2015}. Litter packing, measured as litter-bed density or
+packing ratio, is an important control ignition and spread in models of surface
+fire spread \citep{Weber-1991, Balbi+Santoni+etal-1999, Morvan+Larini-2001}.
+Multiple studies have demonstrated that litter bed density controls the
+flammability of dry litter fuels in seasonally dry climates where higher litter
+density reduces flame spread rate. Across these studies, there is increasing
+consensus that larger less densely packed litter beds increase oxygen
+availability and burn with faster spread rates \citep{Scarf+Westoby-2006,
+  Schwilk+Caprio-2011, Magalhaes+Schwilk-2012, Cornwell+Elvira+etal-2015,
+  Varner+Kane+etal-2015, Kreye+Varner+etal-2018}. The litter packing is driven
+by litter particle size and shape which is in turn influenced by leaf size, how
+leaves or branches are abscised, and subsequent decomposition
+\citep{Fonda+Belanger+etal-1998, Stephens+Finney+etal-2004,
+  Scarff+Westoby-2006, Kane+Varner+etal-2008, Cornwell+Elvira+etal-2015,
+  Zhao+Cornwell+etal-2016, Grootemaat+Wright+etal-2017,
+  Prior+Murphy+etal-2018}.
+
+These species-specific effects are increasingly well established. However, how
+moisture dynamics modulate species specific flammability is poorly known and
+could be driven by by leaf traits, emergent litter traits, or by both. Species
+specific effects on litter flammability are recorded for the forests of the
+southeastern US \citep{Nowacki+Abrams-2008} where species-specific litter
+drying rates play a role in flammability feedbacks
+\citep{Kreye+Varner+etal-2013}. Fire behavior consequences of individual
+species traits have been recorded for western US forests as well: in mixed
+conifer forest in California, where the fire regime is characterized by surface
+fires fueled by plant litter, overstory tree composition drives variation in
+local fire behavior \citep{Schwilk+Caprio-2011} as a result of differential
+litter flammability across tree species and as a result of synergistic
+interactions in multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}.
+These flammability feedbacks may help explain historical fire frequency
+patterns \citep{Schwilk+Caprio-2011}, but current estimates of litter
+flammability in mixed conifer forests are based largely on burning dry litter.
 
@@
 effects as well. Explicit models of fire behavior as a function of time since
-wetting could improve fire prediction, however.  TODO REWRITE: In many litter fuel driven
-systems, natural lightning ignitions are concurrent with precipitation and,
-therefore, fire behavior can depend on rate of drying following a wetting
-event. If species differ in their drying rate, this may influence
-relatively flammability across species and over time following
-wetting. Such differential moisture content across time may contribute to the
-often high spatial heterogeneity of surface fires
-\citep{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.
+wetting could improve fire prediction, however. Drying rates following snowmelt
+or following rainfall during the fire season will influence the length of the
+fire season, flammability over time, and the probability of sustained ignition,
+higher spread rates and fuel consumption when ignition occurs. That species can
+differ in litter drying rates has been demonstrated for at least one forest
+system \citep{Kreye+Varner+etal-2013}. Such differential moisture content
+across time may contribute to the often high spatial heterogeneity of surface
+fires \citep{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.
 
-Litter packing, measured as litter-bed density or packing ratio, controls the
-flammability of dry litter fuels in seasonally dry climates because higher
-litter density reduces flammability as a consequence of oxygen limitation. Leaf
-litter density is driven by leaf size and shape
-\citep{Fonda+Belanger+etal-1998, Scarff+Westoby-2006, Kane+Varner+etal-2008}.
-These litter particle effects on litter bed aeration, in turn, have effects on
-litter bed flammability \citep{Ganteaume+Marielle+etal-2011,
-  Schwilk+Caprio-2011, Kreye+Varner+etal-2013}. These species-specific effects
-are well established. However, how moisture dynamics modulate species specific
-flammability is poorly known and could be driven by by leaf traits, emergent
-litter traits, or both. Species specific effects on litter flammability
-are recorded for the forests of the southeastern US \citep{Nowacki+Abrams-2008}
-where species-specific litter drying rates play a role in flammability
-feedbacks \citep{Kreye+Varner+etal-2013}.
-Fire behavior consequences of individual species traits have been recorded for western US forests as well: in mixed conifer forest in California, where the fire regime is characterized by surface fires fueled
-by plant litter, overstory tree composition drives variation in local fire
-behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
-flammability across tree species and as a result of synergistic interactions in
-multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
-flammability feedbacks may help explain historical fire frequency patterns, but
-current estimates of litter flammability in mixed conifer forests are based
-largely on burning dry litter.
+Fuel moisture content through time is controlled by two processes, absorption
+and retention \citep{Kreye+Kobziar+etal-2013}. Absorptive capacity represents
+the potential maximum fuel moisture and is evaluated by the moisture content of
+litter when saturated. Differences in moisture content over time since
+saturation can indicate different abilities to retain moisture. Litter particle
+geometry may have direct effects on maximum moisture absorption through
+area-mass relationships and we also expect effects of leaf morphology,
+especially leaf cuticle thickness \citep{Van_Wagner-1969}. Like absorption,
+moisture retention should also be influenced by leaf and litter bed traits:
+surface-area-to-volume ratio and litter bulk density will influence air flow
+through the litter bed. Litter packing therefore effects fire behavior in two
+ways: directly by its effect on oxygen availability for combustion, and
+indirectly by its effect on dry-down rate. Both these pathways are affected by
+leaf particle traits, especially leaf size \citep{Scarff+Westoby-2006}.
+Fuel-beds composed of larger leaves with lower moisture retainability should
+favor drying, thus increasing the probability of ignition and fire spread.
 
-Fuel moisture content through time is controlled by two main processes,
-absorption and retention \citep{Kreye+Kobziar+etal-2013}.
-Absorptive capacity represents the potential maximum fuel moisture and is
-evaluated by the moisture content of litter when saturated. Differences in moisture content over time since saturation can indicate different abilities to retain
-moisture. Litter particle geometry may have direct effects on maximum moisture
-absorption through area-mass relationships and we also expect effects of leaf
-morphology, especially leaf cuticle thickness \citep{Van_Wagner-1969}. Like absorption, mmoisture retention should also be influenced by leaf and litter bed traits:
-surface-area-to-volume ratio and litter bulk density will influence air flow
-through the litter bed. Litter packing therefore effects fire behavior in two ways: directly by its effect on oxygen availability for combustion,
-and indirectly by its effect on dry-down rate. Both these pathways are
-affected by leaf particle traits, especially leaf size
-\citep{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves with lower
-moisture retainability should favor drying, thus increasing the probability
-of ignition and fire spread.
+In mixed species communities, species traits may interact such that the
+flammability of multi-species mixtures is not the average of the constituent
+species. Such nonadditive effects are often synergistic in dry litter fuels
+where the flammability of a mixture is driven by the most flammable species
+\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}, but negative
+nonadditivity has also been recorded for litter mixtures
+\citep{Blauw+Wensink+etal-2015, Zhao+Cornwell+etal-2016}. It is possible that
+similar nonadditive effects may influence litter moisture content directly and
+flammability of moist fuels. There is some evidence for nonadditive
+flammability in moist fuels with one study finding increasing nonadditivity
+with moisture \citep{Blauw+Wensink+etal-2015}.
 
-Flammability is a multi-dimensional characteristic of fuels
-\citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
-different components of fire behavior. In litter driven fire, the two major
-axes of variation are described by measurements related to flame spread rate
-and measurements related to heat transfer [CITATIONS]  . Flame spread rate is easy to measure
-in burning trials, but heat transfer can be more difficult and often proxies
-such as duration of heating are used \citep{Magalhaes+Schwilk-2012,
-  Varner+Kane+etal-2015}. These different components of fire behavior have
-differing ecological effects.
-
-TODO
-
-Species traits may interact such that the
-flammability of multi-species mixtures is not the average of the constituent
-species. Such non-additive effects are often synergistic in dry litter fuels where
-the flammability of a mixture is driven by the most flammable species
-\citep{VanAltena+Logtestjin+etal-2012, Magalhaes+Schwilk-2012}. TODO cite negative flammability. Zhao et al 2019]
-
-It is possible
-that similar non-additive effects may influence litter moisture content,
-
-% especially if different leaf traits influence absorptive capacity and drying
-% rate.
-
-There is some evidence for non-additive flammability in moist fuels with
-one study finding increasing non-additivity with moisture
-\citep{Blauw+Wensink+etal-2015}.
+[TODO: reseearch question here]
 
@@
 maximum water retention and rate of moisture loss and examined how these vary
-in relation to two leaf and litter bed traits [CHECK AND FIX]. We then examined the influence
-of moisture on the flammability across these species. Experimental burn trials
-of leaf litter at various moisture levels determined how time since wetting
-influenced two axes of flammability characterized by flame spread rate and
-total fuel consumed. Previous work described the flammability of these species
-in monocultures and in mixtures for oven-dry litter
-\citep{Magalhaes+Schwilk-2012} and found consistent positive non-additivity of
-flammability where fire behavior of mixtures was driven by the most flammable
-species. Here, we examined if this positive non-additivity was maintained in
-mixtures of varying levels of fuel moisture. The burn trials also
-performed with multi-species litter mixtures to evaluate if there are non-additive effects as
-observed in oven-dried litter. We tested if the non-additive effects on
-reported in previous studies occur under [TODO CHECK]
-ecologically-relevant moisture levels,
-and whether fuel moisture effects strengthen or moderate such non-additivity.
+in relation to two leaf and litter bed traits [CHECK AND FIX]. We then examined
+the influence of moisture on the flammability across these species.
+Experimental burn trials of leaf litter at various moisture levels determined
+how time since wetting influenced two axes of flammability characterized by
+flame spread rate and total fuel consumed. Previous work described the
+flammability of these species in monocultures and in mixtures for oven-dry
+litter \citep{Magalhaes+Schwilk-2012} and found consistent positive
+nonadditivity of flammability where fire behavior of mixtures was driven by the
+most flammable species. Here, we examined if this positive nonadditivity was
+maintained in mixtures of varying levels of fuel moisture. The burn trials also
+performed with multi-species litter mixtures to evaluate if there are
+nonadditive effects as observed in oven-dried litter. We tested if the
+nonadditive effects on reported in previous studies occur under [TODO CHECK]
+ecologically-relevant moisture levels, and whether fuel moisture effects
+strengthen or moderate such nonadditivity.
 
@@
 twigs less than 0.625\,cm diameter (= 1-hour fuel) that have fallen in the
-previous year (mostly undecomposed or lightly decomposed). Leaf litter was collected in the summer
-(mid-June to mid-July) in 2012 from 21 separate sites across the study area. We
-collected from a minimum of four populations per species with each population
-9--32\,km distant from its neighbors. An exception to this was \emph{A.
-  magnifica} which is restricted to the higher elevations and could only be
-collected from one area due to designated wilderness boundaries and logistical
-constraints. Within that area, \emph{A. magnifica} litter was collected from
-over 10 different individuals. For the remaining species, the collection
-involved 2--4 individual trees at each site which were separated from one
-another by at least 10\,m. The litter was collected 2\,m away from the trunk to
-obtain a more uniform sample of leaf and fine branchlet litter, because bark
-and heavier branches tend to fall closer to the tree trunk. The top 2.5-5\,cm
-of litter were collected by raking and shoveling into large bags. All litter
-was air dried to $< 5$\% fuel moisture by dry weight, then weighed and divided into moisture trials. The collected litter was dried in large batches, then
-stored in sealed plastic bags inside cardboard boxes in a lab storage space. We
-minimized handling and moving the bagged litter to avoid mechanical damage and
-changes to the particle size distributions. The moisture manipulations and
-subsequent burning trials were conducted in April--August 2015.
+previous year (mostly undecomposed or lightly decomposed). Leaf litter was
+collected in the summer (mid-June to mid-July) in 2012 from 21 separate sites
+across the study area. We collected from a minimum of four populations per
+species with each population 9--32\,km distant from its neighbors. An exception
+to this was \emph{A. magnifica} which is restricted to the higher elevations
+and could only be collected from one area due to designated wilderness
+boundaries and logistical constraints. Within that area, \emph{A. magnifica}
+litter was collected from over 10 different individuals. For the remaining
+species, the collection involved 2--4 individual trees at each site which were
+separated from one another by at least 10\,m. The litter was collected 2\,m
+away from the trunk to obtain a more uniform sample of leaf and fine branchlet
+litter, because bark and heavier branches tend to fall closer to the tree
+trunk. The top 2.5-5\,cm of litter were collected by raking and shoveling into
+large bags. All litter was air dried to $< 5$\% fuel moisture by dry weight,
+then weighed and divided into moisture trials. The collected litter was dried
+in large batches, then stored in sealed plastic bags inside cardboard boxes in
+a lab storage space. We minimized handling and moving the bagged litter to
+avoid mechanical damage and changes to the particle size distributions. The
+moisture manipulations and subsequent burning trials were conducted in
+April--August 2015.
 
@@
 maximum moisture content. The samples were then allowed to dry in a controlled
-environment kept at $21^\circ$C temperature and 30\% relative humidity, assessed via
-an iButton sensor (Maxim Integrated) placed over the samples. Litter baskets
-were arranged in a completely randomized design. During the dry-down process
-three subsamples of 5--10\,g were taken at 24 hour intervals and weighed. These
-subsamples were then oven dried for 24 hours at $100^\circ$C and re-weighed to assess
-moisture content. This was repeated until the final sample from a basket was
-less than 10\% moisture.
+environment kept at $21^\circ$C temperature and 30\% relative humidity,
+assessed via an iButton sensor (Maxim Integrated) placed over the samples.
+Litter baskets were arranged in a completely randomized design. During the
+dry-down process three subsamples of 5--10\,g were taken at 24 hour intervals
+and weighed. These subsamples were then oven dried for 24 hours at $100^\circ$C
+and re-weighed to assess moisture content. This was repeated until the final
+sample from a basket was less than 10\% moisture.
 
@@
 likely to influence drying rate and litter packing density (litter bulk
-density). [TODO check]  Specific leaf area was measured on freshly fallen litter by sampling
-5 individuals per species. This measurement included petiole and for the
-scale-leaved species included the full excised branchlet. Therefore, this is
-not directly comparable to typical specific leaf area measurements on fresh
-leaves. Litter bulk density was obtained in this experiment for each basket by
-dividing the dry mass (450\,g) by the estimated litter volume (basket width x
-length x litter depth).
+density). [TODO check] Specific leaf area was measured on freshly fallen litter
+by sampling 5 individuals per species. This measurement included petiole and
+for the scale-leaved species included the full excised branchlet. Therefore,
+this is not directly comparable to typical specific leaf area measurements on
+fresh leaves. Litter bulk density was obtained in this experiment for each
+basket by dividing the dry mass (450\,g) by the estimated litter volume (basket
+width x length x litter depth).
 
@@
 time to ignition and duration of flaming combustion. The flammability trials
-followed the methodology detailed in 
-\citet{Magalhaes+Schwilk-2012}.
+followed the methodology detailed in \citet{Magalhaes+Schwilk-2012}.
 
@@
 Some trials resulted only in smoldering combustion; for those trials, a time to
-ignition and fuel consumption were recorded, but flame spread rate and flame height
-were recorded as zero. Our thermocouple system failed or data was lost for two
-days of burning trials, therefore we have omitted those data from this
+ignition and fuel consumption were recorded, but flame spread rate and flame
+height were recorded as zero. Our thermocouple system failed or data was lost
+for two days of burning trials, therefore we have omitted those data from this
 analysis. Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that flame
 spread rate captures one important axis of flammability variation
-\citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal02018} and percent fuel consumed
-captures a portion of the second major axis of flammability variation. Our analysis,
-therefore, focused on these two relatively orthogonal measures of flammability
-to avoid examining multiple redundant dependent variables.
+\citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal02018} and
+percent fuel consumed captures a portion of the second major axis of
+flammability variation. Our analysis, therefore, focused on these two
+relatively orthogonal measures of flammability to avoid examining multiple
+redundant dependent variables.
 
@@
 retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
-($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R 3.6.1
-\citep{RCoreTeam-2019}. We fit models with R packages `lme4'
+($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R
+3.6.1 \citep{RCoreTeam-2019}. We fit models with R packages `lme4'
 \citep{Bates_Machler_etal-2015} and `afex' \citep{Singmann_Bolker_etal-2017}.
@@
 Kenward-Roger approximation \citep{Kenward_Roger-1997} which is recommended by
-Luke \citeyear{Luke-2017} as producing acceptable Type I error rates. We tested if
-one litter particle trait (specific leaf area) and one litter bed trait (litter bulk density)
-influenced moisture dynamics by fitting linear models with species means as
-observations. For each response variable we fit a single linear model with two
-predictors and an interaction term.
+Luke \citeyear{Luke-2017} as producing acceptable Type I error rates. We tested
+if one litter particle trait (specific leaf area) and one litter bed trait
+(litter bulk density) influenced moisture dynamics by fitting linear models
+with species means as observations. For each response variable we fit a single
+linear model with two predictors and an interaction term.
 
@@
 genera in the Cupressaceae, \emph{Calocedrus} and \emph{Sequoiadendron}, also
-behaved similarly to one another. Therefore, to preserve power we used taxonomic
-groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus}) as our
-flammability functional groups and treated these as fixed effects in our
+behaved similarly to one another. Therefore, to preserve power we used
+taxonomic groups (\emph{Abies}, Cupressaceae, \emph{Quercus}, and \emph{Pinus})
+as our flammability functional groups and treated these as fixed effects in our
 models. The effect of fuel moisture and time since wetting on flammability
@@
 linear models with a beta regression link with the `glmmADMB' package
-\citep{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function of
-moisture content and of time since wetting.
+\citep{Skaug_Fournier_etal-2016} to model percent fuel consumed as a function
+of moisture content and of time since wetting.
 
-We investigated possible non-additive effects on moisture and on flammability
-by testing if moisture content and flammability of mixtures were predicted by
-the average of the constituent species. To choose mixtures, we first grouped
-the species according to their maximum water retention and drying rate
-according to estimated marginal means by species using the R package `emmeans'
+We investigated possible nonadditive effects on moisture and on flammability by
+testing if moisture content and flammability of mixtures were predicted by the
+average of the constituent species. To choose mixtures, we first grouped the
+species according to their maximum water retention and drying rate according to
+estimated marginal means by species using the R package `emmeans'
 \citep{Lenth-2019}. This resulted in four groups of species and we then
@@
 tests to test if the centered intercept was different than zero (a measure of
-overall non-additivity) and if the slope with time since wetting was different
-from zero (indicating non-additivity changing with drying).
+overall nonadditivity) and if the slope with time since wetting was different
+from zero (indicating nonadditivity changing with drying).
 
@@
   concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa} (SI Fig. 1). There are
-also differences in drying rates across the eight species and estimated
-mixed linear model slopes discriminated species into three groups from lowest
-to most rapid: 1) \emph{C. decurrens}, 2) \emph{Abies spp} and \emph{S.
-  giganteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp} (SI Fig. 2).
+also differences in drying rates across the eight species and estimated mixed
+linear model slopes discriminated species into three groups from lowest to most
+rapid: 1) \emph{C. decurrens}, 2) \emph{Abies spp} and \emph{S. giganteum}, and
+3) \emph{Q. kelloggii} and \emph{Pinus spp} (SI Fig. 2).
 
@@
 and the rate of change with moisture content varied across the taxonomic groups
-(Fig. \ref{fig:consume_moist}, Table \ref{tab:consume_anova}). Although
-the overall ranking of groups was similar for fuel consumption and for flame
-spread rate, in contrast to the pattern with spread rate, for fuel consumption,
+(Fig. \ref{fig:consume_moist}, Table \ref{tab:consume_anova}). Although the
+overall ranking of groups was similar for fuel consumption and for flame spread
+rate, in contrast to the pattern with spread rate, for fuel consumption,
 \emph{Pinus} and \emph{Abies} has a steeper decline in flammability with fuel
@@
 generalized linear mixed beta regression model as indicated by the nested model
-comparisons in Table \ref{tab:consume_anova}. As with flame spread rate
-some pines had the most rapid response to drying. However, unlike in the other
+comparisons in Table \ref{tab:consume_anova}. As with flame spread rate some
+pines had the most rapid response to drying. However, unlike in the other
 flammability analyses, here the pines did not all behave similarly: \emph{P.
   lambertiana} had lower fuel consumption than did \emph{P. ponderosa} and
-\emph{P. jeffreyi}. We noticed this during trials and in model fitting we
-found that placing \emph{P. lambertiana} in its own group allowed model
-convergence that was otherwise impossible. This was at odds with our intent to
-use only \emph{a priori} taxonomic groups, but we could not obtain model
-convergence otherwise. This result should therefore be seen as preliminary
-because the test does not reflect an \emph{a priori} hypothesis.
+\emph{P. jeffreyi}. We noticed this during trials and in model fitting we found
+that placing \emph{P. lambertiana} in its own group allowed model convergence
+that was otherwise impossible. This was at odds with our intent to use only
+\emph{a priori} taxonomic groups, but we could not obtain model convergence
+otherwise. This result should therefore be seen as preliminary because the test
+does not reflect an \emph{a priori} hypothesis.
 
@@
 
-\subsection*{Non-additivity in  litter mixtures}
+\subsection*{Nonadditivity in  litter mixtures}
 
@@
 
-There was evidence for non-additive effects in drying rate. Mixtures tended to
+There was evidence for nonadditive effects in drying rate. Mixtures tended to
 have lower water content than predicted from the mean of the constituent
@@
 
-Flammability measures also demonstrated non-additivity in mixtures. Flame
+Flammability measures also demonstrated nonadditivity in mixtures. Flame
 spread rate was consistently higher than predicted (Fig
@@
 intercept for observed-predicted values greater than zero, $P$ = 0.017) and the
-positive non-additivity increased with time since wetting (slope greater than
+positive nonadditivity increased with time since wetting (slope greater than
 zero, $P$ = 0.005). Fuel consumption, however, showed no overall bias towards
-positive or negative non-additivity, but non-additivity increased with time
+positive or negative nonadditivity, but nonadditivity increased with time
 since wetting (Fig. \ref{fig:mixture_obs_pred_consume}, slope not equal to
@@
 The genera in this study displayed different patterns of both moisture
-absorption and desorption, associated with different litter traits. In
-this sample of species, those that have a greater absorptive capacity are also
-among the most flammable in terms of spread rate when dry. This highlights the
+absorption and desorption, associated with different litter traits. In this
+sample of species, those that have a greater absorptive capacity are also among
+the most flammable in terms of spread rate when dry. This highlights the
 importance of fuel moisture in modulating flammability. One shortcoming of past
@@
 tested only dry fuels. We therefore expected that rankings of species
-flammability might change under more [TODO FIX] realistic moisture levels. For example,
-one presumption of ours prior to this experiment was that the high flammability
-previously reported for \emph{Q. kelloggii} \citep{Magalhaes+Schwilk-2012} would
-disappear under higher fuel moisture contents. This prediction was partly
-supported: \emph{Q. kelloggii} fails to ignite until after 72 hours of drying
-in this study (Figs. \ref{fig:spread_time} and \ref{fig:consume_time}).
-However, high moisture content in this oak was unable to suppress ignition of
-mixtures to which it contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
+flammability might change under more [TODO FIX] realistic moisture levels. For
+example, one presumption of ours prior to this experiment was that the high
+flammability previously reported for \emph{Q. kelloggii}
+\citep{Magalhaes+Schwilk-2012} would disappear under higher fuel moisture
+contents. This prediction was partly supported: \emph{Q. kelloggii} fails to
+ignite until after 72 hours of drying in this study (Figs.
+\ref{fig:spread_time} and \ref{fig:consume_time}). However, high moisture
+content in this oak was unable to suppress ignition of mixtures to which it
+contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
 \ref{fig:mixture_obs_pred_consume}).
@@
 drying rate. This has consequences for flammability as a function of species
-and time since wetting. Our flammability trials demonstrate that both
-flame spread rate and fuel consumption increased with time since wetting but
-that the rate of increase varied across these taxa. Only the pines are able to
-ignite and support flame spread immediately after wetting (Fig.
+and time since wetting. Our flammability trials demonstrate that both flame
+spread rate and fuel consumption increased with time since wetting but that the
+rate of increase varied across these taxa. Only the pines are able to ignite
+and support flame spread immediately after wetting (Fig.
 \ref{fig:spread_time}). In contrast, \emph{Quercus} exhibited a threshold
@@
 ignition is possible, having faster spread rates than the other taxa. The
-Cupressaceae, which in our two species have dense litter but low maximum water absorption had low
-flammability across moisture levels.
-This result is consistent with the low flammability found in an Australian Cupressaceae that have small litter particles and dense packing citep{Scarff+Westoby-2006, Bowman+Haverkamp-2018}.
-However, the densely packed litter is not a consistent trait across the family because \citet{Cornwell+Elvira+etal-2015} found relatively high litter flammability and low litter density for some Asian Cupressaceae.
+Cupressaceae, which in our two species have dense litter but low maximum water
+absorption had low flammability across moisture levels. This result is
+consistent with the low flammability found in an Australian Cupressaceae that
+have small litter particles and dense packing citep{Scarff+Westoby-2006,
+  Bowman+Haverkamp-2018}. However, the densely packed litter is not a
+consistent trait across the family because \citet{Cornwell+Elvira+etal-2015}
+found relatively high litter flammability and low litter density for some Asian
+Cupressaceae.
 
@@
 but we expect that decomposition will affect moisture dynamics. For example,
-Van Wagner \citeyear{Van_Wagner-1969}, investigating fuel weathering, found that
-pine needles had a variation in moisture content 3 to 6 times greater when they
-lacked their protective wax coating when compared to intact needles.
+Van Wagner \citeyear{Van_Wagner-1969}, investigating fuel weathering, found
+that pine needles had a variation in moisture content 3 to 6 times greater when
+they lacked their protective wax coating when compared to intact needles.
 
@@
 behavior types observed in single species trials. Mixtures that contained
-\emph{Quercus} absorbed more water during wetting, and the one mixture
-without any Cupressaceae had the most rapid drying rate (SI Figure 4 and SI
-Table 8). Drydown rates, however, did not behave as the average of the
-constituent species: mixtures tended to be drier than expected early in the
-process and wetter than expected later (Fig. \ref{fig:mixture_obs_pred_mc}).
-The fact that mixtures tended to have lower water content than predicted from
-the mean of the constituent species after wetting, but higher water content
-than expected after long drying is surprising because past work has shown that
-mixtures tend to pack less densely than do litter monocultures \citep[][Table
+\emph{Quercus} absorbed more water during wetting, and the one mixture without
+any Cupressaceae had the most rapid drying rate (SI Figure 4 and SI Table 8).
+Drydown rates, however, did not behave as the average of the constituent
+species: mixtures tended to be drier than expected early in the process and
+wetter than expected later (Fig. \ref{fig:mixture_obs_pred_mc}). The fact that
+mixtures tended to have lower water content than predicted from the mean of the
+constituent species after wetting, but higher water content than expected after
+long drying is surprising because past work has shown that mixtures tend to
+pack less densely than do litter monocultures \citep[][Table
 2]{Magalhaes+Schwilk-2012}. If packing density controls rate of drying, then we
@@
 with the results obtained for the individual species, but there were
-significant non-additive effects. Mixtures containing \emph{Q. kelloggii} and
+significant nonadditive effects. Mixtures containing \emph{Q. kelloggii} and
 \emph{P. jeffreyi} tended to have fast spread rates. These results are likely
 influenced by the lower bulk densities of litter mixtures containing those two
-species. Most interestingly, there were patterns of non-additivity in
-spread rate and in fuel consumption for the four mixtures, indicating that
-positive non-additivity in flammability is not restricted to the completely dry
-litter mixtures studied in previous work \citep{Magalhaes+Schwilk-2012,
-  VanAltena+Logtestjin+etal-2012}. Non-additivity in both spread rate and fuel
+species. Most interestingly, there were patterns of nonadditivity in spread
+rate and in fuel consumption for the four mixtures, indicating that positive
+nonadditivity in flammability is not restricted to the completely dry litter
+mixtures studied in previous work \citep{Magalhaes+Schwilk-2012,
+  VanAltena+Logtestjin+etal-2012}. Nonadditivity in both spread rate and fuel
 consumption became more positive with time since wetting and these patterns
@@
 than that reported by Blauw et al \citeyear{Blauw+Wensink+etal-2015}. Blauw et
-al found that non-additivity was greater in moist mixtures of heathland fuels
-than in dry fuels. In their work, non-additivity was generally negative but
-cases of positive non-additivity in mixtures was often a result of mixtures
-meeting ignition thresholds not met by the constituent species alone [TODO: check wording]. To limit
-the number of models fit, we limited our investigation to two flammability
-parameters that represent two major axes of flammability variation. Therefore,
-our study does not fully capture ignition effects although we do have some
-results consistent with theirs: mixtures with \emph{Q. kellogii} could ignite
-at times since wetting at which the species in monoculture could not.
+al found that nonadditivity was greater in moist mixtures of heathland fuels
+than in dry fuels. In their work, nonadditivity was generally negative but
+cases of positive nonadditivity in mixtures was often a result of mixtures
+meeting ignition thresholds not met by the constituent species alone [TODO:
+check wording]. To limit the number of models fit, we limited our investigation
+to two flammability parameters that represent two major axes of flammability
+variation. Therefore, our study does not fully capture ignition effects
+although we do have some results consistent with theirs: mixtures with \emph{Q.
+  kellogii} could ignite at times since wetting at which the species in
+monoculture could not.
 
-Overall, we report a consistent pattern of increasingly positive
-non-additivity in flammability with litter drying across two largely orthogonal
-measures of flammability \citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal02018}. This result, together with previously published work,
-suggests that litter flammability is prone to non-additive effects, suggesting
-caution in lumping species into fuel categories. Community flammability is
-dependent on and particular to its constituent species and cannot be addressed
-by simply averaging the flammability of all species included. In litter-driven
-fire, positive non-additivity appears to be the rule, and the relationships we
-described here can form a working hypothesis and baseline expectation to adjust
-fire behavior predictions in real communities: in mixtures with fuel moisture
-content below 50\%, observed flame spread rates approached double the predicted
-value based on monoculture trials (Fig. \ref{fig:mixture_obs_pred_spread}). In
-completely dry litter from this same system, past work demonstrates observed
-spread rates approached three times the expectation
-\citep{Magalhaes+Schwilk-2012}.
+Overall, we report a consistent pattern of increasingly positive nonadditivity
+in flammability with litter drying across two largely orthogonal measures of
+flammability \citep{Schwilk-2015, Pausas+Keeley+etal-2017,
+  Prior+Murphy+etal02018}. This result, together with previously published
+work, suggests that litter flammability is prone to nonadditive effects,
+suggesting caution in lumping species into fuel categories. Community
+flammability is dependent on and particular to its constituent species and
+cannot be addressed by simply averaging the flammability of all species
+included. In litter-driven fire, positive nonadditivity appears to be the rule,
+and the relationships we described here can form a working hypothesis and
+baseline expectation to adjust fire behavior predictions in real communities:
+in mixtures with fuel moisture content below 50\%, observed flame spread rates
+approached double the predicted value based on monoculture trials (Fig.
+\ref{fig:mixture_obs_pred_spread}). In completely dry litter from this same
+system, past work demonstrates observed spread rates approached three times the
+expectation \citep{Magalhaes+Schwilk-2012}.
 
@@
 
-Positive non-additivity, whereby more flammable species have effects on fire
+Positive nonadditivity, whereby more flammable species have effects on fire
 behavior disproportionate to their contribution to the litter layer, could lead

### 2019-12-13 ms/response_to_reviews.tex
Progress towards R1

@@
 
+We agree that our prior Discussion was weak and have extensively rewritten it
+
+We argue that dry down rates are of strong ecological importance.
 TODO.
+Our dry down experiment occured under conduitions of very low humidity.
+We would expect the rates we observe in the lab to be faster than those in nature.
+But the differences in rates across species, the mixture effects and the resulting flammability are all important contributions of this and should represent relative differneces under field conditions.
+
+There are very few studeis characterizing drying patterns across speceis, only one ... Kreye.
 
@@
 \begin{quote}
-I support the AE's recommendation to reject but allow resubmission.
-However, the bar for a resubmission will be high.
-My feeling is that your study confirms the essential findings from your 2012 study published in Journal of Ecology with the addition that your methodology of drying down adds some realism.
-Indeed, the confirmatory nature of this study was pointed out by reviewer 2 in their point 6.
-In addition to the points raised by the AE and reviewers, I will be looking in any resubmission for more evidence that this study advances your earlier work (apart from your last paragraph, your discussion is largely an explanation of your results with little context from other studies), that the findings have real ecological relevance (i.e., does it matter to the mixed-conifer communities?) and how is the new dry down treatment related to dry down conditions in litter mixtures in the field.
-If you can adequately address all of these concerns then I'll be happy to consider a resubmission.
+First, apologies to the authors for the long delay in the review process.
+Because of the rather contrasting assessments of the first two reviewers, I was waiting for the third reviewer to produce his/her review, but unfortunately we waited in vain.
+So I have to make a recommendation based on these two contrasting assessments plus my own judgement.
+On the one hand, in agreement with reviewer 1, I can see the novelty of this study.
+So far non-additivity in litter mixture flammability has been studied either at air-dry litter moisture or at fixed levels of litter moisture (e.g.
+Blauw et al 2015) and I do applaud the approach of bringing more field realism into these studies by allowing mixtures to dry out at different rates from expected values based on their single components.
+Having said that, there are also major problems with the current manuscript, as reported especially by reviewer 2 (and some additional ones by myself, see below).
+Besides these major problems there is a rather high degree of inaccuracy in the text.
+Therefore, given the high rejection rates of often even good manuscripts to J.
+Ecol., I have to make a negative recommendation for this manuscript.
+However, given the novelty and principally nice design of the experiment, I would allow the authors to submit a strongly revised new manuscript, addressing all major and minor comments by the reviewers and myself, as a new submission.
+(At least one of the reviewers to check the revision would be new.) In addition to the points raised by the reviewers, I like to add some of my own:
 \end{quote}
 
-Thank you. Add context. Yes, several comments converged on the absolutely correct criticism that our discussion did not engage adequately with the literature.  Wer have added X and X...
+Thank you.
+Yes, several comments converged on the absolutely correct criticism that our discussion did not engage adequately with the literature.
+We have made substantial changes to the Introduction and the Discussion adding more context and citing more of the recent literature.
+We reorganized the Introduction and now do not rush the reader so quickly through the ideas.
 
-inclduing cite Zhoa 2019
+[TODO]
 
@@
 
-We have added a figure showing particle size distributions for each of these species. Yes, Abies litter is a a combination of still attached needes and those that have detached from the branchlet.
+We have added information to the methods explaining the collection in more detail.
+All material was naturally dropped litter which could include fine twigs and leaves.
+Litter in these two \emph{Abies} species is a combination of branchlets with still attached needles and individual needles that have detached from the branchlet and added reference to the Zhao et al 2019 results when discussing the mixture results
 
-We
+We have added a figure (SI fig.
+1) to the SI materials showing litter particle size (containing prism volumes) distributions for each of these species.
+These data were collected from the stored litter.
+Therefore, the data show the particle size distributions of the material that was burned.
+We did not add these data to the main figures as they are not part of any statistical test or main conclusions, but we are happy to do so if the editor suggests.
 
@@
 
-Guided by the drying rates measured in the initail trials, we staggered saturation times across species aiming to allow completely blocked design in time.
-This staggering was mostly successful but Because some ignition attempts were unsuccessful, some days lacked the full complemnt of species while others later in the porcess sometimes had more than one replicate per species burned.
-
-TODO
+Guided by the drying rates measured in the initial trials, we staggered saturation times across species aiming to allow completely blocked design in time.
+Because some ignition attempts were unsuccessful, some days lacked the full complemnt of species while others later in the process sometimes had more than one replicate per species burned.
+We've added this information to the text [Lines X-X]
 
@@
 
-We have added detail as requested by reviewer 2.
-See lines (X-X).
-The litter was dried in the lab (very low hmidity, but not oven dried), for several weeks on open racks.
-Then the litter was bagged and sealed and stored within garbage bags inside cardboard boxes in a lab storage area.
-We saw no signs of mold or ongoing decomposition.
-We minimized handling of the bags to avoid mechanical damage and changes to little particle size.
+We have added detail as requested by reviewer 2. See lines X-X. The litter was initially oven dried in large paper bags, then stored in the the lab (very low humidity), for several weeks on
+open racks. Then the litter was bagged and sealed and stored within garbage
+bags inside cardboard boxes in a lab storage area. We saw no signs of mold or
+ongoing decomposition. We minimized handling of the bags to avoid mechanical
+damage and changes to little particle size.
 
@@
 
-TODO.  Mention.  Maybe due to very small samples that miminimize effects of packing in Cornwell et al 2015?
+Good point and another place where we did not discuss the literature fully. We
+have now added this to the discussion and made the conjecture that these
+differing patterns may result from different branchelt structure across the
+Cupressaceae with Australian \emph{Callitris} and the two Californian species
+packing more tightly and resulting in less flammable litter than the Asian
+Cupressaceae in Cornwell et al (2015).
 
@@
 
-We thank the reviewer for their comments and apologize for the failure in proofreading.
-The reviewer should not have had to do such copy editing and I I sincerely apologize for that.
-We have taken heed of both reviewers' comments and fixed all specific grammar errors and fixed the issue we had with missing symbols.
-For brevity, we've omited response to the very simple suggested grammar fixes and simply made all requested changes.
-We respond to other points individually, below
+We thank the reviewer for their comments and apologize for the failure in
+proofreading. The reviewer should not have had to conduct copy editing and we
+sincerely apologize for that. We have taken heed of the reviewers' comments and
+fixed all specific grammar errors and missing symbols. For brevity, we've
+omitted responses to the very simple suggested grammar fixes and simply made all
+requested changes. We respond to other points individually.
 
@@
 
-TODO.
-
-This portion was poorly written.
-We have rewritten this in response to reviewer 2's comment that most fires (or, at elast burned area) result from dry lightning strikes across the west.
-See our responses below.
-
-TODO
+This portion was poorly written. We have rewritten this in response to reviewer 2's comment that most fires result from dry lightning strikes across the west. See our responses to Reviewer 2, below.
 
@@
 
-Agreed. fixed.
+Agreed and fixed.
 
@@
 Thank you. This was an oversight. We have also added more context from the
-literature to our Discussion
-
-TODO
+literature to our Discussion. See lines XX-XX.
 
 \begin{quote}
-lines 243/244 I was a bit puzzled here -It would help to be more specific here - which aspects of moisture dynamics? Which response variables did you use?  (or 'as described below')
+lines 243/244 I was a bit puzzled here --- It would help to be more specific here --- which aspects of moisture dynamics? Which response variables did you use?  (or 'as described below')
 \end{quote}
 
-TODO. We have rewritten that section to be more precise.
+We have rewritten that section to be more precise.  See lines XX.
 
@@
 
-True. We have edited this as suggested.
+We have edited this as suggested.
 
@@
 
-TODO
+TODO!
 
@@
 
-TODO.
+TODO. Rewrite.
 
@@
 
-Agreed. Fixed.
+Agreed. Fixed. CHECK!
 
@@
 We have added some better discussion of past results and included discussion of these papers, as well as contrasting with the Cornwell et al 2015 result as pointed out by the Associate Editor.
+See lines XX-XX.
 
@@
 
-TODO
+Done. See line XX.
 
@@
 
-We agree this does lack consistency and an early draft used the same concise Fig 6 format for all of them.
-But we decided that having time on the x-axis was worth the space and allows the reader more insight into the
-
+We agree this does lack consistency and an early draft used the same concise format as Fig 6 format for all three figures.
+However, we decided that having time on the x-axis was worth the space and allows the reader more insight into the flammability effects of drying over time.
 Reviewer 2 pointed out some exceptions to the overall statistical pattern and those exceptions would not have been easily visible without an explicit time axis.
+There is a tension here between illustrating simply the results of the statistical model and allowing the reader to get a better idea of variation in flammability consequences of drying time.
 
@@
 
-TODO
-
-Yes. We have added citations here.
+Yes. That was an oversight. We have added citations here. See lines XX-XX
 
@@
 
-Yes, this sentence was poorly written and oversimplified. We thank the reviewer
+This sentence was poorly written and oversimplified. We thank the reviewer
 for pointing this out and we have rewritten this section. However, our badly
@@
 thunderstorm activity often exists (eg wet and dry convection days can occur
-under overlapping conditions, Rorig and Ferguson 1999).
-
---  we should have said that any any precipoation that occurs during seasons when lighting is likely can make ingitions less likely
+under overlapping conditions, Rorig and Ferguson 1999).  We've rewritten this section, see lines XX-XX.
 
@@
 
-TODO.
+We have edited the abstract to make the research questions clearer. Yes, our first submission jumped from stating current reserach gaps to specific methods too quickly.  We added X X
 
@@
 
-Yes, this is a good point and we have added some discussion of this.  We tried to lmit our statements, hwoever, to those supported by the statistical tests, and .....
-
-TODO.
+Yes, this is a good point and we have added some discussion of this.
+We tried to lmit our statements to those supported by the statistical tests (Table X), and therefore focused on the positive effect of drying and the overall trend for positive nonadditivity.
+This, of course, obscures some details which are worth mentioning and we have edited this section in response to the reviewer's criticism.
+We agree that the high variability in nonadditivity across mixtures deserves highlighting.
+Please see our changes on lines X-X.
 
@@
 
-We have made extensive edits in response to both reviewer's suggestions
+We have made extensive edits in response to both reviewer's suggestions.
 
@@
 
-TODO
+We have changed the title.
 
@@
 
-We have rewritten much of the discussion to better include findings from the literature.....
+TODO
+
+We have rewritten much of the discussion to better include findings from the literature and rewritten the concluding paragraph.
 
@@
 
-TODO
+We have editing the abstract to more clearly state the research questions reduce its length.
 
@@
 
-TODO
-
-We disagree in this case.
+OK, w thought this established enough to not need citation but have added one.
 
@@
 
-TODO
+We have rearranged the logic of the introduction as suggested.
 
@@
 
-TODO
+We have rewritten this section and removed that phrase.
 
@@
 
+We have rewritten this for clarity. See line XX.
+
 \begin{quote}
@@
 \end{quote}
+
+Fixed.
 
@@
 
+We were not sure what the reviewer's criticism was here but have rewritten this system to be more detailed and precise.
+
 \begin{quote}
@@
 \end{quote}
+
+Good point. Fixed and in another sentence elsewhere.
 
@@
 
+Fixed.
+
 \begin{quote}
@@
 \end{quote}
+
+We have rewritten this. See line X.
 
@@
 
+We are not sure of the criticism here.
+
 \begin{quote}
@@
 \end{quote}
+
+We have broken these ideas into multiple paragraphs.
 
@@
 
+Yes, excellent point. We have added detail and citations here (lines X-X).
+
 \begin{quote}
@@
 \end{quote}
+
+TODO?
 
@@
 
+TODO.
+
 \begin{quote}
@@
 
+TODO
+
+We have rewritten this. See line X.
+
 \begin{quote}
@@
 \end{quote}
+
+TODO
+
+We assume that the reviewer's complaint is that ``ecologically-relevant'' is vague.
 
@@
 
+Symbols fixed throughout.
+
 \begin{quote}
@@
 \end{quote}
+
+Yes. We have added some detail to this and also now provide particle size distributions for each species measured on the actual stored litter (SI fig 1).
 
@@
 
+We have added information to the methods explaining the collection in more detail. See lines XX and our response to the Associate Editor, above.
+
 \begin{quote}
-Line 157: is it possible that all litter has less than 5% fuel moisture when air-dried?
+Line 157: is it possible that all litter has less than 5\% fuel moisture when air-dried?
 \end{quote}
+
+We have corrected the methods because the litter was initially oven dried in large paper bags, then stored on racks in the low humidity lab.
+Our subsample measurements showed less than 5\% moisture.
+See lines X-X.
 
@@
 \end{quote}
+
+We have added detail on the long term dry storage. See lines X-X.
 
@@
 
+Symbols fixed.
+
 \begin{quote}
@@
 
+Agreed, fixed.
+
 \begin{quote}
@@
 \end{quote}
+
+No, mass was fixed, width and length of the burning tray was fixed (15\,cm by 1.5\,m), and depth was allowed to vary.  We edited this for clarity. See lines X-X
 
@@
 
+TODO check
+
 \begin{quote}
@@
 
+Flammability is a multivariate trait. To avoid collinearity and to simplify
+models, two general approaches are possible: either reduce and transform the
+response variables and create synthetic variables through a method such as PCA,
+or pick a few variables representative of the axes of variation. When the
+experiment was designed, we had intended to use the first method. The loss of
+several days of thermocouple data made that an unattractive and unnecessary
+option, so we opted for the simpler to explain method of picking two traits
+already known to correlate with different axes of litter flammability. I envy
+the scientist who has never dealt with unforeseen events.
+
 \begin{quote}
@@
 \end{quote}
+
+TODO
 
@@
 
+Although we are unsure of the criticism here, we have rewritten the sentence.
+
 \begin{quote}
@@
 \end{quote}
+
+Yes.  This is not surprising in the case of these three pines where one has quite different needles. But our \emph{a priori} groups were based on genus.
 
@@
 
+TODO
+
 \begin{quote}
@@
 \end{quote}
+
+Yes, oops a typo. We have rewritten the heading. Line X
 
@@
 
+Agreed. That was vague. Fixed.
+
 \begin{quote}
@@
 \end{quote}
+
+We removed the offending word.
 
@@
 
+We don't understand the criticism. Perhaps it is the word ``alone'' that is ambiguous?  We changed ``alone'' to ``in monoculture'' although we thought the original wording was clear.
+
 \begin{quote}
@@
 \end{quote}
+
+We don't understand the criticism. Academic writing commonly uses present tense to describe results that still apply. There is no grammatical error.
 
@@
 
+TODO
+
 \closing{Sincerely,}

### 2019-12-13 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Progress towards R1

@@
 
-      \item To fill these gaps in knowledge, determined
+      \item To fill these gaps in knowledge, we determined
     moisture absorption and retention characteristics of the litter from
@@
 Species differ in their litter flammability \cite{Fonda+Belanger+etal-1998,
-  Fonda-2001, Scarf+Westoby-2006, Kane+Varner+etal-2008,
+  Fonda-2001, Scarff+Westoby-2006, Kane+Varner+etal-2008,
   Cornwell+Elvira+etal-2015}. Litter packing, measured as litter-bed density or
@@
 consensus that larger less densely packed litter beds increase oxygen
-availability and burn with faster spread rates \citep{Scarf+Westoby-2006,
+availability and burn with faster spread rates \citep{Scarff+Westoby-2006,
   Schwilk+Caprio-2011, Magalhaes+Schwilk-2012, Cornwell+Elvira+etal-2015,
@@
   Zhao+Cornwell+etal-2016, Grootemaat+Wright+etal-2017,
-  Prior+Murphy+etal-2018}.
+  Prior+Murphy+etal-2018, Zhao+vanLogtestijn+etal-2019}.
 
@@
 nonadditivity has also been recorded for litter mixtures
-\citep{Blauw+Wensink+etal-2015, Zhao+Cornwell+etal-2016}. It is possible that
+\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible that
 similar nonadditive effects may influence litter moisture content directly and
@@
 nonadditive effects on reported in previous studies occur under [TODO CHECK]
-ecologically-relevant moisture levels, and whether fuel moisture effects
+higher moisture levels, and whether fuel moisture effects
 strengthen or moderate such nonadditivity.
@@
 trunk. The top 2.5-5\,cm of litter were collected by raking and shoveling into
-large bags. All litter was air dried to $< 5$\% fuel moisture by dry weight,
-then weighed and divided into moisture trials. The collected litter was dried
-in large batches, then stored in sealed plastic bags inside cardboard boxes in
-a lab storage space. We minimized handling and moving the bagged litter to
-avoid mechanical damage and changes to the particle size distributions. The
-moisture manipulations and subsequent burning trials were conducted in
-April--August 2015.
+large bags. All litter was oven dried to $< 5$\% fuel moisture by dry weight,
+then weighed and divided into moisture trials. The collected litter was stored
+in large paper bags and cardboard boxes in a lab storage space. We minimized
+handling and moving the bagged litter to avoid mechanical damage and changes to
+the particle size distributions. The moisture manipulations and subsequent
+burning trials were conducted in April--August 2015.
 
@@
 likely to influence drying rate and litter packing density (litter bulk
-density). [TODO check] Specific leaf area was measured on freshly fallen litter
+density). Specific leaf area was measured on freshly fallen litter
 by sampling 5 individuals per species. This measurement included petiole and
@@
 width x length x litter depth).
+To examine litter particle size distributions in the stored litter, we subsampled 10 approximately 2\,g samples per species from the stored litter.
+Each individual particle of that subsample was measured for length (longest axis), width (longest axis perpendicular to length) and thickness (perpendicular to length and width, at midrib) using an electronic caliper.
 
@@
 trials all used 450\,g (dry weight) of litter. For three-species mixtures, this
-was 150\,g per species. [TODO give depths] This design allows fire to reach
-constant flame spread rates \citep{Magalhaes+Schwilk-2012}. Two graduated metal
-rulers equally spaced along the apparatus allowed visual determination of
-maximum flame height. Litter was ignited with a propane torch; maximum
-adiabatic flame temperature of 1899$^\circ$C,
+was 150\,g per species. Therefore, fuel mass and burning tray length and width
+were fixed, but fuel depth was allowed to vary. We recorded the actual fuel
+depth to calculate litter bed density before burning. This design allows fire
+to reach constant flame spread rates \citep{Magalhaes+Schwilk-2012}. Two
+graduated metal rulers equally spaced along the apparatus allowed visual
+determination of maximum flame height. Litter was ignited with a propane torch;
+maximum adiabatic flame temperature of 1899$^\circ$C,
 \url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the apparatus
@@
 spread rate captures one important axis of flammability variation
-\citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal02018} and
+\citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal-2018} and
 percent fuel consumed captures a portion of the second major axis of
@@
 (Nielsen-Kellerman, Boothwyn, PA). All trials were conducted on clear days
-between 10 am and 3 pm.
+between 10 am and 3 pm. We scheduled burns in a randomized blocked design
+aiming to burn one replicate of each species or mixture on single burn day.
+This required scheduling species saturation and during start based on the dry
+down curves.
 
@@
 Dry-down curves were created based on the drying experiment, fitted with an
-exponential decay curve to each species.:
+exponential decay curve to each species:
 
@@
 
-There were marked differences in the extent of moisture species retain
-initially following wetting and draining by gravity (all $P < 0.0001$, Fig.
-\ref{fig:drydown} and Table \ref{tab:drydown}). \emph{Q. kelloggii}, the only
-broadleaf, is the species that retained the most moisture initially.
-\emph{Quercus}, like the pines, had a rapid drying rate (Fig. \ref{fig:drydown}
-and SI Fig. 1). Calculating estimated marginal means following the linear
-mixed-effects model discriminated the species into three different maximum
-water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P. lambertiana} and
-\emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S. giganteum}, \emph{A.
-  concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa} (SI Fig. 1). There are
-also differences in drying rates across the eight species and estimated mixed
-linear model slopes discriminated species into three groups from lowest to most
-rapid: 1) \emph{C. decurrens}, 2) \emph{Abies spp} and \emph{S. giganteum}, and
-3) \emph{Q. kelloggii} and \emph{Pinus spp} (SI Fig. 2).
+Past work has demonstrated that green leaf width predicts litter packing and,
+consequently, flame spread rate across these species. Litter particle size
+distributions, however show that there is substantial variation in particle
+sizes within a species (SI fig. 1). There were marked differences in the extent
+of moisture species retain initially following wetting and draining by gravity
+(all $P < 0.0001$, Fig. \ref{fig:drydown} and Table \ref{tab:drydown}).
+\emph{Q. kelloggii}, the only broadleaf, is the species that retained the most
+moisture initially. \emph{Quercus}, like the pines, had a rapid drying rate
+(Fig. \ref{fig:drydown} and SI Fig. 2). Calculating estimated marginal means
+following the linear mixed-effects model discriminated the species into three
+different maximum water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P.
+  lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S.
+  giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa}
+(SI Fig. 2). There are also differences in drying rates across the eight
+species and estimated mixed linear model slopes discriminated species into
+three groups from lowest to most rapid: 1) \emph{C. decurrens}, 2) \emph{Abies
+  spp} and \emph{S. giganteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp}
+(SI Fig. 3).
 
@@
 Across these eight species specific leaf area had a positive association with maximum
-water retention (SI Fig 4, linear model $P$ = 0.003) and litter
-bulk density had a negative effect on desiccation rate (SI Fig 5,
+water retention (SI Fig 5, linear model $P$ = 0.003) and litter
+bulk density had a negative effect on desiccation rate (SI Fig 6,
 linear model $P$ = 0.029). ANOVA results are shown in Tables
@@
 Based on visual inspection of the dry down curves and the estimated marginal
-means for maximum water retainability and desiccation rate (SI Figs 2--3), we
+means for maximum water retainability and desiccation rate (SI Figs 3--4), we
 selected four species to represent the full range of litter moisture dynamics
@@
 Mixture dry down rates and maximum water retention showed some relationship to
-the dry down curves for individual constituent species (see SI Fig. 6). For
+the dry down curves for individual constituent species (see SI Fig. 7). For
 example, the mixture that does not contain \emph{Q. kelloggii} (AbCaPi in SI
-Fig. 6) has the lowest initial retention. Initial water retention and drying rates
-differed across mixtures (linear model based on log of moisture content, SI Tables 8 and 9).
+Fig. 7) has the lowest initial retention. Initial water retention and drying
+rates differed across mixtures (linear model based on log of moisture content,
+SI Tables 8 and 9).
 
@@
 Flammability measures also demonstrated nonadditivity in mixtures. Flame
-spread rate was consistently higher than predicted (Fig
+spread rate was consistently higher than predicted (Fig.
 \ref{fig:mixture_obs_pred_spread}, Wald test of mixed linear model centered
@@
 tested only dry fuels. We therefore expected that rankings of species
-flammability might change under more [TODO FIX] realistic moisture levels. For
+flammability might change under higher moisture levels. For
 example, one presumption of ours prior to this experiment was that the high
@@
 \emph{Quercus} absorbed more water during wetting, and the one mixture without
-any Cupressaceae had the most rapid drying rate (SI Figure 4 and SI Table 8).
+any Cupressaceae had the most rapid drying rate (SI Figure 7 and SI Table 8).
 Drydown rates, however, did not behave as the average of the constituent
@@
 cases of positive nonadditivity in mixtures was often a result of mixtures
-meeting ignition thresholds not met by the constituent species alone [TODO:
-check wording]. To limit the number of models fit, we limited our investigation
+meeting ignition thresholds not met by the constituent species in monoculture. To limit the number of models fit, we limited our investigation
 to two flammability parameters that represent two major axes of flammability
@@
 flammability \citep{Schwilk-2015, Pausas+Keeley+etal-2017,
-  Prior+Murphy+etal02018}. This result, together with previously published
+  Prior+Murphy+etal-2018}.
+Some mixtures, however did exhibit negative nonadditivity.
+For example, at 72 hours following saturation, fuel consumption across mixtures could show neutral, positive, or negative nonadditivity (Fig.
+\ref{fig:mixture_obs_pred_consume}).
+Previous studies have demonstrated the potential for the direction of nonadditivty to shift with fuel moisture \citep{Blauw+Wensink+etal-2015} or with altering mixtures of leaf vs twig litter \citep{Zhao+vanLogtestijn+etal-2019}.
+Our litter represented naturally droppped litter and the ratio of leaves to twigs in such material can vary both by species but also by wind and storm events the previous season.
+With only four distinct litter combinations, it is difficult to attribute cause to these variations within the overall statistical trend, but litter packing alone is complex and needles and twigs can combine such that the addition of a species to a mixture can create a denser litter bed or less dense litter bed than is expected \citep{Zhao+vanLogtestijn+etal-2019}.
+
+% (eg  \emph{Qu)(eg \emph{Abies} branchlets with needles fitting between loosely packed \emph{Pinus} needles) or less dense
+
+[TODO].
+
+This result, together with previously published
 work, suggests that litter flammability is prone to nonadditive effects,

### 2019-12-13 scripts/moisture-ms-figs-tables.R

@@ -1,5 +1,5 @@
 # Figures for moisture manuscript
-# 2019-05-16
+# 2019-12-12
 
 library(xtable)
 library(ggrepel)
@@ -12,7 +12,37 @@ source("dry-down.R")
 source("burn-moist.R")
 source("read-moistmix.R")
 
+# decomp project data for litter particle size distributions
+source("decomp-flam.R") # needed for Fig 2
+
 set.seed(100)
+
+decompsp <- decomp %>% left_join(species)
+
+###############################################################################
+## SI Fig 1: litter particle length distributions
+###############################################################################
+
+SI_fig1 <- ggplot(subset(decompsp, year==0), aes(display.name, l+w+t)) +
+  geom_violin(scale="width", fill="gray") +
+#  geom_jitter(alpha=0.1) +
+  xlab("Species") +
+  ylab(expression("Litter particle volume (" * mm^3 * ")")) +
+#  geom_point(data=mtrait, aes(display.name, leaf.length)) +
+#  scale_fill_brewer(palette="Greys", direction=-1, name="Year") +
+  pubtheme +
+  theme(axis.text.x  = element_text(face="italic", angle=45, vjust=1, hjust=1))
+SI_fig1
+ggsave(file.path(RESULTS, "SI_fig1_litter-particle-volumes-by-species.pdf"), plot=SI_fig1,
+                 width=col2, height=col2/1.5, units="cm")
+
+# sample size for caption=500 - 5800 particles depending on species.
+
+x <- subset(decomp, year==0) %>% group_by(spcode)
+summarize(x, n=length(l))
+
+
+
 
 ###############################################################################
 ## Figure 1: dry-down
@@ -38,10 +68,10 @@ ggsave(file.path(RESULTS, "fig1_drydown-curves.pdf"), plot=fig1,
        width=col1, height=col1, units="cm")
 
 ###############################################################################
-# SI fig 1: dry down on semi log scale
+# SI fig 2: dry down on semi log scale
 ###############################################################################
 
-SI_fig1 <- ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
[…]

### 2019-12-13 ms/SI.tex
Progress towards R2

@@
   \centering
-  \includegraphics[width=8cm]{figs_tables/SI_fig1_drydown-curves_logged.pdf}
+  \includegraphics[width=8cm]{figs_tables/SI_fig1_litter-particle-lengths-by-species.pdf}
    \label{SI_fig1}
+   \caption[Litter particle length distributions]{10 2\,g subsamples of litter per species were assembled and each individual particle of that subsample was measured for length (longest axis), width (longest axis perpendicular to length) and thickness (perpendicular to length and width, at midrib) using an electronic caliper.
+Violin plots here show the distribution of the sum of the three dimensions as a measure of the containing prism volume (N=27--450 per species).}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=8cm]{figs_tables/SI_fig2_drydown-curves_logged.pdf}
+   \label{SI_fig2}
 \caption[Semi-log scale dry down curves for eight litter types.]{Lines show best-fit lines fit to the natural log of moisture content on a dry mass basis.
@@
   \centering
-  \label{SI_fig2}
-  \includegraphics[width=16cm]{figs_tables/SI_fig2_max_water_emmeans.pdf}
+  \label{SI_fig3}
+  \includegraphics[width=16cm]{figs_tables/SI_fig3_max_water_emmeans.pdf}
 \caption{Estimated marginal means by species for maximum water retainability based on mixed model results. Shaded regions indicate 95\% confidence intervals for the marginal means.}
@@
   \centering
-  \label{SI_fig3}
-\includegraphics[width=16cm]{figs_tables/SI_fig3_dessication_emmeans.pdf}
+  \label{SI_fig4}
+\includegraphics[width=16cm]{figs_tables/SI_fig4_dessication_emmeans.pdf}
 \caption{Estimated marginal means by species for dessication rate based on mixed model results. Shaded regions indicate 95\% confidence intervals for the marginal means.}
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/SI_fig4_SLA_maxMC.pdf}
+\includegraphics[width=8cm]{figs_tables/SI_fig5_SLA_maxMC.pdf}
 \caption{Maximum water retention by specific leaf area (SLA) across eight species.}
-  \label{fig:maxmc-di}
+  \label{fig:sla-maxmc}
 \end{figure}
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/SI_fig5_di_bd.pdf}
+\includegraphics[width=8cm]{figs_tables/SI_fig6_di_bd.pdf}
 \caption{Linear model predicting dessication rate as a function of specific leaf area and litter bulk density across eight species.}.
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/SI_fig6_mixture_drydown-curves.pdf}
+\includegraphics[width=8cm]{figs_tables/SI_fig7_mixture_drydown-curves.pdf}
 \caption{Fuel moisture by time since saturation for four litter mixtures.
@@
 Mixed linear model results shown in SI Table 7.}
-  \label{SI_fig6}
+  \label{SI_fig7}
 \end{figure}
 
-% SI fig 5
-\begin{figure}[H]
-  \centering
-\includegraphics[width=8cm]{figs_tables/SI_fig4_SLA_maxMC.pdf}
-\caption{Maximum water retention by specific leaf area across eight species.
-  % ANOVA table in Table \ref{tab:mc_di_anova}.
-}
-  \label{fig:maxmc-di}
-\end{figure}
+% % SI fig 8
+% \begin{figure}[H]
+%   \centering
+% \includegraphics[width=8cm]{figs_tables/SI_fig8_SLA_maxMC.pdf}
+% \caption{Maximum water retention by specific leaf area across eight species.
+%   % ANOVA table in Table \ref{tab:mc_di_anova}.
+% }
+%   \label{fig:maxmc-di}
+% \end{figure}
 
-% SI fig 6
-\begin{figure}[H]
-  \centering
-\includegraphics[width=8cm]{figs_tables/SI_fig5_di_bd.pdf}
-\caption{Drying rate rate as a function of specific leaf area and litter bulk
-  density across eight species.
-  % ANOVA table in Table \ref{tab:mc_di_anova}
-}.
-  \label{fig:bd-di}
-\end{figure}
+% \begin{figure}[H]
+%   \centering
+% \includegraphics[width=8cm]{figs_tables/SI_fig9_di_bd.pdf}
+% \caption{Drying rate rate as a function of specific leaf area and litter bulk
+%   density across eight species.
+%   % ANOVA table in Table \ref{tab:mc_di_anova}
+% }.
+%   \label{fig:bd-di}
+% \end{figure}
 

### 2019-12-13 scripts/read-flam.R

@@ -1,6 +1,6 @@
 ## Read flammability trial data
 
-#library(dplyr)
+library(dplyr)
 
 flam <- read.csv("../data/burn-trials/flamdt.csv", stringsAsFactors=FALSE)
 flam <- flam %>% mutate(mixture=nchar(spcode) > 4,

### 2019-12-17 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
WIP

@@
   Fonda-2001, Scarff+Westoby-2006, Kane+Varner+etal-2008,
-  Cornwell+Elvira+etal-2015}. Litter packing, measured as litter-bed density or
+  Cornwell+Elvira+etal-2015} and fire regime can vary of small spatial scales in response to community composition and species-specific flammability \citep{Stephens-2001, Hille+Stephens-2005, Schwilk+Caprio-2011}.
+Litter packing, measured as litter-bed density or
 packing ratio, is an important control ignition and spread in models of surface
@@
 These species-specific effects are increasingly well established. However, how
-moisture dynamics modulate species specific flammability is poorly known and
-could be driven by by leaf traits, emergent litter traits, or by both. Species
+moisture dynamics modulate species specific flammability is poorly known. Species
 specific effects on litter flammability are recorded for the forests of the
@@
 across time may contribute to the often high spatial heterogeneity of surface
-fires \citep{Knapp_Schwilk_etal-2006, Kreye_Hiers_etal-2018}.
+fires \citep{Hille+Stephens-2005, Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and
+to variation in fire regime across landscapes \citep{Stephens-2001}.
 
-Fuel moisture content through time is controlled by two processes, absorption
-and retention \citep{Kreye+Kobziar+etal-2013}. Absorptive capacity represents
+Fuel moisture content through time is controlled by two processes, wetting (including absorption and adsorption) and drying rate (desorbtion) \citep{Anderson-1990, Kreye+Kobziar+etal-2013}.
+Absorptive capacity represents
 the potential maximum fuel moisture and is evaluated by the moisture content of
 litter when saturated. Differences in moisture content over time since
-saturation can indicate different abilities to retain moisture. Litter particle
+saturation (drying rate) can indicate different abilities to retain moisture. Litter particle
 geometry may have direct effects on maximum moisture absorption through
@@
 leaf particle traits, especially leaf size \citep{Scarff+Westoby-2006}.
-Fuel-beds composed of larger leaves with lower moisture retainability should
+Fuel-beds composed of larger leaves or leaves with lower moisture retainability should
 favor drying, thus increasing the probability of ignition and fire spread.
@@
 
-[TODO: reseearch question here]
-
-We investigated species specific litter moisture dynamics across eight tree
-species in a temperate fire-prone forest in California. We determined the
-maximum water retention and rate of moisture loss and examined how these vary
-in relation to two leaf and litter bed traits [CHECK AND FIX]. We then examined
-the influence of moisture on the flammability across these species.
+We asked if differences in litter moisture absorption and drying rates could
+change the relative flammability of litter across species compared with
+estimates based on dry fuels. We investigated these litter moisture dynamics
+across eight tree species in a temperate fire-prone forest in California. We
+determined the maximum water retention and rate of moisture loss and examined
+how these vary in relation to specific leaf area and litter packing density.
 Experimental burn trials of leaf litter at various moisture levels determined
@@
 nonadditivity of flammability where fire behavior of mixtures was driven by the
-most flammable species. Here, we examined if this positive nonadditivity was
-maintained in mixtures of varying levels of fuel moisture. The burn trials also
-performed with multi-species litter mixtures to evaluate if there are
-nonadditive effects as observed in oven-dried litter. We tested if the
-nonadditive effects on reported in previous studies occur under [TODO CHECK]
-higher moisture levels, and whether fuel moisture effects
-strengthen or moderate such nonadditivity.
+most flammable species. We tested if the nonadditive effects on reported in
+previous studies occur under higher moisture levels, and whether fuel moisture
+effects strengthen or moderate any nonadditivity.
 
@@
 sample from a basket was less than 10\% moisture.
+This experimental setup was designed to accelerate drying compared with field conditions because the baskets were suspended in dry air 2.8 -- 5.3$^\circ$C above the warmest month average temperatures for the closest weather stations to the collection sites (Grant Grove and Lodegpole stations, NOAA 1981-2010 normals, https://www.ncdc.noaa.gov/cdo-web/datatools/normals).
 
@@
 sample of species, those that have a greater absorptive capacity are also among
-the most flammable in terms of spread rate when dry. This highlights the
-importance of fuel moisture in modulating flammability. One shortcoming of past
+the most flammable in terms of spread rate when dry.
+Traits such as large leaves and high SLA may affect both absorptive capacity and flammability with a consequence that flammability rankings across species can change with time since wetting.
+One shortcoming of past
 work in this system \citep{Magalhaes+Schwilk-2012} was that flammability trials
@@
 response to drying with no ignition prior to 72 hours of drying but once
-ignition is possible, having faster spread rates than the other taxa. The
+ignition is possible, it had faster spread rates than the other taxa. The
 Cupressaceae, which in our two species have dense litter but low maximum water

### 2019-12-17 ms/response_to_reviews.tex
WIP

@@
 
-We agree that our prior Discussion was weak and have extensively rewritten it
+We agree that our prior Discussion was weak and have extensively rewritten it.
+We argue that this work is not merely confirmatory. Our revised Introduction
+argues for the ecological significance of dry down rates. The importance of
+fuel moisture is not under debate. Furthermore, fine scale heterogeneity in
+fire behavior has been statistically associated wwith variation in fuel
+moisture (Hille and Stephens 2005) and fire history can vary over short spatial
+scales as a consequence of species composition (Stephens 2001). Although the
+importance of dry-down rates is less well-established, prescribed burns in
+Sierra Nevada mixed conifer forest demonstrate that fuel moisture likely drives
+differences in fire behavior depending on the timing of fire throughout the
+season (Kaufman and Martin 1989, Knapp and Keeley 2006, Hille and Stephens
+2005, Knapp et al 2006). These differences in fire behavior in turn control
+tree mortality ((Stephens et all 2002, Schwilk et al 2006). Despite this, we
+know of only one other study specifically investigating differential drying
+behavior (Kreye et al 2013).
 
-We argue that dry down rates are of strong ecological importance.
-TODO.
-Our dry down experiment occured under conduitions of very low humidity.
-We would expect the rates we observe in the lab to be faster than those in nature.
-But the differences in rates across species, the mixture effects and the resulting flammability are all important contributions of this and should represent relative differneces under field conditions.
-
-There are very few studeis characterizing drying patterns across speceis, only one ... Kreye.
-
+TODO
 Add context also suggested by reviewer 1.
@@
 Thank you.
-Yes, several comments converged on the absolutely correct criticism that our discussion did not engage adequately with the literature.
+Several comments converged on the absolutely correct criticism that our Discussion did not engage adequately with the literature.
 We have made substantial changes to the Introduction and the Discussion adding more context and citing more of the recent literature.
 We reorganized the Introduction and now do not rush the reader so quickly through the ideas.
-
-[TODO]
+We have expanded our justification for the study as argued for by the Senior Editor.
 
@@
 
-We have added information to the methods explaining the collection in more detail.
+We have added information to the methods explaining the litter collection in more detail.
 All material was naturally dropped litter which could include fine twigs and leaves.
-Litter in these two \emph{Abies} species is a combination of branchlets with still attached needles and individual needles that have detached from the branchlet and added reference to the Zhao et al 2019 results when discussing the mixture results
+Litter in these two \emph{Abies} species is a combination of branchlets with still attached needles and individual needles that have detached from the branchlet.
+We now cite the Zhao et al (2019) results when discussing the mixture results.
 
+Yes, when deciding how to limit the number of tests at the comparitive scale (species as data points), we decided to not include size, because we believed its influence on flammability is primarily through litter packing and was therefore captured by litterbed density.
 We have added a figure (SI fig.
-1) to the SI materials showing litter particle size (containing prism volumes) distributions for each of these species.
+1) to the SI materials showing litter particle size (the containing prism volumes) distributions for each of these species.
 These data were collected from the stored litter.
@@
 
-Guided by the drying rates measured in the initial trials, we staggered saturation times across species aiming to allow completely blocked design in time.
-Because some ignition attempts were unsuccessful, some days lacked the full complemnt of species while others later in the process sometimes had more than one replicate per species burned.
-We've added this information to the text [Lines X-X]
+Guided by the drying rates measured in the initial trials, we staggered starting saturation times across species.
+This was aimed at allowing a completely blocked design.
+However, because some ignition attempts were unsuccessful, some days lacked the full complement of species while others, later in the process, sometimes had more than one replicate per species burned.
+We've added this information to the text [Lines X-X].
 
@@
 
-We have added detail as requested by reviewer 2. See lines X-X. The litter was initially oven dried in large paper bags, then stored in the the lab (very low humidity), for several weeks on
-open racks. Then the litter was bagged and sealed and stored within garbage
-bags inside cardboard boxes in a lab storage area. We saw no signs of mold or
-ongoing decomposition. We minimized handling of the bags to avoid mechanical
-damage and changes to little particle size.
+We have added detail as requested by reviewer 2. See lines X-X. All litter was
+initially oven dried in large paper bags. The litter was then stored in the lab
+at low humidity for several weeks on open racks. The litter was then bagged and
+sealed and stored within garbage bags inside cardboard boxes in a lab storage
+area. We saw no signs of mold nor ongoing decomposition. We minimized handling
+of the bags to avoid mechanical damage and changes to litter particle size.
 
@@
 
-Good point and another place where we did not discuss the literature fully. We
+This is a good point and another place where we did not discuss the literature fully. We
 have now added this to the discussion and made the conjecture that these
-differing patterns may result from different branchelt structure across the
+differing patterns may result from different structure across the
 Cupressaceae with Australian \emph{Callitris} and the two Californian species
@@
 
-This portion was poorly written. We have rewritten this in response to reviewer 2's comment that most fires result from dry lightning strikes across the west. See our responses to Reviewer 2, below.
+This portion was poorly written and this was also mentioned by Reviewer 2.
+We have rewritten this in response to reviewer 2's comment that most fires result from dry lightning strikes across the west.
+See our responses to Reviewer 2, below.
 
@@
 
-TODO. Rewrite.
+Yes, we think that our logic must not have been clear here as we don't see the contradiction. We  have rewritten these sentences and we hope they are no longer confusing. See lines X-X.
 
@@
 
-Agreed. Fixed. CHECK!
+Agreed. Fixed.
 
@@
 
-We agree this does lack consistency and an early draft used the same concise format as Fig 6 format for all three figures.
+We agree this does lack consistency and an early draft used the same concise format of Fig 6 for all three figures.
 However, we decided that having time on the x-axis was worth the space and allows the reader more insight into the flammability effects of drying over time.
 Reviewer 2 pointed out some exceptions to the overall statistical pattern and those exceptions would not have been easily visible without an explicit time axis.
-There is a tension here between illustrating simply the results of the statistical model and allowing the reader to get a better idea of variation in flammability consequences of drying time.
+There is a trade-off here between illustrating simply the results of the statistical model and allowing the reader to get a better idea of variation in flammability consequences of drying time.
 
@@
 
-This sentence was poorly written and oversimplified. We thank the reviewer
-for pointing this out and we have rewritten this section. However, our badly
-made point was: lighting ignitions can occur during the same general period as
-precipitation. As the reviewer points out, dry strikes that find dry fuel are
-more likely to lead to fires (Abatzoglou et al 2016). That the total number
-lightning strikes does not well predict burned area and that ignitions occur
-often from dry strikes argues for the importance of fuel moisture. Although
-synoptic weather conditions that create dry and wet storms differ, the
-potential for precipitation and lightning during the same overall period of
-thunderstorm activity often exists (eg wet and dry convection days can occur
-under overlapping conditions, Rorig and Ferguson 1999).  We've rewritten this section, see lines XX-XX.
+This sentence was poorly written and oversimplified. We thank the reviewer for
+pointing this out and we have rewritten this section. However, our badly made
+point was: lighting ignitions can occur during the same general period as
+precipitation; therefore, drying rates are important. As the reviewer points
+out, dry lightning strikes that find dry fuel are more likely to lead to fires
+than are strikes that find damp fuels (Abatzoglou et al 2016). The fact that
+the total number lightning strikes does not well predict burned area and that
+ignitions occur often from dry strikes argues for the importance of fuel
+moisture. Although synoptic weather conditions that create dry and wet storms
+differ, the potential for precipitation and lightning during the same overall
+period of thunderstorm activity often exists (eg wet and dry convection days
+can occur under overlapping conditions, Rorig and Ferguson 1999). We've
+rewritten this section, see lines XX-XX.
 
@@
 
-TODO.  Get data.
-
 \begin{quote}
@@
 
-We have edited the abstract to make the research questions clearer. Yes, our first submission jumped from stating current reserach gaps to specific methods too quickly.  We added X X
+We have edited the abstract to make the research questions clearer. Yes, our
+first submission jumped from stating current reserach gaps to specific methods
+too quickly. See also our response to the Senior Editor, above.
 
@@
 Yes, this is a good point and we have added some discussion of this.
-We tried to lmit our statements to those supported by the statistical tests (Table X), and therefore focused on the positive effect of drying and the overall trend for positive nonadditivity.
+We tried to limit our statements to those supported by the statistical tests (eg, slope of nonaddivity with drying was greater than 1), and therefore focused on the positive effect of drying and the overall trend for positive nonadditivity.
 This, of course, obscures some details which are worth mentioning and we have edited this section in response to the reviewer's criticism.
-We agree that the high variability in nonadditivity across mixtures deserves highlighting.
+We agree that the high variability in nonadditivity across mixtures deserves highlighting and that is the reason we presented the actual predictions and departures from prediction across time.
 Please see our changes on lines X-X.
@@
 
-OK, w thought this established enough to not need citation but have added one.
+We thought this was an established enough fact to not need citation but have added one.
 
@@
 
-Good point. Fixed and in another sentence elsewhere.
+Good point.
+We fixed that here and in another sentence elsewhere.
+Furthermore, we've edited to be more precise about what we mean by ``absorption'' because we are referring to a litter bed and we include adsorption to litter surfaces in that definition.
+We had originally tried to avoid the chemical terms because we are observing a higher-level phenomenon.
 
@@
 
-We are not sure of the criticism here.
+We edited to ``composed of larger leaves or leaves with lower moisture retainability''
 
@@
 
-TODO.
+Yes, this was incorrect. Fixed.
 

### 2019-12-18 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Rewrite conclusions, make edits
based on first proof read.

@@
 
-\nwords{7600}
+\nwords{8200}
 \ntables{4}
 \nfig{8}
-\nref{39}
+\nref{51}
 
@@
     species-specific litter traits influence litter moisture absorption and
-    retention. Characterizing how species’ leaf litter retain moisture through
-    time and how fuel from multiple species interacts are critical to the
-    assessment of fuel’s availability to burn and key steps to adding species
-    or taxon specific information to fire behavior prediction.
+    retention. Characterizing how litter fuels retain moisture through time and
+    how fuels from multiple species interact are key steps to adding species or
+    taxon specific information to fire behavior prediction.
 
-      \item To fill these gaps in knowledge, we determined
-    moisture absorption and retention characteristics of the litter from
-    multiple tree species, investigated what litter traits influenced these
-    processes, and tested for nonadditive moisture dynamics and nonadditive
-    flammability in multi-species mixtures. The litter from eight tree species of a
-    mixed-conifer forest was saturated and allowed to dry to determine moisture
-    absorption capacity and drying rates. Burn trials were performed in moist
-    litter beds of single species and of mixtures to establish flammability
-    response to the dry-down process.
+      \item To fill these gaps in knowledge, we determined moisture absorption
+    and retention characteristics of the litter from multiple tree species,
+    investigated what litter traits influenced these processes, and tested for
+    nonadditive moisture dynamics and nonadditive flammability in multi-species
+    mixtures. The litter from eight tree species of a mixed-conifer forest was
+    saturated and allowed to dry to determine moisture absorption capacity and
+    drying rates. Burn trials were performed in litter beds of single species
+    and of mixtures to establish flammability response to the dry-down process.
 
@@
     flammability across the major litter types typified by different tree taxa.
-    Litter mixtures exhibited consistent nonadditivity in flame spread rate and
+    Litter mixtures exhibited nonadditivity in flame spread rate and
     in fuel consumption. This nonadditivity became more positive with fuel
@@
       \item \emph{Synthesis:} Taxon specific patterns of moisture absorption
-    and retention can change flammability rankings under moist compared with dry fuels.
-    Litter mixtures generally burn with fire behavior more similar to that of
-    the most flammable constituent species and this effect increases as fuels
-    dry. [TODO more]
+    and retention can change flammability rankings under moist compared with
+    dry fuels because traits that increase flame spread rate increase water
+    absorption. Litter mixtures generally burn with fire behavior more similar
+    to that of the most flammable constituent species and this effect increases
+    as fuels dry.
 
@@
 
-\noindent \textbf{Keywords:} California, fire, flame spread rate, flammability, fuel consumption, fuel moisture, leaf litter, nonadditivity
+\noindent \textbf{Keywords:} California, fire, flame spread rate, flammability,
+fuel consumption, fuel moisture, leaf litter, nonadditivity
 
@@
 litter fuels can also vary. In particular, fine dead fuels respond very readily
-to environmental changes \citep{Nelson-2001}. Fuel moisture influences
-flammability and fire behavior through effects on ignition, spread rate, flame
-height, fuel consumption, and heat release, due to the high specific heat
-content of water \citep{Rothermel-1972, Nelson-2001}. Moisture of litter fuels
-varies with climate, with seasons, and with short term weather within a season.
-Therefore, the rate of drying and flammability as a function of moisture
+to environmental changes \citep{Nelson-2001}. Moisture of litter fuels varies
+with climate, with seasons, and with short term weather within a season.
+Therefore, both the rate of drying and flammability as a function of moisture
 content are key parameters that influence fuel flammability throughout the fire
@@
 components of fire behavior have differing ecological effects with spread rate
-being associated with rates of heat release and total burned area and with heat
-transfer to soils or tissues predicting soil impacts of fire and the mortality
-of better insulated plant tissues. A shortcoming in our current knowledge is
-our integration of specific species and plant traits into fuel hazard
-assessment, fire modeling, and fire effects prediction. Developing quantitative
-yet robust and general ways to integrate actual plant communities into fuel
-assessment will require understanding the magnitude of flammability variability
-across species in space and time and parameterizing based on plant
-characteristics. [TODO: get opinion on final sentences]
+being associated with rates of heat release and area burned, and with heat
+transfer predicting soil impacts of fire and the mortality. A shortcoming in
+our current knowledge is our integration of specific species and plant traits
+into fuel hazard assessment, fire modeling, and fire effects prediction.
+Developing quantitative yet robust and general ways to integrate actual plant
+communities into fuel assessment will require understanding the magnitude of
+flammability variability across species in space and time and parameterizing
+based on plant characteristics.
 
-Species differ in their litter flammability \cite{Fonda+Belanger+etal-1998,
+Species differ in their litter flammability \citep{Fonda+Belanger+etal-1998,
   Fonda-2001, Scarff+Westoby-2006, Kane+Varner+etal-2008,
-  Cornwell+Elvira+etal-2015} and fire regime can vary of small spatial scales in response to community composition and species-specific flammability \citep{Stephens-2001, Hille+Stephens-2005, Schwilk+Caprio-2011}.
-Litter packing, measured as litter-bed density or
-packing ratio, is an important control ignition and spread in models of surface
-fire spread \citep{Weber-1991, Balbi+Santoni+etal-1999, Morvan+Larini-2001}.
-Multiple studies have demonstrated that litter bed density controls the
-flammability of dry litter fuels in seasonally dry climates where higher litter
-density reduces flame spread rate. Across these studies, there is increasing
-consensus that larger less densely packed litter beds increase oxygen
-availability and burn with faster spread rates \citep{Scarff+Westoby-2006,
-  Schwilk+Caprio-2011, Magalhaes+Schwilk-2012, Cornwell+Elvira+etal-2015,
-  Varner+Kane+etal-2015, Kreye+Varner+etal-2018}. The litter packing is driven
-by litter particle size and shape which is in turn influenced by leaf size, how
-leaves or branches are abscised, and subsequent decomposition
-\citep{Fonda+Belanger+etal-1998, Stephens+Finney+etal-2004,
-  Scarff+Westoby-2006, Kane+Varner+etal-2008, Cornwell+Elvira+etal-2015,
-  Zhao+Cornwell+etal-2016, Grootemaat+Wright+etal-2017,
-  Prior+Murphy+etal-2018, Zhao+vanLogtestijn+etal-2019}.
+  Cornwell+Elvira+etal-2015} and fire regime can vary over fine spatial scales
+in response to community composition and species-specific flammability
+\citep{Stephens-2001, Hille+Stephens-2005, Schwilk+Caprio-2011}. Litter
+packing, measured as litter-bed density or packing ratio, is an important
+control on ignition and spread in models of surface fire \citep{Weber-1991,
+  Balbi+Santoni+etal-1999, Morvan+Larini-2001}. Multiple studies have
+demonstrated that litter bed density controls the flammability of dry litter
+fuels in seasonally dry climates where higher litter density reduces flame
+spread rate. Across these studies, there is increasing consensus that larger,
+less densely packed litter beds increase oxygen availability and burn with
+faster spread rates \citep{Scarff+Westoby-2006, Schwilk+Caprio-2011,
+  Magalhaes+Schwilk-2012, Cornwell+Elvira+etal-2015, Varner+Kane+etal-2015,
+  Kreye+Varner+etal-2018}. The litter packing is driven by litter particle size
+and shape, which is in turn influenced by leaf size, how leaves or branches are
+abscised, and subsequent decomposition \citep{Fonda+Belanger+etal-1998,
+  Stephens+Finney+etal-2004, Scarff+Westoby-2006, Kane+Varner+etal-2008,
+  Cornwell+Elvira+etal-2015, Zhao+Cornwell+etal-2016,
+  Grootemaat+Wright+etal-2017, Prior+Murphy+etal-2018,
+  Zhao+vanLogtestijn+etal-2019}.
 
-These species-specific effects are increasingly well established. However, how
-moisture dynamics modulate species specific flammability is poorly known. Species
-specific effects on litter flammability are recorded for the forests of the
-southeastern US \citep{Nowacki+Abrams-2008} where species-specific litter
-drying rates play a role in flammability feedbacks
-\citep{Kreye+Varner+etal-2013}. Fire behavior consequences of individual
-species traits have been recorded for western US forests as well: in mixed
-conifer forest in California, where the fire regime is characterized by surface
-fires fueled by plant litter, overstory tree composition drives variation in
-local fire behavior \citep{Schwilk+Caprio-2011} as a result of differential
-litter flammability across tree species and as a result of synergistic
-interactions in multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}.
-These flammability feedbacks may help explain historical fire frequency
-patterns \citep{Schwilk+Caprio-2011}, but current estimates of litter
-flammability in mixed conifer forests are based largely on burning dry litter.
-
-Fire scientists often categorize fuel by size classes that correspond to
+These species-specific flammability effects are increasingly well established.
+However, how moisture dynamics modulate species specific flammability is poorly
+known. Fire scientists often categorize fuel by size classes that correspond to
 moisture exchange rates, although such simplified categories have long been
@@
 
-Fuel moisture content through time is controlled by two processes, wetting (including absorption and adsorption) and drying rate (desorbtion) \citep{Anderson-1990, Kreye+Kobziar+etal-2013}.
-Absorptive capacity represents
+Fuel moisture content through time is controlled by two processes, wetting
+(including absorption and adsorption) and drying rate (desorbtion)
+\citep{Anderson-1990, Kreye+Kobziar+etal-2013}. Absorptive capacity represents
 the potential maximum fuel moisture and is evaluated by the moisture content of
-litter when saturated. Differences in moisture content over time since
-saturation (drying rate) can indicate different abilities to retain moisture. Litter particle
-geometry may have direct effects on maximum moisture absorption through
-area-mass relationships and we also expect effects of leaf morphology,
-especially leaf cuticle thickness \citep{Van_Wagner-1969}. Like absorption,
-moisture retention should also be influenced by leaf and litter bed traits:
+litter when saturated; hereafter we refer to this measurement as ``maximum
+water absorption''. Differences in moisture content over time since saturation
+(drying rate) can indicate different abilities to retain moisture. Litter
+particle geometry may have direct effects on maximum moisture absorption
+through surface area affects on adsoprtion and we also expect effects of leaf
+morphology, especially leaf cuticle thickness \citep{Van_Wagner-1969}. Drying
+rate should also be influenced by leaf and litter bed traits:
 surface-area-to-volume ratio and litter bulk density will influence air flow
-through the litter bed. Litter packing therefore effects fire behavior in two
+through the litter bed. Litter packing, therefore, effects fire behavior in two
 ways: directly by its effect on oxygen availability for combustion, and
-indirectly by its effect on dry-down rate. Both these pathways are affected by
+indirectly by its effect on drying rate. Both these pathways are affected by
 leaf particle traits, especially leaf size \citep{Scarff+Westoby-2006}.
-Fuel-beds composed of larger leaves or leaves with lower moisture retainability should
-favor drying, thus increasing the probability of ignition and fire spread.
+Fuel-beds composed of larger leaves or leaves with lower maximum moisture absorption
+should favor drying, thus increasing the probability of ignition and fire
+spread.
 
@@
 nonadditivity has also been recorded for litter mixtures
-\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible that
-similar nonadditive effects may influence litter moisture content directly and
+\citep{Blauw+Wensink+etal-2015, Zhao+vanLogtestijn+etal-2019}. It is possible
+that similar nonadditive effects may influence litter moisture content and
 flammability of moist fuels. There is some evidence for nonadditive
 flammability in moist fuels with one study finding increasing nonadditivity
-with moisture \citep{Blauw+Wensink+etal-2015}.
+with moisture \citep{Blauw+Wensink+etal-2015}. In mixed conifer forest in
+California, where the fire regime is characterized by surface fires fueled by
+plant litter, overstory tree composition drives variation in local fire
+behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
+flammability across tree species and as a result of synergistic interactions in
+multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
+flammability feedbacks may help explain historical fire frequency patterns
+\citep{Schwilk+Caprio-2011}, but current estimates of litter flammability in
+mixed conifer forests are based largely on burning dry litter.
 
@@
 across eight tree species in a temperate fire-prone forest in California. We
-determined the maximum water retention and rate of moisture loss and examined
+determined the maximum water absorption and rate of moisture loss and examined
 how these vary in relation to specific leaf area and litter packing density.
@@
 across the study area. We collected from a minimum of four populations per
-species with each population 9--32\,km distant from its neighbors. An exception
-to this was \emph{A. magnifica} which is restricted to the higher elevations
-and could only be collected from one area due to designated wilderness
-boundaries and logistical constraints. Within that area, \emph{A. magnifica}
-litter was collected from over 10 different individuals. For the remaining
-species, the collection involved 2--4 individual trees at each site which were
-separated from one another by at least 10\,m. The litter was collected 2\,m
-away from the trunk to obtain a more uniform sample of leaf and fine branchlet
-litter, because bark and heavier branches tend to fall closer to the tree
-trunk. The top 2.5-5\,cm of litter were collected by raking and shoveling into
-large bags. All litter was oven dried to $< 5$\% fuel moisture by dry weight,
-then weighed and divided into moisture trials. The collected litter was stored
-in large paper bags and cardboard boxes in a lab storage space. We minimized
-handling and moving the bagged litter to avoid mechanical damage and changes to
-the particle size distributions. The moisture manipulations and subsequent
-burning trials were conducted in April--August 2015.
+species with each population within a species was 9--32\,km distant from its
+neighbors. An exception to this was \emph{A. magnifica} which is restricted to
+the higher elevations and could only be collected from one area due to
+designated wilderness boundaries and logistical constraints. Within that area,
+\emph{A. magnifica} litter was collected from over 10 different individuals.
+For the remaining species, the collection involved 2--4 individual trees at
+each site which were separated from one another by at least 10\,m. The litter
+was collected 2\,m away from the trunk to obtain a more uniform sample of leaf
+and fine branchlet litter, because bark and heavier branches tend to fall
+closer to the tree trunk. The top 2.5-5\,cm of litter were collected by raking
+and shoveling into large bags. All litter was oven dried to $< 5$\% fuel
+moisture by dry weight, then weighed and divided into moisture trials. The
+collected litter was stored in large paper bags and cardboard boxes in a lab
+storage space. We minimized handling and moving the bagged litter to avoid
+mechanical damage and changes to the particle size distributions. The moisture
+manipulations and subsequent burning trials were conducted in April--August
+2015.
 
@@
 
-We produced dry-down curves to determine maximum water retainability and the
-rate of moisture content loss for each species and for four distinct
+We produced dry-down curves to determine maximum water absorption and the
+rate of drying for each species and for four distinct
 three-species mixtures. These curves allowed us to determine the rate at which
@@
 and re-weighed to assess moisture content. This was repeated until the final
-sample from a basket was less than 10\% moisture.
-This experimental setup was designed to accelerate drying compared with field conditions because the baskets were suspended in dry air 2.8 -- 5.3$^\circ$C above the warmest month average temperatures for the closest weather stations to the collection sites (Grant Grove and Lodegpole stations, NOAA 1981-2010 normals, https://www.ncdc.noaa.gov/cdo-web/datatools/normals).
+sample from a basket was less than 10\% moisture. This experimental setup was
+designed to accelerate drying compared with field conditions because the
+baskets were suspended in air 2.8 -- 5.3$^\circ$C above the warmest month
+average temperatures for the closest weather stations to the collection sites
+(Grant Grove and Lodegpole stations, NOAA 1981-2010 normals,
+https://www.ncdc.noaa.gov/cdo-web/datatools/normals).
 
@@
 the observed moisture dynamics was low, so we decided \emph{a priori} to test
-the effect of one leaf trait likely to influence moisture absorption (specific
-leaf area), and one litter trait known to directly affect flammability and
-likely to influence drying rate and litter packing density (litter bulk
-density). Specific leaf area was measured on freshly fallen litter
-by sampling 5 individuals per species. This measurement included petiole and
-for the scale-leaved species included the full excised branchlet. Therefore,
-this is not directly comparable to typical specific leaf area measurements on
-fresh leaves. Litter bulk density was obtained in this experiment for each
-basket by dividing the dry mass (450\,g) by the estimated litter volume (basket
-width x length x litter depth).
-To examine litter particle size distributions in the stored litter, we subsampled 10 approximately 2\,g samples per species from the stored litter.
-Each individual particle of that subsample was measured for length (longest axis), width (longest axis perpendicular to length) and thickness (perpendicular to length and width, at midrib) using an electronic caliper.
+the effect of one leaf trait likely to influence moisture absorption and
+adsorption (specific leaf area), and one litter bed trait known to directly
+affect flammability and likely to influence drying rate and litter packing
+density (litter bulk density). Specific leaf area was measured on freshly
+fallen litter by sampling 5 individuals per species. This measurement included
+petiole and for the scale-leaved species included the full excised branchlet.
+Litter bulk density was obtained in this experiment for each basket by dividing
+the dry mass (450\,g) by the estimated litter volume (basket width x length x
+litter depth). To examine litter particle size distributions in the stored
+litter, we subsampled 10 approximately 2\,g samples per species from the stored
+litter. Each individual particle of that subsample was measured for length
+(longest axis), width (longest axis perpendicular to length) and thickness
+(perpendicular to length and width at midrib) using an electronic caliper.
 
@@
 were fixed, but fuel depth was allowed to vary. We recorded the actual fuel
-depth to calculate litter bed density before burning. This design allows fire
-to reach constant flame spread rates \citep{Magalhaes+Schwilk-2012}. Two
-graduated metal rulers equally spaced along the apparatus allowed visual
-determination of maximum flame height. Litter was ignited with a propane torch;
-maximum adiabatic flame temperature of 1899$^\circ$C,
-\url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the apparatus
-and allowed to burn until extinction. If a sample failed to ignite after 30
-seconds, no further attempt was made and if three samples of a species at a
-particular moisture level failed to ignite, no further attempts were made on
-that species/moisture combination and the replicate was allowed to continue
-drying until the next scheduled checkpoint. A timer was used to record total
-time to ignition and duration of flaming combustion. The flammability trials
-followed the methodology detailed in \citet{Magalhaes+Schwilk-2012}.
+depth at three points along the tray to calculate litter bed density before
+burning. This design allows fire to reach constant flame spread rates
+\citep{Magalhaes+Schwilk-2012}. Two metal rulers equally spaced along the
+apparatus allowed visual determination of maximum flame height. Litter was
+ignited with a propane torch; maximum adiabatic flame temperature of
+1899$^\circ$C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end
+of the apparatus and allowed to burn until extinction. If a sample failed to
+ignite after 30 seconds, no further attempt was made and if three samples of a
+species at a particular moisture level failed to ignite, no further attempts
+were made on that species/moisture combination and the replicate was allowed to
+continue drying until the next scheduled checkpoint. The flammability
+trials followed the methodology detailed in \citet{Magalhaes+Schwilk-2012}.
 
@@
 height were recorded as zero. Our thermocouple system failed or data was lost
-for two days of burning trials, therefore we have omitted those data from this
-analysis. Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that flame
-spread rate captures one important axis of flammability variation
-\citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal-2018} and
-percent fuel consumed captures a portion of the second major axis of
+for two days of burning trials, therefore we have omitted those temperature
+data from this analysis. Our past work \citep{Magalhaes+Schwilk-2012}
+demonstrated that flame spread rate captures one important axis of flammability
+variation \citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal-2018}
+and percent fuel consumed captures a portion of the second major axis of
 flammability variation. Our analysis, therefore, focused on these two
@@
 decay coefficient. To estimate these coefficients we fit a linear mixed-effects
-model by first taking the natural logarithm of moisture content, with time
-since wetting and species as fixed effects nested within replicate basket. We
-extracted the coefficients of the fitted model to describe maximum water
-retainability ($m_{max}$; g of water per g of dry mass) and desiccation rate
-($\lambda$; hour$^{-1}$) for each species. All analyses were conducted in R
-3.6.1 \citep{RCoreTeam-2019}. We fit models with R packages `lme4'
-\citep{Bates_Machler_etal-2015} and `afex' \citep{Singmann_Bolker_etal-2017}.
-Approximate degrees of freedom and p-values were calculated by the
-Kenward-Roger approximation \citep{Kenward_Roger-1997} which is recommended by
-Luke \citeyear{Luke-2017} as producing acceptable Type I error rates. We tested
-if one litter particle trait (specific leaf area) and one litter bed trait
-(litter bulk density) influenced moisture dynamics by fitting linear models
-with species means as observations. For each response variable we fit a single
-linear model with two predictors and an interaction term.
+model by first taking the natural logarithm of moisture content. Taxonomic
+group was a fixed effect and time since wetting was a fixed effect nested
+within replicate basket. We extracted the coefficients of the fitted model to
+describe maximum water absorption ($m_{max}$; g of water per g of dry mass)
+and drying rate ($\lambda$; hour$^{-1}$) for each species. All analyses
+were conducted in R 3.6.1 \citep{RCoreTeam-2019}. We fit models with R packages
+`lme4' \citep{Bates_Machler_etal-2015} and `afex'
+\citep{Singmann_Bolker_etal-2017}. Approximate degrees of freedom and p-values
+were calculated by the Kenward-Roger approximation \citep{Kenward_Roger-1997}
+which is recommended by Luke \citeyear{Luke-2017} as producing acceptable Type
+I error rates. We tested if one litter particle trait (specific leaf area) and
+one litter bed trait (litter bulk density) influenced moisture dynamics by
+fitting linear models with species means as observations. For each response
+variable we fit a single linear model with two predictors and an interaction
+term.
 
@@
 testing if moisture content and flammability of mixtures were predicted by the
-average of the constituent species. To choose mixtures, we first grouped the
-species according to their maximum water retention and drying rate according to
-estimated marginal means by species using the R package `emmeans'
+average of the constituent species. To choose mixture combination, we first
+grouped the species according to their maximum water retention and drying rate
+according to estimated marginal means by species using the R package `emmeans'
 \citep{Lenth-2019}. This resulted in four groups of species and we then
@@
 each mixture and time since wetting, we predicted moisture content based on the
-mean content of the three constituent species at that time since wetting. We
-predicted flammability parameters based on the average flammability measures
-from individual species trials at each time since wetting and then calculated a
-predicted spread rate and fuel consumption as the average of these three values
-for the mixture. The expectation from the null model was that the difference
-between observed and predicted values of flammability will be zero. We tested
-if there were significant departures from the null with a mixed effects linear
-model. For flammability parameters, we fit linear mixed effect models with the
-observed-predicted flammability value as the dependent variable and used Wald
-tests to test if the centered intercept was different than zero (a measure of
-overall nonadditivity) and if the slope with time since wetting was different
-from zero (indicating nonadditivity changing with drying).
+mean moisture content of the three constituent species at that time since
+wetting. We predicted flammability parameters based on the average flammability
+measures from individual species trials at each time since wetting and then
+calculated a predicted spread rate and fuel consumption as the average of these
+three values for the mixture. The expectation from the null model was that the
+difference between observed and predicted values of flammability will be zero.
+We tested if there were significant departures from the null with a mixed
+effects linear model. For flammability parameters, we fit linear mixed effect
+models with the observed-predicted flammability value as the dependent variable
+and used Wald tests to test if the centered intercept was different than zero
+(a measure of overall nonadditivity) and if the slope with time since wetting
+was different from zero (indicating nonadditivity changing with drying).
 
@@
 distributions, however show that there is substantial variation in particle
-sizes within a species (SI fig. 1). There were marked differences in the extent
+sizes within a species (SI Fig.~1). There were marked differences in the extent
 of moisture species retain initially following wetting and draining by gravity
-(all $P < 0.0001$, Fig. \ref{fig:drydown} and Table \ref{tab:drydown}).
+(all $P < 0.0001$, Fig.~\ref{fig:drydown} and Table~\ref{tab:drydown}).
 \emph{Q. kelloggii}, the only broadleaf, is the species that retained the most
 moisture initially. \emph{Quercus}, like the pines, had a rapid drying rate
-(Fig. \ref{fig:drydown} and SI Fig. 2). Calculating estimated marginal means
+(Fig.~\ref{fig:drydown} and SI Fig.~2). Calculating estimated marginal means
 following the linear mixed-effects model discriminated the species into three
@@
   giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa}
-(SI Fig. 2). There are also differences in drying rates across the eight
+(SI Fig.~2). There are also differences in drying rates across the eight
 species and estimated mixed linear model slopes discriminated species into
@@
   spp} and \emph{S. giganteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp}
-(SI Fig. 3).
+(SI Fig.~3).
 
@@
     statistics and p-values were calculated by the Kenward-Roger approximation.
-    \citep{Kenward_Roger-1997}. Model coefficients reported in SI Table 1.}
+    \citep{Kenward_Roger-1997}. Model coefficients reported in SI Table~1.}
   \label{tab:drydown}
@@
 Across these eight species specific leaf area had a positive association with maximum
-water retention (SI Fig 5, linear model $P$ = 0.003) and litter
-bulk density had a negative effect on desiccation rate (SI Fig 6,
-linear model $P$ = 0.029). ANOVA results are shown in Tables
-\ref{tab:mc_di_anova}.
+water retention (SI Fig~5, linear model $P = 0.003$) and litter
+bulk density had a negative effect on drying rate (SI Fig~6,
+linear model $P = 0.029$). ANOVA results are shown in Table~\ref{tab:mc_di_anova}.
 
@@
     species. ANOVA tables shown with degrees of freedom, sum of squares, mean
-    squares, $F$, and $P$ for each effect. Model coefficients in SI tables 2 and 3}
+    squares, $F$, and $P$ for each effect. Model coefficients in SI Tables~2 and 3}
   \label{tab:mc_di_anova}
@@
 
-Flame spread rate declined with fuel moisture (Fig. \ref{fig:spread_moist}).
+Flame spread rate declined with fuel moisture (Fig.~\ref{fig:spread_moist}).
 \emph{Quercus} had higher spread rates across all fuel moistures and had a
 steeper decline in response to moisture than did the other groups (taxon by
-time since wetting interaction $P$ = 0.022, Table \ref{tab:spreadrate_anova}).
-Fuel moisture negatively influenced fuel consumption. Overall fuel consumption
-and the rate of change with moisture content varied across the taxonomic groups
-(Fig. \ref{fig:consume_moist}, Table \ref{tab:consume_anova}). Although the
+time since wetting interaction $P = 0.022$, Table~\ref{tab:spreadrate_anova}).
+Fuel moisture negatively influenced fuel consumption. Fuel consumption and the
+rate of change with moisture content varied across the taxonomic groups
+(Fig.~\ref{fig:consume_moist}, Table~\ref{tab:consume_anova}). Although the
 overall ranking of groups was similar for fuel consumption and for flame spread
-rate, in contrast to the pattern with spread rate, for fuel consumption,
-\emph{Pinus} and \emph{Abies} has a steeper decline in flammability with fuel
-moisture than did the other groups.
+rate, there were different pattern across species in these two flammability
+parameters. For spread rate, the decline in flammability with moisture was
+shallowest for \emph{Pinus} and Cupressaceae (Fig.~\ref{fig:spread_moist}). For
+fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in
+flammability with fuel moisture than did the other groups
+(Fig.~\ref{fig:consume_moist}).
 
@@
 \includegraphics[width=8cm]{figs_tables/fig2_spread_actualMC.pdf}
-\caption{Flame spread rate by moisture content (model fit $F$= 61.15, df = 9,
-  148, $P$ < 0.0001). Model ANOVA table in Table \ref{tab:spreadrate_anova} and
+\caption{Flame spread rate by moisture content (model fit $F = 61.15$, df = 9,
+  148, $P < 0.0001$). Model ANOVA table in Table \ref{tab:spreadrate_anova} and
   coefficients in SI Table 4.}
@@
 
-As a consequence of the relationships shown above, time since wetting had a
-significant effect on spread rate ($P < 0.0001$) and the slope of this effect
-varied by taxonomic group (Fig. \ref{fig:spread_time}, Table
-\ref{tab:spreadrate_anova}). The pines were able to ignite and burn much sooner
-after wetting than did the other taxa, none of which had measurable spread
-rates until burning was attempted 72 hours after wetting (note that trials that
+As a consequence of the relationships between fuel moisture and flammability,
+time since wetting had a significant effect on spread rate ($P < 0.0001$) and
+the slope of this effect varied by taxonomic group (Fig. \ref{fig:spread_time},
+Table~\ref{tab:spreadrate_anova}). The pines were able to ignite and burn much
+sooner after wetting than did the other taxa, none of which had measurable
+spread rates until burning was attempted 72 hours after wetting. Trials that
 failed to even ignite smoldering combustion are omitted completely from the
-model and figure). At 72 hours, the \emph{Abies}, \emph{Calocedrus}, and
+model and figure. At 72 hours, the \emph{Abies}, \emph{Calocedrus}, and
 \emph{Sequoiadendron} had dryer litter than did the \emph{Quercus}, yet the
 spread rates were slow. \emph{Quercus} had the steepest response to time since
-wetting (Fig. \ref{fig:spread_time}, SI Table 6)).
+wetting (Fig.~\ref{fig:spread_time}, SI Table~6)).
 
@@
 Fuel consumption was highly variable across trials and intermediate values were
-less common than were very low or moderately high consumption (Fig.
-\ref{fig:consume_time}). All parameters significantly contributed to the
+less common than were very low or moderately high consumption
+(Fig.~\ref{fig:consume_time}). All parameters significantly contributed to the
 generalized linear mixed beta regression model as indicated by the nested model
-comparisons in Table \ref{tab:consume_anova}. As with flame spread rate some
-pines had the most rapid response to drying. However, unlike in the other
-flammability analyses, here the pines did not all behave similarly: \emph{P.
-  lambertiana} had lower fuel consumption than did \emph{P. ponderosa} and
-\emph{P. jeffreyi}. We noticed this during trials and in model fitting we found
-that placing \emph{P. lambertiana} in its own group allowed model convergence
-that was otherwise impossible. This was at odds with our intent to use only
-\emph{a priori} taxonomic groups, but we could not obtain model convergence
-otherwise. This result should therefore be seen as preliminary because the test
-does not reflect an \emph{a priori} hypothesis.
+comparisons in Table~\ref{tab:consume_anova}. As was the pattern for flame
+spread rate, pines had rapid flammability increases with drying. However,
+unlike in the other flammability analyses, here the pines did not all behave
+similarly: \emph{P. lambertiana} had lower fuel consumption than did \emph{P.
+  ponderosa} and \emph{P. jeffreyi}. We noticed this during trials and in model
+fitting we found that placing \emph{P. lambertiana} in its own group allowed
+model convergence that was otherwise impossible. This was at odds with our
+intent to use only \emph{a priori} taxonomic groups, but we could not obtain
+model convergence otherwise. This result should therefore be seen as
+preliminary because the test does not reflect an \emph{a priori} hypothesis.
+But probably reflects the fact that \emph{P. lambertiana} has shorter and finer
+needles than the other two pines.
 
@@
 
-Mixture dry down rates and maximum water retention showed some relationship to
-the dry down curves for individual constituent species (see SI Fig. 7). For
+Mixture drying rates and maximum water absorption showed some relationship to
+the dry down curves for individual constituent species (see SI Fig.~7). For
 example, the mixture that does not contain \emph{Q. kelloggii} (AbCaPi in SI
-Fig. 7) has the lowest initial retention. Initial water retention and drying
-rates differed across mixtures (linear model based on log of moisture content,
-SI Tables 8 and 9).
+Fig.~7) has the lowest maximum water absorption. Maximum water absorption and
+drying rates differed across mixtures (linear model based on log of moisture
+content, SI Tables~8 and 9).
 
@@
 species after wetting, but had higher water content than expected after long
-drying (Wald test shows slope of mixed linear model illustrated in Fig.
-\ref{fig:mixture_obs_pred_mc} less than 1, $P < 0.0001$).
+drying (Wald test shows slope of mixed linear model
+[Fig.~\ref{fig:mixture_obs_pred_mc}] less than 1, $P < 0.0001$).
 
@@
 
-Flammability measures also demonstrated nonadditivity in mixtures. Flame
-spread rate was consistently higher than predicted (Fig.
+Flammability measures also demonstrated nonadditivity in mixtures. Flame spread
+rates were generally higher than predicted (Fig.
 \ref{fig:mixture_obs_pred_spread}, Wald test of mixed linear model centered
-intercept for observed-predicted values greater than zero, $P$ = 0.017) and the
+intercept for observed-predicted values greater than zero, $P = 0.017$) and the
 positive nonadditivity increased with time since wetting (slope greater than
-zero, $P$ = 0.005). Fuel consumption, however, showed no overall bias towards
-positive or negative nonadditivity, but nonadditivity increased with time
-since wetting (Fig. \ref{fig:mixture_obs_pred_consume}, slope not equal to
-zero, $P < 0.0001$).
+zero, $P = 0.005$). Fuel consumption, however, showed no overall bias towards
+positive or negative nonadditivity, but nonadditivity increased with time since
+wetting (Fig. \ref{fig:mixture_obs_pred_consume}, slope not equal to zero,
+$P < 0.0001$). Fuel consumption tended to show negative nonadditivity at higher
+fuel moistures and positive nonadditivity following drying
+(Fig.~\ref{fig:mixture_obs_pred_consume}).
 
@@
 
-\subsection*{Genus-specific moisture dynamics and flammability}
+\subsection*{Moisture dynamics and flammability across taxonomic groups}
 
-The genera in this study displayed different patterns of both moisture
-absorption and desorption, associated with different litter traits. In this
-sample of species, those that have a greater absorptive capacity are also among
-the most flammable in terms of spread rate when dry.
-Traits such as large leaves and high SLA may affect both absorptive capacity and flammability with a consequence that flammability rankings across species can change with time since wetting.
-One shortcoming of past
-work in this system \citep{Magalhaes+Schwilk-2012} was that flammability trials
-tested only dry fuels. We therefore expected that rankings of species
-flammability might change under higher moisture levels. For
-example, one presumption of ours prior to this experiment was that the high
-flammability previously reported for \emph{Q. kelloggii}
-\citep{Magalhaes+Schwilk-2012} would disappear under higher fuel moisture
-contents. This prediction was partly supported: \emph{Q. kelloggii} fails to
-ignite until after 72 hours of drying in this study (Figs.
+The taxa in this study displayed different patterns of both moisture absorption
+and desorption, associated with different litter traits. In this sample of
+species, those that have a greater absorptive capacity are also among the most
+flammable in terms of spread rate when dry. Traits such as large leaves and
+high SLA may affect both absorptive capacity and flammability with a
+consequence that flammability rankings across species can change with time
+since wetting. One shortcoming of past work in this system was that
+flammability trials tested only dry fuels \citep{Magalhaes+Schwilk-2012}. We
+therefore expected that rankings of species flammability might change under
+higher moisture levels. For example, one presumption of ours prior to this
+experiment was that the high flammability previously reported for \emph{Q.
+  kelloggii} \citep{Magalhaes+Schwilk-2012} would disappear under higher fuel
+moisture contents. This prediction was partly supported: \emph{Q. kelloggii}
+fails to ignite until after 72 hours of drying in this study (Figs.
 \ref{fig:spread_time} and \ref{fig:consume_time}). However, high moisture
@@
 consistent with the low flammability found in an Australian Cupressaceae that
-have small litter particles and dense packing citep{Scarff+Westoby-2006,
-  Bowman+Haverkamp-2018}. However, the densely packed litter is not a
+have small litter particles and dense packing \citep{Scarff+Westoby-2006,
+  Bowman+Haverkamp+etal-2018}. However, the densely packed litter is not a
 consistent trait across the family because \citet{Cornwell+Elvira+etal-2015}
@@
 Cupressaceae.
-
-[TODO] [more?]
 
@@
 here is probably capturing some joint effect of cuticle thickness and total
-surface area of litter particles, but we cannot test the morphological and
-physical mechanisms of water absorption with these data. The oak has much
-higher initial water retention than do the other species, followed by the
-finest-needled pine, \emph{P. lambertiana}. Our study examines fresh litter,
-but we expect that decomposition will affect moisture dynamics. For example,
-Van Wagner \citeyear{Van_Wagner-1969}, investigating fuel weathering, found
-that pine needles had a variation in moisture content 3 to 6 times greater when
-they lacked their protective wax coating when compared to intact needles.
+surface area of litter particles which infleunces water adsorption, but we
+cannot test the morphological and physical mechanisms with these data. The oak
+has much higher maximum water absorption than do the other species, followed by
+the finest-needled pine, \emph{P. lambertiana}. Our study examines fresh
+litter, but we expect that decomposition will affect moisture dynamics. For
+example, Van Wagner \citeyear{Van_Wagner-1969}, investigating fuel weathering,
+found that pine needles had a variation in moisture content 3 to 6 times
+greater when they lacked their protective wax coating when compared to intact
+needles.
 
@@
 would expect mixtures to dry more rapidly than predicted from monoculture
-trials. This pattern in moisture content through time, however, did not
-directly predict flammability through time.
+trials. This discrepancy may be a consequence of the lightly packing litters
+having high maximum water absorption and of drying being non-linear with time
+(Fig.~\ref{fig:drydown}).
 
@@
 cases of positive nonadditivity in mixtures was often a result of mixtures
-meeting ignition thresholds not met by the constituent species in monoculture. To limit the number of models fit, we limited our investigation
-to two flammability parameters that represent two major axes of flammability
+meeting ignition thresholds not met by the constituent species in monoculture.
+To limit the number of models fit, we limited our investigation to two
+flammability parameters that represent two major axes of flammability
 variation. Therefore, our study does not fully capture ignition effects
-although we do have some results consistent with theirs: mixtures with \emph{Q.
-  kellogii} could ignite at times since wetting at which the species in
+although we do have some results consistent with theirs: mixtures that included
+\emph{Q. kellogii} could ignite at times since wetting at which the species in
 monoculture could not.
@@
 flammability \citep{Schwilk-2015, Pausas+Keeley+etal-2017,
-  Prior+Murphy+etal-2018}.
-Some mixtures, however did exhibit negative nonadditivity.
-For example, at 72 hours following saturation, fuel consumption across mixtures could show neutral, positive, or negative nonadditivity (Fig.
-\ref{fig:mixture_obs_pred_consume}).
-Previous studies have demonstrated the potential for the direction of nonadditivty to shift with fuel moisture \citep{Blauw+Wensink+etal-2015} or with altering mixtures of leaf vs twig litter \citep{Zhao+vanLogtestijn+etal-2019}.
-Our litter represented naturally droppped litter and the ratio of leaves to twigs in such material can vary both by species but also by wind and storm events the previous season.
-With only four distinct litter combinations, it is difficult to attribute cause to these variations within the overall statistical trend, but litter packing alone is complex and needles and twigs can combine such that the addition of a species to a mixture can create a denser litter bed or less dense litter bed than is expected \citep{Zhao+vanLogtestijn+etal-2019}.
+  Prior+Murphy+etal-2018}. Some mixtures, however did exhibit negative
+nonadditivity. For example, at 72 hours following saturation, fuel consumption
+across mixtures could show neutral, positive, or negative nonadditivity (Fig.
+\ref{fig:mixture_obs_pred_consume}). Previous studies have demonstrated the
+potential for the direction of nonadditivty to shift with fuel moisture
+\citep{Blauw+Wensink+etal-2015} or with altering mixtures of leaf vs twig
+litter \citep{Zhao+vanLogtestijn+etal-2019}. Our litter represented naturally
+droppped litter and the ratio of leaves to twigs in such material can vary both
+by species but also by wind and storm events the previous season. With only
+four distinct litter combinations, it is impossible to attribute cause to these
+variations within the overall statistical trend, but litter packing alone is
+complex and needles and twigs can combine such that the addition of a species
+to a mixture can create a denser litter bed or less dense litter bed than is
+expected \citep{Zhao+vanLogtestijn+etal-2019}.
 
-% (eg  \emph{Qu)(eg \emph{Abies} branchlets with needles fitting between loosely packed \emph{Pinus} needles) or less dense
+% (eg \emph{Qu)(eg \emph{Abies} branchlets with needles fitting between loosely
+% packed \emph{Pinus} needles) or less dense
 
-[TODO].
-
-This result, together with previously published
-work, suggests that litter flammability is prone to nonadditive effects,
-suggesting caution in lumping species into fuel categories. Community
-flammability is dependent on and particular to its constituent species and
-cannot be addressed by simply averaging the flammability of all species
-included. In litter-driven fire, positive nonadditivity appears to be the rule,
-and the relationships we described here can form a working hypothesis and
+We draw two main conclusions from this work. First, maximum water absorption is
+high in those species with the most flammability litter under dry conditions.
+As a consequence, under higher fuel moisture, the relative flammability of
+these taxa can change compared to the flammability of dry fuels. Second, litter
+flammability is prone to nonadditive effects which become more positive with
+drying. The relationships we described here can form a working hypothesis and
 baseline expectation to adjust fire behavior predictions in real communities:
@@
 system, past work demonstrates observed spread rates approached three times the
-expectation \citep{Magalhaes+Schwilk-2012}.
-
-[TODO rewrite below]
-
-Positive nonadditivity, whereby more flammable species have effects on fire
-behavior disproportionate to their contribution to the litter layer, could lead
-to vegetation feedbacks. In this temperate mixed conifer forest, these effects
-are likely to include positive feedbacks whereby the shade intolerant species
-such as the long-needled pines and the large-leaved oak, are favored by the
-increased fire to which their fuels contribute \citep{Schwilk+Caprio-2011} and
-similar results are reported for southeastern US forests where species
-composition shifts have led to decreases in fire activity
-\citep{Nowacki+Abrams-2008}. Such positive feedback effects increases the
-potential for species to exert community scale effects even when relatively
-rare and that can influence community assembly and the evolutionary of
-niche-constructing traits \citep{Kerr+Schwilk+etal-1999, Schwilk+Kerr-2002}.
-Ongoing climate change and concurrent changes to fire regimes are linked
-phenomena. Vegetation-fire feedbacks are important at a wide range of scales
-from stand-level species composition shifts to the earth system
-\citep{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}. The patterns we
-demonstrate here show the potential importance of species traits and
-interactions at local and stand scales where management and fire operations
-meet species' biology.
+expectation \citep{Magalhaes+Schwilk-2012}. These numbers provide a starting
+point for parameterizing fire models with mixed fuels. In this temperate mixed
+conifer forest, these effects are likely to include positive feedbacks whereby
+the shade intolerant species such as the long-needled pines and the
+large-leaved oak, are favored by the increased fire to which their fuels
+contribute \citep{Schwilk+Caprio-2011} and similar results are reported for
+southeastern US forests where species composition shifts have led to decreases
+in fire activity \citep{Nowacki+Abrams-2008}. Vegetation-fire feedbacks are
+important at a wide range of scales from stand-level species composition shifts
+to the earth system \citep{Harris+Remenyi+etal-2016,
+  Archibald+Lehmann+etal-2018}. The patterns we demonstrate here show the
+potential importance of species traits and interactions at local and stand
+scales where management and fire operations meet species' biology.
 
@@
 % %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% \conflictsofinterest{The authors declare no conflict of interest.
-The funders had no role in the design of the study; in the collection, analyses, or interpretation of data; in the writing of the manuscript, or in the decision to publish the results.}
+% \conflictsofinterest{The authors declare no conflict of interest. The funders
+% had no role in the design of the study; in the collection, analyses, or
+% interpretation of data; in the writing of the manuscript, or in the decision
+% to publish the results.}
 

### 2019-12-19 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Getting close

@@
 content are key parameters that influence fuel flammability throughout the fire
-season \citep{Nelson-2001}.
+season.
 
@@
 However, how moisture dynamics modulate species specific flammability is poorly
-known. Fire scientists often categorize fuel by size classes that correspond to
-moisture exchange rates, although such simplified categories have long been
-recognized as not sufficient to describe species-specific effects on moisture
-dynamics \citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are
-dependent upon forest structure (shading and rain throughfall effects), litter
-physical structure, and chemical composition of the fuel-beds
-\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
-models describing moisture dynamics of fine fuels are used by fire scientists
-\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
-usually aim to describe equilibrium conditions and at least include temperature
-and relative humidity although some include precipitation and solar radiation
-effects as well. Explicit models of fire behavior as a function of time since
-wetting could improve fire prediction, however. Drying rates following snowmelt
-or following rainfall during the fire season will influence the length of the
-fire season, flammability over time, and the probability of sustained ignition,
+known. Fuel moisture often limits fire activity in the western US
+\citep{Miller+Urban-1999, Parks+Parisien-2014}. Fire scientists often
+categorize fuel by size classes that correspond to moisture exchange rates,
+although such simplified categories have long been recognized as not sufficient
+to describe species-specific effects on moisture dynamics
+\citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are dependent upon
+forest structure (shading and rain throughfall effects), litter physical
+structure, and chemical composition of the fuel-beds \citep{Nelson+Hiers-2008,
+  Matthews-2014, Kreye_Hiers_etal-2018}. A variety of models describing
+moisture dynamics of fine fuels are used by fire scientists \citep{Viney-1991,
+  Nelson-2000, Catchpole+Catchpole+etal-2001}. These models usually aim to
+describe equilibrium conditions and at least include temperature and relative
+humidity although some include precipitation and solar radiation effects as
+well. Explicit models of fire behavior as a function of time since wetting
+could improve fire prediction, however. Drying rates following snowmelt or
+following rainfall during the fire season will influence the length of the fire
+season, flammability over time, and the probability of sustained ignition,
 higher spread rates and fuel consumption when ignition occurs. That species can
 differ in litter drying rates has been demonstrated for at least one forest
-system \citep{Kreye+Varner+etal-2013}. Such differential moisture content
-across time may contribute to the often high spatial heterogeneity of surface
-fires \citep{Hille+Stephens-2005, Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and
-to variation in fire regime across landscapes \citep{Stephens-2001}.
+system \citep{Kreye+Varner+etal-2013}. Differential drying can be especially
+important when precipitation can occur during the same season as ignitions.
+Although synoptic weather conditions that create dry and wet lightning storms
+differ, the potential for precipitation and lightning during the same overall
+period of thunderstorm activity often exists: wet and dry convection days can
+occur under overlapping conditions, \citep{Rorig+Ferguson-1999}. Moisture
+levels of fine fuels respond rapidly to precipitation events during the fire
+season \citep{Estes+Knapp+etal-2012}. Differential moisture content across time
+may contribute to the often high spatial heterogeneity of surface fires
+\citep{Hille+Stephens-2005, Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and to
+variation in fire regime across landscapes \citep{Stephens-2001}.
 
@@
 leaf particle traits, especially leaf size \citep{Scarff+Westoby-2006}.
-Fuel-beds composed of larger leaves or leaves with lower maximum moisture absorption
-should favor drying, thus increasing the probability of ignition and fire
-spread.
+Fuel-beds composed of larger leaves or leaves with lower maximum moisture
+absorption should favor drying, thus increasing the probability of ignition and
+fire spread.
 
@@
 change the relative flammability of litter across species compared with
-estimates based on dry fuels. We investigated these litter moisture dynamics
-across eight tree species in a temperate fire-prone forest in California. We
-determined the maximum water absorption and rate of moisture loss and examined
-how these vary in relation to specific leaf area and litter packing density.
-Experimental burn trials of leaf litter at various moisture levels determined
-how time since wetting influenced two axes of flammability characterized by
-flame spread rate and total fuel consumed. Previous work described the
-flammability of these species in monocultures and in mixtures for oven-dry
-litter \citep{Magalhaes+Schwilk-2012} and found consistent positive
-nonadditivity of flammability where fire behavior of mixtures was driven by the
-most flammable species. We tested if the nonadditive effects on reported in
-previous studies occur under higher moisture levels, and whether fuel moisture
-effects strengthen or moderate any nonadditivity.
+flammability measurements based on dry fuels. We investigated these litter
+moisture dynamics across eight tree species in a temperate fire-prone forest in
+California. We determined the maximum water absorption and rate of moisture
+loss and examined how these traits vary in relation to specific leaf area and
+litter packing density. Experimental burn trials of leaf litter at various
+moisture levels determined how time since wetting influenced two axes of
+flammability characterized by flame spread rate and total fuel consumed.
+Previous work described the flammability of these species in monocultures and
+in mixtures for oven-dry litter \citep{Magalhaes+Schwilk-2012} and found
+consistent positive nonadditivity of flammability where fire behavior of
+mixtures was driven by the most flammable species. We tested if the nonadditive
+effects on reported in previous studies occur under higher moisture levels, and
+whether fuel moisture effects strengthen or moderate any nonadditivity.
 
@@
 Buchholz. In this study, we consider litter the top layer of leaves and small
-twigs less than 0.625\,cm diameter (= 1-hour fuel) that have fallen in the
-previous year (mostly undecomposed or lightly decomposed). Leaf litter was
-collected in the summer (mid-June to mid-July) in 2012 from 21 separate sites
-across the study area. We collected from a minimum of four populations per
-species with each population within a species was 9--32\,km distant from its
-neighbors. An exception to this was \emph{A. magnifica} which is restricted to
-the higher elevations and could only be collected from one area due to
-designated wilderness boundaries and logistical constraints. Within that area,
-\emph{A. magnifica} litter was collected from over 10 different individuals.
-For the remaining species, the collection involved 2--4 individual trees at
-each site which were separated from one another by at least 10\,m. The litter
-was collected 2\,m away from the trunk to obtain a more uniform sample of leaf
-and fine branchlet litter, because bark and heavier branches tend to fall
-closer to the tree trunk. The top 2.5-5\,cm of litter were collected by raking
-and shoveling into large bags. All litter was oven dried to $< 5$\% fuel
-moisture by dry weight, then weighed and divided into moisture trials. The
-collected litter was stored in large paper bags and cardboard boxes in a lab
-storage space. We minimized handling and moving the bagged litter to avoid
-mechanical damage and changes to the particle size distributions. The moisture
-manipulations and subsequent burning trials were conducted in April--August
-2015.
+twigs less than 1\,cm diameter that have fallen in the previous year (mostly
+undecomposed or lightly decomposed litter representing fuels classified as
+1-hour and 10-hour fuels). Leaf litter was collected in the summer (mid-June to
+mid-July) in 2012 from 21 separate sites across the study area. We collected
+from a minimum of four populations per species with each population within a
+species was 9--32\,km distant from its neighbors. An exception to this was
+\emph{A. magnifica} which is restricted to the higher elevations and could only
+be collected from one area due to designated wilderness boundaries and
+logistical constraints. Within that area, \emph{A. magnifica} litter was
+collected from over 10 different individuals. For the remaining species, the
+collection involved 2--4 individual trees at each site which were separated
+from one another by at least 10\,m. The litter was collected 2\,m away from the
+trunk to obtain a more uniform sample of leaf and fine branchlet litter,
+because bark and heavier branches tend to fall closer to the tree trunk. The
+top 2.5-5\,cm of litter were collected by raking and shoveling into large bags.
+All litter was oven dried to $< 5$\% fuel moisture by dry weight, then weighed
+and divided into moisture trials. The collected litter was stored in large
+paper bags and cardboard boxes in a lab storage space. We minimized handling
+and moving the bagged litter to avoid mechanical damage and changes to the
+particle size distributions. The moisture manipulations and subsequent burning
+trials were conducted in April--August 2015.
 
@@
 
-We produced dry-down curves to determine maximum water absorption and the
-rate of drying for each species and for four distinct
-three-species mixtures. These curves allowed us to determine the rate at which
-a species loses moisture, approximated as exponential decay, and to determine
-maximum water content following saturation and gravity draining. Samples from
-the eight species (six replicates per species) and four mixtures (5 replicates
-per mixture type) were placed in baskets approximately 45 x 45\,cm by 15\,cm
-tall constructed of aluminum mesh with 0.52\,cm square openings. Samples were
-450\,g dry weight litter per trial and litter depth was standardized to 10\,cm
-by changing the horizontal dimensions of the basket. Weight was measured by
-using a balance sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus,
-OH). Samples were immersed in plastic storage bins filled with water for 24
-hours to fully saturate them. After saturation, the baskets were removed from
-the water, allowed to drain for 3 minutes, and samples weighed to measure
-maximum moisture content. The samples were then allowed to dry in a controlled
-environment kept at $21^\circ$C temperature and 30\% relative humidity,
-assessed via an iButton sensor (Maxim Integrated) placed over the samples.
-Litter baskets were arranged in a completely randomized design. During the
-dry-down process three subsamples of 5--10\,g were taken at 24 hour intervals
-and weighed. These subsamples were then oven dried for 24 hours at $100^\circ$C
-and re-weighed to assess moisture content. This was repeated until the final
-sample from a basket was less than 10\% moisture. This experimental setup was
-designed to accelerate drying compared with field conditions because the
-baskets were suspended in air 2.8 -- 5.3$^\circ$C above the warmest month
-average temperatures for the closest weather stations to the collection sites
-(Grant Grove and Lodegpole stations, NOAA 1981-2010 normals,
-https://www.ncdc.noaa.gov/cdo-web/datatools/normals).
+We produced dry-down curves to determine maximum water absorption and the rate
+of drying for each species and for four distinct three-species mixtures. These
+curves allowed us to determine the rate at which a species loses moisture,
+approximated as exponential decay, and to determine maximum water content
+following saturation and gravity draining. Samples from the eight species (six
+replicates per species) and four mixtures (5 replicates per mixture type) were
+placed in baskets approximately 45 x 45\,cm by 15\,cm tall constructed of
+aluminum mesh with 0.52\,cm square openings. Samples were 450\,g dry weight
+litter per trial and litter depth was standardized to 10\,cm by changing the
+horizontal dimensions of the basket. Weight was measured by using a balance
+sensitive to 0.1\,g (model XS16001L, Mettler Toledo, Columbus, OH). Samples
+were immersed in plastic storage bins filled with water for 24 hours to fully
+saturate them. After saturation, the baskets were removed from the water,
+allowed to drain for 3 minutes, and samples weighed to measure maximum moisture
+content. The samples were then allowed to dry in a controlled environment kept
+at $21^\circ$C temperature and 30\% relative humidity, assessed via an iButton
+sensor (Maxim Integrated) placed over the samples. Litter baskets were arranged
+in a completely randomized design. During the dry-down process three subsamples
+of 5--10\,g were taken at 24 hour intervals and weighed. These subsamples were
+then oven dried for 24 hours at $100^\circ$C and re-weighed to assess moisture
+content. This was repeated until the final sample from a basket was less than
+10\% moisture. This experimental setup was designed to accelerate drying
+compared with field conditions because the baskets were suspended in air 2.8 --
+5.3$^\circ$C above the warmest month average temperatures for the closest
+weather stations to the collection sites (Grant Grove and Lodegpole stations,
+NOAA 1981-2010 normals, https://www.ncdc.noaa.gov/cdo-web/datatools/normals).
 
@@
 response to drying with no ignition prior to 72 hours of drying but once
-ignition is possible, it had faster spread rates than the other taxa. The
-Cupressaceae, which in our two species have dense litter but low maximum water
-absorption had low flammability across moisture levels. This result is
+ignition is possible, it had faster spread rates than the other taxa. Our two 
+Cupressaceae, which have dense litter and low maximum water
+absorption, had low flammability across moisture levels. This result is
 consistent with the low flammability found in an Australian Cupressaceae that

### 2019-12-19 ms/response_to_reviews.tex
Rewrite conclusions, make edits
based on first proof read.
Getting close

@@
 
-Enclosed is our revised manuscript research article, ``''.
+Enclosed is our revised manuscript research article, ``Moisture effect on leaf
+litter alters non-additive flammability in mixed conifer forest''.
 
@@
 \begin{quote}
-I support the AE's recommendation to reject but allow resubmission.
-However, the bar for a resubmission will be high.
-My feeling is that your study confirms the essential findings from your 2012 study published in Journal of Ecology with the addition that your methodology of drying down adds some realism.
-Indeed, the confirmatory nature of this study was pointed out by reviewer 2 in their point 6.
-In addition to the points raised by the AE and reviewers, I will be looking in any resubmission for more evidence that this study advances your earlier work (apart from your last paragraph, your discussion is largely an explanation of your results with little context from other studies), that the findings have real ecological relevance (i.e., does it matter to the mixed-conifer communities?) and how is the new dry down treatment related to dry down conditions in litter mixtures in the field.
-If you can adequately address all of these concerns then I'll be happy to consider a resubmission.
+  I support the AE's recommendation to reject but allow resubmission. However,
+  the bar for a resubmission will be high. My feeling is that your study
+  confirms the essential findings from your 2012 study published in Journal of
+  Ecology with the addition that your methodology of drying down adds some
+  realism. Indeed, the confirmatory nature of this study was pointed out by
+  reviewer 2 in their point 6. In addition to the points raised by the AE and
+  reviewers, I will be looking in any resubmission for more evidence that this
+  study advances your earlier work (apart from your last paragraph, your
+  discussion is largely an explanation of your results with little context from
+  other studies), that the findings have real ecological relevance (i.e., does
+  it matter to the mixed-conifer communities?) and how is the new dry down
+  treatment related to dry down conditions in litter mixtures in the field. If
+  you can adequately address all of these concerns then I'll be happy to
+  consider a resubmission.
 \end{quote}
@@
 argues for the ecological significance of dry down rates. The importance of
-fuel moisture is not under debate. Furthermore, fine scale heterogeneity in
-fire behavior has been statistically associated wwith variation in fuel
+fuel moisture is well established and fire in the western US is often limited
+by fuel moisture (Parks et al 2014). Furthermore, fine scale heterogeneity in
+fire behavior has been statistically associated with variation in fuel
 moisture (Hille and Stephens 2005) and fire history can vary over short spatial
@@
 2005, Knapp et al 2006). These differences in fire behavior in turn control
-tree mortality ((Stephens et all 2002, Schwilk et al 2006). Despite this, we
-know of only one other study specifically investigating differential drying
-behavior (Kreye et al 2013).
+tree mortality (Stephens et al 2002, Schwilk et al 2006). Despite this, we know
+of only one other study specifically investigating differential drying behavior
+(Kreye et al 2013).
 
-TODO
-Add context also suggested by reviewer 1.
-
-\begin{quote}
-
-\end{quote}
-
-TODO
+We did not conduct a field experiment to compare drying rates in nature, but we
+expect our experimental design to accelerate drying and explain that now in the
+text. In terms of comparing fuel moisture with field measurements, Fuels can
+vary from single digits to over 100\% fuel moisture throughout the fire season.
+For example, 10-hour fuels in northern California varied from less than 10\% to
+over 80\% depending on precipitation events in a study following fuel moisture
+throughout the fire season (Estes et al 2012). Most measurements come from
+prescribed burns where prescription guidelines mean that fires are rarely
+attempted when 1 and 10 hour fuels are greater than 15\% moisture. For example,
+early season burns in this mixed conifer system show evidence of moisture
+limiting fire behavior at 12-14\% fuel moisture (Knapp et al 2005).
 
@@
 
-We have added information to the methods explaining the litter collection in more detail.
-All material was naturally dropped litter which could include fine twigs and leaves.
-Litter in these two \emph{Abies} species is a combination of branchlets with still attached needles and individual needles that have detached from the branchlet.
-We now cite the Zhao et al (2019) results when discussing the mixture results.
+We have added information to the methods explaining the litter collection in
+more detail. All material was naturally dropped litter which could include fine
+twigs and leaves. Litter in these two \emph{Abies} species is a combination of
+branchlets with still attached needles and individual needles that have
+detached from the branchlet. We now cite the Zhao et al (2019) results when
+discussing the mixture results.
 
-Yes, when deciding how to limit the number of tests at the comparitive scale (species as data points), we decided to not include size, because we believed its influence on flammability is primarily through litter packing and was therefore captured by litterbed density.
-We have added a figure (SI fig.
-1) to the SI materials showing litter particle size (the containing prism volumes) distributions for each of these species.
-These data were collected from the stored litter.
-Therefore, the data show the particle size distributions of the material that was burned.
-We did not add these data to the main figures as they are not part of any statistical test or main conclusions, but we are happy to do so if the editor suggests.
+When deciding how to limit the number of tests at the comparitive scale
+(species as data points), we decided to not include size, because we believed
+its influence on flammability is primarily through litter packing and was
+therefore captured by litterbed density. However, in response to the request to
+show particle sizes, we have added a figure (SI fig. 1) to the SI materials
+showing litter particle size (the containing prism volumes) distributions for
+each of these species. These data were collected from the stored litter. We did
+not add these data to the main figures as they are not part of any statistical
+test or main conclusions, but we are happy to do so if the editor suggests.
 
 \begin{quote}
-In the Methods, it was unclear how the various treatments and replicates were grouped over time, given that temperature and humidity in the fire facility could vary.
-Were there blocks in time with one replicate for each of the treatments (in random burning order) within a block?
-Or was there a different statistical design in time?
+  In the Methods, it was unclear how the various treatments and replicates were
+  grouped over time, given that temperature and humidity in the fire facility
+  could vary. Were there blocks in time with one replicate for each of the
+  treatments (in random burning order) within a block? Or was there a different
+  statistical design in time?
 \end{quote}
 
-Guided by the drying rates measured in the initial trials, we staggered starting saturation times across species.
-This was aimed at allowing a completely blocked design.
-However, because some ignition attempts were unsuccessful, some days lacked the full complement of species while others, later in the process, sometimes had more than one replicate per species burned.
+Guided by the drying rates measured in the initial trials, we staggered
+starting saturation times across species. This was aimed at allowing a
+completely blocked design. However, because some ignition attempts were
+unsuccessful, some days lacked the full complement of species while others,
+later in the process, sometimes had more than one replicate per species burned.
 We've added this information to the text [Lines X-X].
@@
 \begin{quote}
-Line 196 (see also reviewer 2): how can we assume that the quality of the stored litter is the same as that used in the previous study 3 years earlier, i.e. can these two datasets be safely combined?
+  Line 196 (see also reviewer 2): how can we assume that the quality of the
+  stored litter is the same as that used in the previous study 3 years earlier,
+  i.e. can these two datasets be safely combined?
 \end{quote}
@@
 at low humidity for several weeks on open racks. The litter was then bagged and
-sealed and stored within garbage bags inside cardboard boxes in a lab storage
-area. We saw no signs of mold nor ongoing decomposition. We minimized handling
-of the bags to avoid mechanical damage and changes to litter particle size.
+sealed and stored within cardboard boxes in a lab storage area. We saw no signs
+of mold nor ongoing decomposition. We minimized handling of the bags to avoid
+mechanical damage and changes to litter particle size.
 
 \begin{quote}
-Line 365 (see also reviewer 1): on the other hand Cornwell et al. 2015 New Phyt reported mostly high flammability in a range of Cupressaceae.
+  Line 365 (see also reviewer 1): on the other hand Cornwell et al. 2015 New
+  Phyt reported mostly high flammability in a range of Cupressaceae.
 \end{quote}
@@
 \begin{quote}
-This work compares mixtures of fuels, relative to their individual components, in terms of moisture retention and flammability.
-Moisture dynamics were also investigated in relation to properties of leaves of the species forming litter.
-This information is important in understanding flammability of surface fuels in the field.
+  This work compares mixtures of fuels, relative to their individual
+  components, in terms of moisture retention and flammability. Moisture
+  dynamics were also investigated in relation to properties of leaves of the
+  species forming litter. This information is important in understanding
+  flammability of surface fuels in the field.
 
-The study has been well thought through and well executed, and the ms is generally well written.
-Stylistically, the main problem is that commas are missing in many sentences.
-This might sound trivial, but will really help the reader.
+  The study has been well thought through and well executed, and the ms is
+  generally well written. Stylistically, the main problem is that commas are
+  missing in many sentences. This might sound trivial, but will really help the
+  reader.
 \end{quote}
 
-We thank the reviewer for their comments and apologize for the failure in
+We thank the reviewer for their comments and apologize for our failure in
 proofreading. The reviewer should not have had to conduct copy editing and we
@@
 fixed all specific grammar errors and missing symbols. For brevity, we've
-omitted responses to the very simple suggested grammar fixes and simply made all
-requested changes. We respond to other points individually.
+omitted responses to the very simple suggested grammar fixes and simply made
+all requested changes. We respond to other points individually.
 
 \begin{quote}
-line 72 - the comment about natural lightning ignitions comes from nowhere, and is a  bit of a jolt for the reader - rephrase slightly, e.g.
-many fires result from lightning ignitions, which are concurrent with ....
+  line 72 - the comment about natural lightning ignitions comes from nowhere,
+  and is a bit of a jolt for the reader - rephrase slightly, e.g. many fires
+  result from lightning ignitions, which are concurrent with ....
 \end{quote}
 
-This portion was poorly written and this was also mentioned by Reviewer 2.
-We have rewritten this in response to reviewer 2's comment that most fires result from dry lightning strikes across the west.
-See our responses to Reviewer 2, below.
+This portion was poorly written and this was also mentioned by Reviewer 2. We
+have rewritten this in response to reviewer 2's comment that most fires that
+spread result from dry lightning strikes across the west. See our responses to
+Reviewer 2, below.
 
 \begin{quote}
-line 103 - 'litter packing therefore affects fire behaviour in two ways' is easier to read
+  line 103 - 'litter packing therefore affects fire behaviour in two ways' is
+  easier to read
 \end{quote}
 
-Yes, agreed. We have tried to clean up the writing throughout and make the prose more active and precise.
+Yes, agreed. We have tried to clean up the writing throughout and make the
+prose more active and precise.
 
@@
 \begin{quote}
-line 220 - see also Prior et al (2018)  Conceptualizing ecological flammability ....Fire 1, article 14
+  line 220 - see also Prior et al (2018) Conceptualizing ecological
+  flammability ....Fire 1, article 14
 \end{quote}
@@
 \begin{quote}
-lines 243/244 I was a bit puzzled here --- It would help to be more specific here --- which aspects of moisture dynamics? Which response variables did you use?  (or 'as described below')
+  lines 243/244 I was a bit puzzled here --- It would help to be more specific
+  here --- which aspects of moisture dynamics? Which response variables did you
+  use? (or 'as described below')
 \end{quote}
@@
 \begin{quote}
-line 307 and line 326 - a minor inconsistency is that you analyse present species groups, yet discuss the results in terms of individual species
+  line 307 and line 326 - a minor inconsistency is that you analyse present
+  species groups, yet discuss the results in terms of individual species
 \end{quote}
 
-TODO!
-
-Yes. We have made edits to
-
-TODO. Check wording.
+We've tried to increase consistency throughout.
 
 \begin{quote}
-lines 348-349 - logically, this does not necessarily follow - factors such as large leaves or high SLA could independently affect both absorptive capacity and flammability - although fuel moisture does of course have a strong effect
+  lines 348-349 - logically, this does not necessarily follow - factors such as
+  large leaves or high SLA could independently affect both absorptive capacity
+  and flammability - although fuel moisture does of course have a strong effect
 \end{quote}
 
-Yes, we think that our logic must not have been clear here as we don't see the contradiction. We  have rewritten these sentences and we hope they are no longer confusing. See lines X-X.
+We think that our logic must not have been clear here as we don't see the
+contradiction. HOwever, We have rewritten these sentences and we hope they are
+no longer confusing. See lines X-X.
 
@@
 \begin{quote}
-line 365 - wording is problematic - change to '...ignition is possible, it had faster spread rates ...'
+  line 365 - wording is problematic - change to '...ignition is possible, it
+  had faster spread rates ...'
 \end{quote}
@@
 \begin{quote}
-line 365 - The low flammability of litter from your Cupressaceae species is interesting - Scarff and Westoby (2006) also found this for the Australian Cupressaceae genus Callitris, as did Bowman et al.
-(2018) J.
-Ecol 106, 1010 -1022
+  line 365 - The low flammability of litter from your Cupressaceae species is
+  interesting - Scarff and Westoby (2006) also found this for the Australian
+  Cupressaceae genus Callitris, as did Bowman et al. (2018) J. Ecol 106, 1010
+  -1022
 \end{quote}
 
-We have added some better discussion of past results and included discussion of these papers, as well as contrasting with the Cornwell et al 2015 result as pointed out by the Associate Editor.
-See lines XX-XX.
+We have added some better discussion of past results and included discussion of
+these papers. We also now contrast our result with that of Cornwell et al
+(2015) as suggested out by the Associate Editor. See lines XX-XX.
 
@@
 \begin{quote}
-Figures 7 and 8 - is there a reason that you don't present the information in the same way as in Figure 6, if your main point is the discrepancy between the mixtures and the averaged individual values?
+  Figures 7 and 8 - is there a reason that you don't present the information in
+  the same way as in Figure 6, if your main point is the discrepancy between
+  the mixtures and the averaged individual values?
 \end{quote}
 
-We agree this does lack consistency and an early draft used the same concise format of Fig 6 for all three figures.
-However, we decided that having time on the x-axis was worth the space and allows the reader more insight into the flammability effects of drying over time.
-Reviewer 2 pointed out some exceptions to the overall statistical pattern and those exceptions would not have been easily visible without an explicit time axis.
-There is a trade-off here between illustrating simply the results of the statistical model and allowing the reader to get a better idea of variation in flammability consequences of drying time.
+We agree this does lack consistency and an early draft used the same concise
+format of Fig 6 for all three figures. However, we decided that having time on
+the x-axis was worth the space and allows the reader more insight into the
+flammability effects of drying over time. Reviewer 2 pointed out some
+exceptions to the overall statistical pattern and those exceptions would not
+have been easily visible without an explicit time axis. There is a trade-off
+here between illustrating simply the results of the statistical model and
+allowing the reader to get a better idea of variation in flammability
+consequences of drying time.
 
 \begin{quote}
-line 416 - Prior et al (2018) also show that rate of spread and consumability are largely independent in litter fuels
+  line 416 - Prior et al (2018) also show that rate of spread and consumability
+  are largely independent in litter fuels
 \end{quote}
 
-Yes. That was an oversight. We have added citations here. See lines XX-XX
+Yes. That omission was an oversight. We have added citations here. See lines
+XX-XX
 
@@
   My major concerns are:
-1.
-One motivation for the study is not valid.
-I do not completely agree with the authors that natural lightning ignitions are concurrent with precipitation and, therefore, fire behavior can depend on rate of drying following a wetting event.
-The efficiency of individual lighting strikes in igniting a fire is affected by the moisture properties of forest fuel at the moment of lighting (Krawchuket et al.
-2006, Ecology).
-Dry lighting that occurs without any accompanied rainfall nearby is most likely to cause wildfire because the fuel is dry enough to be ignited.
+
+  1. One motivation for the study is not valid. I do not completely agree with
+  the authors that natural lightning ignitions are concurrent with
+  precipitation and, therefore, fire behavior can depend on rate of drying
+  following a wetting event. The efficiency of individual lighting strikes in
+  igniting a fire is affected by the moisture properties of forest fuel at the
+  moment of lighting (Krawchuk et al. 2006, Ecology). Dry lighting that occurs
+  without any accompanied rainfall nearby is most likely to cause wildfire
+  because the fuel is dry enough to be ignited.
 \end{quote}
 
-This sentence was poorly written and oversimplified. We thank the reviewer for
+Our sentence was poorly written and oversimplified. We thank the reviewer for
 pointing this out and we have rewritten this section. However, our badly made
@@
 can occur under overlapping conditions, Rorig and Ferguson 1999). We've
-rewritten this section, see lines XX-XX.
-
-Cite:
-
-Article{rorig1999characteristics,
-  title={Characteristics of lightning and wildland fire ignition in the Pacific Northwest},
-  author={Rorig, Miriam L and Ferguson, Sue A},
-  journal={Journal of Applied Meteorology},
-  volume={38},
-  number={11},
-  pages={1565--1575},
-  year={1999}
-}
-
-@article{abatzoglou2016controls,
-  title={Controls on interannual variability in lightning-caused fire activity in the western US},
-  author={Abatzoglou, John T and Kolden, Crystal A and Balch, Jennifer K and Bradley, Bethany A},
-  journal={Environmental Research Letters},
-  volume={11},
-  number={4},
-  pages={045005},
-  year={2016},
-  publisher={IOP Publishing}
-}
-
-http://www.publish.csiro.au/wf/WF08117
-
-@article{van2008temporal,
-  title={Temporal and spatial distribution of lightning strikes in California in relation to large-scale weather patterns},
-  author={Van Wagtendonk, Jan W and Cayan, Daniel R},
-  journal={Fire Ecology},
-  volume={4},
-  number={1},
-  pages={34--56},
-  year={2008},
-  publisher={Springer}
-}
+rewritten this section and added details justifying the importance of drying
+rates in fine fuels as suggested by the Senior Editor. See lines XX-XX.
 
 \begin{quote}
-2.
-The research question of the research is not explicitly stated both in the abstract and in the introduction.
-In the end of the introduction, the authors focus on specific data-collection goals, while leave the reader to figure out what the research question is.
+  2. The research question of the research is not explicitly stated both in the
+  abstract and in the introduction. In the end of the introduction, the authors
+  focus on specific data-collection goals, while leave the reader to figure out
+  what the research question is.
 \end{quote}
 
-We have edited the abstract to make the research questions clearer. Yes, our
-first submission jumped from stating current reserach gaps to specific methods
-too quickly. See also our response to the Senior Editor, above.
+We have edited the abstract to make the research questions clearer. We agree
+that our first submission jumped from stating current reserach gaps to specific
+methods too quickly. See also our response to the Senior Editor, above.
 
 \begin{quote}
-3.
-The authors concluded that positive non-additivity appeared to be the rule in litter-driven fire and this non-additivity became more positive with fuel drying.
-But if we look closely at figure 8, non-additivity was negative within 50 hours since wetting, non-additivity was both positive and negative at about 75 hours since wetting, and non-additivity was positive at about 100 hours since wetting.
-The high variability of non-additive effects at about 75 hours since wetting might have important ecological effects on fire ignition and behavior, and are important information cannot be ignored.
+  3. The authors concluded that positive non-additivity appeared to be the rule
+  in litter-driven fire and this non-additivity became more positive with fuel
+  drying. But if we look closely at figure 8, non-additivity was negative
+  within 50 hours since wetting, non-additivity was both positive and negative
+  at about 75 hours since wetting, and non-additivity was positive at about 100
+  hours since wetting. The high variability of non-additive effects at about 75
+  hours since wetting might have important ecological effects on fire ignition
+  and behavior, and are important information cannot be ignored.
 \end{quote}
 
-Yes, this is a good point and we have added some discussion of this.
-We tried to limit our statements to those supported by the statistical tests (eg, slope of nonaddivity with drying was greater than 1), and therefore focused on the positive effect of drying and the overall trend for positive nonadditivity.
-This, of course, obscures some details which are worth mentioning and we have edited this section in response to the reviewer's criticism.
-We agree that the high variability in nonadditivity across mixtures deserves highlighting and that is the reason we presented the actual predictions and departures from prediction across time.
-Please see our changes on lines X-X.
+This is a good point and we have added some discussion of this. We tried to
+limit our statements to those supported by the statistical tests (eg, the slope of
+nonaddivity with drying was greater than 1), Therefore, we focused on the
+positive effect of drying and the overall trend for positive nonadditivity.
+This, of course, obscures some details which are worth mentioning and we have
+edited this section in response to the reviewer's criticism. We agree that the
+high variability in nonadditivity across mixtures deserves highlighting and
+that is the reason we presented the actual predictions and departures from
+prediction across time. Please see our changes on lines X-X.
 
 \begin{quote}
-4. The writing of this manuscript needs to be improved for clarity and accuracy. There are many typos in the text, and the using to scientific terms lack consistency through the text.
+  4. The writing of this manuscript needs to be improved for clarity and
+  accuracy. There are many typos in the text, and the using to scientific terms
+  lack consistency through the text.
 \end{quote}
@@
 We have made extensive edits in response to both reviewer's suggestions.
+Furthermore, we have standardized our usage of some terms. We use the term
+``maximum water absorption'' to indicate water content following saturation. Of
+course, this is slightly inaccurate as the processes involved include both
+absorption and adsorption. Our earlier draft often used the term ``maximum
+water retention''. But we think this earlier wording was confusing as retention
+implied and inverse to drying (and we were inconsistent in usage as the
+reviewer points out).
 
 \begin{quote}
-5. The title does not precisely communicate the key message of the study. I suggest it is more precise to say “moisture effect … and alter non-additive flammability…”
+  5. The title does not precisely communicate the key message of the study. I
+  suggest it is more precise to say “moisture effect … and alter non-additive
+  flammability…”
 \end{quote}
@@
 \begin{quote}
-6. The last paragraph of the discussion has a significant overlap with the last paragraph of the discussion in the authors’ previous paper (de Magalhães \& Schwilk 2012) in terms of the main content.
+  6. The last paragraph of the discussion has a significant overlap with the
+  last paragraph of the discussion in the authors’ previous paper (de Magalhães
+  \& Schwilk 2012) in terms of the main content.
 \end{quote}
 
-TODO
-
-We have rewritten much of the discussion to better include findings from the literature and rewritten the concluding paragraph.
+We have rewritten much of the discussion to better include findings from the
+literature and rewritten the concluding paragraph.
 
@@
 
-We have editing the abstract to more clearly state the research questions reduce its length.
+We have editing the abstract to more clearly state the research questions
+reduce its length.
 
@@
 
-We thought this was an established enough fact to not need citation but have added one.
+We have added one.
 
 \begin{quote}
-Lines 50-61: in this paragraph, after the first sentence I think it is more logic to first explain the mechanism of moisture’s negative effects on fire, then explain the reason of the driving importance of dead fuel moisture content to surface fire ignition and behavior because fine dead fuels respond more readily to environmental changes than living fuels.
+  Lines 50-61: in this paragraph, after the first sentence I think it is more
+  logic to first explain the mechanism of moisture’s negative effects on fire,
+  then explain the reason of the driving importance of dead fuel moisture
+  content to surface fire ignition and behavior because fine dead fuels respond
+  more readily to environmental changes than living fuels.
 \end{quote}
@@
 \begin{quote}
-Line 72: revise the description “ litter fuel driven system”.
+Line 72: revise the description ``litter fuel driven system''.
 \end{quote}
@@
 \begin{quote}
-Line 85: revise “driven by by” and “emergent litter litter”.
+Line 85: revise ``driven by by'' and ``emergent litter litter''.
 \end{quote}
@@
 
-We were not sure what the reviewer's criticism was here but have rewritten this system to be more detailed and precise.
+We were not sure what the reviewer's criticism was here but have rewritten this
+section to be more detailed and precise.
 
 \begin{quote}
-Line 95: “absorption capacity” and “rate of moisture loss” are not correct descriptions of processes. Better say “absorption and retention”.
+  Line 95: “absorption capacity” and “rate of moisture loss” are not correct
+  descriptions of processes. Better say “absorption and retention”.
 \end{quote}
 
-Good point.
-We fixed that here and in another sentence elsewhere.
-Furthermore, we've edited to be more precise about what we mean by ``absorption'' because we are referring to a litter bed and we include adsorption to litter surfaces in that definition.
-We had originally tried to avoid the chemical terms because we are observing a higher-level phenomenon.
+Good point. We fixed that here and in another sentence elsewhere. Furthermore,
+we've edited to explain what we mean by ``absorption'' because we are referring
+to a litter bed and we include adsorption to litter surfaces in that
+definition. We had originally tried to avoid the chemical terms because we are
+observing a higher-level phenomenon but our previous language was confusing.
 
 \begin{quote}
-Line 98: revise “after saturation” to “dried certain time after saturation” for clarity.
+  Line 98: revise “after saturation” to “dried certain time after saturation”
+  for clarity.
 \end{quote}
@@
 \begin{quote}
-Lines 102-103: litter bulk density is not a litter particle trait. Delete “will influence air flow through the litter bed”.
+  Lines 102-103: litter bulk density is not a litter particle trait. Delete
+  “will influence air flow through the litter bed”.
 \end{quote}
@@
 \begin{quote}
-Line 107: “fuel-beds composed of larger leaves with lower moisture retainability”?
+  Line 107: “fuel-beds composed of larger leaves with lower moisture
+  retainability”?
 \end{quote}
 
-We edited to ``composed of larger leaves or leaves with lower moisture retainability''
+We edited to ``composed of larger leaves or leaves with lower moisture
+retainability''
 
@@
 \begin{quote}
-Lines 117-119: better specify the system, because negative non-additive effects on litter
-flammability were also found.
+  Lines 117-119: better specify the system, because negative non-additive
+  effects on litter flammability were also found.
 \end{quote}
@@
 \begin{quote}
-Line 120: “especially if different leaf traits influence absorptive capacity and drying rate”?
+  Line 120: “especially if different leaf traits influence absorptive capacity
+  and drying rate”?
 \end{quote}
 
-TODO?
+We rewrote these sentences.
 
 \begin{quote}
-Line 125: revise “two leaf and litter bed traits”, and better directly say “SLA and packing density”.
+  Line 125: revise “two leaf and litter bed traits”, and better directly say
+  “SLA and packing density”.
 \end{quote}
@@
 
-TODO
-
 We have rewritten this. See line X.
@@
 
-TODO
-
-We assume that the reviewer's complaint is that ``ecologically-relevant'' is vague.
+We assume that the reviewer's complaint is that ``ecologically-relevant'' is
+vague. We have revised this and removed that term.
 
@@
 \begin{quote}
-Line 147: at least for Quercus, leaves fallen in the previous year will be partly decomposed in the summer.
+  Line 147: at least for Quercus, leaves fallen in the previous year will be
+  partly decomposed in the summer.
 \end{quote}
 
-Yes. We have added some detail to this and also now provide particle size distributions for each species measured on the actual stored litter (SI fig 1).
+We have added some detail to this and also now provide particle size
+distributions for each species measured on the actual stored litter (SI fig 1).
 
@@
 
-We have added information to the methods explaining the collection in more detail. See lines XX and our response to the Associate Editor, above.
+We have added information to the methods explaining the collection in more
+detail. See lines XX and our response to the Associate Editor, above.
 
 \begin{quote}
-Line 157: is it possible that all litter has less than 5\% fuel moisture when air-dried?
+  Line 157: is it possible that all litter has less than 5\% fuel moisture when
+  air-dried?
 \end{quote}
 
-We have corrected the methods because the litter was initially oven dried in large paper bags, then stored on racks in the low humidity lab.
-Our subsample measurements showed less than 5\% moisture.
-See lines X-X.
+We have corrected the methods because the litter was initially oven dried in
+large paper bags, then stored on racks in the low humidity lab. Our subsample
+measurements showed less than 5\% moisture. See lines X-X.
 
 \begin{quote}
-Line 158: the collecting of the leaf litter is in 2012. The conduction of the experiment is in 2015. Can the author explain how do they save the litter?
+  Line 158: the collecting of the leaf litter is in 2012. The conduction of the
+  experiment is in 2015. Can the author explain how do they save the litter?
 \end{quote}
@@
 
-Yes, this was a typo. Thanks. The aluminum mesh openings were 0.203 inches. We have clarified and corrected to ``0.52\,cm''
+Yes, this was a mistake. Thanks. The aluminum mesh openings were 0.203 inches.
+We have clarified and corrected to ``0.52\,cm''
 
@@
 \begin{quote}
-Line 182: revise “effect” to “affect”. 	Revise “litter bulk density” to “litter packing density”
+  Line 182: revise “effect” to “affect”. Revise “litter bulk density” to
+  “litter packing density”
 \end{quote}
@@
 \begin{quote}
-Line 185: I do not think this sentence is necessary because whether or not petioles should be included in SLA measurements depends on the research question at hand.
+  Line 185: I do not think this sentence is necessary because whether or not
+  petioles should be included in SLA measurements depends on the research
+  question at hand.
 \end{quote}
@@
 
-No, mass was fixed, width and length of the burning tray was fixed (15\,cm by 1.5\,m), and depth was allowed to vary.  We edited this for clarity. See lines X-X
+No, mass was fixed, width and length of the burning tray was fixed (15\,cm by
+1.5\,m), and depth was allowed to vary. We edited this section for clarity. See
+lines X-X
 
@@
 \begin{quote}
-Line 208: revise “species/moisture combination” to “species and moisture combination”.
+  Line 208: revise “species/moisture combination” to “species and moisture
+  combination”.
 \end{quote}
@@
 \begin{quote}
-Line 224: if not all flammability data was used in the analysis, then why measure them?
+  Line 224: if not all flammability data was used in the analysis, then why
+  measure them?
 \end{quote}
@@
 \begin{quote}
-Line 242: revise “one leaf trait and one litter trait” to “one litter particle trait and one litter bed traits”.
+  Line 242: revise “one leaf trait and one litter trait” to “one litter
+  particle trait and one litter bed traits”.
 \end{quote}
@@
 \begin{quote}
-Lines 314-320: this indicates that even in the same genus, species vary in flammability.
+  Lines 314-320: this indicates that even in the same genus, species vary in
+  flammability.
 \end{quote}
 
-Yes.  This is not surprising in the case of these three pines where one has quite different needles. But our \emph{a priori} groups were based on genus.
+Yes. This is not surprising in the case of these three pines where one has
+quite different needles. But our \emph{a priori} groups were based on genus.
 
@@
 
-TODO
+We have revised.
 
@@
 
-Agreed. That was vague. Fixed.
+We agree such wording was vague. Fixed.
 
@@
 
-TODO.
+We've rewritten these sentences as ``Our two Cupressaceae, which have dense
+litter and low maximum water absorption, had low flammability across moisture
+levels.''
 
@@
 
-We removed the offending word.
+We removed the word.
 
@@
 
-We don't understand the criticism. Perhaps it is the word ``alone'' that is ambiguous?  We changed ``alone'' to ``in monoculture'' although we thought the original wording was clear.
+We don't understand the criticism. Perhaps the word ``alone'' is ambiguous? We
+changed ``alone'' to ``in monoculture'' and hope that is clearer.
 
@@
 
-We don't understand the criticism. Academic writing commonly uses present tense to describe results that still apply. There is no grammatical error.
+We don't understand the criticism. Is the complaint with present tense?
+Academic writing commonly uses present tense to describe results that still
+apply. There is no grammatical error.
 
@@
 
-TODO
+Fixed.
 

### 2019-12-19 ms/response_to_reviews.tex
fixup

@@
 \usepackage{times, fullpage}
+\usepackage[font=itshape]{quoting}
 \usepackage{url}
@@
 
-We thank the editor and reviewer for their helpful comments and suggestions. We
-have itemized each comment below and followed each with our reply and a list of
-changes we made in response to the comment.
+We thank the editors and reviewers for their substantial comments and
+suggestions. We have itemized each comment below and followed each with our
+reply and a description of changes we made in response to the comment.
 
-{\bf Responses to Senior Editor's comments:}
+{\bf Responses to the Senior Editor's comments:}
 
-\begin{quote}
+\begin{quoting}
   I support the AE's recommendation to reject but allow resubmission. However,
@@
   consider a resubmission.
-\end{quote}
+\end{quoting}
 
@@
 by fuel moisture (Parks et al 2014). Furthermore, fine scale heterogeneity in
-fire behavior has been statistically associated with variation in fuel
-moisture (Hille and Stephens 2005) and fire history can vary over short spatial
-scales as a consequence of species composition (Stephens 2001). Although the
+fire behavior has been statistically associated with variation in fuel moisture
+(Hille and Stephens 2005) and fire history can vary over short spatial scales
+as a consequence of species composition (Stephens 2001). Although the
 importance of dry-down rates is less well-established, prescribed burns in
@@
 
-We did not conduct a field experiment to compare drying rates in nature, but we
-expect our experimental design to accelerate drying and explain that now in the
-text. In terms of comparing fuel moisture with field measurements, Fuels can
-vary from single digits to over 100\% fuel moisture throughout the fire season.
-For example, 10-hour fuels in northern California varied from less than 10\% to
-over 80\% depending on precipitation events in a study following fuel moisture
-throughout the fire season (Estes et al 2012). Most measurements come from
-prescribed burns where prescription guidelines mean that fires are rarely
-attempted when 1 and 10 hour fuels are greater than 15\% moisture. For example,
-early season burns in this mixed conifer system show evidence of moisture
-limiting fire behavior at 12-14\% fuel moisture (Knapp et al 2005).
+The editor asks that we compare fuel moisture with field measurements. Fuels
+can vary from single digits to over 100\% fuel moisture throughout the fire
+season. For example, relatively fine fuels in northern California varied from
+less than 10\% to over 80\% depending on precipitation events in a study
+following fuel moisture throughout the fire season (Estes et al 2012). Most
+measurements come from prescribed burns where prescription guidelines mean that
+fires are rarely attempted when 1-hour and 10-hour fuels are greater than 15\%
+moisture. For example, early season burns in this mixed conifer system show
+evidence of moisture limiting fire behavior at 12-14\% fuel moisture (Knapp et
+al 2005).
 
-{\bf Responses to Associate Editor's comments:}
+{\bf Responses to the Associate Editor's comments:}
 
-\begin{quote}
+\begin{quoting}
   First, apologies to the authors for the long delay in the review process.
-Because of the rather contrasting assessments of the first two reviewers, I was waiting for the third reviewer to produce his/her review, but unfortunately we waited in vain.
-So I have to make a recommendation based on these two contrasting assessments plus my own judgement.
-On the one hand, in agreement with reviewer 1, I can see the novelty of this study.
-So far non-additivity in litter mixture flammability has been studied either at air-dry litter moisture or at fixed levels of litter moisture (e.g.
-Blauw et al 2015) and I do applaud the approach of bringing more field realism into these studies by allowing mixtures to dry out at different rates from expected values based on their single components.
-Having said that, there are also major problems with the current manuscript, as reported especially by reviewer 2 (and some additional ones by myself, see below).
-Besides these major problems there is a rather high degree of inaccuracy in the text.
-Therefore, given the high rejection rates of often even good manuscripts to J.
-Ecol., I have to make a negative recommendation for this manuscript.
-However, given the novelty and principally nice design of the experiment, I would allow the authors to submit a strongly revised new manuscript, addressing all major and minor comments by the reviewers and myself, as a new submission.
-(At least one of the reviewers to check the revision would be new.) In addition to the points raised by the reviewers, I like to add some of my own:
-\end{quote}
+  Because of the rather contrasting assessments of the first two reviewers, I
+  was waiting for the third reviewer to produce his/her review, but
+  unfortunately we waited in vain. So I have to make a recommendation based on
+  these two contrasting assessments plus my own judgement. On the one hand, in
+  agreement with reviewer 1, I can see the novelty of this study. So far
+  non-additivity in litter mixture flammability has been studied either at
+  air-dry litter moisture or at fixed levels of litter moisture (e.g. Blauw et
+  al 2015) and I do applaud the approach of bringing more field realism into
+  these studies by allowing mixtures to dry out at different rates from
+  expected values based on their single components. Having said that, there are
+  also major problems with the current manuscript, as reported especially by
+  reviewer 2 (and some additional ones by myself, see below). Besides these
+  major problems there is a rather high degree of inaccuracy in the text.
+  Therefore, given the high rejection rates of often even good manuscripts to
+  J. Ecol., I have to make a negative recommendation for this manuscript.
+  However, given the novelty and principally nice design of the experiment, I
+  would allow the authors to submit a strongly revised new manuscript,
+  addressing all major and minor comments by the reviewers and myself, as a new
+  submission. (At least one of the reviewers to check the revision would be
+  new.) In addition to the points raised by the reviewers, I like to add some
+  of my own:
+\end{quoting}
 
-Thank you.
-Several comments converged on the absolutely correct criticism that our Discussion did not engage adequately with the literature.
-We have made substantial changes to the Introduction and the Discussion adding more context and citing more of the recent literature.
-We reorganized the Introduction and now do not rush the reader so quickly through the ideas.
-We have expanded our justification for the study as argued for by the Senior Editor.
+Thank you. Several comments converged on the absolutely correct criticism that
+our Discussion did not engage adequately with the literature. We have made
+substantial changes to the Introduction and the Discussion adding more context
+and citing more of the recent literature. We reorganized the Introduction and
+now do not rush the reader so quickly through the ideas. We have expanded our
+justification for the study as argued for by the Senior Editor.
 
-\begin{quote}
-The authors mention leaf size explicitly as an important parameter in non-additivity (with some references including their own 2012 paper) but they do not include it in the current study.
-Especially at low moisture, the small individual needles of Abies should cause low ventilation and a high likelihood of negative non-additivity on flammability, as was shown by Zhao et al 2016 Ecol.
-Evol.
-Figure 4).
-In the Methods, as observed also by reviewer 2, there is very little detail on the extent to which litter of the respective species was picked up as individual leaves, branchlets without leaves versus branchlets with leaves still attached.
-This is really important, also because positive versus negative non-additivity of leaf-twig mixtures have been reported recently (Zhao et al.
-2019 Plant and Soil).
-I feel that data and/or photos indicating the size and shape of the different litter types (and if possible variability in these) are important to add and to be used in the interpretation of the results.
-\end{quote}
+\begin{quoting}
+  The authors mention leaf size explicitly as an important parameter in
+  non-additivity (with some references including their own 2012 paper) but they
+  do not include it in the current study. Especially at low moisture, the small
+  individual needles of Abies should cause low ventilation and a high
+  likelihood of negative non-additivity on flammability, as was shown by Zhao
+  et al 2016 Ecol. Evol. Figure 4). In the Methods, as observed also by
+  reviewer 2, there is very little detail on the extent to which litter of the
+  respective species was picked up as individual leaves, branchlets without
+  leaves versus branchlets with leaves still attached. This is really
+  important, also because positive versus negative non-additivity of leaf-twig
+  mixtures have been reported recently (Zhao et al. 2019 Plant and Soil). I
+  feel that data and/or photos indicating the size and shape of the different
+  litter types (and if possible variability in these) are important to add and
+  to be used in the interpretation of the results.
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   In the Methods, it was unclear how the various treatments and replicates were
@@
   statistical design in time?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 196 (see also reviewer 2): how can we assume that the quality of the
@@
   i.e. can these two datasets be safely combined?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 365 (see also reviewer 1): on the other hand Cornwell et al. 2015 New
   Phyt reported mostly high flammability in a range of Cupressaceae.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   This work compares mixtures of fuels, relative to their individual
@@
   reader.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   line 72 - the comment about natural lightning ignitions comes from nowhere,
@@
   result from lightning ignitions, which are concurrent with ....
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   line 103 - 'litter packing therefore affects fire behaviour in two ways' is
   easier to read
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   line 133 - 'in mixtures of litter with varying levels..' is more explicit
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 line 182 - affect, not effect
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   line 220 - see also Prior et al (2018) Conceptualizing ecological
   flammability ....Fire 1, article 14
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   lines 243/244 I was a bit puzzled here --- It would help to be more specific
@@
   use? (or 'as described below')
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 line 290 - you show an association, not necessarily an effect
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   line 307 and line 326 - a minor inconsistency is that you analyse present
   species groups, yet discuss the results in terms of individual species
-\end{quote}
+\end{quoting}
 
-We've tried to increase consistency throughout.
+We've tried to increase consistency throughout. We refer to taxonomic groups
+when appropriate and have standardized the subsection headings.
 
-\begin{quote}
+\begin{quoting}
   lines 348-349 - logically, this does not necessarily follow - factors such as
@@
   and flammability - although fuel moisture does of course have a strong effect
-\end{quote}
+\end{quoting}
 
-We think that our logic must not have been clear here as we don't see the
-contradiction. HOwever, We have rewritten these sentences and we hope they are
-no longer confusing. See lines X-X.
+We have rewritten these sentences and we hope they are no longer confusing. See
+lines X-X.
 
-\begin{quote}
+\begin{quoting}
 Table 4 - get rid of the unnecessary zeroes in the df column
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   line 365 - wording is problematic - change to '...ignition is possible, it
   had faster spread rates ...'
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   line 365 - The low flammability of litter from your Cupressaceae species is
@@
   -1022
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 line 387 - capitalise 'I' in SI Table  8
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 line 403 - change to 'to the completely dry litter mixtures studied in previous work'
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Figures 7 and 8 - is there a reason that you don't present the information in
@@
   the mixtures and the averaged individual values?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   line 416 - Prior et al (2018) also show that rate of spread and consumability
   are largely independent in litter fuels
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 
@@
   because the fuel is dry enough to be ignited.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   2. The research question of the research is not explicitly stated both in the
@@
   what the research question is.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   3. The authors concluded that positive non-additivity appeared to be the rule
@@
   and behavior, and are important information cannot be ignored.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   4. The writing of this manuscript needs to be improved for clarity and
@@
   lack consistency through the text.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   5. The title does not precisely communicate the key message of the study. I
@@
   flammability…”
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   6. The last paragraph of the discussion has a significant overlap with the
@@
   \& Schwilk 2012) in terms of the main content.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 The abstract can be shorter. It exceeds 350 words limit. 
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 Line 50: add reference to the first sentence.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Lines 50-61: in this paragraph, after the first sentence I think it is more
@@
   more readily to environmental changes than living fuels.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 Line 72: revise the description ``litter fuel driven system''.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 Line 75: “at any particular time”?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 Line 85: revise ``driven by by'' and ``emergent litter litter''.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 Line 88: “similarly”?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 95: “absorption capacity” and “rate of moisture loss” are not correct
   descriptions of processes. Better say “absorption and retention”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 98: revise “after saturation” to “dried certain time after saturation”
   for clarity.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 101: delete “also”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Lines 102-103: litter bulk density is not a litter particle trait. Delete
   “will influence air flow through the litter bed”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Lines 103-105: leaf size is not leaf litter trait.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 107: “fuel-beds composed of larger leaves with lower moisture
   retainability”?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 115: start a new paragraph here.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Lines 117-119: better specify the system, because negative non-additive
   effects on litter flammability were also found.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 120: “especially if different leaf traits influence absorptive capacity
   and drying rate”?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 125: revise “two leaf and litter bed traits”, and better directly say
   “SLA and packing density”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 133: delete “also”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 134: “if species contribute equally to litter moisture dynamics”?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 136: delete “on”. Revise “ecologically-relevant”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 Line 141: revise “(36 36N, 118 42 W)”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 147: at least for Quercus, leaves fallen in the previous year will be
   partly decomposed in the summer.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 155: add more details about how the litter was collected?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 157: is it possible that all litter has less than 5\% fuel moisture when
   air-dried?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 158: the collecting of the leaf litter is in 2012. The conduction of the
   experiment is in 2015. Can the author explain how do they save the litter?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 Line 167: 1 cm ? aluminum screen.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 173: revise “21C temperature”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 182: revise “effect” to “affect”. Revise “litter bulk density” to
   “litter packing density”
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 185: I do not think this sentence is necessary because whether or not
@@
   question at hand.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 201: the fuel depth was controlled or not?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 Line 205: revise “1899C”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 208: revise “species/moisture combination” to “species and moisture
   combination”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 209: “the replicate was allowed to continue drying for …”
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 216: revise “mass loss” to “fuel consumption”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 224: if not all flammability data was used in the analysis, then why
   measure them?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 242: revise “one leaf trait and one litter trait” to “one litter
   particle trait and one litter bed traits”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 307: revise “at that point”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Lines 314-320: this indicates that even in the same genus, species vary in
   flammability.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 Line 330: “rates”?
-\end{quote}
+\end{quoting}
 
-TODO
+Fixed.
 
-\begin{quote}
+\begin{quoting}
   Line 344: “moisture dynamics flammability” is difficult to understand.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 346: revise “apparently controlled”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 350: revise “very dry”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
 Line 351: revise “more realistic moisture levels”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 361: revise “that that”.
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 366: “but”?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 401: “striking”？
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 410: “not met by the constituent species alone”?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 415: “demonstrates”?
-\end{quote}
+\end{quoting}
 
@@
 
-\begin{quote}
+\begin{quoting}
   Line 439: revise “to”.
-\end{quote}
+\end{quoting}
 

### 2019-12-20 ms/response_to_reviews.tex
Close to final R2?

@@
+% Response letter 2019-12-20
 
@@
 
-Enclosed is our revised manuscript research article, ``Moisture effect on leaf
-litter alters non-additive flammability in mixed conifer forest''.
+Enclosed is our revised manuscript research article, ``Moisture absorption and drying alters nonadditive litter flammability in mixed conifer forest''.
 
@@
 
-We agree that our prior Discussion was weak and have extensively rewritten it.
-We argue that this work is not merely confirmatory. Our revised Introduction
-argues for the ecological significance of dry down rates. The importance of
-fuel moisture is well established and fire in the western US is often limited
-by fuel moisture (Parks et al 2014). Furthermore, fine scale heterogeneity in
-fire behavior has been statistically associated with variation in fuel moisture
-(Hille and Stephens 2005) and fire history can vary over short spatial scales
-as a consequence of species composition (Stephens 2001). Although the
-importance of dry-down rates is less well-established, prescribed burns in
-Sierra Nevada mixed conifer forest demonstrate that fuel moisture likely drives
-differences in fire behavior depending on the timing of fire throughout the
-season (Kaufman and Martin 1989, Knapp and Keeley 2006, Hille and Stephens
-2005, Knapp et al 2006). These differences in fire behavior in turn control
-tree mortality (Stephens et al 2002, Schwilk et al 2006). Despite this, we know
-of only one other study specifically investigating differential drying behavior
-(Kreye et al 2013).
+We agree that our Discussion was weak and we have rewritten it. We argue,
+however, that this work is not merely confirmatory. Our revised Introduction
+argues for the ecological significance of drying rates. The importance of fuel
+moisture is well established. At large scales, fire in the western US is often
+limited by fuel moisture (Parks et al 2014). At fine scales, variation in fuel
+moisture influences heterogeneity in fire behavior (Hille and Stephens 2005)
+and may contribute to fire history variation over short distances (Stephens
+2001). Although the importance of drying rate is less well-established,
+prescribed burns in Sierra Nevada mixed conifer forest demonstrate that fuel
+moisture drives differences in fire behavior depending on the timing of fire
+throughout the season (Kaufman and Martin 1989, Knapp and Keeley 2006, Hille
+and Stephens 2005, Knapp et al 2006). These differences in fire behavior in
+turn cause spatial variation in tree mortality (Stephens et al 2002, Schwilk et
+al 2006). Despite this, we know of only one other study specifically
+investigating differential drying behavior (Kreye et al 2013b) and none that
+included drying rates with flammability measurement.
 
-The editor asks that we compare fuel moisture with field measurements. Fuels
-can vary from single digits to over 100\% fuel moisture throughout the fire
-season. For example, relatively fine fuels in northern California varied from
-less than 10\% to over 80\% depending on precipitation events in a study
-following fuel moisture throughout the fire season (Estes et al 2012). Most
-measurements come from prescribed burns where prescription guidelines mean that
-fires are rarely attempted when 1-hour and 10-hour fuels are greater than 15\%
-moisture. For example, early season burns in this mixed conifer system show
-evidence of moisture limiting fire behavior at 12-14\% fuel moisture (Knapp et
-al 2005).
+The editor asks that we compare fuel moisture with field measurements. Litter
+fuels can vary from single digits to over 100\% fuel moisture throughout the
+fire season. For example, relatively fine fuels in northern California varied
+from less than 10\% to over 80\% in response to repeated precipitation events
+throughout the fire season (Estes et al 2012). Early season burns in this mixed
+conifer system show evidence of moisture limiting fire behavior at 12-14\% fuel
+moisture (Knapp et al 2005). We have added this information to the Discussion
+(lines 396--404).
 
@@
 
-Thank you. Several comments converged on the absolutely correct criticism that
-our Discussion did not engage adequately with the literature. We have made
-substantial changes to the Introduction and the Discussion adding more context
-and citing more of the recent literature. We reorganized the Introduction and
-now do not rush the reader so quickly through the ideas. We have expanded our
-justification for the study as argued for by the Senior Editor.
+Thank you for the consideration and comments. Several reviewer comments
+converged on the correct criticism that our Discussion did not engage
+adequately with the literature. We have made substantial changes to the
+Introduction and the Discussion adding more context and citing more of the
+recent literature. We reorganized the Introduction and now do not rush the
+reader so quickly through the ideas. We have expanded our justification for the
+study as argued for by the Senior Editor and others.
 
@@
   likelihood of negative non-additivity on flammability, as was shown by Zhao
-  et al 2016 Ecol. Evol. Figure 4). In the Methods, as observed also by
-  reviewer 2, there is very little detail on the extent to which litter of the
-  respective species was picked up as individual leaves, branchlets without
-  leaves versus branchlets with leaves still attached. This is really
-  important, also because positive versus negative non-additivity of leaf-twig
-  mixtures have been reported recently (Zhao et al. 2019 Plant and Soil). I
-  feel that data and/or photos indicating the size and shape of the different
-  litter types (and if possible variability in these) are important to add and
-  to be used in the interpretation of the results.
+  et al 2016 Ecol. Evol. Figure 4).
+\end{quoting}
+
+We have added a new figure in response to this comment (SI Fig 1).
+
+When deciding how to limit the number of tests at the comparative scale
+(species as data points), we decided to not include size, because its influence
+on flammability is primarily through litter packing and was captured by
+litterbed density. However, in response to the request to show particle sizes,
+we have added a figure (SI Fig. 1) to the SI materials showing litter particle
+size (the containing prism volumes) distributions for each of these species.
+These data were collected from the stored litter. We did not add these data to
+the main figures as they are not part of any statistical test or main
+conclusions, but we are happy to do so if the editor suggests.
+
+\begin{quoting}
+  In the Methods, as observed also by reviewer 2, there is very little detail
+  on the extent to which litter of the respective species was picked up as
+  individual leaves, branchlets without leaves versus branchlets with leaves
+  still attached. This is really important, also because positive versus
+  negative non-additivity of leaf-twig mixtures have been reported recently
+  (Zhao et al. 2019 Plant and Soil). I feel that data and/or photos indicating
+  the size and shape of the different litter types (and if possible variability
+  in these) are important to add and to be used in the interpretation of the
+  results.
 \end{quoting}
@@
 We have added information to the methods explaining the litter collection in
-more detail. All material was naturally dropped litter which could include fine
+more detail. All material was naturally dropped litter which could include
 twigs and leaves. Litter in these two \emph{Abies} species is a combination of
 branchlets with still attached needles and individual needles that have
-detached from the branchlet. We now cite the Zhao et al (2019) results when
-discussing the mixture results.
-
-When deciding how to limit the number of tests at the comparitive scale
-(species as data points), we decided to not include size, because we believed
-its influence on flammability is primarily through litter packing and was
-therefore captured by litterbed density. However, in response to the request to
-show particle sizes, we have added a figure (SI fig. 1) to the SI materials
-showing litter particle size (the containing prism volumes) distributions for
-each of these species. These data were collected from the stored litter. We did
-not add these data to the main figures as they are not part of any statistical
-test or main conclusions, but we are happy to do so if the editor suggests.
+detached. We now cite the Zhao et al (2019) results when discussing the mixture
+results (lines 466--470).
 
@@
 starting saturation times across species. This was aimed at allowing a
-completely blocked design. However, because some ignition attempts were
-unsuccessful, some days lacked the full complement of species while others,
-later in the process, sometimes had more than one replicate per species burned.
-We've added this information to the text [Lines X-X].
+completely blocked design. This was largely successful, but because some
+ignition attempts failed, some days lacked the full complement of species.
+We've added this information to the text (lines 218-221 and see lines 230-232
+for our explanation of how we used weather as a covariate).
 
@@
 
-We have added detail as requested by reviewer 2. See lines X-X. All litter was
-initially oven dried in large paper bags. The litter was then stored in the lab
-at low humidity for several weeks on open racks. The litter was then bagged and
-sealed and stored within cardboard boxes in a lab storage area. We saw no signs
-of mold nor ongoing decomposition. We minimized handling of the bags to avoid
-mechanical damage and changes to litter particle size.
+We have added detail as requested by reviewer 2. See lines 170-177. All litter
+was initially oven dried in large paper bags. The litter was then stored in the
+lab at low humidity for several weeks on open racks. The litter was then bagged
+and sealed and stored within cardboard boxes in a lab storage area. We saw no
+signs of mold nor ongoing decomposition. We minimized handling of the bags to
+avoid mechanical damage and changes to litter particle size.
 
@@
 
-This is a good point and another place where we did not discuss the literature fully. We
-have now added this to the discussion and made the conjecture that these
-differing patterns may result from different structure across the
+This is a good point and another place where we did not discuss the literature
+fully. We have now added this to the Discussion and made the conjecture that
+these differing patterns may result from different structure across the
 Cupressaceae with Australian \emph{Callitris} and the two Californian species
 packing more tightly and resulting in less flammable litter than the Asian
-Cupressaceae in Cornwell et al (2015).
+Cupressaceae in Cornwell et al (2015). See lines 409-415.
 
@@
 We thank the reviewer for their comments and apologize for our failure in
-proofreading. The reviewer should not have had to conduct copy editing and we
-sincerely apologize for that. We have taken heed of the reviewers' comments and
-fixed all specific grammar errors and missing symbols. For brevity, we've
-omitted responses to the very simple suggested grammar fixes and simply made
-all requested changes. We respond to other points individually.
+proofreading. We have taken heed of the reviewers' comments and fixed all
+specific grammar errors and missing symbols. For brevity, we've omitted
+responses to the straightforward grammar fixes and simply made all requested
+changes. We respond to other points individually below.
 
@@
 This portion was poorly written and this was also mentioned by Reviewer 2. We
-have rewritten this in response to reviewer 2's comment that most fires that
+have rewritten this in response to Reviewer 2's comment that most fires that
 spread result from dry lightning strikes across the west. See our responses to
-Reviewer 2, below.
+Reviewer 2, below and the paragraph now at lines 100-108.
 
@@
 
-Yes, agreed. We have tried to clean up the writing throughout and make the
-prose more active and precise.
+Done. We have tried to clean up the writing throughout and make the prose more
+active and precise.
 
@@
 
-Thank you. This was an oversight. We have also added more context from the
-literature to our Discussion. See lines XX-XX.
+Thank you. This was an oversight. Fixed on line 260. We have also added more
+context from the literature to our Discussion. See the paragraph that begins on
+line 462.
 
@@
 
-We have rewritten that section to be more precise.  See lines XX.
+We have rewritten that section to be more precise (lines 276-278).
 
@@
 
-We have edited this as suggested.
+We have edited as suggested (line 324)
 
@@
 
-We've tried to increase consistency throughout. We refer to taxonomic groups
-when appropriate and have standardized the subsection headings.
+We've tried to increase consistency throughout. Yes, we examined dry down
+curves by species because we did not have strong prior hypothesis about
+groups and because we had decided to find moisture dynamic groups empirically
+for deciding on mixtures. For the flammability analyses, we had decided to
+examine patterns by taxonomic group becuase we thought that would allow clearer
+presentation and aid generalizability and understanding. We now refer to
+taxonomic groups when appropriate and have standardized the subsection
+headings. We could do all analyses on a species basis, but that would be
+re-running new models on those data and we have tried to be very
+straightforward on what we did and honest about our analyses.
 
 \begin{quoting}
-  lines 348-349 - logically, this does not necessarily follow - factors such as
+  lines 348--349 - logically, this does not necessarily follow - factors such as
   large leaves or high SLA could independently affect both absorptive capacity
@@
 We have rewritten these sentences and we hope they are no longer confusing. See
-lines X-X.
+lines 384--388.
 
@@
 
-Fixed.
+Done.
 
@@
 
-Agreed. Fixed.
+Done.
 
@@
 
-We have added some better discussion of past results and included discussion of
-these papers. We also now contrast our result with that of Cornwell et al
-(2015) as suggested out by the Associate Editor. See lines XX-XX.
+We have added discussion of past results and included these papers. We also now
+contrast our result with that of Cornwell et al (2015) as suggested out by the
+Associate Editor. See lines 409--415.
 
@@
 
-Done. See line XX.
+Done.
 
@@
 have been easily visible without an explicit time axis. There is a trade-off
-here between illustrating simply the results of the statistical model and
+here between only illustrating the results of the statistical model and
 allowing the reader to get a better idea of variation in flammability
@@
 Yes. That omission was an oversight. We have added citations here. See lines
-XX-XX
+462--465
 
@@
 rewritten this section and added details justifying the importance of drying
-rates in fine fuels as suggested by the Senior Editor. See lines XX-XX.
+rates in fine fuels as suggested by the Senior Editor. See lines 100--108 and
+other changes to the Introduction.
 
@@
 
-This is a good point and we have added some discussion of this. We tried to
-limit our statements to those supported by the statistical tests (eg, the slope of
-nonaddivity with drying was greater than 1), Therefore, we focused on the
-positive effect of drying and the overall trend for positive nonadditivity.
+This is a good point and we have added some discussion of this pattern. We
+tried to limit our statements to those supported by the statistical tests.
 This, of course, obscures some details which are worth mentioning and we have
@@
 that is the reason we presented the actual predictions and departures from
-prediction across time. Please see our changes on lines X-X.
+prediction across time. Please see our revised Discussion and lines 462--474.
 
@@
 ``maximum water absorption'' to indicate water content following saturation. Of
-course, this is slightly inaccurate as the processes involved include both
-absorption and adsorption. Our earlier draft often used the term ``maximum
-water retention''. But we think this earlier wording was confusing as retention
-implied and inverse to drying (and we were inconsistent in usage as the
-reviewer points out).
+course, this is slightly inaccurate and includes both absorption into leaves
+and adsorption to surfaces. Our earlier draft used the term ``maximum water
+retention''. But we think this earlier wording was more confusing (and we were
+inconsistent in usage as the reviewer points out).
 
@@
 
-We have rewritten this section and removed that phrase.
+We have rewritten this section and removed that phrase. See lines 100--108.
 
@@
 
-We have rewritten this for clarity. See line XX.
+We have rewritten this for clarity. See line response above.
 
@@
 
-We were not sure what the reviewer's criticism was here but have rewritten this
-section to be more detailed and precise.
+We were not sure what the reviewer's criticism was here but have rewritten the
+Introduction to bmore detailed and precise.
 
@@
 Good point. We fixed that here and in another sentence elsewhere. Furthermore,
-we've edited to explain what we mean by ``absorption'' because we are referring
-to a litter bed and we include adsorption to litter surfaces in that
-definition. We had originally tried to avoid the chemical terms because we are
-observing a higher-level phenomenon but our previous language was confusing.
+we've edited to explain what we mean by ``absorption.'
 
@@
 
-Fixed.
+Edited. See paragraph that begins on line 109.
 
@@
 
-We have rewritten this. See line X.
+We have rewritten this. See lines 113--125.
 
@@
 
-We edited to ``composed of larger leaves or leaves with lower moisture
-retainability''
+We edited this section. See above.
 
@@
 
-Yes, excellent point. We have added detail and citations here (lines X-X).
+Yes, excellent point. We have added detail and citations here (lines 128--131).
 
@@
 
-Yes, this was incorrect. Fixed.
+Done. Lines 145--146.
 
@@
 
-We have rewritten this. See line X.
+We have rewritten this. See lines 148--151.
 
@@
 We have added some detail to this and also now provide particle size
-distributions for each species measured on the actual stored litter (SI fig 1).
+distributions for each species measured on the actual stored litter (SI Fig.~1).
 
@@
 We have added information to the methods explaining the collection in more
-detail. See lines XX and our response to the Associate Editor, above.
+detail. See lines 169--177 and our response to the Associate Editor, above.
 
@@
 large paper bags, then stored on racks in the low humidity lab. Our subsample
-measurements showed less than 5\% moisture. See lines X-X.
+measurements showed less than 5\% moisture. See lines 173--177.
 
@@
 
-We have added detail on the long term dry storage. See lines X-X.
+We have added detail on the long term dry storage. See lines 173--177.
 
@@
 
-Agreed, fixed.
+Agreed and done.
 
@@
 
-No, mass was fixed, width and length of the burning tray was fixed (15\,cm by
-1.5\,m), and depth was allowed to vary. We edited this section for clarity. See
-lines X-X
+No, in flammability trials mass was fixed, width and length of the burning tray
+was fixed (15\,cm by 1.5\,m), and depth was allowed to vary. We edited this
+section for clarity. This decision was absed on discussion with materials
+engineers who study fire that recommend a burning tray length of 10 times width
+to best measure steady flame spread. See lines 236--250.
 
@@
 
-TODO check
+Edited for clarity (Lines 248--249).
 
@@
 option, so we opted for the simpler to explain method of picking two traits
-already known to correlate with different axes of litter flammability. I envy
-the scientist who has never dealt with unforeseen events.
+already known to correlate with different axes of litter flammability.
+
+We could eliminate any mention of the other measurements if the reviewer and
+editor believes that is clearer. In this draft have erred always towards full
+disclosure.
 
@@
 
-TODO
+Done.
 
@@
 
-Although we are unsure of the criticism here, we have rewritten the sentence.
+We have rewritten that paragraph (lines 338--345)
 
@@
 
-Fixed.
+Corrected.
 
@@
 
-Yes, oops a typo. We have rewritten the heading. Line X
+Yes, oops a typo. We have revised the headings.
 
@@
 
-We have revised.
+Done.
 
@@
 litter and low maximum water absorption, had low flammability across moisture
-levels.''
+levels.'' Lines 409--411.
 
@@
 
-We don't understand the criticism. Perhaps the word ``alone'' is ambiguous? We
-changed ``alone'' to ``in monoculture'' and hope that is clearer.
+Perhaps the word ``alone'' is ambiguous? We edited these sentences (lines 459--461).
 
@@
 
-We don't understand the criticism. Is the complaint with present tense?
-Academic writing commonly uses present tense to describe results that still
-apply. There is no grammatical error.
+We rewrote this sentence.
 

### 2019-12-20 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
fixup
Close to final R2?

@@
 
-\title{Moisture effect on leaf litter alters non-additive flammability in mixed conifer forest}
+\title{Moisture absorption and drying alters nonadditive litter flammability
+  in mixed conifer forest}
 
-\running{Leaf litter flammability}
+\running{Litter flammability}
 
@@
 \nfig{8}
-\nref{51}
+\nref{55}
 
@@
     flammability across the major litter types typified by different tree taxa.
-    Litter mixtures exhibited nonadditivity in flame spread rate and
+
+    \item Litter mixtures exhibited nonadditivity in flame spread rate and
     in fuel consumption. This nonadditivity became more positive with fuel
@@
       \item \emph{Synthesis:} Taxon specific patterns of moisture absorption
-    and retention can change flammability rankings under moist compared with
-    dry fuels because traits that increase flame spread rate increase water
+    and drying can change flammability rankings under moist compared with dry
+    fuels because traits that increase flame spread rate increase water
     absorption. Litter mixtures generally burn with fire behavior more similar
@@
 being associated with rates of heat release and area burned, and with heat
-transfer predicting soil impacts of fire and the mortality. A shortcoming in
+transfer predicting soil impacts of fire and tree mortality. A shortcoming in
 our current knowledge is our integration of specific species and plant traits
@@
 known. Fuel moisture often limits fire activity in the western US
-\citep{Miller+Urban-1999, Parks+Parisien-2014}. Fire scientists often
-categorize fuel by size classes that correspond to moisture exchange rates,
-although such simplified categories have long been recognized as not sufficient
-to describe species-specific effects on moisture dynamics
-\citep{Anderson-1985}. Fuel moisture dynamics of leaf litter are dependent upon
-forest structure (shading and rain throughfall effects), litter physical
-structure, and chemical composition of the fuel-beds \citep{Nelson+Hiers-2008,
-  Matthews-2014, Kreye_Hiers_etal-2018}. A variety of models describing
-moisture dynamics of fine fuels are used by fire scientists \citep{Viney-1991,
-  Nelson-2000, Catchpole+Catchpole+etal-2001}. These models usually aim to
-describe equilibrium conditions and at least include temperature and relative
-humidity although some include precipitation and solar radiation effects as
-well. Explicit models of fire behavior as a function of time since wetting
-could improve fire prediction, however. Drying rates following snowmelt or
-following rainfall during the fire season will influence the length of the fire
-season, flammability over time, and the probability of sustained ignition,
-higher spread rates and fuel consumption when ignition occurs. That species can
-differ in litter drying rates has been demonstrated for at least one forest
-system \citep{Kreye+Varner+etal-2013}. Differential drying can be especially
-important when precipitation can occur during the same season as ignitions.
-Although synoptic weather conditions that create dry and wet lightning storms
-differ, the potential for precipitation and lightning during the same overall
-period of thunderstorm activity often exists: wet and dry convection days can
-occur under overlapping conditions, \citep{Rorig+Ferguson-1999}. Moisture
-levels of fine fuels respond rapidly to precipitation events during the fire
-season \citep{Estes+Knapp+etal-2012}. Differential moisture content across time
-may contribute to the often high spatial heterogeneity of surface fires
-\citep{Hille+Stephens-2005, Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and to
-variation in fire regime across landscapes \citep{Stephens-2001}.
+\citep{Miller+Urban-1999, Parks+Parisien-2014}. Fire scientists categorize fuel
+by size classes that correspond to moisture exchange rates, although such
+simplified categories have long been recognized as not sufficient to describe
+species-specific effects on moisture dynamics \citep{Anderson-1985} which are
+dependent upon forest structure (shading and rain through-fall effects), litter
+physical structure, and chemical composition of the fuel-beds
+\citep{Nelson+Hiers-2008, Matthews-2014, Kreye_Hiers_etal-2018}. A variety of
+models describing moisture dynamics of fine fuels are used by fire scientists
+\citep{Viney-1991, Nelson-2000, Catchpole+Catchpole+etal-2001}. These models
+usually aim to describe equilibrium conditions and at least include temperature
+and relative humidity, although some include precipitation and solar radiation
+effects as well. Explicit models of fire behavior as a function of time since
+wetting could improve fire prediction, however. Long term drying following snow
+melt will influence the length of the fire season. Short term drying following
+rainfall during the fire season will influence the probability of sustained
+ignition, spread rates, and fuel consumption when ignition occurs. That species
+can differ in litter drying rates has been demonstrated for at least one forest
+system \citep{Kreye+Varner+etal-2013}.
+
+Differential drying is especially important when precipitation occurs
+during the same season as ignitions. Although synoptic weather conditions that
+create dry and wet lightning storms differ, the potential for precipitation and
+lightning during the same overall period of thunderstorm activity often exists:
+wet and dry convection days can occur under overlapping conditions,
+\citep{Rorig+Ferguson-1999}. Moisture levels of fine fuels respond rapidly to
+precipitation events during the fire season \citep{Estes+Knapp+etal-2012}.
+Differential moisture content across time may contribute to the often high
+spatial heterogeneity of surface fires \citep{Hille+Stephens-2005,
+  Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and to variation in fire regime
+across landscapes \citep{Stephens-2001}.
 
@@
 water absorption''. Differences in moisture content over time since saturation
-(drying rate) can indicate different abilities to retain moisture. Litter
-particle geometry may have direct effects on maximum moisture absorption
-through surface area affects on adsoprtion and we also expect effects of leaf
-morphology, especially leaf cuticle thickness \citep{Van_Wagner-1969}. Drying
-rate should also be influenced by leaf and litter bed traits:
-surface-area-to-volume ratio and litter bulk density will influence air flow
-through the litter bed. Litter packing, therefore, effects fire behavior in two
-ways: directly by its effect on oxygen availability for combustion, and
-indirectly by its effect on drying rate. Both these pathways are affected by
-leaf particle traits, especially leaf size \citep{Scarff+Westoby-2006}.
-Fuel-beds composed of larger leaves or leaves with lower maximum moisture
-absorption should favor drying, thus increasing the probability of ignition and
-fire spread.
+can indicate different abilities to retain moisture. Leaf morphology,
+especially leaf cuticle thickness, can influence water absorption
+\citep{Van_Wagner-1969}. Litter particle geometry may have effects on maximum
+moisture content because a larger total surface area will increase water
+adsorption. Drying rate should also be influenced by leaf and litter bed traits
+because surface-area-to-volume ratio and litter bulk density will influence air
+flow through the litter bed. Therefore, we expect litter packing to effect fire
+behavior in two ways: directly by its effect on oxygen availability for
+combustion, and indirectly by its effect on drying rate. Both these pathways
+are affected by leaf particle traits, especially leaf size
+\citep{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves or leaves with
+thin cuticles should favor moisture absorption. On the other hand, large leaves
+lower litter density and should favor drying, thus increasing the probability
+of ignition and fire spread. The interaction of mositure absorption and drying
+deserves investigation.
 
@@
 moisture levels determined how time since wetting influenced two axes of
-flammability characterized by flame spread rate and total fuel consumed.
-Previous work described the flammability of these species in monocultures and
-in mixtures for oven-dry litter \citep{Magalhaes+Schwilk-2012} and found
-consistent positive nonadditivity of flammability where fire behavior of
-mixtures was driven by the most flammable species. We tested if the nonadditive
-effects on reported in previous studies occur under higher moisture levels, and
-whether fuel moisture effects strengthen or moderate any nonadditivity.
+flammability characterized by flame spread rate and total fuel consumed. To
+determine the potential magnitude of mixture effects, we tested if the
+nonadditive effects on reported in previous studies occur under higher moisture
+levels, and whether fuel moisture effects strengthen or moderate any
+nonadditivity.
 
@@
 at $21^\circ$C temperature and 30\% relative humidity, assessed via an iButton
-sensor (Maxim Integrated) placed over the samples. Litter baskets were arranged
-in a completely randomized design. During the dry-down process three subsamples
-of 5--10\,g were taken at 24 hour intervals and weighed. These subsamples were
-then oven dried for 24 hours at $100^\circ$C and re-weighed to assess moisture
-content. This was repeated until the final sample from a basket was less than
-10\% moisture. This experimental setup was designed to accelerate drying
-compared with field conditions because the baskets were suspended in air 2.8 --
-5.3$^\circ$C above the warmest month average temperatures for the closest
-weather stations to the collection sites (Grant Grove and Lodegpole stations,
-NOAA 1981-2010 normals, https://www.ncdc.noaa.gov/cdo-web/datatools/normals).
+sensor (Maxim Integrated) placed 1\,m above the samples. Litter baskets were
+arranged in a completely randomized design. During the dry-down process three
+subsamples of 5--10\,g were taken at 24 hour intervals and weighed. These
+subsamples were then oven dried for 24 hours at $100^\circ$C and re-weighed to
+assess moisture content. This was repeated until the final sample from a basket
+was less than 10\% moisture. This experimental setup was designed to accelerate
+drying compared with field conditions because the baskets were suspended in air
+2.8 -- 5.3$^\circ$C above the warmest month average temperatures for the
+closest weather stations to the collection sites (Grant Grove and Lodegpole
+stations, NOAA 1981-2010 normals,
+https://www.ncdc.noaa.gov/cdo-web/datatools/normals).
 
@@
 adsorption (specific leaf area), and one litter bed trait known to directly
-affect flammability and likely to influence drying rate and litter packing
-density (litter bulk density). Specific leaf area was measured on freshly
-fallen litter by sampling 5 individuals per species. This measurement included
-petiole and for the scale-leaved species included the full excised branchlet.
-Litter bulk density was obtained in this experiment for each basket by dividing
-the dry mass (450\,g) by the estimated litter volume (basket width x length x
-litter depth). To examine litter particle size distributions in the stored
-litter, we subsampled 10 approximately 2\,g samples per species from the stored
-litter. Each individual particle of that subsample was measured for length
-(longest axis), width (longest axis perpendicular to length) and thickness
+affect flammability and likely to influence drying rate (litter bulk density).
+Specific leaf area was measured on freshly fallen litter by sampling 5
+individuals per species. This measurement included the petiole and for the
+scale-leaved species included the full excised branchlet. Litter bulk density
+was obtained in this experiment for each basket by dividing the dry mass
+(450\,g) by the estimated litter volume (basket width x length x litter depth).
+To examine litter particle size distributions in the stored litter, we
+subsampled 10 approximately 2\,g samples per species from the stored litter.
+Each individual particle of that subsample was measured for length (longest
+axis), width (longest axis perpendicular to length) and thickness
 (perpendicular to length and width at midrib) using an electronic caliper.
@@
 our timing so as to attempt to burn samples over a range of moisture contents.
-We aimed for moisture contents from 10\% to 100\% on a dry mass basis as
+We staggered saturation and start of drying so we could conduct our burning in
+a blocked design with multiple species at the same moisture content per burning
+day. We aimed for moisture contents from 10\% to 100\% on a dry mass basis because
 preliminary trials on damp fuels suggested that across these fuels ignition
@@
 We assigned these replicates a time since wetting value of 144 hours equivalent
-to the longest time since drying for the wetted samples. Flammability was
-assessed using a 150\,cm long burn table, 15\,cm wide and 15\,cm tall, in which
-leaf litter was placed and the table gently shaken to allow settling. Burning
-trials all used 450\,g (dry weight) of litter. For three-species mixtures, this
-was 150\,g per species. Therefore, fuel mass and burning tray length and width
-were fixed, but fuel depth was allowed to vary. We recorded the actual fuel
-depth at three points along the tray to calculate litter bed density before
-burning. This design allows fire to reach constant flame spread rates
-\citep{Magalhaes+Schwilk-2012}. Two metal rulers equally spaced along the
-apparatus allowed visual determination of maximum flame height. Litter was
-ignited with a propane torch; maximum adiabatic flame temperature of
-1899$^\circ$C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end
-of the apparatus and allowed to burn until extinction. If a sample failed to
-ignite after 30 seconds, no further attempt was made and if three samples of a
-species at a particular moisture level failed to ignite, no further attempts
-were made on that species/moisture combination and the replicate was allowed to
-continue drying until the next scheduled checkpoint. The flammability
-trials followed the methodology detailed in \citet{Magalhaes+Schwilk-2012}.
+to the longest time since drying for the wetted samples.
+
+Burn trials occurred in a cement structure used to simulate house fires at the
+Fire Department of the City of Lubbock. This structure minimized wind and
+helped regulate temperature and relative humidity. Temperature and relative
+humidity were measured before every trial using a Kestrel 3000
+(Nielsen-Kellerman, Boothwyn, PA) and these values were used to calculate vapor
+pressure deficit as a single weather covariate at each trial. All trials were
+conducted on clear days between 10 am and 3 pm. We scheduled burns in a
+randomized blocked design aiming to burn one replicate of each species or
+mixture on single burn day. This required scheduling species saturation and
+during start based on the dry down curves.
+
+Flammability was assessed using a 150\,cm long burn table, 15\,cm wide and
+15\,cm tall, in which leaf litter was placed and the table gently shaken to
+allow settling. Burning trials all used 450\,g (dry weight) of litter. For
+three-species mixtures, this was 150\,g per species. Therefore, fuel mass and
+burning tray length and width were fixed, but fuel depth was allowed to vary.
+We recorded the actual fuel depth at three points along the tray to calculate
+litter bed density before burning. This design allows fire to reach constant
+flame spread rates \citep{Magalhaes+Schwilk-2012}. Two metal rulers equally
+spaced along the apparatus allowed visual determination of maximum flame
+height. Litter was ignited with a propane torch; maximum adiabatic flame
+temperature of 1899$^\circ$C,
+\url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the apparatus
+and allowed to burn until extinction. If a sample failed to ignite after 30
+seconds, no further attempt was made and if three samples of a species at a
+particular moisture level failed to ignite, no further attempts were made on
+that species/moisture combination and the replicate was allowed to continue
+drying until the next scheduled checkpoint. The flammability trials followed
+the methodology detailed in \citet{Magalhaes+Schwilk-2012}.
 
@@
 
-Burn trials occurred in a cement structure used to simulate house fires at the
-Fire Department of the City of Lubbock. This structure minimized wind and
-helped regulate temperature and relative humidity. Temperature and relative
-humidity were measured before every trial using a Kestrel 3000
-(Nielsen-Kellerman, Boothwyn, PA). All trials were conducted on clear days
-between 10 am and 3 pm. We scheduled burns in a randomized blocked design
-aiming to burn one replicate of each species or mixture on single burn day.
-This required scheduling species saturation and during start based on the dry
-down curves.
-
 \subsection*{Statistical analyses}
@@
 within replicate basket. We extracted the coefficients of the fitted model to
-describe maximum water absorption ($m_{max}$; g of water per g of dry mass)
-and drying rate ($\lambda$; hour$^{-1}$) for each species. All analyses
-were conducted in R 3.6.1 \citep{RCoreTeam-2019}. We fit models with R packages
+describe maximum water absorption ($m_{max}$; g of water per g of dry mass) and
+drying rate ($\lambda$; hour$^{-1}$) for each species. All analyses were
+conducted in R 3.6.1 \citep{RCoreTeam-2019}. We fit models with R packages
 `lme4' \citep{Bates_Machler_etal-2015} and `afex'
@@
 I error rates. We tested if one litter particle trait (specific leaf area) and
-one litter bed trait (litter bulk density) influenced moisture dynamics by
-fitting linear models with species means as observations. For each response
-variable we fit a single linear model with two predictors and an interaction
-term.
+one litter bed trait (litter bulk density) influenced maximum water absorption
+and drying rate by fitting linear models with species means as observations.
+For each response variable we fit a single linear model with two predictors and
+an interaction term.
 
@@
 average of the constituent species. To choose mixture combination, we first
-grouped the species according to their maximum water retention and drying rate
-according to estimated marginal means by species using the R package `emmeans'
+grouped the species according to their maximum water absorption and drying rate
+according to estimated marginal means using the R package `emmeans'
 \citep{Lenth-2019}. This resulted in four groups of species and we then
@@
 difference between observed and predicted values of flammability will be zero.
-We tested if there were significant departures from the null with a mixed
-effects linear model. For flammability parameters, we fit linear mixed effect
-models with the observed-predicted flammability value as the dependent variable
-and used Wald tests to test if the centered intercept was different than zero
-(a measure of overall nonadditivity) and if the slope with time since wetting
-was different from zero (indicating nonadditivity changing with drying).
+We fit linear mixed effect models with the observed-predicted flammability
+value as the dependent variable and used Wald tests to test if the centered
+intercept was different than zero (a measure of overall nonadditivity) and if
+the slope with time since wetting was different from zero (indicating
+nonadditivity changing with drying).
 
@@
 
-\subsection*{Species specific moisture dynamics and leaf traits}
+\subsection*{Moisture dynamics across species}
 
-Past work has demonstrated that green leaf width predicts litter packing and,
+Past work has demonstrated that green leaf length predicts litter packing and,
 consequently, flame spread rate across these species. Litter particle size
 distributions, however show that there is substantial variation in particle
-sizes within a species (SI Fig.~1). There were marked differences in the extent
-of moisture species retain initially following wetting and draining by gravity
-(all $P < 0.0001$, Fig.~\ref{fig:drydown} and Table~\ref{tab:drydown}).
-\emph{Q. kelloggii}, the only broadleaf, is the species that retained the most
-moisture initially. \emph{Quercus}, like the pines, had a rapid drying rate
-(Fig.~\ref{fig:drydown} and SI Fig.~2). Calculating estimated marginal means
-following the linear mixed-effects model discriminated the species into three
-different maximum water retention groups: 1) \emph{Q. kelloggii}; 2) \emph{P.
-  lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica}, \emph{S.
-  giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P. ponderosa}
-(SI Fig.~2). There are also differences in drying rates across the eight
-species and estimated mixed linear model slopes discriminated species into
-three groups from lowest to most rapid: 1) \emph{C. decurrens}, 2) \emph{Abies
-  spp} and \emph{S. giganteum}, and 3) \emph{Q. kelloggii} and \emph{Pinus spp}
-(SI Fig.~3).
+sizes within a species (SI Fig.~1). There were differences in the maximum water
+absorption (all $P < 0.0001$, Fig.~\ref{fig:drydown} and
+Table~\ref{tab:drydown}). \emph{Q. kelloggii}, the only broadleaf, is the
+species that absorbed the most moisture initially. \emph{Quercus}, like the
+pines, had a rapid drying rate (Fig.~\ref{fig:drydown} and SI Fig.~2).
+Calculating estimated marginal means following the linear mixed-effects model
+discriminated the species into three different maximum water absorption groups:
+1) \emph{Q. kelloggii}; 2) \emph{P. lambertiana} and \emph{C. decurrens}; and
+3) \emph{A. magnifica}, \emph{S. giganteum}, \emph{A. concolor}, \emph{P.
+  jeffreyi} and \emph{P. ponderosa} (SI Fig.~2). There are also differences in
+drying rates across the eight species and estimated mixed linear model slopes
+discriminated species into three groups from lowest to most rapid: 1) \emph{C.
+  decurrens}, 2) \emph{Abies spp} and \emph{S. giganteum}, and 3) \emph{Q.
+  kelloggii} and \emph{Pinus spp} (SI Fig.~3).
 
@@
   gravity for 3 min before initial weighing. Colors indicate genera and there
-  were significant differences in both maximum water retention and in rate of
+  were significant differences in both maximum water absorption and in rate of
   drying (based linear model on log of water content). A small amount of
@@
 
-Across these eight species specific leaf area had a positive association with maximum
-water retention (SI Fig~5, linear model $P = 0.003$) and litter
-bulk density had a negative effect on drying rate (SI Fig~6,
-linear model $P = 0.029$). ANOVA results are shown in Table~\ref{tab:mc_di_anova}.
+Across these eight species specific leaf area had a positive association with
+maximum water absorption (Table~\ref{tab:mc_di_anova}, SI Fig~5, linear model
+$P = 0.003$) and litter bulk density had a negative effect on drying rate
+(Table~\ref{tab:mc_di_anova}, SI Fig~6, linear model $P = 0.029$).
 
 \begin{table}[h]
-  \caption{Linear model results for maximum water retention (\% moisture
+  \caption{Linear model results for maximum water absorption (\% moisture
     content on dry mass basis) and for estimated drying rate (hr$^{-1}$) as
@@
 
-\subsection*{Moisture effects on flammability}
+\subsection{Flammability across taxonomic groups}
 
@@
 overall ranking of groups was similar for fuel consumption and for flame spread
-rate, there were different pattern across species in these two flammability
+rate, there were different pattern across taxa in these two flammability
 parameters. For spread rate, the decline in flammability with moisture was
@@
 model convergence otherwise. This result should therefore be seen as
-preliminary because the test does not reflect an \emph{a priori} hypothesis.
-But probably reflects the fact that \emph{P. lambertiana} has shorter and finer
+preliminary because the test does not reflect an \emph{a priori} hypothesis,
+but probably reflects the fact that \emph{P. lambertiana} has shorter and finer
 needles than the other two pines.
@@
 Based on visual inspection of the dry down curves and the estimated marginal
-means for maximum water retainability and desiccation rate (SI Figs 3--4), we
-selected four species to represent the full range of litter moisture dynamics
-for use in the litter mixtures and examined all four three-species combinations
-of those four. The four species were \emph{Abies concolor}, \emph{Calocedrus
-  decurrens}, \emph{Pinus jeffreyi} and \emph{Quercus kelloggii}.
+means for maximum water absorption and drying rate (SI Figs 3--4), we selected
+four species to represent the full range of litter moisture dynamics for use in
+the litter mixtures and examined all four three-species combinations of those
+four. The four species were \emph{Abies concolor}, \emph{Calocedrus decurrens},
+\emph{Pinus jeffreyi} and \emph{Quercus kelloggii}.
 
@@
 of different species will exhibit different moisture contents over time as a
-consequence of both differential initial absorption and of differences in
-drying rate. This has consequences for flammability as a function of species
-and time since wetting. Our flammability trials demonstrate that both flame
-spread rate and fuel consumption increased with time since wetting but that the
-rate of increase varied across these taxa. Only the pines are able to ignite
-and support flame spread immediately after wetting (Fig.
-\ref{fig:spread_time}). In contrast, \emph{Quercus} exhibited a threshold
-response to drying with no ignition prior to 72 hours of drying but once
-ignition is possible, it had faster spread rates than the other taxa. Our two 
-Cupressaceae, which have dense litter and low maximum water
-absorption, had low flammability across moisture levels. This result is
-consistent with the low flammability found in an Australian Cupressaceae that
-have small litter particles and dense packing \citep{Scarff+Westoby-2006,
-  Bowman+Haverkamp+etal-2018}. However, the densely packed litter is not a
-consistent trait across the family because \citet{Cornwell+Elvira+etal-2015}
-found relatively high litter flammability and low litter density for some Asian
-Cupressaceae.
+consequence of both differential initial absorption and differences in drying
+rate. This leads to flammability varying both by taxon and time since wetting.
+In these mixed conifer forests, fine fuels can vary in fuel moisture throughout
+the fire season. For example, 10-hour fuels in northern California varied from
+less than 10\% to over 80\% depending on precipitation events in a study
+following fuel moisture throughout the fire season
+\citep{Estes+Knapp+etal-2012}. Early summer prescribed burns at our study sites
+in Sequoia National Park show evidence of moisture limiting fire behavior at
+12-14\% fuel moisture (Knapp et al 2005).
 
-With eight species, power to detect species-specific trait effects on these
-moisture patterns was weak, but we examined one litter trait known to directly
-influence flammability: litter bulk density and one leaf trait, specific leaf
-area. Litter density is largely driven by leaf size in this system and, in
-turn, controls the major axis of flammability variation in dry litter
-\citep{Magalhaes+Schwilk-2012}. Our results suggest that moisture dynamics are
-controlled by both traits intrinsic to the leaf particles (specific leaf area)
-and an emergent property of the litter bed (bulk density). Specific leaf area
-here is probably capturing some joint effect of cuticle thickness and total
-surface area of litter particles which infleunces water adsorption, but we
-cannot test the morphological and physical mechanisms with these data. The oak
-has much higher maximum water absorption than do the other species, followed by
-the finest-needled pine, \emph{P. lambertiana}. Our study examines fresh
-litter, but we expect that decomposition will affect moisture dynamics. For
-example, Van Wagner \citeyear{Van_Wagner-1969}, investigating fuel weathering,
-found that pine needles had a variation in moisture content 3 to 6 times
-greater when they lacked their protective wax coating when compared to intact
-needles.
+Our flammability trials demonstrate that both flame spread rate and fuel
+consumption increased with time since wetting but that the rate of increase
+varied across these taxa. Only the pines are able to ignite and support flame
+spread immediately after wetting (Fig. \ref{fig:spread_time}). In contrast,
+\emph{Quercus} exhibited a threshold response to drying with no ignition prior
+to 72 hours of drying but once ignition is possible, it had faster spread rates
+than the other taxa. Our two Cupressaceae, which have dense litter but have low
+maximum water absorption, had low flammability across moisture levels. This
+result is consistent with the low flammability found in an Australian
+Cupressaceae that have small litter particles and dense packing
+\citep{Scarff+Westoby-2006, Bowman+Haverkamp+etal-2018}. However, the densely
+packed litter is not a consistent trait across the family because
+\citet{Cornwell+Elvira+etal-2015} found relatively high litter flammability and
+low litter density for some Asian Cupressaceae.
 
-\subsection*{Effects of moisture and flammability in litter mixtures}
+We examined one litter bed trait known to directly influence flammability,
+litter bulk density, and one leaf trait, specific leaf area. Litter density is
+largely driven by leaf size in this system and, in turn, controls the major
+axis of flammability variation in dry litter \citep{Magalhaes+Schwilk-2012}.
+Our results suggest that moisture dynamics are controlled by both traits
+intrinsic to the leaf particles (specific leaf area) and an emergent property
+of the litter bed (bulk density). Specific leaf area here is probably capturing
+some joint effect of cuticle thickness that may influence water absorption and
+total surface area of litter particles which increases water adsorption. The
+oak has much higher maximum water absorption than do the other species,
+followed by the finest-needled pine, \emph{P. lambertiana}. Our study examines
+fresh litter, but we expect that decomposition will affect moisture dynamics.
+For example, Van Wagner \citeyear{Van_Wagner-1969}, investigating fuel
+weathering, found that pine needles had a variation in moisture content 3 to 6
+times greater when they lacked their protective wax coating when compared to
+intact needles.
+
+\subsection*{Nonadditivity in litter mixtures}
 
@@
 any Cupressaceae had the most rapid drying rate (SI Figure 7 and SI Table 8).
-Drydown rates, however, did not behave as the average of the constituent
+Drying rates, however, did not behave as the average of the constituent
 species: mixtures tended to be drier than expected early in the process and
-wetter than expected later (Fig. \ref{fig:mixture_obs_pred_mc}). The fact that
-mixtures tended to have lower water content than predicted from the mean of the
-constituent species after wetting, but higher water content than expected after
-long drying is surprising because past work has shown that mixtures tend to
-pack less densely than do litter monocultures \citep[][Table
-2]{Magalhaes+Schwilk-2012}. If packing density controls rate of drying, then we
-would expect mixtures to dry more rapidly than predicted from monoculture
-trials. This discrepancy may be a consequence of the lightly packing litters
-having high maximum water absorption and of drying being non-linear with time
+wetter than expected later (Fig. \ref{fig:mixture_obs_pred_mc}). Past work has
+shown that mixtures tend to pack less densely than do litter monocultures
+\citep{Magalhaes+Schwilk-2012}. If packing density controls rate of drying,
+then we would expect mixtures to dry more rapidly than predicted from
+monoculture trials. Therefore, it is surprising that mixtures tended to have
+lower maximum water content than predicted from the mean of the constituent
+species, but higher water content than expected after long drying. This
+discrepancy may be a consequence of the lightly packing litters having high
+maximum water absorption and of drying being non-linear with time
 (Fig.~\ref{fig:drydown}).
@@
 were consistent across mixture identities. This is a slightly different result
-than that reported by Blauw et al \citeyear{Blauw+Wensink+etal-2015}. Blauw et
+than that reported by \citet{Blauw+Wensink+etal-2015}. Blauw et
 al found that nonadditivity was greater in moist mixtures of heathland fuels
@@
 although we do have some results consistent with theirs: mixtures that included
-\emph{Q. kellogii} could ignite at times since wetting at which the species in
-monoculture could not.
+\emph{Q. kelloggii} could ignite at times since wetting at which \emph{Quercus}
+in monoculture could not.
 
@@
 nonadditivity. For example, at 72 hours following saturation, fuel consumption
-across mixtures could show neutral, positive, or negative nonadditivity (Fig.
+across mixtures showed neutral, positive, or negative nonadditivity (Fig.
 \ref{fig:mixture_obs_pred_consume}). Previous studies have demonstrated the
-potential for the direction of nonadditivty to shift with fuel moisture
+potential for the direction of nonadditivity to shift with fuel moisture
 \citep{Blauw+Wensink+etal-2015} or with altering mixtures of leaf vs twig
 litter \citep{Zhao+vanLogtestijn+etal-2019}. Our litter represented naturally
-droppped litter and the ratio of leaves to twigs in such material can vary both
+dropped litter and the ratio of leaves to twigs in such material can vary both
 by species but also by wind and storm events the previous season. With only
@@
 We draw two main conclusions from this work. First, maximum water absorption is
-high in those species with the most flammability litter under dry conditions.
-As a consequence, under higher fuel moisture, the relative flammability of
+highest in those species with the most flammability litter under dry conditions.
+As a consequence, under higher fuel moisture, the flammability ranking of
 these taxa can change compared to the flammability of dry fuels. Second, litter
@@
 
-conceptualization, RMQM and DWS; methodology, RMQM and DWS; experiments, RMQM;
+Conceptualization, RMQM and DWS; methodology, RMQM and DWS; experiments, RMQM;
 coding and data analysis, DWS and RMQM; writing DWS and RMQM; writing--review

### 2019-12-20 scripts/moisture-ms-figs-tables.R

@@ -100,7 +100,7 @@ print(tab1, file=file.path(RESULTS, "tab1_drydown_anova.ltx"),  booktabs=TRUE, f
 
 
 ###############################################################################
-## SI Table 1: drydown by species model coefficients
+## SI Table 1: drydown by genus model coefficients
 ###############################################################################
 SI_tab1 <- summary(dry.mod)$coefficient
 print(xtable(SI_tab1), file=file.path(RESULTS, "SI_tab1_drydown_coef.ltx"),  booktabs=TRUE, floating=FALSE)

### 2019-12-20 ms/response_to_reviews.tex
Oops, typo in title fixed.

@@
 
-Enclosed is our revised manuscript research article, ``Moisture absorption and drying alters nonadditive litter flammability in mixed conifer forest''.
+Enclosed is our revised manuscript research article, ``Moisture absorption and drying alter nonadditive litter flammability in mixed conifer forest''.
 

### 2019-12-20 ms/response_to_reviews.tex
Edits to response letter

@@
 throughout the season (Kaufman and Martin 1989, Knapp and Keeley 2006, Hille
-and Stephens 2005, Knapp et al 2006). These differences in fire behavior in
-turn cause spatial variation in tree mortality (Stephens et al 2002, Schwilk et
-al 2006). Despite this, we know of only one other study specifically
-investigating differential drying behavior (Kreye et al 2013b) and none that
-included drying rates with flammability measurement.
+and Stephens 2005, Knapp et al 2006). These differences in fire behavior cause
+spatial variation in tree mortality (Stephens et al 2002, Schwilk et al 2006).
+Despite this, we know of only one other study specifically investigating
+differential drying behavior (Kreye et al 2013b) and none that included drying
+rates with flammability measurement.
 
@@
 
-When deciding how to limit the number of tests at the comparative scale
+When considering how to limit the number of tests at the comparative scale
 (species as data points), we decided to not include size, because its influence
@@
 we have added a figure (SI Fig. 1) to the SI materials showing litter particle
-size (the containing prism volumes) distributions for each of these species.
-These data were collected from the stored litter. We did not add these data to
-the main figures as they are not part of any statistical test or main
-conclusions, but we are happy to do so if the editor suggests.
+size distributions for each of these species. These data were collected from
+the stored litter. We did not add these data to the main figures as they are
+not part of any statistical test or main conclusions, but we are happy to do so
+if the editor suggests.
 
@@
 We've tried to increase consistency throughout. Yes, we examined dry down
-curves by species because we did not have strong prior hypothesis about
-groups and because we had decided to find moisture dynamic groups empirically
-for deciding on mixtures. For the flammability analyses, we had decided to
-examine patterns by taxonomic group becuase we thought that would allow clearer
-presentation and aid generalizability and understanding. We now refer to
-taxonomic groups when appropriate and have standardized the subsection
-headings. We could do all analyses on a species basis, but that would be
-re-running new models on those data and we have tried to be very
-straightforward on what we did and honest about our analyses.
+curves by species because we did not have prior hypothesis about groups and
+because we had decided to find moisture dynamic groups empirically for deciding
+on mixtures. For the flammability analyses, we had decided to examine patterns
+by taxonomic group because we thought that would allow clearer presentation and
+aid generalizability and understanding. We now refer to taxonomic groups when
+appropriate and have standardized the subsection headings. We could do all
+analyses on a species basis, but that would be re-running new models on those
+data.
 
@@
 the total number lightning strikes does not well predict burned area and that
-ignitions occur often from dry strikes argues for the importance of fuel
-moisture. Although synoptic weather conditions that create dry and wet storms
-differ, the potential for precipitation and lightning during the same overall
-period of thunderstorm activity often exists (eg wet and dry convection days
-can occur under overlapping conditions, Rorig and Ferguson 1999). We've
+observed ignitions occur most often from dry strikes argues for the importance
+of fuel moisture. Although synoptic weather conditions that create dry and wet
+storms differ, the potential for precipitation and lightning during the same
+overall period of thunderstorm activity often exists (eg wet and dry convection
+days can occur under overlapping conditions, Rorig and Ferguson 1999). We've
 rewritten this section and added details justifying the importance of drying
@@
 ``maximum water absorption'' to indicate water content following saturation. Of
-course, this is slightly inaccurate and includes both absorption into leaves
+course, this is slightly inaccurate and includes both absorption into particles
 and adsorption to surfaces. Our earlier draft used the term ``maximum water
@@
 
-We have rearranged the logic of the introduction as suggested.
+We have rearranged the logic of the Introduction as suggested.
 
@@
 We were not sure what the reviewer's criticism was here but have rewritten the
-Introduction to bmore detailed and precise.
+Introduction.
 
@@
 
-No, in flammability trials mass was fixed, width and length of the burning tray
-was fixed (15\,cm by 1.5\,m), and depth was allowed to vary. We edited this
-section for clarity. This decision was absed on discussion with materials
-engineers who study fire that recommend a burning tray length of 10 times width
-to best measure steady flame spread. See lines 236--250.
+In flammability trials mass was fixed, width and length of the burning tray was
+fixed (15\,cm by 1.5\,m), and depth was allowed to vary. We edited this section
+for clarity. This decision was based on discussion with materials engineers who
+study fire that recommend a burning tray length of 10 times width to best
+measure steady flame spread. See lines 236--250.
 
@@
 several days of thermocouple data made that an unattractive and unnecessary
-option, so we opted for the simpler to explain method of picking two traits
-already known to correlate with different axes of litter flammability.
+option, so we opted for the simpler method of picking two traits already known
+to correlate with different axes of litter flammability.
 
 We could eliminate any mention of the other measurements if the reviewer and
-editor believes that is clearer. In this draft have erred always towards full
-disclosure.
+editors wish. In this draft have erred towards full disclosure.
 

### 2019-12-20 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Oops, typo in title fixed.
Fix typos

@@
 
-\title{Moisture absorption and drying alters nonadditive litter flammability
+\title{Moisture absorption and drying alter nonadditive litter flammability
   in mixed conifer forest}
@@
     retention. Characterizing how litter fuels retain moisture through time and
-    how fuels from multiple species interact are key steps to adding species or
-    taxon specific information to fire behavior prediction.
+    how fuels from multiple species interact are key steps to adding taxon
+    specific information to fire behavior prediction.
 
@@
     flammability variation and found that these differing moisture dynamics
-    result in time since wetting having strongly differing effects on
+    resulted in time since wetting having strongly differing effects on
     flammability across the major litter types typified by different tree taxa.

### 2019-12-20 ms/response_to_reviews.tex
fixup

@@
 
-Enclosed is our revised manuscript research article, ``Moisture absorption and drying alter nonadditive litter flammability in mixed conifer forest''.
+Enclosed is our revised manuscript research article, ``Moisture absorption and drying alter nonadditive litter flammability in a mixed conifer forest''.
 
@@
 We could eliminate any mention of the other measurements if the reviewer and
-editors wish. In this draft have erred towards full disclosure.
+editors wish. In this draft have erred on the side of full disclosure.
 

### 2020-03-27 ms/response_to_reviews.tex
Start R2 for J of Ecol

@@
 adequately with the literature. We have made substantial changes to the
-Introduction and the Discussion adding more context and citing more of the
+Introduction and the Discussion. We added context and cited more of the
 recent literature. We reorganized the Introduction and now do not rush the
@@
 When considering how to limit the number of tests at the comparative scale
-(species as data points), we decided to not include size, because its influence
-on flammability is primarily through litter packing and was captured by
-litterbed density. However, in response to the request to show particle sizes,
-we have added a figure (SI Fig. 1) to the SI materials showing litter particle
-size distributions for each of these species. These data were collected from
-the stored litter. We did not add these data to the main figures as they are
-not part of any statistical test or main conclusions, but we are happy to do so
-if the editor suggests.
+(species as data points), we decided to not include size, because it affects
+flammability primarily through litter packing and was captured by litterbed
+density. However, in response to the request to show particle sizes, we have
+added a figure (SI Fig. 1) to the SI materials showing litter particle size
+distributions for each of these species. These data were collected from the
+stored litter. We did not add these data to the main figures as they are not
+part of any statistical test or main conclusions, but we are happy to do so if
+the editor suggests.
 

### 2020-03-27 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
fixup
Start R2 for J of Ecol

@@
 \title{Moisture absorption and drying alter nonadditive litter flammability
-  in mixed conifer forest}
+  in a mixed conifer forest}
 
@@
   drying (based linear model on log of water content). A small amount of
-  horizontal jitter (0.5 hr) was added to this plot to aid in visualizing
-  because, although collections were staggered, there were sometimes multiple
-  samples collected at a single time point.}
+  horizontal jitter (0.5 hr) was added to this plot because, although
+  collections were staggered, there were sometimes multiple samples collected
+  at a single time point.}
  \label{fig:drydown}
@@
 
-\subsection{Flammability across taxonomic groups}
+\subsection*{Flammability across taxonomic groups}
 
@@
   \ref{tab:spreadrate_anova} and coefficients in SI Table 6. Horizontal jitter
-  (up to 2 hr) was added to the plot for visualization and to avoid
-  overplotting. All collections were at the times labeled on the x-axis. A lack
-  of data at earlier time points for a taxon indicates no ignitions occurred at
-  those time points.}
+  (up to 2 hr) was added to the plot to avoid overplotting. All collections
+  were at the times labeled on the x-axis. A lack of data at earlier time
+  points for a taxon indicates no ignitions occurred at those time points.}
 \label{fig:spread_time}
@@
   table in Table \ref{tab:consume_anova} and coefficients in SI Table 7.
-  Horizontal jitter (up to 2 hr) was added to the plot for visualization and to
-  avoid overplotting. All collections were at the times labeled on the x-axis..
-  A lack of data at earlier time points for a taxon indicates no ignitions
-  occurred at those time points.}
+  Horizontal jitter (up to 2 hr) was added to the plot to avoid overplotting.
+  All collections were at the times labeled on the x-axis. A lack of data at
+  earlier time points for a taxon indicates no ignitions occurred at those time
+  points.}
   \label{fig:consume_time}

### 2020-04-02 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Combine figures as requested by reviews on R1

@@
-%% Revision 2019-12-10
 \documentclass[letterpaper,12pt]{article}
@@
 \ntables{4}
-\nfig{8}
+\nfig{5}
 \nref{55}
@@
 lower litter density and should favor drying, thus increasing the probability
-of ignition and fire spread. The interaction of mositure absorption and drying
+of ignition and fire spread. The interaction of moisture absorption and drying
 deserves investigation.
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/fig1_drydown-curves.pdf}
-\caption{Dry down curves for eight litter types. Lines show best-fit
-  exponential decay curves fit to moisture content on a dry mass basis. Litter
-  samples we saturated by immersion for 24 hours and then allowed to drain by
-  gravity for 3 min before initial weighing. Colors indicate genera and there
-  were significant differences in both maximum water absorption and in rate of
-  drying (based linear model on log of water content). A small amount of
-  horizontal jitter (0.5 hr) was added to this plot because, although
-  collections were staggered, there were sometimes multiple samples collected
-  at a single time point.}
+\includegraphics[width=5cm]{figs_tables/fig1_dry_effects_by_taxon.pdf}
+\caption{Effects of litter drying on fuel moisture and flammability. A) Dry
+  down curves for eight species. Lines show best-fit exponential decay curves
+  fit to moisture content on a dry mass basis. B) Flame spread rate by time
+  since wetting. Lines indicate best fit linear model fits by taxonomic group.
+  Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI
+  Table 6. C) Percent fuel consumed by time since wetting. Curves shown
+  indicate best fit beta regression lines by taxonomic group. Model ANOVA table
+  in Table \ref{tab:consume_anova} and coefficients in SI Table 7. For all
+  plots, horizontal jitter (0.5 to 2 hr) was added to avoid overplotting. A
+  lack of data at earlier time points for a taxon indicates no ignitions
+  occurred at those time points. }
  \label{fig:drydown}
@@
 
-Flame spread rate declined with fuel moisture (Fig.~\ref{fig:spread_moist}).
+Flame spread rate declined with fuel moisture (Fig.~\ref{fig:flam_moist}).
 \emph{Quercus} had higher spread rates across all fuel moistures and had a
@@
 rate of change with moisture content varied across the taxonomic groups
-(Fig.~\ref{fig:consume_moist}, Table~\ref{tab:consume_anova}). Although the
+(Fig.~\ref{fig:flam_moist}, Table~\ref{tab:consume_anova}). Although the
 overall ranking of groups was similar for fuel consumption and for flame spread
@@
 parameters. For spread rate, the decline in flammability with moisture was
-shallowest for \emph{Pinus} and Cupressaceae (Fig.~\ref{fig:spread_moist}). For
+shallowest for \emph{Pinus} and Cupressaceae (Fig.~\ref{fig:flam_moist}). For
 fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in
 flammability with fuel moisture than did the other groups
-(Fig.~\ref{fig:consume_moist}).
+(Fig.~\ref{fig:flam_moist}).
 
@@
   \centering
-\includegraphics[width=8cm]{figs_tables/fig2_spread_actualMC.pdf}
-\caption{Flame spread rate by moisture content (model fit $F = 61.15$, df = 9,
+\includegraphics[width=8cm]{figs_tables/fig2_moisture_flam.pdf}
+\caption{Fuel moisture effects on flammability. A) Flame spread rate by moisture content (model fit $F = 61.15$, df = 9,
   148, $P < 0.0001$). Model ANOVA table in Table \ref{tab:spreadrate_anova} and
-  coefficients in SI Table 4.}
-  \label{fig:spread_moist}
+  coefficients in SI Table 4. B) Percent fuel consumed by moisture content. Lines indicate best fit
+  beta regression lines. Model deviance table in Table \ref{tab:consume_anova}
+  and coefficients are in SI Table 5.}
+  \label{fig:flam_moist}
 \end{figure}
@@
 
-%% fuel consumption by moisture:
-\begin{figure}[h]
-  \centering
-\includegraphics[width=8cm]{figs_tables/fig3_consume_actualMC.pdf}
-\caption{Percent fuel consumed by moisture content. Lines indicate best fit
-  beta regression lines. Model deviance table in Table \ref{tab:consume_anova}
-  and coefficients are in SI Table 5.}
-  \label{fig:consume_moist}
-\end{figure}
-
 \begin{table}[h]
@@
 time since wetting had a significant effect on spread rate ($P < 0.0001$) and
-the slope of this effect varied by taxonomic group (Fig. \ref{fig:spread_time},
+the slope of this effect varied by taxonomic group (Fig. \ref{fig:drydown},
 Table~\ref{tab:spreadrate_anova}). The pines were able to ignite and burn much
@@
 model and figure. At 72 hours, the \emph{Abies}, \emph{Calocedrus}, and
-\emph{Sequoiadendron} had dryer litter than did the \emph{Quercus}, yet the
+\emph{Sequoiadendron} had drier litter than did the \emph{Quercus}, yet the
 spread rates were slow. \emph{Quercus} had the steepest response to time since
-wetting (Fig.~\ref{fig:spread_time}, SI Table~6)).
-
-\begin{figure}[h]
-  \centering
-\includegraphics[width=8cm]{figs_tables/fig4_spread_time.pdf}
-\caption{Flame spread rate by time since wetting for litter of eight species.
-  Lines indicate best fit linear model fits. Model ANOVA table in Table
-  \ref{tab:spreadrate_anova} and coefficients in SI Table 6. Horizontal jitter
-  (up to 2 hr) was added to the plot to avoid overplotting. All collections
-  were at the times labeled on the x-axis. A lack of data at earlier time
-  points for a taxon indicates no ignitions occurred at those time points.}
-\label{fig:spread_time}
-\end{figure}
+wetting (Fig.~\ref{fig:drydown}, SI Table~6)).
 
@@
 less common than were very low or moderately high consumption
-(Fig.~\ref{fig:consume_time}). All parameters significantly contributed to the
+(Fig.~\ref{fig:drydown}). All parameters significantly contributed to the
 generalized linear mixed beta regression model as indicated by the nested model
@@
 needles than the other two pines.
-
-\begin{figure}[h]
-  \centering
-\includegraphics[width=8cm]{figs_tables/fig5_consume_time.pdf}
-\caption{Percent fuel consumed by time since wetting for litter of eight
-  species. Curves shown indicate best fit beta regression lines. Model ANOVA
-  table in Table \ref{tab:consume_anova} and coefficients in SI Table 7.
-  Horizontal jitter (up to 2 hr) was added to the plot to avoid overplotting.
-  All collections were at the times labeled on the x-axis. A lack of data at
-  earlier time points for a taxon indicates no ignitions occurred at those time
-  points.}
-  \label{fig:consume_time}
-\end{figure}
 
@@
   \centering
-\includegraphics[width=16cm]{figs_tables/fig6_mixture_obs_vs_pred_mc.pdf}
+\includegraphics[width=16cm]{figs_tables/fig3_mixture_obs_vs_pred_mc.pdf}
 \caption{Observed vs predicted fuel moisture content for litter mixtures.
@@
   \centering
-\includegraphics[width=16cm]{figs_tables/fig7_mixture_obs_vs_pred_spread.pdf}
+\includegraphics[width=16cm]{figs_tables/fig4_mixture_obs_vs_pred_spread.pdf}
 \caption{Observed vs. predicted flame spread rate for litter mixtures. Filled
@@
   \centering
-\includegraphics[width=16cm]{figs_tables/fig8_mixture_obs_vs_pred_consume.pdf}
+\includegraphics[width=16cm]{figs_tables/fig5_mixture_obs_vs_pred_consume.pdf}
 \caption{Observed vs predicted fuel consumption for litter mixtures. Filled
@@
 fails to ignite until after 72 hours of drying in this study (Figs.
-\ref{fig:spread_time} and \ref{fig:consume_time}). However, high moisture
+\ref{fig:drydown} and \ref{fig:drydown}). However, high moisture
 content in this oak was unable to suppress ignition of mixtures to which it
@@
 varied across these taxa. Only the pines are able to ignite and support flame
-spread immediately after wetting (Fig. \ref{fig:spread_time}). In contrast,
+spread immediately after wetting (Fig. \ref{fig:drydown}). In contrast,
 \emph{Quercus} exhibited a threshold response to drying with no ignition prior

### 2020-04-02 scripts/moisture-ms-figs-tables.R

@@ -4,6 +4,7 @@
 library(xtable)
 library(ggrepel)
 library(glmmADMB)
+library(cowplot)
 
 RESULTS <- "../ms/figs_tables/"
 
@@ -45,17 +46,19 @@ summarize(x, n=length(l))
 
 
 ###############################################################################
-## Figure 1: dry-down
+## Figure 1a: dry-down
 ###############################################################################
 xbreaks <- seq(0, 144, 24)
 ybreaks <- seq(0, 700, 50)
 
-fig1 <- ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
+fig1a <- ggplot(mc, aes(hour, MC_dry, group=display.name, color=taxon)) +
   geom_jitter(height=0, width=0.5, size=1.2, alpha=0.7, stroke=0) +
   geom_smooth(method="glm",
-              method.args=list(family=gaussian(link="log")), se=FALSE, size =0.6) +
+              method.args=list(family=gaussian(link="log")), se=FALSE, size=1) +
   scale_colour_manual(values=schwilkcolors) +
-  xlab("Hours since wetting") + ylab("Moisture by dry weight (%)") +
+  #xlab("Time since wetting (hr)") +
+  xlab("") +
+  ylab("Moisture by dry weight (%)") +
   scale_x_continuous(breaks=xbreaks) +
   scale_y_continuous(breaks=ybreaks) +
   pubtheme.nogridlines +
@@ -64,8 +67,8 @@ fig1 <- ggplot(mc, aes(hour, MC_dry, group=display.name, color=genus)) +
         legend.text = element_text(family=fontfamily, size=smsize, face="italic"))
 #       legend.key.height=unit(smsize,"pt"))
 
-ggsave(file.path(RESULTS, "fig1_drydown-curves.pdf"), plot=fig1,
-       width=col1, height=col1, units="cm")
+## ggsave(file.path(RESULTS, "fig1a_drydown-curves.pdf"), plot=fig1a,
+##        width=col1, height=col1, units="cm")
 
 ###############################################################################
 # SI fig 2: dry down on semi log scale
@@ -250,13 +253,13 @@ oldflam <- read.csv("../data/burn-trials/flamdt.csv") %>% right_join(species) %>
   mutate(hour=144, vpd=2.44, t2ignit=ignit, combust=combus) #, actualMC_dry=5)
 b <- oldflam %>% filter(nchar(as.character(spcode)) < 5) %>% bind_rows(burnt)
 
[…]

### 2020-04-23 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Start on response letter and intro edits

@@
 
-      \item \emph{Synthesis:} Taxon specific patterns of moisture absorption
-    and drying can change flammability rankings under moist compared with dry
-    fuels because traits that increase flame spread rate increase water
-    absorption. Litter mixtures generally burn with fire behavior more similar
-    to that of the most flammable constituent species and this effect increases
-    as fuels dry.
+      \item \emph{Synthesis:} This study is the first to examine ecologically
+    relevant drying rates and their interaction with non additive misture
+    effects on flammability. We find that taxon specific patterns of moisture
+    absorption and drying change flammability rankings under moist compared
+    with dry fuels because traits that increase flame spread rate increase
+    water absorption. Litter mixtures generally burn with fire behavior more
+    similar to that of the most flammable constituent species and this effect
+    increases as fuels dry.
 
@@
 flammability in moist fuels with one study finding increasing nonadditivity
-with moisture \citep{Blauw+Wensink+etal-2015}. In mixed conifer forest in
-California, where the fire regime is characterized by surface fires fueled by
-plant litter, overstory tree composition drives variation in local fire
-behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
-flammability across tree species and as a result of synergistic interactions in
+with moisture \citep{Blauw+Wensink+etal-2015}, but this previous work did not
+include estimation of the dry-down process and had an unclear relationship to
+fire behavior int he field.. In mixed conifer forest in California, where the
+fire regime is characterized by surface fires fueled by plant litter, overstory
+tree composition drives variation in local fire behavior
+\citep{Schwilk+Caprio-2011} as a result of differential litter flammability
+across tree species and as a result of synergistic interactions in
 multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
@@
 
-We asked if differences in litter moisture absorption and drying rates could
-change the relative flammability of litter across species compared with
-flammability measurements based on dry fuels. We investigated these litter
-moisture dynamics across eight tree species in a temperate fire-prone forest in
-California. We determined the maximum water absorption and rate of moisture
-loss and examined how these traits vary in relation to specific leaf area and
-litter packing density. Experimental burn trials of leaf litter at various
-moisture levels determined how time since wetting influenced two axes of
-flammability characterized by flame spread rate and total fuel consumed. To
-determine the potential magnitude of mixture effects, we tested if the
-nonadditive effects on reported in previous studies occur under higher moisture
-levels, and whether fuel moisture effects strengthen or moderate any
-nonadditivity.
+This study is the first to examine ecologically relevant drying rates and their
+interaction with non additive moisture effects on flammability. We asked if
+differences in litter moisture absorption and drying rates could change the
+relative flammability of litter across species compared with flammability
+measurements based on dry fuels. We investigated these litter moisture dynamics
+across eight tree species in a temperate fire-prone forest in California. We
+determined the maximum water absorption and rate of moisture loss and examined
+how these traits vary in relation to specific leaf area and litter packing
+density. Experimental burn trials of leaf litter at various moisture levels
+determined how time since wetting influenced two axes of flammability
+characterized by flame spread rate and total fuel consumed. To determine the
+potential magnitude of mixture effects, we tested if the nonadditive effects on
+reported in previous studies occur under higher moisture levels, and whether
+fuel moisture effects strengthen or moderate any nonadditivity.
 

### 2020-04-29 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Edits in response to reviews

@@
 \documentclass[letterpaper,12pt]{article}
+\usepackage{caption}
+\DeclareCaptionLabelFormat{adja-page}{\hrulefill\\#1 #2 \emph{(previous page)}}
 \usepackage{jecol}
+\usepackage{endfloat}
 
@@
     drying rates. Burn trials were performed in litter beds of single species
-    and of mixtures to establish flammability response to the dry-down process.
+    and of mixtures to establish flammability response to the dry-down process
+    and to species interactions.
+
+      \item Litter mixtures exhibited nonadditivity in flame spread rate and in
+    fuel consumption. This nonadditivity became more positive with fuel drying
 
@@
     area being associated with higher maximum absorption and lower litter bulk
-    density being associated with faster drying rates. We examined two axes of
-    flammability variation and found that these differing moisture dynamics
-    resulted in time since wetting having strongly differing effects on
-    flammability across the major litter types typified by different tree taxa.
-
-      \item Litter mixtures exhibited nonadditivity in flame spread rate and
-    in fuel consumption. This nonadditivity became more positive with fuel
-    drying.
+    density being associated with faster drying rates. We found that these
+    differing moisture dynamics resulted in time since wetting having strongly
+    differing effects on flammability across the major litter types typified by
+    different tree taxa..
 
@@
 living fuels has strong effects on flammability, but moisture of dead excised
-litter fuels can also vary. In particular, fine dead fuels respond very readily
+litter fuels can also vary. In particular, fine dead fuels respond readily
 to environmental changes \citep{Nelson-2001}. Moisture of litter fuels varies
@@
 variation are described by measurements related to flame spread rate and
-measurements related to heat release or heat transfer
+measurements related to heat release or heat total transfer
 \citep{Magalhaes+Schwilk-2012, Prior+Murphy+etal-2018}. These different
 components of fire behavior have differing ecological effects with spread rate
-being associated with rates of heat release and area burned, and with heat
+being associated with flame heights and area burned, and with total heat
 transfer predicting soil impacts of fire and tree mortality. A shortcoming in
@@
 combustion, and indirectly by its effect on drying rate. Both these pathways
-are affected by leaf particle traits, especially leaf size
-\citep{Scarff+Westoby-2006}. Fuel-beds composed of larger leaves or leaves with
-thin cuticles should favor moisture absorption. On the other hand, large leaves
-lower litter density and should favor drying, thus increasing the probability
-of ignition and fire spread. The interaction of moisture absorption and drying
-deserves investigation.
+are influenced by leaf particle traits, especially leaf size
+\citep{Scarff+Westoby-2006}. Fuel-beds comprising larger leaves or leaves with
+thin cuticles should favor moisture absorption. Conversely, large leaves lower
+litter density and should favor drying, thus increasing the probability of
+ignition and fire spread. The interaction of these processes, especially in
+species mixtures, is poorly known.
 
@@
 flammability of moist fuels. There is some evidence for nonadditive
-flammability in moist fuels with one study finding increasing nonadditivity
+flammability in moist fuels with a single study finding increasing nonadditivity
 with moisture \citep{Blauw+Wensink+etal-2015}, but this previous work did not
 include estimation of the dry-down process and had an unclear relationship to
-fire behavior int he field.. In mixed conifer forest in California, where the
+fire behavior in the field. In mixed conifer forest in California, where the
 fire regime is characterized by surface fires fueled by plant litter, overstory
@@
 top 2.5-5\,cm of litter were collected by raking and shoveling into large bags.
-All litter was oven dried to $< 5$\% fuel moisture by dry weight, then weighed
-and divided into moisture trials. The collected litter was stored in large
-paper bags and cardboard boxes in a lab storage space. We minimized handling
-and moving the bagged litter to avoid mechanical damage and changes to the
-particle size distributions. The moisture manipulations and subsequent burning
-trials were conducted in April--August 2015.
+All litter was oven dried at $60^\circ$C to $< 5$\% fuel moisture by dry
+weight, then weighed and divided into moisture trials. The collected litter was
+stored in large paper bags and cardboard boxes in a lab storage space. We
+minimized handling and moving the bagged litter to avoid mechanical damage and
+changes to the particle size distributions. The moisture manipulations and
+subsequent burning trials were conducted in April--August 2015.
 
@@
 following saturation and gravity draining. Samples from the eight species (six
-replicates per species) and four mixtures (5 replicates per mixture type) were
+replicates per species) and four mixtures (five replicates per mixture type) were
 placed in baskets approximately 45 x 45\,cm by 15\,cm tall constructed of
@@
 was less than 10\% moisture. This experimental setup was designed to accelerate
-drying compared with field conditions because the baskets were suspended in air
+drying compared with field conditions because the baskets were suspended in air that was
 2.8 -- 5.3$^\circ$C above the warmest month average temperatures for the
@@
 (Nielsen-Kellerman, Boothwyn, PA) and these values were used to calculate vapor
-pressure deficit as a single weather covariate at each trial. All trials were
+pressure deficit as a weather covariate at each trial. All trials were
 conducted on clear days between 10 am and 3 pm. We scheduled burns in a
@@
 litter bed density before burning. This design allows fire to reach constant
-flame spread rates \citep{Magalhaes+Schwilk-2012}. Two metal rulers equally
+flame spread rates within the range of rates reported for field burns \citep{Magalhaes+Schwilk-2012}. Two metal rulers equally
 spaced along the apparatus allowed visual determination of maximum flame
@@
 height were recorded as zero. Our thermocouple system failed or data was lost
-for two days of burning trials, therefore we have omitted those temperature
+for two days of burning trials, therefore we have omitted temperature
 data from this analysis. Our past work \citep{Magalhaes+Schwilk-2012}
@@
 packing. These previously reported flammability trials demonstrated that
-species within a genus tended to behave similarly and the two scale-leaved
+species within a genus tended to behave similarly and that the two scale-leaved
 genera in the Cupressaceae, \emph{Calocedrus} and \emph{Sequoiadendron}, also
@@
 as our flammability functional groups and treated these as fixed effects in our
-models. The effect of fuel moisture and time since wetting on flammability
+models. The effects of fuel moisture and time since wetting on flammability
 parameters were modeled using linear models with fuel moisture or time since
@@
 
-We investigated possible nonadditive effects on moisture and on flammability by
+We investigated nonadditive effects on moisture and on flammability by
 testing if moisture content and flammability of mixtures were predicted by the
-average of the constituent species. To choose mixture combination, we first
+average of the constituent species. To choose mixture combinations, we first
 grouped the species according to their maximum water absorption and drying rate
@@
 
-\subsection*{Moisture dynamics across species}
+\subsection*{Moisture dynamics}
 
@@
 discriminated species into three groups from lowest to most rapid: 1) \emph{C.
-  decurrens}, 2) \emph{Abies spp} and \emph{S. giganteum}, and 3) \emph{Q.
-  kelloggii} and \emph{Pinus spp} (SI Fig.~3).
+  decurrens}, 2) \emph{Abies spp.} and \emph{S. giganteum}, and 3) \emph{Q.
+  kelloggii} and \emph{Pinus spp.} (SI Fig.~3).
 
-\begin{figure}[h]
+\begin{figure}
   \centering
-\includegraphics[width=5cm]{figs_tables/fig1_dry_effects_by_taxon.pdf}
+  \includegraphics[width=8cm]{figs_tables/fig1_dry_effects_by_taxon.pdf}
+  \caption{}
+\end{figure}
+%%\clearpage
+\begin{figure}
+  \captionsetup{labelformat=adja-page}
+  \ContinuedFloat
 \caption{Effects of litter drying on fuel moisture and flammability. A) Dry
@@
 \end{figure}
+%% \clearpage
 
-\begin{table}[h]
+\begin{table}
   \caption{Linear mixed model results for moisture content as a function of
@@
     statistics and p-values were calculated by the Kenward-Roger approximation.
-    \citep{Kenward_Roger-1997}. Model coefficients reported in SI Table~1.}
+    \citep{Kenward_Roger-1997}. Model coefficients are reported in SI Table~1.}
   \label{tab:drydown}
@@
 
-Across these eight species specific leaf area had a positive association with
+Across these eight species, specific leaf area had a positive association with
 maximum water absorption (Table~\ref{tab:mc_di_anova}, SI Fig~5, linear model
@@
 
-\begin{table}[h]
+\begin{table}
   \caption{Linear model results for maximum water absorption (\% moisture
@@
     species. ANOVA tables shown with degrees of freedom, sum of squares, mean
-    squares, $F$, and $P$ for each effect. Model coefficients in SI Tables~2 and 3}
+    squares, $F$, and $P$ for each effect. Model coefficients in SI Tables~2 and 3.}
   \label{tab:mc_di_anova}
@@
 %% spread rate by moisture
-\begin{figure}[h]
+\begin{figure}
   \centering
@@
 %% Table 3 
-\begin{table}[h]
+\begin{table}
   \caption{Generalized linear mixed model results for two models predicting
@@
     approximation \citep{Kenward_Roger-1997}. Estimated coefficients are in SI
-    Tables 4 and 6}
+    Tables 4 and 6.}
   \label{tab:spreadrate_anova}
@@
 
-\begin{table}[h]
+\begin{table}
   \caption{Generalized linear mixed model results for two sets of models
@@
 spread rate, pines had rapid flammability increases with drying. However,
-unlike in the other flammability analyses, here the pines did not all behave
-similarly: \emph{P. lambertiana} had lower fuel consumption than did \emph{P.
+in contrast to the other flammability analyses, here the pines did not behave
+consistently: \emph{P. lambertiana} had lower fuel consumption than did \emph{P.
   ponderosa} and \emph{P. jeffreyi}. We noticed this during trials and in model
@@
 
-Based on visual inspection of the dry down curves and the estimated marginal
+Based on the estimated marginal
 means for maximum water absorption and drying rate (SI Figs 3--4), we selected
-four species to represent the full range of litter moisture dynamics for use in
+four species for use in
 the litter mixtures and examined all four three-species combinations of those
@@
 
-Mixture drying rates and maximum water absorption showed some relationship to
-the dry down curves for individual constituent species (see SI Fig.~7). For
-example, the mixture that does not contain \emph{Q. kelloggii} (AbCaPi in SI
-Fig.~7) has the lowest maximum water absorption. Maximum water absorption and
-drying rates differed across mixtures (linear model based on log of moisture
-content, SI Tables~8 and 9).
-
-\begin{figure}[h]
+\begin{figure}
   \centering
@@
 [Fig.~\ref{fig:mixture_obs_pred_mc}] less than 1, $P < 0.0001$). 
+Mixture drying rates and maximum water absorption showed some relationship to
+the dry down curves for individual constituent species (see SI Fig.~7). For
+example, the mixture that does not contain \emph{Q. kelloggii} (AbCaPi in SI
+Fig.~7) has the lowest maximum water absorption. Maximum water absorption and
+drying rates differed across mixtures (linear model based on log of moisture
+content, SI Tables~8 and 9).
 
-\begin{figure}[h]
+\begin{figure}
   \centering
@@
 
-\begin{figure}[h]
+\begin{figure}
   \centering
@@
 contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
-\ref{fig:mixture_obs_pred_consume}).
+\ref{fig:mixture_obs_pred_consume}), indicating the importance of non-additive effects.
 
@@
 oak has much higher maximum water absorption than do the other species,
-followed by the finest-needled pine, \emph{P. lambertiana}. Our study examines
-fresh litter, but we expect that decomposition will affect moisture dynamics.
-For example, Van Wagner \citeyear{Van_Wagner-1969}, investigating fuel
-weathering, found that pine needles had a variation in moisture content 3 to 6
-times greater when they lacked their protective wax coating when compared to
-intact needles.
+followed by the finest-needled pine, \emph{P. lambertiana}. We expect that
+decomposition will affect these dynamics. For example, Van Wagner
+\citeyear{Van_Wagner-1969}, investigating fuel weathering, found that pine
+needles had a variation in moisture content 3 to 6 times greater when they
+lacked their protective wax coating when compared to intact needles.
 
@@
 mixtures. These mixtures contained species representative of all dry-down
-behavior types observed in single species trials. Mixtures that contained
-\emph{Quercus} absorbed more water during wetting, and the one mixture without
-any Cupressaceae had the most rapid drying rate (SI Figure 7 and SI Table 8).
-Drying rates, however, did not behave as the average of the constituent
-species: mixtures tended to be drier than expected early in the process and
-wetter than expected later (Fig. \ref{fig:mixture_obs_pred_mc}). Past work has
-shown that mixtures tend to pack less densely than do litter monocultures
-\citep{Magalhaes+Schwilk-2012}. If packing density controls rate of drying,
-then we would expect mixtures to dry more rapidly than predicted from
-monoculture trials. Therefore, it is surprising that mixtures tended to have
-lower maximum water content than predicted from the mean of the constituent
-species, but higher water content than expected after long drying. This
-discrepancy may be a consequence of the lightly packing litters having high
-maximum water absorption and of drying being non-linear with time
-(Fig.~\ref{fig:drydown}).
+behavior types observed in single species trials. Drying rates, however, did
+not behave as the average of the constituent species: mixtures tended to be
+drier than expected early in the process and wetter than expected later (Fig.
+\ref{fig:mixture_obs_pred_mc}). Past work has shown that mixtures tend to pack
+less densely than do litter monocultures \citep{Magalhaes+Schwilk-2012}. If
+packing density controls rate of drying, then we would expect mixtures to dry
+more rapidly than predicted from monoculture trials. Therefore, it is
+surprising that mixtures tended to have lower maximum water content than
+predicted from the mean of the constituent species, but higher water content
+than expected after long drying. This discrepancy may be a consequence of the
+lightly packing litters having high maximum water absorption and of drying
+being non-linear with time (Fig.~\ref{fig:drydown}).
 
-The overall ranking of flammability in litter mixtures showed some consistency
-with the results obtained for the individual species, but there were
-significant nonadditive effects. Mixtures containing \emph{Q. kelloggii} and
-\emph{P. jeffreyi} tended to have fast spread rates. These results are likely
-influenced by the lower bulk densities of litter mixtures containing those two
-species. Most interestingly, there were patterns of nonadditivity in spread
-rate and in fuel consumption for the four mixtures, indicating that positive
-nonadditivity in flammability is not restricted to the completely dry litter
-mixtures studied in previous work \citep{Magalhaes+Schwilk-2012,
-  VanAltena+Logtestjin+etal-2012}. Nonadditivity in both spread rate and fuel
-consumption became more positive with time since wetting and these patterns
-were consistent across mixture identities. This is a slightly different result
-than that reported by \citet{Blauw+Wensink+etal-2015}. Blauw et
-al found that nonadditivity was greater in moist mixtures of heathland fuels
-than in dry fuels. In their work, nonadditivity was generally negative but
-cases of positive nonadditivity in mixtures was often a result of mixtures
-meeting ignition thresholds not met by the constituent species in monoculture.
-To limit the number of models fit, we limited our investigation to two
-flammability parameters that represent two major axes of flammability
-variation. Therefore, our study does not fully capture ignition effects
-although we do have some results consistent with theirs: mixtures that included
-\emph{Q. kelloggii} could ignite at times since wetting at which \emph{Quercus}
-in monoculture could not.
+%  Mixtures that contained
+% \emph{Quercus} absorbed more water during wetting, and the one mixture without
+% any Cupressaceae had the most rapid drying rate (SI Figure 7 and SI Table 8).
+
+% The overall ranking of flammability in litter mixtures showed some consistency
+% with the results obtained for the individual species, but there were
+% significant nonadditive effects. Mixtures containing \emph{Q. kelloggii} and
+% \emph{P. jeffreyi} tended to have fast spread rates. These results are likely
+% influenced by the lower bulk densities of litter mixtures containing those two
+% species.
+
+There were patterns of nonadditivity in spread rate and in fuel consumption for
+the four mixtures, indicating that positive nonadditivity in flammability is
+not restricted to the completely dry litter mixtures studied in previous work
+\citep{Magalhaes+Schwilk-2012, VanAltena+Logtestjin+etal-2012}. Nonadditivity
+in both spread rate and fuel consumption became more positive with time since
+wetting and these patterns were consistent across mixture identities. This is a
+slightly different result than that reported by
+\citet{Blauw+Wensink+etal-2015}. Blauw et al found that nonadditivity was
+greater in moist mixtures of heathland fuels than in dry fuels. In their work,
+nonadditivity was generally negative but cases of positive nonadditivity in
+mixtures was often a result of mixtures meeting ignition thresholds not met by
+the constituent species in monoculture. We do have some results consistent with
+theirs: mixtures that included \emph{Q. kelloggii} could ignite at times since
+wetting at which \emph{Quercus} in monoculture could not.
 
@@
 expected \citep{Zhao+vanLogtestijn+etal-2019}.
-
-% (eg \emph{Qu)(eg \emph{Abies} branchlets with needles fitting between loosely
-% packed \emph{Pinus} needles) or less dense
 
@@
 
-Conceptualization, RMQM and DWS; methodology, RMQM and DWS; experiments, RMQM;
+Conceptualization, DWS and RMQM; methodology, RMQM and DWS; experiments, RMQM;
 coding and data analysis, DWS and RMQM; writing DWS and RMQM; writing--review

### 2020-04-29 scripts/moisture-ms-figs-tables.R

@@ -513,8 +513,9 @@ fig1c
 ## ggsave(file.path(RESULTS, "fig5_consume_time.pdf"), plot=fig5,
 ##        width=col1, height=col1, unit="cm")
 
-
-## New Fig1 ###
+###############################################################################
+## Fig1 combined a,b,c  ###
+##############################################################################
 fig1 <- plot_grid(fig1a, fig1b, fig1c, labels=c("A", "B", "C"), ncol=1, align="h")
 fig1
 

### 2020-05-01 ms/SI.tex
Change SI fig order

@@
   \centering
-  \includegraphics[width=8cm]{figs_tables/SI_fig1_litter-particle-lengths-by-species.pdf}
+  \includegraphics[width=8cm]{figs_tables/SI_fig2_drydown-curves_logged.pdf}
    \label{SI_fig1}
-   \caption[Litter particle length distributions]{10 2\,g subsamples of litter per species were assembled and each individual particle of that subsample was measured for length (longest axis), width (longest axis perpendicular to length) and thickness (perpendicular to length and width, at midrib) using an electronic caliper.
-Violin plots here show the distribution of the sum of the three dimensions as a measure of the containing prism volume (N=27--450 per species).}
-\end{figure}
-
-\begin{figure}[H]
-  \centering
-  \includegraphics[width=8cm]{figs_tables/SI_fig2_drydown-curves_logged.pdf}
-   \label{SI_fig2}
-\caption[Semi-log scale dry down curves for eight litter types.]{Lines show best-fit lines fit to the natural log of moisture content on a dry mass basis.
-The slope of the line shows the dessication rate ($hr^{-1}$).
+\caption{Semi-log scale dry down curves for eight litter types.
+Lines show best-fit lines fit to the natural log of moisture content on a dry mass basis.
+The slope of the line shows the drying rate ($hr^{-1}$).
 Litter samples we saturated by immersion for 24 hours and then allowed to drain by gravity for 3 min before initial weighing.
-Colors indicate genera and there were significant differences in both }
+Colors indicate genera and there were significant differences in both maximum absorption and drying rate (main text Table 1.}
 
@@
   \centering
-  \label{SI_fig3}
+  \label{SI_fig1}
   \includegraphics[width=16cm]{figs_tables/SI_fig3_max_water_emmeans.pdf}
@@
   \centering
-  \label{SI_fig4}
+  \label{SI_fig3}
 \includegraphics[width=16cm]{figs_tables/SI_fig4_dessication_emmeans.pdf}
-\caption{Estimated marginal means by species for dessication rate based on mixed model results. Shaded regions indicate 95\% confidence intervals for the marginal means.}
+\caption{Estimated marginal means by species for drying rate based on mixed model results. Shaded regions indicate 95\% confidence intervals for the marginal means.}
 \end{figure}
@@
 \includegraphics[width=8cm]{figs_tables/SI_fig5_SLA_maxMC.pdf}
-\caption{Maximum water retention by specific leaf area (SLA) across eight species.}
-  \label{fig:sla-maxmc}
+\caption{Maximum water absorption by specific leaf area (SLA) across eight species.}
+  \label{SI_fig4}
 \end{figure}
@@
 \includegraphics[width=8cm]{figs_tables/SI_fig6_di_bd.pdf}
-\caption{Linear model predicting dessication rate as a function of specific leaf area and litter bulk density across eight species.}.
-  \label{fig:bd-di}
+\caption{Linear model predicting drying rate as a function of specific leaf area and litter bulk density across eight species.}.
+  \label{SI_fig5}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=8cm]{figs_tables/SI_fig1_litter-particle-volumes-by-species.pdf}
+   \label{SI_fig6}
+   \caption{Litter particle volume distributions.
+10 2\,g subsamples of litter per species were assembled and each individual particle of that subsample was measured for length (longest axis), width (longest axis perpendicular to length) and thickness (perpendicular to length and width, at midrib) using an electronic caliper.
+Violin plots here show the distribution of the sum of the three dimensions as a measure of the containing prism volume (N=27--450 per species).}
 \end{figure}

### 2020-05-01 scripts/moisture-ms-figs-tables.R

@@ -21,7 +21,7 @@ set.seed(100)
 decompsp <- decomp %>% left_join(species)
 
 ###############################################################################
-## SI Fig 1: litter particle length distributions
+## SI Fig 6: litter particle length distributions
 ###############################################################################
 
 SI_fig1 <- ggplot(subset(decompsp, year==0), aes(display.name, l+w+t)) +

### 2020-05-01 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
fixup

@@
 \usepackage{jecol}
-\usepackage{endfloat}
+\usepackage[notablist,nofiglist]{endfloat}
+%\usepackage{color}
+%\usepackage{soul}
+%\renewcommand{\hl}[1]{#1}
 
@@
         \item  Litter mixtures exhibited nonadditivity in flame spread rate and in
-    fuel consumption. This nonadditivity became more positive with fuel drying
+    fuel consumption. This nonadditivity became more positive with fuel drying.
 
@@
 vapor will both cool the flaming front and dilute flammable gases
-\citep{Albini-1976, Shafizadeh-1977}. Environmental conditions such as
-temperature, relative humidity, and wind drive the moisture content of the fuel
-available to burn \citep{Kreye+Varner+etal-2018}. Variation in moisture of
-living fuels has strong effects on flammability, but moisture of dead excised
-litter fuels can also vary. In particular, fine dead fuels respond readily
-to environmental changes \citep{Nelson-2001}. Moisture of litter fuels varies
-with climate, with seasons, and with short term weather within a season.
-Therefore, both the rate of drying and flammability as a function of moisture
-content are key parameters that influence fuel flammability throughout the fire
-season.
+\citep{Albini-1976, Shafizadeh-1977}. Variation in moisture of living fuels has
+strong effects on flammability, but moisture of dead excised litter fuels can
+also vary. In particular, fine dead fuels respond readily to environmental
+conditions such as temperature, relative humidity, and wind \citep{Nelson-2001,
+  Kreye+Varner+etal-2018}. Moisture of litter fuels varies with climate, with
+seasons, and with short term weather within a season. Therefore, both the rate
+of drying and flammability as a function of moisture content are key parameters
+that influence fuel flammability throughout the fire season. However, we lack
+information on species-specific flammability responses to moisture dynamics. We
+neither understand the relative importance of species differences in such
+dynamics nor how these might interact to influence flammability in litter beds
+comprising multiple species in forest communities.
 
@@
 \citep{Schwilk-2015, Pausas+Keeley+etal-2017} with different traits influencing
-different components of fire behavior. In many forests that experience surface
-fire, litter properties influence fire behavior
-\citep{Ganteaume+Marielle+etal-2011, Schwilk+Caprio-2011,
-  Varner+Kane+etal-2015}. In litter driven fire, the two major axes of
-variation are described by measurements related to flame spread rate and
-measurements related to heat release or heat total transfer
+different components of fire behavior. In forests that experience surface fire,
+litter properties influence fire behavior \citep{Ganteaume+Marielle+etal-2011,
+  Schwilk+Caprio-2011, Varner+Kane+etal-2015}. In litter driven fire, the two
+major axes of variation are described by measurements related to flame spread
+rate and measurements capturing heat release or total heat transfer
 \citep{Magalhaes+Schwilk-2012, Prior+Murphy+etal-2018}. These different
 components of fire behavior have differing ecological effects with spread rate
-being associated with flame heights and area burned, and with total heat
-transfer predicting soil impacts of fire and tree mortality. A shortcoming in
-our current knowledge is our integration of specific species and plant traits
-into fuel hazard assessment, fire modeling, and fire effects prediction.
-Developing quantitative yet robust and general ways to integrate actual plant
-communities into fuel assessment will require understanding the magnitude of
-flammability variability across species in space and time and parameterizing
-based on plant characteristics.
+being associated with flame heights and area burned, and with heat transfer
+predicting soil impacts of fire and tree mortality. A shortcoming in our
+current knowledge is our integration of specific species and plant traits into
+fuel hazard assessment, fire modeling, and fire effects prediction. Developing
+quantitative yet robust and general ways to integrate actual plant communities
+into fuel assessment will require understanding the magnitude of flammability
+variability across species in space and time and parameterizing based on plant
+characteristics.
 
@@
   Cornwell+Elvira+etal-2015} and fire regime can vary over fine spatial scales
-in response to community composition and species-specific flammability
-\citep{Stephens-2001, Hille+Stephens-2005, Schwilk+Caprio-2011}. Litter
-packing, measured as litter-bed density or packing ratio, is an important
-control on ignition and spread in models of surface fire \citep{Weber-1991,
-  Balbi+Santoni+etal-1999, Morvan+Larini-2001}. Multiple studies have
-demonstrated that litter bed density controls the flammability of dry litter
-fuels in seasonally dry climates where higher litter density reduces flame
-spread rate. Across these studies, there is increasing consensus that larger,
-less densely packed litter beds increase oxygen availability and burn with
-faster spread rates \citep{Scarff+Westoby-2006, Schwilk+Caprio-2011,
-  Magalhaes+Schwilk-2012, Cornwell+Elvira+etal-2015, Varner+Kane+etal-2015,
-  Kreye+Varner+etal-2018}. The litter packing is driven by litter particle size
-and shape, which is in turn influenced by leaf size, how leaves or branches are
-abscised, and subsequent decomposition \citep{Fonda+Belanger+etal-1998,
+in response to such species-specific flammability \citep{Stephens-2001,
+  Hille+Stephens-2005, Schwilk+Caprio-2011}. Litter packing, measured as
+litter-bed density or packing ratio, is an important control on ignition and
+spread in models of surface fire \citep{Weber-1991, Balbi+Santoni+etal-1999,
+  Morvan+Larini-2001}. Multiple studies have demonstrated that litter bed
+density controls the flammability of dry litter fuels in seasonally dry
+climates: higher litter density reduces flame spread rate. Across these
+studies, there is increasing consensus that less densely packed litter beds
+increase oxygen availability and burn with faster spread rates
+\citep{Scarff+Westoby-2006, Schwilk+Caprio-2011, Magalhaes+Schwilk-2012,
+  Cornwell+Elvira+etal-2015, Varner+Kane+etal-2015, Kreye+Varner+etal-2018}.
+The litter packing is driven by litter particle size and shape, which is in
+turn influenced by leaf size, how leaves or branches are abscised, and
+subsequent decomposition \citep{Fonda+Belanger+etal-1998,
   Stephens+Finney+etal-2004, Scarff+Westoby-2006, Kane+Varner+etal-2008,
@@
 
-Differential drying is especially important when precipitation occurs
-during the same season as ignitions. Although synoptic weather conditions that
-create dry and wet lightning storms differ, the potential for precipitation and
+Differential drying is especially important when precipitation occurs during
+the same season as ignitions. Although synoptic weather conditions that create
+dry and wet lightning storms differ, the potential for precipitation and
 lightning during the same overall period of thunderstorm activity often exists:
-wet and dry convection days can occur under overlapping conditions,
+wet and dry convection days can occur under overlapping conditions
 \citep{Rorig+Ferguson-1999}. Moisture levels of fine fuels respond rapidly to
-precipitation events during the fire season \citep{Estes+Knapp+etal-2012}.
-Differential moisture content across time may contribute to the often high
-spatial heterogeneity of surface fires \citep{Hille+Stephens-2005,
-  Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and to variation in fire regime
-across landscapes \citep{Stephens-2001}.
+precipitation events during the fire season \citep{Estes+Knapp+etal-2012} and
+this can lead to repeated wetting and drying periods during which lightning
+ignition may occur. Such differential moisture content across time may
+contribute to the often high spatial heterogeneity of surface fires
+\citep{Hille+Stephens-2005, Knapp+Keeley-2006, Kreye_Hiers_etal-2018} and to
+variation in fire regime across landscapes \citep{Stephens-2001}.
 
@@
 litter density and should favor drying, thus increasing the probability of
-ignition and fire spread. The interaction of these processes, especially in
-species mixtures, is poorly known.
+ignition and fire spread. The interactions among these processes, especially in
+species mixtures, are poorly known.
 
@@
 interaction with non additive moisture effects on flammability. We asked if
-differences in litter moisture absorption and drying rates could change the
+differences in litter moisture absorption and drying rates change the
 relative flammability of litter across species compared with flammability
@@
 \emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
-Buchholz. In this study, we consider litter the top layer of leaves and small
+Buchholz. In this study, we consider ``litter'' the top layer of leaves and small
 twigs less than 1\,cm diameter that have fallen in the previous year (mostly
@@
 following saturation and gravity draining. Samples from the eight species (six
-replicates per species) and four mixtures (five replicates per mixture type) were
-placed in baskets approximately 45 x 45\,cm by 15\,cm tall constructed of
+replicates per species) and four mixtures (five replicates per mixture type)
+were placed in baskets approximately 45 x 45\,cm by 15\,cm tall constructed of
 aluminum mesh with 0.52\,cm square openings. Samples were 450\,g dry weight
@@
 was less than 10\% moisture. This experimental setup was designed to accelerate
-drying compared with field conditions because the baskets were suspended in air that was
-2.8 -- 5.3$^\circ$C above the warmest month average temperatures for the
-closest weather stations to the collection sites (Grant Grove and Lodegpole
-stations, NOAA 1981-2010 normals,
+drying compared with field conditions because the baskets were suspended in air
+that was 2.8 -- 5.3$^\circ$C warmer than the historical average of the warmest
+summer month at the closest weather stations to the collection sites (Grant
+Grove and Lodegpole stations, NOAA 1981-2010 temperature means,
 https://www.ncdc.noaa.gov/cdo-web/datatools/normals).
@@
 litter bed density before burning. This design allows fire to reach constant
-flame spread rates within the range of rates reported for field burns \citep{Magalhaes+Schwilk-2012}. Two metal rulers equally
-spaced along the apparatus allowed visual determination of maximum flame
-height. Litter was ignited with a propane torch; maximum adiabatic flame
-temperature of 1899$^\circ$C,
-\url{http://www.benzomatic.com/products/fuel.aspx}) at one end of the apparatus
-and allowed to burn until extinction. If a sample failed to ignite after 30
-seconds, no further attempt was made and if three samples of a species at a
-particular moisture level failed to ignite, no further attempts were made on
-that species/moisture combination and the replicate was allowed to continue
-drying until the next scheduled checkpoint. The flammability trials followed
-the methodology detailed in \citet{Magalhaes+Schwilk-2012}.
+flame spread rates within the range of rates reported for field burns
+\citep{Magalhaes+Schwilk-2012}. Two metal rulers equally spaced along the
+apparatus allowed visual determination of maximum flame height. Litter was
+ignited with a propane torch; maximum adiabatic flame temperature of
+1899$^\circ$C, \url{http://www.benzomatic.com/products/fuel.aspx}) at one end
+of the apparatus and allowed to burn until extinction. If a sample failed to
+ignite after 30 seconds, no further attempt was made and if three samples of a
+species at a particular moisture level failed to ignite, no further attempts
+were made on that species/moisture combination and the replicate was allowed to
+continue drying until the next scheduled checkpoint. The flammability trials
+followed the methodology detailed in \citet{Magalhaes+Schwilk-2012}.
 
@@
 height were recorded as zero. Our thermocouple system failed or data was lost
-for two days of burning trials, therefore we have omitted temperature
-data from this analysis. Our past work \citep{Magalhaes+Schwilk-2012}
-demonstrated that flame spread rate captures one important axis of flammability
-variation \citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal-2018}
-and percent fuel consumed captures a portion of the second major axis of
+for two days of burning trials, therefore we have omitted temperature data from
+this analysis. Our past work \citep{Magalhaes+Schwilk-2012} demonstrated that
+flame spread rate captures one important axis of flammability variation
+\citep{Schwilk-2015, Pausas+Keeley+etal-2017, Prior+Murphy+etal-2018} and
+percent fuel consumed captures a portion of the second major axis of
 flammability variation. Our analysis, therefore, focused on these two
@@
 
-Past work has demonstrated that green leaf length predicts litter packing and,
-consequently, flame spread rate across these species. Litter particle size
-distributions, however show that there is substantial variation in particle
-sizes within a species (SI Fig.~1). There were differences in the maximum water
-absorption (all $P < 0.0001$, Fig.~\ref{fig:drydown} and
-Table~\ref{tab:drydown}). \emph{Q. kelloggii}, the only broadleaf, is the
-species that absorbed the most moisture initially. \emph{Quercus}, like the
-pines, had a rapid drying rate (Fig.~\ref{fig:drydown} and SI Fig.~2).
-Calculating estimated marginal means following the linear mixed-effects model
-discriminated the species into three different maximum water absorption groups:
-1) \emph{Q. kelloggii}; 2) \emph{P. lambertiana} and \emph{C. decurrens}; and
-3) \emph{A. magnifica}, \emph{S. giganteum}, \emph{A. concolor}, \emph{P.
-  jeffreyi} and \emph{P. ponderosa} (SI Fig.~2). There are also differences in
-drying rates across the eight species and estimated mixed linear model slopes
-discriminated species into three groups from lowest to most rapid: 1) \emph{C.
+Across these species there were significant differences in the maximum water
+absorption and drying rate (all $P < 0.0001$, Fig.~\ref{fig:drydown},
+Table~\ref{tab:drydown} and SI Figs.~1 -- 3). Calculating estimated marginal
+means following the linear mixed-effects model discriminated the species into
+three different maximum water absorption groups: 1) \emph{Q. kelloggii}; 2)
+\emph{P. lambertiana} and \emph{C. decurrens}; and 3) \emph{A. magnifica},
+\emph{S. giganteum}, \emph{A. concolor}, \emph{P. jeffreyi} and \emph{P.
+  ponderosa} (SI Fig.~2). Estimated mixed linear model slopes discriminated
+species into three groups from lowest to most rapid drying rate: 1) \emph{C.
   decurrens}, 2) \emph{Abies spp.} and \emph{S. giganteum}, and 3) \emph{Q.
@@
 Across these eight species, specific leaf area had a positive association with
-maximum water absorption (Table~\ref{tab:mc_di_anova}, SI Fig~5, linear model
+maximum water absorption (Table~\ref{tab:mc_di_anova}, SI Fig~4, linear model
 $P = 0.003$) and litter bulk density had a negative effect on drying rate
-(Table~\ref{tab:mc_di_anova}, SI Fig~6, linear model $P = 0.029$).
+(Table~\ref{tab:mc_di_anova}, SI Fig~5, linear model $P = 0.029$). Past work
+has demonstrated that green leaf length predicts litter packing and,
+consequently, flame spread rate across these species. Litter particle size
+distributions, however show that there is substantial variation in particle
+sizes within a species (SI Fig.~6).
 
@@
 \centering
-%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
-
 \begin{tabular}{ccrrrrr}
@@
   \midrule
-
 Drying rate & density & 1 & $3.92 \times 10^{-4}$ & $3.92 \times 10^{-4}$ & 11.01 & 0.029 \\ 
@@
   \bottomrule
-
 \end{tabular}
@@
 
-\subsection*{Flammability across taxonomic groups}
+\subsection*{Flammability}
 
-Flame spread rate declined with fuel moisture (Fig.~\ref{fig:flam_moist}).
-\emph{Quercus} had higher spread rates across all fuel moistures and had a
-steeper decline in response to moisture than did the other groups (taxon by
-time since wetting interaction $P = 0.022$, Table~\ref{tab:spreadrate_anova}).
-Fuel moisture negatively influenced fuel consumption. Fuel consumption and the
-rate of change with moisture content varied across the taxonomic groups
-(Fig.~\ref{fig:flam_moist}, Table~\ref{tab:consume_anova}). Although the
-overall ranking of groups was similar for fuel consumption and for flame spread
-rate, there were different pattern across taxa in these two flammability
-parameters. For spread rate, the decline in flammability with moisture was
+Flame spread rate declined with fuel moisture (Fig.~\ref{fig:flam_moist}) and
+maximum rates and the steepness of decline differed by taxonomic group
+(Table~\ref{tab:spreadrate_anova}). Similarly, fuel moisture negatively
+influenced fuel consumption (Fig.~\ref{fig:flam_moist},
+Table~\ref{tab:consume_anova}). Although the overall ranking of groups was
+similar for fuel consumption and for flame spread rate, there were different
+pattern across taxa in these two flammability parameters. For spread rate, the
+decline in flammability with moisture was steepest for \emph{Quercus} and
 shallowest for \emph{Pinus} and Cupressaceae (Fig.~\ref{fig:flam_moist}). For
@@
 \includegraphics[width=8cm]{figs_tables/fig2_moisture_flam.pdf}
-\caption{Fuel moisture effects on flammability. A) Flame spread rate by moisture content (model fit $F = 61.15$, df = 9,
-  148, $P < 0.0001$). Model ANOVA table in Table \ref{tab:spreadrate_anova} and
-  coefficients in SI Table 4. B) Percent fuel consumed by moisture content. Lines indicate best fit
-  beta regression lines. Model deviance table in Table \ref{tab:consume_anova}
-  and coefficients are in SI Table 5.}
+\caption{Fuel moisture effects on flammability. A) Flame spread rate by
+  moisture content (model fit $F = 61.15$, df = 9, 148, $P < 0.0001$). Model
+  ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI Table
+  4. B) Percent fuel consumed by moisture content. Lines indicate best fit beta
+  regression lines. Model deviance table in Table \ref{tab:consume_anova} and
+  coefficients are in SI Table 5.}
   \label{fig:flam_moist}
@@
 
-\subsection*{Nonadditivity in  litter mixtures}
+\subsection*{Nonadditivity in moisture dynamics and flammability}
 
-Based on the estimated marginal
-means for maximum water absorption and drying rate (SI Figs 3--4), we selected
-four species for use in
-the litter mixtures and examined all four three-species combinations of those
-four. The four species were \emph{Abies concolor}, \emph{Calocedrus decurrens},
-\emph{Pinus jeffreyi} and \emph{Quercus kelloggii}.
+Based on the estimated marginal means for maximum water absorption and drying
+rate (SI Figs 2--3), we selected four species for use in the litter mixtures
+and examined all three-species combinations of those four. The four species
+were \emph{Abies concolor}, \emph{Calocedrus decurrens}, \emph{Pinus jeffreyi}
+and \emph{Quercus kelloggii}.
 
@@
 
-There was evidence for nonadditive effects in drying rate. Mixtures tended to
-have lower water content than predicted from the mean of the constituent
-species after wetting, but had higher water content than expected after long
-drying (Wald test shows slope of mixed linear model
-[Fig.~\ref{fig:mixture_obs_pred_mc}] less than 1, $P < 0.0001$). 
-Mixture drying rates and maximum water absorption showed some relationship to
-the dry down curves for individual constituent species (see SI Fig.~7). For
-example, the mixture that does not contain \emph{Q. kelloggii} (AbCaPi in SI
-Fig.~7) has the lowest maximum water absorption. Maximum water absorption and
-drying rates differed across mixtures (linear model based on log of moisture
-content, SI Tables~8 and 9).
+Drying rate exhibited nonadditive effects. Mixtures tended to have lower water
+content than predicted from the mean of the constituent species after wetting,
+but had higher water content than expected after long drying (Wald test shows
+slope of mixed linear model less than 1, $P < 0.0001$,
+Fig.~\ref{fig:mixture_obs_pred_mc}). Mixture drying rates and maximum water
+absorption showed some relationship to the dry down curves for individual
+constituent species (see SI Fig.~7). For example, the mixture that does not
+contain \emph{Q. kelloggii} (AbCaPi in SI Fig.~7) has the lowest maximum water
+absorption. Maximum water absorption and drying rates differed across mixtures
+(linear model based on log of moisture content, SI Tables~8 and 9).
 
@@
 The taxa in this study displayed different patterns of both moisture absorption
-and desorption, associated with different litter traits. In this sample of
-species, those that have a greater absorptive capacity are also among the most
-flammable in terms of spread rate when dry. Traits such as large leaves and
-high SLA may affect both absorptive capacity and flammability with a
-consequence that flammability rankings across species can change with time
-since wetting. One shortcoming of past work in this system was that
-flammability trials tested only dry fuels \citep{Magalhaes+Schwilk-2012}. We
-therefore expected that rankings of species flammability might change under
-higher moisture levels. For example, one presumption of ours prior to this
-experiment was that the high flammability previously reported for \emph{Q.
-  kelloggii} \citep{Magalhaes+Schwilk-2012} would disappear under higher fuel
-moisture contents. This prediction was partly supported: \emph{Q. kelloggii}
-fails to ignite until after 72 hours of drying in this study (Figs.
-\ref{fig:drydown} and \ref{fig:drydown}). However, high moisture
-content in this oak was unable to suppress ignition of mixtures to which it
-contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
-\ref{fig:mixture_obs_pred_consume}), indicating the importance of non-additive effects.
+and desorption, associated with different litter traits. Those species that had
+a greater absorptive capacity were also among the most flammable in terms of
+spread rate when dry. Traits such as large leaves and high SLA may affect both
+absorptive capacity and flammability with a consequence that flammability
+rankings across species can change with time since wetting. One shortcoming of
+past work in this system was that flammability trials tested only dry fuels
+\citep{Magalhaes+Schwilk-2012}. We therefore expected that rankings of species
+flammability might change under higher moisture levels. For example, one
+presumption of ours prior to this experiment was that the high flammability
+previously reported for \emph{Q. kelloggii} \citep{Magalhaes+Schwilk-2012}
+would lessen significantly under higher fuel moisture. This prediction
+was partly supported: \emph{Q. kelloggii} fails to ignite until after 72 hours
+of drying in this study (Figs. \ref{fig:drydown} and \ref{fig:drydown}).
+However, high moisture content in this oak was unable to suppress ignition of
+mixtures to which it contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
+\ref{fig:mixture_obs_pred_consume}), indicating the importance of non-additive
+effects.
 
@@
 consequence of both differential initial absorption and differences in drying
-rate. This leads to flammability varying both by taxon and time since wetting.
-In these mixed conifer forests, fine fuels can vary in fuel moisture throughout
-the fire season. For example, 10-hour fuels in northern California varied from
-less than 10\% to over 80\% depending on precipitation events in a study
-following fuel moisture throughout the fire season
+rate. This leads to local litter flammability varying both by taxon and time
+since wetting. In these mixed conifer forests, fine fuels can vary in fuel
+moisture throughout the fire season. For example, 10-hour fuels in northern
+California varied from less than 10\% to over 80\% depending on precipitation
+events in a study following fuel moisture throughout the fire season
 \citep{Estes+Knapp+etal-2012}. Early summer prescribed burns at our study sites
 in Sequoia National Park show evidence of moisture limiting fire behavior at
-12-14\% fuel moisture (Knapp et al 2005).
+12-14\% fuel moisture \citep{Knapp+Keeley+etal-2005}.
 
@@
 than the other taxa. Our two Cupressaceae, which have dense litter but have low
-maximum water absorption, had low flammability across moisture levels. This
+maximum water absorption, had low flammability across all moisture levels. This
 result is consistent with the low flammability found in an Australian
-Cupressaceae that have small litter particles and dense packing
-\citep{Scarff+Westoby-2006, Bowman+Haverkamp+etal-2018}. However, the densely
+Cupressaceae that have small litter particles that pack densely
+\citep{Scarff+Westoby-2006, Bowman+Haverkamp+etal-2018}. However, densely
 packed litter is not a consistent trait across the family because
@@
 axis of flammability variation in dry litter \citep{Magalhaes+Schwilk-2012}.
-Our results suggest that moisture dynamics are controlled by both traits
-intrinsic to the leaf particles (specific leaf area) and an emergent property
-of the litter bed (bulk density). Specific leaf area here is probably capturing
-some joint effect of cuticle thickness that may influence water absorption and
-total surface area of litter particles which increases water adsorption. The
-oak has much higher maximum water absorption than do the other species,
-followed by the finest-needled pine, \emph{P. lambertiana}. We expect that
-decomposition will affect these dynamics. For example, Van Wagner
-\citeyear{Van_Wagner-1969}, investigating fuel weathering, found that pine
-needles had a variation in moisture content 3 to 6 times greater when they
-lacked their protective wax coating when compared to intact needles.
+Our results suggest that moisture dynamics are controlled both by a trait
+intrinsic to the leaf particles (specific leaf area) and by an emergent
+property of the litter bed (bulk density). Specific leaf area here is probably
+capturing some joint effect of cuticle thickness, which may influence water
+absorption, and total surface area of litter particles, which increases water
+adsorption. The oak has much higher maximum water absorption than do the other
+species, followed by the finest-needled pine, \emph{P. lambertiana}.
+
+% We expect that decomposition will affect these dynamics. For example, Van
+% Wagner \citeyear{Van_Wagner-1969}, investigating fuel weathering, found that
+% pine needles had a variation in moisture content 3 to 6 times greater when
+% they lacked their protective wax coating when compared to intact needles.
 
@@
 mixtures. These mixtures contained species representative of all dry-down
-behavior types observed in single species trials. Drying rates, however, did
-not behave as the average of the constituent species: mixtures tended to be
-drier than expected early in the process and wetter than expected later (Fig.
+behavior types observed in single species trials. Drying rates did not behave
+as the average of the constituent species: mixtures tended to be drier than
+expected early in the process and wetter than expected later (Fig.
 \ref{fig:mixture_obs_pred_mc}). Past work has shown that mixtures tend to pack
@@
 by species but also by wind and storm events the previous season. With only
-four distinct litter combinations, it is impossible to attribute cause to these
-variations within the overall statistical trend, but litter packing alone is
-complex and needles and twigs can combine such that the addition of a species
-to a mixture can create a denser litter bed or less dense litter bed than is
-expected \citep{Zhao+vanLogtestijn+etal-2019}.
+four distinct litter combinations, it is impossible to attribute mechanism to
+the trend. Litter packing alone is a complex trait: needles and twigs can
+combine such that the addition of a species to a mixture can create a denser
+litter bed or less dense litter bed than is expected
+\citep{Zhao+vanLogtestijn+etal-2019}.
 
 We draw two main conclusions from this work. First, maximum water absorption is
-highest in those species with the most flammability litter under dry conditions.
-As a consequence, under higher fuel moisture, the flammability ranking of
-these taxa can change compared to the flammability of dry fuels. Second, litter
-flammability is prone to nonadditive effects which become more positive with
-drying. The relationships we described here can form a working hypothesis and
-baseline expectation to adjust fire behavior predictions in real communities:
-in mixtures with fuel moisture content below 50\%, observed flame spread rates
-approached double the predicted value based on monoculture trials (Fig.
+highest in those species with the most flammability litter under dry conditions
+(higher SLA, larger leaf area, less dense litter). As a consequence, under
+higher fuel moisture, the flammability ranking of these taxa can change
+compared to the flammability of dry fuels (for example, crossing lines in Fig.
+\ref{fig:flam_moist} B and C). Second, litter flammability is prone to
+nonadditive effects which become more positive with drying. The relationships
+we described here can form a working hypothesis and baseline expectation to
+adjust fire behavior predictions in real communities: in mixtures with fuel
+moisture content below 50\%, observed flame spread rates approached double the
+predicted value based on monoculture trials (Fig.
 \ref{fig:mixture_obs_pred_spread}). In completely dry litter from this same
-system, past work demonstrates observed spread rates approached three times the
-expectation \citep{Magalhaes+Schwilk-2012}. These numbers provide a starting
-point for parameterizing fire models with mixed fuels. In this temperate mixed
-conifer forest, these effects are likely to include positive feedbacks whereby
-the shade intolerant species such as the long-needled pines and the
-large-leaved oak, are favored by the increased fire to which their fuels
-contribute \citep{Schwilk+Caprio-2011} and similar results are reported for
-southeastern US forests where species composition shifts have led to decreases
-in fire activity \citep{Nowacki+Abrams-2008}. Vegetation-fire feedbacks are
-important at a wide range of scales from stand-level species composition shifts
-to the earth system \citep{Harris+Remenyi+etal-2016,
-  Archibald+Lehmann+etal-2018}. The patterns we demonstrate here show the
-potential importance of species traits and interactions at local and stand
-scales where management and fire operations meet species' biology.
+system, observed flame spread rates in mixtures approached three times the
+expectation based on the average of constituent species
+\citep{Magalhaes+Schwilk-2012}. These numbers provide a starting point for
+parameterizing fire models with mixed fuels. In this temperate mixed conifer
+forest, these effects are likely to include positive feedbacks whereby the
+shade intolerant species such as the long-needled pines and the large-leaved
+oak, are favored by the increased fire spread to which their fuels contribute
+\citep{Schwilk+Caprio-2011}. Similar results are reported for southeastern US
+forests where species composition shifts have led to decreases in fire activity
+\citep{Nowacki+Abrams-2008}. Vegetation-fire feedbacks are important at a wide
+range of scales from stand-level species composition shifts to the earth system
+\citep{Harris+Remenyi+etal-2016, Archibald+Lehmann+etal-2018}. The patterns we
+demonstrate here show the importance of species traits and interactions at
+local and stand scales where management and fire operations meet species'
+biology.
 

### 2020-05-01 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
fixup
Update number of words

@@
 \documentclass[letterpaper,12pt]{article}
+\usepackage{jecol}
+\usepackage[notablist,nofiglist]{endfloat}
 \usepackage{caption}
 \DeclareCaptionLabelFormat{adja-page}{\hrulefill\\#1 #2 \emph{(previous page)}}
-\usepackage{jecol}
-\usepackage[notablist,nofiglist]{endfloat}
-%\usepackage{color}
-%\usepackage{soul}
-%\renewcommand{\hl}[1]{#1}
 
@@
 
-\nwords{8200}
+\nwords{6330 (main text only), 8000 (abstract, text, captions) }
 \ntables{4}
@@
 collected from over 10 different individuals. For the remaining species, the
-collection involved 2--4 individual trees at each site which were separated
+collection involved 2 -- 4 individual trees at each site which were separated
 from one another by at least 10\,m. The litter was collected 2\,m away from the
@@
 because bark and heavier branches tend to fall closer to the tree trunk. The
-top 2.5-5\,cm of litter were collected by raking and shoveling into large bags.
+top 2.5 -- 5\,cm of litter were collected by raking and shoveling into large bags.
 All litter was oven dried at $60^\circ$C to $< 5$\% fuel moisture by dry
@@
 changes to the particle size distributions. The moisture manipulations and
-subsequent burning trials were conducted in April--August 2015.
+subsequent burning trials were conducted in April -- August 2015.
 
@@
 We performed flammability tests across multiple moisture levels on monospecific
-litter beds and on mixed litter beds with each species or mixture replicated 5
-times at each moisture level. We used the initial dry-down experiment to guide
-our timing so as to attempt to burn samples over a range of moisture contents.
-We staggered saturation and start of drying so we could conduct our burning in
-a blocked design with multiple species at the same moisture content per burning
-day. We aimed for moisture contents from 10\% to 100\% on a dry mass basis because
-preliminary trials on damp fuels suggested that across these fuels ignition
-will not occur above 100\% fuel moisture. To expand our data on the dry end of
-this gradient, we supplemented these data with data from an earlier experiment
-on oven-dried litter \citep{Magalhaes+Schwilk-2012}. This supplemental data
-added five replicates per species at an oven dried moisture content of $<$ 5\%.
-We assigned these replicates a time since wetting value of 144 hours equivalent
-to the longest time since drying for the wetted samples.
+litter beds and on mixed litter beds with each species or mixture replicated
+five times at each moisture level. We used the initial dry-down experiment to
+guide our timing so as to attempt to burn samples over a range of moisture
+contents. We staggered saturation and start of drying so we could conduct our
+burning in a blocked design with multiple species at the same moisture content
+per burning day. We aimed for moisture contents from 10\% to 100\% on a dry
+mass basis because preliminary trials on damp fuels suggested that across these
+fuels ignition will not occur above 100\% fuel moisture. To expand our data on
+the dry end of this gradient, we supplemented these data with data from an
+earlier experiment on oven-dried litter \citep{Magalhaes+Schwilk-2012}. This
+supplemental data added five replicates per species at an oven dried moisture
+content of $<$ 5\%. We assigned these replicates a time since wetting value of
+144 hours equivalent to the longest time since drying for the wetted samples.
 
@@
     statistics and p-values were calculated by the Kenward-Roger approximation.
-    \citep{Kenward_Roger-1997}. Model coefficients are reported in SI Table~1.}
-  \label{tab:drydown}
+    \citep{Kenward_Roger-1997}. Model coefficients are reported in SI Table~1.
+  \label{tab:drydown}}
 \centering
@@
 
-\subsection*{Moisture dynamics and flammability across taxonomic groups}
+\subsection*{Moisture dynamics and flammability}
 
@@
 coding and data analysis, DWS and RMQM; writing DWS and RMQM; writing--review
-and editing, DWS
+and editing, DWS.
 

### 2020-05-01 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
fixup

@@
 \nfig{5}
-\nref{55}
+\nref{57}
 
@@
     differing effects on flammability across the major litter types typified by
-    different tree taxa..
+    different tree taxa.
 

### 2020-05-01 scripts/moisture-ms-figs-tables.R
Slight edits to figs 1 and 2
Make fig 1 less tall so it will fit on page in one column.

@@ -1,6 +1,13 @@
 # Figures for moisture manuscript
 # 2019-12-12
 
+## install.packages("R2admb")
+## install.packages("glmmADMB", 
+##     repos=c("http://glmmadmb.r-forge.r-project.org/repos",
+##             getOption("repos")),
+##     type="source")
+
+
 library(xtable)
 library(ggrepel)
 library(glmmADMB)
@@ -44,7 +51,6 @@ summarize(x, n=length(l))
 
 
 
-
 ###############################################################################
 ## Figure 1a: dry-down
 ###############################################################################
@@ -64,7 +70,9 @@ fig1a <- ggplot(mc, aes(hour, MC_dry, group=display.name, color=taxon)) +
   pubtheme.nogridlines +
   theme(legend.position=c(0.75, 0.86),
         legend.title=element_blank(),
-        legend.text = element_text(family=fontfamily, size=smsize, face="italic"))
+        legend.text = element_text(family=fontfamily, size=smsize, face="italic"),
+        plot.margin = unit(c(0, 5.5, 3, 5.5), "pt"),
+        axis.title.x = element_blank()
 #       legend.key.height=unit(smsize,"pt"))
 
 ## ggsave(file.path(RESULTS, "fig1a_drydown-curves.pdf"), plot=fig1a,
@@ -274,13 +282,15 @@ b.scaled <- b %>% mutate_if(is.numeric, scale)
 fig2a <- ggplot(b, aes(actualMC_dry, spread, color=taxon)) +
   geom_point(size=1.5, alpha=0.7, stroke=0) +
   scale_colour_manual(values=schwilkcolors) +
-  xlab("Moisture content (%)") + ylab("Spread rate (cm/s)") +
+  ylab("Spread rate (cm/s)") +
   pubtheme.nogridlines +
   stat_smooth(method="lm", se=FALSE, size=1, drop=TRUE) +
   theme(legend.position=c(.76, .86),
         legend.spacing.y=unit(0,"cm"),
         legend.text = element_text(family=fontfamily, size=smsize-1, face="italic"),
-        legend.title=element_blank())
+        legend.title=element_blank(),
+        plot.margin = unit(c(0, 5.5, 3, 5.5), "pt"),
+        axis.title.x = element_blank())
[…]

### 2020-05-01 scripts/theme-opts.R

@@ -4,8 +4,8 @@ library(ggplot2)
 library(gridExtra)
 library(scales)
 library(extrafont)
-#font_import(pattern="Arial")
-loadfonts()
+font_import(pattern="Arial")
+#loadfonts()
 
 # constants and theme for publications
 #schwilkcolors <- c("#D68D18", "#836B43", "#A0AE6A", "#437683", "#18B0D6")

### 2020-05-22 scripts/decomp-flam.R
Combine figures for R3
Combined figures 1 and 2 and 4 and 4 as requested in reviews on R2

@@ -18,7 +18,7 @@ library(AICcmodavg)
 ## any real reason for selecting columns there?
 # New dataframe with the predicted spread rate after decomposition (year 1)
 decomp.by.species <- decomp.sum %>%  ungroup() %>%
-  select(spcode, year, l_mean, w_mean, t_mean, larea_mean, lt_mean) %>%
+  dplyr::select(spcode, year, l_mean, w_mean, t_mean, larea_mean, lt_mean) %>%
   group_by(spcode, year) %>% summarise(l_mean_sd = sd(l_mean, na.rm=TRUE),
                                  l_mean = mean(l_mean),
                                  t_mean_sd = sd(t_mean, na.rm=TRUE),
@@ -35,7 +35,7 @@ decomp.by.species <- decomp.sum %>%  ungroup() %>%
 ## Year 0 decomp by flam summary for figure 2
 flam.by.species <- flam.sp.avg %>%
   mutate(spread_lc = spread.mean-2*spread.sd, spread_uc = spread.mean+2*spread.sd) %>%
-  select(spcode, spread_mean=spread.mean, spread_lc, spread_uc)
+  dplyr::select(spcode, spread_mean=spread.mean, spread_lc, spread_uc)
 
 flamdecomp <- decomp.by.species %>% filter(year == 0) %>% left_join(flam.by.species)
 ## decomp.Y0.conifers <- filter(decomp.Y0, spcode %in% 
@@ -74,7 +74,7 @@ pred.y1 <- decomp.by.species %>% filter(year == 1)
 #pred.ci <- predict(lmfit2, newdata=pred.y1, interval="confidence", level=.95, se.fit=TRUE)
 pred.pi <- predict(lmfit2, newdata=pred.y1, interval="confidence", #"prediction",
                    level=.95, se.fit=TRUE)$fit %>% as.data.frame() %>%
-            select(spread_mean = fit, spread_lc = lwr, spread_uc = upr)
+            dplyr::select(spread_mean = fit, spread_lc = lwr, spread_uc = upr)
 
 pred.y1 <- bind_cols(pred.y1, pred.pi)
 

### 2020-05-22 scripts/moisture-ms-figs-tables.R

@@ -64,16 +64,16 @@ fig1a <- ggplot(mc, aes(hour, MC_dry, group=display.name, color=taxon)) +
   scale_colour_manual(values=schwilkcolors) +
   #xlab("Time since wetting (hr)") +
   xlab("") +
-  ylab("Moisture by dry weight (%)") +
+  ylab("Fuel moisture (%)") +
   scale_x_continuous(breaks=xbreaks) +
   scale_y_continuous(breaks=ybreaks) +
   pubtheme.nogridlines +
-  theme(legend.position=c(0.75, 0.86),
+  theme(legend.position=c(0.79, 0.86),
         legend.title=element_blank(),
-        legend.text = element_text(family=fontfamily, size=smsize, face="italic"),
-        plot.margin = unit(c(0, 5.5, 3, 5.5), "pt"),
-        axis.title.x = element_blank()
-#       legend.key.height=unit(smsize,"pt"))
+        legend.text = element_text(family=fontfamily, size=smsize-1, face="italic"),
+        plot.margin = unit(c(0, 3, 3, 5.5), "pt"),
+        axis.title.x = element_blank())
+#        legend.key.height=unit(smsize,"pt"))
 
 ## ggsave(file.path(RESULTS, "fig1a_drydown-curves.pdf"), plot=fig1a,
 ##        width=col1, height=col1, units="cm")
@@ -277,24 +277,25 @@ b.scaled <- b %>% mutate_if(is.numeric, scale)
 
 
 ###############################################################################
-## Fig 2: Spread rate and moisture content
+## Fig 1d: Spread rate and moisture content
 ###############################################################################
-fig2a <- ggplot(b, aes(actualMC_dry, spread, color=taxon)) +
+fig1d <- ggplot(b, aes(actualMC_dry, spread, color=taxon)) +
   geom_point(size=1.5, alpha=0.7, stroke=0) +
   scale_colour_manual(values=schwilkcolors) +
   ylab("Spread rate (cm/s)") +
   pubtheme.nogridlines +
   stat_smooth(method="lm", se=FALSE, size=1, drop=TRUE) +
-  theme(legend.position=c(.76, .86),
+  theme(legend.position=c(.79, .86),
         legend.spacing.y=unit(0,"cm"),
         legend.text = element_text(family=fontfamily, size=smsize-1, face="italic"),
[…]

### 2020-05-22 scripts/theme-opts.R

@@ -4,8 +4,8 @@ library(ggplot2)
 library(gridExtra)
 library(scales)
 library(extrafont)
-font_import(pattern="Arial")
-#loadfonts()
+#font_import(pattern="Arial")
+loadfonts()
 
 # constants and theme for publications
 #schwilkcolors <- c("#D68D18", "#836B43", "#A0AE6A", "#437683", "#18B0D6")

### 2020-05-22 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Add comment with "tweetable abstract ?!"
Edits to MS in response to reviews on R2

@@
 \ntables{4}
-\nfig{5}
+\nfig{3}
 \nref{57}
@@
 \maketitle
+
+% Tweetable:
+
+% Moisture dynamics changes flammability rankings. Litter mixtures nonadditivel and this effect increases as fuels dry.
+
+% 50 word summary
+
+% This study examined ecologically relevant drying rates and their
+% interaction with on litter flammability. We demonstrated that fuel
+% moisture changes flammability and interactions in mixed
+% communities: litter mixtures generally burn with fire behavior
+% more similar to that of the most flammable constituent species
+% and this effect increases as fuels dry.
 
@@
       \item \emph{Synthesis:} This study is the first to examine ecologically
-    relevant drying rates and their interaction with non additive misture
+    relevant drying rates and their interaction with nonadditive mixture
     effects on flammability. We find that taxon specific patterns of moisture
@@
 flammability of moist fuels. There is some evidence for nonadditive
-flammability in moist fuels with a single study finding increasing nonadditivity
-with moisture \citep{Blauw+Wensink+etal-2015}, but this previous work did not
-include estimation of the dry-down process and had an unclear relationship to
-fire behavior in the field. In mixed conifer forest in California, where the
-fire regime is characterized by surface fires fueled by plant litter, overstory
-tree composition drives variation in local fire behavior
-\citep{Schwilk+Caprio-2011} as a result of differential litter flammability
-across tree species and as a result of synergistic interactions in
+flammability in moist fuels with a single study finding increasing
+nonadditivity with moisture \citep{Blauw+Wensink+etal-2015}, but this previous
+work did not include estimation of mixture effect on the dry-down process and
+had an unclear relationship to fire behavior in the field. In mixed conifer
+forest in California, where the fire regime is characterized by surface fires
+fueled by plant litter, overstory tree composition drives variation in local
+fire behavior \citep{Schwilk+Caprio-2011} as a result of differential litter
+flammability across tree species and as a result of synergistic interactions in
 multi-species litter mixtures \citep{Magalhaes+Schwilk-2012}. These
@@
 This study is the first to examine ecologically relevant drying rates and their
-interaction with non additive moisture effects on flammability. We asked if
+interaction with nonadditive moisture effects on flammability. We asked if
 differences in litter moisture absorption and drying rates change the
@@
 Across these species there were significant differences in the maximum water
-absorption and drying rate (all $P < 0.0001$, Fig.~\ref{fig:drydown},
+absorption and drying rate (all $P < 0.0001$, Fig.~\ref{fig:drydown}A,
 Table~\ref{tab:drydown} and SI Figs.~1 -- 3). Calculating estimated marginal
@@
   \centering
-  \includegraphics[width=8cm]{figs_tables/fig1_dry_effects_by_taxon.pdf}
+  \includegraphics[width=16cm]{figs_tables/fig1_dry_effects_by_taxon.pdf}
   \caption{}
@@
   indicate best fit beta regression lines by taxonomic group. Model ANOVA table
-  in Table \ref{tab:consume_anova} and coefficients in SI Table 7. For all
-  plots, horizontal jitter (0.5 to 2 hr) was added to avoid overplotting. A
+  in Table \ref{tab:consume_anova} and coefficients in SI Table 7.
+  D) Flame spread rate by moisture content (model fit $F = 61.15$, df = 9, 148,
+  $P < 0.0001$). Model ANOVA table in Table \ref{tab:spreadrate_anova} and
+  coefficients in SI Table 4. E) Percent fuel consumed by moisture content.
+  Lines indicate best fit beta regression lines. Model deviance table in Table
+  \ref{tab:consume_anova} and coefficients are in SI Table 5. For plots in left
+  column, horizontal jitter (0.5 to 2 hr) was added to avoid overplotting. A
   lack of data at earlier time points for a taxon indicates no ignitions
-  occurred at those time points. }
+  occurred at those time points.}
  \label{fig:drydown}
@@
 has demonstrated that green leaf length predicts litter packing and,
-consequently, flame spread rate across these species. Litter particle size
-distributions, however show that there is substantial variation in particle
-sizes within a species (SI Fig.~6).
+consequently, flame spread rate across these species
+\citep{Magalhaes+Schwilk-2012}. Litter particle size distributions, however
+show that there is substantial variation in particle sizes within a species (SI
+Fig.~6).
 
@@
 
-Flame spread rate declined with fuel moisture (Fig.~\ref{fig:flam_moist}) and
+Flame spread rate declined with fuel moisture (Fig.~\ref{fig:drydown}D) and
 maximum rates and the steepness of decline differed by taxonomic group
 (Table~\ref{tab:spreadrate_anova}). Similarly, fuel moisture negatively
-influenced fuel consumption (Fig.~\ref{fig:flam_moist},
+influenced fuel consumption (Fig.~\ref{fig:drydown}E,
 Table~\ref{tab:consume_anova}). Although the overall ranking of groups was
@@
 decline in flammability with moisture was steepest for \emph{Quercus} and
-shallowest for \emph{Pinus} and Cupressaceae (Fig.~\ref{fig:flam_moist}). For
+shallowest for \emph{Pinus} and Cupressaceae (Fig.~\ref{fig:drydown}D). For
 fuel consumption, \emph{Pinus} and \emph{Abies} has a steeper decline in
 flammability with fuel moisture than did the other groups
-(Fig.~\ref{fig:flam_moist}).
-
-%% spread rate by moisture
-\begin{figure}
-  \centering
-\includegraphics[width=8cm]{figs_tables/fig2_moisture_flam.pdf}
-\caption{Fuel moisture effects on flammability. A) Flame spread rate by
-  moisture content (model fit $F = 61.15$, df = 9, 148, $P < 0.0001$). Model
-  ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI Table
-  4. B) Percent fuel consumed by moisture content. Lines indicate best fit beta
-  regression lines. Model deviance table in Table \ref{tab:consume_anova} and
-  coefficients are in SI Table 5.}
-  \label{fig:flam_moist}
-\end{figure}
+(Fig.~\ref{fig:drydown}E).
 
@@
 spread rates were slow. \emph{Quercus} had the steepest response to time since
-wetting (Fig.~\ref{fig:drydown}, SI Table~6)).
+wetting (Fig.~\ref{fig:drydown}B, SI Table~6)).
 
@@
 less common than were very low or moderately high consumption
-(Fig.~\ref{fig:drydown}). All parameters significantly contributed to the
+(Fig.~\ref{fig:drydown}C and E). All parameters significantly contributed to the
 generalized linear mixed beta regression model as indicated by the nested model
@@
   \centering
-\includegraphics[width=16cm]{figs_tables/fig3_mixture_obs_vs_pred_mc.pdf}
+\includegraphics[width=16cm]{figs_tables/fig2_mixture_obs_vs_pred_mc.pdf}
 \caption{Observed vs predicted fuel moisture content for litter mixtures.
@@
   \centering
-\includegraphics[width=16cm]{figs_tables/fig4_mixture_obs_vs_pred_spread.pdf}
-\caption{Observed vs. predicted flame spread rate for litter mixtures. Filled
-  circles show observed spread rate values. Open diamonds indicate predicted
-  spread rates based on the average spread rate at each time since wetting for
-  the three constituent species in each mixture. Codes for species in mixtures:
-  Ab = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi =
-  \emph{Pinus jeffreyi} and Qu = \emph{Quercus kellogii}.}
-  \label{fig:mixture_obs_pred_spread}
-\end{figure}
-
-\begin{figure}
-  \centering
-\includegraphics[width=16cm]{figs_tables/fig5_mixture_obs_vs_pred_consume.pdf}
-\caption{Observed vs predicted fuel consumption for litter mixtures. Filled
-  circles show observed fuel consumption percentages. Open diamonds indicate
-  predicted fuel consumption based on the average fuel consumption at each
-  particular time since wetting for the three constituent species in each
-  mixture. Codes for species in mixtures: Ab = \emph{Abies concolor}, Ca =
-  \emph{Calocedrus decurrens}, Pi = \emph{Pinus jeffreyi} and Qu =
-  \emph{Quercus kellogii}.}
-  \label{fig:mixture_obs_pred_consume}
+\includegraphics[width=16cm]{figs_tables/fig3_mixture_obs_vs_pred_flam.pdf}
+\caption{Observed vs. predicted flammability for litter mixtures. Filled
+  circles show observed values. Open diamonds indicate predicted values based
+  on the average spread or fuel consumption at each time since wetting for the
+  three constituent species in each mixture. Codes for species in mixtures: Ab
+  = \emph{Abies concolor}, Ca = \emph{Calocedrus decurrens}, Pi = \emph{Pinus
+    jeffreyi} and Qu = \emph{Quercus kellogii}.}
+  \label{fig:mixture_obs_pred_flam}
 \end{figure}
@@
 rates were generally higher than predicted (Fig.
-\ref{fig:mixture_obs_pred_spread}, Wald test of mixed linear model centered
+\ref{fig:mixture_obs_pred_flam}A, Wald test of mixed linear model centered
 intercept for observed-predicted values greater than zero, $P = 0.017$) and the
@@
 positive or negative nonadditivity, but nonadditivity increased with time since
-wetting (Fig. \ref{fig:mixture_obs_pred_consume}, slope not equal to zero,
+wetting (Fig. \ref{fig:mixture_obs_pred_flam}B, slope not equal to zero,
 $P < 0.0001$). Fuel consumption tended to show negative nonadditivity at higher
 fuel moistures and positive nonadditivity following drying
-(Fig.~\ref{fig:mixture_obs_pred_consume}).
+(Fig.~\ref{fig:mixture_obs_pred_flam}B).
 
@@
 previously reported for \emph{Q. kelloggii} \citep{Magalhaes+Schwilk-2012}
-would lessen significantly under higher fuel moisture. This prediction
-was partly supported: \emph{Q. kelloggii} fails to ignite until after 72 hours
-of drying in this study (Figs. \ref{fig:drydown} and \ref{fig:drydown}).
-However, high moisture content in this oak was unable to suppress ignition of
-mixtures to which it contributed (Figs. \ref{fig:mixture_obs_pred_spread} and
-\ref{fig:mixture_obs_pred_consume}), indicating the importance of non-additive
-effects.
+would lessen significantly under higher fuel moisture. This prediction was
+partly supported: \emph{Q. kelloggii} fails to ignite until after 72 hours of
+drying in this study (Figs. \ref{fig:drydown} and \ref{fig:drydown}). However,
+high moisture content in this oak was unable to suppress ignition of mixtures
+to which it contributed (Fig. \ref{fig:mixture_obs_pred_flam}), indicating the
+importance of nonadditive effects.
 
@@
 across mixtures showed neutral, positive, or negative nonadditivity (Fig.
-\ref{fig:mixture_obs_pred_consume}). Previous studies have demonstrated the
+\ref{fig:mixture_obs_pred_flam}B). Previous studies have demonstrated the
 potential for the direction of nonadditivity to shift with fuel moisture
@@
 compared to the flammability of dry fuels (for example, crossing lines in Fig.
-\ref{fig:flam_moist} B and C). Second, litter flammability is prone to
+\ref{fig:drydown} D and E). Second, litter flammability is prone to
 nonadditive effects which become more positive with drying. The relationships
@@
 predicted value based on monoculture trials (Fig.
-\ref{fig:mixture_obs_pred_spread}). In completely dry litter from this same
+\ref{fig:mixture_obs_pred_flam}A). In completely dry litter from this same
 system, observed flame spread rates in mixtures approached three times the
@@
 All data and code for this project is at
-\url{https://github.com/schwilklab/trait-flam}. DOI to be provided. For review,
-use ``joe'' branch which contains this manuscript and code needed to create the
-figures and tables.
+\url{https://github.com/schwilklab/trait-flam}. Checkout git tag ``JOE.R3''
+which contains this manuscript and code needed to create the figures and
+tables.
 

### 2020-05-25 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Fix caption

@@
     fit to moisture content on a dry mass basis. B) Flame spread rate by time
-  since wetting. Lines indicate best fit linear model fits by taxonomic group.
-  Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in SI
-  Table 6. C) Percent fuel consumed by time since wetting. Curves shown
-  indicate best fit beta regression lines by taxonomic group. Model ANOVA table
-  in Table \ref{tab:consume_anova} and coefficients in SI Table 7.
-  D) Flame spread rate by moisture content (model fit $F = 61.15$, df = 9, 148,
-  $P < 0.0001$). Model ANOVA table in Table \ref{tab:spreadrate_anova} and
-  coefficients in SI Table 4. E) Percent fuel consumed by moisture content.
-  Lines indicate best fit beta regression lines. Model deviance table in Table
-  \ref{tab:consume_anova} and coefficients are in SI Table 5. For plots in left
-  column, horizontal jitter (0.5 to 2 hr) was added to avoid overplotting. A
-  lack of data at earlier time points for a taxon indicates no ignitions
-  occurred at those time points.}
+    since wetting. Lines indicate best fit linear model by taxonomic group.
+    Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in
+    SI Table 6. C) Percent fuel consumed by time since wetting. Curves indicate
+    best fit beta regression lines by taxonomic group. Model ANOVA table in
+    Table \ref{tab:consume_anova} and coefficients in SI Table 7. D) Flame
+    spread rate by fuel moisture content. Model ANOVA table in Table
+    \ref{tab:spreadrate_anova} and coefficients in SI Table 4. E) Percent fuel
+    consumed by moisture content. Lines indicate best fit beta regression
+    lines. Model deviance table in Table \ref{tab:consume_anova} and
+    coefficients are in SI Table 5. For plots in left column, horizontal jitter
+    was added to avoid overplotting. for flammability plots, a lack of data at
+    earlier time points for a taxon indicates no ignitions occurred at those
+    time points.}
  \label{fig:drydown}

### 2020-05-25 scripts/moisture-ms-figs-tables.R
Best i can do on space between graphs with cowplot
May need to move to patchwork package instead.

see https://stackoverflow.com/questions/41834093/decrease-margins-between-plots-when-using-cowplot

@@ -55,7 +55,7 @@ summarize(x, n=length(l))
 ## Figure 1a: dry-down
 ###############################################################################
 xbreaks <- seq(0, 144, 24)
-ybreaks <- seq(0, 700, 50)
+ybreaks <- seq(0, 700, 100)
 
 fig1a <- ggplot(mc, aes(hour, MC_dry, group=display.name, color=taxon)) +
   geom_jitter(height=0, width=0.5, size=1.2, alpha=0.7, stroke=0) +
@@ -71,7 +71,7 @@ fig1a <- ggplot(mc, aes(hour, MC_dry, group=display.name, color=taxon)) +
   theme(legend.position=c(0.79, 0.86),
         legend.title=element_blank(),
         legend.text = element_text(family=fontfamily, size=smsize-1, face="italic"),
-        plot.margin = unit(c(0, 3, 3, 5.5), "pt"),
+        plot.margin = unit(c(5, 1, 3, 5), "pt"),
         axis.title.x = element_blank())
 #        legend.key.height=unit(smsize,"pt"))
 
@@ -352,7 +352,7 @@ fig1e <- ggplot(b, aes(actualMC_dry, consum, color=taxon)) +
         legend.text = element_text(family=fontfamily, size=smsize-1, face="italic"),
         legend.title=element_blank(),
         axis.title.y = element_blank(),
-        plot.margin = unit(c(0, 5.5, 3, 0), "pt"))
+        plot.margin = unit(c(0, 5, 5, 0), "pt"))
 fig1e
 
 ###############################################################################
@@ -415,7 +415,7 @@ fig1b <- ggplot(b, aes(hour, spread, color=taxon)) +
         legend.title=element_blank(),
         axis.title.x = element_blank(),
         legend.text = element_text(family=fontfamily, size=smsize-1, face="italic"),
-        plot.margin = unit(c(0, 3, 3, 5.5), "pt"))
+        plot.margin = unit(c(0, 1, 3, 5), "pt"))
 fig1b
 ## ggsave(file.path(RESULTS, "fig4_spread_time.pdf"), plot=fig4,
 ##        width=col1, height=col1, units="cm")
@@ -510,7 +510,7 @@ fig1c <- ggplot(b, aes(hour, consum, color=taxon2)) +
         legend.spacing.y=unit(0,"cm"),
[…]

### 2020-05-25 scripts/moisture-ms-figs-tables.R
Change multiplots to patchwork library

@@ -11,7 +11,7 @@
 library(xtable)
 library(ggrepel)
 library(glmmADMB)
-library(cowplot)
+library(patchwork)
 
 RESULTS <- "../ms/figs_tables/"
 
@@ -519,8 +519,14 @@ fig1c
 ###############################################################################
 ## Fig1 combined a,b,c,d,e  ###
 ##############################################################################
-fig1 <- plot_grid(fig1a, NULL, fig1b, fig1d, fig1c, fig1e, labels=c("A", "", "B", "D", "C", "E"),
-                  ncol=2, align="hv", hjust=c(-0.5,-2,-0.5,-2, -0.5, -2))
+## fig1 <- plot_grid(fig1a, NULL, fig1b, fig1d, fig1c, fig1e, labels=c("A", "", "B", "D", "C", "E"),
+##                   ncol=2, align="hv", hjust=c(-0.5,-2,-0.5,-2, -0.5, -2))
+fig1 <- fig1a + fig1b + fig1c + plot_spacer() + fig1d + fig1e +
+  plot_layout(nrow = 3, byrow = FALSE) +
+  plot_annotation(tag_levels = 'A') &
+  theme(plot.margin = unit(c(4, 4, 4, 4), "pt"),
+        plot.tag.position = c(0, 1),
+        plot.tag = element_text(hjust = -0.5, vjust = 0.3))
 fig1
 
 ggsave(file.path(RESULTS, "fig1_dry_effects_by_taxon.pdf"), plot=fig1,

### 2020-05-25 scripts/moisture-ms-figs-tables.R
Change panel plots to use patchwork package

@@ -71,12 +71,7 @@ fig1a <- ggplot(mc, aes(hour, MC_dry, group=display.name, color=taxon)) +
   theme(legend.position=c(0.79, 0.86),
         legend.title=element_blank(),
         legend.text = element_text(family=fontfamily, size=smsize-1, face="italic"),
-        plot.margin = unit(c(5, 1, 3, 5), "pt"),
         axis.title.x = element_blank())
-#        legend.key.height=unit(smsize,"pt"))
-
-## ggsave(file.path(RESULTS, "fig1a_drydown-curves.pdf"), plot=fig1a,
-##        width=col1, height=col1, units="cm")
 
 ###############################################################################
 # SI fig 2: dry down on semi log scale
@@ -283,17 +278,15 @@ fig1d <- ggplot(b, aes(actualMC_dry, spread, color=taxon)) +
   geom_point(size=1.5, alpha=0.7, stroke=0) +
   scale_colour_manual(values=schwilkcolors) +
   ylab("Spread rate (cm/s)") +
+  scale_x_continuous(limits=c(0,80), breaks=c(0,20,40,60,80)) +
   pubtheme.nogridlines +
   stat_smooth(method="lm", se=FALSE, size=1, drop=TRUE) +
   theme(legend.position=c(.79, .86),
         legend.spacing.y=unit(0,"cm"),
         legend.text = element_text(family=fontfamily, size=smsize-1, face="italic"),
         legend.title=element_blank(),
-        plot.margin = unit(c(0, 5.5, 3, 0), "pt"),
         axis.title.x = element_blank(),
         axis.title.y = element_blank())
-## ggsave(file.path(RESULTS, "fig2_spread_actualMC.pdf"), plot=fig2a,
-##        width=col1, height=col1, unit="cm")
 
 fig1d
 spread.moist.mod <- lmer(spread ~ actualMC_dry + taxon + taxon:actualMC_dry + vpd + (1 | spcode),
@@ -342,7 +335,7 @@ newdata$consum <- predict(consume.moist.mod, newdata, type="response")*100
 fig1e <- ggplot(b, aes(actualMC_dry, consum, color=taxon)) +
   geom_jitter(height=1, width=1, size=1.5, alpha=0.7, stroke=0) +
   scale_colour_manual(values=schwilkcolors) +
-  xlab("Fuel moisture (%)") + 
[…]

### 2020-05-25 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Minor clean up edits

@@
 \running{Litter flammability}
-
-\author{Rita Quinones de Magalhaes$^{1,2}$ and Dylan W. Schwilk$^1$}
+\author{Rita Quinones de Magalh\~{a}es$^{1,2}$ and Dylan W. Schwilk$^1$}
 
@@
 
-\section*{Data and code accessibility}
+\section*{Data and code availability}
 All data and code for this project is at
-\url{https://github.com/schwilklab/trait-flam}. Checkout git tag ``JOE.R3''
+\url{https://github.com/schwilklab/trait-flam}. Checkout git tag ``JoE.1.0''
 which contains this manuscript and code needed to create the figures and

### 2020-06-23 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Edits for R4 and response to R3

@@
 flammability in moist fuels with a single study finding increasing
-nonadditivity with moisture \citep{Blauw+Wensink+etal-2015}, but this previous
-work did not include estimation of mixture effect on the dry-down process and
+nonadditivity with moisture \citep{Blauw+Wensink+etal-2015}. However, this study
+did not include estimation of mixture effect on the dry-down process and
 had an unclear relationship to fire behavior in the field. In mixed conifer
@@
 that are representative of this type of forest: \emph{Pinus jeffreyi} Grev. \&
-Balf., \emph{Pinus lambertiana} Dougl., \emph{Pinus ponderosa} Dougl. ex Laws.,
-\emph{Abies concolor} (Gord. \& Glend.) Lindl. ex Hildebr., \emph{Abies
-  magnifica} A. Murr., \emph{Calocedrus decurrens} (Torr.) Florin,
+Balf., \emph{Pinus lambertiana} Douglas, \emph{Pinus ponderosa} Douglas ex Lawson \& C. Lawson,
+\emph{Abies concolor} (Gordon \& Glend.) Lindl. ex Hildebr., \emph{Abies
+  magnifica} A. Murray, \emph{Calocedrus decurrens} (Torr.) Florin,
 \emph{Quercus kelloggii} Newb., and \emph{Sequoiadendron giganteum} (Lindl.) J.
-Buchholz. In this study, we consider ``litter'' the top layer of leaves and small
+Buchholz (identification and authorities from the Jepson Flora Project 2020). In this study, we consider ``litter'' the top layer of leaves and small
 twigs less than 1\,cm diameter that have fallen in the previous year (mostly
@@
 as our flammability functional groups and treated these as fixed effects in our
-models. The effects of fuel moisture and time since wetting on flammability
+models (Table \ref{tab:species}). The effects of fuel moisture and time since wetting on flammability
 parameters were modeled using linear models with fuel moisture or time since
@@
 of moisture content and of time since wetting.
+
+\begin{table}
+  \caption{Species included in this study and the taxonomic groups used as fixed effects in statistical models. Names and authorities according to \citet{Jepson:2020}.
+  \label{tab:species}}
+\centering
+%% \tablesize{} %% You can specify the fontsize here, e.g., \tablesize{\footnotesize}. If commented out \small will be used.
+\begin{tabular}{lll}
+  \toprule
+  Species & Authority & Taxonomic group \\
+  \midrule
+    \emph{Pinus jeffreyi}    & Grev. \& Balf.  & \emph{Pinus}\\
+    \emph{Pinus lambertiana} & Douglas          & \emph{Pinus} \\
+    \emph{Pinus ponderosa}   &  Douglas ex Lawson \& C. Lawson & \emph{Pinus} \\
+    \emph{Abies concolor}    & (Gordon \& Glend.) Lindl. ex Hildebr. & \emph{Abies} \\
+    \emph{Abies magnifica}   & A. Murray   & \emph{Abies} \\
+    \emph{Calocedrus decurrens} & (Torr.) Florin & Cuppressaceae \\
+    \emph{Sequoiadendron giganteum} & (Lindl.) J. Buchholz & Cuppressaceae \\
+    \emph{Quercus kelloggii}    & Newb. & \emph{Quercus}\\ 
+  \bottomrule
+\end{tabular}
+\end{table}
 
@@
     down curves for eight species. Lines show best-fit exponential decay curves
-    fit to moisture content on a dry mass basis. B) Flame spread rate by time
-    since wetting. Lines indicate best fit linear model by taxonomic group.
-    Model ANOVA table in Table \ref{tab:spreadrate_anova} and coefficients in
-    SI Table 6. C) Percent fuel consumed by time since wetting. Curves indicate
-    best fit beta regression lines by taxonomic group. Model ANOVA table in
-    Table \ref{tab:consume_anova} and coefficients in SI Table 7. D) Flame
-    spread rate by fuel moisture content. Model ANOVA table in Table
-    \ref{tab:spreadrate_anova} and coefficients in SI Table 4. E) Percent fuel
+    fit to moisture content on a dry mass basis. The lines for \emph{P.
+      jeffreyi} and for \emph{P. ponderosa} overlap and appear to be a single
+    line. B) Flame spread rate by time since wetting. Lines indicate best fit
+    linear model by taxonomic group (Table \ref{tab:species}). Model ANOVA
+    table is in Table \ref{tab:spreadrate_anova} and coefficients are in SI Table 6.
+    C) Percent fuel consumed by time since wetting. Curves indicate best fit
+    beta regression by taxonomic group, except that \emph{P. lambertiana}
+    is fit separately from \emph{P. jeffrey} and \emph{P. ponderosa} as
+    explained in the text (Table \ref{tab:species}). Model ANOVA table is in Table
+    \ref{tab:consume_anova} and coefficients are in SI Table 7. D) Flame spread
+    rate by fuel moisture content. Lines indicate best fit linear model by
+    taxonomic group (Table \ref{tab:species}). Model ANOVA table is in Table
+    \ref{tab:spreadrate_anova} and coefficients are in SI Table 4. E) Percent fuel
     consumed by moisture content. Lines indicate best fit beta regression
-    lines. Model deviance table in Table \ref{tab:consume_anova} and
-    coefficients are in SI Table 5. For plots in left column, horizontal jitter
-    was added to avoid overplotting. for flammability plots, a lack of data at
-    earlier time points for a taxon indicates no ignitions occurred at those
-    time points.}
+    lines. Curves indicate best fit linear model by taxonomic group (Table
+    \ref{tab:species}). Model deviance is in Table \ref{tab:consume_anova}
+    and coefficients are in SI Table 5. For plots in left column, horizontal
+    jitter was added to avoid overplotting. For flammability plots, a lack of
+    data at earlier time points for a taxon indicates no ignitions occurred at
+    those time points.}
  \label{fig:drydown}

### 2020-06-23 scripts/moisture-ms-figs-tables.R

@@ -27,6 +27,8 @@ set.seed(100)
 
 decompsp <- decomp %>% left_join(species)
 
+fig_line_size<-0.9
+
 ###############################################################################
 ## SI Fig 6: litter particle length distributions
 ###############################################################################
@@ -60,7 +62,7 @@ ybreaks <- seq(0, 700, 100)
 fig1a <- ggplot(mc, aes(hour, MC_dry, group=display.name, color=taxon)) +
   geom_jitter(height=0, width=0.5, size=1.2, alpha=0.7, stroke=0) +
   geom_smooth(method="glm",
-              method.args=list(family=gaussian(link="log")), se=FALSE, size=1) +
+              method.args=list(family=gaussian(link="log")), se=FALSE, size=fig_line_size) +
   scale_colour_manual(values=schwilkcolors) +
   #xlab("Time since wetting (hr)") +
   xlab("") +
@@ -280,7 +282,7 @@ fig1d <- ggplot(b, aes(actualMC_dry, spread, color=taxon)) +
   ylab("Spread rate (cm/s)") +
   scale_x_continuous(limits=c(0,80), breaks=c(0,20,40,60,80)) +
   pubtheme.nogridlines +
-  stat_smooth(method="lm", se=FALSE, size=1, drop=TRUE) +
+  stat_smooth(method="lm", se=FALSE, size=fig_line_size, drop=TRUE) +
   theme(legend.position=c(.79, .86),
         legend.spacing.y=unit(0,"cm"),
         legend.text = element_text(family=fontfamily, size=smsize-1, face="italic"),
@@ -339,7 +341,7 @@ fig1e <- ggplot(b, aes(actualMC_dry, consum, color=taxon)) +
   scale_y_continuous("Fuel consumed (%)", breaks=c(0, 25, 50, 75, 100), labels=c("0","25","50","75","")) +
   pubtheme.nogridlines +
   #stat_smooth(method="lm", se=FALSE, size=0.8) +
-  geom_line(data=newdata, size=1) +
+  geom_line(data=newdata, size=fig_line_size) +
   theme(legend.position=c(.78, .86),
         legend.spacing.y=unit(0,"cm"),
         legend.text = element_text(family=fontfamily, size=smsize-1, face="italic"),
@@ -397,7 +399,7 @@ print(xtable(tab.consume.moist.coef), file=file.path(RESULTS, "SI_tab5_consume_m
 
[…]

### 2020-07-01 ms/Magalhaes_Schwilk_litter_flam_moisture.tex
Finish R4
Final submission after acceptance

@@
 \nwords{6330 (main text only), 8000 (abstract, text, captions) }
-\ntables{4}
+\ntables{5}
 \nfig{3}
-\nref{57}
+\nref{58}
 
@@
       \item Moisture content is a strong determinant of forest fuel
-    flammability, however, little is known about the mechanisms by which
+    flammability. However, little is known about the mechanisms by which
     species-specific litter traits influence litter moisture absorption and
@@
 \section*{Data and code availability}
-All data and code for this project is at
+All data and code for this project is at \url{http://doi.org/10.5281/zenodo.3905213} and
 \url{https://github.com/schwilklab/trait-flam}. Checkout git tag ``JoE.1.0''
